### 2020-02-17 chapters/abstract.tex
Initial Overleaf Import

\chapter*{Abstract}

Data pattern shift detection has been researched within diverse areas and application domains and refers to the process of finding patterns in data that do not conform to expected or usual behavior. Accurate and timely detection of data pattern deviations allows for immediate measures to be taken, preventing loss of income and maintaining the companies' reputation among customers.
As such, many large companies take this into consideration and deploy data anomaly components in their systems. In this way, system administrators can proactively adjust and prevent potential service failures.

For example, Yahoo's anomaly detection system fulfills its purpose of monitoring and raising alerts on time-series events for several of Yahoo's use cases. At Microsoft, their custom, internally made anomaly detection system monitors millions of metrics coming from a variety of different internal services, enabling engineers to solve live on-site issues before they scale and become bigger ones.

Many real-time stream monitoring systems are static once deployed in a production environment. Engineers of said systems usually configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data. They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system. Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.

With this thesis, we set out to design and evaluate a system that makes use of lightweight streaming analytical methods to detect deviations in data patterns while in the presence of large volumes of high velocity, highly skewed, seasonal data.

\vspace*{10mm}\noindent
\textbf{Keywords}: real-time, streaming, lightweight, monitoring, volume, velocity, variety, seasonal

### 2020-02-17 chapters/acknows.tex
Initial Overleaf Import

\chapter*{Acknowledgements}

\vspace{10mm}
\flushleft{João Dias Conde Azevedo}

### 2020-02-17 chapters/appendix1.tex
Initial Overleaf Import

\chapter{Loren Ipsum} \label{ap1:loren}

If you are going to use a passage of Lorem Ipsum, you need to be sure
there isn't anything embarrassing hidden in the middle of text.

\section{What is \emph{Loren Ipsum}?}

\emph{\textbf{Lorem Ipsum}} is simply dummy text of the printing and
typesetting industry. Lorem Ipsum has been the industry's standard
dummy text ever since the 1500s, when an unknown printer took a galley
of type and scrambled it to make a type specimen book.
It has survived not only five centuries, but also the leap into
electronic typesetting, remaining essentially unchanged.
It was popularised in the 1960s with the release of Letraset sheets
containing Lorem Ipsum passages, and more recently with desktop
publishing software like Aldus PageMaker including versions of Lorem
Ipsum\footnote{Available at \url{http://www.lipsum.com/}}.

\section{Where does Loren come from?}

Contrary to popular belief, Lorem Ipsum is not simply random text.
It has roots in a piece of classical Latin literature from 45 BC,
making it over 2000 years old.
Richard McClintock, a Latin professor at Hampden-Sydney College in
Virginia, looked up one of the more obscure Latin words, consectetur,
from a Lorem Ipsum passage, and going through the cites of the word in
classical literature, discovered the undoubtable source.
Lorem Ipsum comes from sections 1.10.32 and 1.10.33 of ``de Finibus
Bonorum et Malorum'' (The Extremes of Good and Evil) by Cicero,
written in 45 BC.
This book is a treatise on the theory of ethics, very popular during
the Renaissance. T
he first line of Lorem Ipsum, ``Lorem ipsum dolor sit amet\ldots'',
comes from a line in section 1.10.32.

The standard chunk of Lorem Ipsum used since the 1500s is reproduced
below for those interested.
Sections 1.10.32 and 1.10.33 from ``de Finibus Bonorum et Malorum'' by
Cicero are also reproduced in their exact original form, accompanied
by English versions from the 1914 translation by H. Rackham.

\section{Why using Loren?}

It is a long established fact that a reader will be distracted by the
readable content of a page when looking at its layout.
The point of using Lorem Ipsum is that it has a more-or-less normal
distribution of letters, as opposed to using ``Content here, content
here'', making it look like readable English.
Many desktop publishing packages and web page editors now use Lorem
Ipsum as their default model text, and a search for ``lorem ipsum''
will uncover many web sites still in their infancy.
Various versions have evolved over the years, sometimes by accident,
sometimes on purpose (injected humour and the like).

\section{Where to Find Examples?}

There are many variations of passages of Lorem Ipsum available, but
the majority have suffered alteration in some form, by injected
humour, or randomised words which don't look even slightly
believable.
If you are going to use a passage of Lorem Ipsum, you need to be sure
there isn't anything embarrassing hidden in the middle of text.
All the Lorem Ipsum generators on the Internet tend to repeat
predefined chunks as necessary, making this the first true generator 
on the Internet.
It uses a dictionary of over 200 Latin words, combined with a handful
of model sentence structures, to generate Lorem Ipsum which looks
reasonable.
The generated Lorem Ipsum is therefore always free from repetition,
injected humour, or non-characteristic words etc.

### 2020-02-17 chapters/background.tex
Initial Overleaf Import

\chapter{Background}\label{chap:background} \minitoc

\section{Stream processing as a superset of batch processing} \label{sec:stream-superset}

%1st paradigm, processing of large volumes of data
%oldest processing paradigm, simplest
Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine. For instance, grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes. Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop \cite{hadoop} was one of the first widely known technologies for large-scale batch processing, created in 2005. Nearly a decade later, Apache Spark \cite{spark} was released. Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times. As more and more data became available, the need for stream processing increased. Stream processing refers to the processing of data streams. A data stream is essentially an unbounded data set. This is in contrast to batch processing which works on a finite data set. This distinction is important since some computations are impossible to compute accurately on unbounded data sets. Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements.
%On the other hand, stream processing refers to the processing of data streams. A data stream is essentially an unbounded data set. This is in contrast to batch processing which works on a finite data set. This distinction is important since some computations are impossible to compute accurately on unbounded data sets. Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements. An example of such Stream Processing Engines (SPE) is Apache Flink \cite{flink}.

% background on batch systems, micro-batch, true streaming
%Batch processing is considered the oldest between both processing paradigms. Historically speaking, Apache Hadoop \cite{hadoop} was one of the first widely known technologies for large-scale batch processing, created in 2005. Nearly a decade later, Apache Spark \cite{spark} was released. Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times. As more and more data became available, the more need for stream processing increased. % join this with previous batch intro
In order to provide a streaming solution, Apache Spark created the Spark Streaming module. Being a batch system in its core, Spark implements streaming as micro-batching. This approach handles streaming data by grouping incoming events into very small batches. Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations. However, this creates an artificial barrier that does not really exist in the streaming context. More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues. First, since we are indeed working with batches, results will never be real-time updated per event. For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events. To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed. As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.

% streaming as a proper superset of batch
Distinguishing between batch and stream processing is important to understand how a monitoring system on top of the latter would work. The Venn diagram \ref{fig:stream-superset} was taken from \cite{batch-is-a-special-case-of-streaming} and illustrates the relation of both processing techniques. A data stream is unbounded or infinite while a bounded data stream has a beginning and an end. Using these definitions, the content of an unbounded data stream contains that of a bounded data set. That is, from an infinite stream, it is possible to fix a left and right boundary in order to obtain a bounded and finite data set. This relation is visible in the \ref{fig:unbounded-bounded-superset} diagram from \cite{batch-is-a-special-case-of-streaming}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.7]{figures/streaming-subset-batch.png}
      \caption{Stream processing as a superset of batch processing}
      \label{fig:stream-superset}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.7]{figures/unbounded-superset-bounded.png}
      \caption{Unbounded data streams as a superset of bounded data sets}
      \label{fig:unbounded-bounded-superset}
    \end{center}
\end{figure}

In Section 2.2 we will cover exactly how an unbounded data set (also known as a data stream) can be converted into a bounded data set with the use of windows.

%\section{Anatomy of data analytics}
% data stream is composed by, that may exist or not
%window
%aggregation
%group_by

\section{From unbounded streams to finite data sets} \label{sec:windows}

In Section 2.1 we have defined a data stream as an unbounded data set. But what exactly are the issues of working with an infinite stream of data? Why might we need to convert it into a bounded data set and how to do so? In this Section, we will analyze such issues and present streaming windows as the solution.

% boundless aggregations no need for windows
Computations over an unbounded data set are not impossible. For example, computing the \textit{average} of an event's field for all incoming events is computationally feasible. However, is this boundless average over the stream of events useful? While in some scenarios it may be, more often than not we would like to introduce some logical context into our computations. For instance, consider the monitoring of machines spread out across multiple racks in a large data center. Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream. Is the average of all temperatures of all machines over an infinite period useful for the detection of overheating equipment? In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for actionable context, in this case identifying a likely-to-fail rack. To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end. By creating such a conceptual range, we realize that in practice we need a bounded data set. Partitioning an infinite data stream to obtain such a finite data set is made through the use of windows.

The authors of \cite{Wang-Windows-Stream-Processing} claim that \textit{"recent history is often managed using windows"} and that \textit{"windowing makes it possible to implement streaming versions of the traditionally blocking relational operators, such as streaming aggregations"}. More formally, the authors of \cite{Botan-SECRET} state that \textit{"a window W over a stream S is a finite subset of S"}.

% 2 dimensions of windows
Windows can be categorized by two dimensions: the window size \textit{W} and the sliding step size \textit{S}, both either tuple or time-based. Tuple-based windows' size is specified by the number of items the window can contain and time-based windows specify the time interval worth of incoming data a window can store. For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items. In contrast, a time-based window storing the last 5 minutes of events can contain a varying number of tuples but will always keep the events arrived in the last 5 minutes. A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events. The step size \textit{S} of a windowing technique is formally defined by \cite{Botan-SECRET} as \textit{"the distance between consecutive windows"}. This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away. As mentioned, these units can be units of time or a number of tuples.

There are three types of windows: true sliding, stepping or tumbling windows.
% true sliding: unitary steps
A true sliding window is defined by any given window size \textit{W} and a unitary step (\textit{S}=1), which can be defined as time or tuple-based. At each step, the window advances one unit, which may be one event for tuple-based steps, or one time-unit whether that is in milliseconds, seconds, days, weeks or another magnitude of time, measured based on the system clock or event time. Such a true sliding movement can be observed in Figure \ref{fig:sliding-window} with a tuple-based window of size \textit{W} of three tuples and unitary step.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/sliding.png}
      \caption[True sliding window]{True sliding window, W=3, S=1}
      \label{fig:sliding-window}
    \end{center}
\end{figure}

% stepping windows (per x units over Y window size)
% stepping: step > 1 < window size
A stepping window behaves exactly like a true sliding one but the step size \textit{S} is not unitary. In other words, a stepping window is defined by any window size \textit{W} and steps \textit{S} greater than a unit. Once again, both the window and step sizes, \textit{W} and \textit{S} respectively, can be time or tuple-based. Figure \ref{fig:stepping-window} represents a stepping tuple-based window of size \textit{W} of three and a stepping size \textit{S} of two.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/stepping.png}
      \caption[Stepping window]{Stepping window, W=3, S=2}
      \label{fig:stepping-window}
    \end{center}
\end{figure}

% tumbling: step = window size
A tumbling window, also defined by its window size \textit{W} and step size \textit{S}, is a window where W is equal to S. Hence, tumbling windows partition the stream into non-overlapping chunks with the same size. Since the intersection of different tumbling windows is always empty, each event is said to belong to exactly one tumbling window. An example of a tumbling window can be seen in Figure \ref{fig:tumbling-window} with a tuple-based window of size \textit{W} of three and an equally large step size \textit{S}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/tumbling.png}
      \caption[Tumbling window]{Tumbling window, W=S=3}
      \label{fig:tumbling-window}
    \end{center}
\end{figure}

The mathematical expression below summarizes our categorization of sliding windows.
\begin{equation} 
  Window \,\, of \,\, size \,\, W, \,\, step \,\, S =
    \begin{cases}
      true \,\, sliding & \text{if \textit{S}=1}\\
      stepping & \text{if \textit{S}>1}\\
      tumbling & \text{if \textit{S}=\textit{W}}
    \end{cases}
\end{equation}

For the scope of this Thesis, we will work under a time-based sliding window framework with a unitary step size. Despite focusing on time-based true sliding windows, our approach is just as valid for tuple sized windows and for other step values as well.

\section{Aggregations over data streams} \label{sec:aggregations}

\subsection{Sliding Window Aggregations}

Whether working with a static bounded data set or a dynamic data stream bounded by windows, there might be a need to reduce the collection of data to a single value. The operation of reducing a collection of items to a single value is known by many names: fold, reduce, accumulate or aggregate. Aggregate operations are present in most programming languages and frameworks like Apache Hadoop and its MapReduce paradigm. Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}. Hadoop's MapReduce paradigm relies on applying a map and reduce functions to the collection of data, in this order. The map task will transform each item and the reduce task will merge the collection into a single output. Reduce methods make use of a \textit{combine} function and recursively apply it. The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregation state. Doing this recursively for each item in the collection results in a single value: the aggregation value for the data set.

Aggregations over data are fundamental in overall data processing, not just stream processing. There are multiple examples of aggregations, such as computing counts, averages, maxima or minima and more. In stream processing, aggregation plays a central role, but unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion. While some aggregations can be computed on an infinite data stream, often we want to compute these over a finite data set. As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques, thus aggregating data in the sliding windows. The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs).

Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs. How would we compute the average temperature per rack in the data center in the last 5 minutes? To better understand and later on examine this aggregation, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}. Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:

\begin{verbatim}
SELECT rack_id, AVG(temperature)
FROM Temperatures
GROUP BY TUMBLE(timestamp, Time.minutes(5), rack_id)
\end{verbatim}

In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to filter out events not in this time interval. The resulting list of rows has \textit{rack\_id} as its key and a corresponding value a list of temperatures, from the past 5 minutes. Finally, we compute an \textit{AVG} aggregation for each of the racks, resulting in the average temperature per rack in the past 5 minutes.

%The distinction between the aggregation algorithm and the aggregation operation is important. For example, the operation we are interested in might be a sum, an average, a maximum, a minimum or something more complex such as a membership test or the count of distinct elements. The aggregation algorithm comes with its own data structures for storing the intermediate aggregation state and a specific algorithm to update this data structure. For example, consider the computation of an aggregation like a sum over data grouped by a true sliding window of any size. One way of computing said sum would be to, after each unitary step, compute from scratch the sum of all window values. However, a more efficient way of doing so would be to compute the total sum of the window once and then, at each step, evict the oldest element and subtract it from the total sum state and include the new event in the window while adding it up to the aggregation state. In this scenario, even though the aggregation operator is the same --- \textit{i.e.} a sum --- the algorithms differ in that one recalculates everything from scratch and the other incrementally changes the aggregation state. Hence, the operation and the algorithm differ greatly from each other.

\subsection{Aggregation properties}

Different aggregations have different properties. These properties are listed in \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}, where the authors group different aggregations according to the properties they satisfy. Table \ref{tbl:aggregations-properties} was adapted from this work and categorizes different families of operations based on a set of properties.

\begin{table}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/aggregation-operations-properties.png}
      \caption[Aggregation operations and properties]{Aggregation operations and properties. Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
      \label{tbl:aggregations-properties}
    \end{center}
\end{table}

In order to properly explain these properties it is necessary to first define the \textit{combine} and \textit{lower} functions. Both these functions are first introduced in the work of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. The \textit{combine} function merges the incoming event with the incrementally computed aggregation state. The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state. For instance, consider the use case of computing a sum aggregation over a window of data. The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event. The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.

The following are formal definitions of these properties and were summarized from \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. An aggregation operation is said to have:

\begin{enumerate}
    \item  an \textbf{\textit{invertible combine}} if a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$

    \item  an \textbf{\textit{associative combine}} if $(x \oplus y) \oplus z = x \oplus (y \oplus z)$, for all $x$ and $y$

    \item  a \textbf{\textit{commutative combine}} if $x \oplus y = y \oplus x$, for all $x$ and $y$

    \item  a \textbf{\textit{size-preserving combine}} if the aggregation state resulting from $\oplus$ always occupies the same space in memory

    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter
\end{enumerate}

These properties are important to evaluate SWAG algorithms. Different algorithms might only apply to certain families of aggregations operations. This way, the multiple SWAG algorithms presented in Chapter 3 will be bench-marked not only on their time and space complexity but also on the restrictions they impose regarding the aggregation operator properties.

\subsection{Algorithms for Sliding Window Aggregations}

% Generic SWAG algorithms - Generic algorithms for the computation of exact or approximate aggregations over a data stream in a sliding window fashion (DABA, Subtract On Evict, Recalculate From Scratch, Two Stacks)
% Aggregations - reduce a collection of values to one (sum, min, avg, cardinality of a set, membership query)
  % Exact Aggregations - an aggregation whose value is exact, in that there is no error or loss of resolution associated with the computation
  % Approximate Aggregations -  an aggregation whose value has some error margin. Usually controllable.

% \section{Concept drift}
% seasonal and evolutionary
% static ML models
% data pattern shifts -> in this context, concept drift
% data pattern shifts can be formally defined and introduced as concept drift IN THIS PARTICULAR CASE of monitoring a stream that feeds into ML models

### 2020-02-17 chapters/conclusion.tex
Initial Overleaf Import

\chapter{Conclusion} \label{chap:conclusion} \minitoc

We introduced background concepts on large scale data processing, windowing techniques applied to data streams and sliding window aggregations. We studied algorithms for change detection in data streams, pattern mining, anomaly detection and sliding window aggregation algorithms, specifically a subset of these named probabilistic data structures.

%%hypothesis validated?
The monitoring of data streams in real-time is applicable across many domains and applications. To the best of our knowledge, we claim that there has not been a real attempt at doing so in a lightweight way.

%%future work
Preliminary work shows that there exist constant space and time complexity algorithms to aggregate data, thus reducing memory usage. We intend to research more about lightweight sliding window aggregations and sliding implementations of probabilistic data structures. We also intend to search for new methods of summarizing data that do not necessarily explicitly work in a sliding window fashion. After identifying a set of algorithms that meet our requirements, the focus will be on implementing the best theoretical solution and benchmark our streaming system using data sets from the financial fraud space versus previously performed batch analysis.

%%summary of contributions
The main contribution of this Thesis will be the developed system code as well as a set of tests that measure the system's usefulness based on several parameters, such as recall, precision, memory used and latency.

### 2020-02-17 chapters/introduction.tex
Initial Overleaf Import

\chapter{Introduction} \label{chap:intro} \minitoc

\section{Context} \label{sec:context}
In the past decade, applications have become increasingly data-driven, placing data at the center of application design. Different use cases process data in different ways for different purposes. For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's clients while ensuring that sales run smoothly and that products are delivered to customers' homes. Streaming services, like YouTube and Netflix, try to guarantee that media content reaches up to millions of users simultaneously. Furthermore, this media content may have been previously recorded and stored, or may be generated on the fly. Social networks are also responsible for generating large volumes of data. A social network like Twitter, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day \cite{twitter-throughput}. \textit{Tweets} may contain text, media content or both. Cybersecurity applications are yet another example of time-sensitive, data-driven applications. These need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the protected system.

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized in terms of its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former case is best known as stream processing, while the latter as batch processing. E-commerce applications have to process endless streams of credit card transactions. Live streaming services need to distribute the media content created by the \textit{streamer} (content creator) and deliver it across multiple users. Social networks process and analyze hundreds of posts per second to find out trending keywords across their network and suggest them to users. Computer network monitor systems want to detect intruders and revoke them system access before they have time to corrupt the system or steal private information. All of the previously given use cases rely on real-time processing of data for near real-time actions. Hence, between batch and stream processing, the latter fits these use cases better.

As mentioned, the processing of data can be divided into two main categories: batch and stream processing. However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, stream processing demand is increasing. Data stream processing refers to the handling of huge volumes of high variety of data generated at high-velocity from multiple sources in real-time. Unlike batch processing on static data sets, where assumptions on the underlying data distribution can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery}. Furthermore, in the streaming paradigm, the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}. \textit{Recency} is measured under different scales for different use cases. For instance, considering the monitoring of geological data to forecast possible natural disasters such as earthquakes versus the monitoring of a computer network for intruder detection. In the former case, after the forecast of a future earthquake, governmental authorities need a couple of days to launch an evacuation plan and keep the population safe. However, in the intruder detection scenario, the decision of removing access to a user must be done as soon as possible, preferably in a few seconds. Hence, while in the first scenario information retrieved within a day would still be recent, in the second scenario, information is considered recent and relevant if delivered within seconds (or even milliseconds).

Both in batch and stream processing, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements. Operations applied to data sets that produce a single result are commonly known as aggregate operations. In stream processing, we apply these aggregations to data streams to obtain valuable information. However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques. A thorough analysis to stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last $10^6$ events, respectively.

Consider the case is the detection of credit card fraud on the data stream of all the transactions made on Amazon by every US citizen on a single day. In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one. Such a system would have to process all the incoming transactions and decide in a fraction of a second if it is a fraudulent one or not. It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone. Thus justifying the need for autonomous real-time data stream monitoring systems.

Many market solutions monitor streams of credit card transactions with the intent of fraud detection. In the context of this Thesis, we will work side by side with Feedzai. Feedzai offers fraud prevention solutions for the previously presented use case of monitoring an infinite stream of credit card transactions and detecting fraud. We will work side by side with them, taking advantage of the multiple real-life data sets from the financial fraud space they can provide.

Feedzai's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained before deployment making assumptions on the statistical distributions of future incoming data. Thus their model performance is linked to the maintenance of these assumptions. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.} Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.} new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, the performance decay caused by data pattern shifts is also known as concept drift. %UNCOMMENT WHEN BACKGROUND SECTION IS DONE: Such a concept will be further explained in Section \ref{sec:concept-drift}.

\section{Objectives} \label{sec:objectives}
Feedzai's data pattern shift challenge is not unique to them nor their field. Many analytical systems are static once deployed and are configured \textit{a priori} under the assumption that future data flowing through the system will roughly follow the same distribution as previously seen data.
The problem is that data pattern shifts will lead to poor performance by the previously configured system. Monitoring the data patterns themselves and alerting for changes in the underlying distribution would allow users to reconfigure the system before its performance declines.

The objective of this Thesis is to build a lightweight real-time system that detects these shifts in data patterns in systems that handle high volumes of high velocity, highly skewed and seasonal data.

\section{Motivation} \label{sec:motivation}
For any organization, the quality of their provided services is paramount. Therefore, it is common for companies to have systems monitoring the former, testing their usage rate, availability and performance, just to name a few. The motivation behind this Thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in performance.

We want to build a lightweight enough solution that can be integrated into existing workflows. Ensuring constant time and space complexity means that the system will remain usable for the ever-growing volumes of data. Since such a monitoring system is not mission-critical --- \textit{i.e.} the provided services don't rely on it --- the operational cost associated with it must be kept to a minimum. Having a real-time system monitoring other real-time systems consuming as many resources as the latter would be far too costly.

\section{Hypothesis} \label{sec:hypothesis}
The question this Thesis aims to answer is whether sliding window aggregations and probabilistic data structures can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data with a low memory footprint.

\section{Hypothesis Validation} \label{sec:validation}
The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
\begin{itemize}
    \item high precision ratio --- \textit{i.e.} at least 80\%
    \item high recall ratio --- \textit{i.e.} at least 80\%
    \item low latency --- \textit{i.e.} constant time complexity, \textit{O(1)}
    \item memory efficiency --- \textit{i.e.} constant memory complexity, \textit{O(1)}
\end{itemize}

Experiments will be made using multiple real data sets from the financial fraud space and compared to existing batch analysis results.

\section{Document structure}
This section merely sums up how each of the following Chapters contributes to this Thesis.

In Chapter \ref{chap:background} of this document we brief the reader with the necessary background and relevant concepts needed to understand the following Chapters.

In Chapter \ref{chap:sota} we examine the state of the art in the area, analyzing the most recent work done and classifying it into subgroups.

In Chapter \ref{chap:approach} we define the scope of this project, the research questions we intend to answer, our experimental methodology and our future planning.

In Chapter \ref{chap:conclusion} we summarize our contributions.

### 2020-02-17 chapters/problem.tex
Initial Overleaf Import

\chapter{Problem and proposed solution} \label{chap:approach} \minitoc

\section{Scope}
The focus of this Thesis is the development of a lightweight real-time data monitoring system working in the presence of high volumes of high velocity, high variety, highly skewed and seasonal data. The resultant system should detect and report data pattern shifts in an unsupervised fashion.

Essentially, the system will summarize incoming data with sliding window aggregations and probabilistic data structures in a sliding window fashion and compare current feature values with an initial given static configuration. Any big enough deviations between these two states should be reported as a data pattern shift.

Furthermore, the resulting system shall strive to achieve high precision and recall ratios as well as low latencies with a low memory footprint.

\section{Research Questions}
With this Thesis, we intend to build a system that monitors real-time data streams in a lightweight fashion. As such, in the end, to determine if the system is successful in doing so or not, the following research questions must be answered:

\begin{enumerate}[start=1, label={\textbf{RQ\arabic*:}}]
    \item Can the system accurately detect data pattern shifts --- \textit{i.e.} does it have \textbf{high precision ratio}?

    \item Can the system detect most of the data pattern shifts --- \textit{i.e.} does it have \textbf{high recall ratio}?

    \item Can the system detect data pattern shifts in real-time --- \textit{i.e.} does it work under extremely \textbf{low latencies}?

    \item Can the system detect data pattern shifts in a lightweight fashion --- \textit{i.e.} does it have a \textbf{low memory footprint}?

\end{enumerate}

\section{Experimental Methodology}

In order to validate or reject our proposed hypothesis, a set of tests will be carried out to determine the system's usefulness. The data sets used will belong to the financial fraud space. The output of the system will be compared to the results of an existing batch analysis tool, for the same data sets, both provided by Feedzai.

To test the stream monitoring system we will need a streaming engine. As such, either a custom-built streaming engine or an existing one --- \textit{e.g.} like Apache Flink \cite{Rabl-Apache-Flink} --- will be used.

In brief, it is desired that our in real-time system monitoring report is as close to the batch analysis in terms of accuracy as possible.

\section{Planning}
To accomplish the proposed experiments, the following steps are required:

\begin{enumerate}[start=1, label={\textbf{T\arabic*:}}]
    \item \textbf{Research more state of the art aggregations} - search for more memory efficient sliding window aggregations to use in the final system.

    \item \textbf{Data set and batch results analysis} - explore chosen data sets and the resulting batch analysis made on them.

    \item \textbf{Stream Processing Engine (SPE) setup} - setup the stream processing engine chosen --- \textit{e.g.} Apache Flink \cite{Rabl-Apache-Flink} --- or build a custom very basic one.

    \item \textbf{Implementation of chosen algorithms} - implement the selected algorithms in an efficient programming language --- \textit{e.g.} Java or C++ --- in a sliding window fashion and integrate it with the SPE.

    \item \textbf{Test the implementation with the chosen data sets} - compare the batch analysis with our streaming analysis and measure precision and recall ratios as well as latency and memory usage.

    \item \textbf{Thesis writing} - the writing of the Thesis document will be an on-going process but ultimately finished last.
\end{enumerate}

In Image \ref{fig:tasks} we present the expected start and end dates for the main tasks presented before.

\begin{figure}
    \begin{center}
      \includegraphics[scale=0.75]{figures/gantt-labels.png}
      \caption{Tasks and deadlines}
      \label{fig:tasks}
    \end{center}
\end{figure}

Furthermore, we built a Gantt Chart \ref{fig:gantt} that illustrates the expected time frame of completion for each task, in weeks.

\begin{figure}
    \begin{center}
      \includegraphics[scale=0.50]{figures/gantt-thesis.png}
      \caption[Thesis research plan]{Gantt Chart of Thesis research plan}
      \label{fig:gantt}
    \end{center}
\end{figure}

### 2020-02-17 chapters/resumo.tex
Initial Overleaf Import

\chapter*{Resumo}

A deteção de mudanças de padrões nos dados tem sido investigada nas mais diversas áreas e aplicações e diz respeito ao processo de encontrar padrões nos dados que não correspondam ao esperado ou usual. A deteção atempada e correta destas mudanças de padrões permite que medidas imediatas sejam tomadas, prevenindo perdas de rendimento e mantendo a reputação da empresa entre os seus clientes. Como tal e tendo isto em conta, muitas empresas constroem sistemas de deteção de anomalias. Desta maneira, administradores do sistema são capazes de proactivamente ajustar e prevenir potenciais falhas nos seus serviços.

Por exemplo, o sistema de deteção de anomalias da Yahoo serve o propósito de monitorizar e alertar \textit{streams} de eventos temporais em múltiplos casos de uso da Yahoo. Na Microsoft, o seu sistema interno de deteção de anomalias monitoriza milhões de métricas provenientes de diferentes serviçoes internos, permitindo que os engenheiros os resolvam em tempo real antes que estes problemas adquiram proporções maiores.

Muitos sistemas de monitorização em tempo real são estáticos quando implementados num ambiente de produção. Os engenheiros responsáveis por tais sistemas configuram-nos assumindo que os dados que futuramente fluirão pelo sistema se assemelham aos dados observados até ao momento. Eles fazem-no conscientemente e o melhor que sabem com as ferramentas de que dispoem, mantendo sempre em mente que no futuro possivelmente terão de reconfigurar o sistema. Assim, apesar da configuração inicial do sistema ser a que melhor se adequa, ao longo do tempo, devido a desvios nos padrões de dados, o desempenho do sistema estático incialmente implementado decai gradualmente.

Desta forma, pretendemos avaliar o uso de métodos de análise de \textit{streams} eficientes em termos de memória --- \textit{i.e} métodos que usam pouca memóra e calculam os resultados esperados em tempo real --- com o intuito de detetar desvios nos padrões de dados na presença de grandes volumes de dados com padrões sazonais que fluem pelo sistema a grande velocidade.

\vspace*{10mm}\noindent
\textbf{Keywords}: tempo real, streaming, leves, monitorização, volume, velocidade, variedade, sazonalidade

### 2020-02-17 chapters/sota.tex
Initial Overleaf Import

\chapter{State of the Art} \label{chap:sota} \minitoc

In this Chapter, we will discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time computations over true sliding windows using resource-lightweight approaches on streaming systems. We present a wide analysis of several categories of algorithms that seemed like viable options to build the desired system.

\section{Detecting Change in Data Streams} \label{sec:change-detect}

For static data sets, it is reasonable to assume that the data was generated by a fixed process, for example, the data is a sample from a static distribution. But a data stream has necessarily a temporal dimension, and the underlying process that generates the data stream can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}. It is this change in the underlying distribution that we want to detect and report as a data pattern shift.

\subsection{A two window paradigm algorithm for change detection}
\label{subsec:2-window}

The authors of \cite{Kifer-Detecting-Change} propose an algorithm for change detection in data streams. The only assumption made is that data points are generated independently. The authors propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions. Thus the change detection algorithm works on a two-window paradigm.

Figure \ref{fig:change-detection-2-windows} contains the pseudocode presented by the authors. The proposed method uses two windows, a reference \textit{W\textsubscript{1}} and a sliding one \textit{W\textsubscript{2}}. The authors used tuple-based windows of size \textit{k} and the sliding window was a true sliding one, as described in Section \ref{sec:windows}. The reference window \textit{W\textsubscript{2}} works as a \textit{baseline} and contains the first \textit{k} points of the stream that occurred after the last detected change.

The algorithm begins by filling both windows with the first \textit{k} tuples. Then it slides window \textit{W\textsubscript{2}} (by one, since it is a true sliding window). It does this \textit{k} times. In between slides, the algorithm checks for change by applying the \textit{d} function, which measures the discrepancy between the two windows contents, \textit{W\textsubscript{1}} and \textit{W\textsubscript{2}}. Whenever this discrepancy is above a certain threshold $\alpha$, the algorithm reports a change in the data stream. Then it repeats the whole process, re-initializing both windows with the next \textit{k} items and proceeding as already described.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/2-windows-change-pseudocode.png}
      \caption{Two window method for data stream change detection}
      \label{fig:change-detection-2-windows}
    \end{center}
\end{figure}

This algorithm reduces change detection in data streams to testing whether two samples, the \textit{reference} and the \textit{sliding window} contents, are generated by different distributions. Consequently, the authors study the case of detecting differences in distribution between two samples. This is the purpose of the \textit{d} function. The authors evaluate many statistical tests as possible implementations of this \textit{d} function that must truly quantify an intuitive notion of change.

Experiments were made using the following statistical tests as the \textit{d} function: Wilcoxon signed-rank, $\phi$\textsubscript{A} and $\Xi$\textsubscript{A}. The Wilcoxon test is well known but the $\phi$\textsubscript{A} and the $\Xi$\textsubscript{A} statistical tests were defined by the authors. Since computing these three statistical tests for our windows contents takes identical time and space, we consider that a thorough analysis of each one is out of scope for this Thesis, as it does not provide relevant information.

The authors conclude that no test is best in all of the experiments done. However, the $\phi$\textsubscript{A} and $\Xi$\textsubscript{A} statistics vastly outperformed the others in some tests and did not perform that much worse in other test cases.

Diving deeper in the algorithm presented in \ref{fig:change-detection-2-windows} we now analyze the computation of the \textit{d} function value. In the authors' work, efficient computation of the previously mentioned statistical tests --- \textit{i.e.} the value of function \textit{d} --- is further explored and an algorithm for such is presented. First, the \textit{KS structure} is presented. The formal definition given by the authors is shown in Figure \ref{fig:ks-structure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/ks-structure.png}
      \caption{Formal definition of the KS structure}
      \label{fig:ks-structure}
    \end{center}
\end{figure}

The KS structure is built from the reference \textit{W\textsubscript{1}} and sliding \textit{W\textsubscript{2}} windows. Consider the pair \textit{(\textit{W\textsubscript{1}}, \textit{W\textsubscript{2}})} where the size of both windows is \textit{k}. Creating the KS structure \textit{Z} of size \textit{2k} is done by joining the elements from \textit{W\textsubscript{1}} and \textit{W\textsubscript{2}}. As stated in \ref{fig:ks-structure}, each element of \textit{Z} will have a value and a weight. The values will be the actual elements joined. The weights of elements coming from \textit{W\textsubscript{1}} will be \textit{-1/k} while the weights of elements coming from \textit{W\textsubscript{2}} will be \textit{1/k}. The final KS structure \textit{Z} must be sorted in increasing order by value. The KS structure \textit{Z} can be maintained throughout running time by making use of a balanced tree, such as a B-tree, in \textit{O(log(2k))} time. By making use of the KS structure, all of the mentioned statistics can be computed in \textit{O(2k)} time.

\subsubsection{Time complexity analysis}
%k to initialize and when change is detected
%k * d
%d 
 %nlog(2k) to initialize tree and when change is detected
 %log(2k) to maintain the tree
The algorithm begins by collecting the first \textit{k} tuples. Both windows will have size \textit{k}. This takes \textit{O(k)} time but is a task that will only occur on initialization and on change detection, upon which the window contents must reset, as described in \ref{fig:change-detection-2-windows}. For each of the next \textit{k} iterations the function \textit{d} will be computed and compared to the threshold $\alpha$. As previously mentioned, computing the value of \textit{d} can be done in \textit{O(log(2k))} time per iteration. However, the initial time cost of initializing the balanced tree is of \textit{O(k log(2k))}.

Initialization is not frequent. It occurs once at the beginning of execution and once per change detection upon which the algorithm resets the windows' contents. To provide an accurate time complexity, we take into account the probability \textit{P} of change occurring and being detected leading to a reset. Assuming \textit{P} is within $[0, 1]$, we have a time complexity of \textit{O(P(k + k log(2k)))} per reset --- \textit{i.e.} the probability of a reset occurring times \textit{O(k)} to initialize the windows' contents plus \textit{O(k log(2k))} to build the balanced tree.

In runtime, the algorithm will perform \textit{k} iterations where it computes the \textit{d} function value, taking \textit{O(log(2k))} per iteration, amounting to a time complexity of \textit{O(k log(2k)}.

Putting both initialization and runtime complexities we have a total time complexity of \textit{O(P(k + k log(2k)) + k log(2k))}.

\subsubsection{Space complexity analysis}

The algorithm presented makes use of two \textit{k} sized windows and one balanced tree of size \textit{k} as well. This represents a space complexity of \textit{O(k+k+k)} or \textit{O(3k)}.

\subsubsection{Applicability to our Hypothesis}

The algorithm for change detection presented in Subsection \ref{subsec:2-window} has time complexity of \textit{O(P(k + k log(2k)) + k log(2k))}. and space complexity of \textit{O(3k)}, with \textit{k} as a window size constant.

In this Thesis, we want to detect and alert data pattern shifts so detecting change in data streams is directly applicable to our use case. However, our focus is on a lightweight solution and this algorithm has a linear space complexity, so it does not fit our scope as a possible solution.

\section{Pattern Mining}

Taking on a new approach, in this Section, we explore pattern mining algorithms rather than algorithms that detect change in data streams. The idea behind using pattern mining algorithms being that if we can maintain a list of frequent patterns we would be able to detect when said patterns change --- \textit{i.e.} the list changes. Change in the frequent mined patterns would be associated with a data pattern shift and immediately reported.

\subsection{Mining Frequent Patterns: the FP-Stream structure}

The authors of \cite{Giannella-Mining-Frequent-Patterns} developed a data structure named \textit{FP-Stream} --- that summarizes frequent data patterns -- and algorithms to build and incrementally maintain it.

We agree with the author's statement that it is \textit{"unrealistic to hold all streaming data in the limited main memory"}. With this in mind, they divide patterns into three categories: frequent, sub-frequent and infrequent patterns. The focus of their work is the mining and maintenance of frequent and sub-frequent patterns since the latter might become frequent later. Thus, infrequent patterns are discarded, using less memory.

The FP-Stream structure consists of a frequent pattern tree (FP-Tree \cite{Han-FP-tree}) with tilted-time windows in each of the nodes.

The authors are not very explicit on the definition of a tilted-time window. They claim we are \textit{"often interested in recent changes at a fine granularity, but long term changes at a coarse granularity"} and that tilted-time windows fit such use case. Figure \ref{fig:tilted-time-window} shows such a window and was withdrawn from the paper. In it, we see the 4 quarters of the last hour, then the last 24 hours and finally 31 days. They claim that \textit{"one can compute frequent itemsets in the last hour with the precision of quarter of an hour, the last day with the precision of hour, and so on, until the whole month"}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1]{figures/tilted-time-window.png}
      \caption{Tilted-time window}
      \label{fig:tilted-time-window}
    \end{center}
\end{figure}

An FP-Tree \cite{Han-FP-tree} as shown in Figure \ref{fig:fptree}, is a tree representation of frequent patterns. Each node in the frequent pattern tree represents a pattern and its frequency (\textit{support} column in the Figure), recorded in the node. The same tree construction algorithm from \cite{Han-FP-tree} is used in the FP-Stream algorithm.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=1]{figures/fptree.png}
      \caption{FP-Tree}
      \label{fig:fptree}
    \end{center}
\end{figure}

The authors propose using only one frequent pattern tree, where at each node, the frequency for each tilted-time window is maintained. Figure \ref{fig:fpstream} shows the FP-Stream structure, as an example of a frequent pattern tree with tilted-time windows embedded in each node.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/fp-stream.png}
      \caption{FP-Stream}
      \label{fig:fpstream}
    \end{center}
\end{figure}

The algorithm for constructing and maintaining the FP-Stream structure is given as a high-level list of instructions. The algorithm groups incoming streaming data into batches. Initialization is done only when the first batch is complete. As the transactions for the first batch arrived, the frequencies for all items are computed. Once all transactions for the first batch have arrived, the batch is scanned to create an FP-tree structure, pruning all infrequent items -- \textit{i.e.} below a certain frequency threshold. Finally, an FP-Stream structure is created by mining the frequent items from the FP-Tree.

The incremental update of the FP-Stream as data arrives is described on a very high level and does not add relevant information to understand time and space complexity without knowledge of the FP-Tree algorithm, on which it heavily relies. Further research will be done to perform a more thorough analysis of this algorithm, starting with the analysis of \cite{Han-FP-tree}.

\subsubsection{Applicability to our Hypothesis}

No definitive conclusions can be made without the full time and space complexity analysis. However, the maintenance of a frequent pattern tree and tilted-time windows for each of the nodes might reveal to memory intensive, discarding this as a valid solution.

\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm}

The authors of \cite{Yang-Neighbor-Based-Pattern-Detection} propose a method for incremental detection of neighbor based patterns, namely density-based clusters and distance-based outliers, specifically for sliding windows scenarios. Figure \ref{fig:plot-cluster-outlier} is an example of the previous two mentioned neighbor based patterns. Formal definitions for both distance-based outliers \ref{fig:distance-outlier-def} and density-based clusters \ref{fig:density-cluster-def} are provided.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/density-distance-based.png}
      \caption[Clustering and outlier identification plot]{Two density-based clusters and one distance-based outlier determined by neighbour-based pattern detection algorithms}
      \label{fig:plot-cluster-outlier}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/distance-based-outlier-def.png}
      \caption{Formal definition of distance-based outliers}
      \label{fig:distance-outlier-def}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/density-based-cluster-def.png}
      \caption{Formal definition of density-based clusters}
      \label{fig:density-cluster-def}
    \end{center}
\end{figure}

Density-based clusters label data points as \textit{core points}, \textit{edge points} or \textit{noise}. These classifications are explained in \ref{fig:density-cluster-def}. Data points are classified based on the number of neighbors they have within a range $\theta\textsuperscript{range}$, measured versus a $\theta\textsuperscript{count}$. In brief, if a data point has more than $\theta\textsuperscript{count}$ neighbors within $\theta\textsuperscript{range}$ it is classified as a \textit{core point}. If it does not but it is a neighbor of a core point, then it is an \textit{edge point}. Otherwise, it is considered \textit{noise}.

The authors make use of the \textit{“predictability"} property of the expiration of existing objects. In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point in the window on arrival --- \textit{i.e.} the future windows it will belong to. This leads to the notion of \textit{predicted views}. Given the objects in a current window, they predict the pattern structures that will persist in subsequent windows by considering the objects in the current window and abstract these predicted pattern structures into \textit{“predicted views"} of each future window. This technique allows the efficient maintenance of neighborship counts by pre-handling the eventual expiring effect of new data points.

In the paper, the authors first propose the Abstract-C and Abstract-M algorithms. Ultimately, they combine all the desirable properties of each algorithm into Extra-N.

\subsubsection{Abstract-C}
Abstract-C was the first proposed algorithm. The main idea is to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has rather than the actual list of neighbors. The main advantage of this approach is the reduced memory footprint since only a count is stored --- a single integer --- instead of a list of pointers --- to all neighbors.

When a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one --- since \textit{p\textsubscript{i}} no longer exists. However, we are unable to reach \textit{p\textsubscript{i}}'s neighbors since in Abstract-C only a count of neighbors is stored --- rather than pointers to each of the neighbors.

The solution to this problem is exploiting the previously presented \textit{predictability} property. The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbours. The authors introduce the concept of \textit{"lifetime neighbour counts"} or \textit{lt\_cnt}. The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbour counts} --- \textit{i.e.} the number of \textit{predicted neighbours} \textit{p\textsubscript{i}} has in any future window where he exists. More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s current neighbours that are known to survive in the corresponding future window. Furthermore, the length of \textit{p\textsubscript{i}.lt\_cnt} is kept equal to \textit{p\textsubscript{i}.lifespan}, decreasing by one after each window slide, removing the left most entry. The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbours, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}. Applying the \textit{predictability} property makes it possible to compute the lifespan of all points. Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}. This implies that at \textit{W\textsubscript{i}}, \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\}.

According to the definition given in \ref{fig:distance-outlier-def}, the lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute distance-based outliers: for each data point \textit{p\textsubscript{i}} check if the first (current) element of \textit{p\textsubscript{i}.lt\_cnt} is less than  \textit{$\theta\textsuperscript{fra}$} $\times$ \textit{N} to decide whether a data point is an outlier or not, with \textit{N} as the number of data points. Similarly, the core objects for the density-based clusters can be computed by comparing the first element of \textit{p\textsubscript{i}.lt\_cnt} with \textit{$\theta$\textsuperscript{count}}. However, the \textit{lt\_cnt} array does not carry enough information to generate the density-based clusters. Despite being able  to find all core points in the window, we can not group them into clusters. In order to compute that, Abstract-C analyzes each core point and the surrounding $\theta\textsuperscript{range}$ area in the window to reconstruct the clusters.

In conclusion, Abstract-C achieves linear memory consumption in the number of data points in the window. However, since it reconstructs the clusters for each core object in the window, it has \textit{O(CP)} time complexity, with \textit{CP} as the total number of core points. Hence, its performance largely depends on  \textit{CP}, which can vary from 0 -- no core points --- to the total window size --- all core points.

\subsubsection{Abstract-M}
Abstract-M is proposed as an Abstract-C enhancement by introducing the notion of \textit{cluster memberships}, achieved by attributing a \textit{clusterID} to each data point.

The main challenge of this approach is discounting the effect of expired data points. The removal of any data point on any cluster may result in the total break of the cluster into multiple smaller ones, which may persist or degrade to noise. Furthermore, when a cluster is split, the newly formed clusters must be attributed with new \textit{clusterID}s --- \textit{i.e.} every single data point in those clusters must be relabeled with the new \textit{clusterID}.

Once again, the solution to handle expiring data points relies on the predictability property. Knowing the data points that exist in each window means we can also know the set of future clusters and points that belong. This is included in the \textit{predicted views} concept previously mentioned. In brief, the \textit{predicted clusters} are created using such predicted views for each of the future windows.

Addition of new data points may cause three changes to the clusters: \textit{birth} of a new cluster, \textit{expansion} of an existing one and \textit{merge} of multiple existing clusters. When adding a new data point, mark it with a new \textit{clusterID} in case of a \textit{birth} or in an \textit{expansion} case mark it with the \textit{clusterID} of the cluster it belongs to. Handling the \textit{merge} of multiple clusters is done by using a tree structure where two or more sibling \textit{clusterID}s are in fact part of a bigger cluster with the \textit{clusterID} stored in their parent. For example, if cluster \textit{ID1} and cluster \textit{ID2} are merged into a new cluster \textit{ID3} by the arrival of a new data point, they are inserted as children of the node containing \textit{ID3}.

New data points may join existing neighborhoods of existing data points and may promote them to core points by making the size of their neighborhood greater or equal to \textit{$\theta$\textsuperscript{count}}. Once such promotion happens, the promoted core point needs to communicate to its neighbors its new role, since noise in its neighborhood becomes edge points and is given a \textit{clusterID}. Since there are no connections from the promoted data point to its neighbors, Abstract-M analyzes each of the promoted core points and the $\theta\textsuperscript{range}$ area around them to find the points requiring state update.

In conclusion, Abstract-M remains linear in memory regarding the number of window data points. However, since the number of promoted core points tends to be smaller and is always less or equal to the number of core points, this is an improvement on Abstract-C.

\subsubsection{Extra-N}

Abstract-M still performs range queries (area with $\theta\textsuperscript{range}$ diameter) for each of the promoted core points. To improve upon this, Extra-N combines the neighborship maintenance mechanisms proposed in Abstract-C and Abstract-M.

The authors state that different information is required from different classes of data points (\textit{core}, \textit{edge} or \textit{noise}). For example, maintaining a list of points to neighbors is required for a non-core point while neighbor counts are sufficient for core points. More precisely, Extra-N marks each data point \textit{p\textsubscript{i}} with a \textit{clusterID} in all windows where \textit{p\textsubscript{i}} is predicted to be a core point while keeping the exact list of neighbors (pointers) of \textit{p\textsubscript{i}} for all windows where he is predicted to be noise or an edge point. This way, the Extra-N algorithm stores the amount of information needed to compute both distance-based outliers and density-based clusters according to their definitions, \ref{fig:distance-outlier-def} and \ref{fig:density-cluster-def} respectively.

\subsubsection{Time complexity analysis}
Extra-N only performs range queries for incoming data points. This means that for each new data point \textit{p\textsubscript{i}}, Extra-N will search the area of diameter $\theta\textsuperscript{range}$ around it and update on it neighbors. In conclusion, the time complexity of Extra-N is \textit{O(n)} where \textit{n} is the number of new data points.

\subsubsection{Space complexity analysis}
In summary, Extra-N keeps the memory consumption linear to the number of data points in the window without performing unnecessary range queries.

\subsubsection{Applicability to our Hypothesis}
While a linear complexity fits most of the scenarios, it does not fit ours since we aim to build a low memory footprint solution, ideally with space complexity of \textit{O(1)}. Hence, this algorithm will be of no use to us.

\section{Anomaly Detection}
Anomaly detection refers to the process of finding anomalies in data. An anomaly relates to a point in time where the behavior of the system does not match the expected one. Hence, under this definition, an anomaly does not necessarily imply a problem.

With the study of anomaly detection algorithms, we intend to search for algorithms that identify data pattern shifts as anomalies. If any of the approaches searched for is not computationally intensive it will be considered valid for the final solution.

\subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}

In \cite{Ahmad-HTM}, the authors present an anomaly detection technique based on an on-line algorithm called Hierarchical Temporal Memory (HTM). HTM has a few properties that are desirable for data stream anomaly detection scenarios. First, it is an unsupervised algorithm that requires no labels. This property comes in handy when monitoring a data stream in real-time, where labels for incoming data will likely be absent. Additionally, HTM adapts to noisy domains. In other words, random variations in the underlying distribution of incoming data that do not persist throughout time will be ignored. Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.} changes in magnitudes of data or unusual timing for particular patterns, respectively.

HTM is also an online learning algorithm, meaning that it adapts to changing statistics of the underlying data stream. In other words, if there is a sudden random increase in a specific data point attribute, the algorithm will produce an alert. However, if that initially thought to be a random spike becomes frequent then the algorithm adapts and assimilates this as "a new reality". Hence, it stops classifying it as an anomaly. Unfortunately, this property is not desirable in our system. Our stream monitoring system must report changes relatively to an initial static configuration and should not accept a new distribution as the new standard.

\subsubsection{Applicability to our Hypothesis}
Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again. Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made. In conclusion, we do not deem HTM applicable to our solution.

\subsection{DeepAnT}

DeepAnT \cite{Munir-DeepAnT} is a deep learning-based approach for the detection of anomalies in time series data, that uses unlabeled data to capture and learn the data distribution of the stream used to forecast the normal behavior of the time series data. It is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a data streaming scenario where real-time decisions have to be made before labels are effectively-known. It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.

DeepAnT employs a Convolutional Neural Network (CNN) as its \textit{time series predictor} module. Before monitoring a data stream, the CNN must be trained. The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN. Additionally, the data set used to train the model thus not require labels and may up to 5\% of the data set may correspond to anomalous data.

\subsubsection{Applicability to our Hypothesis}

Similar to the HTM algorithm presented, DeepAnT adapts to the underlying distribution of the data stream. In other words, if the data patterns shift from the initial configuration but stabilize, DeepAnT will accept it as it is. As mentioned, our goal is to build a monitoring system that alerts data pattern shifts continuously, until they match the initial configuration again. Furthermore, DeepAnT uses deep learning which is actually a subset of machine learning and just as inexplicable. In summary, we do not deem DeepAnT usable for our use case.

\iffalse

\section{Sliding Window Aggregations}
Aggregations and Sliding Window Aggregations (SWAGs) have been introduced in Section \ref{sec:aggregations}. In this Section, we analyze algorithms that allow for sliding window aggregations utilizing a set of different aggregation operators. In other words, different algorithms work with different aggregation categories based on their properties. The properties of different categories of aggregations were defined in Section \ref{sec:aggregations} and summarized in Table \ref{tbl:operations-properties}.

\subsection{DABA}
\subsubsection{Time complexity analysis}
\subsubsection{Space complexity analysis}
\subsubsection{Applicability to our Hypothesis}

\fi

\section{Probabilistic Data Structures} \label{sec:pds}
Probabilistic data structures use hash functions to compactly represent a set of items while approximately answering queries with fixed error bounds. This way, these structures are capable of processing huge volumes of data. They require a single pass through the data, which is appropriate for a streaming scenario. Probabilistic data structures have constant space and time complexity \cite{Singh-PDS-BIGD} making them candidate methods to build our real-time and low memory footprint system.

\subsection{Membership Queries and the Bloom Filter}
A Bloom filter \cite{BLOOM-BLOOMFILTER} is a probabilistic data structure that allows membership queries on a set of elements. Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure, Bloom filters do not always give a certain answer. \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is. In other words, a bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

A Bloom filter is an array of bits of size \textit{m}. A Bloom Filter makes use of \textit{k} hash functions. The choice of the \textit{m} and \textit{k} constants will determine the false positive rate.

Initially, all bits are set to 0. Adding an element is done by determining which positions should be set to 1. To that end, the new element is fed into each hash function that maps the element to an array position. The resultant \textit{k} outputs are used as the \textit{k} positions of the array to be set to 1.

Performing a membership query --- \textit{i.e.} testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions. If \textit{el} was already in the set then all \textit{k} positions should be 1. If any position contains a 0 then the element is \textit{definitely not in set}, as shown in Figure \ref{fig:bloom-filter} from \cite{bloom-filter-wikipedia}. If all positions are 1 then the element is \textit{possibly in the set}.

The reason a Bloom Filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements. For example, hash functions \textit{k\textsubscript{1}}, \textit{k\textsubscript{2}} and \textit{k\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ and an element \textit{el2} to positions $[4,5,6]$. This way, bits in positions $[1,2,3,4,5,6]$ are set to 1. Given a new element \textit{el3} not yet seen, hash functions \textit{k\textsubscript{1}}, \textit{k\textsubscript{2}} and \textit{k\textsubscript{3}} may map it to $[1,3,5]$ and all of these positions are already 1 because of the previous elements. Thus the Bloom filter returns \textit{"possibly in set"} because all bits are set to 1 when in reality \textit{el3} was not in the set. This is considered a \textit{false positive}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/bloom-filter.png}
      \caption[Bloom filter membership test]{Bloom filter membership test of element 'W'. Finding one '0' indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

An improvement to this structure was defined in \cite{Kirsch-Better-Bloom} where the authors attempt to reduce the overhead of bloom filters operations such as adding a new element or answering a query by focusing on the hash functions. The proposed method is based on hashing literature. It has been shown that multiple hash functions can be created from the combination of only two hash functions. This means that a new hash function \textit{g} can be produced based on  two other independent and uniform existing hash functions, \textit{h1} and \textit{h2}. The new hash function will map elements to a universe ranging from \textit{0} to \textit{p-1} and will be defined as \textit{$g(x) = h1(x) + ih2(x) mod p$}. Since the only nontrivial computations performed by a Bloom filter are the evaluations of pseudo-random hash functions, any reduction in the required number of pseudo-random hash functions yields a reduction in the time required to add an element or query a Bloom filter.

Bloom filters are not the only approximate aggregators that use hash functions as a basis of their algorithm. All of the studied probabilistic data structures do so. Therefore, the technique proposed is beneficial in HyperLogLogs, Count-Min Sketches and Bloom filters.

\subsubsection{Time complexity analysis}
Considering a Bloom filter with \textit{m} bits and \textit{k} hash functions, insertion and search will both take \textit{O(k)} time because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1. Note the time complexity does not at all depend on the number of elements in it and that \textit{k} will be a rather small constant (magnitude of $10^1$). Hence, the Bloom filter is said to have constant time complexity (\textit{O(1)}).

\subsubsection{Space complexity analysis}
Considering a Bloom filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity. Similarly to the time complexity analysis, we point out that \textit{m} will be constant and that each of the array elements occupies 1 bit. Thus the Bloom filter has a constant space complexity (\textit{O(1)}).

\subsubsection{Applicability to our Hypothesis}
Bloom filters exhibit constant space and time complexity making it great methods to test our hypothesis. Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}.

\subsection{Item Frequency and the Count-Min Sketch}

A Count–Min Sketch (CMS) is a probabilistic data structure that gives an estimate of the frequency of each element in the data set. The estimate is a minimum count. The real frequency of an element will never be lower than the given estimate, hence the name \textit{count min}. The CMS is first introduced in \cite{Cormode-CMS}.

A CMS is represented by a two-dimensional array (or matrix) with width \textit{w} and depth \textit{d}. A CMS will make use of \textit{d} hash functions --- \textit{i.e.} each associated with one row of the matrix. Each hash function will map elements to a position between 1 and \textit{w} --- \textit{i.e.} the column of the two-dimensional array.

Initially, all of the matrix positions are 0. When adding an element, for each row \textit{i} of the \textit{d} rows in the matrix, hash the element using that row's hash function to obtain \textit{j}. Lastly, for all obtained pairs of \textit{i} and \textit{j}, increment matrix cells \textit{(i, j)} value by 1, as seen in Figure \ref{fig:cms}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/cms.png}
      \caption[Count-Min Sketch update]{CMS mapping element \textit{el} using \textit{d} hash functions to determine the column position for each row and increment it}
      \label{fig:cms}
    \end{center}
\end{figure}

Retrieving the minimum frequency of an element \textit{el} can be done by taking the minimum value of all row counts for \textit{el}. Mathematically, the frequency estimate is given by
\[ min \{matrix[i][\textit{h\textsubscript{i}}(el)]\} \]
for each row \textit{i} of the total \textit{d} rows.

%The CMS is not always correct because...

\subsubsection{Time complexity analysis}
When adding an element or querying the frequency of one, the Count-Min Sketch loops through each row and applies a constant in time hash function, proceeding to a position update or retrieval, respectively. Hence, the time complexity is of \textit{O(d)}. Given \textit{d} is a constant, the CMS has constant time complexity --- \textit{O(1)}.

\subsubsection{Space complexity analysis}
The Count-Min Sketch relies on a two-dimensional matrix as its only data structure. The memory used will correspond to the \textit{wd} counts, hence \textit{O(wd)} space complexity. Since both \textit{w} and \textit{d} are constants, the CMS has constant space complexity --- \textit{O(1)}.

\subsubsection{Applicability to our Hypothesis}
Count-Min Sketches are constant in both time and space. Similarly to Bloom filters, this makes them ideal candidates for our solution.

\subsection{Cardinality Estimation and the HyperLogLog}
In the paper \cite{Flajolet-PCA}, a class of probabilistic counting algorithms is introduced to estimate the number of distinct elements based on \textit{"bit pattern observables"} in the binary representation of the hashed values --- \textit{i.e.} patterns related to the number and positions of 0's and 1's in a binary string.

\subsubsection{LogLog}
Later on, the same authors propose the LogLog algorithm \cite{Flajolet-LogLog}. In LogLog, the \textit{bit pattern observable} recorded from each item's hash value is the position of the leftmost 1-bit. The authors claim it is related to the total number of distinct elements in the data set.

\subsubsection{HyperLogLog}
A couple of years later, the same authors propose HyperLogLog (HLL) in \cite{Flajolet-HLL}. HLL solves the problem of counting distinct elements in a data set, also known as the cardinality of the data set \cite{Flajolet-HLL}. HLL gives an approximation of the cardinality of the data set.

In HyperLogLog, an item is hashed into a binary string. Based on the first \textit{n} bits, the string is passed on to one of the \textit{$2^n$ buckets}. After discarding the first \textit{n} bits, the number of trailing 0's plus 1 is counted. In other words, the position of the leftmost 1-bit is recorded, like in LogLog. This is what is stored in each bucket: the maximum value of the position of the leftmost 1-bit of all seen items so far.

%TODO \textbf{HYPERLOGLOG HOW IT WORKS IMAGE}

Finally, the estimate for the number of distinct elements will be $2^p$, where \textit{p} is the harmonic mean of all bucket values holding the maximum leftmost 1-bit position value. This computation of a harmonic mean over \textit{m} buckets is called stochastic averaging and is the main difference between HyperLogLog and LogLog. Given the nature of streaming systems, it is also important to notice that this algorithm is suitable for a distributed system since each of the buckets can be updated independently.

\subsubsection{Time complexity analysis}
Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, count the leftmost 1-bit position and update the bucket value if need be. Estimating the cardinality of the set amounts to finding the maximum value \textit{p} between all buckets and computing $2^p$. Hence, both inserting a new element and estimating the cardinality take constant time --- \textit{O(1)}.

\subsubsection{Space complexity analysis}
Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in  \cite{Flajolet-HLL}: 
\say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}

HyperLogLog uses \textit{m} buckets that store a single integer value. This gives it complexity of \textit{O(m)} where \textit{m} is a small constant. Hence, HyperLogLog has constant space complexity --- \textit{O(1)}.

\subsubsection{Applicability to our Hypothesis}
Similar to the previously studied probabilistic data structures, HyperLogLog (HLL) has constant space and time complexity while giving an approximation of a set's cardinality. Hence, HLLs are considered valuable methods for our work.

\iffalse

\section{Sliding Window Aggregation with Probabilistic Data Structures}
In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space complexity. These structures aggregate an endless stream of data. However, when working under a sliding window context there is the need not only to insert new events but to evict old ones. Probabilistic data structures aggregate all incoming data but have no way of removing elements.

In this Section, we explore and analyze sliding window implementations of the previously discussed probabilistic data structures.

\subsection{Sliding HyperLogLog}
HyperLogLog  (HLL) will only be useful in our system if it can be implemented efficiently in a sliding window fashion. That is precisely what \cite{Chabchoub-Sliding-HLL} presents us with. In this paper, the original HLL algorithm from \cite{Flajolet-HLL} is adapted to a sliding window scenario with the use of slightly more memory. More importantly, the estimate is as accurate as in the original HLL algorithm.

It consumes additional space compared to the original algorithm of at most $5\textit{m}ln(\textit{n}/\textit{m})$ bytes, with \textit{m} being the number of buckets and \textit{n} the number of distinct elements actually flowing through the stream. As explained by the authors, storing additional context is required so that the algorithm can work in a sliding window fashion. Thus they propose maintaining a shortlist of packets. A packet consists of a pair \textit{(ti, Ri)}, where the \textit{ti} is the arrival time of the packet and \textit{Ri} is the position of the leftmost 1-bit in the binary representation of the hashed value associated with that packet. The idea is to store only packets that are a possible maximum over a future window. Hence, this list is called \textit{List of Future Possible Maxima (LFPM)}. Furthermore, they propose efficient methods for the insertion and eviction of new and old elements from the window, respectively. In this sliding version, one LFPM is kept per bucket. For that reason, estimating the set cardinality is similar to the original HyperLogLog: compute the harmonic mean for all buckets. The value of each bucket is the highest \textit{Ri} for that bucket. Hence, the extra required memory in the Sliding HyperLogLog algorithm presented is given by the size of the \textit{m} lists of future possible maxima.

\subsubsection{Time complexity analysis}

\subsubsection{Space complexity analysis}

\subsubsection{Applicability to our Hypothesis}

\fi

%\textbf{image for sliding hyperloglog}

### 2020-02-17 chapters/background.tex
Aggregations,  Aggregations properties, Sliding Window Aggregations (SWAGs), SWAG algorithms

@@
 
-\subsection{Sliding Window Aggregations}
+An aggregation --- also known as a fold, reduce or accumulate --- is an operation that reduces a collection of values to a single one.
+Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality).
+An aggregation may be exact or approximate.
+Exact aggregations have an exact value or state in that there is no error or loss of resolution associated.
+Approximate aggregations have an error margin to the associated aggregation value.
+This error is usually controllable by parameter optimization.
 
-Whether working with a static bounded data set or a dynamic data stream bounded by windows, there might be a need to reduce the collection of data to a single value.
-The operation of reducing a collection of items to a single value is known by many names: fold, reduce, accumulate or aggregate.
-Aggregate operations are present in most programming languages and frameworks like Apache Hadoop and its MapReduce paradigm.
+Aggregate operations are present in most programming languages and frameworks.
 Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}.
-Hadoop's MapReduce paradigm relies on applying a map and reduce functions to the collection of data, in this order.
-The map task will transform each item and the reduce task will merge the collection into a single output.
+Hadoop's MapReduce paradigm consists of first applying a map function to the collection of data and then a reduce one.
+The map task will transform each item as desired and the reduce task will merge the collection into a single value.
 Reduce methods make use of a \textit{combine} function and recursively apply it.
@@
 
-Aggregations over data are fundamental in overall data processing, not just stream processing.
-There are multiple examples of aggregations, such as computing counts, averages, maxima or minima and more.
+\subsection{Sliding Window Aggregations}
+
 In stream processing, aggregation plays a central role, but unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion.
 While some aggregations can be computed on an infinite data stream, often we want to compute these over a finite data set.
-As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques, thus aggregating data in the sliding windows.
+As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques, thus being able to aggregate the data in the sliding window.
 The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs).
-
 Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs.
@@
 In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to filter out events not in this time interval.
-The resulting list of rows has \textit{rack\_id} as its key and a corresponding value a list of temperatures, from the past 5 minutes.
-Finally, we compute an \textit{AVG} aggregation for each of the racks, resulting in the average temperature per rack in the past 5 minutes.
-
-%The distinction between the aggregation algorithm and the aggregation operation is important.
-For example, the operation we are interested in might be a sum, an average, a maximum, a minimum or something more complex such as a membership test or the count of distinct elements.
-The aggregation algorithm comes with its own data structures for storing the intermediate aggregation state and a specific algorithm to update this data structure.
-For example, consider the computation of an aggregation like a sum over data grouped by a true sliding window of any size.
-One way of computing said sum would be to, after each unitary step, compute from scratch the sum of all window values.
-However, a more efficient way of doing so would be to compute the total sum of the window once and then, at each step, evict the oldest element and subtract it from the total sum state and include the new event in the window while adding it up to the aggregation state.
-In this scenario, even though the aggregation operator is the same --- \textit{i.e.} a sum --- the algorithms differ in that one recalculates everything from scratch and the other incrementally changes the aggregation state.
-Hence, the operation and the algorithm differ greatly from each other.
+The resulting list of rows has \textit{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes.
+Finally, for each rack and its list of temperatures, we compute an \textit{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.
 
@@
 In order to properly explain these properties it is necessary to first define the \textit{combine} and \textit{lower} functions.
-Both these functions are first introduced in the work of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
 The \textit{combine} function merges the incoming event with the incrementally computed aggregation state.
@@
 
-    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter
+    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
 \end{enumerate}
 
-These properties are important to evaluate SWAG algorithms.
-Different algorithms might only apply to certain families of aggregations operations.
-This way, the multiple SWAG algorithms presented in Chapter 3 will be bench-marked not only on their time and space complexity but also on the restrictions they impose regarding the aggregation operator properties.
+These properties are used to evaluate SWAG algorithms.
+Different algorithms might only apply to certain families of aggregations.
+This way, the multiple SWAG algorithms presented throughout this Thesis will be bench-marked not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.
 
-\subsection{Algorithms for Sliding Window Aggregations}
+\subsection{Sliding Window Aggregation Algorithms}
 
@@
 
+A sliding window aggregation (SWAG) algorithm allows the computation of an exact or approximate aggregation over a sliding window.
+A SWAG algorithm may restrict the set of possible aggregations based on their properties.
+Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state.
+Despite storing information using different structures, every SWAG algorithm must implement three methods:
+
+\begin{itemize}
+    \item \textit{insert} - inserts new value in the window
+    \item \textit{evict} - evicts all values that no longer fit in the window
+    \item \textit{query} - returns the aggregation state/value
+\end{itemize}
+
+Tangwongsan et al. present two of the most basic SWAG algorithms: \textit{Recalculate From Scratch (RFS)} and \textit{Subtract On Evict (SOE)}.
+
+\subsubsection{Recalculate From Scratch}
+Recalculate From Scratch (RFS) is the simplest and most general SWAG algorithm.
+As the name suggests, this algorithm recomputes the aggregation over the entire window for each insertion and eviction.
+For instance, the computation of a sum over a sliding window using RFS means that upon each insertion or eviction all elements are summed.
+
+This approach does not restrain the aggregation in terms of properties: it works for every family of aggregations.
+However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing.
+
+\subsubsection{Subtract On Evict}
+Subtract On Evict (SOE) only works for aggregations whose  \textit{combine} function is invertible.
+That is, as defined in Table \ref{tbl:aggregations-properties}, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$.
+
+SOE inserts new elements into the window and combines the previous aggregation value with the new element.
+On eviction, SOE applies the inverse of the combine function, $\ominus$, to expire the evicted element from the aggregation state.
+
+With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity.
+However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable.
+
+\subsection{Summary}
+This subsection adds nothing new to the previous ones. However, it concisely summarizes the most relevant previously discussed concepts.
+
+An aggregation is an operation that reduces a collection of values to a single one.
+Examples of such are sums, minima, averages or finding the number of distinct elements in a set (cardinality).
+An aggregation may be exact or approximate.
+Exact aggregations have an exact value or state in that there is no error or loss of resolution associated.
+Approximate aggregations have an error margin to the associated aggregation value.
+This error is usually controllable by parameter optimization.
+
+Sliding Window Aggregation algorithms allow the computation of exact or approximate aggregations over a data stream in a sliding window fashion.
+Each algorithm must implement three methods: \textit{insert}, \textit{evict} and \textit{query}.
+A SWAG algorithm may restrict the set of possible aggregations based on their properties.
+
 % \section{Concept drift}

### 2020-02-20 chapters/background.tex
Bloom Filter

@@
 An aggregation may be exact or approximate.
-Exact aggregations have an exact value or state in that there is no error or loss of resolution associated.
-Approximate aggregations have an error margin to the associated aggregation value.
-This error is usually controllable by parameter optimization.
+Exact aggregations compute a value from a sequence of elements, where no error or loss of resolution exists.
+On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters.
 
-Aggregate operations are present in most programming languages and frameworks.
+Aggregation is present in most programming languages and frameworks.
 Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}.
-Hadoop's MapReduce paradigm consists of first applying a map function to the collection of data and then a reduce one.
-The map task will transform each item as desired and the reduce task will merge the collection into a single value.
+Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one.
+\textit{Map} task will transform each item as desired and the \textit{reduce} task will merge the collection into a single value.
 Reduce methods make use of a \textit{combine} function and recursively apply it.
-The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregation state.
-Doing this recursively for each item in the collection results in a single value: the aggregation value for the data set.
-
-\subsection{Sliding Window Aggregations}
-
-In stream processing, aggregation plays a central role, but unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion.
-While some aggregations can be computed on an infinite data stream, often we want to compute these over a finite data set.
-As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques, thus being able to aggregate the data in the sliding window.
-The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs).
-Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs.
-How would we compute the average temperature per rack in the data center in the last 5 minutes?
-To better understand and later on examine this aggregation, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}.
-Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:
-
-\begin{verbatim}
-SELECT rack_id, AVG(temperature)
-FROM Temperatures
-GROUP BY TUMBLE(timestamp, Time.minutes(5), rack_id)
-\end{verbatim}
-
-In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to filter out events not in this time interval.
-The resulting list of rows has \textit{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes.
-Finally, for each rack and its list of temperatures, we compute an \textit{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.
+The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state.
+Doing this recursively for each item in the collection results in a single value: the aggregate value for the data set.
 
@@
 These properties are listed in \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}, where the authors group different aggregations according to the properties they satisfy.
-Table \ref{tbl:aggregations-properties} was adapted from this work and categorizes different families of operations based on a set of properties.
+Table \ref{tbl:aggregations-properties} was adapted from this work and categorizes different families of operations based on said properties.
 
@@
       \includegraphics[scale=0.5]{figures/aggregation-operations-properties.png}
-      \caption[Aggregation operations and properties]{Aggregation operations and properties.
+      \caption[Aggregation properties]{Aggregation properties.
 Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
@@
 
-In order to properly explain these properties it is necessary to first define the \textit{combine} and \textit{lower} functions.
-The \textit{combine} function merges the incoming event with the incrementally computed aggregation state.
+To understand these properties, we must be familiar with the \textit{combine} and \textit{lower} functions.
+The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state.
 The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state.
@@
 
-\subsection{Sliding Window Aggregation Algorithms}
+\subsection{Sliding Window Aggregations}
 
-% Generic SWAG algorithms - Generic algorithms for the computation of exact or approximate aggregations over a data stream in a sliding window fashion (DABA, Subtract On Evict, Recalculate From Scratch, Two Stacks)
-% Aggregations - reduce a collection of values to one (sum, min, avg, cardinality of a set, membership query)
-  % Exact Aggregations - an aggregation whose value is exact, in that there is no error or loss of resolution associated with the computation
-  % Approximate Aggregations -  an aggregation whose value has some error margin. Usually controllable.
+Unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion.
+While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one.
+As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques.
+In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it.
+The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs).
+Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs.
+How would we compute the average temperature per rack in the data center in the last 5 minutes?
+To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}.
+Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:
+
+\begin{verbatim}
+SELECT rack_id, AVG(temperature)
+FROM Temperatures
+GROUP BY TUMBLE(timestamp, Time.minutes(5), rack_id)
+\end{verbatim}
+
+In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to evict events not in this time interval.
+The resulting list of rows has \textit{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes.
+Finally, for each rack and its list of temperatures, we compute an \textit{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.
+
+\subsection{Sliding Window Aggregation Algorithms} \label{sec:back-swag-algs}
 
 A sliding window aggregation (SWAG) algorithm allows the computation of an exact or approximate aggregation over a sliding window.
-A SWAG algorithm may restrict the set of possible aggregations based on their properties.
+A SWAG algorithm may restrict the set of possible to compute aggregations based on their properties.
 Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state.
@@
 \subsubsection{Subtract On Evict}
-Subtract On Evict (SOE) only works for aggregations whose  \textit{combine} function is invertible.
+Subtract On Evict (SOE) only works for aggregations whose  \textit{combine} $\oplus$ function is invertible.
 That is, as defined in Table \ref{tbl:aggregations-properties}, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$.
@@
 
-\subsection{Summary}
-This subsection adds nothing new to the previous ones. However, it concisely summarizes the most relevant previously discussed concepts.
-
-An aggregation is an operation that reduces a collection of values to a single one.
-Examples of such are sums, minima, averages or finding the number of distinct elements in a set (cardinality).
-An aggregation may be exact or approximate.
-Exact aggregations have an exact value or state in that there is no error or loss of resolution associated.
-Approximate aggregations have an error margin to the associated aggregation value.
-This error is usually controllable by parameter optimization.
-
-Sliding Window Aggregation algorithms allow the computation of exact or approximate aggregations over a data stream in a sliding window fashion.
-Each algorithm must implement three methods: \textit{insert}, \textit{evict} and \textit{query}.
-A SWAG algorithm may restrict the set of possible aggregations based on their properties.
-
 % \section{Concept drift}

### 2020-02-20 chapters/sota.tex
Bloom Filter

@@
 
-\subsubsection{Time complexity analysis}
+\subsubsection*{Time complexity analysis}
 %k to initialize and when change is detected
@@
 
-\subsubsection{Space complexity analysis}
+\subsubsection*{Space complexity analysis}
 
@@
 
-\subsubsection{Applicability to our Hypothesis}
+\subsubsection*{Applicability to our Hypothesis}
 
@@
 
-\subsubsection{Time complexity analysis}
+\subsubsection*{Time complexity analysis}
 Extra-N only performs range queries for incoming data points.
@@
 
-\subsubsection{Space complexity analysis}
+\subsubsection*{Space complexity analysis}
 In summary, Extra-N keeps the memory consumption linear to the number of data points in the window without performing unnecessary range queries.
 
-\subsubsection{Applicability to our Hypothesis}
+\subsubsection*{Applicability to our Hypothesis}
 While a linear complexity fits most of the scenarios, it does not fit ours since we aim to build a low memory footprint solution, ideally with space complexity of \textit{O(1)}.
@@
 
-\subsubsection{Applicability to our Hypothesis}
+\subsubsection*{Applicability to our Hypothesis}
 Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again.
@@
 
-\iffalse
-
-\section{Sliding Window Aggregations}
-Aggregations and Sliding Window Aggregations (SWAGs) have been introduced in Section \ref{sec:aggregations}.
-In this Section, we analyze algorithms that allow for sliding window aggregations utilizing a set of different aggregation operators.
-In other words, different algorithms work with different aggregation categories based on their properties.
-The properties of different categories of aggregations were defined in Section \ref{sec:aggregations} and summarized in Table \ref{tbl:operations-properties}.
-
+\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
+As discussed in Section \ref{sec:back-swag-algs} ...
+Besides RFS and SOE other methods were proposed like DABA...
 \subsection{DABA}
-\subsubsection{Time complexity analysis}
-\subsubsection{Space complexity analysis}
-\subsubsection{Applicability to our Hypothesis}
-
-\fi
+\subsubsection*{Time complexity analysis}
+\subsubsection*{Space complexity analysis}
+\subsubsection*{Applicability to our Hypothesis}
 
 \section{Probabilistic Data Structures} \label{sec:pds}
-Probabilistic data structures use hash functions to compactly represent a set of items while approximately answering queries with fixed error bounds.
-This way, these structures are capable of processing huge volumes of data.
+In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs).
+However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods.
+These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds.
+Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.
+
+Probabilistic data structures use hash functions to compactly represent a set of items.
 They require a single pass through the data, which is appropriate for a streaming scenario.
-Probabilistic data structures have constant space and time complexity \cite{Singh-PDS-BIGD} making them candidate methods to build our real-time and low memory footprint system.
+Probabilistic data structures have constant space and time complexity \cite{Singh-PDS-BIGD} making them ideal aggregators to build our real-time and low memory footprint system.
+
+% FOR EACH SKETCH DO:
+% approx aggregation to compute
+% data structures
+% algorithm for insertion
+% algorithm for query
+% error bound related with parameters
+% time complexity analysis
+% space complexity analysis
+% Applicability to our Hypothesis
 
 \subsection{Membership Queries and the Bloom Filter}
-A Bloom filter \cite{BLOOM-BLOOMFILTER} is a probabilistic data structure that allows membership queries on a set of elements.
+% approx aggregation to compute
+A Bloom filter \cite{BLOOM-BLOOMFILTER} aggregator allows the computation of a membership query approximate aggregation on a set of elements.
 Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?".
-Being a probabilistic data structure, Bloom filters do not always give a certain answer.
+Being a probabilistic data structure and approximate aggregator, Bloom filters' answer has an error associated.
 \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is.
@@
 
-A Bloom filter is an array of bits of size \textit{m}.
-A Bloom Filter makes use of \textit{k} hash functions.
-The choice of the \textit{m} and \textit{k} constants will determine the false positive rate.
+% data structures
+\subsubsection*{Data structure}
+A Bloom filter is an array of bits of size \textit{m}, initially all set to 0.
+Figure \ref{fig:initial-bloom-filter} represents a Bloom filter with a bit array of size \textit{m}=7 with all bits initially set to 0.
 
-Initially, all bits are set to 0.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=1.1]{figures/initial-bloom.png}
+      \caption[Bloom filter bit array initial state]{Bloom filter bit array initial state of \textit{m}=7}
+      \label{fig:initial-bloom-filter}
+    \end{center}
+\end{figure}
+
+% algorithm for insertion
+\subsubsection*{Element Insertion}
+Element insertion and query answering make use of \textit{k} hash functions, each mapping elements to a position in the bit array.
 Adding an element is done by determining which positions should be set to 1.
@@
 
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/insert-bloom.png}
+      \caption[Bloom filter insertion]{Insertion of element \textit{el} into a Bloom filter with \textit{m}=7 and \textit{k}=3}
+      \label{fig:insertion-bloom-filter}
+    \end{center}
+\end{figure}
+
+% algorithm for query
+\subsubsection*{Membership Query}
 Performing a membership query --- \textit{i.e.} testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions.
-If \textit{el} was already in the set then all \textit{k} positions should be 1.
-If any position contains a 0 then the element is \textit{definitely not in set}, as shown in Figure \ref{fig:bloom-filter} from \cite{bloom-filter-wikipedia}.
-If all positions are 1 then the element is \textit{possibly in the set}.
+If \textit{el} was already in the set, all \textit{k} positions should be 1.
+If any position contains a 0 then the element is \textit{definitely not in set}, as shown in Figure \ref{fig:bloom-filter}.
 
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/query-bloom.png}
+      \caption[Bloom filter membership test]{Bloom filter membership test of element \textit{el}. Finding at least one '0' indicates that it is not in the set.}
+      \label{fig:bloom-filter}
+    \end{center}
+\end{figure}
+
+However, if all positions are 1 then the element is said to be \textit{possibly in the set}.
 The reason a Bloom Filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements.
-For example, hash functions \textit{k\textsubscript{1}}, \textit{k\textsubscript{2}} and \textit{k\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ and an element \textit{el2} to positions $[4,5,6]$.
+For example, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ as illustrated in Figure \ref{fig:bloom-filter-fp1} and an element \textit{el2} to positions $[4,5,6]$ like in Figure \ref{fig:bloom-filter-fp2}.
 This way, bits in positions $[1,2,3,4,5,6]$ are set to 1.
-Given a new element \textit{el3} not yet seen, hash functions \textit{k\textsubscript{1}}, \textit{k\textsubscript{2}} and \textit{k\textsubscript{3}} may map it to $[1,3,5]$ and all of these positions are already 1 because of the previous elements.
-Thus the Bloom filter returns \textit{"possibly in set"} because all bits are set to 1 when in reality \textit{el3} was not in the set.
+When querying about the membership of a new element \textit{el3}, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map it to positions $[1,3,5]$ as exemplified in Figure \ref{fig:bloom-filter-fp3}.
+Since all of these positions are 1 because of the previous elements inserted, the Bloom filter returns \textit{"possibly in set"} when in reality \textit{el3} was not in the set.
 This is considered a \textit{false positive}.
@@
     \begin{center}
-      \includegraphics[scale=0.8]{figures/bloom-filter.png}
-      \caption[Bloom filter membership test]{Bloom filter membership test of element 'W'. Finding one '0' indicates that it is not in the set.}
-      \label{fig:bloom-filter}
+      \includegraphics[scale=0.4]{figures/fp-bloom-1.png}
+      \caption[Bloom filter insertion]{Insertion of \textit{el1} in the Bloom filter}
+      \label{fig:bloom-filter-fp1}
     \end{center}
@@
 
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/fp-bloom-2.png}
+      \caption[Bloom filter insertion]{Insertion of \textit{el2} in the Bloom filter}
+      \label{fig:bloom-filter-fp2}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/fp-bloom-3.png}
+      \caption[Bloom filter false positive]{Membership query of \textit{el3} and false positive result}
+      \label{fig:bloom-filter-fp3}
+    \end{center}
+\end{figure}
+
+\subsubsection*{False Positive rate}
+% error bound related with parameters
+The false positive rate is a function of the bit array size \textit{m}, the number of hash functions \textit{k} and the number of elements in the set \textit{n}.
+It is assumed that the hash functions produce an independent and random index value for each element.
+Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula.
+
+The probability of one arbitrary bit being set to 1 is
+\begin{equation}
+    \frac{1}{m}
+\end{equation}
+because there are \textit{m} bits. Likewise, the probability of one bit not being set to 1 is
+\begin{equation}
+    1-\frac{1}{m}
+\end{equation}
+
+Mullin states the probability that an arbitrary bit is not set to 1 after one element insertion and corresponding \textit{k} bit updates is
+\begin{equation}
+    (1-\frac{1}{m})^\textit{k}
+\end{equation}
+
+For \textit{n} insertions, the probability a bit is not set to 1 is
+\begin{equation}
+    (1-\frac{1}{m})^\textit{kn}
+\end{equation}
+
+Therefore, the probability that an arbitrary bit is set to 1 after \textit{n} insertions is 
+\begin{equation}
+    1-(1-\frac{1}{m})^\textit{kn}
+\end{equation}
+
+Hence, the probability of false positive of a membership query for a new element for \textit{k} hash functions is given by
+\begin{equation}
+    (1-(1-\frac{1}{m})^\textit{kn})^\textit{k}
+\end{equation}
+
+This false positive rate can be approximated by 
+\begin{equation}
+    (1-e^\textit{-kn/m})^\textit{k}
+\end{equation}
+
+\subsubsection*{Optimal number of hash functions}
+We want to choose the number of hash functions \textit{k} in order to minimize the false positive rate.
+In order to do so, we must first allocate \textit{m} bits for the Bloom filter, choosing \textit{m} based on available memory.
+The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by
+\begin{equation}
+    \frac{m}{n}ln(2)
+\end{equation}
+
+\iffalse
+%TODO hash improvement probably better declared outside of bloom
 An improvement to this structure was defined in \cite{Kirsch-Better-Bloom} where the authors attempt to reduce the overhead of bloom filters operations such as adding a new element or answering a query by focusing on the hash functions.
@@
 Therefore, the technique proposed is beneficial in HyperLogLogs, Count-Min Sketches and Bloom filters.
+\fi
 
-\subsubsection{Time complexity analysis}
-Considering a Bloom filter with \textit{m} bits and \textit{k} hash functions, insertion and search will both take \textit{O(k)} time because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1.
-Note the time complexity does not at all depend on the number of elements in it and that \textit{k} will be a rather small constant (magnitude of $10^1$).
-Hence, the Bloom filter is said to have constant time complexity (\textit{O(1)}).
+\subsubsection*{Time complexity analysis}
+Considering a Bloom filter with \textit{m} bits and \textit{k} hash functions, insertion has \textit{O(k)} time complexity because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1.
+Similarly, query answering has \textit{O(k)} time complexity because it needs only to run the input through the \textit{k} hash functions and check if all bits are set to 1.
+If so, it returns \textit{possibly in set}.
+Otherwise it returns \textit{definitely not in set}.
 
-\subsubsection{Space complexity analysis}
+Note the time complexity does not at all depend on the number of elements in it and that \textit{k} will be a rather small constant. Hence, the Bloom filter is said to have constant time complexity.
+
+\subsubsection*{Space complexity analysis}
 Considering a Bloom filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity.
 Similarly to the time complexity analysis, we point out that \textit{m} will be constant and that each of the array elements occupies 1 bit.
-Thus the Bloom filter has a constant space complexity (\textit{O(1)}).
+Thus the Bloom filter has a constant space complexity.
 
-\subsubsection{Applicability to our Hypothesis}
+\subsubsection*{Applicability to our Hypothesis}
 Bloom filters exhibit constant space and time complexity making it great methods to test our hypothesis.
@@
 
-\subsubsection{Time complexity analysis}
+\subsubsection*{Time complexity analysis}
 When adding an element or querying the frequency of one, the Count-Min Sketch loops through each row and applies a constant in time hash function, proceeding to a position update or retrieval, respectively.
@@
 
-\subsubsection{Space complexity analysis}
+\subsubsection*{Space complexity analysis}
 The Count-Min Sketch relies on a two-dimensional matrix as its only data structure.
@@
 
-\subsubsection{Applicability to our Hypothesis}
+\subsubsection*{Applicability to our Hypothesis}
 Count-Min Sketches are constant in both time and space. Similarly to Bloom filters, this makes them ideal candidates for our solution.
@@
 
-\subsubsection{Time complexity analysis}
+\subsubsection*{Time complexity analysis}
 Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, count the leftmost 1-bit position and update the bucket value if need be.
@@
 
-\subsubsection{Space complexity analysis}
+\subsubsection*{Space complexity analysis}
 Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in  \cite{Flajolet-HLL}: 
@@
 
-\subsubsection{Applicability to our Hypothesis}
+\subsubsection*{Applicability to our Hypothesis}
 Similar to the previously studied probabilistic data structures, HyperLogLog (HLL) has constant space and time complexity while giving an approximation of a set's cardinality.
@@
 
+% PDS have no way to expire old data (must for SWAG)
+% they could be implemented using one of the previously seen SWAG algorithms but lose the space efficiency
+% present sliding versions as solution
+
 \iffalse
-
 \section{Sliding Window Aggregation with Probabilistic Data Structures}

### 2020-02-20 chapters/sota.tex
Count-Min Sketch

@@
 
-% FOR EACH SKETCH DO:
+\subsection{Membership Queries and the Bloom filter}
 % approx aggregation to compute
-% data structures
-% algorithm for insertion
-% algorithm for query
-% error bound related with parameters
-% time complexity analysis
-% space complexity analysis
-% Applicability to our Hypothesis
-
-\subsection{Membership Queries and the Bloom Filter}
-% approx aggregation to compute
-A Bloom filter \cite{BLOOM-BLOOMFILTER} aggregator allows the computation of a membership query approximate aggregation on a set of elements.
+A Bloom filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements.
 Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?".
-Being a probabilistic data structure and approximate aggregator, Bloom filters' answer has an error associated.
+Being a probabilistic data structure and approximate aggregator, Bloom filters' answers have an error associated.
 \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is.
-In other words, a bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set.
+In other words, a Bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set.
 Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.
@@
 A Bloom filter is an array of bits of size \textit{m}, initially all set to 0.
-Figure \ref{fig:initial-bloom-filter} represents a Bloom filter with a bit array of size \textit{m}=7 with all bits initially set to 0.
+Figure \ref{fig:initial-bloom-filter} represents a Bloom filter with a bit array of size \textit{m}=7 in its initial state.
 
@@
 The resultant \textit{k} outputs are used as the \textit{k} positions of the array to be set to 1.
+Figure \ref{fig:insertion-bloom-filter} shows the insertion of element \textit{el} in the Bloom filter, making use of three hash functions (\textit{k}=3), \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}}.
+The outputs of the hash functions correspond to the array indexes to set to 1.
 
@@
 
-However, if all positions are 1 then the element is said to be \textit{possibly in the set}.
-The reason a Bloom Filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements.
+If all positions are 1 then the element is said to be \textit{possibly in the set}.
+The reason a Bloom filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements.
 For example, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ as illustrated in Figure \ref{fig:bloom-filter-fp1} and an element \textit{el2} to positions $[4,5,6]$ like in Figure \ref{fig:bloom-filter-fp2}.
@@
 
+% FOR EACH SKETCH DO:
+% approx aggregation to compute
+% data structures
+% algorithm for insertion
+% algorithm for query
+% error bound related with parameters
+% time complexity analysis
+% space complexity analysis
+% Applicability to our Hypothesis
 \subsection{Item Frequency and the Count-Min Sketch}
-
-A Count–Min Sketch (CMS) is a probabilistic data structure that gives an estimate of the frequency of each element in the data set.
-The estimate is a minimum count.
-The real frequency of an element will never be lower than the given estimate, hence the name \textit{count min}.
+% approx aggregation to compute
+A Count–Min Sketch (CMS) is an approximate aggregator that estimates the frequency of each element in the data set.
+It is named after the two basic operations used to provide the estimate: counting (\textit{count}) and computing the minimum (\textit{min}).
 The CMS is first introduced in \cite{Cormode-CMS}.
 
+% data structures
+\subsubsection*{Data structure}
 A CMS is represented by a two-dimensional array (or matrix) with width \textit{w} and depth \textit{d}.
@@
 Each hash function will map elements to a position between 1 and \textit{w} --- \textit{i.e.} the column of the two-dimensional array.
-
 Initially, all of the matrix positions are 0.
-When adding an element, for each row \textit{i} of the \textit{d} rows in the matrix, hash the element using that row's hash function to obtain \textit{j}.
-Lastly, for all obtained pairs of \textit{i} and \textit{j}, increment matrix cells \textit{(i, j)} value by 1, as seen in Figure \ref{fig:cms}.
+Figure \ref{fig:initial-cms} represents the two-dimensional array used by the CMS and corresponding dimensions in its initial state.
 
@@
     \begin{center}
-      \includegraphics[scale=0.8]{figures/cms.png}
-      \caption[Count-Min Sketch update]{CMS mapping element \textit{el} using \textit{d} hash functions to determine the column position for each row and increment it}
+      \includegraphics[scale=0.5]{figures/initial-cms.png}
+      \caption[Count-Min Sketch matrix initial state]{Count-Min Sketch matrix initial state of \textit{d}=4 and \textit{w}=9}
+      \label{fig:initial-cms}
+    \end{center}
+\end{figure}
+
+% algorithm for insertion
+\subsubsection*{Element Insertion}
+When adding an element, for each row \textit{i} of the \textit{d} rows in the matrix, hash the element using that row's hash function to obtain \textit{j}.
+Lastly, for all obtained pairs of \textit{i} and \textit{j}, increment matrix cells \textit{(i, j)} value by 1, as seen in Figure \ref{fig:cms}.
+If elements have different weights these can be added instead of incrementing one unit.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/insertion-cms.png}
+      \caption[Count-Min Sketch insertion]{CMS mapping element \textit{el} using \textit{d}=4 hash functions to determine the column position for each row and increment it}
       \label{fig:cms}
@@
 
-Retrieving the minimum frequency of an element \textit{el} can be done by taking the minimum value of all row counts for \textit{el}. Mathematically, the frequency estimate is given by
+% algorithm for query
+\subsubsection*{Frequency Query}
+Retrieving the minimum frequency of an element \textit{el} can be done by taking the minimum value of all row counts for \textit{el}.
+Mathematically, the frequency estimate is the minimum value of the set of all row values, for each row \textit{i} of the total \textit{d} rows, given by
 \[ min \{matrix[i][\textit{h\textsubscript{i}}(el)]\} \]
-for each row \textit{i} of the total \textit{d} rows.
+
+Figure \ref{fig:query-cms} illustrates this procedure.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/query-cms.png}
+      \caption[Count-Min Sketch frequency query]{Count-Min Sketch frequency query of element \textit{el}}
+      \label{fig:query-cms}
+    \end{center}
+\end{figure}
 
 %The CMS is not always correct because...
-
 \subsubsection*{Time complexity analysis}
-When adding an element or querying the frequency of one, the Count-Min Sketch loops through each row and applies a constant in time hash function, proceeding to a position update or retrieval, respectively.
-Hence, the time complexity is of \textit{O(d)}.
-Given \textit{d} is a constant, the CMS has constant time complexity --- \textit{O(1)}.
+When adding an element, the Count-Min Sketch loops through each row and applies a constant in time hash function, updating the value of the target cell.
+Hence, insertion loops through \textit{d} rows and has time complexity of \textit{O(d)}.
+Similarly, querying the frequency of an element \textit{el} consists in computing the minimum of all row values.
+Since both insertion and query have constant time complexity, the CMS is constant in time.
 
@@
 The memory used will correspond to the \textit{wd} counts, hence \textit{O(wd)} space complexity.
-Since both \textit{w} and \textit{d} are constants, the CMS has constant space complexity --- \textit{O(1)}.
+Since both \textit{w} and \textit{d} are constants, the CMS has constant space complexity.
 

### 2020-02-22 chapters/background.tex
HyperLogLog. PDIS delivery.

@@
 
-%1st paradigm, processing of large volumes of data
-%oldest processing paradigm, simplest
 Batch processing is the processing of data grouped by batches.
@@
 Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements.
-%On the other hand, stream processing refers to the processing of data streams.
-A data stream is essentially an unbounded data set.
-This is in contrast to batch processing which works on a finite data set.
-This distinction is important since some computations are impossible to compute accurately on unbounded data sets.
-Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements.
-An example of such Stream Processing Engines (SPE) is Apache Flink \cite{flink}.
 
-% background on batch systems, micro-batch, true streaming
-%Batch processing is considered the oldest between both processing paradigms.
-Historically speaking, Apache Hadoop \cite{hadoop} was one of the first widely known technologies for large-scale batch processing, created in 2005.
-Nearly a decade later, Apache Spark \cite{spark} was released.
-Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times.
-As more and more data became available, the more need for stream processing increased.
-% join this with previous batch intro
 In order to provide a streaming solution, Apache Spark created the Spark Streaming module.
@@
 On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters.
+Approximate aggregations trade precision of results for efficient time and space usage.
+Such trade-off between accuracy and time/space efficiency will be made clear when discussing some of the existing approximate aggregators in Section \ref{sec:pds}.
 
-Aggregation is present in most programming languages and frameworks.
+Aggregations are present in most programming languages and frameworks.
 Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}.
 Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one.
-\textit{Map} task will transform each item as desired and the \textit{reduce} task will merge the collection into a single value.
+\textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value.
 Reduce methods make use of a \textit{combine} function and recursively apply it.
@@
 \subsection{Aggregation properties}
+\label{sec:agg-properties}
 
@@
 
-These properties are used to evaluate SWAG algorithms.
+%These properties are used to evaluate SWAG algorithms.
 Different algorithms might only apply to certain families of aggregations.
-This way, the multiple SWAG algorithms presented throughout this Thesis will be bench-marked not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.
+This way, the multiple SWAG algorithms presented throughout this Thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.
 
@@
 To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}.
-Assuming all the temperatures can be queried from the \textit{Temperatures} table and that to each rack corresponds a different \textit{rack\_id} we have the following query:
+Assuming all the temperatures can be queried from the \textit{Temperatures} table and that each rack corresponds a different \textit{rack\_id} we have the following query:
 
@@
 Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state.
-Despite storing information using different structures, every SWAG algorithm must implement three methods:
 
-\begin{itemize}
-    \item \textit{insert} - inserts new value in the window
-    \item \textit{evict} - evicts all values that no longer fit in the window
-    \item \textit{query} - returns the aggregation state/value
-\end{itemize}
+The aggregation properties analyzed in Section \ref{sec:agg-properties} are used to evaluate SWAG algorithms.
+Different algorithms might only apply to certain families of aggregations.
+This way, the multiple SWAG algorithms presented throughout this Thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.
 

### 2020-02-22 chapters/sota.tex
HyperLogLog. PDIS delivery.

@@
 
-\subsection{A two window paradigm algorithm for change detection}
+\subsection*{A two window paradigm algorithm for change detection}
 \label{subsec:2-window}
@@
 This is the purpose of the \textit{d} function.
-The authors evaluate many statistical tests as possible implementations of this \textit{d} function that must truly quantify an intuitive notion of change.
+The authors evaluate several statistical methods as possible implementations of this \textit{d} function in order to encode an intuitive notion of change.
 
-Experiments were made using the following statistical tests as the \textit{d} function: Wilcoxon signed-rank, $\phi$\textsubscript{A} and $\Xi$\textsubscript{A}.
+The following statistical tests for the \textit{d} function were used: Wilcoxon signed-rank, $\phi$\textsubscript{A} and $\Xi$\textsubscript{A}.
 The Wilcoxon test is well known but the $\phi$\textsubscript{A} and the $\Xi$\textsubscript{A} statistical tests were defined by the authors.
@@
 It occurs once at the beginning of execution and once per change detection upon which the algorithm resets the windows' contents.
-To provide an accurate time complexity, we take into account the probability \textit{P} of change occurring and being detected leading to a reset.
+To provide an accurate time complexity, we define \textit{P} as the probability of a change occurring and being detected leading to a reset.
 Assuming \textit{P} is within $[0, 1]$, we have a time complexity of \textit{O(P(k + k log(2k)))} per reset --- \textit{i.e.} the probability of a reset occurring times \textit{O(k)} to initialize the windows' contents plus \textit{O(k log(2k))} to build the balanced tree.
@@
 
-\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
-As discussed in Section \ref{sec:back-swag-algs} ...
-Besides RFS and SOE other methods were proposed like DABA...
-\subsection{DABA}
-\subsubsection*{Time complexity analysis}
-\subsubsection*{Space complexity analysis}
-\subsubsection*{Applicability to our Hypothesis}
+%\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
+%As discussed in Section \ref{sec:back-swag-algs} ...
+%Besides RFS and SOE other methods were proposed like DABA...
+%\subsection{DABA}
+%\subsubsection*{Time complexity analysis}
+%\subsubsection*{Space complexity analysis}
+%\subsubsection*{Applicability to our Hypothesis}
 
 \section{Probabilistic Data Structures} \label{sec:pds}
-In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs).
+%TODO: UNCOMMENT
+%In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs).
 However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods.
@@
 
-\subsection{Membership Queries and the Bloom filter}
+These aggregators are commonly denoted as sketches and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds.
+Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.
+
+\subsection{Membership Queries and the Bloom filter} \label{sec:bloom}
 % approx aggregation to compute
@@
       \includegraphics[scale=1.1]{figures/initial-bloom.png}
-      \caption[Bloom filter bit array initial state]{Bloom filter bit array initial state of \textit{m}=7}
+      \caption[Bloom filter initial state]{Bloom filter bit array initial state of \textit{m}=7}
       \label{fig:initial-bloom-filter}
@@
       \includegraphics[scale=0.4]{figures/query-bloom.png}
-      \caption[Bloom filter membership test]{Bloom filter membership test of element \textit{el}. Finding at least one '0' indicates that it is not in the set.}
+      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one '0' indicates that it is not in the set.}
       \label{fig:bloom-filter}
@@
       \includegraphics[scale=0.4]{figures/fp-bloom-1.png}
-      \caption[Bloom filter insertion]{Insertion of \textit{el1} in the Bloom filter}
+      \caption[Bloom filter insertion of \textit{el1}]{Insertion of \textit{el1} in the Bloom filter}
       \label{fig:bloom-filter-fp1}
@@
       \includegraphics[scale=0.4]{figures/fp-bloom-2.png}
-      \caption[Bloom filter insertion]{Insertion of \textit{el2} in the Bloom filter}
+      \caption[Bloom filter insertion of \textit{el2}]{Insertion of \textit{el2} in the Bloom filter}
       \label{fig:bloom-filter-fp2}
@@
 
-% FOR EACH SKETCH DO:
-% approx aggregation to compute
-% data structures
-% algorithm for insertion
-% algorithm for query
-% error bound related with parameters
-% time complexity analysis
-% space complexity analysis
-% Applicability to our Hypothesis
 \subsection{Item Frequency and the Count-Min Sketch}
 % approx aggregation to compute
-A Count–Min Sketch (CMS) is an approximate aggregator that estimates the frequency of each element in the data set.
+The Count-Min Sketch (CMS) uses two principles from the discussed Bloom filters in Section \ref{sec:bloom}.
+First, Bloom filters show that precision can be sacrificed to achieve space savings.
+Second, they are very similar at a technical level.
+
+A Count–Min Sketch \cite{Cormode-CMS} is an approximate aggregator that estimates the frequency of each element in the data set.
 It is named after the two basic operations used to provide the estimate: counting (\textit{count}) and computing the minimum (\textit{min}).
-The CMS is first introduced in \cite{Cormode-CMS}.
 
@@
       \includegraphics[scale=0.5]{figures/initial-cms.png}
-      \caption[Count-Min Sketch matrix initial state]{Count-Min Sketch matrix initial state of \textit{d}=4 and \textit{w}=9}
+      \caption[Count-Min Sketch initial state]{Count-Min Sketch matrix initial state of \textit{d}=4 and \textit{w}=9}
       \label{fig:initial-cms}
@@
 
-Figure \ref{fig:query-cms} illustrates this procedure.
+Figure \ref{fig:query-cms} illustrates this procedure, where the result of the query "What is the frequency of \textit{el}?" returns "2".
 
@@
 
-%The CMS is not always correct because...
+% error bound related with parameters
+\subsubsection*{Optimal number of hash functions} 
+For an element \textit{el}, each row will have one position incremented by \textit{el}'s true frequency.
+However, multiple elements may be mapped to the same position within a row, thus overlapping their frequencies.
+Hence, for an element \textit{el}, the CMS provides a frequency estimate that is greater or equal to the true frequency of \textit{el} in the set.
+
+The number of hash functions \textit{d} to be used depends on the desired error probability. To achieve an error probability $\delta$, \textit{d} is given by the following expression:
+\begin{equation}
+    d = \lceil ln \frac{1}{\delta} \rceil
+\end{equation}
+For example, to achieve an error $\delta$ = 1\%, \textit{d} = 5 hash functions are necessary.
+
 \subsubsection*{Time complexity analysis}
@@
 \subsection{Cardinality Estimation and the HyperLogLog}
-In the paper \cite{Flajolet-PCA}, a class of probabilistic counting algorithms is introduced to estimate the number of distinct elements based on \textit{"bit pattern observables"} in the binary representation of the hashed values --- \textit{i.e.} patterns related to the number and positions of 0's and 1's in a binary string.
+A set of probabilistic counting algorithms is introduced by Flajolet et al.
+in \cite{Flajolet-PCA} to estimate the number of distinct elements of a set --- also known as the cardinality of the set.
+These techniques are based on \textit{"bit-pattern observables"} in the binary representation of the hashed values.
+Bit-pattern observables are defined by the authors as \textit{"patterns of bits occurring at the beginning of the (binary) S-values"}, with \textit{S} as the target set.
 
-\subsubsection{LogLog}
-Later on, the same authors propose the LogLog algorithm \cite{Flajolet-LogLog}.
-In LogLog, the \textit{bit pattern observable} recorded from each item's hash value is the position of the leftmost 1-bit.
-The authors claim it is related to the total number of distinct elements in the data set.
+Later on, the authors propose the LogLog algorithm \cite{Flajolet-LogLog}.
+In LogLog, the bit-pattern observable recorded for each item is the position of the leftmost 1-bit, denoted as \textit{P}.
+The authors claim it is related to the total number of distinct elements in the data set in that it is \textit{"more or less a likely indication that the cardinality of S is at least $2^P$"}.
 
-\subsubsection{HyperLogLog}
-A couple of years later, the same authors propose HyperLogLog (HLL) in \cite{Flajolet-HLL}.
-HLL solves the problem of counting distinct elements in a data set, also known as the cardinality of the data set \cite{Flajolet-HLL}.
-HLL gives an approximation of the cardinality of the data set.
+% FOR EACH SKETCH DO:
+% approx aggregation to compute
+% data structures
+% algorithm for insertion
+% algorithm for query
+% error bound related with parameters
+% time complexity analysis
+% space complexity analysis
+% Applicability to our Hypothesis
+\subsubsection*{HyperLogLog}
+A couple of years later, the same authors propose HyperLogLog (HLL) \cite{Flajolet-HLL}, an approximate aggregator to compute the cardinality of a set.
 
-In HyperLogLog, an item is hashed into a binary string.
-Based on the first \textit{n} bits, the string is passed on to one of the \textit{$2^n$ buckets}.
-After discarding the first \textit{n} bits, the number of trailing 0's plus 1 is counted.
-In other words, the position of the leftmost 1-bit is recorded, like in LogLog.
-This is what is stored in each bucket: the maximum value of the position of the leftmost 1-bit of all seen items so far.
+HLL uses the same bit-pattern observable as LogLog --- \textit{P}, the position of the leftmost 1-bit.
+With LogLog, the authors realize that storing only one observation produces extremely inaccurate predictions due to the high variability of a single value.
+The proposed solution is to apply \textit{stochastic averaging}.
+As the authors state, stochastic averaging \textit{"consists in emulating the effect of m experiments"}.
+Effectively speaking, the authors propose dividing the input stream into \textit{m} substreams or buckets, maintaining one observable per bucket and computing \textit{P} as an average of all bucket values (stochastic averaging).
 
-%TODO \textbf{HYPERLOGLOG HOW IT WORKS IMAGE}
+\subsubsection*{Data Structure}
+HyperLogLog (HLL) maintains \textit{m} bit-pattern observables, each being the leftmost 1-bit position, \textit{P}.
+Since each position is a single integer, HLL maintains \textit{m} integer counters, as seen in Figure \ref{fig:hll-ds}.
+Each of these counters holds the maximum value for the bit-pattern observable seen so far.
+The first bit is assumed to be in position "1" and all counters begin with "0".
 
-Finally, the estimate for the number of distinct elements will be $2^p$, where \textit{p} is the harmonic mean of all bucket values holding the maximum leftmost 1-bit position value.
-This computation of a harmonic mean over \textit{m} buckets is called stochastic averaging and is the main difference between HyperLogLog and LogLog.
-Given the nature of streaming systems, it is also important to notice that this algorithm is suitable for a distributed system since each of the buckets can be updated independently.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/hll-ds.png}
+      \caption[HyperLogLog initial state]{HyperLogLog with \textit{m}=4 observable buckets}
+      \label{fig:hll-ds}
+    \end{center}
+\end{figure}
+
+\subsubsection*{Element Insertion}
+In HyperLogLog, new elements are hashed into a binary string.
+The first \textit{n} bits of the hashed representation will map to one of the \textit{m}=$2^n$ buckets.
+Effectively speaking, the first \textit{n} bits work as a \textit{bucket ID}.
+The bit-pattern observable \textit{P} --- the position of the leftmost 1-bit --- is counted from the binary string discarding the first \textit{n} bits.
+The observable \textit{P} is stored in the pre-determined bucket if it is greater than the value the bucket contains.
+
+Figure \ref{fig:hll-insert} demonstrates such procedure for a new element \textit{el}.
+The hash representation of \textit{el}, using \textit{n}=2 bits, maps to bucket \textit{10}.
+Discarding the first \textit{n}=2 bits and counting from 1, the position of the leftmost 1-bit is \textit{P}=4.
+The previous value of bucket \textit{10} was 0 and because \textit{P}=4 is greater than 0 the bucket value is updated.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=1]{figures/hll-insertion.png}
+      \caption[HyperLogLog insertion]{HyperLogLog \textit{el} insertion with \textit{m}=4 and \textit{n}=2 bits}
+      \label{fig:hll-insert}
+    \end{center}
+\end{figure}
+
+\subsubsection*{Cardinality Query}
+The cardinality of the set is estimated to be $2^\textit{P}$ where the bit-pattern observable \textit{P} is the position of the leftmost 1-bit.
+As discussed, HyperLogLog (HLL) differentiates from LogLog in that it applies a stochastic averaging process.
+Hence, the value of \textit{P} will be the harmonic mean of all \textit{m} buckets.
 
 \subsubsection*{Time complexity analysis}
-Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, count the leftmost 1-bit position and update the bucket value if need be.
-Estimating the cardinality of the set amounts to finding the maximum value \textit{p} between all buckets and computing $2^p$.
-Hence, both inserting a new element and estimating the cardinality take constant time --- \textit{O(1)}.
+Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, counting the leftmost 1-bit position and updating the bucket value if need be.
+Estimating the cardinality of the set amounts to finding the maximum value \textit{P} between all buckets and computing $2^P$.
+Hence, both inserting a new element and estimating the cardinality of the set take constant time.
 
 \subsubsection*{Space complexity analysis}
-Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in  \cite{Flajolet-HLL}: 
+Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in the original HyperLogLog paper \cite{Flajolet-HLL}: 
 \say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}
 
-HyperLogLog uses \textit{m} buckets that store a single integer value.
-This gives it complexity of \textit{O(m)} where \textit{m} is a small constant.
-Hence, HyperLogLog has constant space complexity --- \textit{O(1)}.
+HyperLogLog uses \textit{m} buckets that store a single integer value, thus the space complexity is \textit{O(m)} where \textit{m} is a rather small constant.
+Hence, HyperLogLog has constant space complexity.
 
 \subsubsection*{Applicability to our Hypothesis}
-Similar to the previously studied probabilistic data structures, HyperLogLog (HLL) has constant space and time complexity while giving an approximation of a set's cardinality.
-Hence, HLLs are considered valuable methods for our work.
+Similar to the previously studied approximate aggregators, HyperLogLog (HLL) has constant space and time complexity and provides an estimate of a set's cardinality.
+Hence, HLLs are considered valuable aggregators for our Thesis.
 

### 2020-02-24 chapters/background.tex
Background edit. Two-Stacks.

@@
 
-\section{Aggregations over data streams} \label{sec:aggregations}
+\section{Aggregations over data} \label{sec:aggregations}
 
@@
 On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters.
-Approximate aggregations trade precision of results for efficient time and space usage.
-Such trade-off between accuracy and time/space efficiency will be made clear when discussing some of the existing approximate aggregators in Section \ref{sec:pds}.
+Approximate aggregations are required when memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time.
+Thus, approximate aggregators trade precision of results for efficient time and space usage.
+In Section \ref{sec:pds} we explore approximate aggregators that make such a trade-off.
 
@@
 
-\subsection{Aggregation properties}
-\label{sec:agg-properties}
-
-Different aggregations have different properties.
-These properties are listed in \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}, where the authors group different aggregations according to the properties they satisfy.
-Table \ref{tbl:aggregations-properties} was adapted from this work and categorizes different families of operations based on said properties.
-
-\begin{table}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.5]{figures/aggregation-operations-properties.png}
-      \caption[Aggregation properties]{Aggregation properties.
-Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
-      \label{tbl:aggregations-properties}
-    \end{center}
-\end{table}
-
-To understand these properties, we must be familiar with the \textit{combine} and \textit{lower} functions.
-The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state.
-The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state.
-For instance, consider the use case of computing a sum aggregation over a window of data.
-The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event.
-The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.
-
-The following are formal definitions of these properties and were summarized from \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. An aggregation operation is said to have:
-
-\begin{enumerate}
-    \item  an \textbf{\textit{invertible combine}} if a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$
-
-    \item  an \textbf{\textit{associative combine}} if $(x \oplus y) \oplus z = x \oplus (y \oplus z)$, for all $x$ and $y$
-
-    \item  a \textbf{\textit{commutative combine}} if $x \oplus y = y \oplus x$, for all $x$ and $y$
-
-    \item  a \textbf{\textit{size-preserving combine}} if the aggregation state resulting from $\oplus$ always occupies the same space in memory
-
-    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
-\end{enumerate}
-
-%These properties are used to evaluate SWAG algorithms.
-Different algorithms might only apply to certain families of aggregations.
-This way, the multiple SWAG algorithms presented throughout this Thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.
-
 \subsection{Sliding Window Aggregations}
 
-Unlike many other scenarios that focus on data at rest, streaming aggregations handle data in motion.
 While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one.
@@
 The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs).
-Once again let us consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs.
+Considering the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs.
 How would we compute the average temperature per rack in the data center in the last 5 minutes?
@@
 
-A sliding window aggregation (SWAG) algorithm allows the computation of an exact or approximate aggregation over a sliding window.
-A SWAG algorithm may restrict the set of possible to compute aggregations based on their properties.
+A sliding window aggregation algorithm allows the computation of an exact or approximate aggregation over a sliding window.
+A SWAG algorithm may restrict the set of computable aggregations based on their properties.
 Different SWAG algorithms use different data structures for the window contents and the incremental aggregation state.
 
-The aggregation properties analyzed in Section \ref{sec:agg-properties} are used to evaluate SWAG algorithms.
+Two of the most basic SWAG algorithms are \textit{Recalculate-From-Scratch (RFS)} and \textit{Subtract-On-Evict (SOE)}.
+
+Recalculate-From-Scratch (RFS) is the simplest and most general SWAG algorithm.
+As the name suggests, this algorithm recomputes the aggregation over the entire window for each window update.
+For instance, the computation of a sum over a sliding window using RFS means that whenever the window changes all elements are summed.
+This approach works for every aggregation.
+However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing.
+
+Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value.
+For each element insertion or eviction from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation.
+SOE relies on expiring the evicted elements from the aggregation state.
+Hence, SOE only works for \textit{invertible} aggregations.
+To illustrate, consider the computation of a sum aggregation over a sliding window.
+The total sum of the window will be the aggregation state or value.
+When inserting an element, the updated total window sum equals the total sum plus the new element.
+The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed this far minus the evicted element.
+With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity.
+However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable.
+
+Subtract-On-Evict is an improvement over the Recalculate-From-Scratch algorithm that  based on the idea of aggregation invertibility.
+The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.
+
+\subsection{Aggregation properties}
+\label{sec:agg-properties}
+
+Tangwongsan et al. construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfied properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+
+\begin{table}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/aggregation-operations-properties.png}
+      \caption[Aggregation properties]{Aggregation properties.
+Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
+      \label{tbl:aggregations-properties}
+    \end{center}
+\end{table}
+
+To understand these properties, we must be familiar with the \textit{combine} and \textit{lower} functions.
+The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state.
+The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state.
+For instance, consider the use case of computing a sum aggregation over a window of data.
+The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event.
+The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.
+
+Based on the \textit{combine} and \textit{lower} functions, the following are formal definitions of these properties. An aggregation is said to have:
+
+\begin{enumerate}
+    \item  an \textbf{\textit{invertible combine}} if a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$
+
+    \item  an \textbf{\textit{associative combine}} if $(x \oplus y) \oplus z = x \oplus (y \oplus z)$, for all $x$ and $y$
+
+    \item  a \textbf{\textit{commutative combine}} if $x \oplus y = y \oplus x$, for all $x$ and $y$
+
+    \item  a \textbf{\textit{size-preserving combine}} if the aggregation state resulting from $\oplus$ always occupies the same space in memory
+
+    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
+\end{enumerate}
+
+The aggregation properties analyzed in this Section will be used to evaluate SWAG algorithms.
 Different algorithms might only apply to certain families of aggregations.
 This way, the multiple SWAG algorithms presented throughout this Thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.
-
-Tangwongsan et al. present two of the most basic SWAG algorithms: \textit{Recalculate From Scratch (RFS)} and \textit{Subtract On Evict (SOE)}.
-
-\subsubsection{Recalculate From Scratch}
-Recalculate From Scratch (RFS) is the simplest and most general SWAG algorithm.
-As the name suggests, this algorithm recomputes the aggregation over the entire window for each insertion and eviction.
-For instance, the computation of a sum over a sliding window using RFS means that upon each insertion or eviction all elements are summed.
-
-This approach does not restrain the aggregation in terms of properties: it works for every family of aggregations.
-However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing.
-
-\subsubsection{Subtract On Evict}
-Subtract On Evict (SOE) only works for aggregations whose  \textit{combine} $\oplus$ function is invertible.
-That is, as defined in Table \ref{tbl:aggregations-properties}, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$.
-
-SOE inserts new elements into the window and combines the previous aggregation value with the new element.
-On eviction, SOE applies the inverse of the combine function, $\ominus$, to expire the evicted element from the aggregation state.
-
-With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity.
-However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable.
+For example, under this framework, to implement Subtract-On-Evict (SOE), both the \textit{combine} $\oplus$ and its inverse $\ominus$ functions must be implemented so that SOE applies $\ominus$ to expire the evicted element from the aggregation state.
+Since SOE only restriction is for the aggregation to be invertible it works for the \textit{sum-like}, \textit{collect-like} and \textit{median-like} families in Table \ref{tbl:aggregations-properties}.
 

### 2020-02-24 chapters/sota.tex
Background edit. Two-Stacks.

@@
 
-In this Chapter, we will discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time computations over true sliding windows using resource-lightweight approaches on streaming systems.
+In this Chapter, we discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time computations over true sliding windows using resource-lightweight approaches on streaming systems.
 We present a wide analysis of several categories of algorithms that seemed like viable options to build the desired system.
 
-\section{Detecting Change in Data Streams} \label{sec:change-detect}
+\section{Data Mining} \label{sec:change-detect}
 
-For static data sets, it is reasonable to assume that the data was generated by a fixed process, for example, the data is a sample from a static distribution.
-But a data stream has necessarily a temporal dimension, and the underlying process that generates the data stream can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}.
+A data stream has a temporal dimension and the underlying process that generates the data stream can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}.
 It is this change in the underlying distribution that we want to detect and report as a data pattern shift.
 
-\subsection*{A two window paradigm algorithm for change detection}
+\subsection{A two window paradigm algorithm for change detection}
 \label{subsec:2-window}
 
-The authors of \cite{Kifer-Detecting-Change} propose an algorithm for change detection in data streams.
-The only assumption made is that data points are generated independently.
+Kifer et al.
+propose an algorithm for change detection in data streams \cite{Kifer-Detecting-Change}.
 The authors propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions.
@@
 The proposed method uses two windows, a reference \textit{W\textsubscript{1}} and a sliding one \textit{W\textsubscript{2}}.
-The authors used tuple-based windows of size \textit{k} and the sliding window was a true sliding one, as described in Section \ref{sec:windows}.
+The authors use tuple-based windows of size \textit{k} and a true sliding window.
 The reference window \textit{W\textsubscript{2}} works as a \textit{baseline} and contains the first \textit{k} points of the stream that occurred after the last detected change.
@@
 The algorithm begins by filling both windows with the first \textit{k} tuples.
-Then it slides window \textit{W\textsubscript{2}} (by one, since it is a true sliding window).
-It does this \textit{k} times.
-In between slides, the algorithm checks for change by applying the \textit{d} function, which measures the discrepancy between the two windows contents, \textit{W\textsubscript{1}} and \textit{W\textsubscript{2}}.
+Then it slides window \textit{W\textsubscript{2}} (by one tuple, since it is a true sliding window).
+It does this \textit{k} times and in between slides checks for change by applying the \textit{d} function, which measures the discrepancy between the two windows, \textit{W\textsubscript{1}} and \textit{W\textsubscript{2}}.
 Whenever this discrepancy is above a certain threshold $\alpha$, the algorithm reports a change in the data stream.
@@
 
-This algorithm reduces change detection in data streams to testing whether two samples, the \textit{reference} and the \textit{sliding window} contents, are generated by different distributions.
-Consequently, the authors study the case of detecting differences in distribution between two samples.
-This is the purpose of the \textit{d} function.
-The authors evaluate several statistical methods as possible implementations of this \textit{d} function in order to encode an intuitive notion of change.
+This algorithm reduces change detection in data streams to testing whether two samples --- \textit{reference} and \textit{sliding} windows --- are generated by different distributions.
+The authors study methods for identifying differences in distribution between two samples and implement them as the \textit{d} function.
+Furthermore, they develop their own data structure and algorithm for efficient computation of \textit{d}.
 
-The following statistical tests for the \textit{d} function were used: Wilcoxon signed-rank, $\phi$\textsubscript{A} and $\Xi$\textsubscript{A}.
-The Wilcoxon test is well known but the $\phi$\textsubscript{A} and the $\Xi$\textsubscript{A} statistical tests were defined by the authors.
-Since computing these three statistical tests for our windows contents takes identical time and space, we consider that a thorough analysis of each one is out of scope for this Thesis, as it does not provide relevant information.
+%init - \textit{O(klog(k))}    
+%runtime - \textit{O(klog(k)}
 
-The authors conclude that no test is best in all of the experiments done.
-However, the $\phi$\textsubscript{A} and $\Xi$\textsubscript{A} statistics vastly outperformed the others in some tests and did not perform that much worse in other test cases.
-
-Diving deeper in the algorithm presented in \ref{fig:change-detection-2-windows} we now analyze the computation of the \textit{d} function value.
-In the authors' work, efficient computation of the previously mentioned statistical tests --- \textit{i.e.} the value of function \textit{d} --- is further explored and an algorithm for such is presented.
-First, the \textit{KS structure} is presented.
-The formal definition given by the authors is shown in Figure \ref{fig:ks-structure}
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/ks-structure.png}
-      \caption{Formal definition of the KS structure}
-      \label{fig:ks-structure}
-    \end{center}
-\end{figure}
-
-The KS structure is built from the reference \textit{W\textsubscript{1}} and sliding \textit{W\textsubscript{2}} windows.
-Consider the pair \textit{(\textit{W\textsubscript{1}}, \textit{W\textsubscript{2}})} where the size of both windows is \textit{k}.
-Creating the KS structure \textit{Z} of size \textit{2k} is done by joining the elements from \textit{W\textsubscript{1}} and \textit{W\textsubscript{2}}.
-As stated in \ref{fig:ks-structure}, each element of \textit{Z} will have a value and a weight.
-The values will be the actual elements joined.
-The weights of elements coming from \textit{W\textsubscript{1}} will be \textit{-1/k} while the weights of elements coming from \textit{W\textsubscript{2}} will be \textit{1/k}.
-The final KS structure \textit{Z} must be sorted in increasing order by value.
-The KS structure \textit{Z} can be maintained throughout running time by making use of a balanced tree, such as a B-tree, in \textit{O(log(2k))} time.
-By making use of the KS structure, all of the mentioned statistics can be computed in \textit{O(2k)} time.
-
-\subsubsection*{Time complexity analysis}
 %k to initialize and when change is detected
@@
  %log(2k) to maintain the tree
-The algorithm begins by collecting the first \textit{k} tuples.
-Both windows will have size \textit{k}.
-This takes \textit{O(k)} time but is a task that will only occur on initialization and on change detection, upon which the window contents must reset, as described in \ref{fig:change-detection-2-windows}.
-For each of the next \textit{k} iterations the function \textit{d} will be computed and compared to the threshold $\alpha$.
-As previously mentioned, computing the value of \textit{d} can be done in \textit{O(log(2k))} time per iteration.
-However, the initial time cost of initializing the balanced tree is of \textit{O(k log(2k))}.
-
-Initialization is not frequent.
-It occurs once at the beginning of execution and once per change detection upon which the algorithm resets the windows' contents.
-To provide an accurate time complexity, we define \textit{P} as the probability of a change occurring and being detected leading to a reset.
-Assuming \textit{P} is within $[0, 1]$, we have a time complexity of \textit{O(P(k + k log(2k)))} per reset --- \textit{i.e.} the probability of a reset occurring times \textit{O(k)} to initialize the windows' contents plus \textit{O(k log(2k))} to build the balanced tree.
-
-In runtime, the algorithm will perform \textit{k} iterations where it computes the \textit{d} function value, taking \textit{O(log(2k))} per iteration, amounting to a time complexity of \textit{O(k log(2k)}.
-
-Putting both initialization and runtime complexities we have a total time complexity of \textit{O(P(k + k log(2k)) + k log(2k))}.
+\subsubsection*{Time complexity analysis}
+Computing the value of \textit{d} can be done in \textit{O(log(k))} time per iteration. Hence, given \textit{k} iterations, the time complexity of this algorithm is \textit{O(k log(k))}.
 
 \subsubsection*{Space complexity analysis}
-
-The algorithm presented makes use of two \textit{k} sized windows and one balanced tree of size \textit{k} as well. This represents a space complexity of \textit{O(k+k+k)} or \textit{O(3k)}.
+The presented algorithm makes use of two \textit{k} sized windows and a custom data structure used for the efficient computation of \textit{d} of size \textit{k} as well.
+Since \textit{k} is constant, this method has constant space complexity.
 
 \subsubsection*{Applicability to our Hypothesis}
-
-The algorithm for change detection presented in Subsection \ref{subsec:2-window} has time complexity of \textit{O(P(k + k log(2k)) + k log(2k))}.
-and space complexity of \textit{O(3k)}, with \textit{k} as a window size constant.
-
 In this Thesis, we want to detect and alert data pattern shifts so detecting change in data streams is directly applicable to our use case.
@@
 
-\section{Pattern Mining}
-
-Taking on a new approach, in this Section, we explore pattern mining algorithms rather than algorithms that detect change in data streams.
-The idea behind using pattern mining algorithms being that if we can maintain a list of frequent patterns we would be able to detect when said patterns change --- \textit{i.e.} the list changes.
-Change in the frequent mined patterns would be associated with a data pattern shift and immediately reported.
+%\subsection{Pattern Mining}
 
 \subsection{Mining Frequent Patterns: the FP-Stream structure}
+An alternative approach to detecting changes in data streams is to perform pattern mining.
+The idea behind using pattern mining algorithms being that if we can maintain a list of frequent patterns we would be able to detect when said patterns change --- \textit{i.e.} the list changes.
+Changes in the frequent mined patterns would be associated with a data pattern shift and immediately reported.
 
@@
 \subsubsection*{Applicability to our Hypothesis}
-While a linear complexity fits most of the scenarios, it does not fit ours since we aim to build a low memory footprint solution, ideally with space complexity of \textit{O(1)}.
+While a linear complexity fits many use cases, it does not fit ours since we aim to build a low memory footprint solution, ideally with constant space complexity.
 Hence, this algorithm will be of no use to us.
@@
 
-%\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
-%As discussed in Section \ref{sec:back-swag-algs} ...
-%Besides RFS and SOE other methods were proposed like DABA...
-%\subsection{DABA}
-%\subsubsection*{Time complexity analysis}
-%\subsubsection*{Space complexity analysis}
-%\subsubsection*{Applicability to our Hypothesis}
+\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
+
+As discussed in Section \ref{sec:back-swag-algs}, a sliding window aggregation algorithm allows the computation of an exact or approximate aggregation over a sliding window.
+We have presented Recalculate-From-Scratch (RFS) and Subtract-On-Evict (SOE) as the two most basic generic stream processing algorithms.
+In this Section we analyze other more sophisticated SWAG algorithms that fit different use-cases, restraining the set of computable aggregations to certain families, as defined in Table \ref{tbl:aggregations-properties}.
+
+\subsection{Two-Stacks}
+The Two-Stacks sliding window aggregation (SWAG) algorithm works for scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.} elements are evicted in the same order they were inserted.
+Two-Stacks \cite{Tangwongsan-DABA} implements the FIFO queue using two stacks, namely the front stack F and the back stack B.
+Each element in each of the stacks is a pair \textit{<val, agg>} where \textit{val} is the item's value and \textit{agg} is the aggregation value of everything below it on the stack.
+The pseudocode for Two-Stacks \cite{Tangwongsan-DABA} can be seen in Figure \ref{fig:pseudo-2-stacks}.
+
+Insertion of a new item is always made on stack B.
+The updated aggregation state is computed using the \textit{combine} function $\oplus$, taking as arguments the aggregation value \textit{agg} of the top element of stack B and the item's value \textit{val}.
+The pair pushed onto stack B is a pair with the item's value and the new aggregation value.
+
+Evictions correspond to a simple pop from stack F.
+When F is empty, the elements from stack B are flipped onto stack F, re-applying $\oplus$ to make sure each element in F holds the aggregation value of everything below in the stack.
+Given \textit{n} as the size of stack B, the occasional flip operation takes \textit{O(n)} time.
+
+Computing the sliding window aggregation is done by applying the \textit{combine} function $\oplus$ to the aggregation values of each of the stacks top element.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.45]{figures/2-stacks.png}
+      \caption{Two-Stacks algorithm}
+      \label{fig:pseudo-2-stacks}
+    \end{center}
+\end{figure}
+
+\subsubsection*{Time complexity analysis}
+Querying the aggregation state of the window and inserting a new element can be performed in constant time.
+However, while evicting an item is usually done in constant time, occasionally, a flip operation is required, consuming \textit{O(n)} time, with \textit{n} as the size of stack B.
+Because this flip is occasionally performed, the time complexity of Two-Stacks is said to be \textit{amortized} constant.
+
+\subsubsection*{Space complexity analysis}
+Two-Stacks uses two stacks as auxiliary data structures.
+The combined size of both stacks is at most the size of the sliding window \textit{n}.
+Hence, Two-Stacks has linear space complexity \textit{O(n)}.
+
+\subsubsection*{Applicability to our Hypothesis}
+Two-Stacks is a general sliding window aggregation algorithm to compute \textit{associative} aggregations.
+However, it is linear in space and occasionally performs the computationally intensive operation of flipping stack B onto stack F, making it unfit for our use case.
+
+\subsection{DABA}
+\subsubsection*{Time complexity analysis}
+\subsubsection*{Space complexity analysis}
+\subsubsection*{Applicability to our Hypothesis}
 
 \section{Probabilistic Data Structures} \label{sec:pds}
-%TODO: UNCOMMENT
-%In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs).
+In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs).
 However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods.
@@
 Probabilistic data structures have constant space and time complexity \cite{Singh-PDS-BIGD} making them ideal aggregators to build our real-time and low memory footprint system.
-
-These aggregators are commonly denoted as sketches and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds.
-Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.
 
@@
 
-\iffalse
-\section{Sliding Window Aggregation with Probabilistic Data Structures}
-In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space complexity.
+\section{Sliding Window Aggregations with Probabilistic Data Structures}
+In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space.
 These structures aggregate an endless stream of data.
 However, when working under a sliding window context there is the need not only to insert new events but to evict old ones.
-Probabilistic data structures aggregate all incoming data but have no way of removing elements.
+Probabilistic data structures aggregate all incoming data but have no way of expiring old elements from the aggregation state.
+In this Section, we explore sliding window implementations of the previously discussed probabilistic data structures.
 
-In this Section, we explore and analyze sliding window implementations of the previously discussed probabilistic data structures.
+\subsection{Sliding Bloom filter}
+\subsubsection*{Time complexity analysis}
+\subsubsection*{Space complexity analysis}
+\subsubsection*{Applicability to our Hypothesis}
+
+\subsection{Sliding Count-Min Sketch}
+\subsubsection*{Time complexity analysis}
+\subsubsection*{Space complexity analysis}
+\subsubsection*{Applicability to our Hypothesis}
 
@@
 
-\subsubsection{Time complexity analysis}
-
-\subsubsection{Space complexity analysis}
-
-\subsubsection{Applicability to our Hypothesis}
-
-\fi
-
-%\textbf{image for sliding hyperloglog}
+\subsubsection*{Time complexity analysis}
+\subsubsection*{Space complexity analysis}
+\subsubsection*{Applicability to our Hypothesis}

### 2020-02-26 chapters/sota.tex
Two-Stacks, DABA

@@
 \subsection{Two-Stacks}
-The Two-Stacks sliding window aggregation (SWAG) algorithm works for scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.} elements are evicted in the same order they were inserted.
+The Two-Stacks SWAG algorithm works for \textit{associative} aggregations in scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.} elements are evicted in the same order they were inserted.
 Two-Stacks \cite{Tangwongsan-DABA} implements the FIFO queue using two stacks, namely the front stack F and the back stack B.
 Each element in each of the stacks is a pair \textit{<val, agg>} where \textit{val} is the item's value and \textit{agg} is the aggregation value of everything below it on the stack.
-The pseudocode for Two-Stacks \cite{Tangwongsan-DABA} can be seen in Figure \ref{fig:pseudo-2-stacks}.
+The pseudocode for Two-Stacks can be seen in Figure \ref{fig:pseudo-2-stacks}, with $\oplus$ as the \textit{combine} function and $\overline{\theta}$ as the identity value of the \textit{combine} function, such that $\overline{\theta} \oplus x$ = $x \oplus \overline{\theta}$ = x.
+For example, if $\oplus$ is the arithmetic sum then $\overline{\theta}$ is 0.
+If $\oplus$ is the arithmetic multiplication then $\overline{\theta}$ is 1.
 
 Insertion of a new item is always made on stack B.
-The updated aggregation state is computed using the \textit{combine} function $\oplus$, taking as arguments the aggregation value \textit{agg} of the top element of stack B and the item's value \textit{val}.
-The pair pushed onto stack B is a pair with the item's value and the new aggregation value.
+The updated aggregation state is computed using the \textit{combine} function $\oplus$ taking as arguments the aggregation value \textit{agg} of the top element of stack B and the item's value \textit{val}.
+The pair pushed onto stack B is a pair with \textit{val} and the new aggregation value $\textit{agg} \oplus \textit{val}$.
 
 Evictions correspond to a simple pop from stack F.
-When F is empty, the elements from stack B are flipped onto stack F, re-applying $\oplus$ to make sure each element in F holds the aggregation value of everything below in the stack.
+However, when F is empty, the elements from stack B are flipped onto stack F, re-applying $\oplus$ to make sure each element in F holds the aggregation value of everything below in the stack.
 Given \textit{n} as the size of stack B, the occasional flip operation takes \textit{O(n)} time.
 
-Computing the sliding window aggregation is done by applying the \textit{combine} function $\oplus$ to the aggregation values of each of the stacks top element.
+Computing the sliding window aggregation is done by applying the \textit{combine} function $\oplus$ to the aggregation values of each of the stacks' top element.
 
@@
 
-\subsection{DABA}
+\subsection{De-Amortized Banker’s Aggregator}
+
+The De-Amortized Banker’s Aggregator (DABA) \cite{Tangwongsan-DABA} is a SWAG algorithm that allows the computation of \textit{associative} aggregations.
+Quoting the authors: \textit{"Deamortization is a method that turns the average-case behavior into the worst-case behavior, usually by carefully spreading out expensive operations"}.
+Hence the name \textit{"De-Amortized"}, as the main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout run time.
+
+\subsection*{Data Structure}
+
+DABA uses two queues, \textit{vals} and \textit{aggs}, both of size \textit{n}, seen in Figure \ref{fig:daba-ds}.
+Both queues share six pointers --- \textit{F, L, R, A, B} and \textit{E} --- that point to the same position for each queue.
+These pointers are always ordered as follows: $\textit{F} \leq \textit{L} \leq \textit{R} \leq \textit{A} \leq \textit{B} \leq \textit{E}.$
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/daba-ds.png}
+      \caption{DABA data structure}
+      \label{fig:daba-ds}
+    \end{center}
+\end{figure}
+
+Queue \textit{vals} holds the actual sliding window contents.
+New values are pushed in the back and old values are popped from the front of the queue.
+Queue \textit{aggs} stores aggregations of sub-ranges of \textit{vals}.
+Aside from pointer \textit{E}, each pointer defines a sub-list and each sub-list is aggregated either to the left \textbullet--- or to the right ---\textbullet.
+If a sub-list is aggregated to the left \textbullet---, each \textit{aggs} element will be the aggregation value of all elements in \textit{vals} from that position to the right.
+On the other hand, if a sub-list is aggregated to the right ---\textbullet, each \textit{aggs} element will be the aggregation value of all elements in \textit{vals} from that position to the left.
+Values from \textit{vals} can be aggregated to the left or to the right because DABA restricts the aggregation to be \textit{associative} (refer to Table \ref{tbl:aggregations-properties}).
+
+\textit{l\textsubscript{F}} is aggregated to the left for easier eviction.
+After an element is evicted, the new head of the queue \textit{aggs} --- pointed by \textit{F} --- already contains the aggregation value of all elements from \textit{vals} to its right side.
+Hence, evicting an element amounts to evicting from both \textit{vals} and \textit{aggs} queue.
+Similarly, \textit{l\textsubscript{B}} is aggregated to the right to facilitate insertion.
+Sub-lists \textit{l\textsubscript{L}}, \textit{l\textsubscript{R}} and \textit{l\textsubscript{A}} allow for incremental reversal of the right-aggregated \textit{l\textsubscript{B}} to the left-aggregated \textit{l\textsubscript{F}}.
+This incremental reversal is tantamount to adjusting pointers that causes shifts in sub-list boundaries.
+Thus, elements of \textit{aggs} may change from one sub-list to another and need to have its aggregation value recomputed.
+
+Inserting a new element corresponds to pushing a new value \textit{val} in \textit{vals} and pushing the new aggregation value \textit{agg} in \textit{aggs}.
+The new aggregation value \textit{agg} is computed by applying the \textit{combine} function $\oplus$ to the back element of \textit{aggs} and the new value \textit{val} --- \textit{i.e.} pushing \textit{val} to \textit{vals} and $\textit{aggs[E]} \oplus \textit{val}$ to \textit{aggs}.
+
+Evicting old elements is done by evicting the same number of elements from \textit{vals} and \textit{aggs}.
+
+\textit{l\textsubscript{F}} and \textit{l\textsubscript{B}} combined cover the whole \textit{vals} queue --- \textit{i.e.} every element in the sliding window.
+\textit{l\textsubscript{F}} is aggregated to the left, thus \textit{aggs[F]} contains the aggregation value of everything in \textit{l\textsubscript{F}}.
+On the other hand, \textit{l\textsubscript{B}} is aggregated to the right, thus \textit{aggs[E]} stores the aggregation value of everything in \textit{l\textsubscript{B}}.
+Hence, the aggregation value of the entire window will be the result of \textit{$aggs[F] \oplus aggs[E]$}.
+
+\subsection*{Incremental Reversal of \textit{l\textsubscript{B}} to \textit{l\textsubscript{F}}}
+
+The main conceptual difference between DABA and Two-Stacks is that in Two-Stacks an occasional expensive \textit{flip} operation takes place --- flipping stack B onto stack F --- and in DABA the reversal of \textit{l\textsubscript{B}} to \textit{l\textsubscript{F}} is incremental.
+This allows DABA to achieve constant time complexity (on average) rather than Two-Stacks' amortized constant time complexity.
+
+Figure \ref{fig:pseudo-daba} shows the pseudocode presented by the authors for the DABA algorithm.
+In reality, after each insertion and eviction, besides the procedure already described, function \textit{fixup} is executed.
+The \textit{fixup} function performs the incremental reversal of \textit{l\textsubscript{B}} to \textit{l\textsubscript{F}}.
+The pseudocode \ref{fig:pseudo-daba} shows how such reversal is done but we will not elaborate on such procedure.
+However, it is important to stress out that this is the core idea of DABA: avoid occasional expensive operations, like Two-Stacks' \textit{flip}, by spreading them out across run time, in the form of DABA's \textit{fixup}.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.7]{figures/daba-pseudocode.png}
+      \caption{DABA algorithm}
+      \label{fig:pseudo-daba}
+    \end{center}
+\end{figure}
+
 \subsubsection*{Time complexity analysis}
+Querying the aggregation state as described is done in constant time.
+Insertion and eviction of elements take constant time and invoke function \textit{fixup}.
+The authors develop Theorems that prove function \textit{fixup} is executed in constant time as well.
+Hence, DABA has constant time complexity.
+
 \subsubsection*{Space complexity analysis}
+DABA stores the window contents in the \textit{vals} queue.
+Queue \textit{aggs} is of the same size as \textit{vals} and stores partial aggregations over \textit{vals}.
+Hence, if the window size is \textit{n}, DABA uses two \textit{n}-sized queues.
+Hence, DABA has \textit{O(n)} space complexity, meaning memory consumption grows linearly with window size.
+
 \subsubsection*{Applicability to our Hypothesis}
+Even though DABA improves on the time complexity of Two-Stacks, it still shows linear space complexity regarding window size.
+In our monitoring system, we need to analyze large windows of data, requiring at least logarithmic space complexity, thus making DABA inapplicable for our Hypothesis.
 
@@
 The reason a Bloom filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements.
-For example, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ as illustrated in Figure \ref{fig:bloom-filter-fp1} and an element \textit{el2} to positions $[4,5,6]$ like in Figure \ref{fig:bloom-filter-fp2}.
+For example, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ and an element \textit{el2} to positions $[4,5,6]$.
 This way, bits in positions $[1,2,3,4,5,6]$ are set to 1.
-When querying about the membership of a new element \textit{el3}, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map it to positions $[1,3,5]$ as exemplified in Figure \ref{fig:bloom-filter-fp3}.
+When querying about the membership of a new element \textit{el3}, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map it to positions $[1,3,5]$ as exemplified in Figure \ref{fig:bloom-filter-fp}.
 Since all of these positions are 1 because of the previous elements inserted, the Bloom filter returns \textit{"possibly in set"} when in reality \textit{el3} was not in the set.
@@
       \includegraphics[scale=0.4]{figures/fp-bloom-1.png}
-      \caption[Bloom filter insertion of \textit{el1}]{Insertion of \textit{el1} in the Bloom filter}
-      \label{fig:bloom-filter-fp1}
-    \end{center}
-\end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
       \includegraphics[scale=0.4]{figures/fp-bloom-2.png}
-      \caption[Bloom filter insertion of \textit{el2}]{Insertion of \textit{el2} in the Bloom filter}
-      \label{fig:bloom-filter-fp2}
-    \end{center}
-\end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
       \includegraphics[scale=0.4]{figures/fp-bloom-3.png}
-      \caption[Bloom filter false positive]{Membership query of \textit{el3} and false positive result}
-      \label{fig:bloom-filter-fp3}
+      \caption{Bloom filter false positive}
+      \label{fig:bloom-filter-fp}
     \end{center}

### 2020-02-27 chapters/sota.tex
Sliding HyperLogLog with Lists of Future Possible Maxima

@@
 \section{Data Mining} \label{sec:change-detect}
-
 A data stream has a temporal dimension and the underlying process that generates the data stream can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}.
@@
 
-\subsection{De-Amortized Banker’s Aggregator}
+\subsection{De-Amortized Banker’s Aggregator} \label{sec:daba}
 
@@
 
-\subsubsection*{Time complexity analysis}
+\subsection*{Time complexity analysis}
 Querying the aggregation state as described is done in constant time.
@@
 
-\subsubsection*{Space complexity analysis}
+\subsection*{Space complexity analysis}
 DABA stores the window contents in the \textit{vals} queue.
@@
 
-\subsubsection*{Applicability to our Hypothesis}
+\subsection*{Applicability to our Hypothesis}
 Even though DABA improves on the time complexity of Two-Stacks, it still shows linear space complexity regarding window size.
@@
 % data structures
-\subsubsection*{Data structure}
+\subsection*{Data structure}
 A Bloom filter is an array of bits of size \textit{m}, initially all set to 0.
@@
 % algorithm for insertion
-\subsubsection*{Element Insertion}
+\subsection*{Element Insertion}
 Element insertion and query answering make use of \textit{k} hash functions, each mapping elements to a position in the bit array.
@@
 % algorithm for query
-\subsubsection*{Membership Query}
+\subsection*{Membership Query}
 Performing a membership query --- \textit{i.e.} testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions.
@@
 
-\subsubsection*{False Positive rate}
+\subsection*{False Positive rate}
 % error bound related with parameters
@@
 
-\subsubsection*{Optimal number of hash functions}
+\subsection*{Optimal number of hash functions}
 We want to choose the number of hash functions \textit{k} in order to minimize the false positive rate.
@@
 
-\subsubsection*{Time complexity analysis}
+\subsection*{Time complexity analysis}
 Considering a Bloom filter with \textit{m} bits and \textit{k} hash functions, insertion has \textit{O(k)} time complexity because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1.
@@
 
-\subsubsection*{Space complexity analysis}
+\subsection*{Space complexity analysis}
 Considering a Bloom filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity.
@@
 
-\subsubsection*{Applicability to our Hypothesis}
+\subsection*{Applicability to our Hypothesis}
 Bloom filters exhibit constant space and time complexity making it great methods to test our hypothesis.
@@
 % data structures
-\subsubsection*{Data structure}
+\subsection*{Data structure}
 A CMS is represented by a two-dimensional array (or matrix) with width \textit{w} and depth \textit{d}.
@@
 % algorithm for insertion
-\subsubsection*{Element Insertion}
+\subsection*{Element Insertion}
 When adding an element, for each row \textit{i} of the \textit{d} rows in the matrix, hash the element using that row's hash function to obtain \textit{j}.
@@
 % algorithm for query
-\subsubsection*{Frequency Query}
+\subsection*{Frequency Query}
 Retrieving the minimum frequency of an element \textit{el} can be done by taking the minimum value of all row counts for \textit{el}.
@@
 % error bound related with parameters
-\subsubsection*{Optimal number of hash functions} 
+\subsection*{Optimal number of hash functions} 
 For an element \textit{el}, each row will have one position incremented by \textit{el}'s true frequency.
@@
 
-\subsubsection*{Time complexity analysis}
+\subsection*{Time complexity analysis}
 When adding an element, the Count-Min Sketch loops through each row and applies a constant in time hash function, updating the value of the target cell.
@@
 
-\subsubsection*{Space complexity analysis}
+\subsection*{Space complexity analysis}
 The Count-Min Sketch relies on a two-dimensional matrix as its only data structure.
@@
 
-\subsubsection*{Applicability to our Hypothesis}
+\subsection*{Applicability to our Hypothesis}
 Count-Min Sketches are constant in both time and space. Similarly to Bloom filters, this makes them ideal candidates for our solution.
 
-\subsection{Cardinality Estimation and the HyperLogLog}
+\subsection{Cardinality Estimation and the HyperLogLog} \label{sec:hyper-log-log}
+
 A set of probabilistic counting algorithms is introduced by Flajolet et al.
@@
 % Applicability to our Hypothesis
-\subsubsection*{HyperLogLog}
+\subsection*{HyperLogLog}
 A couple of years later, the same authors propose HyperLogLog (HLL) \cite{Flajolet-HLL}, an approximate aggregator to compute the cardinality of a set.
@@
 
-\subsubsection*{Data Structure}
+\subsection*{Data Structure}
 HyperLogLog (HLL) maintains \textit{m} bit-pattern observables, each being the leftmost 1-bit position, \textit{P}.
@@
 
-\subsubsection*{Element Insertion}
+\subsection*{Element Insertion}
 In HyperLogLog, new elements are hashed into a binary string.
@@
 
-\subsubsection*{Cardinality Query}
+\subsection*{Cardinality Query}
 The cardinality of the set is estimated to be $2^\textit{P}$ where the bit-pattern observable \textit{P} is the position of the leftmost 1-bit.
@@
 
-\subsubsection*{Time complexity analysis}
+\subsection*{Time complexity analysis}
 Adding an item is tantamount to hashing it, determining the corresponding bucket based on the first \textit{n} bits, counting the leftmost 1-bit position and updating the bucket value if need be.
@@
 
-\subsubsection*{Space complexity analysis}
+\subsection*{Space complexity analysis}
 Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in the original HyperLogLog paper \cite{Flajolet-HLL}: 
@@
 
-\subsubsection*{Applicability to our Hypothesis}
+\subsection*{Applicability to our Hypothesis}
 Similar to the previously studied approximate aggregators, HyperLogLog (HLL) has constant space and time complexity and provides an estimate of a set's cardinality.
 Hence, HLLs are considered valuable aggregators for our Thesis.
-
-% PDS have no way to expire old data (must for SWAG)
-% they could be implemented using one of the previously seen SWAG algorithms but lose the space efficiency
-% present sliding versions as solution
 
@@
 Probabilistic data structures aggregate all incoming data but have no way of expiring old elements from the aggregation state.
-In this Section, we explore sliding window implementations of the previously discussed probabilistic data structures.
 
-\subsection{Sliding Bloom filter}
-\subsubsection*{Time complexity analysis}
-\subsubsection*{Space complexity analysis}
-\subsubsection*{Applicability to our Hypothesis}
+The probabilistic data structures studied can be implemented under generic sliding window aggregation algorithms like Recalculate-From-Scratch or DABA.
+However, doing so means we lose the main advantage of using these approximate aggregators: the constant space complexity.
+For example, cardinality estimation of a sliding window can be done employing the HyperLogLog heuristic of estimating cardinality as $2^\textit{p}$, with \textit{p} as the maximum position seen for the leftmost 1-bit, using DABA \ref{sec:daba} for the computation of the value \textit{p} per sliding window.
+With this approach, despite computing the aggregation state in constant time, we have a linear in window size space complexity solution.
+Thus, in this Section, we explore sliding window implementations of the previously discussed probabilistic data structures that keep the space and time complexity constant.
 
-\subsection{Sliding Count-Min Sketch}
-\subsubsection*{Time complexity analysis}
-\subsubsection*{Space complexity analysis}
-\subsubsection*{Applicability to our Hypothesis}
+%\subsection{Sliding Bloom filter}
+%\subsection*{Data Structure}
+%\subsection*{Element Insertion}
+%\subsection*{Element Eviction}
+%\subsection*{Time complexity analysis}
+%\subsection*{Space complexity analysis}
+%\subsection*{Applicability to our Hypothesis}
 
-\subsection{Sliding HyperLogLog}
-HyperLogLog  (HLL) will only be useful in our system if it can be implemented efficiently in a sliding window fashion.
-That is precisely what \cite{Chabchoub-Sliding-HLL} presents us with.
-In this paper, the original HLL algorithm from \cite{Flajolet-HLL} is adapted to a sliding window scenario with the use of slightly more memory.
-More importantly, the estimate is as accurate as in the original HLL algorithm.
+%\subsection{Sliding Count-Min Sketch}
+%\subsection*{Data Structure}
+%\subsection*{Element Insertion}
+%\subsection*{Element Eviction}
+%\subsection*{Time complexity analysis}
+%\subsection*{Space complexity analysis}
+%\subsection*{Applicability to our Hypothesis}
 
-It consumes additional space compared to the original algorithm of at most $5\textit{m}ln(\textit{n}/\textit{m})$ bytes, with \textit{m} being the number of buckets and \textit{n} the number of distinct elements actually flowing through the stream.
-As explained by the authors, storing additional context is required so that the algorithm can work in a sliding window fashion.
-Thus they propose maintaining a shortlist of packets.
-A packet consists of a pair \textit{(ti, Ri)}, where the \textit{ti} is the arrival time of the packet and \textit{Ri} is the position of the leftmost 1-bit in the binary representation of the hashed value associated with that packet.
-The idea is to store only packets that are a possible maximum over a future window.
-Hence, this list is called \textit{List of Future Possible Maxima (LFPM)}.
-Furthermore, they propose efficient methods for the insertion and eviction of new and old elements from the window, respectively.
-In this sliding version, one LFPM is kept per bucket.
-For that reason, estimating the set cardinality is similar to the original HyperLogLog: compute the harmonic mean for all buckets.
-The value of each bucket is the highest \textit{Ri} for that bucket.
-Hence, the extra required memory in the Sliding HyperLogLog algorithm presented is given by the size of the \textit{m} lists of future possible maxima.
+\subsection{Sliding Window Cardinality Estimation and the Sliding HyperLogLog}
 
-\subsubsection*{Time complexity analysis}
-\subsubsection*{Space complexity analysis}
-\subsubsection*{Applicability to our Hypothesis}
+The HyperLogLog (HLL) aggregator presented in Section \ref{sec:hyper-log-log} will only be useful to us if it can be implemented efficiently in a sliding window fashion.
+To that end, we analyze the work of Chabchoub et al.
+that adapts the original HLL aggregator from Flajolet et al.
+to work under a sliding window framework  \cite{Chabchoub-Sliding-HLL}.
+
+The HLL aggregator estimates on the set's cardinality assuming it is close to $2^\textit{p}$, where \textit{p} is the maximum position seen for the leftmost 1-bit in the binary values of all of the set items' hashes.
+In practice, all the aggregator has to compute is \textit{p}, the maximum value between a set of positions.
+Thus, the main idea behind Sliding HyperLogLog is to keep only the positions for the leftmost 1-bit that are possible future maximum values for \textit{p}.
+
+\subsection*{Data Structure}
+
+The adaptation of the original HLL to work under a sliding window scenario requires the storage of additional information.
+The original HLL maintained \textit{m} integer counters, each storing the maximum for the leftmost 1-bit position of the analyzed binary strings.
+The counter held the maximum value for all the seen binary strings so far and there was no way to expire old items.
+To allow eviction, the proposed data structure is very similar to the original \textit{m} integer counters but keeps a list of values per bucket and associates a timestamp to each.
+
+The stochastic averaging process of HLL remains unchanged: incoming items are hashed and distributed across \textit{m} buckets.
+However, the authors propose keeping a short list of \textit{packets} per bucket instead of a single integer count.
+A \textit{packet} is defined as a pair <\textit{t\textsubscript{i}, p\textsubscript{i}}> where \textit{t\textsubscript{i}} is the arrival timestamp of the packet and \textit{p\textsubscript{i}} the position of the leftmost 1-bit in the binary hash representation of the associated value.
+Packets in the list are possible maxima over future windows.
+Because of that, the authors name this list a \textit{"List of Future Possible Maxima (LFPM)"}.
+
+\subsection*{Element Insertion \& Eviction}
+Similarly to the original HLL, incoming elements are hashed and the first \textit{n} bits of the string are used to map the new element to one of the \textit{m} buckets.
+The first \textit{n} bits are then discarded and the bit-pattern observable \textit{p} is computed from the remaining string.
+
+Since each bucket now holds a LFPM instead of a single integer, the procedure for bucket update is different.
+Upon insertion of an element \textit{k}, a packet <t\textsubscript{k}, p\textsubscript{k}> is created with t\textsubscript{k} as the timestamp of arrival of the new element \textit{k} and p\textsubscript{k} as the position of the leftmost 1-bit position of \textit{k}'s binary representation.
+Given the new packet <t\textsubscript{k}, p\textsubscript{k}>, the authors propose the following LFPM update method, that checks for and evicts old packets at each insertion:
+
+\begin{itemize}
+    \item Delete \textit{old packets}: packets that do not belong in the window anymore --- \textit{i.e.} packets <t\textsubscript{i}, p\textsubscript{i}> with t\textsubscript{i} $\leq$ t\textsubscript{k} - WindowSize
+
+    \item Delete less or equally valued packets: packets that will never be a maximum --- \textit{i.e.} packets <t\textsubscript{i}, p\textsubscript{i}> with p\textsubscript{i} $\leq$ p\textsubscript{k}
+
+    \item Add the new packet <t\textsubscript{k}, p\textsubscript{k}>
+\end{itemize}
+
+\subsection*{Sliding Window Cardinality Query}
+
+Estimation of cardinality is done similarly to the original HLL cardinality estimation algorithm.
+To estimate the cardinality at a timestamp \textit{t}, select packets that are in the window --- \textit{i.e.} packets <t\textsubscript{i}, p\textsubscript{i}> from the LFPM where t\textsubscript{i} > \textit{t} - WindowSize.
+This filtering is necessary since the proposed algorithm evicts old packets only at element insertion.
+However, it would be possible to do so when estimating the cardinality or even periodically.
+From the filtered list of packets, compute the highest p\textsubscript{i} among them.
+Do this for each bucket and in the end compute the harmonic mean between all values, just like the original HLL.
+The result will be the averaged bit-pattern observable value \textit{p} and the cardinality is estimated to be $2^\textit{p}$.
+
+\subsection*{Time complexity analysis}
+When performing insertions or evictions, the algorithm requires scanning through the Lists of Future Possible Maxima (LFPMs) to delete old packets or packets that will never be window maxima.
+Similarly, we need to scan the lists when estimating the cardinality to filter for packets within our window.
+Denote \textit{L\textsubscript{n}} as the mean size in packets for each LFPM.
+There are \textit{m} buckets, each with a list.
+To scan through all lists we need to analyze \textit{m L\textsubscript{n}} elements.
+
+The authors prove that \textit{L\textsubscript{n}} has an upper bound of \textit{ln(n/m)}, with \textit{n} as the exact cardinality of the set and \textit{m} as the number of buckets used.
+Hence, in the worst-case scenario, we need to analyze \textit{m ln(n/m)} elements.
+This means that implementing a Sliding HLL using Lists of Future Possible Maxima has \textit{O(ln(n))} time complexity, given the number of buckets \textit{m} is constant.
+
+\subsection*{Space complexity analysis}
+The memory used in the Sliding HLL aggregator implemented with Lists of Future Possible Maxima (LFPMs) amounts to the total size of the \textit{m} lists.
+Assuming 4 byte timestamps and 1 byte bit-pattern observables for each packet, the aggregator uses \textit{(4+1) L\textsubscript{n} m} bytes.
+
+Given \textit{ln(n/m)} as the upper bound for \textit{L\textsubscript{n}}, this sliding implementation of the HLL aggregator consumes up to \textit{(4+1) ln(n/m) m} bytes of memory.
+Considering that the number of buckets \textit{m} is constant, the space complexity of a Sliding HLL implementation with LFPM is \textit{O(ln(n))}.
+Thus, the growth of the window in size is followed by a logarithmic increase in used memory.
+
+The authors claim that the proposed sliding version remains as accurate as the original HLL aggregator.
+In the context of their work, each element in the set is denoted as a "flow" and they state that \textit{"In practice, with an additional memory of only 35kB, a standard error of about 3\% can be achieved for a data stream of several million flows"}.
+
+\subsection*{Applicability to our Hypothesis}
+Implementing a Sliding HLL using Lists of Future Possible Maxima (LFPMs) shows logarithmic time and space complexity while maintaining the same accuracy as the original HLL approximate aggregator.
+Hence, this is a valid approach to test our Hypothesis.

### 2020-03-06 chapters/background.tex
Two-Stacks, DABA
Aggregation on data types

@@
 However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing.
+Figure \ref{fig:pseudo-rfs} shows pseudocode for RFS \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.45]{figures/rfs.png}
+      \caption{Recalculate-From-Scratch algorithm}
+      \label{fig:pseudo-rfs}
+    \end{center}
+\end{figure}
 
@@
 However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable.
+Figure \ref{fig:pseudo-soe} shows pseudocode for SOE \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.45]{figures/soe.png}
+      \caption{Subtract-On-Evict algorithm}
+      \label{fig:pseudo-soe}
+    \end{center}
+\end{figure}
 
@@
 
+%TODO: change the sketch-like from the aggregator to the aggregation: bloom filter to membership query, CMS to frequency estimation and HLL to cardinality estimation
 \begin{table}[!htb]
@@
 
+\section{Aggregations on data types}
+
+Some aggregations are only possible on certain data types.
+For instance, computing an average on a set of strings is not possible.
+In this Section we present the data types we will work with and the set of possible aggregations to compute on them.
+
+We consider four data types: \textit{numericals}, \textit{strings}, \textit{dates} and \textit{categoricals}.
+Numerical variables can be measured.
+When compared to classical programming data types, for example those of the C programming language, numericals comprise variable types \textit{short}, \textit{long}, \textit{int}, \textit{float}, \textit{double} and valid combinations of the previous --- \textit{e.g.} \textit{long int}.
+On the other hand, Strings are sequences of characters.
+The \textit{string} type is the most generic data type in that any other type can be converted to a string.
+Every variable, for example \textit{x}=3, can be turned into a string \textit{s}, for example \textit{s}="3".
+Similarly, a date is a sequence of characters and thus a string.
+However, despite being a string, it has semantics unlike a regular one.
+Knowing the date format, one can extract the day, month and year.
+For that reason we make the distinction between the \textit{string} and \textit{date} data types.
+Lastly, a \textit{categorical} is a variable that takes one of a relatively small and known set of possible values.
+A variable is assigned to a particular category based on its properties.
+The categorical variable is also a sequence of characters and therefore a string variable.
+The main difference between the categorical data type and the string is that we know the set of possible labels the categorical variable might take and this affects the type of computable aggregations.
+
+On a set of numerical variables, we can compute sums, minima, maxima, averages, standard deviations and counts (of zeros, infinities and others).
+When operating on a set of strings, we can aggregate the collection into a concatenation of strings, a count, a membership query, cardinality.
+
 % \section{Concept drift}

### 2020-03-06 chapters/conclusion.tex
Aggregation on data types

@@
 
-We introduced background concepts on large scale data processing, windowing techniques applied to data streams and sliding window aggregations.
-We studied algorithms for change detection in data streams, pattern mining, anomaly detection and sliding window aggregation algorithms, specifically a subset of these named probabilistic data structures.
-
-%%hypothesis validated?
-The monitoring of data streams in real-time is applicable across many domains and applications.
-To the best of our knowledge, we claim that there has not been a real attempt at doing so in a lightweight way.
-
-%%future work
-Preliminary work shows that there exist constant space and time complexity algorithms to aggregate data, thus reducing memory usage.
-We intend to research more about lightweight sliding window aggregations and sliding implementations of probabilistic data structures.
-We also intend to search for new methods of summarizing data that do not necessarily explicitly work in a sliding window fashion.
-After identifying a set of algorithms that meet our requirements, the focus will be on implementing the best theoretical solution and benchmark our streaming system using data sets from the financial fraud space versus previously performed batch analysis.
-
-%%summary of contributions
-The main contribution of this Thesis will be the developed system code as well as a set of tests that measure the system's usefulness based on several parameters, such as recall, precision, memory used and latency.
+Yeeeeeeetttttt

### 2020-03-06 chapters/sota.tex
Aggregation on data types

@@
 
+\textbf{TODO: Add table with SWAG algorithms, time and space complexity and applicable restrictions on the aggregation}
+
 \section{Probabilistic Data Structures} \label{sec:pds}
@@
 \end{equation}
-
-\iffalse
-%TODO hash improvement probably better declared outside of bloom
-An improvement to this structure was defined in \cite{Kirsch-Better-Bloom} where the authors attempt to reduce the overhead of bloom filters operations such as adding a new element or answering a query by focusing on the hash functions.
-The proposed method is based on hashing literature.
-It has been shown that multiple hash functions can be created from the combination of only two hash functions.
-This means that a new hash function \textit{g} can be produced based on  two other independent and uniform existing hash functions, \textit{h1} and \textit{h2}.
-The new hash function will map elements to a universe ranging from \textit{0} to \textit{p-1} and will be defined as \textit{$g(x) = h1(x) + ih2(x) mod p$}.
-Since the only nontrivial computations performed by a Bloom filter are the evaluations of pseudo-random hash functions, any reduction in the required number of pseudo-random hash functions yields a reduction in the time required to add an element or query a Bloom filter.
-
-Bloom filters are not the only approximate aggregators that use hash functions as a basis of their algorithm.
-All of the studied probabilistic data structures do so.
-Therefore, the technique proposed is beneficial in HyperLogLogs, Count-Min Sketches and Bloom filters.
-\fi
 
@@
 
+\iffalse
+%TODO: Worth to show? HLL uses only one hash, only CMS and Bloom benefit from this.
+\subsection{Hash improvement}
+An improvement to this structure was defined in \cite{Kirsch-Better-Bloom} where the authors attempt to reduce the overhead of bloom filters operations such as adding a new element or answering a query by focusing on the hash functions.
+The proposed method is based on hashing literature.
+It has been shown that multiple hash functions can be created from the combination of only two hash functions.
+This means that a new hash function \textit{g} can be produced based on  two other independent and uniform existing hash functions, \textit{h1} and \textit{h2}.
+The new hash function will map elements to a universe ranging from \textit{0} to \textit{p-1} and will be defined as \textit{$g(x) = h1(x) + ih2(x) mod p$}.
+Since the only nontrivial computations performed by a Bloom filter are the evaluations of pseudo-random hash functions, any reduction in the required number of pseudo-random hash functions yields a reduction in the time required to add an element or query a Bloom filter.
+
+Bloom filters are not the only approximate aggregators that use hash functions as a basis of their algorithm.
+All of the studied probabilistic data structures do so.
+Therefore, the technique proposed is beneficial in Count-Min Sketches and Bloom filters.
+\fi
+
 \section{Sliding Window Aggregations with Probabilistic Data Structures}
@@
 These structures aggregate an endless stream of data.
-However, when working under a sliding window context there is the need not only to insert new events but to evict old ones.
+However, when working with sliding windows there is the need not only to insert new events but to evict old ones.
 Probabilistic data structures aggregate all incoming data but have no way of expiring old elements from the aggregation state.
@@
 However, doing so means we lose the main advantage of using these approximate aggregators: the constant space complexity.
-For example, cardinality estimation of a sliding window can be done employing the HyperLogLog heuristic of estimating cardinality as $2^\textit{p}$, with \textit{p} as the maximum position seen for the leftmost 1-bit, using DABA \ref{sec:daba} for the computation of the value \textit{p} per sliding window.
+For example, cardinality estimation of a sliding window can be done employing the HyperLogLog heuristic of estimating cardinality as $2^\textit{p}$, with \textit{p} as the maximum position seen for the leftmost 1-bit, using DABA \ref{sec:daba} for the computation of \textit{p} per sliding window.
 With this approach, despite computing the aggregation state in constant time, we have a linear in window size space complexity solution.
-Thus, in this Section, we explore sliding window implementations of the previously discussed probabilistic data structures that keep the space and time complexity constant.
+Thus, in this Section, we explore sliding window implementations of the previously discussed probabilistic data structures that keep the space and time complexity at the very least logarithmic.
 
@@
 
-\subsection{Sliding Window Cardinality Estimation and the Sliding HyperLogLog}
+\subsection{Sliding HyperLogLog}
 
 The HyperLogLog (HLL) aggregator presented in Section \ref{sec:hyper-log-log} will only be useful to us if it can be implemented efficiently in a sliding window fashion.
-To that end, we analyze the work of Chabchoub et al.
-that adapts the original HLL aggregator from Flajolet et al.
-to work under a sliding window framework  \cite{Chabchoub-Sliding-HLL}.
+The HLL aggregator estimates the set's cardinality assuming it is close to $2^\textit{p}$, where \textit{p} is the maximum position seen for the leftmost 1-bit in the binary values of all of the set items' hashes.
+In practice, all the aggregator has to compute is \textit{p}, the maximum value of all observables.
+Stochastic averaging is applied by emulating the effects of \textit{m} experiments using \textit{m} buckets to store \textit{m} bit-pattern observables.
 
-The HLL aggregator estimates on the set's cardinality assuming it is close to $2^\textit{p}$, where \textit{p} is the maximum position seen for the leftmost 1-bit in the binary values of all of the set items' hashes.
-In practice, all the aggregator has to compute is \textit{p}, the maximum value between a set of positions.
-Thus, the main idea behind Sliding HyperLogLog is to keep only the positions for the leftmost 1-bit that are possible future maximum values for \textit{p}.
+\subsubsection{Sliding HyperLogLog using Lists of Future Possible Maxima} \label{sec:hll-lfpm}
 
-\subsection*{Data Structure}
+The work of Chabchoub et al. adapts the original HLL aggregator from Flajolet et al. to work under a sliding window framework  \cite{Chabchoub-Sliding-HLL}.
+The main idea behind this implementation is to replace every HLL counter with a \textit{"List of Future Possible Maxima (LFPM)"}.
+
+\subsubsection*{Data Structure}
 
@@
 The original HLL maintained \textit{m} integer counters, each storing the maximum for the leftmost 1-bit position of the analyzed binary strings.
-The counter held the maximum value for all the seen binary strings so far and there was no way to expire old items.
+The counter held the maximum value for all the seen binary strings so far but there was no way to expire old items.
 To allow eviction, the proposed data structure is very similar to the original \textit{m} integer counters but keeps a list of values per bucket and associates a timestamp to each.
@@
 A \textit{packet} is defined as a pair <\textit{t\textsubscript{i}, p\textsubscript{i}}> where \textit{t\textsubscript{i}} is the arrival timestamp of the packet and \textit{p\textsubscript{i}} the position of the leftmost 1-bit in the binary hash representation of the associated value.
-Packets in the list are possible maxima over future windows.
+Packets are only kept in the list if they are possible maxima over future windows.
 Because of that, the authors name this list a \textit{"List of Future Possible Maxima (LFPM)"}.
 
-\subsection*{Element Insertion \& Eviction}
+\subsubsection*{Element Insertion \& Eviction}
 Similarly to the original HLL, incoming elements are hashed and the first \textit{n} bits of the string are used to map the new element to one of the \textit{m} buckets.
@@
 
-\subsection*{Sliding Window Cardinality Query}
+\subsubsection*{Sliding Window Cardinality Query}
 
 Estimation of cardinality is done similarly to the original HLL cardinality estimation algorithm.
-To estimate the cardinality at a timestamp \textit{t}, select packets that are in the window --- \textit{i.e.} packets <t\textsubscript{i}, p\textsubscript{i}> from the LFPM where t\textsubscript{i} > \textit{t} - WindowSize.
-This filtering is necessary since the proposed algorithm evicts old packets only at element insertion.
-However, it would be possible to do so when estimating the cardinality or even periodically.
-From the filtered list of packets, compute the highest p\textsubscript{i} among them.
-Do this for each bucket and in the end compute the harmonic mean between all values, just like the original HLL.
+To estimate the cardinality at a timestamp \textit{t}, from all \textit{m} buckets and all \textit{m} lists, select packets that are in the window --- \textit{i.e.} packets <t\textsubscript{i}, p\textsubscript{i}> where t\textsubscript{i} > \textit{t} - WindowSize.
+This filtering is necessary since the proposed algorithm evicts old packets only at element insertion and at the time of the query some packets may no longer belong to the window.
+Next, from the filtered set of packets, for each bucket and its list, compute the maximum p\textsubscript{i} value.
+In the end, compute the harmonic mean between all values, just like the original HLL.
 The result will be the averaged bit-pattern observable value \textit{p} and the cardinality is estimated to be $2^\textit{p}$.
 
-\subsection*{Time complexity analysis}
+\subsubsection*{Time complexity analysis}
 When performing insertions or evictions, the algorithm requires scanning through the Lists of Future Possible Maxima (LFPMs) to delete old packets or packets that will never be window maxima.
@@
 
-The authors prove that \textit{L\textsubscript{n}} has an upper bound of \textit{ln(n/m)}, with \textit{n} as the exact cardinality of the set and \textit{m} as the number of buckets used.
+The authors prove that \textit{L\textsubscript{n}} has an upper bound of \textit{ln(n/m)}, with \textit{n} as the cardinality of the set and \textit{m} as the number of buckets used.
 Hence, in the worst-case scenario, we need to analyze \textit{m ln(n/m)} elements.
-This means that implementing a Sliding HLL using Lists of Future Possible Maxima has \textit{O(ln(n))} time complexity, given the number of buckets \textit{m} is constant.
+This means that a Sliding HLL implementation using Lists of Future Possible Maxima has \textit{O(ln(n))} time complexity, given the number of buckets \textit{m} is constant.
 
-\subsection*{Space complexity analysis}
+\subsubsection*{Space complexity analysis}
 The memory used in the Sliding HLL aggregator implemented with Lists of Future Possible Maxima (LFPMs) amounts to the total size of the \textit{m} lists.
@@
 
-Given \textit{ln(n/m)} as the upper bound for \textit{L\textsubscript{n}}, this sliding implementation of the HLL aggregator consumes up to \textit{(4+1) ln(n/m) m} bytes of memory.
+The lists are dynamic thus they shrink and grow at run time.
+However, \textit{ln(n/m)} was proved to be an upper bound for \textit{L\textsubscript{n}}.
+Hence, this sliding implementation of the HLL aggregator consumes up to \textit{(4+1) ln(n/m) m} bytes of memory.
 Considering that the number of buckets \textit{m} is constant, the space complexity of a Sliding HLL implementation with LFPM is \textit{O(ln(n))}.
-Thus, the growth of the window in size is followed by a logarithmic increase in used memory.
 
@@
 
-\subsection*{Applicability to our Hypothesis}
-Implementing a Sliding HLL using Lists of Future Possible Maxima (LFPMs) shows logarithmic time and space complexity while maintaining the same accuracy as the original HLL approximate aggregator.
-Hence, this is a valid approach to test our Hypothesis.
+\subsubsection*{Applicability to our Hypothesis}
+Implementing a Sliding HLL using Lists of Future Possible Maxima (LFPMs) shows logarithmic time and space complexity while providing estimates with the same accuracy as the original HLL approximate aggregator.
+However, the memory consumption is dynamic due to the variable size of the lists.
+
+Despite the fact memory consumption dynamically changes during runtime, this is a valid approach to test our Hypothesis due to the worst-case scenario being a logarithmic space complexity.
+However, in Section \ref{sec:hll-drv} we analyze an implementation that is not dynamic in memory during run time making it a preferred approach.
+
+\subsubsection{Sliding HyperLogLog using Distance Recorder Vectors} \label{sec:hll-drv}
+
+In Section \ref{sec:hll-lfpm} we have analyzed a Sliding HLL implementation where each of the \textit{m} buckets holds a list of values (LFPM) instead of a single integer count.
+However, the LFPM structure is dynamic and its size varies at run time.
+
+The approach proposed by Xu uses a smaller amount of memory that does not change at run time \cite{Xu-hll-sliding-drv}.
+The author proposes a special counter: a \textit{"Distance Recorder Vector (DRV)"}.
+
+\subsubsection*{Data Structure}
+Much like the original HLL and the LFPM Sliding implementation, the Distance Recorder Vector (DRV) approach uses \textit{m} buckets.
+However, instead of integer counts or lists, each of the buckets stores a DRV.
+
+A DRV is composed by \textit{$\lceil ln(n/m) \rceil$} records, with \textit{n} as the cardinality of the set.
+Notice that \textit{n/m} is the average number of distinct elements each bucket \textit{m} is expected to process and that a binary string of size \textit{$\lceil ln(n/m) \rceil$} can represent \textit{n/m} distinct elements.
+Also recall that the bit-pattern observable \textit{p} used by the HLL is the position of the leftmost 1-bit.
+In a binary string of size \textit{$\lceil ln(n/m) \rceil$}, \textit{p} will range from 1 to \textit{$\lceil ln(n/m) \rceil$}.
+The DRV is indexed exactly this way, from 1 to \textit{$\lceil ln(n/m) \rceil$}, and each bit-pattern observable \textit{p} will be used as an index in this vector of records.
+Essentially, each \textit{p} will have an associated record in a DRV of a bucket.
+
+Initially, all records in the DRV will be set to be infinity or a large number.
+
+\subsubsection*{Element Insertion}
+When an element is inserted, the same procedure as in the original HLL is followed: the element is hashed and \textit{p} is computed from the hashed binary representation.
+Denote the bit-pattern observable of the new element as \textit{p\textsubscript{new}}.
+For the DRV of that bucket, increment the value of all records by one.
+Next, set position \textit{p\textsubscript{new}} in the DRV to 0 (DRV[\textit{p\textsubscript{new}}] = 0).
+
+\subsubsection*{Sliding Window Cardinality Query} 
+Element eviction is not performed.
+Instead, the query algorithm takes into account that some records in some of the DVR's might not be valid anymore.
+To estimate the sliding window cardinality is to find the value \textit{p}.
+
+To find the value \textit{p} for a bucket, scan the DRV of that bucket backward.
+Start at the last element --- positioned at \textit{$\lceil ln(n/m) \rceil$} --- and scan towards the head of the vector.
+For each record, check if it is active in the current window.
+To do so, check if its value is less than the size of the window.
+If it is, the record is active.
+Otherwise, the record is inactive.
+
+If the record is inactive continue scanning backward.
+If the record is active, then the position or index currently being scanned is the value of \textit{p} for that bucket.
+If all records are inactive, the bit-pattern observable \textit{p} is 0.
+
+Stochastic averaging is done just like in the original HLL: compute the harmonic mean of all \textit{p\textsubscript{m}} values from all \textit{m} buckets.
+The cardinality of the sliding window is then estimated to be $2^\textit{p}$.
+
+\subsubsection*{Time complexity analysis}
+When inserting an element, determining its bucket, hashing the element and computing \textit{p} is done in constant time.
+However, all the  \textit{$\lceil ln(n/m) \rceil$} records must have their value increased by one and this means iterating over the DRV.
+Setting the element of the DRV indexed by \textit{p} to 0 is done in constant time as well.
+
+Estimating cardinality is done by scanning the DRV backward until an active record is found. In the worst-case scenario, this means scanning all \textit{$\lceil ln(n/m) \rceil$} records.
+
+Both inserting an element and estimating sliding window cardinality require a scan through \textit{$\lceil ln(n/m) \rceil$} records.
+This gives us a time complexity of \textit{O(\textit{ln(n)})}, given the number of buckets \textit{m} is constant.
+
+\subsubsection*{Space complexity analysis}
+This approach still uses \textit{m} buckets but each one contains a Distance Recorder Vector (DRV).
+A DRV is a vector of \textit{$\lceil ln(n/m) \rceil$} elements.
+Hence, \textit{$m \lceil ln(n/m) \rceil$} elements must be kept in memory, giving us a space complexity of \textit{O(ln(n))} with \textit{n} as the cardinality of the set.
+
+\subsubsection*{Applicability to our Hypothesis}
+Despite having the same time and space complexities, the Sliding HLL DRV approach uses a smaller amount of memory that does not change during runtime, while the LFPM requires heap resizing because the lists dynamically shrink and grow.
+Hence, the DRV approach is a better fit to integrate our lightweight real-time solution.

### 2020-06-05 chapters/background.tex
Test

--- 2020-02-24 chapters/background.tex
@@
 Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times.
-As more and more data became available, the need for stream processing increased.
+As more and more data became available in the data stream format, the need for stream processing increased.
 Stream processing refers to the processing of data streams.
@@
 That is, from an infinite stream, it is possible to fix a left and right boundary in order to obtain a bounded and finite data set.
-This relation is visible in the \ref{fig:unbounded-bounded-superset} diagram from \cite{batch-is-a-special-case-of-streaming}.
+This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from \cite{batch-is-a-special-case-of-streaming}.
 
@@
     \begin{center}
-      \includegraphics[scale=0.7]{figures/streaming-subset-batch.png}
+      \includegraphics[scale=0.5]{figures/streaming-subset-batch.png}
       \caption{Stream processing as a superset of batch processing}
@@
     \begin{center}
-      \includegraphics[scale=0.7]{figures/unbounded-superset-bounded.png}
+      \includegraphics[scale=0.5]{figures/unbounded-superset-bounded.png}
       \caption{Unbounded data streams as a superset of bounded data sets}
@@
 Recalculate-From-Scratch (RFS) is the simplest and most general SWAG algorithm.
-As the name suggests, this algorithm recomputes the aggregation over the entire window for each window update.
+As the name suggests, this algorithm recomputes the aggregation over the entire window for each update.
 For instance, the computation of a sum over a sliding window using RFS means that whenever the window changes all elements are summed.
@@
 
+%\section{Aggregations on data types}
+
+%Some aggregations are only possible on certain data types.
+For instance, computing an average on a set of strings is not possible.
+In this Section we present the data types we will work with and the set of possible aggregations to compute on them.
+
+%We consider four data types: \textit{numericals}, \textit{strings}, \textit{dates} and \textit{categoricals}.
+Numerical variables can be measured.
+When compared to classical programming data types, for example those of the C programming language, numericals comprise variable types \textit{short}, \textit{long}, \textit{int}, \textit{float}, \textit{double} and valid combinations of the previous --- \textit{e.g.} \textit{long int}.
+On the other hand, Strings are sequences of characters.
+The \textit{string} type is the most generic data type in that any other type can be converted to a string.
+Every variable, for example \textit{x}=3, can be turned into a string \textit{s}, for example \textit{s}="3".
+Similarly, a date is a sequence of characters and thus a string.
+However, despite being a string, it has semantics unlike a regular one.
+Knowing the date format, one can extract the day, month and year.
+For that reason we make the distinction between the \textit{string} and \textit{date} data types.
+Lastly, a \textit{categorical} is a variable that takes one of a relatively small and known set of possible values.
+A variable is assigned to a particular category based on its properties.
+The categorical variable is also a sequence of characters and therefore a string variable.
+The main difference between the categorical data type and the string is that we know the set of possible labels the categorical variable might take and this affects the type of computable aggregations.
+
+%On a set of numerical variables, we can compute sums, minima, maxima, averages, standard deviations and counts (of zeros, infinities and others).
+When operating on a set of strings, we can aggregate the collection into a concatenation of strings, a count, a membership query, cardinality.
+
+\section{Exponential Moving Averages}
+Both SOE and RFS require saving the entire window contents. IS THERE A WAY WE CAN EMULATE AN WINDOW....
+
 % \section{Concept drift}

### 2020-06-05 chapters/introduction.tex
Aggregation on data types
Test

@@
 Different use cases process data in different ways for different purposes.
-For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's clients while ensuring that sales run smoothly and that products are delivered to customers' homes.
+For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's clients, while ensuring that sales run smoothly and that products are delivered to customers' homes.
 Streaming services, like YouTube and Netflix, try to guarantee that media content reaches up to millions of users simultaneously.
@@
 However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques.
-A thorough analysis to stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last $10^6$ events, respectively.
+A thorough analysis to stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.
 
@@
 Feedzai offers fraud prevention solutions for the previously presented use case of monitoring an infinite stream of credit card transactions and detecting fraud.
-We will work side by side with them, taking advantage of the multiple real-life data sets from the financial fraud space they can provide.
+We will work with them, taking advantage of the multiple real-life data sets they can provide from the financial fraud space.
 
@@
 
-In Chapter \ref{chap:approach} we define the scope of this project, the research questions we intend to answer, our experimental methodology and our future planning.
+%In Chapter \ref{chap:approach} we define the scope of this project, the research questions we intend to answer, our experimental methodology and our future planning.
 

### 2020-06-08 chapters/introduction.tex
...

@@
 Different use cases process data in different ways for different purposes.
-For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's clients, while ensuring that sales run smoothly and that products are delivered to customers' homes.
-Streaming services, like YouTube and Netflix, try to guarantee that media content reaches up to millions of users simultaneously.
-Furthermore, this media content may have been previously recorded and stored, or may be generated on the fly.
+For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's\footnote{https://alibaba.com} clients, while ensuring that sales run smoothly and that products are delivered to customers' homes.
+Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously.
+This media content may have been previously recorded and stored, or may be generated on the fly.
 Social networks are also responsible for generating large volumes of data.
-A social network like Twitter, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day \cite{twitter-throughput}.
+A social network like Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day \cite{twitter-throughput}.
 \textit{Tweets} may contain text, media content or both.
@@
 The former case is best known as stream processing, while the latter as batch processing.
+
 E-commerce applications have to process endless streams of credit card transactions.
@@
 As mentioned, the processing of data can be divided into two main categories: batch and stream processing.
-However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, stream processing demand is increasing.
+However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, the demand for stream processing systems is increasing.
 Data stream processing refers to the handling of huge volumes of high variety of data generated at high-velocity from multiple sources in real-time.
-Unlike batch processing on static data sets, where assumptions on the underlying data distribution can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery}.
+Unlike batch processing on static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery}.
 Furthermore, in the streaming paradigm, the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}.
@@
 However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques.
-A thorough analysis to stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.
+A thorough analysis of stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.
 
-Consider the case is the detection of credit card fraud on the data stream of all the transactions made on Amazon by every US citizen on a single day.
+Consider the use-case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day.
 In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one.
 Such a system would have to process all the incoming transactions and decide in a fraction of a second if it is a fraudulent one or not.
-It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone.
-Thus justifying the need for autonomous real-time data stream monitoring systems.
+It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.
 
 Many market solutions monitor streams of credit card transactions with the intent of fraud detection.
-In the context of this Thesis, we will work side by side with Feedzai.
+In the context of this Thesis, we will work side by side with Feedzai\footnote{https://feedzai.com}.
 Feedzai offers fraud prevention solutions for the previously presented use case of monitoring an infinite stream of credit card transactions and detecting fraud.

### 2020-06-08 chapters/background.tex
Reviewed Intro and Background. Added PDIS pdf

@@
 Batch processing is the oldest paradigm between batch and stream processing.
-Historically speaking, Apache Hadoop \cite{hadoop} was one of the first widely known technologies for large-scale batch processing, created in 2005.
-Nearly a decade later, Apache Spark \cite{spark} was released.
+Historically speaking, Apache Hadoop\footnote{https://hadoop.apache.org} was one of the first widely known technologies for large-scale batch processing, created in 2005.
+Nearly a decade later, Apache Spark\footnote{https://spark.apache.org} was released.
 Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times.
+
 As more and more data became available in the data stream format, the need for stream processing increased.
@@
 This is in contrast to batch processing which works on a finite data set.
-This distinction is important since some computations are impossible to compute accurately on unbounded data sets.
-Such an example of an impossible computation on an infinite data set is grouping incoming data by an ID and counting distinct elements.
+This distinction is important since some aggregations are impossible to compute with 100\% accuracy in an efficient manner on unbounded data sets.
+An example of such is counting distinct elements that flow through the data stream.
+An immediate solution would be to create a regular set of items or strings and have its length be the number of distinct elements processed.
+However, this approach is not very efficient and has a linear memory usage growth regarding the number of distinct elements.
+Therefore, as the number of distinct elements approaches large orders of magnitude we run out of memory.
 
-In order to provide a streaming solution, Apache Spark created the Spark Streaming module.
+In order to provide a streaming solution, Apache Spark created the Spark Streaming\footnote{https://spark.apache.org/docs/latest/streaming-programming-guide.html} module.
 Being a batch system in its core, Spark implements streaming as micro-batching.
@@
 
-% streaming as a proper superset of batch
 Distinguishing between batch and stream processing is important to understand how a monitoring system on top of the latter would work.
-The Venn diagram \ref{fig:stream-superset} was taken from \cite{batch-is-a-special-case-of-streaming} and illustrates the relation of both processing techniques.
+The Venn diagram \ref{fig:stream-superset} \cite{batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
 A data stream is unbounded or infinite while a bounded data stream has a beginning and an end.
-Using these definitions, the content of an unbounded data stream contains that of a bounded data set.
-That is, from an infinite stream, it is possible to fix a left and right boundary in order to obtain a bounded and finite data set.
-This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from \cite{batch-is-a-special-case-of-streaming}.
+Using these definitions, the concept of an unbounded data stream contains that of a bounded data set.
+That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite data set.
+This relation is visible in Figure \ref{fig:unbounded-bounded-superset} \cite{batch-is-a-special-case-of-streaming}.
 
@@
 
-%\section{Anatomy of data analytics}
-% data stream is composed by, that may exist or not
-%window
-%aggregation
-%group_by
-
 \section{From unbounded streams to finite data sets} \label{sec:windows}
@@
 While in some scenarios it may be, more often than not we would like to introduce some logical context into our computations.
-For instance, consider the monitoring of machines spread out across multiple racks in a large data center.
+For instance, consider the monitoring of machines spread out across multiple racks in a large data center and its temperatures.
 Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream.
@@
 In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for actionable context, in this case identifying a likely-to-fail rack.
-To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end.
+To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}.
 By creating such a conceptual range, we realize that in practice we need a bounded data set.
@@
 
-The authors of \cite{Wang-Windows-Stream-Processing} claim that \textit{"recent history is often managed using windows"} and that \textit{"windowing makes it possible to implement streaming versions of the traditionally blocking relational operators, such as streaming aggregations"}.
-More formally, the authors of \cite{Botan-SECRET} state that \textit{"a window W over a stream S is a finite subset of S"}.
+Gedik et.
+al claim that \textit{"recent history is often managed using windows"} and that \textit{"windowing makes it possible to implement streaming versions of the traditionally blocking relational operators, such as streaming a\cite{Wang-Windows-Stream-Processing}ggregations"} \cite{Wang-Windows-Stream-Processing}.
+More formally, Botan et.
+al define that \textit{"a window W over a stream S is a finite subset of S"} \cite{Botan-SECRET}.
 
@@
 A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events.
-The step size \textit{S} of a windowing technique is formally defined by \cite{Botan-SECRET} as \textit{"the distance between consecutive windows"}.
+The step size \textit{S} of a windowing technique is formally defined as \textit{"the distance between consecutive windows"} \cite{Botan-SECRET}.
 This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away.
@@
 
-For the scope of this Thesis, we will work under a time-based sliding window framework with a unitary step size.
-Despite focusing on time-based true sliding windows, our approach is just as valid for tuple sized windows and for other step values as well.
+For the scope of this Thesis, we will work under a tuple-based sliding window framework with a unitary step size.
+Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and for other step values as well.
 
@@
 Aggregations are present in most programming languages and frameworks.
-Quoting from the JavaSE13 documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}.
-Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one.
+Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}.
+Similarly, Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one.
 \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value.
@@
 The \textit{combine} function takes the current aggregation value and combines it with the next item in the set, returning the new aggregated state.
-Doing this recursively for each item in the collection results in a single value: the aggregate value for the data set.
+Doing this recursively for each item in the collection results in a single value: the aggregate value.
 
@@
 In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it.
-The authors of \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs).
-Considering the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs.
+Tangwongsan et.
+al define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+Consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs.
 How would we compute the average temperature per rack in the data center in the last 5 minutes?
-To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink \cite{flink}.
+To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink\footnote{https://flink.apache.org}.
 Assuming all the temperatures can be queried from the \textit{Temperatures} table and that each rack corresponds a different \textit{rack\_id} we have the following query:
@@
 The total sum of the window will be the aggregation state or value.
-When inserting an element, the updated total window sum equals the total sum plus the new element.
-The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed this far minus the evicted element.
+When inserting an element, the updated total window sum equals the current total sum plus the new element.
+The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus the evicted element.
 With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity.
@@
 The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state.
-For instance, consider the use case of computing a sum aggregation over a window of data.
+For instance, consider once again the use case of computing a sum aggregation over a window of data.
 The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event.

### 2020-06-08 chapters/introduction.tex
Reviewed Intro and Background. Added PDIS pdf

@@
 In the Machine Learning field, the performance decay caused by data pattern shifts is also known as concept drift.
-%UNCOMMENT WHEN BACKGROUND SECTION IS DONE: Such a concept will be further explained in Section \ref{sec:concept-drift}.
 
@@
 
-The objective of this Thesis is to build a lightweight real-time system that detects these shifts in data patterns in systems that handle high volumes of high velocity, highly skewed and seasonal data.
+The objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relativily to a reference period.
 
@@
 Therefore, it is common for companies to have systems monitoring the former, testing their usage rate, availability and performance, just to name a few.
-The motivation behind this Thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in performance.
+The motivation behind this Thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in performance of the system analyzing said data stream.
 
@@
 \section{Hypothesis} \label{sec:hypothesis}
-The question this Thesis aims to answer is whether sliding window aggregations and probabilistic data structures can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data with a low memory footprint.
+%The question this Thesis aims to answer is whether sliding window aggregations and probabilistic data structures can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data with a low memory footprint.
+
+The question this Thesis aims to answer is whether a method that relies on comparing an aggregation of the reference period and the aggregation effectively seen during the stream analysis can be used to detect data pattern shifts in the stream, in the presence of high volumes of high velocity, highly skewed and seasonal data while ensuring a low memory footprint.
 
@@
 \begin{itemize}
-    \item high precision ratio --- \textit{i.e.} at least 80\%
     \item high recall ratio --- \textit{i.e.} at least 80\%
@@
 
-Experiments will be made using multiple real data sets from the financial fraud space and compared to existing batch analysis results.
+Experiments will be made using multiple synthetic and real data sets from the financial fraud space.
 
 \section{Document structure} 
+% TODO at the end rewrite and relink this
 This section merely sums up how each of the following Chapters contributes to this Thesis.
@@
 
-In Chapter \ref{chap:conclusion} we summarize our contributions.
+%In Chapter \ref{chap:conclusion} we summarize our contributions.

### 2020-06-09 chapters/background.tex
Re-iterated document structure
Old Sota backup, new sliding window image style

@@
     \begin{center}
-      \includegraphics[scale=0.3]{figures/sliding.png}
+      \hspace*{0.6in}
+      \includegraphics[scale=0.4]{figures/sliding.png}
       \caption[True sliding window]{True sliding window, W=3, S=1}
@@
 Aggregations are present in most programming languages and frameworks.
-Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}: \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}.
+Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}:
+\textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}.
 Similarly, Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one.
@@
 Both SOE and RFS require saving the entire window contents. IS THERE A WAY WE CAN EMULATE AN WINDOW....
-
-% \section{Concept drift}
-% seasonal and evolutionary
-% static ML models
-% data pattern shifts -> in this context, concept drift
-% data pattern shifts can be formally defined and introduced as concept drift IN THIS PARTICULAR CASE of monitoring a stream that feeds into ML models

### 2020-06-09 chapters/sota.tex
Test
Outlier detection method taxonomy

@@
 
-In this Chapter, we discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time computations over true sliding windows using resource-lightweight approaches on streaming systems.
-We present a wide analysis of several categories of algorithms that seemed like viable options to build the desired system.
+A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}.
+In this Chapter, we discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems.
+We present a wide analysis of several categories of algorithms that seem viable to build the desired system.
 
-\section{Data Mining} \label{sec:change-detect}
-A data stream has a temporal dimension and the underlying process that generates the data stream can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}.
-It is this change in the underlying distribution that we want to detect and report as a data pattern shift.
+\section{Outlier detection in time-series data}
+Blázquez-García et.
+al review state-of-the-art outlier detection techniques in the context of time-series data and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
+From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented in \textit{"A review on outlier/anomaly detection in time-series data"} by Blázquez-García et.
+al due to its focus on time-series data.
+
+The authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}.
+Regarding the concept of \textit{outlier} the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}.
+This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
+generated by a different mechanism"} \cite{Hawkins-Outliers}.
+Furthermore, Blázquez-García et.
+al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"}.
+For the purpose of their review and this Thesis, all the previous terms refer to outliers.
+
+The authors evaluate an outlier detection method on three different axes: the input data type, the outlier type to detect and the uni or multivariate nature of the method.
+Figure \ref{fig:outlier-taxonomy} presented by the authors summarizes these categories.
+
+\begin{figure}[!htb]
+  \begin{center}
+    \includegraphics[scale=0.55]{figures/outlier-method-taxonomy.png}
+    \caption{Outlier detection method taxonomy}
+    \label{fig:outlier-taxonomy}
+  \end{center}
+\end{figure}
+
+The \textit{input data} axis refers to the type of time-series the method uses as input.
+According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one.
+A univariate time-series refers to a time-series consisting of a list of scalars generated by a single time-dependent variable.
+To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}.
+On the other hand, a multivariate time-series has more than one time-dependent variable.
+Most importantly, not only is each variable time-dependant but also may have dependencies on other variables.
+With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}.
+Each \textit{k-dimensional} vector consists of \textit{k}  time-dependent variables, possibly correlated.
+
+The \textit{nature of the method} axis refers to the type of analysis performed by the outlier detection method.
+An outlier detection method may perform a univariate or multivariate analysis.
+A univariate method works only with a single time-dependent variable whereas a multivariate one simultaneously works with more than one variable.
+An outlier detection method can take as input a multivariate time-series but perform a univariate analysis, one per time-series of the multivariate input.
+On the other hand, to perform multivariate analysis, the input data type must be multivariate as well.
+
+The third and final axis is the \textit{outlier type} the method aims to report.
+The authors identify three distinct outlier types: point outliers, subsequence outliers and time-series outliers.
+A point outlier corresponds to a single data point in time that deviates from the expected value.
+These point outliers can be uni or multivariate, depending on the nature of the method.
+A univariate point outlier is a single time-dependent variable that stands out from the rest of the time-series.
+On the other hand, a multivariate point outlier takes into account multiple time-dependent variables in comparison to the remaining time-series.
+Subsequence outliers refer to a sequence of points in time that together show anomalous behavior, but may not be individually considered outliers.
+Similarly to point outliers, a subsequence outlier may take into account a single time-dependent variable or multiple ones.
+The authors classify them as univariate subsequence outliers and multivariate subsequence outliers, respectively.
+Lastly, time-series outliers refer to entire time-series being outliers themselves.
+This type of outliers can only be detected if the input data is multivariate.
 

### 2020-06-09 chapters/sota.tex
Removal of the old poorly written and poorly categoryzed outlier detection methods

@@
 Furthermore, Blázquez-García et.
-al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"}.
-For the purpose of their review and this Thesis, all the previous terms refer to outliers.
+al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers.
+Hence, outlier detection refers to the process of finding anomalies in data.
+An anomaly relates to a point in time where the behavior of the system does not match the expected one.
+Note that under this definition an anomaly does not necessarily imply a problem.
 
@@
 
-\subsection{A two window paradigm algorithm for change detection}
-\label{subsec:2-window}
+Kifer et al. 2 window paradigm \\
+ref window -> some data\\
+sliding window -> current data\\
+apply distance metric (like jsd) to both windows if above threshold alert and refill reference (adapts to new distro)\\
+input data: uni or multi variate\\
+method nature: univariate (doenst check correlation)\\
+outlier type: subsequence outlier (window)\\
 
-Kifer et al.
-propose an algorithm for change detection in data streams \cite{Kifer-Detecting-Change}.
-The authors propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions.
-Thus the change detection algorithm works on a two-window paradigm.
+ExtraN\\
+point outlier?\\
+density based clusters\\
+distance based outliers\\
+predictability/forecast property based\\
 
-Figure \ref{fig:change-detection-2-windows} contains the pseudocode presented by the authors.
-The proposed method uses two windows, a reference \textit{W\textsubscript{1}} and a sliding one \textit{W\textsubscript{2}}.
-The authors use tuple-based windows of size \textit{k} and a true sliding window.
-The reference window \textit{W\textsubscript{2}} works as a \textit{baseline} and contains the first \textit{k} points of the stream that occurred after the last detected change.
+HTM\\
+point outlier, model based, prediction\\
+black box ML algorithm\\
+explain properties\\
+use what authors say in their review\\
 
-The algorithm begins by filling both windows with the first \textit{k} tuples.
-Then it slides window \textit{W\textsubscript{2}} (by one tuple, since it is a true sliding window).
-It does this \textit{k} times and in between slides checks for change by applying the \textit{d} function, which measures the discrepancy between the two windows, \textit{W\textsubscript{1}} and \textit{W\textsubscript{2}}.
-Whenever this discrepancy is above a certain threshold $\alpha$, the algorithm reports a change in the data stream.
-Then it repeats the whole process, re-initializing both windows with the next \textit{k} items and proceeding as already described.
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.8]{figures/2-windows-change-pseudocode.png}
-      \caption{Two window method for data stream change detection}
-      \label{fig:change-detection-2-windows}
-    \end{center}
-\end{figure}
-
-This algorithm reduces change detection in data streams to testing whether two samples --- \textit{reference} and \textit{sliding} windows --- are generated by different distributions.
-The authors study methods for identifying differences in distribution between two samples and implement them as the \textit{d} function.
-Furthermore, they develop their own data structure and algorithm for efficient computation of \textit{d}.
-
-%init - \textit{O(klog(k))}    
-%runtime - \textit{O(klog(k)}
-
-%k to initialize and when change is detected
-%k * d
-%d 
- %nlog(2k) to initialize tree and when change is detected
- %log(2k) to maintain the tree
-\subsubsection*{Time complexity analysis}
-Computing the value of \textit{d} can be done in \textit{O(log(k))} time per iteration. Hence, given \textit{k} iterations, the time complexity of this algorithm is \textit{O(k log(k))}.
-
-\subsubsection*{Space complexity analysis}
-The presented algorithm makes use of two \textit{k} sized windows and a custom data structure used for the efficient computation of \textit{d} of size \textit{k} as well.
-Since \textit{k} is constant, this method has constant space complexity.
-
-\subsubsection*{Applicability to our Hypothesis}
-In this Thesis, we want to detect and alert data pattern shifts so detecting change in data streams is directly applicable to our use case.
-However, our focus is on a lightweight solution and this algorithm has a linear space complexity, so it does not fit our scope as a possible solution.
-
-%\subsection{Pattern Mining}
-
-\subsection{Mining Frequent Patterns: the FP-Stream structure}
-An alternative approach to detecting changes in data streams is to perform pattern mining.
-The idea behind using pattern mining algorithms being that if we can maintain a list of frequent patterns we would be able to detect when said patterns change --- \textit{i.e.} the list changes.
-Changes in the frequent mined patterns would be associated with a data pattern shift and immediately reported.
-
-The authors of \cite{Giannella-Mining-Frequent-Patterns} developed a data structure named \textit{FP-Stream} --- that summarizes frequent data patterns -- and algorithms to build and incrementally maintain it.
-
-We agree with the author's statement that it is \textit{"unrealistic to hold all streaming data in the limited main memory"}.
-With this in mind, they divide patterns into three categories: frequent, sub-frequent and infrequent patterns.
-The focus of their work is the mining and maintenance of frequent and sub-frequent patterns since the latter might become frequent later.
-Thus, infrequent patterns are discarded, using less memory.
-
-The FP-Stream structure consists of a frequent pattern tree (FP-Tree \cite{Han-FP-tree}) with tilted-time windows in each of the nodes.
-
-The authors are not very explicit on the definition of a tilted-time window.
-They claim we are \textit{"often interested in recent changes at a fine granularity, but long term changes at a coarse granularity"} and that tilted-time windows fit such use case.
-Figure \ref{fig:tilted-time-window} shows such a window and was withdrawn from the paper.
-In it, we see the 4 quarters of the last hour, then the last 24 hours and finally 31 days.
-They claim that \textit{"one can compute frequent itemsets in the last hour with the precision of quarter of an hour, the last day with the precision of hour, and so on, until the whole month"}.
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=1]{figures/tilted-time-window.png}
-      \caption{Tilted-time window}
-      \label{fig:tilted-time-window}
-    \end{center}
-\end{figure}
-
-An FP-Tree \cite{Han-FP-tree} as shown in Figure \ref{fig:fptree}, is a tree representation of frequent patterns.
-Each node in the frequent pattern tree represents a pattern and its frequency (\textit{support} column in the Figure), recorded in the node.
-The same tree construction algorithm from \cite{Han-FP-tree} is used in the FP-Stream algorithm.
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=1]{figures/fptree.png}
-      \caption{FP-Tree}
-      \label{fig:fptree}
-    \end{center}
-\end{figure}
-
-The authors propose using only one frequent pattern tree, where at each node, the frequency for each tilted-time window is maintained.
-Figure \ref{fig:fpstream} shows the FP-Stream structure, as an example of a frequent pattern tree with tilted-time windows embedded in each node.
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.8]{figures/fp-stream.png}
-      \caption{FP-Stream}
-      \label{fig:fpstream}
-    \end{center}
-\end{figure}
-
-The algorithm for constructing and maintaining the FP-Stream structure is given as a high-level list of instructions.
-The algorithm groups incoming streaming data into batches.
-Initialization is done only when the first batch is complete.
-As the transactions for the first batch arrived, the frequencies for all items are computed.
-Once all transactions for the first batch have arrived, the batch is scanned to create an FP-tree structure, pruning all infrequent items -- \textit{i.e.} below a certain frequency threshold.
-Finally, an FP-Stream structure is created by mining the frequent items from the FP-Tree.
-
-The incremental update of the FP-Stream as data arrives is described on a very high level and does not add relevant information to understand time and space complexity without knowledge of the FP-Tree algorithm, on which it heavily relies.
-Further research will be done to perform a more thorough analysis of this algorithm, starting with the analysis of \cite{Han-FP-tree}.
-
-\subsubsection*{Applicability to our Hypothesis}
-
-No definitive conclusions can be made without the full time and space complexity analysis.
-However, the maintenance of a frequent pattern tree and tilted-time windows for each of the nodes might reveal to memory intensive, discarding this as a valid solution.
-
-\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm}
-
-The authors of \cite{Yang-Neighbor-Based-Pattern-Detection} propose a method for incremental detection of neighbor based patterns, namely density-based clusters and distance-based outliers, specifically for sliding windows scenarios.
-Figure \ref{fig:plot-cluster-outlier} is an example of the previous two mentioned neighbor based patterns.
-Formal definitions for both distance-based outliers \ref{fig:distance-outlier-def} and density-based clusters \ref{fig:density-cluster-def} are provided.
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.8]{figures/density-distance-based.png}
-      \caption[Clustering and outlier identification plot]{Two density-based clusters and one distance-based outlier determined by neighbour-based pattern detection algorithms}
-      \label{fig:plot-cluster-outlier}
-    \end{center}
-\end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.8]{figures/distance-based-outlier-def.png}
-      \caption{Formal definition of distance-based outliers}
-      \label{fig:distance-outlier-def}
-    \end{center}
-\end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.8]{figures/density-based-cluster-def.png}
-      \caption{Formal definition of density-based clusters}
-      \label{fig:density-cluster-def}
-    \end{center}
-\end{figure}
-
-Density-based clusters label data points as \textit{core points}, \textit{edge points} or \textit{noise}.
-These classifications are explained in \ref{fig:density-cluster-def}.
-Data points are classified based on the number of neighbors they have within a range $\theta\textsuperscript{range}$, measured versus a $\theta\textsuperscript{count}$.
-In brief, if a data point has more than $\theta\textsuperscript{count}$ neighbors within $\theta\textsuperscript{range}$ it is classified as a \textit{core point}.
-If it does not but it is a neighbor of a core point, then it is an \textit{edge point}.
-Otherwise, it is considered \textit{noise}.
-
-The authors make use of the \textit{“predictability"} property of the expiration of existing objects.
-In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point in the window on arrival --- \textit{i.e.} the future windows it will belong to.
-This leads to the notion of \textit{predicted views}.
-Given the objects in a current window, they predict the pattern structures that will persist in subsequent windows by considering the objects in the current window and abstract these predicted pattern structures into \textit{“predicted views"} of each future window.
-This technique allows the efficient maintenance of neighborship counts by pre-handling the eventual expiring effect of new data points.
-
-In the paper, the authors first propose the Abstract-C and Abstract-M algorithms. Ultimately, they combine all the desirable properties of each algorithm into Extra-N.
-
-\subsubsection{Abstract-C}
-Abstract-C was the first proposed algorithm.
-The main idea is to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has rather than the actual list of neighbors.
-The main advantage of this approach is the reduced memory footprint since only a count is stored --- a single integer --- instead of a list of pointers --- to all neighbors.
-
-When a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one --- since \textit{p\textsubscript{i}} no longer exists.
-However, we are unable to reach \textit{p\textsubscript{i}}'s neighbors since in Abstract-C only a count of neighbors is stored --- rather than pointers to each of the neighbors.
-
-The solution to this problem is exploiting the previously presented \textit{predictability} property.
-The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbours.
-The authors introduce the concept of \textit{"lifetime neighbour counts"} or \textit{lt\_cnt}.
-The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbour counts} --- \textit{i.e.} the number of \textit{predicted neighbours} \textit{p\textsubscript{i}} has in any future window where he exists.
-More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s current neighbours that are known to survive in the corresponding future window.
-Furthermore, the length of \textit{p\textsubscript{i}.lt\_cnt} is kept equal to \textit{p\textsubscript{i}.lifespan}, decreasing by one after each window slide, removing the left most entry.
-The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbours, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}.
-Applying the \textit{predictability} property makes it possible to compute the lifespan of all points.
-Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}.
-This implies that at \textit{W\textsubscript{i}}, \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\}.
-
-According to the definition given in \ref{fig:distance-outlier-def}, the lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute distance-based outliers: for each data point \textit{p\textsubscript{i}} check if the first (current) element of \textit{p\textsubscript{i}.lt\_cnt} is less than  \textit{$\theta\textsuperscript{fra}$} $\times$ \textit{N} to decide whether a data point is an outlier or not, with \textit{N} as the number of data points.
-Similarly, the core objects for the density-based clusters can be computed by comparing the first element of \textit{p\textsubscript{i}.lt\_cnt} with \textit{$\theta$\textsuperscript{count}}.
-However, the \textit{lt\_cnt} array does not carry enough information to generate the density-based clusters.
-Despite being able  to find all core points in the window, we can not group them into clusters.
-In order to compute that, Abstract-C analyzes each core point and the surrounding $\theta\textsuperscript{range}$ area in the window to reconstruct the clusters.
-
-In conclusion, Abstract-C achieves linear memory consumption in the number of data points in the window.
-However, since it reconstructs the clusters for each core object in the window, it has \textit{O(CP)} time complexity, with \textit{CP} as the total number of core points.
-Hence, its performance largely depends on  \textit{CP}, which can vary from 0 -- no core points --- to the total window size --- all core points.
-
-\subsubsection{Abstract-M}
-Abstract-M is proposed as an Abstract-C enhancement by introducing the notion of \textit{cluster memberships}, achieved by attributing a \textit{clusterID} to each data point.
-
-The main challenge of this approach is discounting the effect of expired data points.
-The removal of any data point on any cluster may result in the total break of the cluster into multiple smaller ones, which may persist or degrade to noise.
-Furthermore, when a cluster is split, the newly formed clusters must be attributed with new \textit{clusterID}s --- \textit{i.e.} every single data point in those clusters must be relabeled with the new \textit{clusterID}.
-
-Once again, the solution to handle expiring data points relies on the predictability property.
-Knowing the data points that exist in each window means we can also know the set of future clusters and points that belong.
-This is included in the \textit{predicted views} concept previously mentioned.
-In brief, the \textit{predicted clusters} are created using such predicted views for each of the future windows.
-
-Addition of new data points may cause three changes to the clusters: \textit{birth} of a new cluster, \textit{expansion} of an existing one and \textit{merge} of multiple existing clusters.
-When adding a new data point, mark it with a new \textit{clusterID} in case of a \textit{birth} or in an \textit{expansion} case mark it with the \textit{clusterID} of the cluster it belongs to.
-Handling the \textit{merge} of multiple clusters is done by using a tree structure where two or more sibling \textit{clusterID}s are in fact part of a bigger cluster with the \textit{clusterID} stored in their parent.
-For example, if cluster \textit{ID1} and cluster \textit{ID2} are merged into a new cluster \textit{ID3} by the arrival of a new data point, they are inserted as children of the node containing \textit{ID3}.
-
-New data points may join existing neighborhoods of existing data points and may promote them to core points by making the size of their neighborhood greater or equal to \textit{$\theta$\textsuperscript{count}}.
-Once such promotion happens, the promoted core point needs to communicate to its neighbors its new role, since noise in its neighborhood becomes edge points and is given a \textit{clusterID}.
-Since there are no connections from the promoted data point to its neighbors, Abstract-M analyzes each of the promoted core points and the $\theta\textsuperscript{range}$ area around them to find the points requiring state update.
-
-In conclusion, Abstract-M remains linear in memory regarding the number of window data points.
-However, since the number of promoted core points tends to be smaller and is always less or equal to the number of core points, this is an improvement on Abstract-C.
-
-\subsubsection{Extra-N}
-
-Abstract-M still performs range queries (area with $\theta\textsuperscript{range}$ diameter) for each of the promoted core points.
-To improve upon this, Extra-N combines the neighborship maintenance mechanisms proposed in Abstract-C and Abstract-M.
-
-The authors state that different information is required from different classes of data points (\textit{core}, \textit{edge} or \textit{noise}).
-For example, maintaining a list of points to neighbors is required for a non-core point while neighbor counts are sufficient for core points.
-More precisely, Extra-N marks each data point \textit{p\textsubscript{i}} with a \textit{clusterID} in all windows where \textit{p\textsubscript{i}} is predicted to be a core point while keeping the exact list of neighbors (pointers) of \textit{p\textsubscript{i}} for all windows where he is predicted to be noise or an edge point.
-This way, the Extra-N algorithm stores the amount of information needed to compute both distance-based outliers and density-based clusters according to their definitions, \ref{fig:distance-outlier-def} and \ref{fig:density-cluster-def} respectively.
-
-\subsubsection*{Time complexity analysis}
-Extra-N only performs range queries for incoming data points.
-This means that for each new data point \textit{p\textsubscript{i}}, Extra-N will search the area of diameter $\theta\textsuperscript{range}$ around it and update on it neighbors.
-In conclusion, the time complexity of Extra-N is \textit{O(n)} where \textit{n} is the number of new data points.
-
-\subsubsection*{Space complexity analysis}
-In summary, Extra-N keeps the memory consumption linear to the number of data points in the window without performing unnecessary range queries.
-
-\subsubsection*{Applicability to our Hypothesis}
-While a linear complexity fits many use cases, it does not fit ours since we aim to build a low memory footprint solution, ideally with constant space complexity.
-Hence, this algorithm will be of no use to us.
-
-\section{Anomaly Detection}
-Anomaly detection refers to the process of finding anomalies in data.
-An anomaly relates to a point in time where the behavior of the system does not match the expected one.
-Hence, under this definition, an anomaly does not necessarily imply a problem.
-
-With the study of anomaly detection algorithms, we intend to search for algorithms that identify data pattern shifts as anomalies.
-If any of the approaches searched for is not computationally intensive it will be considered valid for the final solution.
-
-\subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
-
-In \cite{Ahmad-HTM}, the authors present an anomaly detection technique based on an on-line algorithm called Hierarchical Temporal Memory (HTM).
-HTM has a few properties that are desirable for data stream anomaly detection scenarios.
-First, it is an unsupervised algorithm that requires no labels.
-This property comes in handy when monitoring a data stream in real-time, where labels for incoming data will likely be absent.
-Additionally, HTM adapts to noisy domains.
-In other words, random variations in the underlying distribution of incoming data that do not persist throughout time will be ignored.
-Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.} changes in magnitudes of data or unusual timing for particular patterns, respectively.
-
-HTM is also an online learning algorithm, meaning that it adapts to changing statistics of the underlying data stream.
-In other words, if there is a sudden random increase in a specific data point attribute, the algorithm will produce an alert.
-However, if that initially thought to be a random spike becomes frequent then the algorithm adapts and assimilates this as "a new reality".
-Hence, it stops classifying it as an anomaly.
-Unfortunately, this property is not desirable in our system.
-Our stream monitoring system must report changes relatively to an initial static configuration and should not accept a new distribution as the new standard.
-
-\subsubsection*{Applicability to our Hypothesis}
-Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again.
-Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made.
-In conclusion, we do not deem HTM applicable to our solution.
-
-\subsection{DeepAnT}
-
-DeepAnT \cite{Munir-DeepAnT} is a deep learning-based approach for the detection of anomalies in time series data, that uses unlabeled data to capture and learn the data distribution of the stream used to forecast the normal behavior of the time series data.
-It is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a data streaming scenario where real-time decisions have to be made before labels are effectively-known.
-It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.
-
-DeepAnT employs a Convolutional Neural Network (CNN) as its \textit{time series predictor} module.
-Before monitoring a data stream, the CNN must be trained.
-The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN.
-Additionally, the data set used to train the model thus not require labels and may up to 5\% of the data set may correspond to anomalous data.
-
-\subsubsection*{Applicability to our Hypothesis}
-
-Similar to the HTM algorithm presented, DeepAnT adapts to the underlying distribution of the data stream.
-In other words, if the data patterns shift from the initial configuration but stabilize, DeepAnT will accept it as it is.
-As mentioned, our goal is to build a monitoring system that alerts data pattern shifts continuously, until they match the initial configuration again.
-Furthermore, DeepAnT uses deep learning which is actually a subset of machine learning and just as inexplicable.
-In summary, we do not deem DeepAnT usable for our use case.
+DeepAnT\\
+point outlier, model based, prediction\\
+black box DL algorithm\\
+explain properties\\
+use what authors say in their review\\
 

### 2020-06-10 chapters/sota.tex
..

@@
 An anomaly relates to a point in time where the behavior of the system does not match the expected one.
-Note that under this definition an anomaly does not necessarily imply a problem.
+Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.
 
@@
 According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one.
-A univariate time-series refers to a time-series consisting of a list of scalars generated by a single time-dependent variable.
+A univariate time-series refers to a time-series consisting of a list of values generated by a single time-dependent variable throughout time.
 To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}.

### 2020-06-10 chapters/sota.tex
Point outlier taxonomy, Partial Subsequence outlier taxonomy

@@
 
+%Probably say we analyze outlier methods and why aggregations? relate to the approaches thought of to solve the problem and the sections
+
 \section{Outlier detection in time-series data}
 Blázquez-García et.
-al review state-of-the-art outlier detection techniques in the context of time-series data and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
+al review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
 From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented in \textit{"A review on outlier/anomaly detection in time-series data"} by Blázquez-García et.
@@
 
-The authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}.
-Regarding the concept of \textit{outlier} the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}.
+A time-series is a collection of data points indexed by time.
+Accordingly, the authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}.
+Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}.
 This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
@@
 On the other hand, a multivariate time-series has more than one time-dependent variable.
-Most importantly, not only is each variable time-dependant but also may have dependencies on other variables.
+Most importantly, each variable is not only time-dependant but may also have dependencies on other variables.
 With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}.
@@
 Lastly, time-series outliers refer to entire time-series being outliers themselves.
-This type of outliers can only be detected if the input data is multivariate.
+This type of outliers can only be detected if the input data is multivariate, so as to be able to compare multiple time-series.
 
-Kifer et al. 2 window paradigm \\
-ref window -> some data\\
-sliding window -> current data\\
-apply distance metric (like jsd) to both windows if above threshold alert and refill reference (adapts to new distro)\\
-input data: uni or multi variate\\
-method nature: univariate (doenst check correlation)\\
-outlier type: subsequence outlier (window)\\
+\subsection*{Univariate point outlier detection}
 
+The authors split point outlier detection methods into uni and multivariate, as seen in Figure \ref{fig:outlier-taxonomy}.
+Furthermore, they split univariate point outlier detection methods into three broad categories: model-based, density-based and histogramming based \ref{fig:point-outlier-taxonomy}.
+
+\begin{figure}[!htb]
+  \begin{center}
+    \includegraphics[scale=1]{figures/taxonomy-point-outlier-uni.png}
+    \caption{Point outlier detection in univariate time-series taxonomy}
+    \label{fig:point-outlier-taxonomy}
+  \end{center}
+\end{figure}
+
+The univariate point outlier detection methods based on models are the most common.
+For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value.
+If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier.
+Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold of $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if $|$x\textsubscript{t}$-$e\textsubscript{t}$|>\alpha$.
+Additionally, the authors split the model-based category into estimation and prediction models.
+The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values.
+Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can't.
+
+Density-based univariate point outlier detection methods rely on clustering of data points and neighbor counting.
+These methods consider a datum an outlier if it has less than $\alpha$ neighbors.
+Distance between two points is usually measured using Euclidean distance.
+Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.
+
+Quoting the authors, histogramming univariate point outlier detection methods are \textit{"based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original"}.
+%TODO: explain this better, paper not clear to me
+
+\iffalse
 ExtraN\\
-point outlier?\\
-density based clusters\\
-distance based outliers\\
+point outlier\\
+density-based outliers\\
 predictability/forecast property based\\
@@
 use what authors say in their review\\
+\fi
+
+\subsection*{Univariate subsequence outlier detection}
+Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones \ref{fig:outlier-taxonomy}.
+Furthermore, the authors split univariate subsequence outlier detection methods into five different axes as seen in Figure \ref{fig:subsequence-outlier-taxonomy}.
+
+\begin{figure}[!htb]
+  \begin{center}
+    \includegraphics[scale=0.8]{figures/taxonomy-subsequence-outlier-uni.png}
+    \caption{Subsequence outlier detection in univariate time-series taxonomy}
+    \label{fig:subsequence-outlier-taxonomy}
+  \end{center}
+\end{figure}
+
+Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series in order to find out the most unusual ones.
+Each subsequence is compared to the others using a distance metric, for instances the Euclidean distance.
+The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in a quadratic time complexity.
+These methods find the most unusual subsequence in a time-series.
+However, as the authors put it, these type of methods \textit{"do not have a reference of normality or a threshold"} and hence they do not accurately report outliers.
+
+Univariate subsequence outlier detection methods based on dissimilarity are identical to the previous discord detection based methods but use a reference of normality.
+The representations used for the subsequences and the reference of normality vary among methods in this class of outlier detection methods but they are based on comparison of the reference or normality subsequence with the ones being analyzed for outlier detection.
+Once again, the metric used to measure the distance between the reference period subsequence and the other ones depends entirely on the method and subsequence aggregation or representation.
+Finally, when the measured distance is above a certain threshold, these methods classify the subsequence as an outlier or an anomaly.
+
+Prediction model-based univariate subsequence outlier detection methods assume that the reference or normality period is composed of past events.
+Similarly to univariate point outlier detection methods based on prediction models, these models train on past data and make predictions of the future.
+Subsequences that do not resemble the prediction made by the model are classified as outliers.
+
+\iffalse
+Kifer et al. 2 window paradigm \\
+ref window -> some data\\
+sliding window -> current data\\
+apply distance metric (like jsd) to both windows if above threshold alert and refill reference (adapts to new distro)\\
+input data: uni or multi variate\\
+method nature: univariate (doenst check correlation)\\
+outlier type: subsequence outlier (window)\\
+\fi
 

### 2020-06-10 chapters/sota.tex
Outlier detection method taxonomy

@@
 
-%Probably say we analyze outlier methods and why aggregations? relate to the approaches thought of to solve the problem and the sections
+%Probably say we analyze outlier methods and why aggregations? relate to the approaches thought to solve the problem and the sections
 
@@
 On the other hand, a multivariate time-series has more than one time-dependent variable.
-Most importantly, each variable is not only time-dependant but may also have dependencies on other variables.
+Most importantly, each variable is not only time-dependent but may also have dependencies on other variables.
 With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}.
@@
 If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier.
-Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold of $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if $|$x\textsubscript{t}$-$e\textsubscript{t}$|>\alpha$.
+Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if
+\begin{equation*}
+    |x\textsubscript{t}-e\textsubscript{t}|>\alpha
+\end{equation*}
 Additionally, the authors split the model-based category into estimation and prediction models.
@@
 
-Density-based univariate point outlier detection methods rely on clustering of data points and neighbor counting.
+Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting.
 These methods consider a datum an outlier if it has less than $\alpha$ neighbors.
@@
 HTM\\
-point outlier, model based, prediction\\
-black box ML algorithm\\
+point outlier, model-based, prediction\\
+black-box ML algorithm\\
 explain properties\\
@@
 DeepAnT\\
-point outlier, model based, prediction\\
+point outlier, model-based, prediction\\
 black box DL algorithm\\
@@
 \subsection*{Univariate subsequence outlier detection}
-Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones \ref{fig:outlier-taxonomy}.
+Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones (Figure \ref{fig:outlier-taxonomy}).
 Furthermore, the authors split univariate subsequence outlier detection methods into five different axes as seen in Figure \ref{fig:subsequence-outlier-taxonomy}.
@@
 
-Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series in order to find out the most unusual ones.
-Each subsequence is compared to the others using a distance metric, for instances the Euclidean distance.
-The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in a quadratic time complexity.
+Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series to find out the most unusual ones.
+Each subsequence is compared to the others using a distance metric, for instance, the Euclidean distance.
+The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in quadratic time complexity.
 These methods find the most unusual subsequence in a time-series.
-However, as the authors put it, these type of methods \textit{"do not have a reference of normality or a threshold"} and hence they do not accurately report outliers.
+However, as the authors put it, these types of methods \textit{"do not have a reference of normality or a threshold"} and hence they do not accurately report outliers.
 
-Univariate subsequence outlier detection methods based on dissimilarity are identical to the previous discord detection based methods but use a reference of normality.
+On the other hand, univariate subsequence outlier detection methods based on dissimilarity are identical to the previous discord detection based methods but use a reference of normality.
 The representations used for the subsequences and the reference of normality vary among methods in this class of outlier detection methods but they are based on comparison of the reference or normality subsequence with the ones being analyzed for outlier detection.
@@
 Subsequences that do not resemble the prediction made by the model are classified as outliers.
+
+Frequency-based univariate subsequence outlier detection methods also use a reference period and compute the frequency of each subsequence in the reference or normality period.
+Hence, a subsequence will be an outlier if it does not appear as frequently as expected in subsequent time periods.
+
+Lastly, we have information theory based methods for univariate subsequence outlier detection, which according to the authors \textit{"are closely related to the frequency-based methods"}.
+This type of method not only measures the frequency of subsequences but relates it with the information carried by the subsequence.
+Mathematically, a subsequence \textit{S} will be an outlier if
+
+\begin{equation*}
+    I(S) \times F(S) > \alpha
+\end{equation*}
+
+with \textit{F(S)} as the frequency of subsequence S, \textit{I(S)} as the information carried by S and $\alpha$ as a given threshold.
+According to the authors, \textit{"I(S) is computed taking into account the number of times the symbols within S are repeated through the series"}.
+The authors explain that if \textit{F(S)} is large --- \textit{i.e.} S occurs frequently --- then \textit{I(S)} will be closer to 0.
 

### 2020-06-10 chapters/background.tex
Footnotes

@@
 Distinguishing between batch and stream processing is important to understand how a monitoring system on top of the latter would work.
-The Venn diagram \ref{fig:stream-superset} \cite{batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
+The Venn diagram \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
 A data stream is unbounded or infinite while a bounded data stream has a beginning and an end.
@@
 That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite data set.
-This relation is visible in Figure \ref{fig:unbounded-bounded-superset} \cite{batch-is-a-special-case-of-streaming}.
+This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.
 

### 2020-06-10 chapters/introduction.tex
Footnotes

@@
 Social networks are also responsible for generating large volumes of data.
-A social network like Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day \cite{twitter-throughput}.
+A social network like Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{$https://blog.twitter.com/engineering/en_us/a/2013/new-tweets-per-second-record-and-how.html$}.
 \textit{Tweets} may contain text, media content or both.

### 2020-06-11 chapters/sota.tex
Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm

@@
 al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers.
-Hence, outlier detection refers to the process of finding anomalies in data.
-An anomaly relates to a point in time where the behavior of the system does not match the expected one.
+In summary, outlier detection refers to the process of finding anomalies in data.
+An anomaly (or outlier) relates to a point in time where the behavior of the system does not match the expected one.
 Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.
@@
 
-\iffalse
-ExtraN\\
-point outlier\\
-density-based outliers\\
-predictability/forecast property based\\
+\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm}
 
+Yang et.
+al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios.
+The authors propose Extra-N, an algorithm that detects point outliers based on the number of neighbors it has.
+The authors make use of the \textit{“predictability"} property of the expiration of existing objects.
+In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to.
+This technique allows the efficient maintenance of neighborship counts by pre-handling the eventual expiring effect of new data points.
+
+Ideally we want to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has rather than the actual list of neighbors.
+The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors.
+The problem is that when a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one but we are unable to reach \textit{p\textsubscript{i}}'s neighbors without a list of neighbor pointers.
+The solution to this problem is exploiting the previously presented \textit{predictability} property.
+The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors.
+The authors introduce the concept of \textit{"lifetime neighbor counts"} or \textit{lt\_cnt}.
+The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where he exists.
+More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window.
+The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}.
+Applying the \textit{predictability} property makes it possible to compute the lifespan of all points.
+Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}.
+This implies that at \textit{W\textsubscript{i}},
+\[ \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\} \]
+The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.
+
+\subsubsection*{Time complexity analysis}
+
+To find outliers, Extra-N has to do a linear search through the list of points and check the neighbor counts.
+Hence, for outlier detection, this method shows \textit{O(n)} complexity with \textit{n} as the number of points.
+Additionally, for each new data point \textit{p\textsubscript{i}}, Extra-N has to perform a range query search centered on it to find its neighbors and increment their lifetime neighbor counts by one.
+The run-time complexity of a range query search is \textit{O(log(n))} \cite{range-query-log} with \textit{n} as the number of data points.
+
+In summary, Extra-N shows logarithmic time complexity when updating the sliding window and linear time complexity when looking for outliers, both relative to the window size.
+
+\subsubsection*{Space complexity analysis}
+Extra-N has to keep a lifetime neighbor count dictionary-like object for each window data point.
+For that reason, we claim that Extra-N achieves linear memory consumption relative to the number of data points in the window.
+
+\subsubsection*{Applicability to our Hypothesis}
+Extra-N memory consumption grows linearly in regards to window size, so it does not fit our desired lightweight system architecture.
+Furthermore, updating the sliding window should be done in constant time, which is not the case for Extra-N.
+Hence, this algorithm will be of no use to us.
+
+\subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
 HTM\\
@@
 
+\subsection{DeepAnT}
 DeepAnT\\
@@
 use what authors say in their review\\
-\fi
 
@@
 The authors explain that if \textit{F(S)} is large --- \textit{i.e.} S occurs frequently --- then \textit{I(S)} will be closer to 0.
+
+\subsection{A two window paradigm algorithm for change detection}
 

### 2020-06-11 chapters/sota.tex
Removed unecessary files

@@
 \subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
+HTM
+\iffalse
 HTM\\
@@
 use what authors say in their review\\
-
+\fi
 \subsection{DeepAnT}
+DeepAnT
+\iffalse
 DeepAnT\\
@@
 use what authors say in their review\\
+\fi
 

### 2020-06-11 chapters/conclusions.tex
Document structure

\chapter{Conclusion} \label{chap:conclusion} \minitoc

\section{Hypothesis Summary}
\section{Answer to the hypothesis}
\section{Contributions}
\section{Future work}
\section{Final remarks}

### 2020-06-11 chapters/my-work.tex
Document structure

\chapter{Lightweight Real-Time Feature Monitoring} \label{chap:my-work} \minitoc

%what we want (reiterate)
%A reference aggregation snapshot to compare with a streaming aggregation that must be kept in real-time and constant in memory usage

\section*{Outlier detection methods}
%using data mining outlier stuff, too heavy

\section*{Sliding window aggregations}
%- build a aggregation snapshot of the reference and compare with streaming aggregations
 %       - challenges: 
  %          generic SWAG algorithms grow linearly in space regarding window size
   %         defining the set of aggregations that can be done at feature level that represent the current state well enough for comparison and change detection
    %        defining the threshold of deviation between the reference aggregation values and current streaming aggregations to raise alerts
     %   - mention challenge 1 can be solved with Probabilistic Data Structures and EMAs
      %  - there is no trivial solution for the others
       %     - present an example (2 time-series same max min median and std. but different)

\section{Numerical feature monitoring: a two-phased method}

%- introduce method as two-phased method (batch and streaming)

\subsection{Final goal}
%- what is our goal in the streaming phase?
           % - alert when streaming fields/features/variables' distribution changes considerably
          %  - need to choose an aggregation that allows us to know the distribution of variables -> histogram for each feature
               % - lightweight -> approximated EMA-based histogram (recall EMAs background), each bin count is an EMA-count aggregation
               % - describe an EMA approximated histogram and how it discounts old events, the simulated window size, ...
           % - we can compute an exact histogram over the reference period because it is done in batch, where resources like RAM or time are not heavily constrained in constrast to streaming systems
           % - given 2 histograms (ref hist and stream hist) we can apply distance metrics (such as a JSD)
             %   - worth mentioning other metrics can be used (Wasserstein distance, Kolmogorov–Smirnov, ...)
         %   - but how to threshold this "distance" value? ----> we need to know the distribution of "distance" values
             %   - how is this accomplished
             %   - agressive sampling within ref period of smaller periods (equal in size to the stream simulated window)
              %  - give a practical use-case where it makes a diference:
                 %   - left-skewed distro vs right-skewed distro ---> threshold is very different (95 or 99 percentile for example)

\subsection{Batch analysis phase}

%- batch phase: build reference exact histogram and distribution of distance values
          %  1. compute ref histogram (for each feature)
                %- equal heights/counts per bin, different size bins ---> better adjusted to wildly varying distributions
         %   2. make samples of length equal to streaming window size, build histograms for each using bins from reference histogram to make later accurate comparisons between ref hist and sample hist
             %   - why multiple sample histograms?
                   % encode distributions over smaller time periods since there may be pattern changes within the large reference period
             %   - why an approximated histogram for the samples as well despite being done in batch?
                 %   mimic the streaming approximated histogram and the test that will be performed:
                     %   - batch: sample app. histogram vs ref hist
                    %    - stream: streaming app. histogram vs ref hist
         %   3. compute the distance between each sample's histogram and the reference one (for each feature)
          %  4. we now have a list of distance values, and know their expected distribution (for each feature)
              %  - this is important since a test value obtained measuring the distance between the reference period and other window
                %can be compared to this distribution (again probably give the left and right skewed example)

          %  By the end of the batch phase we have:
               % for each feature:
                    %reference histogram (for later comparison with streaming histogram)
                    %list/histogram of distance values
                    %last sample's histogram ---> used as burn-in period for target histogram

\subsection{Streaming phase}

%- online phase:
          %  for each feature:
            %    1. initialize the streaming histogram as the last sample's histogram from the batch phase (burn-in period)

         %   for each event:
               % for each field/feature:
                  %  1. update the feature's streaming histogram (recall the update method)
                  %  2. compute the distance between the reference histogram and the target histogram
                   % 3. test the percentile the distance value fits according to the feature's distances list and alert if above a certain percentile (say the 99th)

           % introduce the multiple testing correction problem:
              %  "the more inferences are made, the more likely erroneous inferences are to occur"
             %   we are making multiple hypothesis testing when computing a distance and checking the percentile for each feature

                %introduce the Holm-Bonferroni multiple test correction method

            %So in reality we have:

          %  for each event:
               % for each field/feature:
                 %   1. update the feature's streaming histogram (recall the update method)
                  %  2. compute the distance between the reference histogram and the target histogram
                  %  3. apply holm-bonferroni for all feature's distances:
                     %   - each percentile will be a p-value
                    %    - order features by p-value
                    %    - check if p-value > FWER / (m + 1 - k)
                          %  if so then feature is divergent

\subsection{Summary}

### 2020-06-11 chapters/problem-statement.tex
Document structure

\chapter{Problem Statement} \label{chap:statement} \minitoc

%Detecting changes in distribution between a reference/static period and the observed data flowing through a stream

%Time constraints (real-time)

%Memory constraints (lightweight)

### 2020-06-11 chapters/sota.tex
Document structure

@@
 
-\subsubsection*{Time complexity analysis}
+\subsubsection*{Method classification}
+Extra-N can take both uni or multivariate time-series as input data.
+However, it does not relate multiple time-series and does not take into account possible dependencies between different time-dependent variables.
+In that sense, and according to the presented taxonomy (Figure \ref{fig:outlier-taxonomy}), the method is of univariate nature.
+Last but not least, Extra-N aims to detect univariate point outliers.
+These point outliers are identified based on the density of the region or neighborhood they belong too.
+Hence, we say Extra-N is a density-based method (Figure \ref{fig:point-outlier-taxonomy}).
+
+\subsubsection*{Time complexity}
 
@@
 
-\subsubsection*{Space complexity analysis}
+\subsubsection*{Space complexity}
 Extra-N has to keep a lifetime neighbor count dictionary-like object for each window data point.
@@
 \subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
-HTM
+\subsubsection*{Method classification}
+\subsubsection*{Applicability to our Hypothesis}
+
 \iffalse
@@
 \subsection{DeepAnT}
-DeepAnT
+\subsubsection*{Method classification}
+\subsubsection*{Applicability to our Hypothesis}
+
 \iffalse
@@
 \subsection{A two window paradigm algorithm for change detection}
+\subsubsection*{Method classification}
+\subsubsection*{Time complexity}
+\subsubsection*{Space complexity}
+\subsubsection*{Applicability to our Hypothesis}
 

### 2020-06-11 chapters/validation.tex
Document structure

\chapter{Method Validation} \label{chap:validation} \minitoc

\section{Datasets description}
\subsection{Synthetic datasets}
\subsection{Financial fraud datasets}

\section{Tests description}
\subsection{Single feature}
\subsection{Multi feature}

\section{Tests results}
\subsection{Single feature}
\subsection{Multi feature}

\section{Conclusions}

### 2020-06-12 chapters/abstract.tex
HyperLogLog. PDIS delivery.
Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory (HTM)

@@
 
-Data pattern shift detection has been researched within diverse areas and application domains and refers to the process of finding patterns in data that do not conform to expected or usual behavior.
+%Data pattern shift detection has been researched within diverse areas and application domains and refers to the process of finding patterns in data that do not conform to expected or usual behavior.
 Accurate and timely detection of data pattern deviations allows for immediate measures to be taken, preventing loss of income and maintaining the companies' reputation among customers.
-As such, many large companies take this into consideration and deploy data anomaly components in their systems.
+%As such, many large companies take this into consideration and deploy data anomaly components in their systems.
 In this way, system administrators can proactively adjust and prevent potential service failures.
 
-For example, Yahoo's anomaly detection system fulfills its purpose of monitoring and raising alerts on time-series events for several of Yahoo's use cases.
+%For example, Yahoo's anomaly detection system fulfills its purpose of monitoring and raising alerts on time-series events for several of Yahoo's use cases.
 At Microsoft, their custom, internally made anomaly detection system monitors millions of metrics coming from a variety of different internal services, enabling engineers to solve live on-site issues before they scale and become bigger ones.
 
-Many real-time stream monitoring systems are static once deployed in a production environment.
+%Many real-time stream monitoring systems are static once deployed in a production environment.
 Engineers of said systems usually configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data.
@@
 
-With this thesis, we set out to design and evaluate a system that makes use of lightweight streaming analytical methods to detect deviations in data patterns while in the presence of large volumes of high velocity, highly skewed, seasonal data.
+%With this thesis, we set out to design and evaluate a system that makes use of lightweight streaming analytical methods to detect deviations in data patterns while in the presence of large volumes of high velocity, highly skewed, seasonal data.
 
 \vspace*{10mm}\noindent
-\textbf{Keywords}: real-time, streaming, lightweight, monitoring, volume, velocity, variety, seasonal
+\textbf{Keywords}: real-time, streaming, lightweight, monitoring, volume, velocity, seasonal

### 2020-06-12 chapters/sota.tex
Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory (HTM)

@@
 
-\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm}
+\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm} \label{alg:extran}
 
@@
 However, it does not relate multiple time-series and does not take into account possible dependencies between different time-dependent variables.
-In that sense, and according to the presented taxonomy (Figure \ref{fig:outlier-taxonomy}), the method is of univariate nature.
+In that sense, and according to the presented taxonomy (Figure \ref{fig:outlier-taxonomy}), the method accepts uni or multivariate time-series as input but is of univariate nature.
 Last but not least, Extra-N aims to detect univariate point outliers.
 These point outliers are identified based on the density of the region or neighborhood they belong too.
-Hence, we say Extra-N is a density-based method (Figure \ref{fig:point-outlier-taxonomy}).
+Hence, we say Extra-N is a density-based univariate point outlier detection method (Figure \ref{fig:point-outlier-taxonomy}).
 
@@
 \subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
+Ahmad \& Purdy, 2016, present an anomaly detection technique based on an online algorithm called Hierarchical Temporal Memory (HTM)\footnote{https://github.com/numenta/nupic} \cite{Ahmad-HTM}.
+HTM has a few properties that are desirable for data stream anomaly detection scenarios.
+First, it is an unsupervised algorithm that requires no labels.
+This property comes in handy when monitoring a data stream in real-time, where labels for incoming data will likely be absent.
+Additionally, HTM adapts to noisy domains.
+In other words, random variations in the underlying distribution of incoming data that do not persist throughout time will be ignored.
+Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.} changes in magnitudes of data or unusual timing for particular patterns, respectively.
+
+HTM is an online learning algorithm, meaning that it adapts to changing statistics of the underlying data stream.
+In other words, if there is a sudden abrupt change in the distribution of a specific time-dependent variable, the algorithm will raise an alert.
+However, if that initially thought to be random spike becomes frequent, the algorithm adapts and assimilates this as "a new reality" and stops classifying it as an anomaly.
+Unfortunately, this property is not desirable in our system.
+Our stream monitoring system must report changes relatively to an initial static configuration (reference or normality period) and should not accept a new distribution as the new standard.
+
 \subsubsection*{Method classification}
+Similarly to Extra-N presented in Section \ref{alg:extran}, the Hierarchical Temporal Memory (HTM) algorithm accepts both uni and multivariate time-series as input data but does not perform a multivariate analysis.
+Hence, we say it accepts both uni and multivariate inputs but is univariate in nature.
+HTM focuses on finding point outliers, but unlike Extra-N, which is density-based, HTM is a model-based approach.
+Specifically, HTM is a prediction model-based approach for univariate point outlier detection.
+
 \subsubsection*{Applicability to our Hypothesis}
+Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again.
+Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made.
+In conclusion, we do not deem HTM applicable to our solution.
 
-\iffalse
-HTM\\
-point outlier, model-based, prediction\\
-black-box ML algorithm\\
-explain properties\\
-use what authors say in their review\\
-\fi
 \subsection{DeepAnT}

### 2020-06-12 chapters/sota.tex
DeepAnT: A Deep Learning Approach for Unsupervised Anomaly Detection in Time Series

@@
 
-\subsection*{Univariate point outlier detection}
+\subsection{Univariate point outlier detection}
 
@@
 
-\subsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm} \label{alg:extran}
+\subsubsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm} \label{alg:extran}
 
@@
 
-\subsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
-Ahmad \& Purdy, 2016, present an anomaly detection technique based on an online algorithm called Hierarchical Temporal Memory (HTM)\footnote{https://github.com/numenta/nupic} \cite{Ahmad-HTM}.
+\subsubsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
+Ahmad \& Purdy present an anomaly detection technique based on an online algorithm called Hierarchical Temporal Memory (HTM)\footnote{https://github.com/numenta/nupic} \cite{Ahmad-HTM}.
 HTM has a few properties that are desirable for data stream anomaly detection scenarios.
@@
 HTM focuses on finding point outliers, but unlike Extra-N, which is density-based, HTM is a model-based approach.
-Specifically, HTM is a prediction model-based approach for univariate point outlier detection.
+Specifically, HTM is a prediction model-based approach for univariate point outlier detection (Figures \ref{fig:outlier-taxonomy} and \ref{fig:point-outlier-taxonomy}).
 
@@
 
-\subsection{DeepAnT}
+\subsubsection{DeepAnT}
+Munir et.
+al present DeepAnT, a deep learning-based approach for the detection of anomalies in time series data \cite{Munir-DeepAnT}, that uses unlabeled data to capture and learn the distribution of the data stream used to forecast the normal behavior of the time-series.
+It is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known.
+It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.
+
+DeepAnT employs a Convolutional Neural Network (CNN) as its time-series predictor module.
+Before monitoring a data stream, the CNN model must be trained.
+The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN.
+Additionally, the data set used to train the model thus not require labels.
+
 \subsubsection*{Method classification}
+DeepAnT takes as input a uni or multivariate time-series but is a univariate method by nature --- \textit{i.e.} it does not find correlations between multiple time-dependent variables.
+The model used by DeepAnT is a Convolutional Neural Network (CNN), used to make predictions on the future time-series values and measure differences with the actual observed values.
+Hence, DeepAnT is a prediction model-based univariate point outlier detection method (Figures \ref{fig:outlier-taxonomy} and \ref{fig:point-outlier-taxonomy}).
+
 \subsubsection*{Applicability to our Hypothesis}
+Similar to the HTM algorithm presented, DeepAnT adapts to the underlying distribution of the data stream.
+In other words, if the data patterns shift from the initial configuration but stabilize, DeepAnT will accept it as it is.
+As mentioned, our goal is to build a monitoring system that alerts data pattern shifts continuously, until they match the initial configuration again.
+Furthermore, DeepAnT uses deep learning which is actually a subset of machine learning and just as inexplicable.
+In summary, we do not deem DeepAnT usable for our use case.
 
-\iffalse
-DeepAnT\\
-point outlier, model-based, prediction\\
-black box DL algorithm\\
-explain properties\\
-use what authors say in their review\\
-\fi
-
-\subsection*{Univariate subsequence outlier detection}
+\subsection{Univariate subsequence outlier detection}
 Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones (Figure \ref{fig:outlier-taxonomy}).
@@
 
-\subsection{A two window paradigm algorithm for change detection}
+\subsubsection{A two window paradigm algorithm for change detection}
 \subsubsection*{Method classification}

### 2020-06-12 chapters/sota.tex
Kifer et. al - A two window paradigm algorithm for change detection

@@
 Quoting the authors, histogramming univariate point outlier detection methods are \textit{"based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original"}.
-%TODO: explain this better, paper not clear to me
+
+\textcolor{red}{TODO: explain this better, paper not clear to me}
 
@@
 \subsubsection{A two window paradigm algorithm for change detection}
+Kifer et al.
+propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions \cite{Kifer-Detecting-Change}.
+The proposed method uses two windows, a reference \textit{W\textsubscript{1}} and a sliding one \textit{W\textsubscript{2}}.
+Both windows are tuple-based, have size \textit{k} and the sliding window is a true sliding one (recall Section \ref{sec:windows} and the expression \ref{math:windows}).
+The reference window \textit{W\textsubscript{2}} works as a \textit{reference of normality} and contains the first \textit{k} points of the stream that occurred after the last detected change.
+
+The algorithm begins by filling both windows with the first \textit{k} tuples.
+Then it slides window \textit{W\textsubscript{2}} by one tuple, since it is a true sliding window, for each incoming event from the data stream.
+In between slides, the algorithm checks for change by applying a distance function \textit{d} taking as input both the reference and the sliding windows, which measures the discrepancy between both.
+Whenever this distance is above a certain threshold $\alpha$, the algorithm raises an alert.
+After raising an alert, both windows are re-initialized with the next \textit{k} items.
+
+Note that after a change is detected, the reference window is updated with the next \textit{k} tuples from the stream.
+As we have stated, we want a method that does not adapt to the data stream changes and instead keeps raising alerts until the sliding window data distribution returns to the one seen in the reference period.
+It is trivial to adapt this method to always use the same reference period.
+To do so, we simply keep the original reference window and don't update it after each alert.
+
+This algorithm reduces change detection in data streams to testing whether two samples --- \textit{reference} and \textit{sliding} windows --- are generated by different distributions.
+The authors study methods for identifying differences in distribution between two samples and implement them as the \textit{d} function.
+
+\textcolor{red}{\textbf{TODO: develop this idea, they talk about Kolmogorov–Smirnov and others, maybe relate to model monitoring paper, talk about JSD, Wasserstein, so on so forth}}
+
 \subsubsection*{Method classification}
+The method devised by Kifer et.
+al can take uni or multivariate input (if we use one pair of reference/target windows for each time-series) but performs a univariate analysis as it does not find correlations between time-dependent variables.
+This two-windowed method measures the difference between both windows, the reference and the sliding one.
+Furthermore, instead of detecting point outliers, the alerts produced indicate the entire window (or subsequence) shows anomalous behavior.
+Hence, according to the taxonomy presented in Figure \ref{fig:subsequence-outlier-taxonomy}, this method is a dissimilarity-based univariate subsequence outlier detection method.
+
 \subsubsection*{Time complexity}
+The authors show that it is possible to compute the distance between both windows in logarithmic time regarding window size \textit{k}.
+Because the entire sliding window is held in memory, a window update is tantamount to removing the head of the list (element eviction) or appending a new element to the tail (element insertion), which is done in constant time.
+Thus, for each new event, the method evicts the oldest element, inserts the new one and computes the distance between the reference and sliding windows.
+In big O notation, the largest and determining factor is the computation of the distance function, done in \textit{O(log(k))} time, with \textit{k} as the size of both windows.
+
 \subsubsection*{Space complexity}
+The method makes use of two \textit{k}-sized windows, thus storing \textit{2k} elements. Hence, memory consumption grows linearly with window size.
+
 \subsubsection*{Applicability to our Hypothesis}
-
-\iffalse
-Kifer et al. 2 window paradigm \\
-ref window -> some data\\
-sliding window -> current data\\
-apply distance metric (like jsd) to both windows if above threshold alert and refill reference (adapts to new distro)\\
-input data: uni or multi variate\\
-method nature: univariate (doenst check correlation)\\
-outlier type: subsequence outlier (window)\\
-\fi
+In this Thesis, we want to detect and alert data pattern shifts with a low memory footprint.
+However, the method proposed by Kifer et.
+al grows linearly with window size, making it unfit for large windows, our use-case.
 

### 2020-06-12 chapters/background.tex
Removed unecessary files
Kifer et. al - A two window paradigm algorithm for change detection
Pseudocodes for background

@@
       \hspace*{0.6in}
-      \includegraphics[scale=0.4]{figures/sliding.png}
+      \includegraphics[scale=0.35]{figures/sliding.png}
       \caption[True sliding window]{True sliding window, W=3, S=1}
@@
 Figure \ref{fig:stepping-window} represents a stepping tuple-based window of size \textit{W} of three and a stepping size \textit{S} of two.
+
+\textcolor{red}{TODO change all window drawings to a not horrendous style}
 
@@
 \begin{equation} 
-  Window \,\, of \,\, size \,\, W, \,\, step \,\, S =
+\label{math:windows}
+  Sliding\ window \,\, of \,\, size \,\, W, \,\, step \,\, S =
     \begin{cases}
@@
 Doing this recursively for each item in the collection results in a single value: the aggregate value.
+The combine function is here on forward represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.
 
@@
 Two of the most basic SWAG algorithms are \textit{Recalculate-From-Scratch (RFS)} and \textit{Subtract-On-Evict (SOE)}.
-
 Recalculate-From-Scratch (RFS) is the simplest and most general SWAG algorithm.
@@
 However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing.
-Figure \ref{fig:pseudo-rfs} shows pseudocode for RFS \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+Recall that the combine function is represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.
+Pseudocode for the Recalculate-From-Scratch algorithm is shown in Algorithm \ref{pseudo:rfs}.
 
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.45]{figures/rfs.png}
-      \caption{Recalculate-From-Scratch algorithm}
-      \label{fig:pseudo-rfs}
-    \end{center}
-\end{figure}
+\begin{algorithm}
+    \caption{Recalculate-From-Scratch insert, evict and query methods}
+    \label{pseudo:rfs}
+    \begin{algorithmic}[1]
+        \Function{query}{vals}
+            \State $agg\gets\overline{\theta}$
+            \ForEach {$v \in vals$}
+                \State $agg \gets agg \oplus v$
+            \EndFor
+            \State \Return $agg$
+        \EndFunction
+
+        \Function{insert}{vals, v}
+
+            \State $vals.pushBack(v)$
+        \EndFunction
+
+        \Function{evict}{vals}
+
+            \State $vals.popFront()$
+        \EndFunction
+    \end{algorithmic}
+\end{algorithm}
 
@@
 SOE relies on expiring the evicted elements from the aggregation state.
-Hence, SOE only works for \textit{invertible} aggregations.
+Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$.
 To illustrate, consider the computation of a sum aggregation over a sliding window.
 The total sum of the window will be the aggregation state or value.
-When inserting an element, the updated total window sum equals the current total sum plus the new element.
-The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus the evicted element.
+When inserting an element, the updated total window sum equals the current total sum plus ($\oplus$) the new element.
+The sum aggregation is considered \textit{invertible} because, when evicting an element, the new aggregation state is the total sum computed thus far minus ($\ominus$) the evicted element.
 With window size of \textit{n}, SOE has \textit{O(n)} space complexity and \textit{O(1)} time complexity.
 However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable.
-Figure \ref{fig:pseudo-soe} shows pseudocode for SOE \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+Algorithm \ref{pseudo:soe} shows pseudocode for SOE.
 
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.45]{figures/soe.png}
-      \caption{Subtract-On-Evict algorithm}
-      \label{fig:pseudo-soe}
-    \end{center}
-\end{figure}
+Subtract-On-Evict is an improvement over the Recalculate-From-Scratch algorithm based on the idea of aggregation invertibility.
+The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.
 
-Subtract-On-Evict is an improvement over the Recalculate-From-Scratch algorithm that  based on the idea of aggregation invertibility.
-The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.
+\begin{algorithm}
+    \caption{Subtract-On-Evict insert, evict and query methods}
+    \label{pseudo:soe}
+    \begin{algorithmic}[1]
+        \Function{query}{}
+
+            \State \Return $agg$
+        \EndFunction
+
+        \Function{insert}{vals, v}
+
+            \State $vals.pushBack(v)$
+            \State $agg \gets agg \oplus v$
+        \EndFunction
+
+        \Function{evict}{vals}
+
+            \State $v \gets vals.popFront()$
+            \State $agg \gets agg \ominus val$
+        \EndFunction
+    \end{algorithmic}
+\end{algorithm}
 
@@
 
-Tangwongsan et al. construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfied properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+As it turns out, besides \textit{invertibility}, there are other properties that allow us to group aggregations in different families.
+Tangwongsan et al.
+construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfied properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
 
-%TODO: change the sketch-like from the aggregator to the aggregation: bloom filter to membership query, CMS to frequency estimation and HLL to cardinality estimation
+\textcolor{red}{TODO: change the sketch-like from the aggregator to the aggregation: bloom filter to membership query, CMS to frequency estimation and HLL to cardinality estimation}
+
 \begin{table}[!htb]
@@
 \section{Exponential Moving Averages}
-Both SOE and RFS require saving the entire window contents. IS THERE A WAY WE CAN EMULATE AN WINDOW....
+
+\textcolor{red}{TODO: explain EMAs, give mathematical proofs on error bounds, emulated window size based on nhalf}

### 2020-06-12 chapters/introduction.tex
Pseudocodes for background

--- 2020-06-08 chapters/introduction.tex
@@
 Social networks are also responsible for generating large volumes of data.
-A social network like Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day \cite{twitter-throughput}.
+A social network like Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
 \textit{Tweets} may contain text, media content or both.
@@
 \section{Hypothesis} \label{sec:hypothesis}
-%The question this Thesis aims to answer is whether sliding window aggregations and probabilistic data structures can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data with a low memory footprint.
+The question this Thesis aims to answer is whether sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.
 
-The question this Thesis aims to answer is whether a method that relies on comparing an aggregation of the reference period and the aggregation effectively seen during the stream analysis can be used to detect data pattern shifts in the stream, in the presence of high volumes of high velocity, highly skewed and seasonal data while ensuring a low memory footprint.
+\section{Methodology} \label{sec:methodology}
 
-\section{Hypothesis Validation} \label{sec:validation}
+\textcolor{red}{TODO how to define success criteria without "reverse engineering" this?}
+
 The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
@@
 \section{Document structure}
-% TODO at the end rewrite and relink this
+
+\textcolor{red}{TODO at the end rewrite and relink this}
+
 This section merely sums up how each of the following Chapters contributes to this Thesis.

### 2020-06-12 chapters/sota.tex
More pseudocode

@@
 Each element in each of the stacks is a pair \textit{<val, agg>} where \textit{val} is the item's value and \textit{agg} is the aggregation value of everything below it on the stack.
-The pseudocode for Two-Stacks can be seen in Figure \ref{fig:pseudo-2-stacks}, with $\oplus$ as the \textit{combine} function and $\overline{\theta}$ as the identity value of the \textit{combine} function, such that $\overline{\theta} \oplus x$ = $x \oplus \overline{\theta}$ = x.
+The pseudocode for Two-Stacks can be seen in Algorithm \ref{pseudo:2-stacks}, with $\oplus$ as the \textit{combine} function and $\overline{\theta}$ as the identity value of the \textit{combine} function, such that $\overline{\theta} \oplus x$ = $x \oplus \overline{\theta}$ = x.
 For example, if $\oplus$ is the arithmetic sum then $\overline{\theta}$ is 0.
@@
 
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.45]{figures/2-stacks.png}
-      \caption{Two-Stacks algorithm}
-      \label{fig:pseudo-2-stacks}
-    \end{center}
-\end{figure}
+\begin{algorithm}
+    \caption{Two-Stacks insert, evict and query methods}
+    \label{pseudo:2-stacks}
+    \begin{algorithmic}[1]
+        \Function{getStackAggregationValue}{stack}
+         \If{stack.isEmpty()} 
+            \State \Return{$\overline{\theta}$} 
+        \Else 
+            \State \Return{stack.top().agg} 
+        \EndIf
+        \EndFunction
+
+        \Function{query}{F, B}
+            \State \Return{getStackAggregationValue(F) $\oplus$ getStackAggregationValue(B)}
+        \EndFunction
+
+        \Function{insert}{B, v}
+            \State{B.push(v, getStackAggregationValue(B) $\oplus$ v}
+        \EndFunction
+
+        \Function{evict}{F, B}
+
+            \If{F.isEmpty()} \Comment{// Flip B onto F and update aggregation value}
+                \While {\textbf{not} B.isEmpty()}
+                    \State{F.push(B.top().val, B.top().val $\oplus$ getStackAggregationValue(F))}
+                    \State{B.pop()}
+                \EndWhile
+            \EndIf
+            \State{F.pop()}
+        \EndFunction
+    \end{algorithmic}
+\end{algorithm}
 
@@
 
-\textbf{TODO: Add table with SWAG algorithms, time and space complexity and applicable restrictions on the aggregation}
+\textcolor{red}{TODO: Add table with SWAG algorithms, time and space complexity and applicable restrictions on the aggregation}
 

### 2020-06-12 chapters/sota.tex
DABA pseudocode

@@
         \Function{insert}{B, v}
-            \State{B.push(v, getStackAggregationValue(B) $\oplus$ v}
+            \State{B.push( (v, getStackAggregationValue(B) $\oplus$ v) )}
         \EndFunction
@@
 
-            \If{F.isEmpty()} \Comment{// Flip B onto F and update aggregation value}
+            \If{F.isEmpty()} \Comment{// Flip B onto F and update top aggregation value}
                 \While {\textbf{not} B.isEmpty()}
@@
 
-DABA uses two queues, \textit{vals} and \textit{aggs}, both of size \textit{n}, seen in Figure \ref{fig:daba-ds}.
+DABA uses two queues, \textit{vals} and \textit{aggs}, both of size \textit{n}, seen in Figure \ref{fig:daba-ds}, which was directly taken from \textit{"Low-Latency Sliding-Window Aggregation in Worst-Case Constant Time"} by Tangwongsan et.
+al.
 Both queues share six pointers --- \textit{F, L, R, A, B} and \textit{E} --- that point to the same position for each queue.
@@
 
-Figure \ref{fig:pseudo-daba} shows the pseudocode presented by the authors for the DABA algorithm.
+Algorithm \ref{pseudo:daba} shows the pseudocode presented by the authors for the DABA algorithm.
 In reality, after each insertion and eviction, besides the procedure already described, function \textit{fixup} is executed.
 The \textit{fixup} function performs the incremental reversal of \textit{l\textsubscript{B}} to \textit{l\textsubscript{F}}.
-The pseudocode \ref{fig:pseudo-daba} shows how such reversal is done but we will not elaborate on such procedure.
+Algorithms \ref{pseudo:daba-helpers} and \ref{pseudo:daba} show how such reversal is done but we will not elaborate on such procedure.
 However, it is important to stress out that this is the core idea of DABA: avoid occasional expensive operations, like Two-Stacks' \textit{flip}, by spreading them out across run time, in the form of DABA's \textit{fixup}.
 
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.7]{figures/daba-pseudocode.png}
-      \caption{DABA algorithm}
-      \label{fig:pseudo-daba}
-    \end{center}
-\end{figure}
+\begin{algorithm}
+    \caption{DABA helper functions}
+    \label{pseudo:daba-helpers}
+    \begin{algorithmic}[1]
+        \Function{getListFAggregationValue}{}
+            \If{F == B}
+                \State \Return{$\overline{\theta}$}
+            \Else
+                \State \Return{aggs[F]}
+            \EndIf
+        \EndFunction
+
+        \Function{getListBAggregationValue}{}
+            \If{B == E}
+                \State \Return{$\overline{\theta}$}
+            \Else
+                \State \Return{aggs[E - 1]}
+            \EndIf
+        \EndFunction
+
+        \Function{getListLAggregationValue}{}
+            \If{L == R}
+                \State \Return{$\overline{\theta}$}
+            \Else
+                \State \Return{aggs[L]}
+            \EndIf
+        \EndFunction
+
+        \Function{getListRAggregationValue}{}
+            \If{R == A}
+                \State \Return{$\overline{\theta}$}
+            \Else
+                \State \Return{aggs[A - 1]}
+            \EndIf
+        \EndFunction
+
+        \Function{getListAAggregationValue}{}
+            \If{A == B}
+                \State \Return{$\overline{\theta}$}
+            \Else
+                \State \Return{aggs[A]}
+            \EndIf
+        \EndFunction
+
+    \end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+    \caption{DABA insert, evict and query methods}
+    \label{pseudo:daba}
+    \begin{algorithmic}[1]
+        \Function{query}{}
+            \State \Return{getListFAggregationValue() $\oplus$ getListBAggregationValue()}
+        \EndFunction
+
+        \Function{insert}{v}
+            \State vals.pushBack(v)
+            \State aggs.pushBack(getListBAggregationValue() $\oplus$ v)
+            \State fixup()
+        \EndFunction
+
+        \Function{evict}{}
+            \State vals.popFront()
+            \State aggs.popFront()
+            \State fixup()
+        \EndFunction
+
+        \Function{fixup}{}
+        \If{F == B}
+            \State $B \gets E$
+            \State $A \gets E$
+            \State $R \gets E$
+            \State $L \gets E$
+        \Else
+            \If{L == B}
+                \State $L \gets F$
+                \State $A \gets E$
+                \State $B \gets E$
+            \EndIf
+            \If{L == R}
+                \State $A \gets A + 1$
+                \State $R \gets R + 1$
+                \State $L \gets L + 1$
+            \Else
+                \State $aggs[L] \gets getListLAggregationValue() \oplus getListRAggregationValue() \oplus getListAAggregationValue()$
+                \State $L \gets L + 1$
+                \State $aggs[A - 1] \gets vals[A - 1] \oplus getListAAggregationValue()$
+                \State $A \gets A - 1$
+            \EndIf
+        \EndIf
+        \EndFunction
+    \end{algorithmic}
+\end{algorithm}
 
@@
 Even though DABA improves on the time complexity of Two-Stacks, it still shows linear space complexity regarding window size.
-In our monitoring system, we need to analyze large windows of data, requiring at least logarithmic space complexity, thus making DABA inapplicable for our Hypothesis.
+In our monitoring system, we need to analyze large windows of data, thus making DABA inapplicable for our Hypothesis.
 

### 2020-06-13 chapters/sota.tex
...

@@
 \subsection{Two-Stacks}
-The Two-Stacks SWAG algorithm works for \textit{associative} aggregations in scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.} elements are evicted in the same order they were inserted.
+The Two-Stacks SWAG algorithm works for \textit{associative} aggregations (recall Table \ref{tbl:aggregations-properties}) in scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.} elements are evicted in the same order they were inserted.
 Two-Stacks \cite{Tangwongsan-DABA} implements the FIFO queue using two stacks, namely the front stack F and the back stack B.
@@
         \Function{insert}{B, v}
-            \State{B.push( (v, getStackAggregationValue(B) $\oplus$ v) )}
+            \State{B.push( \{val: v, agg: getStackAggregationValue(B) $\oplus$ v\} }
         \EndFunction
@@
                 \While {\textbf{not} B.isEmpty()}
-                    \State{F.push(B.top().val, B.top().val $\oplus$ getStackAggregationValue(F))}
+                    \State{F.push( \{val: B.top().val, agg: B.top().val $\oplus$ getStackAggregationValue(F) \})}
                     \State{B.pop()}
@@
     \begin{center}
+       \hspace{0.5in}
       \includegraphics[scale=0.6]{figures/hll-ds.png}
@@
     \begin{center}
-      \includegraphics[scale=1]{figures/hll-insertion.png}
+      \includegraphics[scale=0.8]{figures/hll-insertion.png}
       \caption[HyperLogLog insertion]{HyperLogLog \textit{el} insertion with \textit{m}=4 and \textit{n}=2 bits}
@@
 Hence, the DRV approach is a better fit to integrate our lightweight real-time solution.
+
+\textcolor{red}{TODO Find and add versions of Sliding Bloom Filter and Sliding CMS}

### 2020-06-14 chapters/my-work.tex
Proposed approaches before proposed method

@@
 
+Our goal is to detect data pattern shifts in an online fashion.
+Such shifts will be measured between a reference or normality period versus the observed reality during the streaming phase.
+Note that we need not to directly compare the entire window contents of the reference period with the contents of the sliding window.
+Doing this would mean our memory consumption grew linearly with window or reference period size.
+We want some sort of aggregation over the reference period that can be maintained efficiently throughout run-time and can be used to distinguish both periods of time and alert for changes.
+In short, we want a reference aggregation snapshot to compare with a streaming aggregation that must be updated in real-time and constant in memory usage.
+
+\section{Approach A: Usage of outlier detection methods}
+The first approach considered relied on outlier detection methods such as the ones presented in Section \ref{sec:outliers} and classified by the taxonomy in Figure \ref{fig:outlier-taxonomy}.
+This straightforward approach works with either subsequence or point outlier detection methods to analyze the streaming sliding window and produce data pattern shift alerts.
+
+\subsection*{How}
+
+Using point outlier detection methods, the main idea would be to keep track of the rate of point outliers detected.
+For instance, keep track of how many outliers were alerted in the past hour.
+The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$.
+On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier.
+
+\subsection*{Challenges}
+The analyzed outlier detection algorithms in Section \ref{sec:outliers} were deemed unfit for our use-case due to one or more of the following: \textit{(a)} high memory consumption, \textit{(b)} high time complexity, \textit{(c)} "memory loss" and \textit{(d)} obfuscated insights.
+
+In our lightweight real-time solution we need constant time complexity to process events one by one in real-time.
+Additionally, because our system is not mission-critical, we want a solution as lightweight as possible, which means something that grows linearly with window size is not a good fit.
+Therefore, most of the outlier detection methods presented become unfit for our use due to high memory consumption and time complexity.
+Some of the methods shown also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period.
+In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream.
+While in some scenarios this is a much-needed feature, in our use-case it is not.
+How clear the alerts and insights retrieved from the methods are is also an important property.
+Some outlier detection methods used Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.
+
+\subsection*{Conclusion}
+We concluded that employing outlier detection methods was not the best course of action and put aside this approach.
+
+\section{Approach B: Thresholding aggregation values}
+This second approach focus on building an aggregation snapshot of the reference period to compare with streaming aggregation values and alert if the difference between both surpasses a threshold $\alpha$.
+
+\subsection*{How}
+
+In this approach we compute sliding window aggregations and compare the aggregations’ current state versus an initial reference, reporting a shift when the difference is substantial.
+For instance, consider the case where we monitor one single feature or variable \textit{x}.
+Assume we define the set of aggregations to compute the mean of \textit{x} and its standard deviation.
+Also assume we save the reference window and maintain a sliding window and the average and standard deviation aggregations for \textit{x} efficiently.
+As seen in Figure \ref{fig:approach2-initial-state}, the initial mean and standard deviation values for \textit{x} presented in the reference window are of 5 and 1, respectively.
+Later on, in \textit{Window 1}, we see the same values for the mean and standard deviation, of 5 and 1, respectively.
+This represents a \textit{normal} case where no alert is produced.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.65]{figures/approach2-normality.png}
+      \caption[]{Normal state for feature \textit{x} in window 1}
+      \label{fig:approach2-initial-state}
+    \end{center}
+\end{figure}
+
+In Figure \ref{fig:approach2-alert-state}, after further sliding steps of our sliding window, we obtain \textit{Window 2}.
+In \textit{Window 2}, the mean and standard deviation aggregation values change from 5 and 1 to 12 and 6, respectively.
+In this case, we would measure the difference between the sliding window aggregations and the reference window ones and if above a certain threshold raise an alert.
+For example, if we used a threshold of $\alpha$ = 6, we would raise an alert, as at least one sliding aggregation, in this case the mean, would differ of at least $\alpha$ when compared to the reference value.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.65]{figures/approach2-alert.png}
+      \caption[]{Alert state for feature \textit{x} in window 2}
+      \label{fig:approach2-alert-state}
+    \end{center}
+\end{figure}
+
+\subsection*{Challenges}
+The first issue with this approach is that generic sliding window aggregation algorithms grow linearly in space regarding window size, such as Recalculate-From-Scratch and Subtract-On-Evict presented in Section \ref{sec:back-swag-algs}, Two-Stacks has seen in Section \ref{sec:2stacks} and DABA in Section \ref{sec:daba}.
+Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.
+Yet another challenge would be to define the maximum deviation threshold $\alpha$  between the reference aggregation state and the streaming aggregation state to produce alerts.
+
+Exponential Moving Averages (Section \ref{sec:emas}), Probabilistic Data Structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}) can be used to solve the first issue and reduce the linear memory complexity of the system relative to sliding window size.
+
+However, there is no trivial solution for the second and third challenges.
+Defining the maximum allowed threshold between reference and sliding window aggregations does not have a straightforward solution.
+Furthermore, defining the set of aggregations to use is also a big challenge.
+To illustrate, consider both time-series show in Figure \ref{fig:approach2-timeseries}.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.65]{figures/approach2-timeseries.png}
+      \caption[]{Alert state for feature \textit{x} in window 2}
+      \label{fig:approach2-timeseries}
+    \end{center}
+\end{figure}
+
+Both time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}.
+Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two clearly different time-series.
+
+\subsection*{Conclusion}
+This approach requires efficient sliding window aggregation state maintenance, which can be done using sliding window probabilistic data structures and/or exponential moving averages.
+However, defining the set of aggregations to use and the deviation threshold itself ultimately led us to discard this approach.
+
+\section{Proposed method: Numerical Feature Monitoring}
 %what we want (reiterate)
@@
 
-\section*{Outlier detection methods}
-%using data mining outlier stuff, too heavy
-
-\section*{Sliding window aggregations}
-%- build a aggregation snapshot of the reference and compare with streaming aggregations
- %       - challenges: 
-  %          generic SWAG algorithms grow linearly in space regarding window size
-   %         defining the set of aggregations that can be done at feature level that represent the current state well enough for comparison and change detection
-    %        defining the threshold of deviation between the reference aggregation values and current streaming aggregations to raise alerts
-     %   - mention challenge 1 can be solved with Probabilistic Data Structures and EMAs
-      %  - there is no trivial solution for the others
-       %     - present an example (2 time-series same max min median and std. but different)
-
-\section{Numerical feature monitoring: a two-phased method}
-
 %- introduce method as two-phased method (batch and streaming)
 
-\subsection{Final goal}
+% only for numerical features for now
+
+\subsection{Stream monitoring goals}
 %- what is our goal in the streaming phase?

### 2020-06-14 chapters/problem-statement.tex
...
Proposed approaches before proposed method

@@
 
-%Detecting changes in distribution between a reference/static period and the observed data flowing through a stream
+With this Chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.
 
-%Time constraints (real-time)
+\section{State of the art issues}
+Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques.
+In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods.
+However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the below characteristics:
 
-%Memory constraints (lightweight)
+\begin{itemize}
+    \item \textbf{Low memory consumption:} need for a low memory footprint system that avoids linear growth relative to the sliding window size used
+
+    \item \textbf{Low time complexity:} need for a time efficient method that updates the sliding window aggregation and raises alerts in constant time
+
+    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout run-time, the opposite of an online learning algorithm
+
+    \item \textbf{Sliding window maintainability:} possibility to evict old events and insert new ones while updating the aggregation state
+
+    \item \textbf{Explainable alerts:} understand why each alert was raised
+\end{itemize}
+
+\section{Research Questions}
+The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time.
+
+To that end, we devise a series of research questions to be answered by the end of this Thesis:
+
+\begin{enumerate}
+    \item \textbf{RQ1:}
+    \item \textbf{RQ2:}
+    \item \textbf{RQ3:}
+\end{enumerate}

### 2020-06-14 chapters/sota.tex
Proposed approaches before proposed method

@@
 
-\section{Outlier detection in time-series data}
+\section{Outlier detection in time-series data} \label{sec:outliers}
 Blázquez-García et.
@@
 
-\subsection{Two-Stacks}
+\subsection{Two-Stacks} \label{sec:2stacks}
 The Two-Stacks SWAG algorithm works for \textit{associative} aggregations (recall Table \ref{tbl:aggregations-properties}) in scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.} elements are evicted in the same order they were inserted.
@@
 
-\section{Sliding Window Aggregations with Probabilistic Data Structures}
+\section{Sliding Window Aggregations with Probabilistic Data Structures} \label{sec:sliding-pds}
 In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space.

### 2020-06-15 chapters/background.tex
Proposed approaches before proposed method
Aggregation properties and families

@@
 Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations.
-However, this creates an artificial barrier that does not really exist in the streaming context.
+However, this creates an artificial barrier that does not truly exist in the streaming context.
 More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues.
@@
 For the scope of this Thesis, we will work under a tuple-based sliding window framework with a unitary step size.
-Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and for other step values as well.
+Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.
 
@@
 
-As it turns out, besides \textit{invertibility}, there are other properties that allow us to group aggregations in different families.
+As it turns out, besides \textit{invertibility}, other properties allow us to group aggregations in different families.
 Tangwongsan et al.
-construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfied properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
 
-\textcolor{red}{TODO: change the sketch-like from the aggregator to the aggregation: bloom filter to membership query, CMS to frequency estimation and HLL to cardinality estimation}
-
-\begin{table}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/aggregation-operations-properties.png}
-      \caption[Aggregation properties]{Aggregation properties.
-Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
-      \label{tbl:aggregations-properties}
-    \end{center}
-\end{table}
-
-To understand these properties, we must be familiar with the \textit{combine} and \textit{lower} functions.
+To understand these properties, we must be familiar with the \textit{combine} and \textit{query} functions.
 The already mentioned \textit{combine} function merges the incoming event with the incrementally computed aggregation state.
-The \textit{lower} function is responsible for returning a result for a given query according to the current aggregation state.
+The \textit{query} function is responsible for returning a result for a given query according to the current aggregation state.
 For instance, consider once again the use case of computing a sum aggregation over a window of data.
 The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event.
-The \textit{lower} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.
+The \textit{query} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.
 
-Based on the \textit{combine} and \textit{lower} functions, the following are formal definitions of these properties. An aggregation is said to have:
+Based on the \textit{combine} and \textit{query} functions, the following are formal definitions of these properties. An aggregation is said to have:
 
@@
 
-    \item  a \textbf{\textit{unary lower}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
+    \item  a \textbf{\textit{unary query}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
 \end{enumerate}
+
+\begin{table}[!htb]
+\begin{center}
+     \begin{tabular}{p{2cm}|p{1.5cm}p{1.5cm}p{1.5cm}p{1.5cm}p{1.6cm}|}
+                 & \rotatebox{35}{\textbf{invertible combine}} & \rotatebox{35}{\textbf{associative combine}} & \rotatebox{35}{\textbf{commutative combine}} & \rotatebox{35}{\textbf{size-preserving combine}} & \rotatebox{35}{\textbf{unary query}}
+                 \\
+                 \hline
+    \textbf{sum-like}     & \checkmark & \checkmark               & \checkmark               & \checkmark                   & \checkmark       \\
+    \textbf{collect-like} & \checkmark              & \checkmark               & $\times$               & $\times$                   & ?           \\
+    \textbf{median-like}  & \checkmark              & \checkmark               & \checkmark               & $\times$                   & ?           \\
+    \textbf{max-like}     & $\times$              & \checkmark               & ?                   & \checkmark                   & \checkmark       \\
+    \textbf{sketch-like}  & $\times$              & \checkmark               & \checkmark               & \checkmark                   & $\times$      
+    \end{tabular}
+    \caption[Aggregation properties]{Aggregation properties.
+Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
+  \label{tbl:aggregations-properties}
+    \end{center}
+\end{table}
+
+According to these properties, Tangwongsan et al.
+define five aggregation families \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+These are the sum, collect, median, max and sketch-like families.
+The sum-like family of aggregations comprises aggregations that have an invertible, associative, commutative and size-preserving \textit{combine} as well as a unary \textit{query} function.
+Examples of such aggregations are the sum, count, mean/average and standard deviation operations.
+The collect-like family of aggregations has an invertible and associative \textit{combine} function but their \textit{query} function is not always unary.
+For example, keeping a string concatenation aggregation is invertible and associative but not commutative nor size-preserving.
+While in the string concatenation case you do not need to pass any arguments to the \textit{query} function, in the case of another collect-like aggregation, for instance, keeping track of the \textit{i-th} youngest element, you need to specify in your query the parameter \textit{i} (\textit{e.g.} the 5th youngest element).
+Hence, their \textit{query} function is not always unary.
+The median-like family of aggregations contains for example the percentile operation.
+For example, the percentile aggregation is invertible, associative and commutative but not size-preserving and requires an argument to the \textit{query} function: the percentile to compute (\textit{e.g.} percentile 95 or 99).
+The max-like family comprises aggregations like the maximum or minimum, operations that are not invertible but associative and size-preserving, as well as unary from a \textit{query} function perspective.
+The final family, the sketch-like family, comprises aggregations that are usually solved by approximate aggregators, also called probabilistic data structures or \textit{sketches}.
+These aggregations can be membership queries, item frequency estimation or cardinality estimation.
 
@@
 Similarly, a date is a sequence of characters and thus a string.
-However, despite being a string, it has semantics unlike a regular one.
+However, despite being a string, it has semantics, unlike a regular one.
 Knowing the date format, one can extract the day, month and year.
@@
 
-\section{Exponential Moving Averages}
+\section{Exponential Moving Averages} \label{sec:emas}
 

### 2020-06-15 chapters/introduction.tex
Aggregation properties and families

@@
 
-The objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relativily to a reference period.
+The objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.
 
@@
 
-\textcolor{red}{TODO how to define success criteria without "reverse engineering" this?}
-
 The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
 \begin{itemize}
-    \item high recall ratio --- \textit{i.e.} at least 80\%
-    \item low latency --- \textit{i.e.} constant time complexity, \textit{O(1)}
-    \item memory efficiency --- \textit{i.e.} constant memory complexity, \textit{O(1)}
+    \item low memory footprint
+
+    \item low latency to work in real-time
+
+    \item fix a reference period as the normal state of the system
+
+    \item explainable alerts
 \end{itemize}
@@
 
-In Chapter \ref{chap:background} of this document we brief the reader with the necessary background and relevant concepts needed to understand the following Chapters.
+%In Chapter \ref{chap:background} of this document we brief the reader with the necessary background and relevant concepts needed to understand the following Chapters.
 
-In Chapter \ref{chap:sota} we examine the state of the art in the area, analyzing the most recent work done and classifying it into subgroups.
+%In Chapter \ref{chap:sota} we examine the state of the art in the area, analyzing the most recent work done and classifying it into subgroups.
 

### 2020-06-15 chapters/my-work.tex
Aggregation properties and families

@@
 \chapter{Lightweight Real-Time Feature Monitoring} \label{chap:my-work} \minitoc
-
-Our goal is to detect data pattern shifts in an online fashion.
-Such shifts will be measured between a reference or normality period versus the observed reality during the streaming phase.
-Note that we need not to directly compare the entire window contents of the reference period with the contents of the sliding window.
-Doing this would mean our memory consumption grew linearly with window or reference period size.
-We want some sort of aggregation over the reference period that can be maintained efficiently throughout run-time and can be used to distinguish both periods of time and alert for changes.
-In short, we want a reference aggregation snapshot to compare with a streaming aggregation that must be updated in real-time and constant in memory usage.
 
@@
     \begin{center}
-      \includegraphics[scale=0.65]{figures/approach2-timeseries.png}
-      \caption[]{Alert state for feature \textit{x} in window 2}
+      \includegraphics[scale=0.7]{figures/approach2-timeseries.png}
+      \caption[]{Two different time-series with same max, min, mean and standard deviation}
       \label{fig:approach2-timeseries}
@@
 \section{Proposed method: Numerical Feature Monitoring}
-%what we want (reiterate)
-%A reference aggregation snapshot to compare with a streaming aggregation that must be kept in real-time and constant in memory usage
 
-%- introduce method as two-phased method (batch and streaming)
+Our goal is to detect data pattern shifts in an online fashion.
+Such shifts will be measured between a reference or normality period versus the observed reality during the streaming phase.
+Note that we need not to directly compare the entire window contents of the reference period with the contents of the sliding window.
+Doing this would require storing all window contents hence making our memory consumption grow linearly with sliding window or reference period size.
+We want some sort of aggregation over the reference period that can be maintained efficiently throughout run-time and can be used to distinguish both periods of time and alert for changes.
+In short, we want a reference aggregation snapshot to compare with a streaming aggregation that must be updated in real-time and constant in memory usage.
 
-% only for numerical features for now
+To that end, we devise a two-phased method.
+The first phase relies on a batch analysis that builds our aggregated reference snapshot to be later on compared with the sliding one.
+The second phase focus on efficiently updating the sliding window aggregation snapshot and occasionally perform a similarity test between the static reference aggregation and the sliding one, alerting for changes if need be.
+The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types as well, as we will discuss.
 
 \subsection{Stream monitoring goals}
-%- what is our goal in the streaming phase?
-           % - alert when streaming fields/features/variables' distribution changes considerably
-          %  - need to choose an aggregation that allows us to know the distribution of variables -> histogram for each feature
-               % - lightweight -> approximated EMA-based histogram (recall EMAs background), each bin count is an EMA-count aggregation
-               % - describe an EMA approximated histogram and how it discounts old events, the simulated window size, ...
-           % - we can compute an exact histogram over the reference period because it is done in batch, where resources like RAM or time are not heavily constrained in constrast to streaming systems
-           % - given 2 histograms (ref hist and stream hist) we can apply distance metrics (such as a JSD)
-             %   - worth mentioning other metrics can be used (Wasserstein distance, Kolmogorov–Smirnov, ...)
-         %   - but how to threshold this "distance" value? ----> we need to know the distribution of "distance" values
-             %   - how is this accomplished
-             %   - agressive sampling within ref period of smaller periods (equal in size to the stream simulated window)
+Before diving into the batch and streaming algorithms we must further define our goal.
+In this Section we pose the question: \textit{"what is our goal during the streaming or monitoring phase?"} and answer it.
+
+A data stream is an infinite stream of events.
+Each event is timestamped and contains multiple fields.
+In the context of this Thesis, each of the event's fields are also named features or time-dependent variables.
+You can think of them as columns in a traditional relational database table.
+Given this context, our goal is to produce alerts of the following form:
+
+\[\textit{timestamp: [(x1, 0.23), (x2, 0.51), (x3, 0.55)}]\]
+
+The format used allows us to know the \textit{timestamp} for the given alert, which features diverge (in this case \textit{x1, x2, x3}) and a value representative of how much each feature diverges (for instance, a float value between 0 and 1 that works as a confidence value).
+
+An alert should be raised if a feature distribution changes considerably.
+This immediately poses challenges such as \textit{(a)} finding a suitable aggregation that encodes each feature's distribution and allows for a change detection test and \textit{(b)} defining a deviation threshold $\alpha$.
+
+\subsection*{Approximated Histogram Aggregation}
+
+We chose to encode a feature distribution as a histogram.
+A histogram is an aggregation that represents the distribution of numerical features.
+However, we must recall that we need to keep this histogram in constant time in a sliding window fashion: that implies we need to be able to evict old elements and insert new ones.
+
+An exact histogram aggregation must keep the entire sliding window in memory...
+
+\textcolor{red}{TODO: review this section after EMAs is written}
+
+\subsection*{Defining the deviation threshold}
+
+Assume we use a distance function that takes two histograms, our reference aggregation histogram and our sliding window aggregation histogram, and outputs a distance or difference value.
+How do we threshold this value?
+For instance, consider the following scenario, where ...
+
+\textcolor{red}{TODO: give an example of left and right skewed distros}
 %  - give a practical use-case where it makes a diference:
  %   - left-skewed distro vs right-skewed distro ---> threshold is very different (95 or 99 percentile for example)
+
+Hard-coding a threshold value to be used for all features is cumbersome, impractical and error prone.
+The distance threshold must be computed based on the distribution of distances.
+How do we get to know the distribution of such distances?
+That is one of the goals for the batch phase.
 

### 2020-06-15 chapters/sota.tex
Aggregation properties and families

@@
 Quoting the authors, histogramming univariate point outlier detection methods are \textit{"based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original"}.
-
-\textcolor{red}{TODO: explain this better, paper not clear to me}
 
@@
 The authors study methods for identifying differences in distribution between two samples and implement them as the \textit{d} function.
-
-\textcolor{red}{\textbf{TODO: develop this idea, they talk about Kolmogorov–Smirnov and others, maybe relate to model monitoring paper, talk about JSD, Wasserstein, so on so forth}}
+They provide as examples of a possible \textit{d} function implementation the  Kolmogorov–Smirnov Test \cite{Kolmogorov-Smirnov} and the Wilcoxon-Signed-Rank Test \cite{Wilcoxon}.
 
@@
 
-\textcolor{red}{TODO: Add table with SWAG algorithms, time and space complexity and applicable restrictions on the aggregation}
-
 \section{Probabilistic Data Structures} \label{sec:pds}
@@
 However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods.
-These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of magnitude faster then some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds.
+These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of-magnitude faster than some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds.
 Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.
@@
 % error bound related with parameters
-The false positive rate is a function of the bit array size \textit{m}, the number of hash functions \textit{k} and the number of elements in the set \textit{n}.
+The false-positive rate is a function of the bit array size \textit{m}, the number of hash functions \textit{k} and the number of elements in the set \textit{n}.
 It is assumed that the hash functions produce an independent and random index value for each element.
@@
 Hence, insertion loops through \textit{d} rows and has time complexity of \textit{O(d)}.
-Similarly, querying the frequency of an element \textit{el} consists in computing the minimum of all row values.
+Similarly, querying the frequency of an element \textit{el} consists of computing the minimum of all row values.
 Since both insertion and query have constant time complexity, the CMS is constant in time.
@@
 
-The authors claim that the proposed sliding version remains as accurate as the original HLL aggregator.
+The authors claim that the proposed sliding version remains as accurate as of the original HLL aggregator.
 In the context of their work, each element in the set is denoted as a "flow" and they state that \textit{"In practice, with an additional memory of only 35kB, a standard error of about 3\% can be achieved for a data stream of several million flows"}.

### 2020-06-16 chapters/my-work.tex
Batch-phase method description

@@
 Note that we need not to directly compare the entire window contents of the reference period with the contents of the sliding window.
-Doing this would require storing all window contents hence making our memory consumption grow linearly with sliding window or reference period size.
+Doing this would require storing all window contents hence making our memory consumption grow linearly with sliding window and/or reference period size.
 We want some sort of aggregation over the reference period that can be maintained efficiently throughout run-time and can be used to distinguish both periods of time and alert for changes.
@@
 
-\subsection*{Approximated Histogram Aggregation}
+\subsection*{Approximated Histogram Aggregation} \label{sec:ema-hist}
 
@@
 A histogram is an aggregation that represents the distribution of numerical features.
-However, we must recall that we need to keep this histogram in constant time in a sliding window fashion: that implies we need to be able to evict old elements and insert new ones.
+However, we must recall that we need to update this histogram in constant time in a sliding window fashion, implying eviction of old elements and insertion of new ones must be done with \textit{O(1)} time complexity.
 
@@
 
-Assume we use a distance function that takes two histograms, our reference aggregation histogram and our sliding window aggregation histogram, and outputs a distance or difference value.
+Assume we use a distance function that takes two histograms --- our reference aggregation histogram and our sliding window aggregation histogram --- and outputs a distance value.
 How do we threshold this value?
-For instance, consider the following scenario, where ...
+For instance, consider the following left and right skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.
+Note that these histograms do not represent the reference and/or sliding window aggregation histogram.
+Here we make the assumption we have an histogram that encodes the distribution of distance values measured between the reference aggregation and the sliding window aggregation.
 
-\textcolor{red}{TODO: give an example of left and right skewed distros}
-%  - give a practical use-case where it makes a diference:
- %   - left-skewed distro vs right-skewed distro ---> threshold is very different (95 or 99 percentile for example)
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/skewed-left-distro.png}
+      \caption[]{Left skewed distribution of distance values}
+      \label{fig:skewed-left-distro}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/skewed-right-distro.png}
+      \caption[]{Right skewed distribution of distance values}
+      \label{fig:skewed-right-distro}
+    \end{center}
+\end{figure}
+
+Let's assume Figure \ref{fig:skewed-left-distro} 's histogram encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}.
+Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results.
+For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}.
+Instead what we aim to do is know this distribution so that we can define our threshold as a percentile and not an hard-coded distance constant.
+For instance, in this case we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
 
@@
 \subsection{Batch analysis phase}
+The goal of the batch phase will be to build our reference snapshot and get to know the distribution of "distances" between the reference snapshot and the sliding window one.
+In other words, we obtain a reference aggregation histogram that we later on use to measure its distance to the sliding window aggregation histogram, producing a distance value.
+We compare the measured distance with its distribution to find out how probable it is and alert in a very unlikely scenario.
+We do this for each feature.
 
-%- batch phase: build reference exact histogram and distribution of distance values
-          %  1. compute ref histogram (for each feature)
-                %- equal heights/counts per bin, different size bins ---> better adjusted to wildly varying distributions
-         %   2. make samples of length equal to streaming window size, build histograms for each using bins from reference histogram to make later accurate comparisons between ref hist and sample hist
-             %   - why multiple sample histograms?
-                   % encode distributions over smaller time periods since there may be pattern changes within the large reference period
-             %   - why an approximated histogram for the samples as well despite being done in batch?
-                 %   mimic the streaming approximated histogram and the test that will be performed:
-                     %   - batch: sample app. histogram vs ref hist
-                    %    - stream: streaming app. histogram vs ref hist
-         %   3. compute the distance between each sample's histogram and the reference one (for each feature)
-          %  4. we now have a list of distance values, and know their expected distribution (for each feature)
-              %  - this is important since a test value obtained measuring the distance between the reference period and other window
-                %can be compared to this distribution (again probably give the left and right skewed example)
+\subsubsection*{Building the reference histogram for each feature}
 
-          %  By the end of the batch phase we have:
-               % for each feature:
-                    %reference histogram (for later comparison with streaming histogram)
-                    %list/histogram of distance values
-                    %last sample's histogram ---> used as burn-in period for target histogram
+First, for each feature, we compute the reference histogram from the reference period dataset.
+This is an exact histogram aggregation.
+When building it we ensure the histogram has equal counts (or equal height) for all bins which often results in different sized bins.
+We make no assumption on the dataset distributions and hence use equal height histograms because they adjust better to wildly varying ones.
+Equal height means we have more bins covering very dense regions (regions with many data points) and less bins in lower density regions.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/ref-hist.png}
+      \caption[]{Reference equal-height histogram}
+      \label{fig:ref-hist}
+    \end{center}
+\end{figure}
+
+\subsubsection*{Finding the distribution of distances for each feature}
+Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase.
+To that end, we make \textit{S} samples of transactions, each with the same tuple-based size of the sliding window we will use in streaming.
+Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
+
+For each sample, and for each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/ref-bins.png}
+      \caption[]{Reference equal-height histogram}
+      \label{fig:ref-hist-bins}
+    \end{center}
+\end{figure}
+
+Recall how an approximated histogram is built as explained in Section \ref{sec:ema-hist}.
+We will process each sample event by event and apply a discount factor per event, thus building an approximated histogram aggregation using the reference bins, for each feature of that sample.
+We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period.
+We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.
+
+For each feature, we have now \textit{S} histograms, one for each sample.
+For each of the \textit{S} histograms we compute the distance between it and the reference histogram.
+Figure \ref{fig:compute-sample-distances} illustrates this process as we apply our distance function for each sample histogram and the reference histogram, obtaining \textit{S} distance measurements.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/compute-sample-distances.png}
+      \caption[Compute as many distance values as samples]{Compute as many distance values as samples, one for each sample-reference histogram pair}
+      \label{fig:compute-sample-distances}
+    \end{center}
+\end{figure}
+
+For each feature, we end up with \textit{S} distance values.
+These values are distance measurements between random samples of data and the reference period, and the histograms for each sample were built exactly like they will be maintained in streaming for better fidelity.
+Hence, we claim we now know the distribution of expected distance values.
+Given a new distance value measured between another window and the reference window, we can compute its probability and produce alerts if below a certain threshold.
+
+\subsubsection*{Burn-in period at system initialization}
+When we boot up our system in the streaming phase we initially have empty histogram aggregations.
+The period of time where you discard the alerts produced by the system until it processes enough data to produce accurate reports is commonly denoted as burn-in period.
+In order to avoid this burn-in period we propose that the sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
+
+\subsubsection*{Batch phase artifacts}
+For each feature, by the end of the batch phase, we have a reference histogram, a list of distance values that represents its distribution and the last sample's histogram to be used as burn-in period for the sliding window or target histogram.
 

### 2020-06-17 chapters/background.tex
Proposed method

@@
 A batch is a collection of data points that have been grouped by certain criteria.
-Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user"}.
+Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}.
 Batch data sets are known as static data \cite{Martin-Batch-Defin}.
-An example application of such a technique would be processing all the keywords searched for in the google search engine.
-For instance, grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes.
+An example application of such a technique would be processing all the keywords searched for in the google search engine by grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes.
 Batch processing is the oldest paradigm between batch and stream processing.
@@
 More than conceptually different from a true stream processing system, processing a stream in a batch fashion, even if in small ones, presents some issues.
-First, since we are indeed working with batches, results will never be real-time updated per event.
+First, since we are indeed working with batches, results will never be updated in real-time per event being processed.
 For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events.
@@
 Is the average of all temperatures of all machines over an infinite period useful for the detection of overheating equipment?
-In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for actionable context, in this case identifying a likely-to-fail rack.
+In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for more actionable context, in this case identifying a likely-to-fail rack.
 To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}.
@@
 
-An aggregation --- also known as a fold, reduce or accumulate --- is an operation that reduces a collection of values to a single one.
+An aggregation --- also known as a fold, reduce or accumulate function --- is an operation that reduces a collection of values to a single one.
 Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality).
@@
 On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters.
-Approximate aggregations are required when memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time.
+Approximate aggregations are required where memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time.
 Thus, approximate aggregators trade precision of results for efficient time and space usage.

### 2020-06-17 chapters/introduction.tex
Batch-phase method description
Proposed method

@@
 This media content may have been previously recorded and stored, or may be generated on the fly.
-Social networks are also responsible for generating large volumes of data.
-A social network like Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
+Social networks are also responsible for generating large volumes of data,Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
 \textit{Tweets} may contain text, media content or both.
@@
 
-Both in batch and stream processing, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements.
-Operations applied to data sets that produce a single result are commonly known as aggregate operations.
+In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements.
+Operations applied to data sets that produce a single result are commonly known as aggregation operations.
 In stream processing, we apply these aggregations to data streams to obtain valuable information.
@@
 In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one.
-Such a system would have to process all the incoming transactions and decide in a fraction of a second if it is a fraudulent one or not.
+Such a system would have to process all incoming transactions and decide in a fraction of a second for each one whether it is fraudulent or not.
 It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.
@@
 Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time.
-In the Machine Learning field, the performance decay caused by data pattern shifts is also known as concept drift.
+In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift.
 
@@
 For any organization, the quality of their provided services is paramount.
-Therefore, it is common for companies to have systems monitoring the former, testing their usage rate, availability and performance, just to name a few.
+Therefore, it is common for companies to have auxiliary systems monitoring their services, testing their usage rate, availability and performance, just to name a few.
 The motivation behind this Thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in performance of the system analyzing said data stream.
@@
 
-Experiments will be made using multiple synthetic and real data sets from the financial fraud space.
+Experiments will be made using multiple synthetic data sets and real ones from the financial fraud space.
 

### 2020-06-17 chapters/my-work.tex
Proposed method

@@
 Both time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}.
-Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two clearly different time-series.
+Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two different time-series.
 
@@
 
-To that end, we devise a two-phased method.
+To that end, we devise a two-phase method.
 The first phase relies on a batch analysis that builds our aggregated reference snapshot to be later on compared with the sliding one.
@@
 \subsection{Stream monitoring goals}
+\textcolor{red}{TODO: add some pseudocode}
+
 Before diving into the batch and streaming algorithms we must further define our goal.
-In this Section we pose the question: \textit{"what is our goal during the streaming or monitoring phase?"} and answer it.
+In this Section, we pose the question: \textit{"what is our goal during the streaming or monitoring phase?"} and answer it.
 
@@
 \subsection*{Approximated Histogram Aggregation} \label{sec:ema-hist}
+\textcolor{red}{TODO: add some pseudocode}
 
@@
 A histogram is an aggregation that represents the distribution of numerical features.
-However, we must recall that we need to update this histogram in constant time in a sliding window fashion, implying eviction of old elements and insertion of new ones must be done with \textit{O(1)} time complexity.
+However, we must recall that we need to update this histogram in constant time in a sliding window fashion, implying the eviction of old elements and insertion of new ones must be done with \textit{O(1)} time complexity.
 
@@
 How do we threshold this value?
-For instance, consider the following left and right skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.
-Note that these histograms do not represent the reference and/or sliding window aggregation histogram.
-Here we make the assumption we have an histogram that encodes the distribution of distance values measured between the reference aggregation and the sliding window aggregation.
+For instance, consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.
+Note that these histograms do not represent the reference and/or sliding window aggregation histograms.
+Here we make the assumption we have a histogram that encodes the distribution of distance values measured between the reference aggregation and the sliding window aggregation.
 
@@
 For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}.
-Instead what we aim to do is know this distribution so that we can define our threshold as a percentile and not an hard-coded distance constant.
-For instance, in this case we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
+Instead what we aim to do is know this distribution so that we can define our threshold as a percentile and not a hard-coded distance constant.
+For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
 
-Hard-coding a threshold value to be used for all features is cumbersome, impractical and error prone.
+Hard-coding a threshold value to be used for all features is cumbersome, impractical and error-prone.
 The distance threshold must be computed based on the distribution of distances.
@@
 \subsection{Batch analysis phase}
+\textcolor{red}{TODO: add some pseudocode}
+
 The goal of the batch phase will be to build our reference snapshot and get to know the distribution of "distances" between the reference snapshot and the sliding window one.
-In other words, we obtain a reference aggregation histogram that we later on use to measure its distance to the sliding window aggregation histogram, producing a distance value.
+In other words, we obtain a reference aggregation histogram that we, later on, use to measure its distance to the sliding window aggregation histogram, producing a distance value.
 We compare the measured distance with its distribution to find out how probable it is and alert in a very unlikely scenario.
@@
 We make no assumption on the dataset distributions and hence use equal height histograms because they adjust better to wildly varying ones.
-Equal height means we have more bins covering very dense regions (regions with many data points) and less bins in lower density regions.
+Equal height means we have more bins covering very dense regions (regions with many data points) and fewer bins in lower density regions.
 
@@
 
-\subsubsection*{Finding the distribution of distances for each feature}
+\subsubsection*{Finding the distribution of distances for each feature} \label{sec:sampling-batch}
 Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase.
@@
 For each feature, we have now \textit{S} histograms, one for each sample.
-For each of the \textit{S} histograms we compute the distance between it and the reference histogram.
+For each of the \textit{S} histograms, we compute the distance between it and the reference histogram.
 Figure \ref{fig:compute-sample-distances} illustrates this process as we apply our distance function for each sample histogram and the reference histogram, obtaining \textit{S} distance measurements.
@@
 When we boot up our system in the streaming phase we initially have empty histogram aggregations.
-The period of time where you discard the alerts produced by the system until it processes enough data to produce accurate reports is commonly denoted as burn-in period.
-In order to avoid this burn-in period we propose that the sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
+The period of time where you discard the alerts produced by the system until it processes enough data to produce accurate reports is commonly denoted as the burn-in period.
+To avoid this burn-in period we propose that the sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
 
 \subsubsection*{Batch phase artifacts}
-For each feature, by the end of the batch phase, we have a reference histogram, a list of distance values that represents its distribution and the last sample's histogram to be used as burn-in period for the sliding window or target histogram.
+For each feature, by the end of the batch phase, we have a reference histogram, a list of distance values that represents its distribution and the last sample's histogram to be used as a burn-in period for the sliding window or target histogram.
 
 \subsection{Streaming phase}
+\textcolor{red}{TODO: add some pseudocode}
 
-%- online phase:
-          %  for each feature:
-            %    1. initialize the streaming histogram as the last sample's histogram from the batch phase (burn-in period)
+In the streaming or online phase, the system will process event by event in a true sliding window fashion (recall Section \ref{sec:windows}) and perform a change detection test periodically.
+This phase makes use of all of the batch phase artifacts, namely, for each feature, the reference histogram, the distribution of distance values and the last sample's histogram.
 
-         %   for each event:
-               % for each field/feature:
-                  %  1. update the feature's streaming histogram (recall the update method)
-                  %  2. compute the distance between the reference histogram and the target histogram
-                   % 3. test the percentile the distance value fits according to the feature's distances list and alert if above a certain percentile (say the 99th)
+For each feature, we initialize the histogram we will approximately maintain with the last sample's data, from the batch phase.
+Then, for each incoming event and for each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
+This is all we do, regarding event processing.
 
-           % introduce the multiple testing correction problem:
-              %  "the more inferences are made, the more likely erroneous inferences are to occur"
-             %   we are making multiple hypothesis testing when computing a distance and checking the percentile for each feature
+Periodically, we perform a multi-feature change detection test.
+The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the sliding window histogram of feature \textit{x}.
+This distance value is then tested against the known distribution of distance values for feature \textit{x}.
+In other words, we test to see the percentile of the computed distance in the distribution of distance values and alert if above a certain percentile --- \textit{e.g.} the 99th-percentile.
+When we perform the test, for each feature \textit{x}, we are computing the percentile of the distance measured between the reference and sliding window histogram aggregations.
+The probability value or \textit{p-value} of a distance value is
 
-                %introduce the Holm-Bonferroni multiple test correction method
+\[\textit{p-value\textsubscript{distance\textsubscript{x}}} = 1 - percentile(\textit{distance\textsubscript{x}})\]
 
-            %So in reality we have:
+Hence, if we set to alert states that have 1\% or less probability to happen, we raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01.
+Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
 
-          %  for each event:
-               % for each field/feature:
-                 %   1. update the feature's streaming histogram (recall the update method)
-                  %  2. compute the distance between the reference histogram and the target histogram
-                  %  3. apply holm-bonferroni for all feature's distances:
-                     %   - each percentile will be a p-value
-                    %    - order features by p-value
-                    %    - check if p-value > FWER / (m + 1 - k)
-                          %  if so then feature is divergent
+\[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]
 
-\subsection{Summary}
+or
+
+\begin{equation}
+    \label{eq:alert-test}
+    \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha
+\end{equation}
+
+\subsubsection*{Multiple Test Correction}
+Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis.
+When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives.
+In other words, when repeating a test multiple times the multiple test problem arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
+In the literature, many multiple test correction methods are proposed, but we opt to use the Holm-Bonferroni multiple test correction.
+The Holm-Bonferroni method is one of many approaches for controlling the so called Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
+
+\subsubsection*{Holm-Bonferroni Correction}
+The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.
+This correction is done by dividing the threshold $\alpha$ by a correction factor, as described below.
+
+First, we compute the \textit{p-value} of each feature.
+We then order our features by \textit{p-value}, in ascending order.
+Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if
+
+\[  \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha / (\textit{m} - \textit{k}) \]
+
+with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature.
+We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.
+
+\subsection{Final remarks}
+
+\subsubsection*{Optimal number of samples}
+In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase.
+But what is the optimal number of samples to make?
+How can we mathematically derive a formula for the number of samples to make and avoid \textit{ad hoc} user-defined constants?
+
+We are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests.
+We apply a multiple test correction on the tests performed and want to set the Family Wise Error Rate \textit{(FWER)} to $\alpha$.
+In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.
+
+We use the Holm-Bonferroni correction and, for each feature, we need to reliably estimate the upper tail of the distribution which corresponds to the \textit{$\alpha$ / n\textsubscript{features}} region.
+Each sample we do translates into a distance value and the collection of these data points builds the distribution of distance values.
+We want to be sure that the number of samples \textit{S} that we do generates enough distance points to have a high probability of having one or more points in the upper tail region.
+To that end, we define the probability of not having at least one point in the upper tail of the distance distribution as $\gamma$.
+
+\textcolor{red}{SEE MODEL MONITORING PAPER} We can then show that the optimal number of samples \textit{S} to make is
+
+\begin{equation}
+    \label{eq:optimal-n-samples}
+    \textit{S} = \frac{\log\gamma}{log(1 - \frac{\alpha}{n\textsubscript{features}})}
+\end{equation}

### 2020-06-18 chapters/background.tex
Some Background feedback applied

@@
 Batch processing is the oldest paradigm between batch and stream processing.
-Historically speaking, Apache Hadoop\footnote{https://hadoop.apache.org} was one of the first widely known technologies for large-scale batch processing, created in 2005.
-Nearly a decade later, Apache Spark\footnote{https://spark.apache.org} was released.
-Spark is capable of doing not only on-disk processing like Hadoop but also in-memory processing, achieving lower processing times.
+Historically speaking, Apache Hadoop \cite{borthakur2007hadoop} \cite{Hadoop} \cite{ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005.
+Hadoop uses a distributed computation framework or programming model called MapReduce \cite{MapReduce}.
+This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
+using a reducer function.
+To do this they read and write from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms).
+Nearly a decade later, Apache Spark \cite{ApacheSpark} \cite{Spark} was released.
+Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}.
+An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
+Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.
 
-As more and more data became available in the data stream format, the need for stream processing increased.
-Stream processing refers to the processing of data streams.
-A data stream is essentially an unbounded data set.
+On the other hand, stream processing refers to the processing of data streams.
+A data stream is essentially a continuous or unbounded data set.
+This data needs to be processed sequentially and incrementally on a event-by-event basis or over sliding windows.
+Streaming data includes a wide variety of data such as logs generated by customers using mobile or web applications, sensor data from Internet-of-Things (IoT) devices or credit card transactions.
 This is in contrast to batch processing which works on a finite data set.
-This distinction is important since some aggregations are impossible to compute with 100\% accuracy in an efficient manner on unbounded data sets.
+This distinction is important since some aggregations are very hard and even impossible to compute with 100\% accuracy for several use cases in an efficient manner on unbounded data sets.
 An example of such is counting distinct elements that flow through the data stream.
@@
 
-In order to provide a streaming solution, Apache Spark created the Spark Streaming\footnote{https://spark.apache.org/docs/latest/streaming-programming-guide.html} module.
-Being a batch system in its core, Spark implements streaming as micro-batching.
+In order to provide a streaming solution, Apache Spark created the Spark Streaming \cite{SparkStreaming} module.
+Being a batch system in its core, Spark implements streaming as micro-batching \cite{SparkStreamingPaper}.
 This approach handles streaming data by grouping incoming events into very small batches.
@@
 
-Distinguishing between batch and stream processing is important to understand how a monitoring system on top of the latter would work.
 The Venn diagram \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
@@
 % boundless aggregations no need for windows
-Computations over an unbounded data set are not impossible.
-For example, computing the \textit{average} of an event's field for all incoming events is computationally feasible.
+Computations over an unbounded data set are possible.
+For example, computing the \textit{average} of an event's field for all incoming events is computationally trivial: we need only to store the total number of events (one integer) and the current total sum (one double) of all events arriving to the system.
 However, is this boundless average over the stream of events useful?
-While in some scenarios it may be, more often than not we would like to introduce some logical context into our computations.
+While in some scenarios it may be, more often than not we would like to introduce some boundaries into our computations.
 For instance, consider the monitoring of machines spread out across multiple racks in a large data center and its temperatures.
 Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream.
-Is the average of all temperatures of all machines over an infinite period useful for the detection of overheating equipment?
+The average temperature of a machine over an infinite period (\textit{i.e.} since the machine was deployed) is of little value for detecting if the equipment is currently overheating, and to trigger the appropriate action to prevent a failure.
 In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for more actionable context, in this case identifying a likely-to-fail rack.
@@
 
-Gedik et.
-al claim that \textit{"recent history is often managed using windows"} and that \textit{"windowing makes it possible to implement streaming versions of the traditionally blocking relational operators, such as streaming a\cite{Wang-Windows-Stream-Processing}ggregations"} \cite{Wang-Windows-Stream-Processing}.
-More formally, Botan et.
-al define that \textit{"a window W over a stream S is a finite subset of S"} \cite{Botan-SECRET}.
+\subsection{Streaming Windows}
+
+Botan et. al \cite{Botan-SECRET} formally define a window as:
+
+\begin{definition}
+A window W over a stream S is a finite subset of S"
+\end{definition}
 
 % 2 dimensions of windows
-Windows can be categorized by two dimensions: the window size \textit{W} and the sliding step size \textit{S}, both either tuple or time-based.
-Tuple-based windows' size is specified by the number of items the window can contain and time-based windows specify the time interval worth of incoming data a window can store.
+Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$.
+Windows can be defined as tuple or time-based.
+Tuple-based windows are windows whose size is defined by \emph{number} of items contained in the window, whereas time-based windows are windows whose size is defined by the total \emph{time} a window can store.
 For instance, a tuple-based window of size $10^3$ tuples will store exactly $10^3$ items.
@@
 A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events.
-The step size \textit{S} of a windowing technique is formally defined as \textit{"the distance between consecutive windows"} \cite{Botan-SECRET}.
+The step size \textit{S} of a windowing technique is formally defined as the distance between consecutive windows \cite{Botan-SECRET}.
 This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away.
@@
 
-There are three types of windows: true sliding, stepping or tumbling windows.
-% true sliding: unitary steps
-A true sliding window is defined by any given window size \textit{W} and a unitary step (\textit{S}=1), which can be defined as time or tuple-based.
+Windows can also be characterized by how they move across data, \textit{i.e.}, how the step $S$ is defined.
+The three most common ways are: true sliding, stepping or tumbling windows.
+A true sliding window is a window with an unitary step, i.e., where $S = 1$.
 At each step, the window advances one unit, which may be one event for tuple-based steps, or one time-unit whether that is in milliseconds, seconds, days, weeks or another magnitude of time, measured based on the system clock or event time.
@@
 % stepping: step > 1 < window size
-A stepping window behaves exactly like a true sliding one but the step size \textit{S} is not unitary.
+A stepping window behaves exactly like a true sliding one but the step size \textit{S} is not unitary, \textit{i.e.}, $S > 1$.
 In other words, a stepping window is defined by any window size \textit{W} and steps \textit{S} greater than a unit.
-Once again, both the window and step sizes, \textit{W} and \textit{S} respectively, can be time or tuple-based.
 Figure \ref{fig:stepping-window} represents a stepping tuple-based window of size \textit{W} of three and a stepping size \textit{S} of two.
@@
 % tumbling: step = window size
-A tumbling window, also defined by its window size \textit{W} and step size \textit{S}, is a window where W is equal to S.
+A tumbling window is a window with a stepping size as large as the window, \textit{i.e.}, $S = W$.
 Hence, tumbling windows partition the stream into non-overlapping chunks with the same size.
@@
 
-For the scope of this Thesis, we will work under a tuple-based sliding window framework with a unitary step size.
+For the scope of this Thesis, we will work under a tuple-based true sliding window framework.
 Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.
@@
 
-An aggregation --- also known as a fold, reduce or accumulate function --- is an operation that reduces a collection of values to a single one.
+An aggregation --- also known as a fold, reduce or accumulate function --- is an operation that reduces a collection of values into a single one.
 Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality).
@@
 On the other hand, approximate aggregations have an error margin to the associated aggregation value which is usually controllable by adjusting a set of parameters.
-Approximate aggregations are required where memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time.
+Approximate aggregations are most useful in situations where memory is limited, the volume of data to process is large and/or computing the exact aggregation can not be done in useful time.
 Thus, approximate aggregators trade precision of results for efficient time and space usage.
@@
 
-Aggregations are present in most programming languages and frameworks.
-Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method\footnote{https://docs.oracle.com/en/java/javase/13/docs/api/java.base/java/util/stream/package-summary.html\#Reduction}:
+Aggregations are present in most programming languages and frameworks. Quoting from the Java programming language documentation on the \textit{Stream.reduce()} method \cite{Java13StreamReduce}: 
 \textit{"A reduction operation (also called a fold) takes a sequence of input elements and combines them into a single summary result by repeated application of a combining operation"}.
-Similarly, Hadoop's MapReduce paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one.
+Similarly, the MapReduce \cite{MapReduce} paradigm consists of first applying \textit{map} steps to the collection of data and then a \textit{reduce} one.
 \textit{Map} tasks will transform each item as desired and the \textit{reduce} task will merge the collection into a single value.
@@
 While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one.
-As we have seen in Section \ref{sec:windows}, to obtain a bounded data set from a data stream we apply windowing techniques.
-In doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it.
-Tangwongsan et.
-al define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
-Consider the data center temperature monitoring problem from Section \ref{sec:windows} as an example use-case for SWAGs.
-How would we compute the average temperature per rack in the data center in the last 5 minutes?
-To better understand, let us translate this question to a streaming SQL query, with a similar syntax to the API provided by Apache Flink\footnote{https://flink.apache.org}.
-Assuming all the temperatures can be queried from the \textit{Temperatures} table and that each rack corresponds a different \textit{rack\_id} we have the following query:
+As we have seen in Section \ref{sec:windows}, we can obtain a bounded data set from a data stream by applying windowing techniques.
+When doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it.
+Tangwongsan \emph{et.
+al} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}.
+In this example, we wanted to compute the average temperature per rack in the data cencer in the last 5 minutes.
+If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:
 
@@
 
-In the query above, the \textit{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \textit{timestamp} field to evict events not in this time interval.
-The resulting list of rows has \textit{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes.
-Finally, for each rack and its list of temperatures, we compute an \textit{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.
+In the query above, the
+\texttt{TUMBLE(timestamp, Time.minutes(5), rack\_id)} statement groups incoming data using time-based tumbling windows of 5 minutes duration, using the \texttt{timestamp} field to evict events not in this time interval.
+The resulting list of rows has \texttt{rack\_id} as its key and as corresponding value a list of temperatures, from the past 5 minutes.
+Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.
 
@@
 Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value.
-For each element insertion or eviction from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation.
+For each element inserted or evicted from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation.
 SOE relies on expiring the evicted elements from the aggregation state.
@@
 
-%\section{Aggregations on data types}
-
-%Some aggregations are only possible on certain data types.
-For instance, computing an average on a set of strings is not possible.
-In this Section we present the data types we will work with and the set of possible aggregations to compute on them.
-
-%We consider four data types: \textit{numericals}, \textit{strings}, \textit{dates} and \textit{categoricals}.
-Numerical variables can be measured.
-When compared to classical programming data types, for example those of the C programming language, numericals comprise variable types \textit{short}, \textit{long}, \textit{int}, \textit{float}, \textit{double} and valid combinations of the previous --- \textit{e.g.} \textit{long int}.
-On the other hand, Strings are sequences of characters.
-The \textit{string} type is the most generic data type in that any other type can be converted to a string.
-Every variable, for example \textit{x}=3, can be turned into a string \textit{s}, for example \textit{s}="3".
-Similarly, a date is a sequence of characters and thus a string.
-However, despite being a string, it has semantics, unlike a regular one.
-Knowing the date format, one can extract the day, month and year.
-For that reason we make the distinction between the \textit{string} and \textit{date} data types.
-Lastly, a \textit{categorical} is a variable that takes one of a relatively small and known set of possible values.
-A variable is assigned to a particular category based on its properties.
-The categorical variable is also a sequence of characters and therefore a string variable.
-The main difference between the categorical data type and the string is that we know the set of possible labels the categorical variable might take and this affects the type of computable aggregations.
-
-%On a set of numerical variables, we can compute sums, minima, maxima, averages, standard deviations and counts (of zeros, infinities and others).
-When operating on a set of strings, we can aggregate the collection into a concatenation of strings, a count, a membership query, cardinality.
-
 \section{Exponential Moving Averages} \label{sec:emas}

### 2020-06-18 chapters/introduction.tex
Some Background feedback applied

@@
 In the past decade, applications have become increasingly data-driven, placing data at the center of application design.
-Different use cases process data in different ways for different purposes.
-For instance, e-commerce platforms need to process a large number of transactions, for example, those made daily by all Alibaba's\footnote{https://alibaba.com} clients, while ensuring that sales run smoothly and that products are delivered to customers' homes.
+For different use cases, the data is processed in different ways for different purposes.
+For instance, e-commerce platforms such as Alibaba\footnote{https://alibaba.com} need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes.
 Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously.
-This media content may have been previously recorded and stored, or may be generated on the fly.
 Social networks are also responsible for generating large volumes of data,Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
@@
 The information that flows through such a stream can be analyzed on-the-fly or stored for later processing.
-The former case is best known as stream processing, while the latter as batch processing.
+The former is best known as stream processing, whereas the latter as batch processing.
 
@@
 Social networks process and analyze hundreds of posts per second to find out trending keywords across their network and suggest them to users.
-Computer network monitor systems want to detect intruders and revoke them system access before they have time to corrupt the system or steal private information.
-All of the previously given use cases rely on real-time processing of data for near real-time actions.
-Hence, between batch and stream processing, the latter fits these use cases better.
+Computer network monitoring systems aim to detect intruders and revoke their system access before they have time to corrupt the system or steal private information.
+All of the referred use cases rely on real-time processing of data for near real-time actions.
+Hence, stream processing fits these use cases better than batch processing.
 
@@
 However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, the demand for stream processing systems is increasing.
-Data stream processing refers to the handling of huge volumes of high variety of data generated at high-velocity from multiple sources in real-time.
-Unlike batch processing on static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery}.
+Unlike batch processing of static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery}.
 Furthermore, in the streaming paradigm, the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}.
 \textit{Recency} is measured under different scales for different use cases.
-For instance, considering the monitoring of geological data to forecast possible natural disasters such as earthquakes versus the monitoring of a computer network for intruder detection.
+For instance, consider the monitoring of geological data to forecast possible natural disasters such as earthquakes versus the monitoring of a computer network for intruder detection.
 In the former case, after the forecast of a future earthquake, governmental authorities need a couple of days to launch an evacuation plan and keep the population safe.
@@
 In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements.
-Operations applied to data sets that produce a single result are commonly known as aggregation operations.
+Operations that produce a single result when applied to data sets are commonly known as aggregation operations.
 In stream processing, we apply these aggregations to data streams to obtain valuable information.
@@
 In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one.
-Such a system would have to process all incoming transactions and decide in a fraction of a second for each one whether it is fraudulent or not.
+Such a system would have to process all incoming transactions and decide in a fraction of a second, for each transaction, whether it is fraudulent or not.
 It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.
@@
 Many market solutions monitor streams of credit card transactions with the intent of fraud detection.
-In the context of this Thesis, we will work side by side with Feedzai\footnote{https://feedzai.com}.
-Feedzai offers fraud prevention solutions for the previously presented use case of monitoring an infinite stream of credit card transactions and detecting fraud.
-We will work with them, taking advantage of the multiple real-life data sets they can provide from the financial fraud space.
+In the context of this Thesis, we work side by side with Feedzai\footnote{https://feedzai.com}.
+Feedzai offers fraud prevention solutions for the previously presented use case of monitoring a continuous stream of credit card transactions and detecting fraud.
+We exploit multiple real-life data sets that are provided from the financial fraud space.
 

### 2020-06-18 chapters/sota.tex
Some Background feedback applied

@@
 The authors split point outlier detection methods into uni and multivariate, as seen in Figure \ref{fig:outlier-taxonomy}.
-Furthermore, they split univariate point outlier detection methods into three broad categories: model-based, density-based and histogramming based \ref{fig:point-outlier-taxonomy}.
+Furthermore, they split univariate point outlier detection methods into three categories: model-based, density-based and histogram-based \ref{fig:point-outlier-taxonomy}.
 
@@
 These methods consider a datum an outlier if it has less than $\alpha$ neighbors.
-Distance between two points is usually measured using Euclidean distance.
+Distance between two points is usually measured using the Euclidean distance.
 Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.
 
-Quoting the authors, histogramming univariate point outlier detection methods are \textit{"based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original"}.
+Quoting the authors, histogram-based univariate point outlier detection methods are \textit{"based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original"}.
 
@@
 
-Ideally we want to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has rather than the actual list of neighbors.
+Ideally we want to maintain for each data point \textit{p\textsubscript{i}}, the number of neighbors it has, rather than the actual list of neighbors.
 The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors.

### 2020-06-18 chapters/my-work.tex
...

@@
 
-\subsubsection*{Multiple Test Correction}
+\subsubsection*{Multiple Test Correction} \label{sec:multi-test}
 Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis.
@@
 
-\subsubsection*{Holm-Bonferroni Correction}
+\subsubsection*{Holm-Bonferroni Correction} \label{sec:holmbonferroni}
 The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.

### 2020-06-18 chapters/validation.tex
Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory (HTM)
...

@@
 
-\section{Datasets description}
+\section{Datasets}
+
+% 01: 
+    % ref: N(10,2) for 5M
+    % target: N(10,2) for 2.5M and U(100,200) after for 2.5M
+
+% 02:
+% 03:
+% 04:
+% 05:
+% 06:
+
+We generated multiple synthetic datasets before moving on to real data.
+Furthermore, we first performed single feature analysis before moving to a multi-feature scenario.
+The problem of multiple testing (Section \ref{sec:multi-test}) only arises in the latter case so we only apply the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for that one.
+
 \subsection{Synthetic datasets}
+For each test we generated two datasets: one to be used by our method as the reference period and the other as the target or streaming period.
+
+\subsubsection*{Reference and Target Datasets 01}
+The Reference Dataset 01 (RD01) contained 5 million events each with a single feature \emph{x1} that followed a normal distribution with mean $\mu=10$ and $\sigma=2$.
+The Target Dataset 01 (TD01) was also made up of 5 million events each with a single feature.
+For the first half of the dataset, \textit{i.e.}, for the first 2.5 million events, similarly to the reference dataset, the feature followed a normal distribution with mean $\mu=10$ and $\sigma=2$.
+However, for the other half, the last 2.5 million events, we changed the generating distribution to be a continuous uniform one with $a=100$ and $b=200$.
+
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{figures/01-reference.png}
+  \captionof{figure}{Reference dataset 01 (RD01)}
+  \label{fig:rd01}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=.9\linewidth]{figures/01-target.png}
+  \captionof{figure}{Target dataset 01 (TD01)}
+  \label{fig:td01}
+\end{minipage}
+
 \subsection{Financial fraud datasets}
 
-\section{Tests description}
+\section{Experimental Setup}
 \subsection{Single feature}
@@
 
-\section{Tests results}
+\section{Results}
 \subsection{Single feature}

### 2020-06-19 chapters/my-work.tex
Marco "direct" feedback

@@
 \subsection*{Conclusion}
+The outlier detection methods reviewed in Section \ref{sec:outliers} were discarded as viable solutions to our problem because of the aforementioned problems of memory and/or time complexity, \textit{"memory loss"} and non-explainable alerts.
+
 We concluded that employing outlier detection methods was not the best course of action and put aside this approach.
@@
 Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.
-Yet another challenge would be to define the maximum deviation threshold $\alpha$  between the reference aggregation state and the streaming aggregation state to produce alerts.
 
@@
 
-However, there is no trivial solution for the second and third challenges.
-Defining the maximum allowed threshold between reference and sliding window aggregations does not have a straightforward solution.
-Furthermore, defining the set of aggregations to use is also a big challenge.
-To illustrate, consider both time-series show in Figure \ref{fig:approach2-timeseries}.
+However, there is no trivial solution for the second challenge.
+Defining the set of aggregations to use does not have a straightforward solution.
+To illustrate this problem, consider both time-series show in Figure \ref{fig:approach2-timeseries}.
 
@@
 
+\textcolor{red}{NOTE TO MOPS: You commented here, but arent EMAs senseitive to time correlations?
+Since we put lower weights on older events, the order matters (while in a sliding window every event has the same weight (1))}
 Both time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}.
@@
 This approach requires efficient sliding window aggregation state maintenance, which can be done using sliding window probabilistic data structures and/or exponential moving averages.
-However, defining the set of aggregations to use and the deviation threshold itself ultimately led us to discard this approach.
+However, the need to choose a set of aggregations to use and the deviation threshold itself ultimately led us to discard this approach.
 
-\section{Proposed method: Numerical Feature Monitoring}
+\section{Proposed method: Feature Distribution Monitoring}
 
@@
 Such shifts will be measured between a reference or normality period versus the observed reality during the streaming phase.
-Note that we need not to directly compare the entire window contents of the reference period with the contents of the sliding window.
-Doing this would require storing all window contents hence making our memory consumption grow linearly with sliding window and/or reference period size.
-We want some sort of aggregation over the reference period that can be maintained efficiently throughout run-time and can be used to distinguish both periods of time and alert for changes.
+Note that we do not need to directly compare the entire window contents of the reference period with the contents of the sliding window.
+The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size.
+Instead, we want to use an aggregation that does not grow linearly in memory regarding window size and that can be incrementally maintained during the streaming phase.
+The resulting reference and streaming aggregations should be comparable with the intent of raising divergence alerts.
 In short, we want a reference aggregation snapshot to compare with a streaming aggregation that must be updated in real-time and constant in memory usage.
@@
 The second phase focus on efficiently updating the sliding window aggregation snapshot and occasionally perform a similarity test between the static reference aggregation and the sliding one, alerting for changes if need be.
-The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types as well, as we will discuss.
+The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types, as we will discuss.
 
@@
 
-Before diving into the batch and streaming algorithms we must further define our goal.
-In this Section, we pose the question: \textit{"what is our goal during the streaming or monitoring phase?"} and answer it.
+Before diving into the batch and streaming algorithms we must further define our goal. In this Section, we address the question: \textit{"What is our goal during the streaming or monitoring phase?"}
 
@@
 In the context of this Thesis, each of the event's fields are also named features or time-dependent variables.
-You can think of them as columns in a traditional relational database table.
-Given this context, our goal is to produce alerts of the following form:
+They can be thought of as columns in a traditional relational database table.
+Our goal is to produce alerts of the following form:
 
@@
 
-The format used allows us to know the \textit{timestamp} for the given alert, which features diverge (in this case \textit{x1, x2, x3}) and a value representative of how much each feature diverges (for instance, a float value between 0 and 1 that works as a confidence value).
+This indicates the \textit{timestamp} for the given alert, which features diverge (in this case \textit{x1, x2, x3}) and a value representative of how much each feature diverges (\textit{e.g.} a float value between 0 and 1 that works as a confidence value).
 
@@
 We chose to encode a feature distribution as a histogram.
-A histogram is an aggregation that represents the distribution of numerical features.
+A histogram is an aggregation that represents the distribution of a numerical feature or a set of categories.
 However, we must recall that we need to update this histogram in constant time in a sliding window fashion, implying the eviction of old elements and insertion of new ones must be done with \textit{O(1)} time complexity.

### 2020-06-20 chapters/background.tex
...
EMAs draft

@@
 
-\subsection{Streaming Windows}
+\subsection{Sliding Windows}
 
@@
 Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.
+Note that this is very similar to any aggregation made in a standard Database Management System (DBMS).
+The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.} in this case, a tumbling window).
 
@@
 Subtract-On-Evict (SOE) on the other hand, is a SWAG algorithm that incrementally computes the aggregation value.
-For each element inserted or evicted from the window, the aggregation value is updated, in contrast to RFS which would recompute the entire aggregation.
+For each element inserted or evicted from the window, the aggregation value is updated without recomputing the entire aggregation like in RFS.
 SOE relies on expiring the evicted elements from the aggregation state.
 Hence, SOE only works for \textit{invertible} aggregations, that is, a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$.
-To illustrate, consider the computation of a sum aggregation over a sliding window.
+To illustrate the differences between SOE and RFS, consider the computation of a sum aggregation over a sliding window.
 The total sum of the window will be the aggregation state or value.
@@
 Algorithm \ref{pseudo:soe} shows pseudocode for SOE.
-
-Subtract-On-Evict is an improvement over the Recalculate-From-Scratch algorithm based on the idea of aggregation invertibility.
-The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.
 
@@
 \label{sec:agg-properties}
+
+On the previous Section, we presented Subtract-On-Evict as an improvement over Recalculate-From-Scratch based on the idea of aggregation invertibility.
+The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.
 
@@
     \end{tabular}
+
+    \end{center}
     \caption[Aggregation properties]{Aggregation properties.
@@
   \label{tbl:aggregations-properties}
-    \end{center}
 \end{table}
@@
 
-\textcolor{red}{TODO: explain EMAs, give mathematical proofs on error bounds, emulated window size based on nhalf}
+We talked about sliding window aggregations and algorithms to compute them and observe that the contents of the sliding window must be stored in memory so that when a new element is inserted we can evict the oldest one and discount its effect from the aggregation value.
+In this Section we analyze Exponential Moving Averages (EMAs).
+EMAs process each event only upon insertion.
+Since there is no eviction operation to perform, we do not need to store the whole window of data.
+
+Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011} \cite{EMA-Hunter} \cite{EMA-MarcusB} are applied to a sequence of elements to compute a weighted average, using different weights for each element.
+These weights decrease exponentially with the age of the observations.
+In other words, EMAs give higher weights to more recent data points.
+
+\textcolor{red}{Sliding window Weights image}
+\textcolor{red}{EMA Weights image}
+
+EMAs are recursive: the EMA value at timestamp \textit{t} depends only on the EMA value at timestamp \textit{t-1}.
+This recursive property allows us to store only one EMA value and continuously update it for each new value val\textsubscript{t}, based on the following recursion formula:
+
+\begin{definition}
+EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda$ + val\textsubscript{t}
+
+\label{def:ema}
+\end{definition}
+
+with $\lambda$ as a constant factor applied to the previous EMA value.
+This $\lambda$ constant (0 < $\lambda \leq$ 1) is known as the smoothing coefficient.
+The smoothing or decay factor $\lambda$ is given by:
+
+\[ \lambda = 2^{- \dfrac{1}{\tau}} \]
+
+where $\tau$ is the half-life of the EMA.
+The half-life $\tau$ controls how fast older events are forgotten as the smoothing factor depends on it.
+The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half.
+For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half.
+
+% EMA-count example (later build histogram concept on multiple EMA-counts)

### 2020-06-20 chapters/my-work.tex
EMAs draft

@@
 While in some scenarios this is a much-needed feature, in our use-case it is not.
-How clear the alerts and insights retrieved from the methods are is also an important property.
+How clear the alerts and insights retrieved from the methods are, is also an important property.
 Some outlier detection methods used Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.
@@
 
-Our goal is to detect data pattern shifts in an online fashion.
-Such shifts will be measured between a reference or normality period versus the observed reality during the streaming phase.
-Note that we do not need to directly compare the entire window contents of the reference period with the contents of the sliding window.
+A data stream is a continuous collection of events.
+Each event is timestamped and contains multiple fields.
+In the context of this Thesis, each of the event's fields are time-dependent variables named features.
+They can be thought of as columns in a traditional relational database table.
+Our goal is to detect data pattern shifts for each feature in an online fashion.
+Such shifts will be measured between a reference period versus the observed reality during the streaming phase.
+Figure \ref{fig:timelines} shows the separated reference and streaming timelines.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/timeperiods.png}
+      \caption[]{Reference and Streaming timelines}
+      \label{fig:timelines}
+    \end{center}
+\end{figure}
+
+We do not need to directly compare the entire window contents of the reference and sliding window periods.
 The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size.
-Instead, we want to use an aggregation that does not grow linearly in memory regarding window size and that can be incrementally maintained during the streaming phase.
-The resulting reference and streaming aggregations should be comparable with the intent of raising divergence alerts.
-In short, we want a reference aggregation snapshot to compare with a streaming aggregation that must be updated in real-time and constant in memory usage.
+Instead, we want to use a  \textit{(a)} constant in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
+Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.
 
-To that end, we devise a two-phase method.
-The first phase relies on a batch analysis that builds our aggregated reference snapshot to be later on compared with the sliding one.
-The second phase focus on efficiently updating the sliding window aggregation snapshot and occasionally perform a similarity test between the static reference aggregation and the sliding one, alerting for changes if need be.
-The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types, as we will discuss.
+Histograms are a representation of the distribution of numerical or categorical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}.
+Histograms require the storage of a fixed-size list of bins and a corresponding fixed-size list of counts per bin for ever growing volumes of data, satisfying property \textit{(a)}.
+Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and adding or subtracting one, respectively, thus satisfying property \textit{(b)}.
+However, computing an histogram aggregation over a sliding window of size $W$ implies storing all $W$ elements in memory to know the element to remove after a new one is inserted.
+This violates our constant in memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}.
+EMA-like histograms computed over sliding windows do not require the storage of window contents.
+An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values.
+EMA-like histograms are further explained in Section \ref{sec:ema-hist}.
 
-\subsection{Stream monitoring goals}
-\textcolor{red}{TODO: add some pseudocode}
+Since the reference window is static, \textit{i.e.}, it is not a sliding window, we can use exact histogram aggregations.
+Hence, for each feature, we build an exact reference histogram aggregation.
+On the other hand, in streaming, we use one EMA-like approximate histogram aggregation per feature (Figure \ref{fig:timelines-hists}).
 
-Before diving into the batch and streaming algorithms we must further define our goal. In this Section, we address the question: \textit{"What is our goal during the streaming or monitoring phase?"}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/timelines-hists.png}
+      \caption[]{Reference and Streaming timelines and corresponding histogram aggregations}
+      \label{fig:timelines-hists}
+    \end{center}
+\end{figure}
 
-A data stream is an infinite stream of events.
-Each event is timestamped and contains multiple fields.
-In the context of this Thesis, each of the event's fields are also named features or time-dependent variables.
-They can be thought of as columns in a traditional relational database table.
-Our goal is to produce alerts of the following form:
+To compare each feature's reference and sliding window histogram aggregation we need a measure of divergence.
+We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein.
+\textcolor{red}{defer the why to later section} Given a divergence value for the two aggregations (reference and sliding window ones) we need a measure of normal statistical fluctuations.
+Since we assume a reference period as the normal state we can find out the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
+This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value.
+\textcolor{red}{defer details to later section}
 
-\[\textit{timestamp: [(x1, 0.23), (x2, 0.51), (x3, 0.55)}]\]
+We devise a two-phase method.
+The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section ??).
+The second phase focus on efficiently updating the sliding window histogram for each feature and periodically performs a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be.
+The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.
 
-This indicates the \textit{timestamp} for the given alert, which features diverge (in this case \textit{x1, x2, x3}) and a value representative of how much each feature diverges (\textit{e.g.} a float value between 0 and 1 that works as a confidence value).
-
-An alert should be raised if a feature distribution changes considerably.
-This immediately poses challenges such as \textit{(a)} finding a suitable aggregation that encodes each feature's distribution and allows for a change detection test and \textit{(b)} defining a deviation threshold $\alpha$.
-
-\subsection*{Approximated Histogram Aggregation} \label{sec:ema-hist}
-\textcolor{red}{TODO: add some pseudocode}
-
-We chose to encode a feature distribution as a histogram.
-A histogram is an aggregation that represents the distribution of a numerical feature or a set of categories.
-However, we must recall that we need to update this histogram in constant time in a sliding window fashion, implying the eviction of old elements and insertion of new ones must be done with \textit{O(1)} time complexity.
-
-An exact histogram aggregation must keep the entire sliding window in memory...
-
-\textcolor{red}{TODO: review this section after EMAs is written}
-
+\iffalse
 \subsection*{Defining the deviation threshold}
-
 Assume we use a distance function that takes two histograms --- our reference aggregation histogram and our sliding window aggregation histogram --- and outputs a distance value.
@@
 How do we get to know the distribution of such distances?
-That is one of the goals for the batch phase.
+That is one of the goals of the batch phase.
+\fi
+
+\subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
+The sliding window histogram we use is an Exponential Moving Average (EMA) based histogram. An EMA-like histogram is essentially a collection of EMA-counts as described in Section \ref{sec:emas}.
+
+.....
 
@@
 
-The goal of the batch phase will be to build our reference snapshot and get to know the distribution of "distances" between the reference snapshot and the sliding window one.
-In other words, we obtain a reference aggregation histogram that we, later on, use to measure its distance to the sliding window aggregation histogram, producing a distance value.
-We compare the measured distance with its distribution to find out how probable it is and alert in a very unlikely scenario.
-We do this for each feature.
+The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and the sliding window aggregations.
+In other words, we:
+
+\begin{enumerate}
+    \item obtain a reference histogram for each feature using all data in the reference period
+
+    \item then compute the distribution of distances between the reference histogram and the sliding window histogram by computing the distance (or divergence value) between the reference histogram and various samples of sliding window histograms
+\end{enumerate}
+
+This distribution of values can then later be compared to a given value of divergence for a single sample to find out how probable that value is to occur (if it is very low an alert may be raised).
 
@@
 
-For each sample, and for each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}.
+For each sample, and for each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}).
+We also add two extra bins that will cover the region to the left and to the right of our histogram, respectively.
+In other words, we add a bin that covers the region between $-\infty$ and our first bin and another that covers values from the last bin to $+\infty$.
+This way, observed values outside of the reference period will be placed in these special bins, either left or right.
 
@@
 
-Recall how an approximated histogram is built as explained in Section \ref{sec:ema-hist}.
-We will process each sample event by event and apply a discount factor per event, thus building an approximated histogram aggregation using the reference bins, for each feature of that sample.
+Each of the histograms is an EMA-like histogram.
+An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}.
+Each sample is processed event by event and a discount factor is applied per event, thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample.
 We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period.
@@
 For each feature, we end up with \textit{S} distance values.
-These values are distance measurements between random samples of data and the reference period, and the histograms for each sample were built exactly like they will be maintained in streaming for better fidelity.
+These values are distance measurements between random samples of data and the reference period.
 Hence, we claim we now know the distribution of expected distance values.
-Given a new distance value measured between another window and the reference window, we can compute its probability and produce alerts if below a certain threshold.
+Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if below a certain threshold.
+Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the sliding window histogram.
+This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.
 

### 2020-06-20 chapters/sota.tex
EMAs draft

@@
 
-A time-series is a collection of data points indexed by time.
-Accordingly, the authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}.
+A time-series is a collection of data points sorted by time.
 Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}.
@@
 A univariate time-series refers to a time-series consisting of a list of values generated by a single time-dependent variable throughout time.
-To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}.
 On the other hand, a multivariate time-series has more than one time-dependent variable.
 Most importantly, each variable is not only time-dependent but may also have dependencies on other variables.
-With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}.
-Each \textit{k-dimensional} vector consists of \textit{k}  time-dependent variables, possibly correlated.
 
@@
 An outlier detection method may perform a univariate or multivariate analysis.
-A univariate method works only with a single time-dependent variable whereas a multivariate one simultaneously works with more than one variable.
+A multivariate analysis considers all variables in the time-series simultaneously while univariate analysis looks at each variable independently.
 An outlier detection method can take as input a multivariate time-series but perform a univariate analysis, one per time-series of the multivariate input.
@@
 
-The authors split point outlier detection methods into uni and multivariate, as seen in Figure \ref{fig:outlier-taxonomy}.
-Furthermore, they split univariate point outlier detection methods into three categories: model-based, density-based and histogram-based \ref{fig:point-outlier-taxonomy}.
+Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy}).
 
@@
 \end{equation*}
-Additionally, the authors split the model-based category into estimation and prediction models.
+Model-based point outlier detection methods can be one of two types: estimation or prediction models.
 The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values.
-Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can't.
+Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can not.
 
@@
 
-Ideally we want to maintain for each data point \textit{p\textsubscript{i}}, the number of neighbors it has, rather than the actual list of neighbors.
+The authors want to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has, rather than the actual list of neighbors.
 The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors.

### 2020-06-20 chapters/validation.tex
EMAs draft

--- 2020-06-11 chapters/validation.tex
@@
 
-\section{Datasets description}
+\section{Datasets}
+
+We generated multiple synthetic datasets before moving on to real data.
+Furthermore, we first performed single feature analysis before moving to a multi-feature scenario.
+The problem of multiple testing (Section \ref{sec:multi-test}) only arises in the latter case so we only apply the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for that one.
+
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|}
+        \hline
+        \textbf{Dataset} & \textbf{Features} & \textbf{Events} \\ \hline
+        R1               & 1                 & 5.000.000       \\ \hline
+        T1               & 1                 & 5.000.000       \\ \hline
+
+        \end{tabular}
+    \end{center}
+    \caption{Summary statistics for each dataset}
+    \label{tbl:summary-statistics}
+\end{table}
+
 \subsection{Synthetic datasets}
+Dataset R1 had a single feature that followed a normal distribution with mean $\mu=10$ and $\sigma=2$ for the entirety of the dataset.
+Figure \ref{fig:timeseries-r1} shows the time-series of the single feature in dataset R1.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/01-reference.png}
+      \caption[]{Time-series for dataset R1}
+      \label{fig:timeseries-r1}
+    \end{center}
+\end{figure}
+
+Dataset T1 also had 5 million events with a single feature.
+For the first half, \textit{i.e.}, for the first 2.5 million events, the feature followed a normal distribution with mean $\mu=10$ and $\sigma=2$.
+For the other half, we changed the generating distribution to be a continuous uniform one with $a=100$ and $b=200$.
+Figure \ref{fig:timeseries-t1} shows the time-series of the single feature in dataset T1 where we see the abrupt change in feature values.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/01-target.png}
+      \caption[]{Time-series for dataset T1}
+      \label{fig:timeseries-t1}
+    \end{center}
+\end{figure}
+
 \subsection{Financial fraud datasets}
 
-\section{Tests description}
-\subsection{Single feature}
-\subsection{Multi feature}
+\section{Experiments}
+\textcolor{red}{one subsection per experiment with i) purpose, ii) execution, iii) results obtained and what thye mean}
 
-\section{Tests results}
-\subsection{Single feature}
-\subsection{Multi feature}
+\subsection{Single-feature analysis with varying sample sizes}
+In this first experiment, we used dataset R1 as the reference period and dataset T1 as the target or streaming period.
+We used 100 bins for the equal-height reference histogram and then added the two special bins mentioned in Section \ref{sec:sampling-batch}.
+We changed the half-life of each EMA-like sample histogram (recall Section \ref{sec:ema-hist}), effectively changing its size.
+We tested samples with \textit{(a)} 2.500, \textit{(b)} 250.000 and \textit{(c)} 1.000.000 events.
+
+\textcolor{red}{We use tukey threshold, how do we introduce this, ahhhhhhhhhhhh}
 

### 2020-06-21 chapters/background.tex
EMA histograms

@@
 For the scope of this Thesis, we will work under a tuple-based true sliding window framework.
-Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.
+However, as explained in Section \ref{sec:ft-monitoring}, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.
 
@@
 For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half.
+Note that it is also possible to express a tuple-based half-life, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half (\textit{e.g.} $\tau = 1$ tuple, $\lambda = 0.5$, for each new tuple reduce all weights by half).
 

### 2020-06-21 chapters/my-work.tex
EMA histograms

@@
 
-Using point outlier detection methods, the main idea would be to keep track of the rate of point outliers detected.
+Using point outlier detection methods, the main idea is to keep track of the rate of point outliers detected.
 For instance, keep track of how many outliers were alerted in the past hour.
@@
 In our lightweight real-time solution we need constant time complexity to process events one by one in real-time.
-Additionally, because our system is not mission-critical, we want a solution as lightweight as possible, which means something that grows linearly with window size is not a good fit.
+Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit.
 Therefore, most of the outlier detection methods presented become unfit for our use due to high memory consumption and time complexity.
-Some of the methods shown also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period.
+Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period.
 In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream.
@@
 How clear the alerts and insights retrieved from the methods are, is also an important property.
-Some outlier detection methods used Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.
+Some outlier detection methods use Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.
 
@@
 
-\section{Proposed method: Feature Distribution Monitoring}
+\section{Proposed method: Feature Distribution Monitoring} \label{sec:ft-monitoring}
 
@@
 We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein.
-\textcolor{red}{defer the why to later section} Given a divergence value for the two aggregations (reference and sliding window ones) we need a measure of normal statistical fluctuations.
+A more thorough analysis on some statistical tests and why we chose to use the JSD is performed in Section \ref{sec:whyjsds}.
+Given a divergence value for the two aggregations (reference and sliding window ones) we need a measure of normal statistical fluctuations.
 Since we assume a reference period as the normal state we can find out the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
@@
 We devise a two-phase method.
-The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section ??).
-The second phase focus on efficiently updating the sliding window histogram for each feature and periodically performs a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be.
+The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}).
+The second phase (detailed in Section \ref{sec:stream-phase}) focus on efficiently updating the sliding window histogram for each feature and periodically performs a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be.
 The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.
@@
 \subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
-The sliding window histogram we use is an Exponential Moving Average (EMA) based histogram. An EMA-like histogram is essentially a collection of EMA-counts as described in Section \ref{sec:emas}.
+As stated, we compute histogram aggregations for the reference and sliding windows.
+The computation of the reference histogram is done once in batch and hence an exact histogram can be computed.
+However, the sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.
 
-.....
+In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they emulate a sliding window without actually storing window elements....
 
-\subsection{Batch analysis phase}
+\textcolor{red}{TODO: complete this after EMAs in background are corrected}
+
+\subsection{Batch analysis phase} \label{sec:batch-phase}
 \textcolor{red}{TODO: add some pseudocode}
@@
 
-\subsection{Streaming phase}
+\subsection{Streaming phase} \label{sec:stream-phase}
 \textcolor{red}{TODO: add some pseudocode}
@@
 \end{equation}
+
+\subsubsection*{Distribution Divergence Statistical Tests} \label{sec:whyjsds}
+
+We used the Jensen-Shannon Divergence (JSD) but....
+
+\textcolor{red}{Talk about wassertein, kolomogrov, why JSD...}

### 2020-06-21 chapters/problem-statement.tex
EMAs draft
EMA histograms

@@
 \chapter{Problem Statement} \label{chap:statement} \minitoc
+
+\textcolor{red}{INCOMPLETE}
 
@@
 \section{State of the art issues}
-Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques.
+Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques required to monitor streaming data patterns in real-time.
 In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods.

### 2020-06-21 chapters/sota.tex
EMA histograms

@@
 In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to.
-This technique allows the efficient maintenance of neighborship counts by pre-handling the eventual expiring effect of new data points.
 
@@
 \subsection*{Applicability to our Hypothesis}
-Bloom filters exhibit constant space and time complexity making it great methods to test our hypothesis.
+Bloom filters exhibit constant space and time complexity making them possible methods to test our hypothesis.
 Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}.
+However, we need to evict old events, and for that reason, bloom filters are of no use to us (unless we find sliding window implementations).
 

### 2020-06-21 chapters/my-work.tex
Why we build divergence distribution

@@
 Since we assume a reference period as the normal state we can find out the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
-This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value.
-\textcolor{red}{defer details to later section}
+This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).
 
@@
 The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.
-
-\iffalse
-\subsection*{Defining the deviation threshold}
-Assume we use a distance function that takes two histograms --- our reference aggregation histogram and our sliding window aggregation histogram --- and outputs a distance value.
-How do we threshold this value?
-For instance, consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.
-Note that these histograms do not represent the reference and/or sliding window aggregation histograms.
-Here we make the assumption we have a histogram that encodes the distribution of distance values measured between the reference aggregation and the sliding window aggregation.
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/skewed-left-distro.png}
-      \caption[]{Left skewed distribution of distance values}
-      \label{fig:skewed-left-distro}
-    \end{center}
-\end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/skewed-right-distro.png}
-      \caption[]{Right skewed distribution of distance values}
-      \label{fig:skewed-right-distro}
-    \end{center}
-\end{figure}
-
-Let's assume Figure \ref{fig:skewed-left-distro} 's histogram encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}.
-Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results.
-For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}.
-Instead what we aim to do is know this distribution so that we can define our threshold as a percentile and not a hard-coded distance constant.
-For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
-
-Hard-coding a threshold value to be used for all features is cumbersome, impractical and error-prone.
-The distance threshold must be computed based on the distribution of distances.
-How do we get to know the distribution of such distances?
-That is one of the goals of the batch phase.
-\fi
 
@@
 \subsubsection*{Finding the distribution of distances for each feature} \label{sec:sampling-batch}
+
 Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase.
@@
 This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.
+
+It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself.
+For instance, given the JSD value varies between 0 and 1, why not just set to alert if the measured JSD value is 0.7?
+The reason is that a divergence value may fall on different percentiles for different features.
+In other words, the same \textit{i-th} percentile will correspond to a different divergence value for different features.
+Consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/skewed-left-distro.png}
+      \caption[]{Left skewed distribution of distance values}
+      \label{fig:skewed-left-distro}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/skewed-right-distro.png}
+      \caption[]{Right skewed distribution of distance values}
+      \label{fig:skewed-right-distro}
+    \end{center}
+\end{figure}
+
+Let's assume Figure \ref{fig:skewed-left-distro} 's histogram encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}.
+Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results.
+For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}.
+Instead what we aim to do is know this distribution so that we can define our threshold as a percentile and not a hard-coded distance constant.
+For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
+
+Hard-coding a threshold value to be used for all features is cumbersome, impractical and error-prone. The distance threshold must be computed based on the distribution of distances.
 

### 2020-06-22 chapters/background.tex
EMA-like histogram

@@
 
-\textcolor{red}{Sliding window Weights image}
-\textcolor{red}{EMA Weights image}
+\textcolor{red}{TODO: Images for EMA and sliding window weights}
 
@@
 \begin{definition}
-EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda$ + val\textsubscript{t}
-
+EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda\delta\textsubscript{t}$ + val\textsubscript{t}
 \label{def:ema}
@@
 
-with $\lambda$ as a constant factor applied to the previous EMA value.
+with $\lambda$ as a constant factor applied to the previous EMA value and $\delta\textsubscript{t}$ as the time passed between \textit{t} and \textit{t-1}.
 This $\lambda$ constant (0 < $\lambda \leq$ 1) is known as the smoothing coefficient.
 The smoothing or decay factor $\lambda$ is given by:
-
-\[ \lambda = 2^{- \dfrac{1}{\tau}} \]
-
+\begin{equation}
+    \lambda = 2^{- \dfrac{1}{\tau}}
+    \label{eq:ema-decay}
+\end{equation}
 where $\tau$ is the half-life of the EMA.
-The half-life $\tau$ controls how fast older events are forgotten as the smoothing factor depends on it.
+The half-life $\tau$ controls how fast old events are forgotten as the smoothing factor depends on it.
 The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half.
 For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half.
-Note that it is also possible to express a tuple-based half-life, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half (\textit{e.g.} $\tau = 1$ tuple, $\lambda = 0.5$, for each new tuple reduce all weights by half).
 
-% EMA-count example (later build histogram concept on multiple EMA-counts)
+Note that it is also possible to express a tuple-based half-life, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half.
+In this case, the definition does not take into account the time passed between events:
+\begin{definition}
+EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda$ + val\textsubscript{t}
+\label{def:tuple-ema}
+\end{definition}
+For example, with a tuple-based half-life of $\tau = 1$ tuple, the smoothing factor is $\lambda = 0.5$, meaning for each new tuple all weights are reduced by half.
+
+In this Thesis we will work with tuple-based EMAs but our approach can be adjusted to use time-based ones.

### 2020-06-22 chapters/my-work.tex
EMA-like histogram

@@
 
-\section{Approach A: Usage of outlier detection methods}
+\section{State of the art outlier detection methods}
 The first approach considered relied on outlier detection methods such as the ones presented in Section \ref{sec:outliers} and classified by the taxonomy in Figure \ref{fig:outlier-taxonomy}.
-This straightforward approach works with either subsequence or point outlier detection methods to analyze the streaming sliding window and produce data pattern shift alerts.
 
-\subsection*{How}
-
+We reviewed subsequence and point outlier detection methods.
 Using point outlier detection methods, the main idea is to keep track of the rate of point outliers detected.
@@
 The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$.
-On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier.
+On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier when compared to the reference window.
 
@@
 
-In our lightweight real-time solution we need constant time complexity to process events one by one in real-time.
+For this Thesis we need constant time complexity to process events one by one in real-time.
 Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit.
-Therefore, most of the outlier detection methods presented become unfit for our use due to high memory consumption and time complexity.
 Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period.
@@
 
-We concluded that employing outlier detection methods was not the best course of action and put aside this approach.
-
-\section{Approach B: Thresholding aggregation values}
+\section{State of the art aggregation algorithms}
 This second approach focus on building an aggregation snapshot of the reference period to compare with streaming aggregation values and alert if the difference between both surpasses a threshold $\alpha$.
@@
 Assume we define the set of aggregations to compute the mean of \textit{x} and its standard deviation.
-Also assume we save the reference window and maintain a sliding window and the average and standard deviation aggregations for \textit{x} efficiently.
+Also assume we save the reference window and maintain a sliding target window and the associated average and standard deviation aggregations for \textit{x} efficiently.
 As seen in Figure \ref{fig:approach2-initial-state}, the initial mean and standard deviation values for \textit{x} presented in the reference window are of 5 and 1, respectively.
@@
 \subsection*{Challenges}
-The first issue with this approach is that generic sliding window aggregation algorithms grow linearly in space regarding window size, such as Recalculate-From-Scratch and Subtract-On-Evict presented in Section \ref{sec:back-swag-algs}, Two-Stacks has seen in Section \ref{sec:2stacks} and DABA in Section \ref{sec:daba}.
+The first issue with this approach is that generic sliding window aggregation algorithms grow linearly in space regarding window size, such as Recalculate-From-Scratch and Subtract-On-Evict from Section \ref{sec:back-swag-algs}, Two-Stacks from Section \ref{sec:2stacks} and DABA in Section \ref{sec:daba}.
 Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.
@@
 
-\section{Proposed method: Feature Distribution Monitoring} \label{sec:ft-monitoring}
+\section{Method: Feature Distribution Monitoring} \label{sec:ft-monitoring}
 
@@
 
-We do not need to directly compare the entire window contents of the reference and sliding window periods.
+We do not need to directly compare the entire window contents of the reference and sliding windows.
 The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size.
@@
 
-In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they emulate a sliding window without actually storing window elements....
+In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they compute sliding window aggregations without actually storing the window contents.
+Definition \ref{def:tuple-ema} shows how to compute an EMA using its recurrence formula.
+An EMA state or value depends only on the previous one times an exponential decay $\lambda$.
+Recall that this decay is given by Equation \ref{eq:ema-decay}:
+\[ \lambda = 2^{- \dfrac{1}{\tau}} \]
+where $\tau$ is commonly denoted as the half-life.
 
-\textcolor{red}{TODO: complete this after EMAs in background are corrected}
+After processing \textit{x} tuples the oldest tuple is \emph{smoothed} by a factor of $\lambda^x = 2^{- \dfrac{x}{\tau}}$.
+If we process $x = 4\tau$ tuples the oldest tuple would have its contribution reduced by
+\[ 2^{- \dfrac{4\tau}{\tau}} = 2^{-4} \]
+which is approximately 0.06. In other words, the \textit{x-th} event would have 6\% contribution towards the EMA state or value meaning older events would have even less weight.
+
+We choose to use windows with four times the half-life size (or $4\tau$), meaning we discard events whose contribution is less than 6\%.
+In fact, we set the half-life to be one quarter of the size of the window we want to emulate.
+For instance, if we want to compute an EMA aggregation over a period of a month, we set the half-life to be the number of tuples that correspond to one week worth of data.
 

### 2020-06-22 chapters/sota.tex
EMA-like histogram

@@
 These methods consider a datum an outlier if it has less than $\alpha$ neighbors.
-Distance between two points is usually measured using the Euclidean distance.
+Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}.
 Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.
@@
 In this Thesis, we want to detect and alert data pattern shifts with a low memory footprint.
-However, the method proposed by Kifer et.
+The method proposed by Kifer et.
 al grows linearly with window size, making it unfit for large windows, our use-case.
+
+However, this two-window paradigm encodes a static normal or reference period in one window and the observed streaming values in a sliding one.
+Since in this Thesis we want to report streaming data pattern shifts relatively to a reference period, we may use a similar two-windowed model if we find lightweight and incrementally maintained aggregations over both windows.
+
+\subsubsection{Automatic Model Monitoring for Data Streams}
+
+ALSO a 2 WINDOW MODEL
+\textcolor{red}{TODO: write this}
 

### 2020-06-23 chapters/background.tex
EMAs Pedro feedback applied

@@
 
-We talked about sliding window aggregations and algorithms to compute them and observe that the contents of the sliding window must be stored in memory so that when a new element is inserted we can evict the oldest one and discount its effect from the aggregation value.
+We talked about sliding window aggregations and algorithms to compute them.
+We observed that the contents of the sliding window must be stored in memory in order to evict the oldest event and discount its effect from the aggregation when a new one is inserted.
 In this Section we analyze Exponential Moving Averages (EMAs).
@@
 
-Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011} \cite{EMA-Hunter} \cite{EMA-MarcusB} are applied to a sequence of elements to compute a weighted average, using different weights for each element.
+Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011} \cite{EMA-Hunter} \cite{EMA-MarcusB} are a type of reduce functions applied to a sequence of elements, one at a time in a orderly fashion, to compute a weighted average, using different weights for each element.
 These weights decrease exponentially with the age of the observations.
 In other words, EMAs give higher weights to more recent data points.
+Figures \ref{fig:window-weights} and \ref{fig:ema-weights} show the weights applied to past events.
+In a typical sliding window, events either have weight of one (inside the window) or zero (outside the window) whereas EMAs give exponentially smaller weights to older events.
 
-\textcolor{red}{TODO: Images for EMA and sliding window weights}
+\begin{figure}[!htb]
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+      \includegraphics[scale=0.5]{figures/window-weights.png}
+  \captionof{figure}{Window weights}
+  \label{fig:window-weights}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+      \includegraphics[scale=0.5]{figures/ema-weights.png}
+  \captionof{figure}{EMA weights}
+  \label{fig:ema-weights}
+\end{minipage}
+\end{figure}
 
@@
 
-In this Thesis we will work with tuple-based EMAs but our approach can be adjusted to use time-based ones.
+Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones.
+In this Thesis we chose to work with tuple-based EMAs due to its simplicity but our approach can be adjusted to use time-based ones.

### 2020-06-23 chapters/my-work.tex
EMAs Pedro feedback applied

@@
 
-\textcolor{red}{NOTE TO MOPS: You commented here, but arent EMAs senseitive to time correlations?
-Since we put lower weights on older events, the order matters (while in a sliding window every event has the same weight (1))}
+\textcolor{red}{Marco had a problem with this example, dont understand what he meant}
 Both time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}.
@@
 They can be thought of as columns in a traditional relational database table.
-Our goal is to detect data pattern shifts for each feature in an online fashion.
+Our goal is to detect data pattern shifts for each feature in an streaming fashion.
 Such shifts will be measured between a reference period versus the observed reality during the streaming phase.
@@
 The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size.
-Instead, we want to use a  \textit{(a)} constant in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
+Instead, we want to use a  \textit{(a)} fixed cost in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
 Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.
 
-Histograms are a representation of the distribution of numerical or categorical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}.
-Histograms require the storage of a fixed-size list of bins and a corresponding fixed-size list of counts per bin for ever growing volumes of data, satisfying property \textit{(a)}.
-Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and adding or subtracting one, respectively, thus satisfying property \textit{(b)}.
+One aggregation that encodes distributions properly are histograms.
+Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}.
+Histograms require storing a fixed-size list of bins and corresponding counts per bin for ever growing volumes of data, satisfying property \textit{(a)}.
+Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and incrementing or decrementing the associated count, respectively, thus satisfying property \textit{(b)}.
 However, computing an histogram aggregation over a sliding window of size $W$ implies storing all $W$ elements in memory to know the element to remove after a new one is inserted.
@@
 A more thorough analysis on some statistical tests and why we chose to use the JSD is performed in Section \ref{sec:whyjsds}.
-Given a divergence value for the two aggregations (reference and sliding window ones) we need a measure of normal statistical fluctuations.
-Since we assume a reference period as the normal state we can find out the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
+Given a divergence value for the two aggregations (reference and sliding) we need a measure of normal statistical fluctuations.
+Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
 This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).
@@
 \begin{enumerate}
-    \item obtain a reference histogram for each feature using all data in the reference period
+    \item Obtain a reference histogram for each feature using all data in the reference period
 
-    \item then compute the distribution of distances between the reference histogram and the sliding window histogram by computing the distance (or divergence value) between the reference histogram and various samples of sliding window histograms
+    \item Compute the distribution of distances between the reference histogram and the sliding window histogram by computing the distance (or divergence value) between the reference histogram and various samples of sliding window histograms
 \end{enumerate}
@@
 It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself.
-For instance, given the JSD value varies between 0 and 1, why not just set to alert if the measured JSD value is 0.7?
+For instance, given that JSD values varie between 0 and 1, why not just set to alert if the measured JSD value is 0.7?
 The reason is that a divergence value may fall on different percentiles for different features.
@@
 \subsubsection*{Batch phase artifacts}
-For each feature, by the end of the batch phase, we have a reference histogram, a list of distance values that represents its distribution and the last sample's histogram to be used as a burn-in period for the sliding window or target histogram.
+By the end of the batch phase, we should have, for each feature: A reference histogram, a list of distance values that represent its distribution and the last sample's histogram to be used as a burn-in period for the sliding window or target histogram.
 
@@
 
-For each feature, we initialize the histogram we will approximately maintain with the last sample's data, from the batch phase.
+For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase.
 Then, for each incoming event and for each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
@@
 
-Periodically, we perform a multi-feature change detection test.
-The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the sliding window histogram of feature \textit{x}.
+In order to obtain alerts, we periodically perform a multi-feature change detection test.
+The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the sliding window histogram of the same feature.
 This distance value is then tested against the known distribution of distance values for feature \textit{x}.
-In other words, we test to see the percentile of the computed distance in the distribution of distance values and alert if above a certain percentile --- \textit{e.g.} the 99th-percentile.
+In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile --- \textit{e.g.} the 99th-percentile.
 When we perform the test, for each feature \textit{x}, we are computing the percentile of the distance measured between the reference and sliding window histogram aggregations.
@@
 
-Hence, if we set to alert states that have 1\% or less probability to happen, we raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01.
+For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01.
 Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
@@
 In other words, when repeating a test multiple times the multiple test problem arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
-In the literature, many multiple test correction methods are proposed, but we opt to use the Holm-Bonferroni multiple test correction.
+In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction.
 The Holm-Bonferroni method is one of many approaches for controlling the so called Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
@@
 
-We are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests.
+When checking for alerts we are performing one test for each feature, hence we performing \textit{n\textsubscript{features}} tests.
 We apply a multiple test correction on the tests performed and want to set the Family Wise Error Rate \textit{(FWER)} to $\alpha$.

### 2020-06-23 chapters/my-work.tex
SOTA improvements

@@
 \section{State of the art outlier detection methods}
-The first approach considered relied on outlier detection methods such as the ones presented in Section \ref{sec:outliers} and classified by the taxonomy in Figure \ref{fig:outlier-taxonomy}.
+Our first theoretical approach was to use one of the outlier detection methods presented in Section \ref{sec:outliers}.
 
@@
 \section{State of the art aggregation algorithms}
-This second approach focus on building an aggregation snapshot of the reference period to compare with streaming aggregation values and alert if the difference between both surpasses a threshold $\alpha$.
+Dissimilarity-based univariate subsequence outlier detection methods as presented in \ref{sec:uni-sub-out} focus on measuring the distance between two periods of time and report an outlier when the distance is above a certain threshold.
+Keeping both reference and target windows in memory means our memory consumption grows linearly regarding both windows size.
+In an attempt to solve this, we focus on storing only an aggregation of both windows.
 
-\subsection*{How}
+In Section \ref{sec:sota-swag-algs} we review algorithms to compute sliding window aggregations and in Sections \ref{sec:pds} and \ref{sec:sliding-pds} we discussed special aggregators and respective sliding window implementations.
 
-In this approach we compute sliding window aggregations and compare the aggregations’ current state versus an initial reference, reporting a shift when the difference is substantial.
+In our dissimilarity-based approach, we compute sliding window aggregations and compare the aggregations’ current state versus an initial reference, reporting a shift when the difference is substantial.
 For instance, consider the case where we monitor one single feature or variable \textit{x}.
@@
 \subsection*{Challenges}
-The first issue with this approach is that generic sliding window aggregation algorithms grow linearly in space regarding window size, such as Recalculate-From-Scratch and Subtract-On-Evict from Section \ref{sec:back-swag-algs}, Two-Stacks from Section \ref{sec:2stacks} and DABA in Section \ref{sec:daba}.
-Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.
+The first issue with this approach is that generic sliding window aggregation algorithms grow linearly in space regarding window size, such as Recalculate-From-Scratch and Subtract-On-Evict from Section \ref{sec:back-swag-algs}, Two-Stacks from Section \ref{sec:2stacks} and DABA in Section \ref{sec:daba} which does not meet our sub-linear memory growth requirement.
 
@@
 
-However, there is no trivial solution for the second challenge.
-Defining the set of aggregations to use does not have a straightforward solution.
+Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.
+\textcolor{red}{Marco had a problem with this example, dont understand what he meant}
 To illustrate this problem, consider both time-series show in Figure \ref{fig:approach2-timeseries}.
+Time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}.
+Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two different time-series.
 
@@
 \end{figure}
-
-\textcolor{red}{Marco had a problem with this example, dont understand what he meant}
-Both time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}.
-Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two different time-series.
 

### 2020-06-23 chapters/sota.tex
SOTA improvements

@@
 A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}.
-In this Chapter, we discuss the most relevant work for the aforementioned problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems.
+In this Chapter, we discuss the most relevant work for the problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems, presented in Chapter \ref{chap:intro}.
 We present a wide analysis of several categories of algorithms that seem viable to build the desired system.
@@
 \section{Outlier detection in time-series data} \label{sec:outliers}
-Blázquez-García et.
-al review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
-From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented in \textit{"A review on outlier/anomaly detection in time-series data"} by Blázquez-García et.
-al due to its focus on time-series data.
-
-A time-series is a collection of data points sorted by time.
-Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}.
-This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
-generated by a different mechanism"} \cite{Hawkins-Outliers}.
-Furthermore, Blázquez-García et.
-al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers.
+Blázquez-García \emph{et al.} review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
+From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.
+
+In this work, the authors define a time-series as a collection of data points sorted by time; and outliers as observations that do not follow the expected behavior.
+This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is “an observation which deviates so much from other observations as to arouse suspicions that it was
+generated by a different mechanism" \cite{Hawkins-Outliers}.
+Furthermore, Blázquez-García \emph{et al.} allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers.
 In summary, outlier detection refers to the process of finding anomalies in data.
@@
 The authors evaluate an outlier detection method on three different axes: the input data type, the outlier type to detect and the uni or multivariate nature of the method.
-Figure \ref{fig:outlier-taxonomy} presented by the authors summarizes these categories.
+Figure \ref{fig:outlier-taxonomy} presented by the authors in \cite{Blazquez-Garcia-Review-Anomaly-Detection} summarizes these categories.
 
@@
 
-Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy}).
+Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy} \cite{Blazquez-Garcia-Review-Anomaly-Detection}) .
 
@@
 If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier.
-Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if
+Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if:
 \begin{equation*}
@@
 The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values.
-Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can not.
+Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can only be used in batch.
 
@@
 al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios.
-The authors propose Extra-N, an algorithm that detects point outliers based on the number of neighbors it has.
-The authors make use of the \textit{“predictability"} property of the expiration of existing objects.
+The authors propose \emph{Extra-N}, an algorithm that detects point outliers based on the number of neighbors it has.
+The authors make use of the \emph{predictability} property of the expiration of existing objects.
 In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to.
 
-The authors want to maintain for each data point \textit{p\textsubscript{i}} the number of neighbors it has, rather than the actual list of neighbors.
+The authors of Extra-N want to maintain for each data point $p\textsubscript{i}$ the number of neighbors it has, rather than the actual list of neighbors.
 The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors.
@@
 The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors.
-The authors introduce the concept of \textit{"lifetime neighbor counts"} or \textit{lt\_cnt}.
-The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where he exists.
+The authors introduce the concept of \textit{lifetime neighbor counts} or \textit{lt\_cnt}.
+The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where it exists.
 More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window.
-The authors provide the following example: consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors, \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}.
-Applying the \textit{predictability} property makes it possible to compute the lifespan of all points.
+Consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors: \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}.
 Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}.
 This implies that at \textit{W\textsubscript{i}},
-\[ \textit{p\textsubscript{i}.lt\_cnt} = \{\textit{W\textsubscript{i}}: 3, \textit{W\textsubscript{i+1}}: 2, \textit{W\textsubscript{i+2}}: 0\} \]
+\[ p\textsubscript{i}.lt\_cnt = \{W\textsubscript{i}: 3, W\textsubscript{i+1}: 2, W\textsubscript{i+2}: 0\} \]
 The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.
@@
 Extra-N can take both uni or multivariate time-series as input data.
-However, it does not relate multiple time-series and does not take into account possible dependencies between different time-dependent variables.
+However, it does not relate multiple time-series and it does not take into account possible dependencies between different time-dependent variables.
 In that sense, and according to the presented taxonomy (Figure \ref{fig:outlier-taxonomy}), the method accepts uni or multivariate time-series as input but is of univariate nature.
@@
 \subsubsection*{Applicability to our Hypothesis}
-Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable methods to accomplish our goal of building a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again.
-Additionally, insights produced by machine learning models are not human-understandable, rendering the user unable to understand why an anomaly report was made.
+Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable to our goals.
+In fact, we need a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again.
+Additionally, insights produced by machine learning models are not human-understandable, leaving the user confused on why an anomaly report was made.
 In conclusion, we do not deem HTM applicable to our solution.
@@
 Munir et.
-al present DeepAnT, a deep learning-based approach for the detection of anomalies in time series data \cite{Munir-DeepAnT}, that uses unlabeled data to capture and learn the distribution of the data stream used to forecast the normal behavior of the time-series.
-It is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known.
+al present DeepAnT, a deep learning-based approach for the detection of anomalies in time-series data \cite{Munir-DeepAnT}.
+DeepAnT uses unlabeled data to capture and learn the distribution of the data stream used to forecast the normal behavior of the time-series.
+This method is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known.
 It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.
@@
 
-\subsection{Univariate subsequence outlier detection}
-Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones (Figure \ref{fig:outlier-taxonomy}).
+\subsection{Univariate subsequence outlier detection} \label{sec:uni-sub-out}
+Similarly to point outlier detection methods, Blázquez-García \emph{et al.} \cite{Blazquez-Garcia-Review-Anomaly-Detection} split subsequence outlier detection methods into uni and multivariate ones (Figure \ref{fig:outlier-taxonomy}).
 Furthermore, the authors split univariate subsequence outlier detection methods into five different axes as seen in Figure \ref{fig:subsequence-outlier-taxonomy}.
@@
 These methods find the most unusual subsequence in a time-series.
-However, as the authors put it, these types of methods \textit{"do not have a reference of normality or a threshold"} and hence they do not accurately report outliers.
+However, as the authors put it, this type of methods lack a reference or normality state and hence do not accurately report outliers.
 
@@
 Similarly to univariate point outlier detection methods based on prediction models, these models train on past data and make predictions of the future.
-Subsequences that do not resemble the prediction made by the model are classified as outliers.
+Subsequences that deviate from the model's prediction are classified as outliers.
 
@@
 This type of method not only measures the frequency of subsequences but relates it with the information carried by the subsequence.
-Mathematically, a subsequence \textit{S} will be an outlier if
+Mathematically, a subsequence $S$ will be an outlier if
 
@@
 
-with \textit{F(S)} as the frequency of subsequence S, \textit{I(S)} as the information carried by S and $\alpha$ as a given threshold.
+with $F(S)$ as the frequency of subsequence $S$, $I(S)$ as the information carried by $S$ and $\alpha$ as a given threshold.
 According to the authors, \textit{"I(S) is computed taking into account the number of times the symbols within S are repeated through the series"}.
@@
 \subsubsection{A two window paradigm algorithm for change detection}
-Kifer et al.
-propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions \cite{Kifer-Detecting-Change}.
+Kifer \emph{et al.} propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions \cite{Kifer-Detecting-Change}.
 The proposed method uses two windows, a reference \textit{W\textsubscript{1}} and a sliding one \textit{W\textsubscript{2}}.
@@
 It is trivial to adapt this method to always use the same reference period.
-To do so, we simply keep the original reference window and don't update it after each alert.
+To do so, we simply keep the original reference window and do not update it after each alert.
 
@@
 \subsubsection*{Method classification}
-The method devised by Kifer et.
-al can take uni or multivariate input (if we use one pair of reference/target windows for each time-series) but performs a univariate analysis as it does not find correlations between time-dependent variables.
+The method devised by Kifer \emph{et al.} can take uni or multivariate input (if we use one pair of reference/target windows for each time-series) but performs a univariate analysis as it does not find correlations between time-dependent variables.
 This two-windowed method measures the difference between both windows, the reference and the sliding one.
@@
 However, this two-window paradigm encodes a static normal or reference period in one window and the observed streaming values in a sliding one.
-Since in this Thesis we want to report streaming data pattern shifts relatively to a reference period, we may use a similar two-windowed model if we find lightweight and incrementally maintained aggregations over both windows.
+Since in this Thesis we want to report streaming data pattern shifts relatively to a reference period, we use a similar two-windowed model, but use lightweight and incrementally maintainable aggregations over both windows.
 
 \subsubsection{Automatic Model Monitoring for Data Streams}
-
 ALSO a 2 WINDOW MODEL
@@
 The De-Amortized Banker’s Aggregator (DABA) \cite{Tangwongsan-DABA} is a SWAG algorithm that allows the computation of \textit{associative} aggregations.
-Quoting the authors: \textit{"Deamortization is a method that turns the average-case behavior into the worst-case behavior, usually by carefully spreading out expensive operations"}.
-Hence the name \textit{"De-Amortized"}, as the main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout run time.
+This method carefully handles expensive operations and spreads them along the execution.
+As a resut, the authors are able to turn the average-case behavior into the worst-case behavior.
+This is commonly known as \textit{de-amortization}, hence the name.
+The main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout run time.
 
@@
 
-DABA uses two queues, \textit{vals} and \textit{aggs}, both of size \textit{n}, seen in Figure \ref{fig:daba-ds}, which was directly taken from \textit{"Low-Latency Sliding-Window Aggregation in Worst-Case Constant Time"} by Tangwongsan et.
-al.
-Both queues share six pointers --- \textit{F, L, R, A, B} and \textit{E} --- that point to the same position for each queue.
-These pointers are always ordered as follows: $\textit{F} \leq \textit{L} \leq \textit{R} \leq \textit{A} \leq \textit{B} \leq \textit{E}.$
+DABA uses two queues, \textit{vals} and \textit{aggs}, both of size \textit{n}, seen in Figure \ref{fig:daba-ds}, which was directly taken from \textit{"Low-Latency Sliding-Window Aggregation in Worst-Case Constant Time"} by Tangwongsan \emph{et al.} Both queues share six pointers --- \textit{F, L, R, A, B} and \textit{E} --- which point to the same position for each queue.
+These pointers are always ordered as follows: $F \leq L \leq R \leq A \leq B \leq E$.
 
@@
 
-\textit{l\textsubscript{F}} is aggregated to the left for easier eviction.
+Figure \ref{fig:daba-ds} shows sub-lists \textit{l\textsubscript{F}}, \textit{l\textsubscript{B}}, \textit{l\textsubscript{L}}, \textit{l\textsubscript{R}} and \textit{l\textsubscript{A}}.
+Sub-list \textit{l\textsubscript{F}} is aggregated to the left for easier eviction.
 After an element is evicted, the new head of the queue \textit{aggs} --- pointed by \textit{F} --- already contains the aggregation value of all elements from \textit{vals} to its right side.
@@
       \includegraphics[scale=0.4]{figures/query-bloom.png}
-      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one '0' indicates that it is not in the set.}
+      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
       \label{fig:bloom-filter}
@@
 It is assumed that the hash functions produce an independent and random index value for each element.
-Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula.
-
-The probability of one arbitrary bit being set to 1 is
+
+Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula. The probability of one arbitrary bit being set to 1 is:
 \begin{equation}
@@
 
-Mullin states the probability that an arbitrary bit is not set to 1 after one element insertion and corresponding \textit{k} bit updates is
+Mullin states the probability that an arbitrary bit is not set to 1 after one element insertion and corresponding \textit{k} bit updates is:
 \begin{equation}
@@
 
-For \textit{n} insertions, the probability a bit is not set to 1 is
+For \textit{n} insertions, the probability a bit is not set to 1 is:
 \begin{equation}
@@
 
-Therefore, the probability that an arbitrary bit is set to 1 after \textit{n} insertions is 
+Therefore, the probability that an arbitrary bit is set to 1 after \textit{n} insertions is:
 \begin{equation}
@@
 
-This false positive rate can be approximated by 
+Finally, this false positive rate can be approximated by 
 \begin{equation}
@@
 In order to do so, we must first allocate \textit{m} bits for the Bloom filter, choosing \textit{m} based on available memory.
-The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by
+The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by:
 \begin{equation}
@@
 
-Note the time complexity does not at all depend on the number of elements in it and that \textit{k} will be a rather small constant. Hence, the Bloom filter is said to have constant time complexity.
+Note that the time complexity does not at all depend on the number of elements processed by the Bloom filter and that \textit{k} will be a rather small constant.
+Hence, the Bloom filter is said to have constant time complexity.
 
@@
 A CMS is represented by a two-dimensional array (or matrix) with width \textit{w} and depth \textit{d}.
-A CMS will make use of \textit{d} hash functions --- \textit{i.e.} each associated with one row of the matrix.
-Each hash function will map elements to a position between 1 and \textit{w} --- \textit{i.e.} the column of the two-dimensional array.
+A CMS will make use of \textit{d} hash functions, each associated with one row of the matrix.
+Each hash function will map elements to a position between 1 and \textit{w}, the column of the two-dimensional array.
 Initially, all of the matrix positions are 0.
@@
 \subsection*{Element Insertion}
-When adding an element, for each row \textit{i} of the \textit{d} rows in the matrix, hash the element using that row's hash function to obtain \textit{j}.
+When adding an element, for each row \textit{i} of the \textit{d} rows in the matrix, we hash the element using that row's hash function to obtain the result $j$.
 Lastly, for all obtained pairs of \textit{i} and \textit{j}, increment matrix cells \textit{(i, j)} value by 1, as seen in Figure \ref{fig:cms}.
@@
       \includegraphics[scale=0.5]{figures/insertion-cms.png}
-      \caption[Count-Min Sketch insertion]{CMS mapping element \textit{el} using \textit{d}=4 hash functions to determine the column position for each row and increment it}
+      \caption[Count-Min Sketch insertion]{CMS mapping element \textit{el} using \textit{d}=4 hash functions to determine the column position for each row and increment it.}
       \label{fig:cms}
@@
 Retrieving the minimum frequency of an element \textit{el} can be done by taking the minimum value of all row counts for \textit{el}.
-Mathematically, the frequency estimate is the minimum value of the set of all row values, for each row \textit{i} of the total \textit{d} rows, given by
+Mathematically, the frequency estimate is the minimum value of the set of all row values, for each row \textit{i} of the total \textit{d} rows, given by:
 \[ min \{matrix[i][\textit{h\textsubscript{i}}(el)]\} \]
@@
 In HyperLogLog, new elements are hashed into a binary string.
-The first \textit{n} bits of the hashed representation will map to one of the \textit{m}=$2^n$ buckets.
+The first \textit{n} bits of the hashed representation will map to one of the $m=2^n$ buckets.
 Effectively speaking, the first \textit{n} bits work as a \textit{bucket ID}.
@@
 Figure \ref{fig:hll-insert} demonstrates such procedure for a new element \textit{el}.
-The hash representation of \textit{el}, using \textit{n}=2 bits, maps to bucket \textit{10}.
-Discarding the first \textit{n}=2 bits and counting from 1, the position of the leftmost 1-bit is \textit{P}=4.
-The previous value of bucket \textit{10} was 0 and because \textit{P}=4 is greater than 0 the bucket value is updated.
+The hash representation of \textit{el}, using $n=2$ bits, maps to bucket \textit{10}.
+Discarding the first $n=2$ bits and counting from 1, the position of the leftmost 1-bit is $P=4$.
+The previous value of bucket \textit{10} was 0 and because $P=4$ is greater than 0 the bucket value is updated.
 

### 2020-06-23 chapters/sota.tex
SOTA Sofia feedback applied

@@
 Additionally, for each new data point \textit{p\textsubscript{i}}, Extra-N has to perform a range query search centered on it to find its neighbors and increment their lifetime neighbor counts by one.
-The run-time complexity of a range query search is \textit{O(log(n))} \cite{range-query-log} with \textit{n} as the number of data points.
+The runtime complexity of a range query search is \textit{O(log(n))} \cite{range-query-log} with \textit{n} as the number of data points.
 
@@
 However, when working with sliding windows there is the need not only to insert new events but to evict old ones.
-Probabilistic data structures aggregate all incoming data but have no way of expiring old elements from the aggregation state.
+The probabilistic data structures described until now can aggregate new incoming data but do not address the type of temporality needed for sliding windows --- \textit{i.e.} the ability to expire old elements from the aggregated state.
 
@@
 The work of Chabchoub et al. adapts the original HLL aggregator from Flajolet et al. to work under a sliding window framework  \cite{Chabchoub-Sliding-HLL}.
-The main idea behind this implementation is to replace every HLL counter with a \textit{"List of Future Possible Maxima (LFPM)"}.
+The main idea behind this implementation is to replace every HLL counter with a \textit{List of Future Possible Maxima (LFPM)}.
 
@@
 Since each bucket now holds a LFPM instead of a single integer, the procedure for bucket update is different.
-Upon insertion of an element \textit{k}, a packet <t\textsubscript{k}, p\textsubscript{k}> is created with t\textsubscript{k} as the timestamp of arrival of the new element \textit{k} and p\textsubscript{k} as the position of the leftmost 1-bit position of \textit{k}'s binary representation.
+Upon insertion of an element \textit{k}, a packet $<t_k, p_k>$ is created with t\textsubscript{k} as the timestamp of arrival of the new element \textit{k} and p\textsubscript{k} as the position of the leftmost 1-bit position of \textit{k}'s binary representation.
 Given the new packet <t\textsubscript{k}, p\textsubscript{k}>, the authors propose the following LFPM update method, that checks for and evicts old packets at each insertion:
@@
 The authors claim that the proposed sliding version remains as accurate as of the original HLL aggregator.
-In the context of their work, each element in the set is denoted as a "flow" and they state that \textit{"In practice, with an additional memory of only 35kB, a standard error of about 3\% can be achieved for a data stream of several million flows"}.
+With an additional memory of only 35kB it is possible to estimate the number of distinct elements in the set with a standard error of 3\% in the presence of up to a million distinct elements.
 
@@
 The approach proposed by Xu uses a smaller amount of memory that does not change at run time \cite{Xu-hll-sliding-drv}.
-The author proposes a special counter: a \textit{"Distance Recorder Vector (DRV)"}.
+The author proposes a special counter: a \textit{Distance Recorder Vector (DRV)}.
 
@@
 
-Initially, all records in the DRV will be set to be infinity or a large number.
+Initially, all records in the DRV will be set to be infinity or a very large number.
 
@@
 When an element is inserted, the same procedure as in the original HLL is followed: the element is hashed and \textit{p} is computed from the hashed binary representation.
-Denote the bit-pattern observable of the new element as \textit{p\textsubscript{new}}.
-For the DRV of that bucket, increment the value of all records by one.
-Next, set position \textit{p\textsubscript{new}} in the DRV to 0 (DRV[\textit{p\textsubscript{new}}] = 0).
+Let us denote the bit-pattern observable of the new element as \textit{p\textsubscript{new}}.
+For the DRV of that bucket, we increment the value of all records by one.
+Next, we set position \textit{p\textsubscript{new}} in the DRV to 0 --- \textit{i.e.} $\textrm{DRV}[p_{new}] = 0$.
 
 \subsubsection*{Sliding Window Cardinality Query} 
-Element eviction is not performed.
-Instead, the query algorithm takes into account that some records in some of the DVR's might not be valid anymore.
+This method does not perform element eviction.
+Instead, the query algorithm assumes that some records in DVR's entries might be expired.
 To estimate the sliding window cardinality is to find the value \textit{p}.
 
-To find the value \textit{p} for a bucket, scan the DRV of that bucket backward.
-Start at the last element --- positioned at \textit{$\lceil ln(n/m) \rceil$} --- and scan towards the head of the vector.
-For each record, check if it is active in the current window.
-To do so, check if its value is less than the size of the window.
+To find the value \textit{p} for a bucket, we scan the DRV of that bucket backward.
+For that, we start at the last element --- positioned at \textit{$\lceil ln(n/m) \rceil$} --- and scan towards the head of the vector.
+For each record, we check if it is active in the current window.
+To do so, we check if its value is less than the size of the window.
 If it is, the record is active.
@@
 
-If the record is inactive continue scanning backward.
+If the record is inactive we continue scanning backward.
 If the record is active, then the position or index currently being scanned is the value of \textit{p} for that bucket.
@@
 Both inserting an element and estimating sliding window cardinality require a scan through \textit{$\lceil ln(n/m) \rceil$} records.
-This gives us a time complexity of \textit{O(\textit{ln(n)})}, given the number of buckets \textit{m} is constant.
+This gives us a time complexity of \textit{O(\textit{ln(n)})}, if the number of buckets \textit{m} is constant.
 

### 2020-06-24 chapters/background.tex
Marco feedback

@@
 Batch data sets are known as static data \cite{Martin-Batch-Defin}.
-An example application of such a technique would be processing all the keywords searched for in the google search engine by grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes.
+An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes).
 Batch processing is the oldest paradigm between batch and stream processing.
@@
 Why might we need to convert it into a bounded data set and how to do so?
-In this Section, we will analyze such issues and present streaming windows as the solution.
+In this Section, we will analyze such issues and present sliding windows as the solution.
 
@@
 For the scope of this Thesis, we will work under a tuple-based true sliding window framework.
-However, as explained in Section \ref{sec:ft-monitoring}, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.
+However, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.
 
@@
 As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}.
-In this example, we wanted to compute the average temperature per rack in the data cencer in the last 5 minutes.
+In this example, we wanted to compute the average temperature per rack in the data center in the last 5 minutes.
 If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:
@@
 This recursive property allows us to store only one EMA value and continuously update it for each new value val\textsubscript{t}, based on the following recursion formula:
-
 \begin{definition}
-EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda\delta\textsubscript{t}$ + val\textsubscript{t}
+EMA\textsubscript{t} = EMA\textsubscript{t-1} * $2^{-\frac{\delta_t}{\tau_{1/2}}}$ + val\textsubscript{t}
 \label{def:ema}
 \end{definition}
+where $\delta_t$ is the time passed between \textit{t} and \textit{t-1} and $\tau_{1/2}$ is the time-based half-life of the EMA.
 
-with $\lambda$ as a constant factor applied to the previous EMA value and $\delta\textsubscript{t}$ as the time passed between \textit{t} and \textit{t-1}.
-This $\lambda$ constant (0 < $\lambda \leq$ 1) is known as the smoothing coefficient.
-The smoothing or decay factor $\lambda$ is given by:
-\begin{equation}
-    \lambda = 2^{- \dfrac{1}{\tau}}
-    \label{eq:ema-decay}
-\end{equation}
-where $\tau$ is the half-life of the EMA.
-The half-life $\tau$ controls how fast old events are forgotten as the smoothing factor depends on it.
+The half-life $\tau_{1/2}$ controls how fast old events are forgotten.
 The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half.
-For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half.
+For instance, with $\tau_{1/2} = 1s$, for each $\delta_t = 1s$ step, we reduce the weights applied to each event by half.
 
-Note that it is also possible to express a tuple-based half-life, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half.
+Note that it is also possible to express a tuple-based half-life $n_{1/2}$, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half.
 In this case, the definition does not take into account the time passed between events:
 \begin{definition}
-EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda$ + val\textsubscript{t}
+EMA\textsubscript{t} = EMA\textsubscript{t-1} * $2^{-\frac{1}{n_{1/2}}}$ + val\textsubscript{t}
 \label{def:tuple-ema}
 \end{definition}
-For example, with a tuple-based half-life of $\tau = 1$ tuple, the smoothing factor is $\lambda = 0.5$, meaning for each new tuple all weights are reduced by half.
+For example, with a tuple-based half-life of $\tau = 1$ tuple, for each new tuple all weights are reduced by half.
 

### 2020-06-24 chapters/introduction.tex
Marco feedback

@@
 The information that flows through such a stream can be analyzed on-the-fly or stored for later processing.
-The former is best known as stream processing, whereas the latter as batch processing.
+The former is best known as stream processing, whereas the latter is known as batch processing.
 
@@
 Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not.
-ML models are configured and trained before deployment making assumptions on the statistical distributions of future incoming data.
-Thus their model performance is linked to the maintenance of these assumptions.
+ML models are configured and trained,  before deployment, often with assumptions on the statistical distributions of future incoming data.
+Thus their model performance depends on how well those assumptions hold for future data.
 However, fraud patterns are not static over time, but seasonal --- \textit{e.g.} Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.} new fraud attack methods by fraudsters.
@@
 Many analytical systems are static once deployed and are configured \textit{a priori} under the assumption that future data flowing through the system will roughly follow the same distribution as previously seen data.
-The problem is that data pattern shifts will lead to poor performance by the previously configured system.
+The problem is that data pattern shifts will lead to poor performance of the previously configured system.
 Monitoring the data patterns themselves and alerting for changes in the underlying distribution would allow users to reconfigure the system before its performance declines.
@@
 The question this Thesis aims to answer is whether sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.
+In this Thesis, we study some sliding window aggregations that do not keep the sliding window in memory but still give an estimate on the aggregation value.
 

### 2020-06-24 chapters/my-work.tex
Marco feedback

@@
 Our first theoretical approach was to use one of the outlier detection methods presented in Section \ref{sec:outliers}.
-
-We reviewed subsequence and point outlier detection methods.
+For that, we reviewed subsequence and point outlier detection methods.
 Using point outlier detection methods, the main idea is to keep track of the rate of point outliers detected.
@@
 
-In Section \ref{sec:sota-swag-algs} we review algorithms to compute sliding window aggregations and in Sections \ref{sec:pds} and \ref{sec:sliding-pds} we discussed special aggregators and respective sliding window implementations.
+In Section \ref{sec:sota-swag-algs} we reviewed algorithms to compute sliding window aggregations and in Sections \ref{sec:pds} and \ref{sec:sliding-pds} we discussed special aggregators and respective sliding window implementations.
 
@@
 In this case, we would measure the difference between the sliding window aggregations and the reference window ones and if above a certain threshold raise an alert.
-For example, if we used a threshold of $\alpha$ = 6, we would raise an alert, as at least one sliding aggregation, in this case the mean, would differ of at least $\alpha$ when compared to the reference value.
+For example, if we used a threshold of $t = 6$, we would raise an alert, as at least one sliding aggregation, in this case the mean, would differ of at least $t$ when compared to the reference value.
 
@@
 Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.
-\textcolor{red}{Marco had a problem with this example, dont understand what he meant}
 To illustrate this problem, consider both time-series show in Figure \ref{fig:approach2-timeseries}.
@@
 
-A data stream is a continuous collection of events.
+In Section \ref{sec:stream-superset}, we defined a data stream as a continuous collection of events.
 Each event is timestamped and contains multiple fields.
-In the context of this Thesis, each of the event's fields are time-dependent variables named features.
-They can be thought of as columns in a traditional relational database table.
+In the context of this Thesis, each of the event's fields are named features.
 Our goal is to detect data pattern shifts for each feature in an streaming fashion.
-Such shifts will be measured between a reference period versus the observed reality during the streaming phase.
+Such shifts will be measured between a static in time or reference period versus the observed reality during the streaming phase.
 Figure \ref{fig:timelines} shows the separated reference and streaming timelines.
@@
 
-We do not need to directly compare the entire window contents of the reference and sliding windows.
-The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size.
+To detect data pattern shifts, we do not need to directly compare the entire window contents of the reference and sliding windows.
+This approach would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size.
 Instead, we want to use a  \textit{(a)} fixed cost in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
@@
 
-We devise a two-phase method.
+To implement the proposed methodology, we devised a two-phase method.
 The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}).
-The second phase (detailed in Section \ref{sec:stream-phase}) focus on efficiently updating the sliding window histogram for each feature and periodically performs a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be.
+The second phase (detailed in Section \ref{sec:stream-phase}) focuses on efficiently updating the sliding window histogram for each feature and periodically performing a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be.
 The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.
@@
 As stated, we compute histogram aggregations for the reference and sliding windows.
-The computation of the reference histogram is done once in batch and hence an exact histogram can be computed.
+The computation of the reference histogram is done once in batch.
+Since the reference data is fixed and does not change over time, computing this histogram in batch is more efficient and easier to implement in a distributed system like Spark.
+In addition, a batch computation also allows us to compute an exact reference histogram.
 However, the sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.
@@
 In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they compute sliding window aggregations without actually storing the window contents.
-Definition \ref{def:tuple-ema} shows how to compute an EMA using its recurrence formula.
-An EMA state or value depends only on the previous one times an exponential decay $\lambda$.
+Definition \ref{def:tuple-ema} shows how to compute an EMA using its recursive formula.
+An EMA state or value depends only on the previous one multiplied by an exponential decay $\lambda$.
 Recall that this decay is given by Equation \ref{eq:ema-decay}:
@@
 In fact, we set the half-life to be one quarter of the size of the window we want to emulate.
-For instance, if we want to compute an EMA aggregation over a period of a month, we set the half-life to be the number of tuples that correspond to one week worth of data.
+For instance, if we want to monitor a period extending up to one month (at an error level of about 6\%) we setr the half-life to be the number of tuples that correspond to one week worth of data.
+This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.
 
@@
 
-The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and the sliding window aggregations.
+The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and EMA-like aggregations.
 In other words, we:
@@
 Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase.
-To that end, we make \textit{S} samples of transactions, each with the same tuple-based size of the sliding window we will use in streaming.
+To that end, we make $S$ samples of transactions (in Section \ref{sec:nsamples} we discuss the optimal number $S$ of samples to make), each with the same tuple-based size of the sliding window we will use in streaming.
 Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
@@
 For each feature, we end up with \textit{S} distance values.
-These values are distance measurements between random samples of data and the reference period.
+These $S$ measurements are distance measurements between random samples of data and the reference period.
 Hence, we claim we now know the distribution of expected distance values.
-Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if below a certain threshold.
+Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if the probability is below a certain threshold.
 Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the sliding window histogram.
@@
 It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself.
-For instance, given that JSD values varie between 0 and 1, why not just set to alert if the measured JSD value is 0.7?
+For instance, given that JSD values vary between 0 and 1, why not just set to alert if the measured JSD value is 0.7?
 The reason is that a divergence value may fall on different percentiles for different features.
@@
 
-Let's assume Figure \ref{fig:skewed-left-distro} 's histogram encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}.
+Let's assume that the histogram from Figure \ref{fig:skewed-left-distro} encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}.
 Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results.
 For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}.
-Instead what we aim to do is know this distribution so that we can define our threshold as a percentile and not a hard-coded distance constant.
+Instead we define our threshold as a percentile of the distribution and not as a hard-coded distance constant.
 For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
@@
 \subsubsection*{Batch phase artifacts}
-By the end of the batch phase, we should have, for each feature: A reference histogram, a list of distance values that represent its distribution and the last sample's histogram to be used as a burn-in period for the sliding window or target histogram.
+By the end of the batch phase, we should have, for each feature: 
+\begin{itemize}
+    \item a reference histogram
+    \item a list of distance values that represent its distribution
+    \item the last sample's histogram, to be used as burn-in period for the sliding window histogram
+\end{itemize}
 
@@
 This distance value is then tested against the known distribution of distance values for feature \textit{x}.
-In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile --- \textit{e.g.} the 99th-percentile.
-When we perform the test, for each feature \textit{x}, we are computing the percentile of the distance measured between the reference and sliding window histogram aggregations.
-The probability value or \textit{p-value} of a distance value is
+In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile, \textit{e.g.}, the 99th-percentile.
+When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and sliding window histogram aggregations.
+The probability value or \textit{p-value} of a distance value is:
 
@@
 Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
-
 \[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]
-
-or
-
+or:
 \begin{equation}
@@
 The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.
-This correction is done by dividing the threshold $\alpha$ by a correction factor, as described below.
+This correction is done by dividing the threshold $\alpha$ by a correction factor, as described next.
 
@@
 We then order our features by \textit{p-value}, in ascending order.
-Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if
-
+Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if:
 \[  \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha / (\textit{m} - \textit{k}) \]
-
 with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature.
@@
 
-\subsubsection*{Optimal number of samples}
+\subsubsection*{Optimal number of samples} \label{sec:nsamples}
 In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase.
@@
 
-When checking for alerts we are performing one test for each feature, hence we performing \textit{n\textsubscript{features}} tests.
+When checking for alerts we are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests.
 We apply a multiple test correction on the tests performed and want to set the Family Wise Error Rate \textit{(FWER)} to $\alpha$.
@@
 
-\textcolor{red}{SEE MODEL MONITORING PAPER} We can then show that the optimal number of samples \textit{S} to make is
-
+\textcolor{red}{SEE MODEL MONITORING PAPER} We can then show that the minimum number of samples $S$ needed to satisfy these probability requirements is:
 \begin{equation}
     \label{eq:optimal-n-samples}
-    \textit{S} = \frac{\log\gamma}{log(1 - \frac{\alpha}{n\textsubscript{features}})}
+    \textit{S} = \frac{\log\gamma}{log(1 - \frac{\alpha}{n_{features}})}
 \end{equation}

### 2020-06-24 chapters/sota.tex
Marco feedback

@@
 The univariate point outlier detection methods based on models are the most common.
-For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value.
-If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier.
-Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if:
+For each timestamp of the time-series, the models compute the predicted value for the time-dependent variable and compare it with the observed value.
+If the difference between the predicted value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier.
+Mathematically, given the predicted value $p\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if:
 \begin{equation*}
-    |x\textsubscript{t}-e\textsubscript{t}|>\alpha
+    |x\textsubscript{t}-p\textsubscript{t}|>\alpha
 \end{equation*}
@@
 Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting.
-These methods consider a datum an outlier if it has less than $\alpha$ neighbors.
+For example, nearest neighbor based methods consider a datum an outlier if it has less than $\alpha$ neighbors.
 Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}.
@@
 
-\subsubsection{Automatic Model Monitoring for Data Streams}
-ALSO a 2 WINDOW MODEL
-\textcolor{red}{TODO: write this}
+%\subsubsection{Automatic Model Monitoring for Data Streams}
+%ALSO a 2 WINDOW MODEL
+%\textcolor{red}{TODO: write this}
 
@@
 \subsubsection*{Sliding Window Cardinality Query} 
-This method does not perform element eviction.
-Instead, the query algorithm assumes that some records in DVR's entries might be expired.
+This method does not perform element eviction. Instead, the query algorithm assumes that some records in DVR's entries might be expired.
+
 To estimate the sliding window cardinality is to find the value \textit{p}.
-
 To find the value \textit{p} for a bucket, we scan the DRV of that bucket backward.

### 2020-06-24 chapters/my-work.tex
Method description done

@@
 For instance, keep track of how many outliers were alerted in the past hour.
-The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$.
-On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier when compared to the reference window.
+The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold.
+On the other hand, using subsequence outlier detection methods would allow us to report the whole target sliding window as an outlier when compared to the reference window.
 
@@
 \section{State of the art aggregation algorithms}
-Dissimilarity-based univariate subsequence outlier detection methods as presented in \ref{sec:uni-sub-out} focus on measuring the distance between two periods of time and report an outlier when the distance is above a certain threshold.
+Dissimilarity-based univariate subsequence outlier detection methods as presented in Section \ref{sec:uni-sub-out} focus on measuring the distance between two periods of time and report an outlier when the distance is above a certain threshold.
 Keeping both reference and target windows in memory means our memory consumption grows linearly regarding both windows size.
@@
 Assume we define the set of aggregations to compute the mean of \textit{x} and its standard deviation.
-Also assume we save the reference window and maintain a sliding target window and the associated average and standard deviation aggregations for \textit{x} efficiently.
+Also assume we save the reference window and maintain a target sliding window and the associated average and standard deviation aggregations for \textit{x} efficiently.
 As seen in Figure \ref{fig:approach2-initial-state}, the initial mean and standard deviation values for \textit{x} presented in the reference window are of 5 and 1, respectively.
@@
 
-In Figure \ref{fig:approach2-alert-state}, after further sliding steps of our sliding window, we obtain \textit{Window 2}.
+In Figure \ref{fig:approach2-alert-state}, after further sliding steps of our target sliding window, we obtain \textit{Window 2}.
 In \textit{Window 2}, the mean and standard deviation aggregation values change from 5 and 1 to 12 and 6, respectively.
-In this case, we would measure the difference between the sliding window aggregations and the reference window ones and if above a certain threshold raise an alert.
+In this case, we would measure the difference between the target sliding window aggregations and the reference window ones and if above a certain threshold raise an alert.
 For example, if we used a threshold of $t = 6$, we would raise an alert, as at least one sliding aggregation, in this case the mean, would differ of at least $t$ when compared to the reference value.
@@
 
-Exponential Moving Averages (Section \ref{sec:emas}), Probabilistic Data Structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}) can be used to solve the first issue and reduce the linear memory complexity of the system relative to sliding window size.
+Exponential Moving Averages (Section \ref{sec:emas}), Probabilistic Data Structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}) can be used to solve this first issue and reduce the linear memory complexity of the system relative to sliding window size.
 
@@
     \begin{center}
-      \includegraphics[scale=0.4]{figures/timeperiods.png}
-      \caption[]{Reference and Streaming timelines}
+      \includegraphics[scale=0.7]{figures/timeperiods.png}
+      \caption[]{Reference and streaming timelines}
       \label{fig:timelines}
@@
 To detect data pattern shifts, we do not need to directly compare the entire window contents of the reference and sliding windows.
-This approach would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size.
+This approach would require storing all window contents hence making the memory consumption grow linearly with target sliding window and/or reference period size.
 Instead, we want to use a \textit{(a)} fixed cost in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
@@
 Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}.
-Histograms require storing a fixed-size list of bins and corresponding counts per bin for ever growing volumes of data, satisfying property \textit{(a)}.
+Histograms require storing a fixed-size list of bins and corresponding counts per bin regardless of the volume of data, satisfying property \textit{(a)}.
 Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and incrementing or decrementing the associated count, respectively, thus satisfying property \textit{(b)}.
-However, computing an histogram aggregation over a sliding window of size $W$ implies storing all $W$ elements in memory to know the element to remove after a new one is inserted.
+Recall Section \ref{sec:back-swag-algs} where we discussed generic sliding window aggregation algorithms and remember the Subtract-On-Evict (SOE) algorithm (Algorithm \ref{pseudo:soe}).
+Computing an histogram aggregation over a sliding window of size $W$ using SOE implies storing all $W$ elements in memory to know the element to remove after a new one is inserted.
 This violates our constant in memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}.
@@
     \begin{center}
-      \includegraphics[scale=0.4]{figures/timelines-hists.png}
+      \includegraphics[scale=0.7]{figures/timelines-hists.png}
       \caption[]{Reference and Streaming timelines and corresponding histogram aggregations}
@@
 
-To compare each feature's reference and sliding window histogram aggregation we need a measure of divergence.
-We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein.
-A more thorough analysis on some statistical tests and why we chose to use the JSD is performed in Section \ref{sec:whyjsds}.
-Given a divergence value for the two aggregations (reference and sliding) we need a measure of normal statistical fluctuations.
+To compare each feature's reference and target sliding window histogram aggregation we need a measure of divergence.
+We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein \cite{EncyclopaediaMathematics}.
+The JSD takes as input two histograms and outputs a value between 0 and 1.
+The closer this value is to 0 the more similar the distributions are.
+On the other hand, a JSD value of 1 represents two totally different distributions.
+We chose to use the JSD metric due to its simplicity and leave as future work other possible tests.
+
+Given a divergence or distance value for the two aggregations (reference and target) we need a measure of normal statistical fluctuations.
 Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
@@
 The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}).
-The second phase (detailed in Section \ref{sec:stream-phase}) focuses on efficiently updating the sliding window histogram for each feature and periodically performing a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be.
+The second phase (detailed in Section \ref{sec:stream-phase}) focuses on efficiently updating the target sliding window histogram for each feature and periodically performing a divergence test between the static reference aggregation and the target sliding one, alerting for changes if need be.
 The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.
@@
 \subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
-As stated, we compute histogram aggregations for the reference and sliding windows.
+As stated, we compute histogram aggregations for the reference and target sliding windows.
 The computation of the reference histogram is done once in batch.
@@
 In addition, a batch computation also allows us to compute an exact reference histogram.
-However, the sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.
+However, the target sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.
 
@@
 Definition \ref{def:tuple-ema} shows how to compute an EMA using its recursive formula.
-An EMA state or value depends only on the previous one multiplied by an exponential decay $\lambda$.
-Recall that this decay is given by Equation \ref{eq:ema-decay}:
-\[ \lambda = 2^{- \dfrac{1}{\tau}} \]
-where $\tau$ is commonly denoted as the half-life.
+An EMA state or value depends only on the previous one multiplied by an exponential decay.
+Recall that a tuple-based decay is given by:
+\[ 2^{- \frac{1}{n_{1/2}}} \]
+where $n_{1/2}$ is commonly denoted as a tuple-based half-life.
 
-After processing \textit{x} tuples the oldest tuple is \emph{smoothed} by a factor of $\lambda^x = 2^{- \dfrac{x}{\tau}}$.
-If we process $x = 4\tau$ tuples the oldest tuple would have its contribution reduced by
-\[ 2^{- \dfrac{4\tau}{\tau}} = 2^{-4} \]
+After processing \textit{x} tuples the oldest tuple is \emph{smoothed} by a factor of $2^{- \frac{x}{n_{1/2}}}$.
+If we process $x = 4n_{1/2}$ tuples the oldest tuple would have its contribution reduced by
+\[ 2^{- \dfrac{4n_{1/2}}{n_{1/2}}} = 2^{-4} \]
 which is approximately 0.06. In other words, the \textit{x-th} event would have 6\% contribution towards the EMA state or value meaning older events would have even less weight.
 
-We choose to use windows with four times the half-life size (or $4\tau$), meaning we discard events whose contribution is less than 6\%.
+We choose to emulate windows with four times the half-life size (or $4n_{1/2}$), meaning we discard events whose contribution is less than 6\%.
 In fact, we set the half-life to be one quarter of the size of the window we want to emulate.
-For instance, if we want to monitor a period extending up to one month (at an error level of about 6\%) we setr the half-life to be the number of tuples that correspond to one week worth of data.
+We want to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data.
 This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.
@@
 
-    \item Compute the distribution of distances between the reference histogram and the sliding window histogram by computing the distance (or divergence value) between the reference histogram and various samples of sliding window histograms
+    \item Compute the distribution of distances between the reference histogram and random windows of data and corresponding histograms
 \end{enumerate}
@@
 Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase.
-To that end, we make $S$ samples of transactions (in Section \ref{sec:nsamples} we discuss the optimal number $S$ of samples to make), each with the same tuple-based size of the sliding window we will use in streaming.
+To that end, we make $S$ samples of transactions (in Section \ref{sec:nsamples} we discuss the optimal number $S$ of samples to make), each with the same tuple-based size of the target sliding window we will use in streaming.
 Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
@@
     \begin{center}
-      \includegraphics[scale=0.4]{figures/compute-sample-distances.png}
+      \includegraphics[scale=0.7]{figures/compute-sample-distances.png}
       \caption[Compute as many distance values as samples]{Compute as many distance values as samples, one for each sample-reference histogram pair}
@@
 Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if the probability is below a certain threshold.
-Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the sliding window histogram.
+Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the target sliding window histogram.
 This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.
@@
 The period of time where you discard the alerts produced by the system until it processes enough data to produce accurate reports is commonly denoted as the burn-in period.
-To avoid this burn-in period we propose that the sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
+To avoid this burn-in period we propose that the target sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
 
@@
     \item a list of distance values that represent its distribution
-    \item the last sample's histogram, to be used as burn-in period for the sliding window histogram
+    \item the last sample's histogram, to be used as burn-in period for the target sliding window histogram
 \end{itemize}
@@
 In order to obtain alerts, we periodically perform a multi-feature change detection test.
-The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the sliding window histogram of the same feature.
+The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the target sliding window histogram of the same feature.
 This distance value is then tested against the known distribution of distance values for feature \textit{x}.
 In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile, \textit{e.g.}, the 99th-percentile.
-When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and sliding window histogram aggregations.
+When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and target sliding window histogram aggregations.
 The probability value or \textit{p-value} of a distance value is:
@@
 When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives.
-In other words, when repeating a test multiple times the multiple test problem arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
-In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction.
-The Holm-Bonferroni method is one of many approaches for controlling the so called Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
+In other words, when repeating a test multiple times the multiple test or simultaneous statistical inference problem \cite{MultiTestProblem-Pubmed}
+\cite{MultiTestProblem-Dickhaus2014}
+\cite{MultiTestProblem-Miller1966}  \cite{multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
+In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction due to its simplicity and ease of computation and leave as future work the testing of other correction methods.
+In statistics, Family-Wise Error Rate (FWER) \cite{MultivariateMT} \cite{MultitestTamhane2018AdvancesIP} is the probability of making one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, consider one hypothesis true when it is not.
+The Holm-Bonferroni method \cite{HolmBonferroni} is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
 
@@
 
-\subsubsection*{Optimal number of samples} \label{sec:nsamples}
+\subsubsection*{Minimum number of samples} \label{sec:nsamples}
 In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase.
@@
 When checking for alerts we are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests.
-We apply a multiple test correction on the tests performed and want to set the Family Wise Error Rate \textit{(FWER)} to $\alpha$.
+We apply a multiple test correction on the tests performed and want to set the FWER to $\alpha$.
 In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.
@@
 
-\textcolor{red}{SEE MODEL MONITORING PAPER} We can then show that the minimum number of samples $S$ needed to satisfy these probability requirements is:
+In \cite{SAMM}, the authors prove that the minimum number of samples $S$ needed to satisfy these probability requirements is:
 \begin{equation}
@@
 \end{equation}
-
-\subsubsection*{Distribution Divergence Statistical Tests} \label{sec:whyjsds}
-
-We used the Jensen-Shannon Divergence (JSD) but....
-
-\textcolor{red}{Talk about wassertein, kolomogrov, why JSD...}

### 2020-06-24 chapters/my-work.tex
Ran grammarly on method

@@
 
-For this Thesis we need constant time complexity to process events one by one in real-time.
+For this Thesis, we need constant time complexity to process events one by one in real-time.
 Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit.
@@
 In the context of this Thesis, each of the event's fields are named features.
-Our goal is to detect data pattern shifts for each feature in an streaming fashion.
+Our goal is to detect data pattern shifts for each feature in a streaming fashion.
 Such shifts will be measured between a static in time or reference period versus the observed reality during the streaming phase.
@@
 This approach would require storing all window contents hence making the memory consumption grow linearly with target sliding window and/or reference period size.
-Instead, we want to use a \textit{(a)} fixed cost in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
+Instead, we want to use a \textit{(a)} fixed cost in-memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
 Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.
@@
 Recall Section \ref{sec:back-swag-algs} where we discussed generic sliding window aggregation algorithms and remember the Subtract-On-Evict (SOE) algorithm (Algorithm \ref{pseudo:soe}).
-Computing an histogram aggregation over a sliding window of size $W$ using SOE implies storing all $W$ elements in memory to know the element to remove after a new one is inserted.
-This violates our constant in memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}.
+Computing a histogram aggregation over a sliding window of size $W$ using SOE implies storing all $W$ elements in memory to know the element to remove after a new one is inserted.
+This violates our constant in-memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}.
 EMA-like histograms computed over sliding windows do not require the storage of window contents.
@@
 The closer this value is to 0 the more similar the distributions are.
-On the other hand, a JSD value of 1 represents two totally different distributions.
+On the other hand, a JSD value of 1 represents two different distributions.
 We chose to use the JSD metric due to its simplicity and leave as future work other possible tests.
@@
 Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
-This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).
+This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand-picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).
 
@@
 We choose to emulate windows with four times the half-life size (or $4n_{1/2}$), meaning we discard events whose contribution is less than 6\%.
-In fact, we set the half-life to be one quarter of the size of the window we want to emulate.
+We set the half-life to be one-quarter of the size of the window we want to emulate.
 We want to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data.
@@
 
-For each sample, and for each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}).
-We also add two extra bins that will cover the region to the left and to the right of our histogram, respectively.
+For each sample, and each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}).
+We also add two extra bins that will cover the region to the left and the right of our histogram, respectively.
 In other words, we add a bin that covers the region between $-\infty$ and our first bin and another that covers values from the last bin to $+\infty$.
@@
 An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}.
-Each sample is processed event by event and a discount factor is applied per event, thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample.
+Each sample is processed event by event and a discount factor is applied per event to each bin (EMA expiration factor), thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample.
+This procedure is illustrated in Figure \ref{fig:EMA-hist-build}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/ema-hist.png}
+      \caption[]{Reference equal-height histogram}
+      \label{fig:EMA-hist-build}
+    \end{center}
+\end{figure}
 We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period.
@@
 For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}.
-Instead we define our threshold as a percentile of the distribution and not as a hard-coded distance constant.
+Instead, we define our threshold as a percentile of the distribution and not as a hard-coded distance constant.
 For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
@@
 For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase.
-Then, for each incoming event and for each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
+Then, for each incoming event and each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
 This is all we do, regarding event processing.

### 2020-06-24 chapters/validation.tex
Ran grammarly on method

@@
 
-We generated multiple synthetic datasets before moving on to real data.
-Furthermore, we first performed single feature analysis before moving to a multi-feature scenario.
-The problem of multiple testing (Section \ref{sec:multi-test}) only arises in the latter case so we only apply the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for that one.
+In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
+As an initial first step we started by validating our system using a single-feature dataset before moving to multi-feature scenarios in order to facilitate system validation considering the complexities of multi-feature datasets and the problem of multiple test correction.
 

### 2020-06-25 chapters/my-work.tex
Some experiments

@@
 Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase.
-To that end, we make $S$ samples of transactions (in Section \ref{sec:nsamples} we discuss the optimal number $S$ of samples to make), each with the same tuple-based size of the target sliding window we will use in streaming.
+To that end, we make $S$ samples of transactions (in Section \ref{sec:nsamples} we discuss the minimum number $S$ of samples to make), each with the same tuple-based size of the target sliding window we will use in streaming.
 Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
@@
     \begin{center}
+    \hspace{2cm}
       \includegraphics[scale=0.8]{figures/ema-hist.png}

### 2020-06-25 chapters/validation.tex
Some experiments

@@
 
-\section{Datasets}
+In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
+As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios in order to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
 
-In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
-As an initial first step we started by validating our system using a single-feature dataset before moving to multi-feature scenarios in order to facilitate system validation considering the complexities of multi-feature datasets and the problem of multiple test correction.
+\section{Experiments with synthetic data}
 
-\begin{table}[!htb]
-    \begin{center}
-        \begin{tabular}{|c|c|c|}
-        \hline
-        \textbf{Dataset} & \textbf{Features} & \textbf{Events} \\ \hline
-        R1               & 1                 & 5.000.000       \\ \hline
-        T1               & 1                 & 5.000.000       \\ \hline
+\subsection{Varying sample sizes}
+We began with single-feature scenarios and by experimenting with different tuple-based half-lifes $n_{1/2}$ which in turn controls the size of the samples we make.
+In Section \ref{sec:ema-hist}, we mentioned we discarded events whose contribution was less than 6\%.
+We set a half-life $n_{1/2}$ value and then multiply it by a constant factor of 4 to get the size of the samples to make (a larger sample would contain events whose contribution was equal or less than 6\%).
 
-        \end{tabular}
-    \end{center}
-    \caption{Summary statistics for each dataset}
-    \label{tbl:summary-statistics}
-\end{table}
-
-\subsection{Synthetic datasets}
-Dataset R1 had a single feature that followed a normal distribution with mean $\mu=10$ and $\sigma=2$ for the entirety of the dataset.
+The reference dataset R1 used in this first experiment contained a single feature that followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$ for the 5 million rows of the dataset.
 Figure \ref{fig:timeseries-r1} shows the time-series of the single feature in dataset R1.
-
 \begin{figure}[!htb]
@@
 \end{figure}
-
-Dataset T1 also had 5 million events with a single feature.
-For the first half, \textit{i.e.}, for the first 2.5 million events, the feature followed a normal distribution with mean $\mu=10$ and $\sigma=2$.
-For the other half, we changed the generating distribution to be a continuous uniform one with $a=100$ and $b=200$.
-Figure \ref{fig:timeseries-t1} shows the time-series of the single feature in dataset T1 where we see the abrupt change in feature values.
-
+The dataset used to process as a data stream or the target dataset also had 5 million events with a single feature.
+For the first half, \textit{i.e.}, for the first 2.5 million events, the feature followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$, like the reference dataset.
+For the other half, we changed the generating distribution to be a continuous uniform one with lower bound $a=100$ and upper bound $b=200$.
+Figure \ref{fig:timeseries-t1} shows the time-series of the single feature in the target dataset T1 where we see the abrupt change in feature values.
 \begin{figure}[!htb]
@@
 
-\subsection{Financial fraud datasets}
+We conducted three different experiments varying the tuple-based half-life $n_{1/2}$, using the reference dataset R1 and the target dataset T1.
+These experiments, corresponding varying half-lifes and execution times are shown in Table \ref{tbl:tests-half-life}.
+Notice that $Sample Size$ corresponds to the size of each sample made in the batch phase and will be four times the half-life.
+We used 100 bins (plus the two special infinity bins) for the reference histogram and made 1000 samples.
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|c|c|}
+        \hline
+        \multicolumn{1}{|l|}{\textbf{Experiment}} & \multicolumn{1}{l|}{\textbf{Half-life}} & \multicolumn{1}{l|}{\textbf{Sample Size}} & \multicolumn{1}{l|}{\textbf{Executors}} & \multicolumn{1}{l|}{\textbf{Batch execution time (seconds)}} \\ \hline
+        01                                        & 625                                     & 2500                                      & 300                                     & 152                                                          \\ \hline
+        02                                        & 62,500                                  & 250,000                                   & 300                                     & 2,822                                                        \\ \hline
+        03                                        & 250,000                                 & 1,000,000                                 & 300                                     & 10,082                                                       \\ \hline
+        \end{tabular}
+    \end{center}
+    \caption{Experiments with varying half-life}
+    \label{tbl:tests-half-life}
+\end{table}
+7 6 8
+Table \ref{tbl:tests-half-life} shows the execution time for the batch phase in seconds.
+All experiments used 300 Spark executor instances.
+The batch phase comprises more than the sampling process but the remaining tasks take the same time across experiment.
+From the Table we conclude that the larger the half-life, the larger the sample size.
+Larger samples imply longer processing times when computing the EMA-like histogram for each sample.
+With no surprise, we see that the larger the sample the longer it takes to finish the batch phase.
 
-\section{Experiments}
-\textcolor{red}{one subsection per experiment with i) purpose, ii) execution, iii) results obtained and what thye mean}
+From a functional point of view, were we able to detect the abrupt change introduced in the target dataset T1?
+Recall that from the 2.5 millionth event forward we change the distribution from a gaussian to an uniform one (Figure \ref{fig:timeseries-t1}).
+Figure \ref{fig:JSD-signal-01} shows the distance (JSD) values computed during streaming at each event of the 5 million events from dataset T1, using the outputs from the batch phase of experiment 01.
+We clearly see a change in JSD values, and we move from values close to 0 to values close to 1 (or even 1).
+Figure \ref{fig:JSD-signal-zoom-01} shows the same signal but for the first half of the streaming period (essentially a zoom-in).
+We see our JSD values before the change were very low (magnitudes of $10^{-2}$).
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-625.png}
+      \caption[]{Distance (JSD) signal and threshold for experiment 01}
+      \label{fig:JSD-signal-01}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-625.png}
+      \caption[]{Zoomed in distance (JSD) signal and threshold for experiment 01}
+      \label{fig:JSD-signal-zoom-01}
+    \end{center}
+\end{figure}
+However, the change in JSD value is very abrupt as well.
+Almost instantly we go from close to 0 values to 1.
+We also had some false-positives, \textit{i.e.}, some JSD values above the threshold.
+This indicates to us that the window used is too small and sensible to small changes.
 
-\subsection{Single-feature analysis with varying sample sizes}
-In this first experiment, we used dataset R1 as the reference period and dataset T1 as the target or streaming period.
-We used 100 bins for the equal-height reference histogram and then added the two special bins mentioned in Section \ref{sec:sampling-batch}.
-We changed the half-life of each EMA-like sample histogram (recall Section \ref{sec:ema-hist}), effectively changing its size.
-We tested samples with \textit{(a)} 2.500, \textit{(b)} 250.000 and \textit{(c)} 1.000.000 events.
+In experiment 02, we use larger sampling windows.
+Remember our target sliding window in streaming is of the same size as the sampling windows.
+In the second experiment, we increase the size of the sampling and target windows by 100 times.
 
-\textcolor{red}{We use tukey threshold, how do we introduce this, ahhhhhhhhhhhh}
+Figure \ref{fig:JSD-signal-02} shows the JSD signal during streaming for dataset T1, using the outputs from the batch phase of experiment 02.
+Again, we see a change in signal from the middle of the streaming period forward.
+Figure \ref{fig:JSD-signal-zoom-02} shows the same signal for the first half of the streaming period.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-62500.png}
+      \caption[]{Distance (JSD) signal and threshold for experiment 02}
+      \label{fig:JSD-signal-02}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-62500.png}
+      \caption[]{Zoomed in distance (JSD) signal and threshold for experiment 02}
+      \label{fig:JSD-signal-zoom-02}
+    \end{center}
+\end{figure}
+Notice that this time the growth of our signal is slower.
+Notice also that we don not have false positives in the first half of the signal.
+This is because we use a larger window, which means each event has a smaller contribution to the aggregation state.
+Hence, outliers will have less impact on the aggregation and corresponding signal value.
 
-\section{Conclusions}
+In experiment 03, we increase the sampling and target windows size once again, this time four times larger than in experiment 02.
+We would expect an even smoother growth of our signal and, again, no false positives.
+And indeed, judging by the corresponding signal computation and zoom-in, Figures \ref{fig:JSD-signal-03} and \ref{fig:JSD-signal-zoom-03}, respectively, the signal looks smoother and we see no false positives.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-250000.png}
+      \caption[]{Distance (JSD) signal and threshold for experiment 03}
+      \label{fig:JSD-signal-03}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-250000.png}
+      \caption[]{Zoomed in distance (JSD) signal and threshold for experiment 03}
+      \label{fig:JSD-signal-zoom-03}
+    \end{center}
+\end{figure}
+
+\subsection{Multiple distribution changes, each contained within the previous one}
+In this experiment, we wanted to see how our signal behaved when there were multiple change points, with each new distribution's domain as a subset of the previous one.
+Figure \ref{fig:timeseries-r2} shows the time-series for the reference dataset R2 used in this experiment.
+It contained 5 million events with a single feature following a gaussian distribution with mean $\mu=100$ and standard deviation $\sigma=50$.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/timeseries-r2.png}
+      \caption[]{Time-series for dataset R2}
+      \label{fig:timeseries-r2}
+    \end{center}
+\end{figure}
+The time-series for target dataset T2 is shown in Figure \ref{fig:timeseries-t2}.
+Notice that we change distributions at 2, 3 and 4 millionth events.
+Notice also that each subsequent distribution produces values whose domain is a subset of the previous distribution.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/timeseries-t2.png}
+      \caption[]{Time-series for dataset T2}
+      \label{fig:timeseries-t2}
+    \end{center}
+\end{figure}
+
+The target dataset T2 begins with the same distribution as the reference dataset R2, a gaussian distribution with mean $\mu=100$ and standard deviation $\sigma=50$.
+At the 2 millionth event mark we use another gaussian distribution, this time with mean $\mu=100$ and standard deviation $\sigma=30$.
+At the 3 millionth event we change the standard deviation to be $\sigma=10$.
+From the 4 millionth event till the end, we use a standard deviation of $\sigma=5$.
+
+In this experiment, we used a half-life $n_{1/2}=62500$ and corresponding sample size of 250,000 events, like in the previous Section' experiment 02.
+Figure \ref{fig:JSD-signal-test02} shows the computed JSD signal for the target dataset and Figure \ref{fig:JSD-signal-zoom-test02} zooms in on the first 2 millionth events, before any distribution is changed.
+We observe that once again we have no false positives in the first 2 million events, which is ideal.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-test02.png}
+      \caption[]{Distance (JSD) signal and threshold}
+      \label{fig:JSD-signal-test02}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-test02.png}
+      \caption[]{Zoomed in distance (JSD) signal and threshold}
+      \label{fig:JSD-signal-zoom-test02}
+    \end{center}
+\end{figure}
+Furthermore, we observe that the computed JSD signal varies accordingly with the changes in distribution.
+Despite us changing the distributions to subsets of the previous ones to try and deceive our method, our signal accurately represents these changes.
+In Figure \ref{fig:JSD-signal-test02}, we see several growth points that correspond to the points where the generating gaussian distribution parameters changed.
+Hence, we are able to detect changes in distribution parameter changes and quantify how big the change is (the closer to 1 our signal is).

### 2020-06-25 chapters/validation.tex
Synthetic data single-feature analysis experiments

@@
 \section{Experiments with synthetic data}
+\subsection{Single Feature Analysis}
 
-\subsection{Varying sample sizes}
+\subsubsection{Varying sample sizes}
 We began with single-feature scenarios and by experimenting with different tuple-based half-lifes $n_{1/2}$ which in turn controls the size of the samples we make.
@@
 
-\subsection{Multiple distribution changes, each contained within the previous one}
-In this experiment, we wanted to see how our signal behaved when there were multiple change points, with each new distribution's domain as a subset of the previous one.
+\subsubsection{Multiple distribution changes, each contained within the previous one}
+In this experiment, we wanted to see how our signal behaved when there were multiple change points, where each distribution was contained within the previous one.
 Figure \ref{fig:timeseries-r2} shows the time-series for the reference dataset R2 used in this experiment.
@@
 Hence, we are able to detect changes in distribution parameter changes and quantify how big the change is (the closer to 1 our signal is).
+
+\subsubsection{Multiple distribution changes, each a superset of the previous one}
+
+In this experiment, we wanted to do the reverse of the previous one.
+In this one, there are multiple change points but each new distribution is a superset of the previous one.
+Figure \ref{fig:timeseries-r3} shows the reference period used, with 5 million events and one feature following a gaussian distribution of mean $\mu=100$ and standard deviation $\sigma=10$.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/timeseries-r3.png}
+      \caption[]{Time-series for dataset R3}
+      \label{fig:timeseries-r3}
+    \end{center}
+\end{figure}
+Figure \ref{fig:timeseries-t3} shows the target period used, also with 5 million events and one feature.
+Like in the previous experiment, we change distributions at the 2, 3 and 4 millionth event mark.
+On the other hand, unlike the previous experiment, we now make subsequent distributions cover the previous ones.
+We change the reference gaussian used with mean $\mu=100$ and standard deviation $\sigma=10$ at the 2, 3 and 4 millionth mark, to use standard deviations $\sigma=30$, $\sigma=40$ and $\sigma=50$, respectively.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/timeseries-t3.png}
+      \caption[]{Time-series for dataset T3}
+      \label{fig:timeseries-t3}
+    \end{center}
+\end{figure}
+
+Figure \ref{fig:JSD-signal-test03} shows the JSD signal for the target dataset in this experiment.
+Once again, we were able to detect the changes introduced at the 2, 3 and 4 millionth events.
+We are also able to identify bigger distribution changes, relative to the reference, by observing bigger JSD values.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-test03.png}
+      \caption[]{Distance (JSD) signal and threshold}
+      \label{fig:JSD-signal-test03}
+    \end{center}
+\end{figure}
+Figure \ref{fig:JSD-signal-zoom-test03} shows a zoomed in portion of the original signal, where we can verify that there are no false positives.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-test03.png}
+      \caption[]{Zoomed in distance (JSD) signal and threshold}
+      \label{fig:JSD-signal-zoom-test03}
+    \end{center}
+\end{figure}
+
+\subsubsection{Return to normality after change}
+Our goal with this experiment was to check if our signal returned to its normal values (below the alert threshold) if we changed our distribution to the original reference one, as intended.
+
+In this experiment, we used reference dataset R4 (Figure \ref{fig:timeseries-r4}) and target dataset T4 (Figure \ref{fig:timeseries-t4}). 
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/timeseries-r4.png}
+      \caption[]{Time-series for dataset R4}
+      \label{fig:timeseries-r4}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/timeseries-t4.png}
+      \caption[]{Time-series for dataset T4}
+      \label{fig:timeseries-t4}
+    \end{center}
+\end{figure}
+
+The reference dataset R4 had a single feature that followed a gaussian distribution of $\mu=100$ and $\sigma=5$ for all 5 million events.
+Up until the 2 millionth event, the target dataset T4 followed a gaussian distribution of $\mu=100$ and $\sigma=5$, similarly to the reference dataset R4.
+At the 2 millionth event, we change the gaussian parameters to $\mu=100$ and $\sigma=30$.
+At the 3 millionth mark, we once again change the gaussian parameters to $\mu=100$ and $\sigma=50$.
+Finally, on the 4 millionth event, we return to our reference distribution, a gaussian with $\mu=100$ and $\sigma=5$.
+
+Figure \ref{fig:JSD-signal-test04} shows the computed JSD signal for the target period.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-test04.png}
+      \caption[]{Distance (JSD) signal and threshold}
+      \label{fig:JSD-signal-test04}
+    \end{center}
+\end{figure}
+
+As expected, our JSD signal increases at the 2 millionth and 3 millionth event marks, because the distribution parameters change.
+At the 4 millionth event, we use the original reference distribution and notice that the JSD signal returns to values below the threshold, considering it a normal state, as we intended.
+Figure \ref{fig:JSD-signal-zoom-test04} is a zoom in on the first 2 millionth events, before any change occurs, and we see no false positives.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-test04.png}
+      \caption[]{Zoomed in distance (JSD) signal and threshold}
+      \label{fig:JSD-signal-zoom-test04}
+    \end{center}
+\end{figure}
+
+\subsubsection{An attempt to deceive our method}
+In this experiment, we attempt to deceive our method by using a different distribution from the reference period but ensuring its domain is roughly the same as the reference one.
+
+We used the reference dataset R5 (Figure \ref{fig:timeseries-r5}) which had 5 million events, with a single-feature that followed a gaussian with  $\mu=100$ and $\sigma=10$.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/timeseries-r5.png}
+      \caption[]{Time-series for dataset R5}
+      \label{fig:timeseries-r5}
+    \end{center}
+\end{figure}
+
+Our target dataset T5 is represented in Figure \ref{fig:timeseries-t5}.
+For the first 2 million events, the generating distribution used was similar to the one used in the reference dataset, a gaussian with $\mu=100$ and $\sigma=10$.
+However, at the 2 millionth event, we change the gaussian standard deviation to $\sigma=30$.
+Then, at the 4 millionth event, we change the distribution type.
+We use an uniform distribution with lower bound $a=80$ and upper bound $b=120$.
+Note in Figure \ref{fig:timeseries-t5} that this last portion resembles the first one, the reference.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/timeseries-t5.png}
+      \caption[]{Time-series for dataset T5}
+      \label{fig:timeseries-t5}
+    \end{center}
+\end{figure}
+
+Let's take a look at the signal computed for the target period.
+Figure \ref{fig:JSD-signal-test05} shows us the JSD signal for dataset T5 \ref{fig:timeseries-t5}.
+As expected, we see a rapid growth of JSD value after the 2 millionth event, when we change the gaussian parameters.
+We then change the distribution type, from a gaussian to a uniform one.
+Despite the latter having the a domain contained within the former, we see that our signal is accurate and does not drop below the alert threshold.
+This is still a different distribution that we want to alert.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-test05.png}
+      \caption[]{Distance (JSD) signal and threshold}
+      \label{fig:JSD-signal-test05}
+    \end{center}
+\end{figure}
+
+Figure \ref{fig:JSD-signal-zoom-test05} is a zoom in of the first 2 millionth events and once again we have no false positives.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-test05.png}
+      \caption[]{Zoomed in distance (JSD) signal and threshold}
+      \label{fig:JSD-signal-zoom-test05}
+    \end{center}
+\end{figure}
+
+\subsection{Multi Feature Analysis}
+
+...

### 2020-06-25 chapters/validation.tex
Multifeature synthetic validation scenario description

@@
 \subsection{Multi Feature Analysis}
+Before moving to real data, we performed one last experiment, using synthetic data but this time performing a multi-feature analysis.
+Table \ref{tbl:multi-feat-ref-dataset-distros} shows the distributions and their parameters for each of the four features (x1, x2, x3 and x4) present in the reference multi-feature dataset R6.
+Figures \ref{fig:timeseries-r6-x1}, \ref{fig:timeseries-r6-x2}, \ref{fig:timeseries-r6-x3} and \ref{fig:timeseries-r6-x4} show the reference time-series for features x1, x2, x3 and x4 on the reference dataset R6, respectively.
 
-...
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|ll}
+        \cline{1-3}
+        \textbf{Feature} & \textbf{Reference Distribution} & \textbf{Parameters} &  &  \\ \cline{1-3}
+        x1               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
+        x2               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
+        x3               & Uniform                         & $a=200$, $b=250$       &  &  \\ \cline{1-3}
+        x4               & Uniform                         & $a=50$, $b=125$        &  &  \\ \cline{1-3}
+        \end{tabular}
+    \end{center}
+    \caption{Multi feature reference dataset and feature distributions}
+    \label{tbl:multi-feat-ref-dataset-distros}
+\end{table}
+
+\begin{figure}[!htb] 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x1.png} 
+    \caption{Feature x1 reference time-series} 
+    \label{fig:timeseries-r6-x1} 
+    \vspace{4ex}
+  \end{minipage}%%
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x2.png} 
+    \caption{Feature x2 reference time-series} 
+    \label{fig:timeseries-r6-x2} 
+    \vspace{4ex}
+  \end{minipage} 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x3.png} 
+    \caption{Feature x3 reference time-series} 
+    \label{fig:timeseries-r6-x3} 
+    \vspace{4ex}
+  \end{minipage}%% 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x4.png} 
+    \caption{Feature x4 reference time-series} 
+    \label{fig:timeseries-r6-x4} 
+    \vspace{4ex}
+  \end{minipage} 
+\end{figure}
+
+We used a target dataset T6 with 5 million events and the same four features (x1, x2, x3 and x4).
+Feature x1 followed the reference distribution for the 5 million events of the target dataset.
+We did not change this feature underlying distribution so as to serve as control, \textit{i.e.}, we expect no alerts for this one.
+Figure \ref{fig:timeseries-t6-x1} shows the time-series for the target dataset T6 and feature x1.
+
+Feature x2's underlying distribution changed four times and returned to normal, so we expect to see alarms with increasing divergence values but eventually no more alarms.
+Figure \ref{fig:timeseries-t6-x2} shows the time-series for the target dataset T6 and feature x2.
+Feature x2 followed a gaussian with $\mu=100$ and $\sigma=5$ like in the reference period up to the 1 millionth event mark, then a gaussian with $\mu=100$ and $\sigma=10$ up to the 2.5 millionth event, then a gaussian $\mu=100$ and $\sigma=15$ until the 3 millionth event, then a gaussian $\mu=100$ and $\sigma=20$ until the 4 millionth event and finally returned to the original gaussian with $\mu=100$ and $\sigma=5$ until the last event.
+
+Feature x3's underlying distribution changed three times and never returned to normal, so we expect to see alarms with increasing divergence values up until the end of the stream analysis.
+Figure \ref{fig:timeseries-t6-x3} shows the time-series for the target dataset T6 and feature x3.
+Feature x3 followed a uniform distribution with $a=200$ and $b=250$ like in the reference period up to the 1.5 millionth event mark, then a uniform distribution with $a=200$ and $b=220$ up to the 2 millionth event, then a uniform distribution with $a=180$ and $b=200$ until the 3.5 millionth event and finally a gaussian with $\mu=50$ and $\sigma=20$ until the last event.
+
+Feature x4's underlying distribution changed two times and returned to normal, so we expect to see no alarms towards the end of the stream analysis.
+Figure \ref{fig:timeseries-t6-x4} shows the time-series for the target dataset T6 and feature x4.
+Feature x4 followed a uniform distribution like the reference one, with $a=50$ and $b=125$ up until the 1.5 millionth event.
+After that, it followed a uniform distribution with $a=125$ and $b=200$ until the 3 millionth event where it resumed its original uniform distribution (with $a=50$ and $b=125$) until the end.
+
+\begin{figure}[!htb] 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x1.png} 
+    \caption{Feature x1 target time-series} 
+    \label{fig:timeseries-t6-x1} 
+    \vspace{4ex}
+  \end{minipage}%%
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x2.png} 
+    \caption{Feature x2 target time-series} 
+    \label{fig:timeseries-t6-x2} 
+    \vspace{4ex}
+  \end{minipage} 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x3.png} 
+    \caption{Feature x3 target time-series} 
+    \label{fig:timeseries-t6-x3} 
+    \vspace{4ex}
+  \end{minipage}%% 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x4.png} 
+    \caption{Feature x4 target time-series} 
+    \label{fig:timeseries-t6-x4} 
+    \vspace{4ex}
+  \end{minipage} 
+\end{figure}

### 2020-06-25 chapters/validation.tex
End of another hell day

@@
 \end{figure}
+
+In this experiment, we used a tuple-based half-life $n_{1/2}=62500$, \textit{i.e.} sampling and target windows of 250,000 tuples.
+We used 100 bins for the reference histogram plus the two special infinity bins.
+In this multi-feature scenario, we used the Holm-Bonferroni correction discussed in Section \ref{sec:holmbonferroni} and set the Family-Wise Error Rate (FWER) to 1\% (FWER = 0.01).
+Furthermore, instead of performing ad-hoc 1000 samples like we did previosuly, we use Equation \ref{eq:optimal-n-samples} to determine the minimum number of samples to make while ensuring $\gamma=1\%$ (recall Section \ref{sec:nsamples}).
+For our four feature analysis scenario, using as inputs  $\gamma=0.01$ and $FWER=0.01$ in Equation \ref{eq:optimal-n-samples} tells us we need to make 1840 samples minimum.
+
+LOOK INTO EVERY1000.TXT AND COMMENT ALERT RESULTS.
+SPECIFY THAT THIS PERIOD IS AD-HOC

### 2020-06-26 chapters/validation.tex
Finished \section{Experiments with synthetic data}

@@
 In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
-As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios in order to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
+As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
 
@@
 \subsubsection{Varying sample sizes}
-We began with single-feature scenarios and by experimenting with different tuple-based half-lifes $n_{1/2}$ which in turn controls the size of the samples we make.
+We began with single-feature scenarios and by experimenting with different tuple-based half-lives $n_{1/2}$ which in turn controls the size of the samples we make.
 In Section \ref{sec:ema-hist}, we mentioned we discarded events whose contribution was less than 6\%.
@@
 Notice that this time the growth of our signal is slower.
-Notice also that we don not have false positives in the first half of the signal.
+Notice also that we do not have false positives in the first half of the signal.
 This is because we use a larger window, which means each event has a smaller contribution to the aggregation state.
@@
 The target dataset T2 begins with the same distribution as the reference dataset R2, a gaussian distribution with mean $\mu=100$ and standard deviation $\sigma=50$.
-At the 2 millionth event mark we use another gaussian distribution, this time with mean $\mu=100$ and standard deviation $\sigma=30$.
-At the 3 millionth event we change the standard deviation to be $\sigma=10$.
-From the 4 millionth event till the end, we use a standard deviation of $\sigma=5$.
+At the 2 millionth event mark, we use another Gaussian distribution, this time with mean $\mu=100$ and standard deviation $\sigma=30$.
+At the 3 millionth event, we change the standard deviation to be $\sigma=10$.
+From the 4 millionth event to the end, we use a standard deviation of $\sigma=5$.
 
@@
 
+\newpage
+
 \subsubsection{Return to normality after change}
@@
 
-The reference dataset R4 had a single feature that followed a gaussian distribution of $\mu=100$ and $\sigma=5$ for all 5 million events.
-Up until the 2 millionth event, the target dataset T4 followed a gaussian distribution of $\mu=100$ and $\sigma=5$, similarly to the reference dataset R4.
-At the 2 millionth event, we change the gaussian parameters to $\mu=100$ and $\sigma=30$.
-At the 3 millionth mark, we once again change the gaussian parameters to $\mu=100$ and $\sigma=50$.
+The reference dataset R4 had a single feature that followed a Gaussian distribution of $\mu=100$ and $\sigma=5$ for all 5 million events.
+Up until the 2 millionth event, the target dataset T4 followed a Gaussian distribution of $\mu=100$ and $\sigma=5$, similarly to the reference dataset R4.
+At the 2 millionth event, we change the Gaussian parameters to $\mu=100$ and $\sigma=30$.
+At the 3 millionth mark, we once again change the Gaussian parameters to $\mu=100$ and $\sigma=50$.
 Finally, on the 4 millionth event, we return to our reference distribution, a gaussian with $\mu=100$ and $\sigma=5$.
@@
 \end{figure}
-
 As expected, our JSD signal increases at the 2 millionth and 3 millionth event marks, because the distribution parameters change.
@@
 Figure \ref{fig:JSD-signal-test05} shows us the JSD signal for dataset T5 \ref{fig:timeseries-t5}.
-As expected, we see a rapid growth of JSD value after the 2 millionth event, when we change the gaussian parameters.
+As expected, we see a rapid growth of JSD value after the 2 millionth event, when we change the Gaussian parameters.
 We then change the distribution type, from a gaussian to a uniform one.
-Despite the latter having the a domain contained within the former, we see that our signal is accurate and does not drop below the alert threshold.
+Despite the latter having a domain contained within the former, we see that our signal is accurate and does not drop below the alert threshold.
 This is still a different distribution that we want to alert.
@@
 Feature x1 followed the reference distribution for the 5 million events of the target dataset.
-We did not change this feature underlying distribution so as to serve as control, \textit{i.e.}, we expect no alerts for this one.
+We did not change this feature underlying distribution to serve as a control, \textit{i.e.}, we expect no alerts for this one.
 Figure \ref{fig:timeseries-t6-x1} shows the time-series for the target dataset T6 and feature x1.
@@
 In this multi-feature scenario, we used the Holm-Bonferroni correction discussed in Section \ref{sec:holmbonferroni} and set the Family-Wise Error Rate (FWER) to 1\% (FWER = 0.01).
-Furthermore, instead of performing ad-hoc 1000 samples like we did previosuly, we use Equation \ref{eq:optimal-n-samples} to determine the minimum number of samples to make while ensuring $\gamma=1\%$ (recall Section \ref{sec:nsamples}).
+Furthermore, instead of performing ad-hoc 1000 samples as we did previously, we use Equation \ref{eq:optimal-n-samples} to determine the minimum number of samples to make while ensuring $\gamma=1\%$ (recall Section \ref{sec:nsamples}).
 For our four feature analysis scenario, using as inputs  $\gamma=0.01$ and $FWER=0.01$ in Equation \ref{eq:optimal-n-samples} tells us we need to make 1840 samples minimum.
 
-LOOK INTO EVERY1000.TXT AND COMMENT ALERT RESULTS.
-SPECIFY THAT THIS PERIOD IS AD-HOC
+In our stream analysis, we periodically performed our divergence test using the Holm-Bonferroni correction \ref{sec:holmbonferroni}.
+The frequency of the test used was a user-defined constant of 1000 events.
+We generated a list of alerts where each one contained the alert timestamp and a list of divergent features with the corresponding p-value.
+We achieved a throughput of 24,630 transactions per second (TPS).
+This value can be made higher if we decrease the frequency of the test.
+
+The first alert we obtain is:
+\[1001000, [(0.0019999568566211323, x2)]\]
+The previous alert tells us that at event 1,001,000 the only diverging feature in the system is feature \textit{x2} and what the p-value for the feature value at that time is.
+This is consistent with our target period induced changes.
+We introduced a change in distribution at the 1 millionth event for feature x2.
+We caught this change 1000 events later because of our test frequency, which is adjustable to obtain results as recent as we need them to be.
+
+The second alert is:
+\[1501000, [(0.001746900672135232, x2), (0.001989149448289207, x4), (0.0019999899243986485, x3)]\]
+We observe that besides x2, features x3 and x4 are now considered divergent as well.
+This is consistent with the changes we introduced for both x3 and x4 at the 1.5 millionth event.
+Again, we get the alert 1000 events after, exactly then the test is performed.
+
+Note that the p-value gives us a measure of divergence and allows us to order features from most divergent to least.
+The lower the p-value (probability value) the more divergent the feature is considered.
+For example, feature x3 has its distribution further changed at the 2 millionth event.
+At the 2 millionth event, we have the following state:
+\[2000000, [(5.953078867981798e-05, x4), (0.0012242703961621304, x3), (0.0017464046986205295, x2)]\]
+Note that x4 has the lowest p-value, being considered the most divergent feature.
+However, because we change feature x3 at the 2 millionth mark, the p-values observed for feature x3 gradually grow and, at the 2,535,000-th event, feature x3 is now considered the most divergent feature:
+\[2535000, [(2.3389481637825504e-05, x3), (2.342775340591352e-05, x4), (0.0016641704588643558, x2)]\]
+At the 3 millionth event, we change the distribution of feature x4 to its original reference one. x3's p-value gradually increases and after 804,000 events we stop reporting this feature as divergent:
+\[3804000, [(3.0928114358452774e-07, x3), (0.0012947913998524463, x2)]\]
+Similarly, we change feature x2's underlying distribution at the 4 millionth event to its original one.
+Its p-value gradually increases until we no longer report it as divergent.
+It took 674,000 events for the feature p-value to raise above the threshold:
+\[4674000, [(3.0922177229886216e-07, x3)]\]
+Feature x3 never returns to its original distribution, so up until the last event (5th million one) we report it as divergent:
+\[5000000, [(3.0922176674774704e-07, x3)]\]
+
+\section{Experiments with real data}

### 2020-06-27 chapters/my-work.tex
End of another hell day
Some Marco feedback regarding Validation

@@
 \cite{MultiTestProblem-Miller1966}  \cite{multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
-In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction due to its simplicity and ease of computation and leave as future work the testing of other correction methods.
-In statistics, Family-Wise Error Rate (FWER) \cite{MultivariateMT} \cite{MultitestTamhane2018AdvancesIP} is the probability of making one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, consider one hypothesis true when it is not.
+In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the amount of false negatives while having as target to control the amount of false positives.
+Furthermore, it does not require to assume any indepedence between the tested hypothesis (as other methods do) which would almost surely be a wrong assumption for realistic datasets.
+Therefore we leave as future work the testing of other correction methods.
+In statistics, Family-Wise Error Rate (FWER) \cite{MultivariateMT} \cite{MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not.
 The Holm-Bonferroni method \cite{HolmBonferroni} is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
@@
     \label{eq:optimal-n-samples}
-    \textit{S} = \frac{\log\gamma}{log(1 - \frac{\alpha}{n_{features}})}
+    \textit{S} = \ceil[\bigg]{\frac{\log\gamma}{log(1 - \frac{\alpha}{n_{features}})}}
 \end{equation}

### 2020-06-27 chapters/validation.tex
Some Marco feedback regarding Validation

@@
 \subsubsection{Varying sample sizes}
-We began with single-feature scenarios and by experimenting with different tuple-based half-lives $n_{1/2}$ which in turn controls the size of the samples we make.
-In Section \ref{sec:ema-hist}, we mentioned we discarded events whose contribution was less than 6\%.
+We began with single-feature scenarios and by experimenting with different tuple-based half-lives $n_{1/2}$.
+The half-life $n_{1/2}$ parameter in turn controls the size of the samples we make in the batch phase and the target sliding window size of the streaming phase.
+
+In Section \ref{sec:ema-hist}, we mentioned that we choose to discard events whose contribution is less than 6\%.
 We set a half-life $n_{1/2}$ value and then multiply it by a constant factor of 4 to get the size of the samples to make (a larger sample would contain events whose contribution was equal or less than 6\%).
 
+All the synthetic datasets used throughout this Section are a time-series where at each time-step we sample uniformly from a fixed distribution, \textit{i.e.} there are no time correlations in the used datasets.
 The reference dataset R1 used in this first experiment contained a single feature that followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$ for the 5 million rows of the dataset.
@@
 \end{figure}
-The dataset used to process as a data stream or the target dataset also had 5 million events with a single feature.
+The dataset processed in the streaming phase (\textit{i.e.} the target dataset) was generated with a size of 5 million events and a single feature.
 For the first half, \textit{i.e.}, for the first 2.5 million events, the feature followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$, like the reference dataset.
@@
 We conducted three different experiments varying the tuple-based half-life $n_{1/2}$, using the reference dataset R1 and the target dataset T1.
-These experiments, corresponding varying half-lifes and execution times are shown in Table \ref{tbl:tests-half-life}.
-Notice that $Sample Size$ corresponds to the size of each sample made in the batch phase and will be four times the half-life.
+The experiment identifiers, the corresponding half-lives and the execution times are shown in Table \ref{tbl:tests-half-life}.
+Notice that $Sample Size$ corresponds to the size of each sample made in the batch phase and so it is four times the half-life.
 We used 100 bins (plus the two special infinity bins) for the reference histogram and made 1000 samples.
@@
 \end{table}
-7 6 8
 Table \ref{tbl:tests-half-life} shows the execution time for the batch phase in seconds.
 All experiments used 300 Spark executor instances.
-The batch phase comprises more than the sampling process but the remaining tasks take the same time across experiment.
 From the Table we conclude that the larger the half-life, the larger the sample size.
@@
 We also had some false-positives, \textit{i.e.}, some JSD values above the threshold.
-This indicates to us that the window used is too small and sensible to small changes.
+This indicates to us that the window used is too small and sensitive to small changes, or that we can choose a more strict threshold.
 
@@
 
-\subsubsection{Multiple distribution changes, each contained within the previous one}
-In this experiment, we wanted to see how our signal behaved when there were multiple change points, where each distribution was contained within the previous one.
+\subsubsection{Multiple distribution changes}
+In this experiment, we wanted to see how our signal behaved when there were multiple change points, where each gaussian distribution was a scaled down version of the previous one.
 Figure \ref{fig:timeseries-r2} shows the time-series for the reference dataset R2 used in this experiment.
@@
 \subsubsection{Multiple distribution changes, each a superset of the previous one}
-
 In this experiment, we wanted to do the reverse of the previous one.
-In this one, there are multiple change points but each new distribution is a superset of the previous one.
+In this one, there are multiple change points but each new distribution spans a domain that is a superset of the previous one.
 Figure \ref{fig:timeseries-r3} shows the reference period used, with 5 million events and one feature following a gaussian distribution of mean $\mu=100$ and standard deviation $\sigma=10$.
@@
 
-\newpage
+\clearpage
 
@@
 Figure \ref{fig:JSD-signal-zoom-test04} is a zoom in on the first 2 millionth events, before any change occurs, and we see no false positives.
+
 \begin{figure}[!htb]
@@
 \[1001000, [(0.0019999568566211323, x2)]\]
-The previous alert tells us that at event 1,001,000 the only diverging feature in the system is feature \textit{x2} and what the p-value for the feature value at that time is.
+This tells us that at event 1,001,000 the only diverging feature in the system is feature \textit{x2} and what the p-value for the feature value at that time is about 0.2\%.
 This is consistent with our target period induced changes.

### 2020-06-27 chapters/validation.tex
Side by side images

@@
 Figure \ref{fig:timeseries-r1} shows the time-series of the single feature in dataset R1.
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/01-reference.png}
-      \caption[]{Time-series for dataset R1}
-      \label{fig:timeseries-r1}
-    \end{center}
-\end{figure}
 The dataset processed in the streaming phase (\textit{i.e.} the target dataset) was generated with a size of 5 million events and a single feature.
@@
 Figure \ref{fig:timeseries-t1} shows the time-series of the single feature in the target dataset T1 where we see the abrupt change in feature values.
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/01-target.png}
-      \caption[]{Time-series for dataset T1}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/01-reference.png}
+  \captionof{figure}{Time-series for dataset R1}
+  \label{fig:timeseries-r1}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/01-target.png}
+  \captionof{figure}{Time-series for dataset T1}
   \label{fig:timeseries-t1}
+\end{minipage}
 \end{center}
-\end{figure}
 
@@
 We see our JSD values before the change were very low (magnitudes of $10^{-2}$).
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-625.png}
-      \caption[]{Distance (JSD) signal and threshold for experiment 01}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-625.png}
+  \captionof{figure}{JSD signal and threshold}
   \label{fig:JSD-signal-01}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-625.png}
+  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \label{fig:JSD-signal-zoom-01}
+\end{minipage}
 \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-625.png}
-      \caption[]{Zoomed in distance (JSD) signal and threshold for experiment 01}
-      \label{fig:JSD-signal-zoom-01}
-    \end{center}
-\end{figure}
 However, the change in JSD value is very abrupt as well.
@@
 Figure \ref{fig:JSD-signal-zoom-02} shows the same signal for the first half of the streaming period.
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-62500.png}
-      \caption[]{Distance (JSD) signal and threshold for experiment 02}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-62500.png}
+  \captionof{figure}{JSD signal and threshold}
   \label{fig:JSD-signal-02}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-62500.png}
+  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \label{fig:JSD-signal-zoom-02}
+\end{minipage}
 \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-62500.png}
-      \caption[]{Zoomed in distance (JSD) signal and threshold for experiment 02}
-      \label{fig:JSD-signal-zoom-02}
-    \end{center}
-\end{figure}
 Notice that this time the growth of our signal is slower.
@@
 And indeed, judging by the corresponding signal computation and zoom-in, Figures \ref{fig:JSD-signal-03} and \ref{fig:JSD-signal-zoom-03}, respectively, the signal looks smoother and we see no false positives.
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-250000.png}
-      \caption[]{Distance (JSD) signal and threshold for experiment 03}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-250000.png}
+  \captionof{figure}{JSD signal and threshold}
   \label{fig:JSD-signal-03}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-250000.png}
+  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \label{fig:JSD-signal-zoom-03}
+\end{minipage}
 \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-250000.png}
-      \caption[]{Zoomed in distance (JSD) signal and threshold for experiment 03}
-      \label{fig:JSD-signal-zoom-03}
-    \end{center}
-\end{figure}
 
@@
 It contained 5 million events with a single feature following a gaussian distribution with mean $\mu=100$ and standard deviation $\sigma=50$.
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/timeseries-r2.png}
-      \caption[]{Time-series for dataset R2}
-      \label{fig:timeseries-r2}
-    \end{center}
-\end{figure}
 The time-series for target dataset T2 is shown in Figure \ref{fig:timeseries-t2}.
@@
 Notice also that each subsequent distribution produces values whose domain is a subset of the previous distribution.
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/timeseries-t2.png}
-      \caption[]{Time-series for dataset T2}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-r2.png}
+  \captionof{figure}{Time-series for dataset R2}
+  \label{fig:timeseries-r2}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-t2.png}
+  \captionof{figure}{Time-series for dataset T2}
   \label{fig:timeseries-t2}
+\end{minipage}
 \end{center}
-\end{figure}
 
@@
 We observe that once again we have no false positives in the first 2 million events, which is ideal.
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-test02.png}
-      \caption[]{Distance (JSD) signal and threshold}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test02.png}
+  \captionof{figure}{JSD signal and threshold}
   \label{fig:JSD-signal-test02}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test02.png}
+  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \label{fig:JSD-signal-zoom-test02}
+\end{minipage}
 \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-test02.png}
-      \caption[]{Zoomed in distance (JSD) signal and threshold}
-      \label{fig:JSD-signal-zoom-test02}
-    \end{center}
-\end{figure}
 Furthermore, we observe that the computed JSD signal varies accordingly with the changes in distribution.
@@
 Figure \ref{fig:timeseries-r3} shows the reference period used, with 5 million events and one feature following a gaussian distribution of mean $\mu=100$ and standard deviation $\sigma=10$.
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/timeseries-r3.png}
-      \caption[]{Time-series for dataset R3}
-      \label{fig:timeseries-r3}
-    \end{center}
-\end{figure}
 Figure \ref{fig:timeseries-t3} shows the target period used, also with 5 million events and one feature.
@@
 We change the reference gaussian used with mean $\mu=100$ and standard deviation $\sigma=10$ at the 2, 3 and 4 millionth mark, to use standard deviations $\sigma=30$, $\sigma=40$ and $\sigma=50$, respectively.
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/timeseries-t3.png}
-      \caption[]{Time-series for dataset T3}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-r3.png}
+  \captionof{figure}{Time-series for dataset R3}
+  \label{fig:timeseries-r3}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1.2\linewidth]{figures/timeseries-t3.png}
+  \captionof{figure}{Time-series for dataset T3}
   \label{fig:timeseries-t3}
+\end{minipage}
 \end{center}
-\end{figure}
 
@@
 We are also able to identify bigger distribution changes, relative to the reference, by observing bigger JSD values.
-\begin{figure}[!htb]
+Figure \ref{fig:JSD-signal-zoom-test03} shows a zoomed in portion of the original signal, where we can verify that there are no false positives.
 \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-test03.png}
-      \caption[]{Distance (JSD) signal and threshold}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test03.png}
+  \captionof{figure}{JSD signal and threshold}
   \label{fig:JSD-signal-test03}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test03.png}
+  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \label{fig:JSD-signal-zoom-test03}
+\end{minipage}
 \end{center}
-\end{figure}
-Figure \ref{fig:JSD-signal-zoom-test03} shows a zoomed in portion of the original signal, where we can verify that there are no false positives.
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-test03.png}
-      \caption[]{Zoomed in distance (JSD) signal and threshold}
-      \label{fig:JSD-signal-zoom-test03}
-    \end{center}
-\end{figure}
-
-\clearpage
 
@@
 In this experiment, we used reference dataset R4 (Figure \ref{fig:timeseries-r4}) and target dataset T4 (Figure \ref{fig:timeseries-t4}). 
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/timeseries-r4.png}
-      \caption[]{Time-series for dataset R4}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-r4.png}
+  \captionof{figure}{Time-series for dataset R4}
   \label{fig:timeseries-r4}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-t4.png}
+  \captionof{figure}{Time-series for dataset T4}
+  \label{fig:timeseries-t4}
+\end{minipage}
 \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/timeseries-t4.png}
-      \caption[]{Time-series for dataset T4}
-      \label{fig:timeseries-t4}
-    \end{center}
-\end{figure}
 
@@
 Figure \ref{fig:JSD-signal-test04} shows the computed JSD signal for the target period.
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-test04.png}
-      \caption[]{Distance (JSD) signal and threshold}
-      \label{fig:JSD-signal-test04}
-    \end{center}
-\end{figure}
 As expected, our JSD signal increases at the 2 millionth and 3 millionth event marks, because the distribution parameters change.
@@
 
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-test04.png}
-      \caption[]{Zoomed in distance (JSD) signal and threshold}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test04.png}
+  \captionof{figure}{JSD signal and threshold}
+  \label{fig:JSD-signal-test04}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test04.png}
+  \captionof{figure}{Zoomed in JSD signal and threshold}
   \label{fig:JSD-signal-zoom-test04}
+\end{minipage}
 \end{center}
-\end{figure}
 
@@
 We used the reference dataset R5 (Figure \ref{fig:timeseries-r5}) which had 5 million events, with a single-feature that followed a gaussian with  $\mu=100$ and $\sigma=10$.
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/timeseries-r5.png}
-      \caption[]{Time-series for dataset R5}
-      \label{fig:timeseries-r5}
-    \end{center}
-\end{figure}
-
 Our target dataset T5 is represented in Figure \ref{fig:timeseries-t5}.
@@
 Note in Figure \ref{fig:timeseries-t5} that this last portion resembles the first one, the reference.
-\begin{figure}[!htb]
 \begin{center}
-      \includegraphics[scale=0.6]{figures/timeseries-t5.png}
-      \caption[]{Time-series for dataset T5}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-r5.png}
+  \captionof{figure}{Time-series for dataset R5}
+  \label{fig:timeseries-r5}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-t5.png}
+  \captionof{figure}{Time-series for dataset T5}
   \label{fig:timeseries-t5}
+\end{minipage}
 \end{center}
-\end{figure}
 
@@
 This is still a different distribution that we want to alert.
-\begin{figure}[!htb]
+Figure \ref{fig:JSD-signal-zoom-test05} is a zoom in of the first 2 millionth events and once again we have no false positives.
 \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-test05.png}
-      \caption[]{Distance (JSD) signal and threshold}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test05.png}
+  \captionof{figure}{Distance (JSD) signal and threshold}
   \label{fig:JSD-signal-test05}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test05.png}
+  \captionof{figure}{Zoomed in distance (JSD) signal and threshold}
+  \label{fig:JSD-signal-zoom-test05}
+\end{minipage}
 \end{center}
-\end{figure}
-
-Figure \ref{fig:JSD-signal-zoom-test05} is a zoom in of the first 2 millionth events and once again we have no false positives.
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.6]{figures/stream-analysis-viz-zoom-test05.png}
-      \caption[]{Zoomed in distance (JSD) signal and threshold}
-      \label{fig:JSD-signal-zoom-test05}
-    \end{center}
-\end{figure}
 
@@
 \[2535000, [(2.3389481637825504e-05, x3), (2.342775340591352e-05, x4), (0.0016641704588643558, x2)]\]
-At the 3 millionth event, we change the distribution of feature x4 to its original reference one. x3's p-value gradually increases and after 804,000 events we stop reporting this feature as divergent:
+At the 3 millionth event, we change the distribution of feature x4 to its original reference one. x4's p-value gradually increases and after 804,000 events we stop reporting this feature as divergent:
 \[3804000, [(3.0928114358452774e-07, x3), (0.0012947913998524463, x2)]\]

### 2020-06-27 chapters/validation.tex
Adds distribution change tables

@@
 Feature x2 followed a gaussian with $\mu=100$ and $\sigma=5$ like in the reference period up to the 1 millionth event mark, then a gaussian with $\mu=100$ and $\sigma=10$ up to the 2.5 millionth event, then a gaussian $\mu=100$ and $\sigma=15$ until the 3 millionth event, then a gaussian $\mu=100$ and $\sigma=20$ until the 4 millionth event and finally returned to the original gaussian with $\mu=100$ and $\sigma=5$ until the last event.
+Table \ref{tbl:multi-feat-x2-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature x2.
+\begin{table}[!htb]
+    \begin{center}
+    \begin{tabular}{|c|c|c|c|l}
+    \cline{1-4}
+    \textbf{From Event} & \textbf{To Event} & \textbf{Distribution} & \multicolumn{1}{l|}{\textbf{Parameters}} &  \\ \cline{1-4}
+    0                   & 1,000,000         & Gaussian              & $\mu=100, \sigma=5$                      &  \\ \cline{1-4}
+    1,000,000           & 2,500,000         & Gaussian              & $\mu=100, \sigma=10$                     &  \\ \cline{1-4}
+    2,500,000           & 3,000,000         & Gaussian              & $\mu=100, \sigma=15$                     &  \\ \cline{1-4}
+    3,000,000           & 4,000,000         & Gaussian              & $\mu=100, \sigma=5$                      &  \\ \cline{1-4}
+    \end{tabular}
+    \end{center}
+    \caption{Feature x2 generating distributions}
+    \label{tbl:multi-feat-x2-changes}
+\end{table}
 
@@
 Feature x3 followed a uniform distribution with $a=200$ and $b=250$ like in the reference period up to the 1.5 millionth event mark, then a uniform distribution with $a=200$ and $b=220$ up to the 2 millionth event, then a uniform distribution with $a=180$ and $b=200$ until the 3.5 millionth event and finally a gaussian with $\mu=50$ and $\sigma=20$ until the last event.
+Table \ref{tbl:multi-feat-x3-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature x3.
+\begin{table}[!htb]
+    \begin{center}
+    \begin{tabular}{|c|c|c|c|l}
+    \cline{1-4}
+    \textbf{From Event} & \textbf{To Event} & \textbf{Distribution} & \multicolumn{1}{l|}{\textbf{Parameters}} &  \\ \cline{1-4}
+    0                   & 1,500,000         & Uniform               & $a=200, b=250$                           &  \\ \cline{1-4}
+    1,500,000           & 2,000,000         & Uniform               & $a=200, b=220$                           &  \\ \cline{1-4}
+    2,000,000            & 3,500,000         & Uniform               & $a=180, b=200$                           &  \\ \cline{1-4}
+    3,500,000           & 5,000,000         & Gaussian              & $\mu=50, \sigma=20$                      &  \\ \cline{1-4}
+    \end{tabular}
+    \end{center}
+    \caption{Feature x3 generating distributions}
+    \label{tbl:multi-feat-x3-changes}
+\end{table}
 
@@
 After that, it followed a uniform distribution with $a=125$ and $b=200$ until the 3 millionth event where it resumed its original uniform distribution (with $a=50$ and $b=125$) until the end.
+Table \ref{tbl:multi-feat-x4-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature x4.
+\begin{table}[!htb]
+    \begin{center}
+    \begin{tabular}{|c|c|c|c|l}
+    \cline{1-4}
+    \textbf{From Event} & \textbf{To Event} & \textbf{Distribution} & \multicolumn{1}{l|}{\textbf{Parameters}} &  \\ \cline{1-4}
+    0                   & 1,500,000         & Uniform               & $a=50, b=125$                            &  \\ \cline{1-4}
+    1,500,000           & 3,000,000         & Uniform               & $a=125, b=200$                           &  \\ \cline{1-4}
+    3,000,000           & 5,000,000         & Uniform               & $a=50, b=125$                            &  \\ \cline{1-4}
+    \end{tabular}
+    \end{center}
+    \caption{Feature x4 generating distributions}
+    \label{tbl:multi-feat-x4-changes}
+\end{table}
 

### 2020-06-27 chapters/my-work.tex
Refactor with Marco pot suggestion

@@
 Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if:
-\[  \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha / (\textit{m} - \textit{k}) \]
+\begin{equation}
+    p-value_{distance_{x}} < \alpha / (m - k)
+    \label{eq:corrected-pvalue}
+\end{equation}
 with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature.

### 2020-06-27 chapters/validation.tex
Refactor with Marco pot suggestion

@@
 Before moving to real data, we performed one last experiment, using synthetic data but this time performing a multi-feature analysis.
-Table \ref{tbl:multi-feat-ref-dataset-distros} shows the distributions and their parameters for each of the four features (x1, x2, x3 and x4) present in the reference multi-feature dataset R6.
+Table \ref{tbl:multi-feat-ref-dataset-distros} shows the underlying generating distributions and parameters for each of the four features (x1, x2, x3 and x4) in the reference multi-feature dataset R6.
 Figures \ref{fig:timeseries-r6-x1}, \ref{fig:timeseries-r6-x2}, \ref{fig:timeseries-r6-x3} and \ref{fig:timeseries-r6-x4} show the reference time-series for features x1, x2, x3 and x4 on the reference dataset R6, respectively.
@@
     2,500,000           & 3,000,000         & Gaussian              & $\mu=100, \sigma=15$                     &  \\ \cline{1-4}
-    3,000,000           & 4,000,000         & Gaussian              & $\mu=100, \sigma=5$                      &  \\ \cline{1-4}
+    3,000,000           & 4,000,000         & Gaussian              & $\mu=100, \sigma=20$                      &  \\ \cline{1-4}
+    4,000,000           & 5,000,000         & Gaussian              & $\mu=100, \sigma=5$                      &  \\ \cline{1-4}
     \end{tabular}
@@
 In this multi-feature scenario, we used the Holm-Bonferroni correction discussed in Section \ref{sec:holmbonferroni} and set the Family-Wise Error Rate (FWER) to 1\% (FWER = 0.01).
-Furthermore, instead of performing ad-hoc 1000 samples as we did previously, we use Equation \ref{eq:optimal-n-samples} to determine the minimum number of samples to make while ensuring $\gamma=1\%$ (recall Section \ref{sec:nsamples}).
+Furthermore, instead of doing 1000 samples as we did previously in an ad-hoc way, we use Equation \ref{eq:optimal-n-samples} to determine the minimum number of samples to make while ensuring $\gamma=1\%$ (recall Section \ref{sec:nsamples}).
 For our four feature analysis scenario, using as inputs  $\gamma=0.01$ and $FWER=0.01$ in Equation \ref{eq:optimal-n-samples} tells us we need to make 1840 samples minimum.
+Using 300 Spark executors, each given 3G of memory, we finished the batch phase in 418 seconds.
 
-In our stream analysis, we periodically performed our divergence test using the Holm-Bonferroni correction \ref{sec:holmbonferroni}.
-The frequency of the test used was a user-defined constant of 1000 events.
-We generated a list of alerts where each one contained the alert timestamp and a list of divergent features with the corresponding p-value.
-We achieved a throughput of 24,630 transactions per second (TPS).
-This value can be made higher if we decrease the frequency of the test.
+In Section \ref{sec:stream-phase}, we mention that, in our stream analysis, we perform our multiple hypothesis divergence test on a fixed user-defined frequency.
+Finding an optimal frequency value for the test falls out of scope for this Thesis, but ideally we set it to be large enough so that we do less computations and achieve higher throughput but still get alerts in useful time.
+In this experiment, we performed the divergence test with the Holm-Bonferroni correction \ref{sec:holmbonferroni} for every new event (frequency of 1 event).
+We did this to collect a list of ordered corrected p-values for each feature.
+A corrected p-value (derived from Equation \ref{eq:corrected-pvalue}) is given by:
+\begin{equation}
+    p-value_{corrected} = p-value_{distance_{x}} * (m - k)
+    \label{for:corrected-pvalue}
+\end{equation}
+With this list, for each feature, we are able to plot the corrected p-values on each event and compare it with the FWER threshold $\alpha$.
+Note that despite performing the test at maximum frequency (in real-time) we still achieved a throughput of 17,921 transactions per second (TPS), even if only for four features.
 
-The first alert we obtain is:
-\[1001000, [(0.0019999568566211323, x2)]\]
-This tells us that at event 1,001,000 the only diverging feature in the system is feature \textit{x2} and what the p-value for the feature value at that time is about 0.2\%.
-This is consistent with our target period induced changes.
-We introduced a change in distribution at the 1 millionth event for feature x2.
-We caught this change 1000 events later because of our test frequency, which is adjustable to obtain results as recent as we need them to be.
+In the following plots, unlike the JSD signal plots presented before, a feature is in a divergent state if its p-value is below the threshold, according to Equation \ref{eq:corrected-pvalue}.
+Working with corrected p-values (Formula \ref{for:corrected-pvalue}), a feature is considered divergent if its corrected p-value is below the threshold.
 
-The second alert is:
-\[1501000, [(0.001746900672135232, x2), (0.001989149448289207, x4), (0.0019999899243986485, x3)]\]
-We observe that besides x2, features x3 and x4 are now considered divergent as well.
-This is consistent with the changes we introduced for both x3 and x4 at the 1.5 millionth event.
-Again, we get the alert 1000 events after, exactly then the test is performed.
+In Figure \ref{fig:x1-corrected-pvalues} we see the plot of corrected p-values for feature x1 and the probability threshold black horizontal line at the bottom.
+Recall that feature x1' values were generated by a Gaussian distribution for both the reference period (Figure \ref{fig:timeseries-r6-x1}) and the target period (Figure \ref{fig:timeseries-t6-x1}), with the same parameters.
+We observe that the corrected p-values never pass below the threshold, as we expected.
+Hence, this feature is never alerted as divergent.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x1-corrected-pvalues.png}
+      \caption{Feature x1 corrected p-values and threshold}
+      \label{fig:x1-corrected-pvalues}
+    \end{center}
+\end{figure}
 
-Note that the p-value gives us a measure of divergence and allows us to order features from most divergent to least.
-The lower the p-value (probability value) the more divergent the feature is considered.
-For example, feature x3 has its distribution further changed at the 2 millionth event.
-At the 2 millionth event, we have the following state:
-\[2000000, [(5.953078867981798e-05, x4), (0.0012242703961621304, x3), (0.0017464046986205295, x2)]\]
-Note that x4 has the lowest p-value, being considered the most divergent feature.
-However, because we change feature x3 at the 2 millionth mark, the p-values observed for feature x3 gradually grow and, at the 2,535,000-th event, feature x3 is now considered the most divergent feature:
-\[2535000, [(2.3389481637825504e-05, x3), (2.342775340591352e-05, x4), (0.0016641704588643558, x2)]\]
-At the 3 millionth event, we change the distribution of feature x4 to its original reference one. x4's p-value gradually increases and after 804,000 events we stop reporting this feature as divergent:
-\[3804000, [(3.0928114358452774e-07, x3), (0.0012947913998524463, x2)]\]
-Similarly, we change feature x2's underlying distribution at the 4 millionth event to its original one.
-Its p-value gradually increases until we no longer report it as divergent.
-It took 674,000 events for the feature p-value to raise above the threshold:
-\[4674000, [(3.0922177229886216e-07, x3)]\]
-Feature x3 never returns to its original distribution, so up until the last event (5th million one) we report it as divergent:
-\[5000000, [(3.0922176674774704e-07, x3)]\]
+Figure \ref{fig:x2-corrected-pvalues} shows a plot of feature x2's corrected p-values and the FWER threshold.
+Feature x2 underlying distribution was changed four times (Table \ref{tbl:multi-feat-x2-changes}): at the 1, 2.5, 3 and 4 millionth events.
+The first change is consistent with the sudden drop in corrected p-value for this feature, as we cross below the threshold and enter a divergent state.
+Around a thousand events after the change we saw p-values of 0.2\% (or 0.002).
+The values keep dropping as time passes and as we introduce bigger changes in distribution.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x2-corrected-pvalues.png}
+      \caption{Feature x2 corrected p-values and threshold}
+      \label{fig:x2-corrected-pvalues}
+    \end{center}
+\end{figure}
+We resume the reference or original distribution at the 4 millionth event.
+However, as we can see from a close-up on the original plot (Figure \ref{fig:x2-corrected-pvalues-zoom}), it takes around 600,000 events for the target sliding histogram aggregation to once again resemble the reference one, resulting in a high enough p-value and thus stop alerting x2 as divergent.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x2-corrected-pvalues-zoom2.png}
+      \caption{Feature x2 zoomed in corrected p-values and threshold}
+      \label{fig:x2-corrected-pvalues-zoom}
+    \end{center}
+\end{figure}
+
+Feature x3 was the only feature that never returned to its original distribution and is the only one that should be reported until the end as divergent.
+This is in fact the case as illustrated in Figure \ref{fig:x3-corrected-pvalues}.
+The first change to x3's generating distribution was introduced at the 1.5 millionth event and we clearly see the corrected p-value rapidly approaching values close to zero.
+For the rest of the dataset, this x3's p-value remains below the threshold, as we never swap to its original distribution.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x3-corrected-pvalues.png}
+      \caption{Feature x3 corrected p-values and threshold}
+      \label{fig:x3-corrected-pvalues}
+    \end{center}
+\end{figure}
+
+Finally, feature x4's corrected p-value plot is given by Figure \ref{fig:x4-corrected-pvalues}.
+Similarly to feature x2, we resume the original distribution at a certain point, in this case at the 3-millionth event.
+However, as we can see from a close-up on the original plot (Figure \ref{fig:x4-corrected-pvalues-zoom}), it takes around 810,000 events for the target sliding window aggregation to be considered similar enough to the reference one and stop alerting x4 as divergent.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x4-corrected-pvalues.png}
+      \caption{Feature x4 corrected p-values and threshold}
+      \label{fig:x4-corrected-pvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x4-corrected-pvalues-zoom2.png}
+      \caption{Feature x4 zoomed in corrected p-values and threshold}
+      \label{fig:x4-corrected-pvalues-zoom}
+    \end{center}
+\end{figure}
+
+With this experiment, we further validate that the p-value is a measure of how divergent a feature is: the closer to zero, the lower the probability of the statistical test result and hence the more divergent a feature is considered to be.
+At the 1.5 millionth event, we swap both x3 and x4's distribution.
+For the first 2,535,000 events, x4 has a lower p-value than x3 and hence is considered more divergent.
+However, after that, we change x3's distribution so aggressively that x3 starts being reported as most divergent, with a lower p-value than x4.
 

### 2020-06-28 chapters/problem-statement.tex
EMA-like histogram
Method description done
problem statement

@@
 
-\textcolor{red}{INCOMPLETE}
+With this Chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.
 
-With this Chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.
+\section{The problem}
+Many real-time stream monitoring systems are static once deployed in a production environment.
+Engineers of said systems configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data.
+They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system.
+Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.
+The problem at hand is determining when to reconfigure the system, whatever it may be (\textit{e.g.} a Machine Learning model), based on streaming analysis.
 
@@
 
-    \item \textbf{Explainable alerts:} understand why each alert was raised
+    \item \textbf{Explainable alerts:} have some measure of divergence to understand why an alert was raised
 \end{itemize}
+
+\section{Proposal}
+We devise a two-phased (batch and streaming phases) and two-windowed (reference and target sliding windows) method.
+We devise a sliding window approximate histogram aggregation that encodes data distributions and provide a method for incrementally maintaining it with constant time and space complexity for every growing volumes of data, ideal for the streaming scenario.
+Given a user-defined reference period and a stream of data to process, our method guarantees to monitor each event' fields and alert if they are considered divergent from their reference distribution.
+We provide a measure of how divergent each field (or feature) is.
+Based on the number of divergent features and the corresponding divergence measure, a system administrator has an easier time determining when to reconfigure the system.
 
@@
 The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time.
-
 To that end, we devise a series of research questions to be answered by the end of this Thesis:
 
-\begin{enumerate}
-    \item \textbf{RQ1:}
-    \item \textbf{RQ2:}
-    \item \textbf{RQ3:}
+\begin{enumerate}[label=\textbf{(RQ\arabic*)}]
+    \item What aggregation can we use to properly encode data distributions?
+    \item How can we maintain said aggregation in a sliding window fashion in real-time with constant memory usage?
+    \item How can we measure divergence between the reference and sliding window aggregations?
+    \item Based on a divergence measure, when should we raise an alert?
 \end{enumerate}

### 2020-06-28 chapters/validation.tex
problem statement

@@
 
+\clearpage
 \section{Experiments with real data}
+
+Due to time constraints, we experimented with data from only one merchant. The merchant's entity will remain anonymous due to legal reasons. \textcolor{red}{merchancs use-case, how to define?}
+
+Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets from DATEX to DATEY.
+We used the older one as the reference dataset and the other one as the target one.
+Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 182 numerical and categorical features.
+Both datasets were known to suffer a lot from concept drift.
+Additionally, the data scientists at Feedzai performed some data preprocessing and cleaning, removing some of the missing values, but only on the reference dataset, which further augmented the differences between the target and reference datasets.
+Because of these reasons, we expected a lot of divergence alarms.
+
+TALK ABOUT DATASETS (size)
+
+In our first analysis, for the batch phase, we used 100 Spark executors, each with 12G of memory.
+We also determined that one week worth of data for this merchant was on average equivalent to 37,803 events.
+Hence, we set the half-life $n_{1/2}=37,803$ and discard events whose contribution is 6\% or less, which is equivalent to using windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).

### 2020-06-28 chapters/validation.tex
Nearly done with experiments

--- 2020-06-27 chapters/validation.tex
@@
 
+\clearpage
 \section{Experiments with real data}
+
+Due to time constraints, we experimented with data from only one merchant.
+The merchant's entity as well as feature names will remain anonymous due to legal reasons.
+\textcolor{red}{merchancs use-case, how to define?}
+
+Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets that combined covered a period of 1 year and 9 months.
+We used the older one as the reference dataset and the other one as the target one.
+Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus categorical features.
+Table \ref{tbl:merchant1-datasets-summary} shows summary statistics for both datasets.
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|c|}
+        \hline
+        \textbf{Dataset} & \textbf{From}              & \textbf{To}           & \multicolumn{1}{l|}{\textbf{Transactions}} \\ \hline
+        Reference        & Tuesday, September 5, 2017 & Friday, June 29, 2018 & 1,604,509                                  \\ \hline
+        Target           & Friday, June 29, 2018      & Sunday, June 30, 2019 & 4,032,505                                  \\ \hline
+        \end{tabular}
+        \caption{Summary statistics for reference and target merchant datasets}
+        \label{tbl:merchant1-datasets-summary}
+    \end{center}
+\end{table}
+
+In the batch phase, we used 100 Spark executors, each with 12G of memory.
+We determined that one week worth of data for this merchant was on average equivalent to 37,803 events.
+We set the half-life to be $n_{1/2}=37,803$ and discard events whose contribution is 6\% or less, which is equivalent to using windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).
+Hence, our samples' windows had 151,212 events.
+Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:nsamples}) of 10\% we needed to make a minimum of 42,137 samples.
+We conclude the batch analysis in 5 hours and 34 minutes.
+
+After the batch phase, we were in possession of the necessary inputs for the streaming phase (detailed in Section \ref{sec:batch-artifacts-summary}), namely, for each feature, the reference histograms, the list of observed divergence values and the last sample's histogram.
+We then used the target dataset, simulating a stream of ordered events.
+Our monitoring system processed event by event and reported divergent features and associated p-values.
+We performed our multiple hypothesis with the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for each 1000 events.
+It took us 382,363 milliseconds to process the 4,032,505 transactions, amounting to a throughput of 10,546 transactions per second (TPS).
+
+Our system produced a large list of alerts.
+In fact, we had alerts from nearly the beginning of the target period until the end.
+To try and make sense of these alerts, we questioned Feedzai's data scientists that worked directly with these data sources.
+We found out that this merchant and in particular these datasets were known to suffer a lot from concept drift.
+Additionally, the data scientists performed some data preprocessing and cleaning, removing some of the missing values, but only on the reference dataset, which further augmented the differences between the target and reference datasets.
+These seemed like plausible reasons for the large number of alarms generated.
+
+%x1 = completion_rate_per_card_72h
+We set out to validate this hypothesis.
+Due to the large number of features (183) we will only show the plots and visualizations drawn for one.
+Furthermore, we will refer to features with anonymized identifiers.
+Consider this first feature x1.
+Figure \textcolor{red}{THIS FIGURE MUST PIONT TO A CORRECT PVALUES PLOT}
+
+\textcolor{red}{INSERT FIGURE}
+
+From the previous plot, we can see this feature drops below the alarm threshold and never rises above it, effectively being considered divergent for the rest of the target period.
+Furthermore, it reaches p-values very close to zero and eventually becomes zero, throughout the remaining target period.
+Is such divergence possible?
+To find out, let us plot the reference and target time-series for feature x1.
+
+Figure \ref{fig:merchant-x1-reference} shows the reference time-series plot and Figure \ref{fig:merchant-x1-target} shows the target time-series for feature x1.
+\begin{center}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/merchant-x1-reference.png}
+  \captionof{figure}{x1 reference time-series}
+  \label{fig:merchant-x1-reference}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/merchant-x1-target.png}
+  \captionof{figure}{x1 target time-series}
+  \label{fig:merchant-x1-target}
+\end{minipage}
+\end{center}
+It became clear to us why this feature and many others were being alerted from the start of the dataset until the end.
+The domain of values is significantly differene.
+The reference period was cleaned of missing values, usually represented by -1.
+Hence, our reference was built expecting values from 0 to 1 with a certain distribution and in the stream analysis we found a large volume of missing values or -1.
+This explains the high number of alerts.
+
+A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
+We discard this idea and defend that we should not: missing values are indeed something we want to detect.
+For instance, if a certain entity responsible for reporting one feature stops working, we want to know that.
+This scenario is very likely to happen in a distributed Internet-of-Things (IoT) system with hundreds of devices transmitting data.
+
+From this experiment, we conclude that for high concept drift scenarios the system works properly, and alerts all divergent features.
+
+In this next experiment, we set out to test our system in low drift scenario, ....
+
+\textcolor{red}{SPLIT TARGET PERIOD IN TWO, USING SAME PERIOD GUARANTEES SIMILAR PERIODS? BUT MAYBE IT DOES NOT (LOOK AT x1 TIMESERIES, 1/3 has no MISSING VALUES AS WELL)}.

### 2020-06-28 chapters/problem-statement.tex
Experiments written

@@
 
-    \item \textbf{Low time complexity:} need for a time efficient method that updates the sliding window aggregation and raises alerts in constant time
+    \item \textbf{Low time complexity:} need for a time-efficient method that updates the sliding window aggregation and raises alerts in constant time
 
@@
 We devise a sliding window approximate histogram aggregation that encodes data distributions and provide a method for incrementally maintaining it with constant time and space complexity for every growing volumes of data, ideal for the streaming scenario.
-Given a user-defined reference period and a stream of data to process, our method guarantees to monitor each event' fields and alert if they are considered divergent from their reference distribution.
+Given a user-defined reference period and a stream of data to process, our method guarantees to monitor each event's fields and alert if they are considered divergent from their reference distribution.
 We provide a measure of how divergent each field (or feature) is.

### 2020-06-28 chapters/validation.tex
Experiments written

@@
 Figure \ref{fig:JSD-signal-test04} shows the computed JSD signal for the target period.
-As expected, our JSD signal increases at the 2 millionth and 3 millionth event marks, because the distribution parameters change.
+As expected, our JSD signal increases at the 2 millionth and 3 millionth event mark, because the distribution parameters change.
 At the 4 millionth event, we use the original reference distribution and notice that the JSD signal returns to values below the threshold, considering it a normal state, as we intended.
-Figure \ref{fig:JSD-signal-zoom-test04} is a zoom in on the first 2 millionth events, before any change occurs, and we see no false positives.
+Figure \ref{fig:JSD-signal-zoom-test04} is a zoom in on the first 2 millionth events before any change occurs, and we see no false positives.
 
@@
 
-In Section \ref{sec:stream-phase}, we mention that, in our stream analysis, we perform our multiple hypothesis divergence test on a fixed user-defined frequency.
-Finding an optimal frequency value for the test falls out of scope for this Thesis, but ideally we set it to be large enough so that we do less computations and achieve higher throughput but still get alerts in useful time.
+In Section \ref{sec:stream-phase}, we mention that, in our stream analysis, we perform our multiple hypothesis divergence tests on a fixed user-defined frequency.
+Finding an optimal frequency value for the test falls out of scope for this Thesis, but ideally, we set it to be large enough so that we do fewer computations and achieve higher throughput but still get alerts in useful time.
 In this experiment, we performed the divergence test with the Holm-Bonferroni correction \ref{sec:holmbonferroni} for every new event (frequency of 1 event).
@@
 \end{equation}
-With this list, for each feature, we are able to plot the corrected p-values on each event and compare it with the FWER threshold $\alpha$.
-Note that despite performing the test at maximum frequency (in real-time) we still achieved a throughput of 17,921 transactions per second (TPS), even if only for four features.
+With this list, for each feature, we can plot the corrected p-values on each event and compare it with the FWER threshold $\alpha$.
+Note that despite performing the test at the maximum frequency (in real-time) we still achieved a throughput of 17,921 transactions per second (TPS), even if only for four features.
 
@@
 
-Due to time constraints, we experimented with data from only one merchant.
-The merchant's entity as well as feature names will remain anonymous due to legal reasons.
-\textcolor{red}{merchancs use-case, how to define?}
+Due to time constraints, we experimented with data from only one merchant. The merchant's entity as well as feature names will remain anonymous due to legal reasons.
 
-Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets that combined covered a period of 1 year and 9 months.
+Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that combined covered a period of 1 year and 9 months.
 We used the older one as the reference dataset and the other one as the target one.
@@
 In the batch phase, we used 100 Spark executors, each with 12G of memory.
-We determined that one week worth of data for this merchant was on average equivalent to 37,803 events.
+We determined that one week's worth of data for this merchant was on average equivalent to 37,803 events.
 We set the half-life to be $n_{1/2}=37,803$ and discard events whose contribution is 6\% or less, which is equivalent to using windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).
@@
 
-After the batch phase, we were in possession of the necessary inputs for the streaming phase (detailed in Section \ref{sec:batch-artifacts-summary}), namely, for each feature, the reference histograms, the list of observed divergence values and the last sample's histogram.
+After the batch phase, we had the necessary inputs for the streaming phase (detailed in Section \ref{sec:batch-artifacts-summary}), namely, for each feature, the reference histograms, the list of observed divergence values and the last sample's histogram.
 We then used the target dataset, simulating a stream of ordered events.
 Our monitoring system processed event by event and reported divergent features and associated p-values.
-We performed our multiple hypothesis with the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for each 1000 events.
+We performed our multiple hypotheses with the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for every 1000 events.
 It took us 382,363 milliseconds to process the 4,032,505 transactions, amounting to a throughput of 10,546 transactions per second (TPS).
@@
 Our system produced a large list of alerts.
-In fact, we had alerts from nearly the beginning of the target period until the end.
+We had alerts from nearly the beginning of the target period until the end.
 To try and make sense of these alerts, we questioned Feedzai's data scientists that worked directly with these data sources.
@@
 We set out to validate this hypothesis.
-Due to the large number of features (183) we will only show the plots and visualizations drawn for one.
-Furthermore, we will refer to features with anonymized identifiers.
-Consider this first feature x1.
-Figure \textcolor{red}{THIS FIGURE MUST PIONT TO A CORRECT PVALUES PLOT}
-
-\textcolor{red}{INSERT FIGURE}
-
-From the previous plot, we can see this feature drops below the alarm threshold and never rises above it, effectively being considered divergent for the rest of the target period.
+Due to the large number of features (183) we only show the plots and visualizations drawn for one.
+Furthermore, we refer to features with anonymized identifiers, like our example feature, x1.
+Figure \ref{fig:merchant-x1-corrected-pvalues} shows the corrected p-values (Formula \ref{for:corrected-pvalue}) for feature x1 at for each event of the time-series and Figure \ref{fig:merchant-x1-corrected-pvalues-zoom} provides a closer look on the region where the p-values drop below the threshold (FWER = 1\%).
+\begin{center}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/merchant-x1-corrected-pvalues.png}
+  \captionof{figure}{x1 corrected p-values and threshold}
+  \label{fig:merchant-x1-corrected-pvalues}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/merchant-x1-corrected-pvalues-zoom.png}
+  \captionof{figure}{x1 zoomed in corrected p-values and threshold}
+  \label{fig:merchant-x1-corrected-pvalues-zoom}
+\end{minipage}
+\end{center}
+We see that this feature drops below the alarm threshold and never rises above it, effectively being considered divergent for the rest of the target period.
 Furthermore, it reaches p-values very close to zero and eventually becomes zero, throughout the remaining target period.
@@
 To find out, let us plot the reference and target time-series for feature x1.
-
 Figure \ref{fig:merchant-x1-reference} shows the reference time-series plot and Figure \ref{fig:merchant-x1-target} shows the target time-series for feature x1.
@@
 It became clear to us why this feature and many others were being alerted from the start of the dataset until the end.
-The domain of values is significantly differene.
+The domain of values is significantly different.
 The reference period was cleaned of missing values, usually represented by -1.
-Hence, our reference was built expecting values from 0 to 1 with a certain distribution and in the stream analysis we found a large volume of missing values or -1.
+Hence, our reference was built expecting values from 0 to 1 with a certain distribution and in the stream analysis, we found a large volume of missing values or -1.
 This explains the high number of alerts.
@@
 
-In this next experiment, we set out to test our system in low drift scenario, ....
+\subsection{Similar reference and target periods}
+In this next experiment, we set out to test our system in a low drift scenario,...
 
-\textcolor{red}{SPLIT TARGET PERIOD IN TWO, USING SAME PERIOD GUARANTEES SIMILAR PERIODS? BUT MAYBE IT DOES NOT (LOOK AT x1 TIMESERIES, 1/3 has no MISSING VALUES AS WELL)}.
+\textcolor{red}{SPLIT TARGET PERIOD IN TWO, USING SAME PERIOD GUARANTEES SIMILAR PERIODS?
+BUT MAYBE IT DOES NOT (LOOK AT x1 TARGET TIMESERIES, 1/3 has no MISSING VALUES AS WELL.
+STILL VERY DIFFERENT FROM 2/3)}.

### 2020-06-30 chapters/background.tex
Images for mechant1 features

@@
 Historically speaking, Apache Hadoop \cite{borthakur2007hadoop} \cite{Hadoop} \cite{ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005.
-Hadoop uses a distributed computation framework or programming model called MapReduce \cite{MapReduce}.
+Hadoop uses a programming model called MapReduce \cite{MapReduce}.
 This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
 using a reducer function.
-To do this they read and write from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms).
-Nearly a decade later, Apache Spark \cite{ApacheSpark} \cite{Spark} was released.
+To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms).
+Nearly a decade later, Apache Spark \cite{ApacheSpark} \cite{Spark} was released as the industry accepted successor.
 Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}.
@@
 
-On the other hand, stream processing refers to the processing of data streams.
+Stream procesisng, on the other hand, refers to the processing of data streams.
 A data stream is essentially a continuous or unbounded data set.
 This data needs to be processed sequentially and incrementally on a event-by-event basis or over sliding windows.
-Streaming data includes a wide variety of data such as logs generated by customers using mobile or web applications, sensor data from Internet-of-Things (IoT) devices or credit card transactions.
+Streaming data varies widely, from logs generated by customers using mobile or web applications, sensor data from Internet-of-Things (IoT) devices to credit card transactions.
 This is in contrast to batch processing which works on a finite data set.
@@
 Doing this recursively for each item in the collection results in a single value: the aggregate value.
-The combine function is here on forward represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.
+The combine function is here on forward represented as $\oplus$ and the identity or neutral element of the aggregation as $\overline{\theta}$.
 

### 2020-06-30 chapters/introduction.tex
problem statement
Images for mechant1 features

@@
 Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously.
-Social networks are also responsible for generating large volumes of data,Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
+Social networks are responsible for generating large volumes of data of which,  Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
 \textit{Tweets} may contain text, media content or both.
 Cybersecurity applications are yet another example of time-sensitive, data-driven applications.
-These need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the protected system.
+The entities that work in these use-cases need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the underlying system.
 
@@
 However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, the demand for stream processing systems is increasing.
-Unlike batch processing of static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery}.
-Furthermore, in the streaming paradigm, the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}.
+Unlike batch processing of static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery} where the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}.
 \textit{Recency} is measured under different scales for different use cases.
@@
 However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques.
-A thorough analysis of stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.
+A thorough analysis of stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set typically by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.
 
@@
 Feedzai offers fraud prevention solutions for the previously presented use case of monitoring a continuous stream of credit card transactions and detecting fraud.
-We exploit multiple real-life data sets that are provided from the financial fraud space.
+We exploit real-life data sets that are provided from the financial fraud space.
 
@@
 Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time.
-In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift.
+In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift \cite{drift} \cite{gama-drift}.
 
@@
 
-The objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.
+Thus the objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.
 

### 2020-06-30 chapters/my-work.tex
Nearly done with experiments
Images for mechant1 features

@@
 An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values.
-EMA-like histograms are further explained in Section \ref{sec:ema-hist}.
+These EMA-like histograms are further explained in Section \ref{sec:ema-hist}.
 
@@
 
-\subsubsection*{Batch phase artifacts}
+\subsubsection*{Batch phase artifacts} \label{sec:batch-artifacts-summary}
 By the end of the batch phase, we should have, for each feature: 

### 2020-06-30 chapters/problem-statement.tex
Images for mechant1 features

@@
 Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.
-The problem at hand is determining when to reconfigure the system, whatever it may be (\textit{e.g.} a Machine Learning model), based on streaming analysis.
+The problem at hand is to determine when to reconfigure the system, \textit{e.g.} a Machine Learning model, based on streaming analysis.
 
@@
 In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods.
-However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the below characteristics:
+However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the desired characteristics, namely:
 
@@
 
-    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout run-time, the opposite of an online learning algorithm
+    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout run-time
 
@@
 
-    \item \textbf{Explainable alerts:} have some measure of divergence to understand why an alert was raised
+    \item \textbf{Explainable alerts:} have an interpretable measure of divergence to understand why an alert was raised
 \end{itemize}
@@
 We devise a sliding window approximate histogram aggregation that encodes data distributions and provide a method for incrementally maintaining it with constant time and space complexity for every growing volumes of data, ideal for the streaming scenario.
-Given a user-defined reference period and a stream of data to process, our method guarantees to monitor each event's fields and alert if they are considered divergent from their reference distribution.
+Given a user-defined reference period and a stream of data to process, our method guarantees to monitor each event's fields and alert if they are considered divergent relative to their reference distribution.
 We provide a measure of how divergent each field (or feature) is.
@@
 
-\begin{enumerate}[label=\textbf{(RQ\arabic*)}]
+\begin{enumerate}[leftmargin=1.5cm, label=\textbf{(RQ\arabic*)}]
     \item What aggregation can we use to properly encode data distributions?
-    \item How can we maintain said aggregation in a sliding window fashion in real-time with constant memory usage?
+    \item How can we maintain such an aggregation in a sliding window fashion in real-time with constant memory usage?
     \item How can we measure divergence between the reference and sliding window aggregations?
@@
 \end{enumerate}
+
+%Assumptions section
+%
+%
+%
+%
+%

### 2020-06-30 chapters/validation.tex
Images for mechant1 features

@@
 In Section \ref{sec:ema-hist}, we mentioned that we choose to discard events whose contribution is less than 6\%.
-We set a half-life $n_{1/2}$ value and then multiply it by a constant factor of 4 to get the size of the samples to make (a larger sample would contain events whose contribution was equal or less than 6\%).
+We set a half-life $n_{1/2}$ value and then multiply it by a constant factor of 4 to get the size of the samples to make (a larger sample would contain events whose contribution was equal to or less than 6\%).
 
@@
 Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that combined covered a period of 1 year and 9 months.
-We used the older one as the reference dataset and the other one as the target one.
+We used the older one as the reference dataset and the more recent one as the target one.
 Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus categorical features.
@@
 Hence, our samples' windows had 151,212 events.
-Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:nsamples}) of 10\% we needed to make a minimum of 42,137 samples.
-We conclude the batch analysis in 5 hours and 34 minutes.
+Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:nsamples}) of 10\% we needed to make a minimum of 42,137 samples, concluding the batch analysis in 5 hours and 34 minutes.
 
@@
 We then used the target dataset, simulating a stream of ordered events.
-Our monitoring system processed event by event and reported divergent features and associated p-values.
+Our monitoring system processed the dataset event by event in sequential, time-ordered fashion, emulating a stream computation and reported divergent features and associated p-values.
 We performed our multiple hypotheses with the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for every 1000 events.
-It took us 382,363 milliseconds to process the 4,032,505 transactions, amounting to a throughput of 10,546 transactions per second (TPS).
+It took us 382,363 milliseconds (roughly 6 minutes) to process the 4,032,505 transactions, amounting to a throughput of 10,546 transactions per second (TPS).
 
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x1-reference.png}
+      \caption{Merchant feature x1 reference time-series}
+      \label{fig:merchant-x1-reference}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x1-target.png}
+      \caption{Merchant feature x1 target time-series}
+      \label{fig:merchant-x1-target}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x2-reference.png}
+      \caption{Merchant feature x2 reference time-series}
+      \label{fig:merchant-x2-reference}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x2-target.png}
+      \caption{Merchant feature x2 target time-series}
+      \label{fig:merchant-x2-target}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x3-reference.png}
+      \caption{Merchant feature x3 reference time-series}
+      \label{fig:merchant-x3-reference}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x3-target.png}
+      \caption{Merchant feature x3 target time-series}
+      \label{fig:merchant-x3-target}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x4-reference.png}
+      \caption{Merchant feature x4 reference time-series}
+      \label{fig:merchant-x4-reference}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x4-target.png}
+      \caption{Merchant feature x4 target time-series}
+      \label{fig:merchant-x4-target}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x5-reference.png}
+      \caption{Merchant feature x5 reference time-series}
+      \label{fig:merchant-x5-reference}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x5-target.png}
+      \caption{Merchant feature x5 target time-series}
+      \label{fig:merchant-x5-target}
+    \end{center}
+\end{figure}
+
+%This merchant and in particular these datasets were known to suffer a lot from concept drift.
 Our system produced a large list of alerts.
 We had alerts from nearly the beginning of the target period until the end.
-To try and make sense of these alerts, we questioned Feedzai's data scientists that worked directly with these data sources.
-We found out that this merchant and in particular these datasets were known to suffer a lot from concept drift.
-Additionally, the data scientists performed some data preprocessing and cleaning, removing some of the missing values, but only on the reference dataset, which further augmented the differences between the target and reference datasets.
-These seemed like plausible reasons for the large number of alarms generated.
 
-%x1 = completion_rate_per_card_72h
-We set out to validate this hypothesis.
-Due to the large number of features (183) we only show the plots and visualizations drawn for one.
+%low drift:
+    %x1 - max_amt_rate_per_location_72h
+    %x2 - num_preauths_per_card_1w
+    %x3 - user_avg_amount_completions_1h
+    %x4 - completions_rate_per_location_zip3_72h
+
+%high drift:
+    %x5 - completion_rate_per_card_72h
+
+%Due to the large number of features (183) we only show the plots and visualizations drawn for some.
 Furthermore, we refer to features with anonymized identifiers, like our example feature, x1.
 Figure \ref{fig:merchant-x1-corrected-pvalues} shows the corrected p-values (Formula \ref{for:corrected-pvalue}) for feature x1 at for each event of the time-series and Figure \ref{fig:merchant-x1-corrected-pvalues-zoom} provides a closer look on the region where the p-values drop below the threshold (FWER = 1\%).
-\begin{center}
-\begin{minipage}{.5\textwidth}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/merchant-x1-corrected-pvalues.png}
-  \captionof{figure}{x1 corrected p-values and threshold}
-  \label{fig:merchant-x1-corrected-pvalues}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/merchant-x1-corrected-pvalues-zoom.png}
-  \captionof{figure}{x1 zoomed in corrected p-values and threshold}
-  \label{fig:merchant-x1-corrected-pvalues-zoom}
-\end{minipage}
-\end{center}
-We see that this feature drops below the alarm threshold and never rises above it, effectively being considered divergent for the rest of the target period.
+
+%We see that this feature drops below the alarm threshold and never rises above it, effectively being considered divergent for the rest of the target period.
 Furthermore, it reaches p-values very close to zero and eventually becomes zero, throughout the remaining target period.
@@
 Figure \ref{fig:merchant-x1-reference} shows the reference time-series plot and Figure \ref{fig:merchant-x1-target} shows the target time-series for feature x1.
-\begin{center}
-\begin{minipage}{.5\textwidth}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/merchant-x1-reference.png}
-  \captionof{figure}{x1 reference time-series}
-  \label{fig:merchant-x1-reference}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/merchant-x1-target.png}
-  \captionof{figure}{x1 target time-series}
-  \label{fig:merchant-x1-target}
-\end{minipage}
-\end{center}
-It became clear to us why this feature and many others were being alerted from the start of the dataset until the end.
+
+%It became clear to us why this feature and many others were being alerted from the start of the dataset until the end.
 The domain of values is significantly different.
@@
 
-A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
+%A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
 We discard this idea and defend that we should not: missing values are indeed something we want to detect.
@@
 
-From this experiment, we conclude that for high concept drift scenarios the system works properly, and alerts all divergent features.
+%missing values streaming
 
-\subsection{Similar reference and target periods}
+\subsection{Same reference and target periods}
 In this next experiment, we set out to test our system in a low drift scenario,...
-
-\textcolor{red}{SPLIT TARGET PERIOD IN TWO, USING SAME PERIOD GUARANTEES SIMILAR PERIODS?
-BUT MAYBE IT DOES NOT (LOOK AT x1 TARGET TIMESERIES, 1/3 has no MISSING VALUES AS WELL.
-STILL VERY DIFFERENT FROM 2/3)}.

### 2020-06-30 chapters/validation.tex
More visualizations

@@
 
+This merchant and in particular these datasets were known to suffer a lot from concept drift.
+Our system produced a large list of alerts, from nearly the beginning of the target period until the end.
+Due to the large number of features (183) we only show some plots.
+Furthermore, we refer to features with anonymized identifiers, like feature $x_t$.
+In the following plots, we show the corrected p-values (Formula \ref{for:corrected-pvalue}) for some features at each event of the time-series and the FWER threshold set at 1\%.
+A feature is considered divergent at a given timestamp if below that threshold line.
+
+%feats = [max_amt_rate_per_location_72h, num_preauths_per_card_1w, user_avg_amount_completions_1h, completions_rate_per_location_zip3_72h, completion_rate_per_card_72h]
+%low drift:
+    %x1 - max_amt_rate_per_location_72h
+    %x2 - num_preauths_per_card_1w
+    %x3 - user_avg_amount_completions_1h
+    %REMOVED %x4 - completions_rate_per_location_zip3_72h
+%high drift:
+    %x4 - completion_rate_per_card_72h
+
+%We see that this feature drops below the alarm threshold and never rises above it, effectively being considered divergent for the rest of the target period.
+Furthermore, it reaches p-values very close to zero and eventually becomes zero, throughout the remaining target period.
+Is such divergence possible?
+To find out, let us plot the reference and target time-series for feature x1.
+Figure \ref{fig:merchant-x1-reference} shows the reference time-series plot and Figure \ref{fig:merchant-x1-target} shows the target time-series for feature x1.
+
+%It became clear to us why this feature and many others were being alerted from the start of the dataset until the end.
+The domain of values is significantly different.
+The reference period was cleaned of missing values, usually represented by -1.
+Hence, our reference was built expecting values from 0 to 1 with a certain distribution and in the stream analysis, we found a large volume of missing values or -1.
+This explains the high number of alerts.
+
+%A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
+We discard this idea and defend that we should not: missing values are indeed something we want to detect.
+For instance, if a certain entity responsible for reporting one feature stops working, we want to know that.
+This scenario is very likely to happen in a distributed Internet-of-Things (IoT) system with hundreds of devices transmitting data.
+
 \begin{figure}[!htb]
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x5-reference.png}
-      \caption{Merchant feature x5 reference time-series}
-      \label{fig:merchant-x5-reference}
+      \includegraphics[scale=0.5]{figures/merchant-x1-correctedpvalues.png}
+      \caption{Merchant feature x1 corrected p-values}
+      \label{fig:merchant-x1-correctedpvalues}
     \end{center}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x5-target.png}
-      \caption{Merchant feature x5 target time-series}
-      \label{fig:merchant-x5-target}
+      \includegraphics[scale=0.5]{figures/merchant-x2-correctedpvalues.png}
+      \caption{Merchant feature x2 corrected p-values}
+      \label{fig:merchant-x2-correctedpvalues}
     \end{center}
@@
 
-%This merchant and in particular these datasets were known to suffer a lot from concept drift.
-Our system produced a large list of alerts.
-We had alerts from nearly the beginning of the target period until the end.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x3-correctedpvalues.png}
+      \caption{Merchant feature x3 corrected p-values}
+      \label{fig:merchant-x3-correctedpvalues}
+    \end{center}
+\end{figure}
 
-%low drift:
-    %x1 - max_amt_rate_per_location_72h
-    %x2 - num_preauths_per_card_1w
-    %x3 - user_avg_amount_completions_1h
-    %x4 - completions_rate_per_location_zip3_72h
-
-%high drift:
-    %x5 - completion_rate_per_card_72h
-
-%Due to the large number of features (183) we only show the plots and visualizations drawn for some.
-Furthermore, we refer to features with anonymized identifiers, like our example feature, x1.
-Figure \ref{fig:merchant-x1-corrected-pvalues} shows the corrected p-values (Formula \ref{for:corrected-pvalue}) for feature x1 at for each event of the time-series and Figure \ref{fig:merchant-x1-corrected-pvalues-zoom} provides a closer look on the region where the p-values drop below the threshold (FWER = 1\%).
-
-%We see that this feature drops below the alarm threshold and never rises above it, effectively being considered divergent for the rest of the target period.
-Furthermore, it reaches p-values very close to zero and eventually becomes zero, throughout the remaining target period.
-Is such divergence possible?
-To find out, let us plot the reference and target time-series for feature x1.
-Figure \ref{fig:merchant-x1-reference} shows the reference time-series plot and Figure \ref{fig:merchant-x1-target} shows the target time-series for feature x1.
-
-%It became clear to us why this feature and many others were being alerted from the start of the dataset until the end.
-The domain of values is significantly different.
-The reference period was cleaned of missing values, usually represented by -1.
-Hence, our reference was built expecting values from 0 to 1 with a certain distribution and in the stream analysis, we found a large volume of missing values or -1.
-This explains the high number of alerts.
-
-%A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
-We discard this idea and defend that we should not: missing values are indeed something we want to detect.
-For instance, if a certain entity responsible for reporting one feature stops working, we want to know that.
-This scenario is very likely to happen in a distributed Internet-of-Things (IoT) system with hundreds of devices transmitting data.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x4-correctedpvalues.png}
+      \caption{Merchant feature x4 corrected p-values}
+      \label{fig:merchant-x4-correctedpvalues}
+    \end{center}
+\end{figure}
 

### 2020-06-30 chapters/validation.tex
Merchant 1 first experiment written

@@
 
-Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that combined covered a period of 1 year and 9 months.
-We used the older one as the reference dataset and the more recent one as the target one.
+Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that combined covered a period of roughly 1 year and 9 months.
+We used the older one as the reference dataset and the more recent one as the target period.
 Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus categorical features.
@@
 We determined that one week's worth of data for this merchant was on average equivalent to 37,803 events.
-We set the half-life to be $n_{1/2}=37,803$ and discard events whose contribution is 6\% or less, which is equivalent to using windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).
+We set the half-life to be $n_{1/2}=37,803$ and discard events whose contribution is 6\% or less, which is equivalent to using sampling windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).
 Hence, our samples' windows had 151,212 events.
-Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:nsamples}) of 10\% we needed to make a minimum of 42,137 samples, concluding the batch analysis in 5 hours and 34 minutes.
+Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:nsamples}) of 10\% we needed to make a minimum of 42,137 samples.
+Our batch analysis took 5 hours and 34 minutes.
 
@@
 
-%feats = [max_amt_rate_per_location_72h, num_preauths_per_card_1w, user_avg_amount_completions_1h, completions_rate_per_location_zip3_72h, completion_rate_per_card_72h]
-%low drift:
-    %x1 - max_amt_rate_per_location_72h
-    %x2 - num_preauths_per_card_1w
-    %x3 - user_avg_amount_completions_1h
-    %REMOVED %x4 - completions_rate_per_location_zip3_72h
-%high drift:
-    %x4 - completion_rate_per_card_72h
-
-%We see that this feature drops below the alarm threshold and never rises above it, effectively being considered divergent for the rest of the target period.
-Furthermore, it reaches p-values very close to zero and eventually becomes zero, throughout the remaining target period.
-Is such divergence possible?
-To find out, let us plot the reference and target time-series for feature x1.
-Figure \ref{fig:merchant-x1-reference} shows the reference time-series plot and Figure \ref{fig:merchant-x1-target} shows the target time-series for feature x1.
-
-%It became clear to us why this feature and many others were being alerted from the start of the dataset until the end.
-The domain of values is significantly different.
-The reference period was cleaned of missing values, usually represented by -1.
-Hence, our reference was built expecting values from 0 to 1 with a certain distribution and in the stream analysis, we found a large volume of missing values or -1.
-This explains the high number of alerts.
-
-%A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
-We discard this idea and defend that we should not: missing values are indeed something we want to detect.
-For instance, if a certain entity responsible for reporting one feature stops working, we want to know that.
-This scenario is very likely to happen in a distributed Internet-of-Things (IoT) system with hundreds of devices transmitting data.
-
+Figure \ref{fig:merchant-x1-reference} is a plot of the reference time-series of feature x1 and Figure \ref{fig:merchant-x1-target} the corresponding target time-series.
 \begin{figure}[!htb]
@@
 \end{figure}
-
 \begin{figure}[!htb]
@@
 \end{figure}
+In Figure \ref{fig:merchant-x1-correctedpvalues} we plot the corrected p-values for the each event of the target period.
+The reference and target periods look very similar and in fact we have very few alarms for this feature, \textit{i.e.}, the number of times the line crosses to the bottom half of the plot, imposed by the threshold line.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x1-correctedpvalues.png}
+      \caption{Merchant feature x1 corrected p-values}
+      \label{fig:merchant-x1-correctedpvalues}
+    \end{center}
+\end{figure}
+Furthermore, the alarms appear to be roughly in the middle portion of the time-series, where in fact the target period appears to have less points.
 
+Now let us look at features x2 and x3. Figures \ref{fig:merchant-x2-reference} and \ref{fig:merchant-x2-target} show feature x2' time-series plots for the reference and target periods, respectively.
 \begin{figure}[!htb]
@@
 \end{figure}
-
+Figures \ref{fig:merchant-x3-reference} and \ref{fig:merchant-x3-target} show feature x3' time-series plots for the reference and target periods, respectively.
 \begin{figure}[!htb]
@@
 \end{figure}
+\clearpage
+Similarly to x1, features x2 and x3 have very similar reference and target periods (not between each other). Just like feature x1, we expect to have some alarms but still not too many.
 
+Once again, to visualize the alarms, we plot the corrected p-values for each event of the target time-series for each feature.
+Figure \ref{fig:merchant-x2-correctedpvalues} is a plot of feature x2' corrected p-values and the FWER threshold of 1\%.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x2-correctedpvalues.png}
+      \caption{Merchant feature x2 corrected p-values}
+      \label{fig:merchant-x2-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x3-correctedpvalues.png}
+      \caption{Merchant feature x3 corrected p-values}
+      \label{fig:merchant-x3-correctedpvalues}
+    \end{center}
+\end{figure}
+The corrected p-values plot shows us once again that both x2 and x3 features had some alarms but quickly rose above the threshold and resumed their reference distribution.
+This is expected from features whose reference and target time-series plots look so similar.
+
+Next, we analyze a feature with very different reference and target time-series.
+Feature x4' reference time-series is presented in Figure \ref{fig:merchant-x4-reference} and its target period in Figure \ref{fig:merchant-x4-target}.
 \begin{figure}[!htb]
@@
 \end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x1-correctedpvalues.png}
-      \caption{Merchant feature x1 corrected p-values}
-      \label{fig:merchant-x1-correctedpvalues}
-    \end{center}
-\end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x2-correctedpvalues.png}
-      \caption{Merchant feature x2 corrected p-values}
-      \label{fig:merchant-x2-correctedpvalues}
-    \end{center}
-\end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x3-correctedpvalues.png}
-      \caption{Merchant feature x3 corrected p-values}
-      \label{fig:merchant-x3-correctedpvalues}
-    \end{center}
-\end{figure}
-
+Notice that even if for the most part both reference and target periods values belong to a domain between 0 and 1, at a certain point in time, in the target time-series, a lot of -1 values appear.
+In Figure \ref{fig:merchant-x4-correctedpvalues} we plot the corrected p-values for feature x4 target time-series.
 \begin{figure}[!htb]
@@
 \end{figure}
+Like the previous analyzed features, x4 also drops below the threshold some times.
+However, unlike the other features, where feature values return to normal and alarms stopped being raised, x4 remains below the threshold line until the end of the target period, where a small spike occurs.
+The last time feature x4 was ever above the threshold line (and hence considered normal) was before the appearance of the -1 values in the target period.
+This seems to work as expected: -1 values were so off the reference domain that for here on out, as long as these values keep appearing, we will not stop reporting x4 as divergent.
 
-%missing values streaming
+The -1 value is often used as placeholder for missing values, \textit{i.e.}, if some problem occurs and the value for a feature is missing at a given timestamp or event, -1 is used.
+A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
+However, we discard this idea and defend that we should not: missing values are indeed something we want to detect.
+For instance, if a certain device responsible for reporting one feature stops working, we want to know that.
+This scenario is very likely to happen in a distributed Internet-of-Things (IoT) system with hundreds of devices transmitting data.
 
+%feats = [max_amt_rate_per_location_72h, num_preauths_per_card_1w, user_avg_amount_completions_1h, completions_rate_per_location_zip3_72h, completion_rate_per_card_72h]
+%low drift:
+    %x1 - max_amt_rate_per_location_72h
+    %x2 - num_preauths_per_card_1w
+    %x3 - user_avg_amount_completions_1h
+%high drift:
+    %x4 - completion_rate_per_card_72h
 \subsection{Same reference and target periods}

### 2020-06-30 chapters/problem-statement.tex
Assumptions,  small corrections

@@
 We devise a sliding window approximate histogram aggregation that encodes data distributions and provide a method for incrementally maintaining it with constant time and space complexity for every growing volumes of data, ideal for the streaming scenario.
-Given a user-defined reference period and a stream of data to process, our method guarantees to monitor each event's fields and alert if they are considered divergent relative to their reference distribution.
+Given a user-defined reference period and a stream of data to process, we monitor each event's fields and alert if they are considered divergent relative to their reference distribution.
 We provide a measure of how divergent each field (or feature) is.
-Based on the number of divergent features and the corresponding divergence measure, a system administrator has an easier time determining when to reconfigure the system.
+Based on this information, it is our belief that a system administrator would have an easier time determining when to reconfigure the system.
+
+\section{Assumptions}
+Our feature monitoring method assumes only that the monitored features are numerical or categorical and that the events arrive in an orderly fashion during streaming.
+If the latter is not true, our method will still work but results may be imprecise, as our sliding window histogram aggregation relies on Exponential Moving Averages (EMAs), which give different weights to each event according to its age.
 
@@
 
-\begin{enumerate}[leftmargin=1.5cm, label=\textbf{(RQ\arabic*)}]
+\begin{enumerate}[leftmargin=1.75cm, label=\textbf{(RQ\arabic*)}]
     \item What aggregation can we use to properly encode data distributions?
@@
 \end{enumerate}
-
-%Assumptions section
-%
-%
-%
-%
-%

### 2020-06-30 chapters/validation.tex
Assumptions,  small corrections

@@
 Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that combined covered a period of roughly 1 year and 9 months.
-We used the older one as the reference dataset and the more recent one as the target period.
+In this experiment, we used the older one as the reference dataset and the more recent one as the target period.
 Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus categorical features.
@@
 We then used the target dataset, simulating a stream of ordered events.
-Our monitoring system processed the dataset event by event in sequential, time-ordered fashion, emulating a stream computation and reported divergent features and associated p-values.
+Our monitoring system processed the dataset event by event in sequential, time-ordered fashion, emulating a stream computation and reported divergent features with associated p-values.
 We performed our multiple hypotheses with the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for every 1000 events.
@@
 Our system produced a large list of alerts, from nearly the beginning of the target period until the end.
-Due to the large number of features (183) we only show some plots.
+Due to a large number of features (183), we only show some plots.
 Furthermore, we refer to features with anonymized identifiers, like feature $x_t$.
@@
 \clearpage
-Similarly to x1, features x2 and x3 have very similar reference and target periods (not between each other). Just like feature x1, we expect to have some alarms but still not too many.
-
+Similarly to x1, features x2 and x3 have very similar reference and target periods (not between each other).
+Just like feature x1, we expect to have some alarms but still not too many.
 Once again, to visualize the alarms, we plot the corrected p-values for each event of the target time-series for each feature.
 Figure \ref{fig:merchant-x2-correctedpvalues} is a plot of feature x2' corrected p-values and the FWER threshold of 1\%.
+Likewise, Figure \ref{fig:merchant-x3-correctedpvalues} shows the same plot but for feature x3.
 \begin{figure}[!htb]
@@
 
-The -1 value is often used as placeholder for missing values, \textit{i.e.}, if some problem occurs and the value for a feature is missing at a given timestamp or event, -1 is used.
+The -1 value is often used as a placeholder for missing values, \textit{i.e.}, if some problem occurs and the value for a feature is missing at a given timestamp or event, -1 is used.
 A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
@@
 For instance, if a certain device responsible for reporting one feature stops working, we want to know that.
-This scenario is very likely to happen in a distributed Internet-of-Things (IoT) system with hundreds of devices transmitting data.
+This scenario is likely to happen in a distributed Internet-of-Things (IoT) system with hundreds of devices transmitting data.
 
@@
 \subsection{Same reference and target periods}
-In this next experiment, we set out to test our system in a low drift scenario,...
+In this experiment, we set out to test our system in a low drift scenario.
+To guarantee the reference and target periods are as similar as possible, we decide to use the same dataset to work as reference and target.
+We chose to use the reference dataset from the previous experiment (Reference of Table \ref{tbl:merchant1-datasets-summary}) and feed it as input for both our batch and stream analysis as the reference and target periods, respectively.
+
+\textcolor{red}{TODO: visualizations and conclusions of this experiment}

### 2020-07-01 chapters/background.tex
PDF images

@@
 Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}.
-Batch data sets are known as static data \cite{Martin-Batch-Defin}.
+Batch datasets are known as static data \cite{Martin-Batch-Defin}.
 An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes).
@@
 Stream procesisng, on the other hand, refers to the processing of data streams.
-A data stream is essentially a continuous or unbounded data set.
-This data needs to be processed sequentially and incrementally on a event-by-event basis or over sliding windows.
+A data stream is essentially a continuous or unbounded dataset.
+This data needs to be processed sequentially and incrementally on a event-by-event basis.
 Streaming data varies widely, from logs generated by customers using mobile or web applications, sensor data from Internet-of-Things (IoT) devices to credit card transactions.
-This is in contrast to batch processing which works on a finite data set.
-This distinction is important since some aggregations are very hard and even impossible to compute with 100\% accuracy for several use cases in an efficient manner on unbounded data sets.
+These examples of infinite datasets contrast with batch processing which works on finite ones.
+This distinction is important since some aggregations are very hard and even impossible to compute with 100\% accuracy for several use cases in an efficient manner on unbounded datasets.
 An example of such is counting distinct elements that flow through the data stream.
@@
 A data stream is unbounded or infinite while a bounded data stream has a beginning and an end.
-Using these definitions, the concept of an unbounded data stream contains that of a bounded data set.
-That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite data set.
+Using these definitions, the concept of an unbounded data stream contains that of a bounded dataset.
+That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite dataset.
 This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.
@@
       \includegraphics[scale=0.5]{figures/unbounded-superset-bounded.png}
-      \caption{Unbounded data streams as a superset of bounded data sets}
+      \caption{Unbounded data streams as a superset of bounded datasets}
       \label{fig:unbounded-bounded-superset}
@@
 
-In Section 2.2 we will cover exactly how an unbounded data set (also known as a data stream) can be converted into a bounded data set with the use of windows.
+In Section 2.2 we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.
 
-\section{From unbounded streams to finite data sets} \label{sec:windows}
+\section{From unbounded streams to finite datasets} \label{sec:windows}
 
-In Section 2.1 we have defined a data stream as an unbounded data set.
+In Section 2.1 we have defined a data stream as an unbounded dataset.
 But what exactly are the issues of working with an infinite stream of data?
-Why might we need to convert it into a bounded data set and how to do so?
+Why might we need to convert it into a bounded dataset and how to do so?
 In this Section, we will analyze such issues and present sliding windows as the solution.
@@
 % boundless aggregations no need for windows
-Computations over an unbounded data set are possible.
-For example, computing the \textit{average} of an event's field for all incoming events is computationally trivial: we need only to store the total number of events (one integer) and the current total sum (one double) of all events arriving to the system.
+Computations over an unbounded dataset are possible.
+For example, computing the \textit{average} of an event's field for all incoming events is computationally trivial: we need only to store the total number of events and the current total sum of all events arriving to the system.
 However, is this boundless average over the stream of events useful?
@@
 To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}.
-By creating such a conceptual range, we realize that in practice we need a bounded data set.
-Partitioning an infinite data stream to obtain such a finite data set is made through the use of windows.
+By creating such a conceptual range, we realize that in practice we need a bounded dataset.
+Partitioning an infinite data stream to obtain such a finite dataset is made through the use of windows.
 
@@
 An aggregation --- also known as a fold, reduce or accumulate function --- is an operation that reduces a collection of values into a single one.
-Examples of such are the computation of data set sums, minima, averages or the number of distinct elements (cardinality).
+Examples of such are the computation of dataset sums, minima, averages or the number of distinct elements (cardinality).
 An aggregation may be exact or approximate.
@@
 Doing this recursively for each item in the collection results in a single value: the aggregate value.
-The combine function is here on forward represented as $\oplus$ and the identity or neutral element of the aggregation as $\overline{\theta}$.
+The combine function is from here onwards represented as $\oplus$ and the identity or neutral element of the aggregation as $\overline{\theta}$.
 
@@
 
-While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one.
-As we have seen in Section \ref{sec:windows}, we can obtain a bounded data set from a data stream by applying windowing techniques.
-When doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it.
+While some aggregations can be computed on an infinite dataset, often we want to compute these over a finite one.
+As we have seen in Section \ref{sec:windows}, we can obtain a bounded dataset from a data stream by applying windowing techniques.
+When doing so, we obtain a finite dataset denoted as a sliding window and compute aggregations over it.
 Tangwongsan \emph{et.

### 2020-07-01 chapters/introduction.tex
PDF images

@@
 In the context of this Thesis, we work side by side with Feedzai\footnote{https://feedzai.com}.
-Feedzai offers fraud prevention solutions for the previously presented use case of monitoring a continuous stream of credit card transactions and detecting fraud.
-We exploit real-life data sets that are provided from the financial fraud space.
-
-Feedzai's solution for fraud detection consists of a workflow of different components.
+In brief, the company's solution for fraud detection consists of a workflow of different components.
 Incoming events are fed into this workflow and processed accordingly.

### 2020-07-01 chapters/validation.tex
PDF images

@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/01-reference.png}
+  \includegraphics[width=1\linewidth]{figures/01-reference.pdf}
   \captionof{figure}{Time-series for dataset R1}
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/01-target.png}
+  \includegraphics[width=1\linewidth]{figures/01-target.pdf}
   \captionof{figure}{Time-series for dataset T1}
@@
 We see our JSD values before the change were very low (magnitudes of $10^{-2}$).
-\begin{center}
-\begin{minipage}{.5\textwidth}
+\begin{figure}[!htb]
 \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-625.png}
-  \captionof{figure}{JSD signal and threshold}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-625.pdf}
+  \caption{For the whole time-series}
   \label{fig:JSD-signal-01}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
   \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-625.png}
-  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-625.pdf}
+  \caption{Zoom in on the first 2.5 million events}
   \label{fig:JSD-signal-zoom-01}
-\end{minipage}
-\end{center}
+\end{subfigure}
+\caption{Experiment 01: JSD signal and threshold}
+\end{figure}
+
 However, the change in JSD value is very abrupt as well.
@@
 Figure \ref{fig:JSD-signal-zoom-02} shows the same signal for the first half of the streaming period.
-\begin{center}
-\begin{minipage}{.5\textwidth}
+\begin{figure}[!htb]
 \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-62500.png}
-  \captionof{figure}{JSD signal and threshold}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-62500.pdf}
+  \caption{For the whole time-series}
   \label{fig:JSD-signal-02}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
   \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-62500.png}
-  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-62500.pdf}
+  \caption{Zoom in on the first 2.5 million events}
   \label{fig:JSD-signal-zoom-02}
-\end{minipage}
-\end{center}
+\end{subfigure}
+\caption{Experiment 02: JSD signal and threshold}
+\end{figure}
 Notice that this time the growth of our signal is slower.
@@
 And indeed, judging by the corresponding signal computation and zoom-in, Figures \ref{fig:JSD-signal-03} and \ref{fig:JSD-signal-zoom-03}, respectively, the signal looks smoother and we see no false positives.
-\begin{center}
-\begin{minipage}{.5\textwidth}
+\begin{figure}[!htb]
 \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-250000.png}
-  \captionof{figure}{JSD signal and threshold}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-250000.pdf}
+  \caption{For the whole time-series}
   \label{fig:JSD-signal-03}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
   \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-250000.png}
-  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-250000.pdf}
+  \caption{Zoom in on the first 2.5 million events}
   \label{fig:JSD-signal-zoom-03}
-\end{minipage}
-\end{center}
+\end{subfigure}
+\caption{Experiment 03: JSD signal and threshold}
+\end{figure}
 
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-r2.png}
+  \includegraphics[width=1\linewidth]{figures/timeseries-r2.pdf}
   \captionof{figure}{Time-series for dataset R2}
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-t2.png}
+  \includegraphics[width=1\linewidth]{figures/timeseries-t2.pdf}
   \captionof{figure}{Time-series for dataset T2}
@@
 We observe that once again we have no false positives in the first 2 million events, which is ideal.
-\begin{center}
-\begin{minipage}{.5\textwidth}
+\begin{figure}[!htb]
 \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test02.png}
-  \captionof{figure}{JSD signal and threshold}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test02.pdf}
+  \caption{For the whole time-series}
   \label{fig:JSD-signal-test02}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
   \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test02.png}
-  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test02.pdf}
+  \caption{Zoom in on the first 2 million events}
   \label{fig:JSD-signal-zoom-test02}
-\end{minipage}
-\end{center}
+\end{subfigure}
+\caption{JSD signal and threshold}
+\end{figure}
 Furthermore, we observe that the computed JSD signal varies accordingly with the changes in distribution.
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-r3.png}
+  \includegraphics[width=1\linewidth]{figures/timeseries-r3.pdf}
   \captionof{figure}{Time-series for dataset R3}
@@
   \centering
-  \includegraphics[width=1.2\linewidth]{figures/timeseries-t3.png}
+  \includegraphics[width=1\linewidth]{figures/timeseries-t3.pdf}
   \captionof{figure}{Time-series for dataset T3}
@@
 Figure \ref{fig:JSD-signal-zoom-test03} shows a zoomed in portion of the original signal, where we can verify that there are no false positives.
-\begin{center}
-\begin{minipage}{.5\textwidth}
+\begin{figure}[!htb]
 \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test03.png}
-  \captionof{figure}{JSD signal and threshold}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test03.pdf}
+  \caption{For the whole time-series}
   \label{fig:JSD-signal-test03}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
   \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test03.png}
-  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test03.pdf}
+  \caption{Zoom in on the first 2 million events}
   \label{fig:JSD-signal-zoom-test03}
-\end{minipage}
-\end{center}
+\end{subfigure}
+\caption{JSD signal and threshold}
+\end{figure}
 
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-r4.png}
+  \includegraphics[width=1\linewidth]{figures/timeseries-r4.pdf}
   \captionof{figure}{Time-series for dataset R4}
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-t4.png}
+  \includegraphics[width=1\linewidth]{figures/timeseries-t4.pdf}
   \captionof{figure}{Time-series for dataset T4}
@@
 
-\begin{center}
-\begin{minipage}{.5\textwidth}
+\begin{figure}[!htb]
 \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test04.png}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test04.pdf}
   \captionof{figure}{JSD signal and threshold}
   \label{fig:JSD-signal-test04}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
   \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test04.png}
-  \captionof{figure}{Zoomed in JSD signal and threshold}
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test04.pdf}
+  \caption{Zoom in on the first 2 million events}
   \label{fig:JSD-signal-zoom-test04}
-\end{minipage}
-\end{center}
+\end{subfigure}
+\caption{JSD signal and threshold}
+\end{figure}
 
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-r5.png}
+  \includegraphics[width=1\linewidth]{figures/timeseries-r5.pdf}
   \captionof{figure}{Time-series for dataset R5}
@@
   \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-t5.png}
+  \includegraphics[width=1\linewidth]{figures/timeseries-t5.pdf}
   \captionof{figure}{Time-series for dataset T5}
@@
 Figure \ref{fig:JSD-signal-zoom-test05} is a zoom in of the first 2 millionth events and once again we have no false positives.
-\begin{center}
-\begin{minipage}{.5\textwidth}
+\begin{figure}[!htb]
 \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test05.png}
-  \captionof{figure}{Distance (JSD) signal and threshold}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test05.pdf}
+  \captionof{figure}{JSD signal and threshold}
   \label{fig:JSD-signal-test05}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
   \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test05.png}
-  \captionof{figure}{Zoomed in distance (JSD) signal and threshold}
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test05.pdf}
+  \caption{Zoom in on the first 2 million events}
   \label{fig:JSD-signal-zoom-test05}
-\end{minipage}
-\end{center}
+\end{subfigure}
+\caption{JSD signal and threshold}
+\end{figure}
 
@@
     \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x1.png} 
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x1.pdf} 
     \caption{Feature x1 reference time-series} 
@@
     \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x2.png} 
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x2.pdf} 
     \caption{Feature x2 reference time-series} 
@@
     \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x3.png} 
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x3.pdf} 
     \caption{Feature x3 reference time-series} 
@@
     \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x4.png} 
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x4.pdf} 
     \caption{Feature x4 reference time-series} 
@@
     \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x1.png} 
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x1.pdf} 
     \caption{Feature x1 target time-series} 
@@
     \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x2.png} 
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x2.pdf} 
     \caption{Feature x2 target time-series} 
@@
     \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x3.png} 
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x3.pdf} 
     \caption{Feature x3 target time-series} 
@@
     \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x4.png} 
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x4.pdf} 
     \caption{Feature x4 target time-series} 
@@
 \clearpage
-Similarly to x1, features x2 and x3 have very similar reference and target periods (not between each other).
+Relative to x1, features x2 and x3 also have very similar reference and target periods.
 Just like feature x1, we expect to have some alarms but still not too many.

### 2020-07-01 chapters/validation.tex
More PDF images and new plots for last validation section
Last experiment written

@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x1-reference.png}
+      \includegraphics[scale=0.5]{figures/merchant-x1-reference.pdf}
       \caption{Merchant feature x1 reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x1-target.png}
+      \includegraphics[scale=0.5]{figures/merchant-x1-target.pdf}
       \caption{Merchant feature x1 target time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x1-correctedpvalues.png}
+      \includegraphics[scale=0.5]{figures/merchant-x1-correctedpvalues.pdf}
       \caption{Merchant feature x1 corrected p-values}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x2-reference.png}
+      \includegraphics[scale=0.5]{figures/merchant-x2-reference.pdf}
       \caption{Merchant feature x2 reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x2-target.png}
+      \includegraphics[scale=0.5]{figures/merchant-x2-target.pdf}
       \caption{Merchant feature x2 target time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x3-reference.png}
+      \includegraphics[scale=0.5]{figures/merchant-x3-reference.pdf}
       \caption{Merchant feature x3 reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x3-target.png}
+      \includegraphics[scale=0.5]{figures/merchant-x3-target.pdf}
       \caption{Merchant feature x3 target time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x2-correctedpvalues.png}
+      \includegraphics[scale=0.5]{figures/merchant-x2-correctedpvalues.pdf}
       \caption{Merchant feature x2 corrected p-values}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x3-correctedpvalues.png}
+      \includegraphics[scale=0.5]{figures/merchant-x3-correctedpvalues.pdf}
       \caption{Merchant feature x3 corrected p-values}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x4-reference.png}
+      \includegraphics[scale=0.5]{figures/merchant-x4-reference.pdf}
       \caption{Merchant feature x4 reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x4-target.png}
+      \includegraphics[scale=0.5]{figures/merchant-x4-target.pdf}
       \caption{Merchant feature x4 target time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x4-correctedpvalues.png}
+      \includegraphics[scale=0.5]{figures/merchant-x4-correctedpvalues.pdf}
       \caption{Merchant feature x4 corrected p-values}
@@
 
-The -1 value is often used as a placeholder for missing values, \textit{i.e.}, if some problem occurs and the value for a feature is missing at a given timestamp or event, -1 is used.
+The -1 value is often used as a placeholder for missing values, \textit{i.e.} if some problem occurs and the value for a feature is missing at a given timestamp or event, -1 is used.
 A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
@@
     %x4 - completion_rate_per_card_72h
+    %x5 - dif_last_and_current_trx
 \subsection{Same reference and target periods}
@@
 We chose to use the reference dataset from the previous experiment (Reference of Table \ref{tbl:merchant1-datasets-summary}) and feed it as input for both our batch and stream analysis as the reference and target periods, respectively.
+We expect to have less alarms for each feature but not necessarily zero.
 
-\textcolor{red}{TODO: visualizations and conclusions of this experiment}
+We first show the corrected p-value plots when both reference and target periods are the same, for the previous features x1, x2 and x3 in Figures \ref{fig:merchant2-x1-correctedpvalues}, \ref{fig:merchant2-x2-correctedpvalues} and \ref{fig:merchant2-x3-correctedpvalues}, respectively.
+
+Feature x1 has less alerts in this experiment where we use equal reference and target periods.
+This is expected and clear when comparing Figure \ref{fig:merchant2-x1-correctedpvalues} and \ref{fig:merchant-x1-correctedpvalues}.
+When comparing Figures \ref{fig:merchant-x2-correctedpvalues} and \ref{fig:merchant-x3-correctedpvalues} with \ref{fig:merchant2-x2-correctedpvalues} and \ref{fig:merchant2-x3-correctedpvalues} for feature x2 and x3, respectively, we observe that the number of alerts stayed more or less the same.
+We think its because x2 and x3 reference and target periods were already pretty similar, so ensuring both periods were the same did not make much difference.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x1-correctedpvalues.pdf}
+      \caption{Merchant feature x1 corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x1-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x2-correctedpvalues.pdf}
+      \caption{Merchant feature x2 corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x2-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x3-correctedpvalues.pdf}
+      \caption{Merchant feature x3 corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x3-correctedpvalues}
+    \end{center}
+\end{figure}
+
+Now we analyze feature x4 from the previous experiment, as this one was the one with the highest measured drift.
+Recall that feature x4' target time-series (Figure \ref{fig:merchant-x4-target}) was significantly different from the reference one (Figure \ref{fig:merchant-x4-reference}).
+Because of this, the corrected p-values plot from the previous experiment (Figure \ref{fig:merchant-x4-correctedpvalues}) crossed our threshold line multiple times and remained in an alert state for most of the target time-series.
+In this new experiment, we used the reference period twice.
+Because of this, the corrected p-values plot is now different.
+Figure \ref{fig:merchant2-x4-correctedpvalues} shows the corrected p-values plot for feature x4 when we used the reference period for both phases.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x4-correctedpvalues.pdf}
+      \caption{Merchant feature x4 corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x4-correctedpvalues}
+    \end{center}
+\end{figure}
+Using the same period twice theoretically means we have no drift at all.
+Hence, we expected to see no alerts.
+Figure \ref{fig:merchant2-x4-correctedpvalues} shows that we have a few alerts but significantly less than in Figure \ref{fig:merchant-x4-correctedpvalues}.
+Furthermore, in this new experiment, x4 eventually stops being reported at all until the end of the dataset and ends at a high p-value.
+Recall that the higher the p-value the less divergent a feature is.
+Overall we see that this feature has fewer alarms and stabilizes well above our alarm threshold which is coherent since we used the same dataset as a reference and target periods.
+
+We now introduce a new feature, x5.
+Feature x5 also suffers from a large amount of drift.
+In the reference and target datasets presented in Table \ref{tbl:merchant1-datasets-summary}, feature x5 reference time-series plot is given by Figure \ref{fig:merchant-x5-reference} and the target time-series by Figure \ref{fig:merchant-x5-target}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x5-reference.pdf}
+      \caption{Merchant feature x5 reference time-series}
+      \label{fig:merchant-x5-reference}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x5-target.pdf}
+      \caption{Merchant feature x5 target time-series}
+      \label{fig:merchant-x5-target}
+    \end{center}
+\end{figure}
+The density in the upper domain of values is clearly different in both reference and target periods.
+Hence, in our first experiment, this feature had a corrected p-values plot with two alarm states, where the second one lasted until the end of the target period, with the lowest possible p-value (zero), represented in Figure \ref{fig:merchant-x5-correctedpvalues}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x5-correctedpvalues.pdf}
+      \caption{Merchant feature x5 corrected p-values}
+      \label{fig:merchant-x5-correctedpvalues}
+    \end{center}
+\end{figure}
+
+However, when using the same reference and target periods, in this case using the time-series in Figure \ref{fig:merchant-x5-reference} twice, we achieve our lowest possible drift scenario.
+The difference in corrected p-values is clear.
+Figure \ref{fig:merchant2-x5-correctedpvalues} shows this plot when using both reference and target periods.
+We observe that our corrected p-values never go below the threshold line, in fact, they stay well above it.
+This is expected behavior when using equal reference and target periods.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x5-correctedpvalues.pdf}
+      \caption{Merchant feature x5 corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x5-correctedpvalues}
+    \end{center}
+\end{figure}

### 2020-07-02 chapters/background.tex
Restivo feedback applied

@@
 \chapter{Background}\label{chap:background} \minitoc
+
+In this chapter we discuss fundamental concepts for our working context.
+In Section \ref{sec:stream-superset} we introduce batch and stream processing as two large scale data processing techniques.
+In Section \ref{sec:windows} we review windowing mechanisms that essentially turn unbounded data streams to finite datasets.
+In Section \ref{sec:aggregations} we introduce aggregations as a whole.
+Then, we specifically discuss sliding window aggregations, algorithms used for its computation and aggregation properties.
+In Section \ref{sec:emas} we present Exponential Moving Averages as aggregators that replace the sliding window contents for a lighter representation, providing an approximate estimation of the aggregation value.
 
@@
 An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes).
-Batch processing is the oldest paradigm between batch and stream processing.
-Historically speaking, Apache Hadoop \cite{borthakur2007hadoop} \cite{Hadoop} \cite{ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005.
+Between batch and stream processing, batch processing is the older paradigm.
+Historically, Apache Hadoop \cite{borthakur2007hadoop, Hadoop, ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005.
 Hadoop uses a programming model called MapReduce \cite{MapReduce}.
@@
 To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms).
-Nearly a decade later, Apache Spark \cite{ApacheSpark} \cite{Spark} was released as the industry accepted successor.
+Nearly a decade later, Apache Spark \cite{ApacheSpark, Spark} was released as the industry accepted successor.
 Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}.
@@
 
-Stream procesisng, on the other hand, refers to the processing of data streams.
+Stream processing, on the other hand, refers to the processing of data streams.
 A data stream is essentially a continuous or unbounded dataset.
@@
 
-The Venn diagram \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
+The Venn diagram in Figure \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
 A data stream is unbounded or infinite while a bounded data stream has a beginning and an end.
@@
 
-In Section 2.2 we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.
+In Section \ref{sec:windows} we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.
 
@@
 
-In Section 2.1 we have defined a data stream as an unbounded dataset.
+In Section \ref{sec:stream-superset} we have defined a data stream as an unbounded dataset.
 But what exactly are the issues of working with an infinite stream of data?
 Why might we need to convert it into a bounded dataset and how to do so?
-In this Section, we will analyze such issues and present sliding windows as the solution.
+In this section, we will analyze such issues and present sliding windows as the solution.
 
@@
 
-\subsection{Sliding Windows}
-
-Botan et. al \cite{Botan-SECRET} formally define a window as:
-
+Botan et. al \cite{Botan-SECRET} define a window as:
 \begin{definition}
-A window W over a stream S is a finite subset of S"
+A window W over a stream $\mathbb{S}$ is a finite subset of $\mathbb{S}$
 \end{definition}
-
-% 2 dimensions of windows
 Windows can be categorized by two dimensions: the window size $W$ and the sliding step size $S$.
@@
 A window slides over incoming data and groups it, enabling us to compute aggregations over a defined and bounded set of events.
-The step size \textit{S} of a windowing technique is formally defined as the distance between consecutive windows \cite{Botan-SECRET}.
+The step size \textit{S} of a windowing technique is defined as the distance between consecutive windows \cite{Botan-SECRET}.
 This means that two consecutive windows W\textsubscript{1} and W\textsubscript{2} are exactly \textit{S} units away.
@@
 
-\subsection{Sliding Window Aggregations}
+\subsection{Sliding window aggregations}
 
@@
 
-\subsection{Sliding Window Aggregation Algorithms} \label{sec:back-swag-algs}
+\subsection{Sliding window aggregation algorithms} \label{sec:back-swag-algs}
 
@@
 
-Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011} \cite{EMA-Hunter} \cite{EMA-MarcusB} are a type of reduce functions applied to a sequence of elements, one at a time in a orderly fashion, to compute a weighted average, using different weights for each element.
+Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB} are a type of reduce functions applied to a sequence of elements, one at a time in a orderly fashion, to compute a weighted average, using different weights for each element.
 These weights decrease exponentially with the age of the observations.
@@
 \end{definition}
-For example, with a tuple-based half-life of $\tau = 1$ tuple, for each new tuple all weights are reduced by half.
+For example, with a tuple-based half-life of $n_{1/2}=1$ tuple, for each new tuple all weights are reduced by half.
 
 Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones.
-In this Thesis we chose to work with tuple-based EMAs due to its simplicity but our approach can be adjusted to use time-based ones.
+In this Thesis we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.

### 2020-07-02 chapters/introduction.tex
Restivo feedback applied

@@
 All of these applications generate large volumes of data over time which poses data storage and processing challenges.
-This data is characterized in terms of its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems.
+This data is characterized by its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems.
 Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams.
@@
 However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques.
-A thorough analysis of stream windowing techniques is done in the Background Section of this document but for now it suffices to say that windowing transforms an unbounded data stream into a finite data set typically by defining a time or tuple-based window size --- \textit{i.e.} events from the last three days or the last million events, respectively.
+A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size --- \textit{e.g.}, events from the last three days or the last million events, respectively.
 
@@
 In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one.
-Such a system would have to process all incoming transactions and decide in a fraction of a second, for each transaction, whether it is fraudulent or not.
+Such a system would have to process all incoming transactions and decide, in a fraction of a second, if a transaction is fraudulent or not.
 It is clear that manually monitoring such an intensive, latency-sensitive data stream is hard and error-prone, thus justifying the need for autonomous real-time data stream monitoring systems.
@@
 Thus their model performance depends on how well those assumptions hold for future data.
-However, fraud patterns are not static over time, but seasonal --- \textit{e.g.} Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.} new fraud attack methods by fraudsters.
+However, fraud patterns are not static over time, but seasonal --- \textit{e.g.}, Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.}, new fraud attack methods by fraudsters.
 Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time.
-In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift \cite{drift} \cite{gama-drift}.
+In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift \cite{drift, gama-drift}.
 
@@
 \section{Hypothesis} \label{sec:hypothesis}
-The question this Thesis aims to answer is whether sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.
-In this Thesis, we study some sliding window aggregations that do not keep the sliding window in memory but still give an estimate on the aggregation value.
+The question this Thesis aims to answer is whether \textit{sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.} In this Thesis, we study sliding window aggregations that replace the sliding window events in memory by a lighter representation, while still being able to provide an estimate of the aggregation value.
 
@@
 
-The hypothesis will be proven true or false depending on whether the following criteria are met. The resulting data pattern shift reporting system must have the following properties:
+The resulting data pattern shift reporting system must have the following properties:
 \begin{itemize}
@@
 \end{itemize}
-
-Experiments will be made using multiple synthetic data sets and real ones from the financial fraud space.
+The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic data sets and a real dataset from the financial fraud space.
 

### 2020-07-02 chapters/problem-statement.tex
Restivo feedback applied

@@
 
-With this Chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.
+With this chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.
 
@@
 Many real-time stream monitoring systems are static once deployed in a production environment.
-Engineers of said systems configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data.
+The engineers of those systems configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data.
 They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system.
 Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.
-The problem at hand is to determine when to reconfigure the system, \textit{e.g.} a Machine Learning model, based on streaming analysis.
+The problem at hand is to determine when to reconfigure the system, \textit{e.g.} a Machine Learning model, based on an analysis of the drifts in the stream of data.
 
@@
 
-    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout run-time
+    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout runtime
 
@@
 \section{Proposal}
-We devise a two-phased (batch and streaming phases) and two-windowed (reference and target sliding windows) method.
-We devise a sliding window approximate histogram aggregation that encodes data distributions and provide a method for incrementally maintaining it with constant time and space complexity for every growing volumes of data, ideal for the streaming scenario.
-Given a user-defined reference period and a stream of data to process, we monitor each event's fields and alert if they are considered divergent relative to their reference distribution.
-We provide a measure of how divergent each field (or feature) is.
+%2 phases, names and purpose
+We devise a two-phased and two-windowed method: the first phase is a batch analysis that makes use of a reference window; and the second phase is a streaming analysis that makes use of a target sliding window.
+The batch analysis main goal is to compute the reference aggregation for the reference window.
+The stream analysis main goals are to incrementally maintain the same type of aggregation but over the streaming data and alert when the difference between the reference aggregation and the target sliding one is high.
+
+%aggregations to represent distributions
+We choose to use an histogram aggregation with constant memory usage to encode the distribution of the observed values.
+Furthermore, we devise a sliding window approximate histogram aggregation that is maintainable in constant time in a sliding window fashion, through the use of Exponential Moving Averages (EMAs).
+Hence, our EMA-like histogram aggregation is ideal for the streaming scenario because of its constant time and space complexity, for every growing volumes of data.
+
+Given a user-defined reference period and a stream of data to process, we monitor each event's fields and alert if they are considered divergent relative to their reference aggregation.
+We provide a measure of how divergent each field is.
 Based on this information, it is our belief that a system administrator would have an easier time determining when to reconfigure the system.
@@
 
-\section{Research Questions}
+\section{Research questions}
 The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time.
 To that end, we devise a series of research questions to be answered by the end of this Thesis:
+
+\textcolor{red}{TODO: further describe each RQ?}
 

### 2020-07-02 chapters/sota.tex
Restivo feedback applied

@@
 
-A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams} \cite{Domingos-Mining-Time-Data-Streams}.
-In this Chapter, we discuss the most relevant work for the problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems, presented in Chapter \ref{chap:intro}.
-We present a wide analysis of several categories of algorithms that seem viable to build the desired system.
-
-%Probably say we analyze outlier methods and why aggregations? relate to the approaches thought to solve the problem and the sections
+A data stream has a temporal dimension and the underlying process that generates data can change over time \cite{Aggarwal-Evolving-Data-Streams, Domingos-Mining-Time-Data-Streams}.
+In this chapter, we discuss the most relevant work for the problem of detecting data pattern shifts in real-time over true sliding windows using resource-lightweight approaches on streaming systems, presented in Chapter \ref{chap:intro}.
+We present a wide analysis of several categories of algorithms to be able to identify what's the region of this vast landscape of methods that may be useful to build the desired system.
 
@@
 Blázquez-García \emph{et al.} review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
-From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.
+From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey, Aguinis-Outlier-survey, Chandola-Outlier-survey-2009, Hodge-Outlier-survey, Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.
 
@@
 Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy} \cite{Blazquez-Garcia-Review-Anomaly-Detection}) .
-
 \begin{figure}[!htb]
@@
 Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting.
-For example, nearest neighbor based methods consider a datum an outlier if it has less than $\alpha$ neighbors.
+For example, nearest neighbor based methods consider a datum an outlier if it has less than $k$ neighbors.
 Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}.
@@
 
-\subsubsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data: the Extra-N algorithm} \label{alg:extran}
-
-Yang et.
-al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios.
-The authors propose \emph{Extra-N}, an algorithm that detects point outliers based on the number of neighbors it has.
-The authors make use of the \emph{predictability} property of the expiration of existing objects.
-In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to.
-
-The authors of Extra-N want to maintain for each data point $p\textsubscript{i}$ the number of neighbors it has, rather than the actual list of neighbors.
-The main advantage of this approach is the reduced memory footprint since only a count is stored rather than a list of pointers to all neighbors.
-The problem is that when a data point \textit{p\textsubscript{i}} expires, the count of each of \textit{p\textsubscript{i}}'s neighbors must be decremented by one but we are unable to reach \textit{p\textsubscript{i}}'s neighbors without a list of neighbor pointers.
-The solution to this problem is exploiting the previously presented \textit{predictability} property.
-The key idea is to predict the expiration of any data point \textit{p\textsubscript{i}} and pre-handle the impact it has on \textit{p\textsubscript{i}}'s neighbors.
-The authors introduce the concept of \textit{lifetime neighbor counts} or \textit{lt\_cnt}.
-The \textit{lt\_cnt} of a data point \textit{p\textsubscript{i}} corresponds to a sequence of \textit{predicted neighbor counts} --- \textit{i.e.} the number of \textit{predicted neighbors} \textit{p\textsubscript{i}} has in any future window where it exists.
-More precisely, each entry on \textit{pi.lt\_cnt} records the number of \textit{p\textsubscript{i}}'s neighbors that are known to survive in the corresponding future window.
-Consider a window \textit{W\textsubscript{i}} and a data point \textit{p\textsubscript{i}} with three neighbors: \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}.
-Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{i}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{i+1}} and \textit{p\textsubscript{i}} expires after \textit{W\textsubscript{i+2}}.
-This implies that at \textit{W\textsubscript{i}},
-\[ p\textsubscript{i}.lt\_cnt = \{W\textsubscript{i}: 3, W\textsubscript{i+1}: 2, W\textsubscript{i+2}: 0\} \]
-The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point \textit{p\textsubscript{i}} check if the first or current element of \textit{p\textsubscript{i}.lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether a data point is an outlier or not.
+\subsubsection{Neighbor-Based Pattern Detection for Windows Over Streaming Data} \label{alg:extran}
+
+Yang et al.
+propose \emph{Extra-N} \cite{Yang-Neighbor-Based-Pattern-Detection}, a method for incremental detection of point outliers in sliding window scenarios.
+Data points are considered outliers if they have few neighbors, \textit{i.e.}, less than a user-defined threshold $\alpha$ of neighbors within a user-defined range $\theta$.
+Figure \ref{fig:point-outlier} depicts points plotted in their feature space (in this case only two dimensions or features are used, x1 and x2) and a point outlier for $alpha=2$ points and any given $\theta$.
+\begin{figure}[!htb]
+  \begin{center}
+    \includegraphics[scale=0.6]{figures/point-outlier.pdf}
+    \caption{Point outlier for $\alpha=2$ and a given $\theta$}
+    \label{fig:point-outlier}
+  \end{center}
+\end{figure}
+
+The key idea is to predict the expiration of any data point in the sliding window and pre-handle the impact it has on its neighbors.
+The authors introduce the concept of \emph{lifetime neighbor counts} or \textit{lt\_cnt}.
+The \textit{lt\_cnt} of a data point corresponds to a sequence of predicted neighbor counts, \textit{i.e.}, the number of predicted neighbors a data point has in any future window where it exists.
+More precisely, each entry on \textit{lt\_cnt} records the number of neighbors that are known to survive in future windows.
+
+For example, consider a window \textit{W\textsubscript{0}} and a data point \textit{p\textsubscript{0}} with three neighbors: \textit{p\textsubscript{1}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}}.
+Assume \textit{p\textsubscript{1}} expires after \textit{W\textsubscript{0}}, \textit{p\textsubscript{2}} and \textit{p\textsubscript{3}} expire after \textit{W\textsubscript{1}} and \textit{p\textsubscript{0}} expires after \textit{W\textsubscript{2}}.
+This implies that at \textit{W\textsubscript{0}},
+\[ p\textsubscript{0}.lt\_cnt = \{W\textsubscript{0}: 3, W\textsubscript{1}: 2, W\textsubscript{2}: 0\} \]
+The lifetime neighbor counts (\textit{lt\_cnt}) carry enough information to compute density-based point outliers: for each data point check if the first or current element of \textit{lt\_cnt} is less than a user-defined threshold $\alpha$ to decide whether that data point is an outlier or not.
 
@@
 Extra-N has to keep a lifetime neighbor count dictionary-like object for each window data point.
-For that reason, we claim that Extra-N achieves linear memory consumption relative to the number of data points in the window.
+For that reason, Extra-N achieves linear memory consumption relative to the number of data points in the window.
 
@@
 Furthermore, updating the sliding window should be done in constant time, which is not the case for Extra-N.
-Hence, this algorithm will be of no use to us.
+Hence, we will not use this algorithm.
 
@@
 In other words, if there is a sudden abrupt change in the distribution of a specific time-dependent variable, the algorithm will raise an alert.
-However, if that initially thought to be random spike becomes frequent, the algorithm adapts and assimilates this as "a new reality" and stops classifying it as an anomaly.
+However, if an initial spike that was assumed to be an anomaly becomes frequent, then the algorithm adapts and assimilates this as "a new reality" and stops classifying it as anomalous.
 Unfortunately, this property is not desirable in our system.
@@
 The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN.
-Additionally, the data set used to train the model thus not require labels.
+Additionally, the data set used to train the model does not require labels.
 
@@
 As mentioned, our goal is to build a monitoring system that alerts data pattern shifts continuously, until they match the initial configuration again.
-Furthermore, DeepAnT uses deep learning which is actually a subset of machine learning and just as inexplicable.
-In summary, we do not deem DeepAnT usable for our use case.
 
@@
 
-\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
+\section{Sliding window aggregation algorithms} \label{sec:sota-swag-algs}
 
@@
 
-\section{Probabilistic Data Structures} \label{sec:pds}
+\section{Probabilistic data structures} \label{sec:pds}
 In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs).
@@
 
-\subsection{Membership Queries and the Bloom filter} \label{sec:bloom}
+\subsection{Membership queries and the Bloom Filter} \label{sec:bloom}
 % approx aggregation to compute
-A Bloom filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements.
+A Bloom Filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements.
 Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?".
@@
     \begin{center}
-      \includegraphics[scale=1.1]{figures/initial-bloom.png}
+      \includegraphics[scale=0.9]{figures/initial-bloom.png}
       \caption[Bloom filter initial state]{Bloom filter bit array initial state of \textit{m}=7}
@@
     \begin{center}
-      \includegraphics[scale=0.4]{figures/insert-bloom.png}
+      \includegraphics[scale=0.3]{figures/insert-bloom.png}
       \caption[Bloom filter insertion]{Insertion of element \textit{el} into a Bloom filter with \textit{m}=7 and \textit{k}=3}
@@
     \begin{center}
-      \includegraphics[scale=0.4]{figures/query-bloom.png}
+      \includegraphics[scale=0.3]{figures/query-bloom.png}
       \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
@@
 
-\subsection{Item Frequency and the Count-Min Sketch}
+\subsection{Item frequency and the Count-Min Sketch}
 % approx aggregation to compute
@@
 
-\subsection{Cardinality Estimation and the HyperLogLog} \label{sec:hyper-log-log}
+\subsection{Cardinality estimation and the HyperLogLog} \label{sec:hyper-log-log}
 
@@
 
-\section{Sliding Window Aggregations with Probabilistic Data Structures} \label{sec:sliding-pds}
+\section{Sliding window aggregations with probabilistic data structures} \label{sec:sliding-pds}
 In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space.
@@
 \textcolor{red}{TODO Find and add versions of Sliding Bloom Filter and Sliding CMS}
+
+\section{Summary}

### 2020-07-02 chapters/validation.tex
Restivo feedback applied

@@
 
-In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
-As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
+In this section, we provide summary statistics for our method's performance and draw visualizations for the obtained results to validate our alarms and reason about them.
 
 \section{Experiments with synthetic data}
-\subsection{Single Feature Analysis}
+In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
+In this section we present summary statistics for the synthetic datasets created and our generated alarms.
+
+\subsection{Single feature analysis}
+As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
+In this section we validate our results for single-feature synthetic datasets.
 
@@
 Let's take a look at the signal computed for the target period.
-Figure \ref{fig:JSD-signal-test05} shows us the JSD signal for dataset T5 \ref{fig:timeseries-t5}.
+Figure \ref{fig:JSD-signal-test05} shows us the JSD signal for dataset T5 (Figure \ref{fig:timeseries-t5}).
 As expected, we see a rapid growth of JSD value after the 2 millionth event, when we change the Gaussian parameters.
@@
 
-\subsection{Multi Feature Analysis}
+\subsection{Multi feature analysis}
 Before moving to real data, we performed one last experiment, using synthetic data but this time performing a multi-feature analysis.
-Table \ref{tbl:multi-feat-ref-dataset-distros} shows the underlying generating distributions and parameters for each of the four features (x1, x2, x3 and x4) in the reference multi-feature dataset R6.
-Figures \ref{fig:timeseries-r6-x1}, \ref{fig:timeseries-r6-x2}, \ref{fig:timeseries-r6-x3} and \ref{fig:timeseries-r6-x4} show the reference time-series for features x1, x2, x3 and x4 on the reference dataset R6, respectively.
+Table \ref{tbl:multi-feat-ref-dataset-distros} shows the underlying generating distributions and parameters for each of the four features ($x_1$, $x_2$, $x_3$ and $x_4$) in the reference multi-feature dataset R6.
+Figures \ref{fig:timeseries-r6-x1}, \ref{fig:timeseries-r6-x2}, \ref{fig:timeseries-r6-x3} and \ref{fig:timeseries-r6-x4} show the reference time-series for features $x_1$, $x_2$, $x_3$ and $x_4$ on the reference dataset R6, respectively.
 
@@
 
-We used a target dataset T6 with 5 million events and the same four features (x1, x2, x3 and x4).
+We used a target dataset T6 with 5 million events and the same four features ($x_1$, $x_2$, $x_3$ and $x_4$).
 Feature x1 followed the reference distribution for the 5 million events of the target dataset.
 We did not change this feature underlying distribution to serve as a control, \textit{i.e.}, we expect no alerts for this one.
-Figure \ref{fig:timeseries-t6-x1} shows the time-series for the target dataset T6 and feature x1.
+Figure \ref{fig:timeseries-t6-x1} shows the time-series for the target dataset T6 and feature $x_1$.
 
 Feature x2's underlying distribution changed four times and returned to normal, so we expect to see alarms with increasing divergence values but eventually no more alarms.
-Figure \ref{fig:timeseries-t6-x2} shows the time-series for the target dataset T6 and feature x2.
-Feature x2 followed a gaussian with $\mu=100$ and $\sigma=5$ like in the reference period up to the 1 millionth event mark, then a gaussian with $\mu=100$ and $\sigma=10$ up to the 2.5 millionth event, then a gaussian $\mu=100$ and $\sigma=15$ until the 3 millionth event, then a gaussian $\mu=100$ and $\sigma=20$ until the 4 millionth event and finally returned to the original gaussian with $\mu=100$ and $\sigma=5$ until the last event.
+Figure \ref{fig:timeseries-t6-x2} shows the time-series for the target dataset T6 and feature $x_2$.
+Feature $x_2$ followed a gaussian with $\mu=100$ and $\sigma=5$ like in the reference period up to the 1 millionth event mark, then a gaussian with $\mu=100$ and $\sigma=10$ up to the 2.5 millionth event, then a gaussian $\mu=100$ and $\sigma=15$ until the 3 millionth event, then a gaussian $\mu=100$ and $\sigma=20$ until the 4 millionth event and finally returned to the original gaussian with $\mu=100$ and $\sigma=5$ until the last event.
 Table \ref{tbl:multi-feat-x2-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature x2.
@@
 
-Due to time constraints, we experimented with data from only one merchant. The merchant's entity as well as feature names will remain anonymous due to legal reasons.
+Due to time constraints, we experimented with data from only one merchant. The merchant's entity as well as feature names will remain anonymous due to confidentiality reasons.
 
-Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that combined covered a period of roughly 1 year and 9 months.
+Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that, combined, covered a period of roughly 1 year and 9 months.
 In this experiment, we used the older one as the reference dataset and the more recent one as the target period.
-Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus categorical features.
+Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus numerically encoded categorical features (\textit{e.g.} one-hot encoded or ordinal encoded \cite{categoricalencoding}).
 Table \ref{tbl:merchant1-datasets-summary} shows summary statistics for both datasets.
@@
 We determined that one week's worth of data for this merchant was on average equivalent to 37,803 events.
-We set the half-life to be $n_{1/2}=37,803$ and discard events whose contribution is 6\% or less, which is equivalent to using sampling windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).
+We set the half-life to be $n_{1/2}=37,803$ and discard events whose EMA weight is 6\% or less, which is equivalent to using sampling windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).
 Hence, our samples' windows had 151,212 events.
@@
 Due to a large number of features (183), we only show some plots.
-Furthermore, we refer to features with anonymized identifiers, like feature $x_t$.
+Furthermore, we refer to features with anonymized identifiers, like feature $x_j$.
 In the following plots, we show the corrected p-values (Formula \ref{for:corrected-pvalue}) for some features at each event of the time-series and the FWER threshold set at 1\%.
@@
 
-Figure \ref{fig:merchant-x1-reference} is a plot of the reference time-series of feature x1 and Figure \ref{fig:merchant-x1-target} the corresponding target time-series.
+Figure \ref{fig:merchant-x1-reference} is a plot of the reference time-series of feature $x_1$ and Figure \ref{fig:merchant-x1-target} the corresponding target time-series.
 \begin{figure}[!htb]
@@
 Furthermore, the alarms appear to be roughly in the middle portion of the time-series, where in fact the target period appears to have less points.
+We can see roughly three alarms.
+The first one in early January 2019 for a few days, the second one lasting about a week at the end of January 2019 and the last one in the second half of February 2019 again for a few days.
 
-Now let us look at features x2 and x3. Figures \ref{fig:merchant-x2-reference} and \ref{fig:merchant-x2-target} show feature x2' time-series plots for the reference and target periods, respectively.
+Now let us look at features $x_2$ and $x_3$.
+Figures \ref{fig:merchant-x2-reference} and \ref{fig:merchant-x2-target} show feature $x_2$' time-series plots for the reference and target periods, respectively.
 \begin{figure}[!htb]
@@
 \end{figure}
-Figures \ref{fig:merchant-x3-reference} and \ref{fig:merchant-x3-target} show feature x3' time-series plots for the reference and target periods, respectively.
+Figures \ref{fig:merchant-x3-reference} and \ref{fig:merchant-x3-target} show feature $x_3$' time-series plots for the reference and target periods, respectively.
 \begin{figure}[!htb]
@@
 \clearpage
-Relative to x1, features x2 and x3 also have very similar reference and target periods.
+Relative to $x_1$, features $x_2$ and $x_3$ also have very similar reference and target periods.
 Just like feature x1, we expect to have some alarms but still not too many.
@@
 \end{figure}
-The corrected p-values plot shows us once again that both x2 and x3 features had some alarms but quickly rose above the threshold and resumed their reference distribution.
+The corrected p-values plot shows us once again that both $x_2$ and $x_3$ features had some alarms but eventually rose above the threshold and resumed their reference distribution.
 This is expected from features whose reference and target time-series plots look so similar.
@@
 
-We first show the corrected p-value plots when both reference and target periods are the same, for the previous features x1, x2 and x3 in Figures \ref{fig:merchant2-x1-correctedpvalues}, \ref{fig:merchant2-x2-correctedpvalues} and \ref{fig:merchant2-x3-correctedpvalues}, respectively.
+We first show the corrected p-value plots when both reference and target periods are the same, for the previous features $x_1$, $x_2$ and $x_3$ in Figures \ref{fig:merchant2-x1-correctedpvalues}, \ref{fig:merchant2-x2-correctedpvalues} and \ref{fig:merchant2-x3-correctedpvalues}, respectively.
 

### 2020-07-02 chapters/problem-statement.tex
mais cenas

@@
 Furthermore, we devise a sliding window approximate histogram aggregation that is maintainable in constant time in a sliding window fashion, through the use of Exponential Moving Averages (EMAs).
-Hence, our EMA-like histogram aggregation is ideal for the streaming scenario because of its constant time and space complexity, for every growing volumes of data.
+Hence, our EMA-like histogram aggregation is ideal for the streaming scenario because of its constant time and space complexity, for ever growing volumes of data.
 

### 2020-07-02 chapters/method.tex
Small corrections

--- 2020-06-30 chapters/my-work.tex
@@
 
-\section{Method: Feature Distribution Monitoring} \label{sec:ft-monitoring}
+\section{Method: Feature distribution monitoring} \label{sec:ft-monitoring}
 
@@
 We set the half-life to be one-quarter of the size of the window we want to emulate.
-We want to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data.
+The size of the window we want to monitor varies from situation to situation, but in the limited scope of this thesis we set to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data.
 This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.

### 2020-07-02 chapters/validation.tex
Small corrections

@@
 In this experiment, we set out to test our system in a low drift scenario.
-To guarantee the reference and target periods are as similar as possible, we decide to use the same dataset to work as reference and target.
+To guarantee the reference and target periods are as similar as possible, we decided to use the same dataset to work as reference and target.
 We chose to use the reference dataset from the previous experiment (Reference of Table \ref{tbl:merchant1-datasets-summary}) and feed it as input for both our batch and stream analysis as the reference and target periods, respectively.
@@
 Feature x1 has less alerts in this experiment where we use equal reference and target periods.
-This is expected and clear when comparing Figure \ref{fig:merchant2-x1-correctedpvalues} and \ref{fig:merchant-x1-correctedpvalues}.
-When comparing Figures \ref{fig:merchant-x2-correctedpvalues} and \ref{fig:merchant-x3-correctedpvalues} with \ref{fig:merchant2-x2-correctedpvalues} and \ref{fig:merchant2-x3-correctedpvalues} for feature x2 and x3, respectively, we observe that the number of alerts stayed more or less the same.
-We think its because x2 and x3 reference and target periods were already pretty similar, so ensuring both periods were the same did not make much difference.
+However, the alarms last for about two months.
+This is unexpected as we predicted we would see no alarms or few ones with short duration.
+Likewise, we observe that for feature x2 (Figure \ref{fig:merchant2-x2-correctedpvalues}) and x3 (Figure \ref{fig:merchant2-x3-correctedpvalues}) we still have many alarms and some lasting up to one month.
 \begin{figure}[!htb]
@@
 \end{figure}
+We do not have a clear explanation for this and due to time constraints we leave further tests to future work.
+We suspect it may be related to the way EMAs work: they keep every past event in the aggregation state.
+In other words, despite the fact EMAs apply a smoothing factor to the aggregation state for each new event, the previous events never have their contribution set to zero, even though the weights tend to zero.
 
@@
 Recall that the higher the p-value the less divergent a feature is.
-Overall we see that this feature has fewer alarms and stabilizes well above our alarm threshold which is coherent since we used the same dataset as a reference and target periods.
+Overall, we see that this feature has fewer alarms and stabilizes well above our alarm threshold which is coherent since we used the same dataset as reference and target periods.
 
@@
 The difference in corrected p-values is clear.
-Figure \ref{fig:merchant2-x5-correctedpvalues} shows this plot when using both reference and target periods.
+Figure \ref{fig:merchant2-x5-correctedpvalues} shows this plot when using equal reference and target periods.
 We observe that our corrected p-values never go below the threshold line, in fact, they stay well above it.

### 2020-07-02 chapters/background.tex
mais cenas
Capitalization

@@
 
-In this chapter we discuss fundamental concepts for our working context.
+In this chapter we discuss fundamental concepts that will help us identify the set of relevant techniques that are suitable to address the problem to be formulated and tackled in this thesis.
 In Section \ref{sec:stream-superset} we introduce batch and stream processing as two large scale data processing techniques.
@@
 
-\section{Stream processing as a superset of batch processing} \label{sec:stream-superset}
+\section{Stream Processing as a Superset of Batch Processing} \label{sec:stream-superset}
 
@@
 
-\section{From unbounded streams to finite datasets} \label{sec:windows}
+\section{From Unbounded Streams to Finite Datasets} \label{sec:windows}
 
@@
 
-\section{Aggregations over data} \label{sec:aggregations}
+\section{Aggregations over Data} \label{sec:aggregations}
 
@@
 
-\subsection{Sliding window aggregations}
+\subsection{Sliding Window Aggregations}
 
@@
 
-\subsection{Sliding window aggregation algorithms} \label{sec:back-swag-algs}
+\subsection{Sliding Window Aggregation Algorithms} \label{sec:back-swag-algs}
 
@@
 
-\subsection{Aggregation properties}
+\subsection{Aggregation Properties}
 \label{sec:agg-properties}
@@
 In this Thesis we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.
+
+\section{Summary}

### 2020-07-02 chapters/method.tex
Capitalization

@@
 
-\section{State of the art outlier detection methods}
+\section{State of the Art Outlier Detection Methods}
 Our first theoretical approach was to use one of the outlier detection methods presented in Section \ref{sec:outliers}.
@@
 
-\section{State of the art aggregation algorithms}
+\section{State of the Art Aggregation Algorithms}
 Dissimilarity-based univariate subsequence outlier detection methods as presented in Section \ref{sec:uni-sub-out} focus on measuring the distance between two periods of time and report an outlier when the distance is above a certain threshold.
@@
 
-\section{Method: Feature distribution monitoring} \label{sec:ft-monitoring}
+\section{Method: Feature Distribution Monitoring} \label{sec:ft-monitoring}
 
@@
 
-\subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
+\subsection{Exponential Moving Average Based Histogram} \label{sec:ema-hist}
 As stated, we compute histogram aggregations for the reference and target sliding windows.
@@
 
-\subsection{Batch analysis phase} \label{sec:batch-phase}
+\subsection{Batch Analysis Phase} \label{sec:batch-phase}
 \textcolor{red}{TODO: add some pseudocode}
@@
 
-\subsubsection*{Building the reference histogram for each feature}
+\subsubsection*{Building the Reference Histogram for each Feature}
 
@@
 
-\subsubsection*{Finding the distribution of distances for each feature} \label{sec:sampling-batch}
+\subsubsection*{Finding the Distribution of Distances for each Feature} \label{sec:sampling-batch}
 
@@
 
-\subsubsection*{Burn-in period at system initialization}
+\subsubsection*{Burn-In Period at System Initialization}
 When we boot up our system in the streaming phase we initially have empty histogram aggregations.
@@
 
-\subsubsection*{Batch phase artifacts} \label{sec:batch-artifacts-summary}
+\subsubsection*{Batch Phase Artifacts} \label{sec:batch-artifacts-summary}
 By the end of the batch phase, we should have, for each feature: 
@@
 
-\subsection{Streaming phase} \label{sec:stream-phase}
+\subsection{Streaming Phase} \label{sec:stream-phase}
 \textcolor{red}{TODO: add some pseudocode}
@@
 
-\subsection{Final remarks}
+\subsection{Final Remarks}
 
-\subsubsection*{Minimum number of samples} \label{sec:nsamples}
+\subsubsection*{Minimum Number of Samples} \label{sec:nsamples}
 In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase.

### 2020-07-02 chapters/problem-statement.tex
Capitalization

@@
 
-\section{The problem}
+\section{The Problem}
 Many real-time stream monitoring systems are static once deployed in a production environment.
@@
 
-\section{State of the art issues}
+\section{State of the Art Issues}
 Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques required to monitor streaming data patterns in real-time.
@@
 
-\section{Research questions}
+\section{Research Questions}
 The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time.

### 2020-07-02 chapters/sota.tex
Small corrections
Capitalization

@@
 
-\section{Outlier detection in time-series data} \label{sec:outliers}
+\section{Outlier Detection in Time-Series Data} \label{sec:outliers}
 Blázquez-García \emph{et al.} review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
@@
 
-\subsection{Univariate point outlier detection}
+\subsection{Univariate Point Outlier Detection}
 
@@
 
-\subsection{Univariate subsequence outlier detection} \label{sec:uni-sub-out}
+\subsection{Univariate Subsequence Outlier Detection} \label{sec:uni-sub-out}
 Similarly to point outlier detection methods, Blázquez-García \emph{et al.} \cite{Blazquez-Garcia-Review-Anomaly-Detection} split subsequence outlier detection methods into uni and multivariate ones (Figure \ref{fig:outlier-taxonomy}).
@@
 
-\section{Sliding window aggregation algorithms} \label{sec:sota-swag-algs}
+\section{Sliding Window Aggregation Algorithms} \label{sec:sota-swag-algs}
 
@@
 
-\section{Probabilistic data structures} \label{sec:pds}
+\section{Probabilistic Data Structures} \label{sec:pds}
 In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs).
@@
 
-\subsection{Membership queries and the Bloom Filter} \label{sec:bloom}
+\subsection{Membership Queries and the Bloom Filter} \label{sec:bloom}
 % approx aggregation to compute
@@
 Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}.
-However, we need to evict old events, and for that reason, bloom filters are of no use to us (unless we find sliding window implementations).
-
-\subsection{Item frequency and the Count-Min Sketch}
+However, we need to evict old events, and for that reason, bloom filters are of no use to us.
+
+\subsection{Item Frequency and the Count-Min Sketch}
 % approx aggregation to compute
@@
 
-\subsection{Cardinality estimation and the HyperLogLog} \label{sec:hyper-log-log}
+\subsection{Cardinality Estimation and the HyperLogLog} \label{sec:hyper-log-log}
 
@@
 
-\section{Sliding window aggregations with probabilistic data structures} \label{sec:sliding-pds}
+\section{Sliding Window Aggregations with Probabilistic Data Structures} \label{sec:sliding-pds}
 In Section \ref{sec:pds} we presented a set of probabilistic data structures constant in time and space.

### 2020-07-02 chapters/validation.tex
Capitalization

@@
 
-\section{Experiments with synthetic data}
+\section{Experiments with Synthetic Data}
 In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
@@
 
-\subsection{Single feature analysis}
+\subsection{Single Feature Analysis}
 As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
@@
 
-\subsubsection{Varying sample sizes}
+\subsubsection{Varying Sample Sizes}
 We began with single-feature scenarios and by experimenting with different tuple-based half-lives $n_{1/2}$.
@@
 
-\subsubsection{Multiple distribution changes}
+\subsubsection{Multiple Distribution Changes}
 In this experiment, we wanted to see how our signal behaved when there were multiple change points, where each gaussian distribution was a scaled down version of the previous one.
@@
 
-\subsubsection{Multiple distribution changes, each a superset of the previous one}
+\subsubsection{Multiple Distribution Changes, each a Superset of the Previous One}
 In this experiment, we wanted to do the reverse of the previous one.
@@
 
-\subsubsection{Return to normality after change}
+\subsubsection{Return to Normality After Change}
 Our goal with this experiment was to check if our signal returned to its normal values (below the alert threshold) if we changed our distribution to the original reference one, as intended.
@@
 
-\subsubsection{An attempt to deceive our method}
+\subsubsection{An Attempt to Deceive our Method}
 In this experiment, we attempt to deceive our method by using a different distribution from the reference period but ensuring its domain is roughly the same as the reference one.
@@
 
-\subsection{Multi feature analysis}
+\subsection{Multi Feature Analysis}
 Before moving to real data, we performed one last experiment, using synthetic data but this time performing a multi-feature analysis.
@@
 \clearpage
-\section{Experiments with real data}
+\section{Experiments with Real Data}
 
@@
     %x5 - dif_last_and_current_trx
-\subsection{Same reference and target periods}
+\subsection{Same Reference and Target Periods}
 In this experiment, we set out to test our system in a low drift scenario.
@@
 \end{figure}
+
+\section{Conclusions}
+
+With this set of experiments

### 2020-07-02 chapters/background.tex
Background: summary

@@
 \section{Summary}
+In this chapter, we introduced fundamental concepts to understand the set of methods that will allow us to test our hypothesis, presented in Chapter \ref{chap:sota}.
+
+In Section \ref{sec:stream-superset}, we introduced batch and stream processing as two large scale data processing techniques.
+We defined each and elaborated on the fundamental differences between both paradigms.
+We conclude that there is no best technique and that each serves a different use case.
+
+In Section \ref{sec:windows} we discussed windowing mechanisms that allowed us to convert unbounded data streams (or infinite datasets) to finite datasets.
+
+In Section \ref{sec:aggregations} we introduced aggregations over datasets.
+We gave examples of computable aggregations over unbounded datasets at first, but then discussed sliding window aggregations, which make use of windowing techniques to obtain a finite dataset, on which the aggregation is computed.
+We further discussed algorithms used for the efficient computation of sliding window aggregations.
+Lastly, we presented a set of aggregation properties.
+
+While sliding window aggregation algorithms keep the window contents in memory, Exponential Moving Averages, presented in Section \ref{sec:emas}, replace them for a lighter representation, providing an approximate estimation of the aggregation value.

### 2020-07-02 chapters/sota.tex
Sota Summary

@@
 Despite having the same time and space complexities, the Sliding HLL DRV approach uses a smaller amount of memory that does not change during runtime, while the LFPM requires heap resizing because the lists dynamically shrink and grow.
-Hence, the DRV approach is a better fit to integrate our lightweight real-time solution.
+Hence, the DRV approach is a better fit to integrate our lightweight real-time solution, if we need to perform cardinality estimation.
 
@@
 \section{Summary}
+
+In this chapter, we presented state-of-the-art outlier detection methods and classified them according to Blázquez-García \emph{et al.} taxonomy \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
+Additionally, we presented sliding window aggregation algorithms, probabilistic data structures and their respective sliding window implementations, with the intent of researching how to incrementally maintain an aggregation in a streaming environment.
+
+In Section \ref{sec:outliers}, we reviewed outlier detection algorithms.
+However, we found most of them inapplicable in our scenario, because they were resource intensive (both time and memory wise), provided no measure of how divergent and outlier was or because they lost track of their reference period.
+
+In Section \ref{sec:sota-swag-algs}, we presented sliding window aggregation algorithms that we intended on using to compute our aggregations.
+However, their time or memory complexities did not meet our sub-linear requirements.
+
+We then discussed probabilistic data structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}).
+Despite being efficient, both time and memory wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use for us.

### 2020-07-03 chapters/validation.tex
Validation conclusions

@@
 
-All the synthetic datasets used throughout this Section are a time-series where at each time-step we sample uniformly from a fixed distribution, \textit{i.e.} there are no time correlations in the used datasets.
+All the synthetic datasets used throughout this section are a time-series where at each time-step we sample uniformly from a fixed distribution, \textit{i.e.} there are no time correlations in the used datasets.
 The reference dataset R1 used in this first experiment contained a single feature that followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$ for the 5 million rows of the dataset.
@@
 \section{Conclusions}
+In this section, we summarize our experimental tests, results and found issues.
 
-With this set of experiments
+We began by performing single-feature analysis on synthetic data.
+The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported.
+
+Still working with synthetic data, we performed our first multi-feature analysis.
+In this scenario, we applied the Holm-Bonferroni multiple test correction to the p-values of each hypothesis.
+Once again, each feature was correctly reported as divergent at the timestamps where we changed the underlying generating distribution.
+
+Finally, we moved on to real data.
+Due to time constraints and the scope of this thesis, we experimented only with data from one merchant.
+The results with real data were not satisfactory.
+In our first experiment with real data, we ran our algorithm with what we thought was an ideal set of parameters and in return produced a lot of alarms.
+A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift.
+In our last experiment, we attempted to reduce concept drift to a maximum by using the same period as reference and target.
+To our surprise, we still produced a large number of long lasting alarms.
+We defer the investigation of the root cause of the problem to future work and leave a few untested hypothesis as solutions for this issue.

### 2020-07-03 chapters/introduction.tex
Capitalization
Document Structure section

@@
 
-\section{Document structure}
+\section{Document Structure}
 
-\textcolor{red}{TODO at the end rewrite and relink this}
+This section merely sums up how each of the following chapters contributes to this thesis.
 
-This section merely sums up how each of the following Chapters contributes to this Thesis.
+In Chapter \ref{chap:background} of this document, we brief the reader with the necessary background and relevant concepts needed to understand the following work, namely the topics of Large Scale Data Processing, Sliding Window Aggregations and Outlier Detection, as well as all the sub-topics inherent to them.
 
-%In Chapter \ref{chap:background} of this document we brief the reader with the necessary background and relevant concepts needed to understand the following Chapters.
+In Chapter \ref{chap:sota}, we examine state of the art methods related with the previously mentioned topics and discuss shortcomings, strengths and the applicability of such methods to our hypothesis.
 
-%In Chapter \ref{chap:sota} we examine the state of the art in the area, analyzing the most recent work done and classifying it into subgroups.
+In Chapter \ref{chap:statement}, we define the problem this thesis aims to solve, the issues with state of the art solutions, our proposed solution, the assumptions made on it and the research questions we intend to answer.
 
-%In Chapter \ref{chap:approach} we define the scope of this project, the research questions we intend to answer, our experimental methodology and our future planning.
+In Chapter \ref{chap:my-work}, we discuss initial thought solutions and reasons to discard them before presenting our distribution monitoring method.
 
-%In Chapter \ref{chap:conclusion} we summarize our contributions.
+In Chapter \ref{chap:validation}, we describe the set of experiments made and the datasets used.
+We start with single feature experiments on synthetic data, then multi feature analysis and finally move on to multi feature analysis on real data.
+
+In Chapter \ref{chap:conclusion}, we draw our final conclusions.
+We reiterate our hypothesis and research questions and we provide answers to them.
+We summarize our contributions and make explicit current issues to be addressed as future work.

### 2020-07-03 chapters/conclusions.tex
Restivo feedback applied
Capitalization
Validation conclusions
Document Structure section
Cute colors on links lol
Conclusions

@@
 
-\section{Hypothesis Summary}
-\section{Answer to the hypothesis}
+\section{Hypothesis Revisited}
+
+In Section \ref{sec:hypothesis} we defined our hypothesis or main question to address as whether sliding window aggregations and outlier detection methods combined could be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.
+Before discussing our hypothesis, we revisit the research questions posed in Section \ref{sec:rqs} and provide answers to them.
+
+\subsubsection*{(RQ1) What aggregation can we use to properly encode data distributions?}
+First, we searched for a set of aggregations that would encode our distribution in the best way.
+Later on, we concluded that a simple histogram is better than any other set of aggregations.
+Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values.
+
+In brief, the answer to RQ1 is that we can use histogram aggregations to properly encode data distributions.
+
+\subsubsection*{(RQ2) How can we maintain such an aggregation in a sliding window fashion in real-time with constant memory usage?}
+
+This research question really is about two things: maintaining the sliding window aggregation in real-time and keeping its memory usage constant.
+
+The aggregation in question is an histogram aggregation.
+An histogram requires storing a fixed-size list of bins and corresponding counts, regardless of the volume of data.
+This means that for ever growing volumes of data our memory consumption stays the same.
+
+However, maintaining a sliding window histogram in real-time is an harder challenge.
+Traditionally, sliding window maintenance requires evicting the oldest element from the window when a new one is inserted and updating the aggregation state.
+We could maintain the histogram in constant time using a Subtract-On-Evict (SOE) algorithm (Algorithm \ref{pseudo:soe}): for each new event, we would evict the oldest one and decrement the bin it belonged to and insert the new event in the window while incrementing the bin it belongs to.
+However, using this approach, we violate our sublinear memory consumption, because now we have to keep the entire window contents in memory, hence we grow linearly regarding sliding window size.
+We instead propose an EMA-like approximate histogram aggregation, in Section \ref{sec:ema-hist}.
+An EMA-like histogram has constant memory complexity and can be maintained in real-time.
+
+To summarize, EMA-like approximate histogram aggregations, defined in Section \ref{sec:ema-hist}, have constant memory and time complexity and can be maintained in a sliding window fashion.
+
+\subsubsection*{(RQ3) How can we measure divergence between the reference and sliding window aggregations?}
+Statistical tests that measure the similarity between two probability distributions are common in the literature.
+Tests like Kolmogorov–Smirnov, Wassertein or Jensen–Shannon (the one we used) provide a bounded measure of distance between two probability distributions, or histograms.
+
+In brief, since the aggregation we use is an histogram, we can use statistical tests that measure similarity between probability distributions.
+We chose to use the Jensen–Shannon Divergence (JSD).
+The JSD takes as input two histograms and outputs a value between 0 and 1.
+The closer this value is to 0 the more similar the distributions are.
+On the other hand, a JSD value of 1 represents two different distributions.
+We chose to use the JSD metric due to its simplicity and leave as future work the usage of other statistical tests.
+
+\subsubsection*{(RQ4) Based on a divergence measure, when should we raise an alert?}
+Statistical divergence tests provide fixed bounds for the measurement value.
+In the case of the Jensen–Shannon Divergence (JSD), the output ranges from 0 to 1.
+This research question poses the following challenge: how can we threshold a JSD value?
+In other words, do we alert a feature as divergent if the JSD measure is above 0.6?
+Or if above 0.99?
+How can we threshold this value?
+
+In Section \ref{sec:sampling-batch}, we explain how we define the threshold and hence define when we raise an alert.
+In our batch phase, we make a number of samples and measure the JSD distance between the histogram aggregations of those samples and our reference histogram aggregation.
+We end up with a list of JSD values.
+Given a new JSD value, we check against the now known distribution of JSD values and alert if above a certain percentile.
+Note that defining a threshold percentile is easier, because we are defining a probability: alerting above the 99th percentile means alerting for hypothesis with 1\% probability.
+
+In brief, we build the distribution of divergence values in batch and use it to compare a future divergence value and find out its probability.
+If said divergence value has a very low probability (\textit{e.g.}, 1\% or less), then we raise an alert.
+
+\subsubsection*{Hypothesis Discussion}
+In our synthetic data experiments, we were able to detect all of the introduced anomalies.
+Additionally, we were able to do it achieving very high streaming throughputs and low latencies.
+In the real data experiments, once again, we managed to achieve high troughputs and low latencies but the produced alerts were not very accurate.
+We defer further investigation to future work because of the time constraints of this thesis but leave open the possiblity that the approximate aggregation value is more inaccurate than what we thought it would be, due to the fact that the EMA tail holds all past events.
+
+Overall, our set of tests supports the claim that sliding window aggregations and outlier detection methods can indeed be combined to detect data pattern shifts in streaming scenarios, with constant time and memory complexity.
+
 \section{Contributions}
-\section{Future work}
-\section{Final remarks}
+We summarize our contributions as follows:
+
+\begin{enumerate}
+    \item a two-phased and two-windowed univariate subsequence outlier detection method, classified according to the taxonomy in Section \ref{sec:uni-sub-out} and detailed in Chapter \ref{chap:my-work}, that is lightweight and works for real-time stream monitoring
+
+    \item an histogram aggregation based on Exponential Moving Averages, that replaces the entire sliding window contents by a lighter representation, resulting in a constant in time and space complexity sliding window aggregation
+
+    \item a set of synthetic datasets and experiments where the method in question accurately detected anomalies
+
+    \item experiments on real data that did not show reasonable results but where we provide insights and possible hypothesis to test for future work
+\end{enumerate}
+
+\section{Future Work}
+
+First and foremost, we believe the next steps are to reason about the obtained results in the real data experiments and find the root cause for the existence of many and large alert periods.
+As possible hypothesis to test, we suggest testing with EMAs with higher smoothing factors, so that past events are forgotten faster; additionally, we suggest decreasing $\gamma$, which further increases the number of samples to make in the batch phase.
+
+In this thesis we used the Jensen–Shannon Divergence (JSD) as statistical test to measure the distance or divergence between our reference and target sliding window aggregation.
+We chose to use the JSD metric due to its simplicity and because it was used in previous work successfully \cite{SAMM}.
+We believe that future work comprises testing new distances in the literature.
+In particular, we believe the Wassertein distance is worth trying next, and illustrate why in Figure \ref{fig:wasserteinvsjsd}.
+\textcolor{red}{TODO: make this image better}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.6]{figures/wasserstein-vs-jsd.pdf}
+      \caption{Wassertein vs Jensen–Shannon distances}
+      \label{fig:wasserteinvsjsd}
+    \end{center}
+\end{figure}
+The Wassertein distance measures the work required to transport the probability mass from one histogram to the other.
+This measure will be larger the further way the probability mass is.
+In Figure \ref{fig:wasserteinvsjsd} scenario, while in the JSD measure is the same for both cases, the Wassertein distance is larger in case 2 because the probability mass has to be transported further.
+
+In this thesis we had to deal with the multiple test problem and apply a multiple test correction.
+In our case, we used Holm-Bonferroni due to its simplicity and because it was similar to past work \cite{SAMM}.
+There are many other multiple test correction methods, and we leave the testing of each as future work.
+
+Lastly, we propose another type of validation method.
+We mentioned in the beginning that our objective is to detect data pattern shifts before the performance of the systems relying on the data declines.
+We propose a test where we train a Machine Learning model on a reference period and measure its prediction accuracy on a target period.
+Then, we propose running our method on reference and target periods to find out timestamps with a high number of divergent features.
+We propose that the model is retrained at these timestamps (tantamount to system reconfiguration) and then used to predict for the remaining target period.
+Finally, we propose that the accuracy should be compared to the previously observed model performance, expecting it to be better after retraining in one of the high divergent timestamps.

### 2020-07-03 chapters/introduction.tex
Conclusions

@@
 \section{Hypothesis} \label{sec:hypothesis}
-The question this Thesis aims to answer is whether \textit{sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.} In this Thesis, we study sliding window aggregations that replace the sliding window events in memory by a lighter representation, while still being able to provide an estimate of the aggregation value.
+The question this Thesis aims to answer is whether \textit{sliding window aggregations and outlier detection methods combined can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.} In this Thesis, we study sliding window aggregations that replace the sliding window events in memory by a lighter representation, while still being able to provide an estimate of the aggregation value.
 

### 2020-07-03 chapters/problem-statement.tex
Conclusions

@@
 
-\section{Research Questions}
+\section{Research Questions} \label{sec:rqs}
 The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time.
 To that end, we devise a series of research questions to be answered by the end of this Thesis:
-
-\textcolor{red}{TODO: further describe each RQ?}
 

### 2020-07-03 chapters/validation.tex
Conclusions

@@
 The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported.
-
 Still working with synthetic data, we performed our first multi-feature analysis.
@@
 Due to time constraints and the scope of this thesis, we experimented only with data from one merchant.
-The results with real data were not satisfactory.
-In our first experiment with real data, we ran our algorithm with what we thought was an ideal set of parameters and in return produced a lot of alarms.
+In our first experiment with real data, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms.
 A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift.
@@
 To our surprise, we still produced a large number of long lasting alarms.
-We defer the investigation of the root cause of the problem to future work and leave a few untested hypothesis as solutions for this issue.
+We defer the investigation of the root cause of the problem to future work and leave a few untested hypothesis as solutions for this issue, like the possibility that despite applying a smoothing factor to the aggregation state for each new event, EMAs never forget past events.
+In other words, the previous events never have their contribution set to zero, even though the weights tend to zero.
+Additionally, our sampling may have just been insufficient.
+To test this, an experiment with lower $\gamma$ and hence higher number of samples should be carried out.

### 2020-07-03 chapters/acknows.tex
Some Background feedback applied
Ack

@@
 \chapter*{Acknowledgements}
+
+A lot of people contributed to this thesis. I will explicitly thank everyone, in no particular order.
+
+I thank my academic supervisors, Professors André Restivo and Pedro Ribeiro, for their valuable insights on how to conduct a research project, namely a master thesis, and how to structure this final document.
+I thank them for the multiple times where they advised me on which direction to take and the multiple great ideas they contributed with.
+Above all, I thank them for always putting my interests first and ensuring I had the necessary conditions to research, develop and write my thesis.
+
+I thank my company supervisors, Marco Sampaio and Pedro Silva, for the advice they gave me throughout this thesis which made it possible for me to develop this project, step by step.
+I thank them for the hours they spent discussing with me next steps to take, experiments to conduct, new methods to try and for the remaining hours they spent analyzing the results with me.
+Finally, I thank them for the many hours spent reviewing this document.
+
+I thank the entire Feedzai Systems Research team, namely, João Oliveirinha, Pedro Silva and Sofia Gomes for their daily support and suggestions.
+I thank them for the technical tips they gave me to surpass certain tasks that at first seemed impossible.
 

### 2020-07-03 chapters/introduction.tex
Ack

@@
 
-In Chapter \ref{chap:my-work}, we discuss initial thought solutions and reasons to discard them before presenting our distribution monitoring method.
+In Chapter \ref{chap:my-work}, we discuss initial solutions and reasons to discard them before presenting our distribution monitoring method.
 
@@
 In Chapter \ref{chap:conclusion}, we draw our final conclusions.
-We reiterate our hypothesis and research questions and we provide answers to them.
-We summarize our contributions and make explicit current issues to be addressed as future work.
+We reiterate our hypothesis and research questions and provide answers to them.
+Finally, we conclude by summarizing our contributions and make explicit current issues to be addressed as future work.

### 2020-07-03 chapters/background.tex
Small corrections

@@
 In Section \ref{sec:stream-superset}, we introduced batch and stream processing as two large scale data processing techniques.
-We defined each and elaborated on the fundamental differences between both paradigms.
-We conclude that there is no best technique and that each serves a different use case.
+We defined each, elaborated on the fundamental differences between both paradigms but also introduced the superset relation there is between stream and batch processing.
 

### 2020-07-03 chapters/conclusions.tex
Small corrections

@@
 
-    \item experiments on real data that did not show reasonable results but where we provide insights and possible hypothesis to test for future work
+    \item experiments on real data that showed unexpected results but where we provide insights and possible hypothesis to test for future work
 \end{enumerate}
@@
 In this thesis we had to deal with the multiple test problem and apply a multiple test correction.
-In our case, we used Holm-Bonferroni due to its simplicity and because it was similar to past work \cite{SAMM}.
+In our case, we used Holm-Bonferroni due to its simplicity and because it controls the family wise error rate with a larger power than some other methods (thus providing a good trade-off between simplicity and power).
 There are many other multiple test correction methods, and we leave the testing of each as future work.

### 2020-07-03 chapters/introduction.tex
Small corrections

@@
 
-This section merely sums up how each of the following chapters contributes to this thesis.
+This section outlines the content of each chapter of this thesis.
 
-In Chapter \ref{chap:background} of this document, we brief the reader with the necessary background and relevant concepts needed to understand the following work, namely the topics of Large Scale Data Processing, Sliding Window Aggregations and Outlier Detection, as well as all the sub-topics inherent to them.
+In Chapter \ref{chap:background} of this document, we brief the reader with the necessary background and relevant concepts needed to understand the following work, namely the topics of Large Scale Data Processing, Sliding Window Aggregations and Outlier Detection, as well as all relevant sub-topics inherent to them.
 
@@
 
-In Chapter \ref{chap:my-work}, we discuss initial solutions and reasons to discard them before presenting our distribution monitoring method.
+In Chapter \ref{chap:my-work}, we discuss initial approaches and reasons to discard them before presenting our proposed solution.
 
-In Chapter \ref{chap:validation}, we describe the set of experiments made and the datasets used.
+In Chapter \ref{chap:validation}, we describe the datasets used and the set of experiments that were performed to validate our method.
 We start with single feature experiments on synthetic data, then multi feature analysis and finally move on to multi feature analysis on real data.
@@
 We reiterate our hypothesis and research questions and provide answers to them.
-Finally, we conclude by summarizing our contributions and make explicit current issues to be addressed as future work.
+Finally, we conclude by summarizing our contributions and, make explicit current issues to be addressed as future work.

### 2020-07-03 chapters/sota.tex
Small corrections

@@
 
+Track changes is off
+Everyone
+Track changes for everyone
+You
+Track changes for You
+Guests
+Track changes for guests
+pedro.cardoso: How is this error calculated?
+Jun 18, 2020 2:52 PM
+You: Same as above
+Jun 20, 2020 5:09 PM • Edit • Delete
+Hit Enter to reply
+pedro.cardoso: This 3rd option is under-explained compared to the other 2. Please elaborate
+Jun 18, 2020 2:53 PM
+You: Not sure myself, that is why is so poorly written. Im going to revisit this
+Jun 19, 2020 4:42 PM • Edit • Delete
+Hit Enter to reply
+pedro.cardoso: Thought: This reads like leftover from the original work. Is there a way to better relate your investigation into SWAGs into the methodology you ended up taking?
+Jun 30, 2020 10:19 AM
+pedro.cardoso: Please revisit this entire section. Particularly the applicability sections of each algorithm in the context of your current approach.
+Jun 30, 2020 10:20 AM
+Hit Enter to reply
+pedro.cardoso: Remove these TODOs. If you want keep a note of them elsewhere, but not as text render being rendered.
+Jul 3, 2020 5:28 PM
+pedro.cardoso: You will forget to do it with other tasks still needed to be done.
+Jul 3, 2020 5:29 PM
+Hit Enter to reply
+pedro.cardoso: "because they evolved their reference period"
+Jul 3, 2020 5:30 PM
+Hit Enter to reply
+pedro.cardoso: Where is Model Monitoring?
+Jul 3, 2020 5:31 PM
+Hit Enter to reply
+Current file
+Overview
+16
+
 Yang et al.
@@
 %ALSO a 2 WINDOW MODEL
-%\textcolor{red}{TODO: write this}
 
@@
 
-\textcolor{red}{TODO Find and add versions of Sliding Bloom Filter and Sliding CMS}
-
 \section{Summary}
@@
 In Section \ref{sec:outliers}, we reviewed outlier detection algorithms.
-However, we found most of them inapplicable in our scenario, because they were resource intensive (both time and memory wise), provided no measure of how divergent and outlier was or because they lost track of their reference period.
+However, we found most of them inapplicable in our scenario, because they were resource intensive (both time and memory wise), provided no measure of how divergent and outlier was or because they evolved their reference period.
 
@@
 We then discussed probabilistic data structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}).
-Despite being efficient, both time and memory wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use for us.
+Despite being efficient, both time and memory wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use to us.

### 2020-07-03 chapters/validation.tex
Small corrections

@@
 We do not have a clear explanation for this and due to time constraints we leave further tests to future work.
-We suspect it may be related to the way EMAs work: they keep every past event in the aggregation state.
-In other words, despite the fact EMAs apply a smoothing factor to the aggregation state for each new event, the previous events never have their contribution set to zero, even though the weights tend to zero.
+We suspect the issue may be that there is a difference between the batch and streaming phases.
+In batch we have subsets of events and compute EMA-like histogram aggregations based on those.
+In streaming, however, we have a continuous stream of data and never expire the 6\% of data that in batch is not present, because it is a finite window.
 
@@
 \section{Conclusions}
-In this section, we summarize our experimental tests, results and found issues.
+In this section, we summarize our experimental tests, results and uncovered issues.
 
@@
 To our surprise, we still produced a large number of long lasting alarms.
-We defer the investigation of the root cause of the problem to future work and leave a few untested hypothesis as solutions for this issue, like the possibility that despite applying a smoothing factor to the aggregation state for each new event, EMAs never forget past events.
-In other words, the previous events never have their contribution set to zero, even though the weights tend to zero.
-Additionally, our sampling may have just been insufficient.
+We defer the investigation of the root cause of the problem to future work and leave a few untested hypothesis as solutions for this issue, like the possibility that the EMA-like histograms in batch work with subsets of data and the streaming ones work with infinite amounts of data.
+In other words, for the streaming EMAs, the previous events never have their contribution set to zero, even though the weights tend to zero, whereas the batch EMAs worked with a finite dataset.
+Additionally, our sampling may have been insufficient.
 To test this, an experiment with lower $\gamma$ and hence higher number of samples should be carried out.

### 2020-07-04 chapters/abstract.tex
Some Background feedback applied
mais cenas
Small corrections
Abstract

--- 2020-02-17 chapters/abstract.tex
@@
 
-Data pattern shift detection has been researched within diverse areas and application domains and refers to the process of finding patterns in data that do not conform to expected or usual behavior.
+%The Abstract
+%State the problem, your approach and solution, and the main contributions of the paper.
+Include little if any background and motivation.
+Be factual but comprehensive.
+The material in the abstract should not be repeated later word for word in the paper.
+
+%Neste contexto, tínhamos este problema, desenvolvemos isto, avaliamos desta forma e chegamos a estas conclusões
+
+% The problem
+%Data pattern shift detection has been researched within diverse areas and application domains and refers to the process of finding patterns in data that do not conform to expected or usual behavior.
 Accurate and timely detection of data pattern deviations allows for immediate measures to be taken, preventing loss of income and maintaining the companies' reputation among customers.
-As such, many large companies take this into consideration and deploy data anomaly components in their systems.
+As such, many large companies take this into consideration and deploy data anomaly detection components in their systems.
 In this way, system administrators can proactively adjust and prevent potential service failures.
 
-For example, Yahoo's anomaly detection system fulfills its purpose of monitoring and raising alerts on time-series events for several of Yahoo's use cases.
-At Microsoft, their custom, internally made anomaly detection system monitors millions of metrics coming from a variety of different internal services, enabling engineers to solve live on-site issues before they scale and become bigger ones.
-
+% The problem
 Many real-time stream monitoring systems are static once deployed in a production environment.
-Engineers of said systems usually configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data.
+The engineers of those systems configure them under the assumption that future data flowing through the system roughly follows the same distribution as previously seen data.
 They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system.
 Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.
+Data pattern shift detection refers to the process of finding patterns in data that do not conform to expected or usual behavior.
+Accurate and timely detection of data pattern deviations allows for immediate measures to be taken.
+Thus, the problem at hand is to determine when to reconfigure the system, \textit{e.g.}, a Machine Learning model, based on an analysis of the drifts in the stream of data.
 
-With this thesis, we set out to design and evaluate a system that makes use of lightweight streaming analytical methods to detect deviations in data patterns while in the presence of large volumes of high velocity, highly skewed, seasonal data.
+%Approach/Solution and %Main contributions
+In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data.
+We develop a two-phased and two-windowed method: first we perform a batch analysis on a reference window and then a stream analysis over a sliding streaming window.
+We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages.
+We do this for each event and for each of its fields, also denominated features in our context.
+We measure the divergence between both reference and sliding window histograms for each feature with the Jensen–Shannon Divergence function, but other functions like the Wassertein distance could be used.
+For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and the associated features.
+We generate alerts based on a user-defined probability threshold, for each feature.
+
+We evaluate our method through a series of tests, some with synthetic datasets and some with real data.
+We further split the tests into single and multi feature analysis.
+Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi feature analysis.
+Experiments with real data were not as accurate.
+Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypothesis that might explain it and hence are worthy of pursuing next.
+
+Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
 
 \vspace*{10mm}\noindent
-\textbf{Keywords}: real-time, streaming, lightweight, monitoring, volume, velocity, variety, seasonal
+\textbf{Keywords}: streaming, real-time, lightweight, monitoring, volume, velocity, seasonal

### 2020-07-04 chapters/conclusions.tex
Abstract

@@
 In our synthetic data experiments, we were able to detect all of the introduced anomalies.
-Additionally, we were able to do it achieving very high streaming throughputs and low latencies.
-In the real data experiments, once again, we managed to achieve high troughputs and low latencies but the produced alerts were not very accurate.
+Additionally, we were able to do it achieving very high streaming throughputs.
+In the real data experiments, once again, we managed to achieve high troughputs but the produced alerts were not very accurate.
 We defer further investigation to future work because of the time constraints of this thesis but leave open the possiblity that the approximate aggregation value is more inaccurate than what we thought it would be, due to the fact that the EMA tail holds all past events.

### 2020-07-04 chapters/abstract.tex
Changed images

@@
 We do this for each event and for each of its fields, also denominated features in our context.
-We measure the divergence between both reference and sliding window histograms for each feature with the Jensen–Shannon Divergence function, but other functions like the Wassertein distance could be used.
+We measure the divergence between both reference and sliding window histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used.
 For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and the associated features.
@@
 Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi feature analysis.
-Experiments with real data were not as accurate.
+However, experiments with real data were not as accurate.
 Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypothesis that might explain it and hence are worthy of pursuing next.
 
-Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
+Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity regarding window size.
 

### 2020-07-04 chapters/abstract.tex
SAMM

@@
 \chapter*{Abstract}
-
-%The Abstract
-%State the problem, your approach and solution, and the main contributions of the paper.
-Include little if any background and motivation.
-Be factual but comprehensive.
-The material in the abstract should not be repeated later word for word in the paper.
-
-%Neste contexto, tínhamos este problema, desenvolvemos isto, avaliamos desta forma e chegamos a estas conclusões
-
-% The problem
-%Data pattern shift detection has been researched within diverse areas and application domains and refers to the process of finding patterns in data that do not conform to expected or usual behavior.
-Accurate and timely detection of data pattern deviations allows for immediate measures to be taken, preventing loss of income and maintaining the companies' reputation among customers.
-As such, many large companies take this into consideration and deploy data anomaly detection components in their systems.
-In this way, system administrators can proactively adjust and prevent potential service failures.
 
@@
 %Approach/Solution and %Main contributions
-In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data.
+In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data.
 We develop a two-phased and two-windowed method: first we perform a batch analysis on a reference window and then a stream analysis over a sliding streaming window.
@@
 We measure the divergence between both reference and sliding window histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used.
-For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and the associated features.
+For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and their associated features.
 We generate alerts based on a user-defined probability threshold, for each feature.
@@
 We evaluate our method through a series of tests, some with synthetic datasets and some with real data.
-We further split the tests into single and multi feature analysis.
+We further split the tests into single and multi-feature analysis.
 Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi feature analysis.
@@
 Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypothesis that might explain it and hence are worthy of pursuing next.
-
 Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity regarding window size.
@@
 \vspace*{10mm}\noindent
-\textbf{Keywords}: streaming, real-time, lightweight, monitoring, volume, velocity, seasonal
+\textbf{Keywords}: streaming, real-time, lightweight, monitoring, concept drift, anomaly detection

### 2020-07-04 chapters/acknows.tex
Abstract
SAMM

@@
 
-A lot of people contributed to this thesis. I will explicitly thank everyone, in no particular order.
+A lot of people contributed to this thesis and I will explicitly thank everyone, in no particular order.
 
@@
 I thank my company supervisors, Marco Sampaio and Pedro Silva, for the advice they gave me throughout this thesis which made it possible for me to develop this project, step by step.
-I thank them for the hours they spent discussing with me next steps to take, experiments to conduct, new methods to try and for the remaining hours they spent analyzing the results with me.
+I thank them for the hours they spent discussing with me the next steps to take, experiments to conduct, new methods to try and for the remaining hours they spent analyzing the results with me.
 Finally, I thank them for the many hours spent reviewing this document.
@@
 
+Last but certainly not least, I thank my family, friends and girlfriend for keeping me sane while I worked on this thesis during the COVID-19 lockdown.
+
 \vspace{10mm}

### 2020-07-04 chapters/background.tex
Changed images
SAMM

@@
 
-The Venn diagram in Figure \ref{fig:stream-superset} from a blog post by Ververica \footnote{https://www.ververica.com/blog/batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
+The Venn diagram in Figure \ref{fig:stream-superset} \cite{batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
 A data stream is unbounded or infinite while a bounded data stream has a beginning and an end.
@@
       \includegraphics[scale=0.5]{figures/streaming-subset-batch.png}
-      \caption{Stream processing as a superset of batch processing}
+      \caption[Stream processing as a superset of batch processing]{Stream processing as a superset of batch processing \cite{batch-is-a-special-case-of-streaming}}
       \label{fig:stream-superset}
@@
       \includegraphics[scale=0.5]{figures/unbounded-superset-bounded.png}
-      \caption{Unbounded data streams as a superset of bounded datasets}
+      \caption[Unbounded data streams as a superset of bounded datasets]{Unbounded data streams as a superset of bounded datasets \cite{batch-is-a-special-case-of-streaming}}
       \label{fig:unbounded-bounded-superset}
@@
     \begin{center}
-      \hspace*{0.6in}
-      \includegraphics[scale=0.35]{figures/sliding.png}
+      \includegraphics[scale=1.1]{figures/sliding.pdf}
       \caption[True sliding window]{True sliding window, W=3, S=1}
@@
 
-\textcolor{red}{TODO change all window drawings to a not horrendous style}
-
 \begin{figure}[!htb]
     \begin{center}
-      \includegraphics[scale=0.3]{figures/stepping.png}
+      \includegraphics[scale=1.1]{figures/stepping.pdf}
       \caption[Stepping window]{Stepping window, W=3, S=2}
@@
     \begin{center}
-      \includegraphics[scale=0.3]{figures/tumbling.png}
+      \includegraphics[scale=1.1]{figures/tumbling.pdf}
       \caption[Tumbling window]{Tumbling window, W=S=3}

### 2020-07-04 chapters/introduction.tex
SAMM

@@
 Ensuring constant time and space complexity means that the system will remain usable for the ever-growing volumes of data.
-Since such a monitoring system is not mission-critical --- \textit{i.e.} the provided services don't rely on it --- the operational cost associated with it must be kept to a minimum.
+Since such a monitoring system is not mission-critical --- \textit{i.e.} the provided services do not rely on it --- the operational cost associated with it must be kept to a minimum.
 Having a real-time system monitoring other real-time systems consuming as many resources as the latter would be far too costly.

### 2020-07-04 chapters/method.tex
Conclusions
SAMM

@@
 On the other hand, a JSD value of 1 represents two different distributions.
-We chose to use the JSD metric due to its simplicity and leave as future work other possible tests.
+We chose to use the JSD metric due to its simplicity and because it was used in previous work successfully \cite{SAMM}, leaving as future work other possible tests.
 
@@
 \subsection{Batch Analysis Phase} \label{sec:batch-phase}
-\textcolor{red}{TODO: add some pseudocode}
-
 The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and EMA-like aggregations.
@@
 \subsection{Streaming Phase} \label{sec:stream-phase}
-\textcolor{red}{TODO: add some pseudocode}
-
 In the streaming or online phase, the system will process event by event in a true sliding window fashion (recall Section \ref{sec:windows}) and perform a change detection test periodically.

### 2020-07-04 chapters/problem-statement.tex
SAMM

@@
 Many real-time stream monitoring systems are static once deployed in a production environment.
-The engineers of those systems configure them under the assumption that future data flowing through the system follows roughly the same distribution as previously seen data.
-They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system.
-Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.
-The problem at hand is to determine when to reconfigure the system, \textit{e.g.} a Machine Learning model, based on an analysis of the drifts in the stream of data.
+These systems are configured \textit{a priori} and make assumptions about future streaming data based on the observable past events.
+The problem is that if these assumptions no longer hold in the future, the stream monitoring systems' performance decays.
+Thus, the problem at hand is to determine when to reconfigure the system based on an analysis of the drifts in the stream of data.
+For example, consider a Machine Learning (ML) model that monitors a data stream of credit card transactions and detects fraud.
+The ML model trains on a reference period and learns from this training period, \textit{i.e.}, makes assumptions based on the analyzed data.
+In streaming, throughout time, the data patterns will change and the assumptions made will no longer hold (\textit{e.g.}, buying patterns change on holidays), which leads to a decay in model performance.
+Our objective is to alert data pattern shifts.
+Our motivation is doing it so that system administrators can reconfigure the system before the system's performance decays.
+In this example, system reconfiguration is tantamount to model retraining.
 
@@
 \begin{itemize}
-    \item \textbf{Low memory consumption:} need for a low memory footprint system that avoids linear growth relative to the sliding window size used
+    \item \textbf{Low memory consumption:} need for a low memory footprint system that avoids linear growth relative to the sliding window size used;
 
-    \item \textbf{Low time complexity:} need for a time-efficient method that updates the sliding window aggregation and raises alerts in constant time
+    \item \textbf{Low time complexity:} need for a time-efficient method that updates the sliding window aggregation and raises alerts in constant time;
 
-    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout runtime
+    \item \textbf{Fixed reference period:} the reference window must be kept the same throughout runtime;
 
-    \item \textbf{Sliding window maintainability:} possibility to evict old events and insert new ones while updating the aggregation state
+    \item \textbf{Sliding window maintainability:} possibility to evict old events and insert new ones while updating the aggregation state;
 
-    \item \textbf{Explainable alerts:} have an interpretable measure of divergence to understand why an alert was raised
+    \item \textbf{Explainable alerts:} have an interpretable measure of divergence to understand why an alert was raised.
 \end{itemize}

### 2020-07-04 chapters/sota.tex
SAMM

@@
 
-Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy} \cite{Blazquez-Garcia-Review-Anomaly-Detection}) .
+Univariate point outlier detection methods are further broken down into three categories: model-based, density-based and histogram-based (Figure \ref{fig:point-outlier-taxonomy}) .
 \begin{figure}[!htb]
@@
     \includegraphics[scale=1]{figures/taxonomy-point-outlier-uni.png}
-    \caption{Point outlier detection in univariate time-series taxonomy}
+    \caption[Point outlier detection in univariate time-series taxonomy]{Point outlier detection in univariate time-series taxonomy \cite{Blazquez-Garcia-Review-Anomaly-Detection}}
     \label{fig:point-outlier-taxonomy}
@@
     \includegraphics[scale=0.8]{figures/taxonomy-subsequence-outlier-uni.png}
-    \caption{Subsequence outlier detection in univariate time-series taxonomy}
+    \caption[Subsequence outlier detection in univariate time-series taxonomy]{Subsequence outlier detection in univariate time-series taxonomy \cite{Blazquez-Garcia-Review-Anomaly-Detection}}
     \label{fig:subsequence-outlier-taxonomy}
@@
 
-\subsubsection{A two window paradigm algorithm for change detection}
+\subsubsection{A two window paradigm algorithm for change detection} \label{kifer}
 Kifer \emph{et al.} propose that detecting change in data streams can be reduced to testing if two windows have different underlying distributions \cite{Kifer-Detecting-Change}.
@@
 
-%\subsubsection{Automatic Model Monitoring for Data Streams}
-%ALSO a 2 WINDOW MODEL
+\subsubsection{SAMM: Automatic Model Monitoring for Data Streams}
+
+Sampaio et al.
+present SAMM \cite{SAMM}, an automatic model monitoring system for data streams.
+Similarly to the method presented in Section \ref{kifer}, SAMM uses two windows: the reference window R and the target window T.
+Window T contains the last $n_t$ elemnts and window R the elements of a reference period, which is prior to T.
+SAMM has two main components that interest us: the one responsible for computing the signal and the one that defines the signal threshold.
+
+The Machine Learning (ML) model being monitored outputs a series of scores for each transaction.
+The authors build two histograms: one with the model scores for the reference window R and another for the target window T.
+The signal is a measure of similarity between both histograms $H_R$ and $H_T$.
+The authors considered using the Kolmogorov-Smirnov, Kuiper and Anderson-Darling test but ultimately used the Jensen-Shannon Divergence (JSD) \cite{JSD} to measure similarity between both histograms.
+The JSD outputs a value between 0 and 1, with totally different histograms being assigned a value of 1.
+The signal is then computed by applying the JSD metric to both histograms $H_R$ and $H_T$ at each new event.
+
+To compute the threshold for the JSD signal, the authors maintain an histogram of observed JSD values during streaming.
+To efficiently maintain the JSD histogram, they devise an algorithm called SPEAR which stands for Streaming Percentiles EstimAtoR.
+SPEAR is linear in time and space complexity regarding the number of histogram bins, which is a small and fixed constant.
+This way, for a new JSD value measured between the histogram $H_R$ and histogram $H_T$, they compare it with the histogram and alert if above a certain percentile (\textit{e.g.}, the 99th percentile).
+
+\subsubsection*{Method classification}
+SAMM can take uni or multivariate time-series as input but performs a univariate analysis.
+This method is also a two-windowed method that measures the difference between reference and target sliding windows.
+The alerts produced report the entire target window as anomalous.
+According to the taxonomy presented in Figure \ref{fig:subsequence-outlier-taxonomy}, we classify this method as a dissimilarity-based univariate subsequence outlier detection method.
+
+\subsubsection*{Time complexity}
+For each new event, SAMM updates the target window T, updates the target histogram $H_T$, computes the JSD between $H_T$ and $H_R$, checks if this value is above a certain percentile and finally updates the histogram of JSDs with the new one.
+
+Maintaining the target window, \textit{i.e.}, evicting old events and inserting new ones, updating the target histogram $H_T$, computing the JSD between $H_T$ and $H_R$ and checking if it is above a given percentile are all done in constant time.
+Updating the JSD histogram and dinamically resizing the bins with the SPEAR algorithm takes $O(n)$ time with $n$ as the number of bins.
+Because the number of bins is small and constant, the update of the JSD histogram is also done in constant time.
+
+Overall, SAMM can process each event in constant time and is suitable for data stream monitoring.
+
+\subsubsection*{Space complexity}
+SAMM keeps in memory a reference window, a target window, a reference histogram, a target histogram and a JSD histogram.
+The histograms can have a small and fixed number of bins so they are not the dominant factor contributing to SAMM's space complexity.
+Instead, because SAMM keeps two windows in memory, it stores all the events belonging to them.
+Assuming $n_t$ as the size of target window and $n_r$ as the size of the reference window, SAMM has $O(n_r + n_t)$ space complexity.
+
+\subsubsection*{Applicability to our Hypothesis}
+SAMM was designed to work for small time periods.
+Hence, for the authors, keeping the reference and target windows in memory was no issue.
+However, in our solution, we need sublinear time and space complexities to handle large windows of data.
+This implies we can not directly use SAMM to solve our problem.
+
+However, SAMM has interesting properties we can use in our final solution.
+First, SAMM also uses a two-window paradigm: a reference and a target sliding one.
+This is similar to the method seen in Section \ref{kifer}, which may indicate this is a valid and tested approach to follow.
+Additionally, SAMM aggregates the contents of both windows as histograms and utilizes a statistical divergence test, the Jensen-Shannon Divergence, to measure the similarity (or dissimilarity) between both.
+This gives them a measure of divergence which they use to raise alerts, which is another great idea to test.
+Finally, and contrarily to the previous method (Section \ref{kifer}), SAMM does not simply threshold this divergence value: it builds a distribution of divergence values and measures percentiles for new event's divergence values.
+This allows a user to easier define a threshold: instead of specifying a threshold between 0 and 1 for the JSD value, the user defines only a probability value, \textit{e.g.}, alert when the probability for a JSD value is equal or less than 1\% (percentile 99th).
+
+Even though SAMM is not directly applicable in our context, it provides us with valid and promissing ideas to test.
 
@@
 As discussed in Section \ref{sec:back-swag-algs}, a sliding window aggregation algorithm allows the computation of an exact or approximate aggregation over a sliding window.
+We want to find an aggregation that is suitable to encode our reference and streaming periods so we avoid keeping the entire sliding window in memory.
+Furthermore, we need an efficient algorithm to incrementally maintain this aggregation.
 We have presented Recalculate-From-Scratch (RFS) and Subtract-On-Evict (SOE) as the two most basic generic stream processing algorithms.
@@
       \includegraphics[scale=0.6]{figures/daba-ds.png}
-      \caption{DABA data structure}
+      \caption[DABA data structure]{DABA data structure \cite{Tangwongsan-DABA}}
       \label{fig:daba-ds}

### 2020-07-04 chapters/validation.tex
SAMM

@@
 
-\subsection{Single Feature Analysis}
+\subsection{Single-Feature Analysis}
 As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
@@
 
-\subsection{Multi Feature Analysis}
+\subsection{Multi-Feature Analysis}
 Before moving to real data, we performed one last experiment, using synthetic data but this time performing a multi-feature analysis.

### 2020-07-04 chapters/abstract.tex
Small corrections

@@
 In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data.
-We develop a two-phased and two-windowed method: first we perform a batch analysis on a reference window and then a stream analysis over a sliding streaming window.
+We develop a two-phased and two-windowed method: first, we perform batch analysis on a reference window and then a stream analysis over a sliding streaming window.
 We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages.
-We do this for each event and for each of its fields, also denominated features in our context.
+We do this for each event and each of its fields also denominated features in our context.
 We measure the divergence between both reference and sliding window histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used.
@@
 We further split the tests into single and multi-feature analysis.
-Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi feature analysis.
+Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi-feature analysis.
 However, experiments with real data were not as accurate.
-Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypothesis that might explain it and hence are worthy of pursuing next.
+Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next.
 Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity regarding window size.

### 2020-07-04 chapters/background.tex
Small corrections

@@
 
-In this chapter we discuss fundamental concepts that will help us identify the set of relevant techniques that are suitable to address the problem to be formulated and tackled in this thesis.
+In this chapter, we discuss fundamental concepts that will help us identify the set of relevant techniques that are suitable to address the problem to be formulated and tackled in this thesis.
 In Section \ref{sec:stream-superset} we introduce batch and stream processing as two large scale data processing techniques.
@@
 A data stream is essentially a continuous or unbounded dataset.
-This data needs to be processed sequentially and incrementally on a event-by-event basis.
+This data needs to be processed sequentially and incrementally on an event-by-event basis.
 Streaming data varies widely, from logs generated by customers using mobile or web applications, sensor data from Internet-of-Things (IoT) devices to credit card transactions.
@@
 This approach handles streaming data by grouping incoming events into very small batches.
-Then, in order to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations.
+Then, to process the entire defined window of data, it joins all micro-batches into a larger one and performs the desired computations.
 However, this creates an artificial barrier that does not truly exist in the streaming context.
@@
 Using these definitions, the concept of an unbounded data stream contains that of a bounded dataset.
-That is, from an infinite stream, it is possible to fix a left and right limits in order to obtain a bounded and finite dataset.
+That is, from an infinite stream, it is possible to fix a left and right limits to obtain a bounded and finite dataset.
 This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.
@@
 Computations over an unbounded dataset are possible.
-For example, computing the \textit{average} of an event's field for all incoming events is computationally trivial: we need only to store the total number of events and the current total sum of all events arriving to the system.
+For example, computing the \textit{average} of an event's field for all incoming events is computationally trivial: we need only to store the total number of events and the current total sum of all events arriving in the system.
 However, is this boundless average over the stream of events useful?
@@
 The average temperature of a machine over an infinite period (\textit{i.e.} since the machine was deployed) is of little value for detecting if the equipment is currently overheating, and to trigger the appropriate action to prevent a failure.
-In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for more actionable context, in this case identifying a likely-to-fail rack.
+In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for a more actionable context, in this case identifying a likely-to-fail rack.
 To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}.
@@
 The three most common ways are: true sliding, stepping or tumbling windows.
-A true sliding window is a window with an unitary step, i.e., where $S = 1$.
+A true sliding window is a window with a unitary step, i.e., where $S = 1$.
 At each step, the window advances one unit, which may be one event for tuple-based steps, or one time-unit whether that is in milliseconds, seconds, days, weeks or another magnitude of time, measured based on the system clock or event time.
@@
 
-On the previous Section, we presented Subtract-On-Evict as an improvement over Recalculate-From-Scratch based on the idea of aggregation invertibility.
+In the previous section, we presented Subtract-On-Evict as an improvement over Recalculate-From-Scratch based on the idea of aggregation invertibility.
 The question then becomes if there are other aggregation properties besides invertibility that must be taken into account when analyzing general stream processing algorithms.
@@
 We talked about sliding window aggregations and algorithms to compute them.
-We observed that the contents of the sliding window must be stored in memory in order to evict the oldest event and discount its effect from the aggregation when a new one is inserted.
-In this Section we analyze Exponential Moving Averages (EMAs).
+We observed that the contents of the sliding window must be stored in memory to evict the oldest event and discount its effect from the aggregation when a new one is inserted.
+In this section, we analyze Exponential Moving Averages (EMAs).
 EMAs process each event only upon insertion.
@@
 
-Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB} are a type of reduce functions applied to a sequence of elements, one at a time in a orderly fashion, to compute a weighted average, using different weights for each element.
+Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB} are a type of reduce functions applied to a sequence of elements, one at a time in an orderly fashion, to compute a weighted average, using different weights for each element.
 These weights decrease exponentially with the age of the observations.
@@
 Figures \ref{fig:window-weights} and \ref{fig:ema-weights} show the weights applied to past events.
-In a typical sliding window, events either have weight of one (inside the window) or zero (outside the window) whereas EMAs give exponentially smaller weights to older events.
+In a typical sliding window, events either have a weight of one (inside the window) or zero (outside the window) whereas EMAs give exponentially smaller weights to older events.
 
@@
 Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones.
-In this Thesis we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.
+In this Thesis, we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.
 
@@
 While sliding window aggregation algorithms keep the window contents in memory, Exponential Moving Averages, presented in Section \ref{sec:emas}, replace them for a lighter representation, providing an approximate estimation of the aggregation value.
+
+In the next chapter, we review several outlier detection methods for time-series data and review state of the art sliding window aggregations and algorithms to summarize our data windows in a lighter representation.

### 2020-07-04 chapters/conclusions.tex
SAMM
Small corrections

@@
 
-In Section \ref{sec:hypothesis} we defined our hypothesis or main question to address as whether sliding window aggregations and outlier detection methods combined could be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.
+In Section \ref{sec:hypothesis} we defined our hypothesis or main question to address as to whether sliding window aggregations and outlier detection methods combined could be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.
 Before discussing our hypothesis, we revisit the research questions posed in Section \ref{sec:rqs} and provide answers to them.
@@
 
-This research question really is about two things: maintaining the sliding window aggregation in real-time and keeping its memory usage constant.
+This research question is about two things: maintaining the sliding window aggregation in real-time and keeping its memory usage constant.
 
-The aggregation in question is an histogram aggregation.
-An histogram requires storing a fixed-size list of bins and corresponding counts, regardless of the volume of data.
-This means that for ever growing volumes of data our memory consumption stays the same.
+The aggregation in question is a histogram aggregation.
+A histogram requires storing a fixed-size list of bins and corresponding counts, regardless of the volume of data.
+This means that for ever-growing volumes of data our memory consumption stays the same.
 
-However, maintaining a sliding window histogram in real-time is an harder challenge.
+However, maintaining a sliding window histogram in real-time is a harder challenge.
 Traditionally, sliding window maintenance requires evicting the oldest element from the window when a new one is inserted and updating the aggregation state.
@@
 Statistical tests that measure the similarity between two probability distributions are common in the literature.
-Tests like Kolmogorov–Smirnov, Wassertein or Jensen–Shannon (the one we used) provide a bounded measure of distance between two probability distributions, or histograms.
+Tests like Kolmogorov–Smirnov, Wasserstein or Jensen–Shannon (the one we used) provide a bounded measure of distance between two probability distributions, or histograms.
 
-In brief, since the aggregation we use is an histogram, we can use statistical tests that measure similarity between probability distributions.
+In brief, since the aggregation we use is a histogram, we can use statistical tests that measure the similarity between probability distributions.
 We chose to use the Jensen–Shannon Divergence (JSD).
@@
 Given a new JSD value, we check against the now known distribution of JSD values and alert if above a certain percentile.
-Note that defining a threshold percentile is easier, because we are defining a probability: alerting above the 99th percentile means alerting for hypothesis with 1\% probability.
+Note that defining a threshold percentile is easier because we are defining a probability: alerting above the 99th percentile means alerting for hypotheses with 1\% probability.
 
@@
 Additionally, we were able to do it achieving very high streaming throughputs.
-In the real data experiments, once again, we managed to achieve high troughputs but the produced alerts were not very accurate.
-We defer further investigation to future work because of the time constraints of this thesis but leave open the possiblity that the approximate aggregation value is more inaccurate than what we thought it would be, due to the fact that the EMA tail holds all past events.
+In the real data experiments, once again, we managed to achieve high throughput but the produced alerts were not very accurate.
+We defer further investigation to future work because of the time constraints of this thesis but leave open the possibility that the approximate aggregation value is more inaccurate than what we thought it would be since the EMA tail holds all past events.
 
@@
 
-    \item an histogram aggregation based on Exponential Moving Averages, that replaces the entire sliding window contents by a lighter representation, resulting in a constant in time and space complexity sliding window aggregation
+    \item a histogram aggregation based on Exponential Moving Averages, that replaces the entire sliding window contents by a lighter representation, resulting in a constant in time and space complexity sliding window aggregation
 
@@
 First and foremost, we believe the next steps are to reason about the obtained results in the real data experiments and find the root cause for the existence of many and large alert periods.
-As possible hypothesis to test, we suggest testing with EMAs with higher smoothing factors, so that past events are forgotten faster; additionally, we suggest decreasing $\gamma$, which further increases the number of samples to make in the batch phase.
+As a possible hypothesis to test, we suggest testing with EMAs with higher smoothing factors, so that past events are forgotten faster; additionally, we suggest decreasing $\gamma$, which further increases the number of samples to make in the batch phase.
 
-In this thesis we used the Jensen–Shannon Divergence (JSD) as statistical test to measure the distance or divergence between our reference and target sliding window aggregation.
+In this thesis, we used the Jensen–Shannon Divergence (JSD) as a statistical test to measure the distance or divergence between our reference and target sliding window aggregation.
 We chose to use the JSD metric due to its simplicity and because it was used in previous work successfully \cite{SAMM}.
 We believe that future work comprises testing new distances in the literature.
-In particular, we believe the Wassertein distance is worth trying next, and illustrate why in Figure \ref{fig:wasserteinvsjsd}.
-\textcolor{red}{TODO: make this image better}
+In particular, we believe the Wasserstein distance is worth trying next, and illustrate why in Figure \ref{fig:wasserteinvsjsd}.
 \begin{figure}[!htb]
@@
 \end{figure}
-The Wassertein distance measures the work required to transport the probability mass from one histogram to the other.
+The Wasserstein distance measures the work required to transport the probability mass from one histogram to the other.
 This measure will be larger the further way the probability mass is.
-In Figure \ref{fig:wasserteinvsjsd} scenario, while in the JSD measure is the same for both cases, the Wassertein distance is larger in case 2 because the probability mass has to be transported further.
+In Figure \ref{fig:wasserteinvsjsd} scenario, while in the JSD measure is the same for both cases, the Wasserstein distance is larger in case 2 because the probability mass has to be transported further.
 
-In this thesis we had to deal with the multiple test problem and apply a multiple test correction.
-In our case, we used Holm-Bonferroni due to its simplicity and because it controls the family wise error rate with a larger power than some other methods (thus providing a good trade-off between simplicity and power).
+In this thesis, we had to deal with the multiple test problem and apply a multiple test correction.
+In our case, we used Holm-Bonferroni due to its simplicity and because it controls the family-wise error rate with larger power than some other methods (thus providing a good trade-off between simplicity and power).
 There are many other multiple test correction methods, and we leave the testing of each as future work.

### 2020-07-04 chapters/introduction.tex
Small corrections

@@
 
-Thus the objective of this Thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.
+Thus the objective of this thesis is to build a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.
 
@@
 Therefore, it is common for companies to have auxiliary systems monitoring their services, testing their usage rate, availability and performance, just to name a few.
-The motivation behind this Thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in performance of the system analyzing said data stream.
+The motivation behind this thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in the performance of the system analyzing said data stream.
 
@@
 \section{Hypothesis} \label{sec:hypothesis}
-The question this Thesis aims to answer is whether \textit{sliding window aggregations and outlier detection methods combined can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.} In this Thesis, we study sliding window aggregations that replace the sliding window events in memory by a lighter representation, while still being able to provide an estimate of the aggregation value.
+The question this thesis aims to answer is whether \textit{sliding window aggregations and outlier detection methods combined can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.} In this thesis, we study sliding window aggregations that replace the sliding window events in memory by a lighter representation, while still being able to provide an estimate of the aggregation value.
 
@@
 
-In Chapter \ref{chap:sota}, we examine state of the art methods related with the previously mentioned topics and discuss shortcomings, strengths and the applicability of such methods to our hypothesis.
+In Chapter \ref{chap:sota}, we examine state of the art methods related to the previously mentioned topics and discuss shortcomings, strengths and the applicability of such methods to our hypothesis.
 
@@
 In Chapter \ref{chap:validation}, we describe the datasets used and the set of experiments that were performed to validate our method.
-We start with single feature experiments on synthetic data, then multi feature analysis and finally move on to multi feature analysis on real data.
+We start with single feature experiments on synthetic data, then multi-feature analysis and finally move on to multi-feature analysis on real data.
 

### 2020-07-04 chapters/method.tex
Small corrections

@@
 \cite{MultiTestProblem-Miller1966}  \cite{multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
-In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the amount of false negatives while having as target to control the amount of false positives.
-Furthermore, it does not require to assume any indepedence between the tested hypothesis (as other methods do) which would almost surely be a wrong assumption for realistic datasets.
+In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives.
+Furthermore, it does not require to assume any independence between the tested hypothesis (as other methods do) which would almost surely be a wrong assumption for realistic datasets.
 Therefore we leave as future work the testing of other correction methods.

### 2020-07-04 chapters/problem-statement.tex
Small corrections

@@
 
-With this chapter, we further discuss the problem we aim to solve with this Thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.
+With this chapter, we further discuss the problem we aim to solve with this thesis making use of concepts discussed in Chapter \ref{chap:background} and explaining why none of the reviewed methods from Chapter \ref{chap:sota} works for our use case.
 
-\section{The Problem}
+\section{The Problem} \label{sec:theproblem}
 Many real-time stream monitoring systems are static once deployed in a production environment.
@@
 
-\section{State of the Art Issues}
+\section{State of the Art Issues} \label{sec:sotaissues}
 Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques required to monitor streaming data patterns in real-time.
@@
 
-\section{Proposal}
+\section{Proposal} \label{sec:proposal}
 %2 phases, names and purpose
@@
 %aggregations to represent distributions
-We choose to use an histogram aggregation with constant memory usage to encode the distribution of the observed values.
+We choose to use a histogram aggregation with constant memory usage to encode the distribution of the observed values.
 Furthermore, we devise a sliding window approximate histogram aggregation that is maintainable in constant time in a sliding window fashion, through the use of Exponential Moving Averages (EMAs).
-Hence, our EMA-like histogram aggregation is ideal for the streaming scenario because of its constant time and space complexity, for ever growing volumes of data.
+Hence, our EMA-like histogram aggregation is ideal for the streaming scenario because of its constant time and space complexity, for ever-growing volumes of data.
 
@@
 We provide a measure of how divergent each field is.
-Based on this information, it is our belief that a system administrator would have an easier time determining when to reconfigure the system.
+Based on this information, we believe that a system administrator would have an easier time determining when to reconfigure the system.
 
-\section{Assumptions}
+\section{Assumptions} \label{sec:assumptions}
 Our feature monitoring method assumes only that the monitored features are numerical or categorical and that the events arrive in an orderly fashion during streaming.
@@
 
-\begin{enumerate}[leftmargin=1.75cm, label=\textbf{(RQ\arabic*)}]
-    \item What aggregation can we use to properly encode data distributions?
-    \item How can we maintain such an aggregation in a sliding window fashion in real-time with constant memory usage?
-    \item How can we measure divergence between the reference and sliding window aggregations?
-    \item Based on a divergence measure, when should we raise an alert?
+\begin{enumerate}[leftmargin=1.6cm, label=\textbf{(RQ\arabic*)}]
+    \item \textbf{What aggregation can we use to properly encode data distributions?}
+
+    We need an aggregation to reduce the entire window contents to a lighter representation and we need one that encodes the distribution of feature values correctly so we can measure the difference between the reference aggregation and the target sliding window aggregation accurately.
+
+    \item \textbf{How can we maintain such an aggregation in a sliding window fashion in real-time with constant memory usage?}
+
+    We need an aggregation that shows constant space complexity so that we can scale the size of our windows as we wish without increasing memory usage.
+Additionally, we need to find a way of incrementally maintaining this aggregation in real-time.
+
+    \item \textbf{How can we measure divergence between the reference and sliding window aggregations?}
+
+    After we aggregate the contents of the reference and target sliding windows, we need a function that takes both aggregation states and returns a measure of divergence.
+
+    \item \textbf{Based on a divergence measure, when should we raise an alert?}
+
+    Hard-coding a threshold value for this divergence measure is cumbersome, impractical and error-prone.
+In other words, we need to find a way of automatically defining the divergence threshold that we use to raise alerts.
+
 \end{enumerate}
+
+\section{Summary}
+In Section \ref{sec:theproblem}, we formally defined the problem we aim to solve with this thesis: alert streaming data pattern shifts in regards to a static reference period with the intent of helping system administrators reconfigure their systems before their performance declines.
+Furthermore, we make explicit we work under time and space constraints, inherent to the streaming paradigm.
+
+In Section \ref{sec:sotaissues}, we summarize the main identified issues regarding state of the art solutions and in Section \ref{sec:proposal} we detail how we propose to implement our solution along with the assumptions made (Section \ref{sec:assumptions}).
+
+In Section \ref{sec:hypothesis}, we defined our hypothesis or main question to address as to whether sliding window aggregations and outlier detection methods combined could be used to detect data pattern shifts in data streaming scenarios in real-time and with a low memory footprint.
+To better discuss our hypothesis, we pose a series of research questions in Section \ref{sec:rqs}.
+
+In Chapter \ref{chap:my-work}, we discuss initial approaches we thought of and reasons to why we discarded them.
+Finally, we present and detail each phase and component of our stream monitoring method.

### 2020-07-04 chapters/resumo.tex
Small corrections

@@
 
-A deteção de mudanças de padrões nos dados tem sido investigada nas mais diversas áreas e aplicações e diz respeito ao processo de encontrar padrões nos dados que não correspondam ao esperado ou usual.
-A deteção atempada e correta destas mudanças de padrões permite que medidas imediatas sejam tomadas, prevenindo perdas de rendimento e mantendo a reputação da empresa entre os seus clientes.
-Como tal e tendo isto em conta, muitas empresas constroem sistemas de deteção de anomalias.
-Desta maneira, administradores do sistema são capazes de proactivamente ajustar e prevenir potenciais falhas nos seus serviços.
-
-Por exemplo, o sistema de deteção de anomalias da Yahoo serve o propósito de monitorizar e alertar \textit{streams} de eventos temporais em múltiplos casos de uso da Yahoo.
-Na Microsoft, o seu sistema interno de deteção de anomalias monitoriza milhões de métricas provenientes de diferentes serviçoes internos, permitindo que os engenheiros os resolvam em tempo real antes que estes problemas adquiram proporções maiores.
-
-Muitos sistemas de monitorização em tempo real são estáticos quando implementados num ambiente de produção.
-Os engenheiros responsáveis por tais sistemas configuram-nos assumindo que os dados que futuramente fluirão pelo sistema se assemelham aos dados observados até ao momento.
-Eles fazem-no conscientemente e o melhor que sabem com as ferramentas de que dispoem, mantendo sempre em mente que no futuro possivelmente terão de reconfigurar o sistema.
-Assim, apesar da configuração inicial do sistema ser a que melhor se adequa, ao longo do tempo, devido a desvios nos padrões de dados, o desempenho do sistema estático incialmente implementado decai gradualmente.
-
-Desta forma, pretendemos avaliar o uso de métodos de análise de \textit{streams} eficientes em termos de memória --- \textit{i.e} métodos que usam pouca memóra e calculam os resultados esperados em tempo real --- com o intuito de detetar desvios nos padrões de dados na presença de grandes volumes de dados com padrões sazonais que fluem pelo sistema a grande velocidade.
-
 \vspace*{10mm}\noindent
-\textbf{Keywords}: tempo real, streaming, leves, monitorização, volume, velocidade, variedade, sazonalidade
+\textbf{Keywords}:

### 2020-07-04 chapters/sota.tex
Small corrections

@@
 Lastly, time-series outliers refer to entire time-series being outliers themselves.
-This type of outliers can only be detected if the input data is multivariate, so as to be able to compare multiple time-series.
+This type of outliers can only be detected if the input data is multivariate, to be able to compare multiple time-series.
 
@@
 
-Track changes is off
-Everyone
-Track changes for everyone
-You
-Track changes for You
-Guests
-Track changes for guests
-pedro.cardoso: How is this error calculated?
-Jun 18, 2020 2:52 PM
-You: Same as above
-Jun 20, 2020 5:09 PM • Edit • Delete
-Hit Enter to reply
-pedro.cardoso: This 3rd option is under-explained compared to the other 2. Please elaborate
-Jun 18, 2020 2:53 PM
-You: Not sure myself, that is why is so poorly written. Im going to revisit this
-Jun 19, 2020 4:42 PM • Edit • Delete
-Hit Enter to reply
-pedro.cardoso: Thought: This reads like leftover from the original work. Is there a way to better relate your investigation into SWAGs into the methodology you ended up taking?
-Jun 30, 2020 10:19 AM
-pedro.cardoso: Please revisit this entire section. Particularly the applicability sections of each algorithm in the context of your current approach.
-Jun 30, 2020 10:20 AM
-Hit Enter to reply
-pedro.cardoso: Remove these TODOs. If you want keep a note of them elsewhere, but not as text render being rendered.
-Jul 3, 2020 5:28 PM
-pedro.cardoso: You will forget to do it with other tasks still needed to be done.
-Jul 3, 2020 5:29 PM
-Hit Enter to reply
-pedro.cardoso: "because they evolved their reference period"
-Jul 3, 2020 5:30 PM
-Hit Enter to reply
-pedro.cardoso: Where is Model Monitoring?
-Jul 3, 2020 5:31 PM
-Hit Enter to reply
-Current file
-Overview
-16
-
 Yang et al.
@@
 Online learning algorithms that adapt to data pattern shifts and eventually consider them regular patterns are not viable to our goals.
-In fact, we need a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again.
-Additionally, insights produced by machine learning models are not human-understandable, leaving the user confused on why an anomaly report was made.
+We need a monitoring system that alerts data pattern shifts continuously until they match the initial configuration again.
+Additionally, insights produced by machine learning models are not human-understandable, leaving the user confused about why an anomaly report was made.
 In conclusion, we do not deem HTM applicable to our solution.
@@
 These methods find the most unusual subsequence in a time-series.
-However, as the authors put it, this type of methods lack a reference or normality state and hence do not accurately report outliers.
+However, as the authors put it, these methods lack a reference or normality state and hence do not accurately report outliers.
 
@@
 Frequency-based univariate subsequence outlier detection methods also use a reference period and compute the frequency of each subsequence in the reference or normality period.
-Hence, a subsequence will be an outlier if it does not appear as frequently as expected in subsequent time periods.
+Hence, a subsequence will be an outlier if it does not appear as frequently as expected in subsequent periods.
 
@@
 Similarly to the method presented in Section \ref{kifer}, SAMM uses two windows: the reference window R and the target window T.
-Window T contains the last $n_t$ elemnts and window R the elements of a reference period, which is prior to T.
+Window T contains the last $n_t$ elements and window R the $n_r$ elements of a reference period, which is prior to T.
 SAMM has two main components that interest us: the one responsible for computing the signal and the one that defines the signal threshold.
@@
 The signal is a measure of similarity between both histograms $H_R$ and $H_T$.
-The authors considered using the Kolmogorov-Smirnov, Kuiper and Anderson-Darling test but ultimately used the Jensen-Shannon Divergence (JSD) \cite{JSD} to measure similarity between both histograms.
+The authors considered using the Kolmogorov-Smirnov \cite{KolomogorovSmirnov, Kolmogorov-Smirnov}, Kuiper \cite{KuiperTest} and Anderson-Darling \cite{AndersonDarling} test but ultimately used the Jensen-Shannon Divergence (JSD) \cite{JSD} to measure the similarity between both histograms.
 The JSD outputs a value between 0 and 1, with totally different histograms being assigned a value of 1.
@@
 
-To compute the threshold for the JSD signal, the authors maintain an histogram of observed JSD values during streaming.
+To compute the threshold for the JSD signal, the authors maintain a histogram of observed JSD values during streaming.
 To efficiently maintain the JSD histogram, they devise an algorithm called SPEAR which stands for Streaming Percentiles EstimAtoR.
@@
 Maintaining the target window, \textit{i.e.}, evicting old events and inserting new ones, updating the target histogram $H_T$, computing the JSD between $H_T$ and $H_R$ and checking if it is above a given percentile are all done in constant time.
-Updating the JSD histogram and dinamically resizing the bins with the SPEAR algorithm takes $O(n)$ time with $n$ as the number of bins.
+Updating the JSD histogram and dynamically resizing the bins with the SPEAR algorithm takes $O(n)$ time with $n$ as the number of bins.
 Because the number of bins is small and constant, the update of the JSD histogram is also done in constant time.
@@
 \subsubsection*{Applicability to our Hypothesis}
-SAMM was designed to work for small time periods.
+SAMM was designed to work for small-time periods.
 Hence, for the authors, keeping the reference and target windows in memory was no issue.
@@
 
-Even though SAMM is not directly applicable in our context, it provides us with valid and promissing ideas to test.
+Even though SAMM is not directly applicable in our context, it provides us with valid and promising ideas to test.
 
@@
 This method carefully handles expensive operations and spreads them along the execution.
-As a resut, the authors are able to turn the average-case behavior into the worst-case behavior.
+As a result, the authors are able to turn the average-case behavior into worst-case behavior.
 This is commonly known as \textit{de-amortization}, hence the name.
-The main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout run time.
+The main idea behind DABA's design is to avoid occasional expensive operations --- such as the flip operation in Two-Stacks --- by gradually executing them throughout runtime.
 
@@
 \subsection*{Time complexity analysis}
-Considering a Bloom filter with \textit{m} bits and \textit{k} hash functions, insertion has \textit{O(k)} time complexity because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1.
+Considering a Bloom Filter with \textit{m} bits and \textit{k} hash functions, insertion has \textit{O(k)} time complexity because all there is to do is run the input through all of the \textit{k} hash functions and set the bits in the given positions to 1.
 Similarly, query answering has \textit{O(k)} time complexity because it needs only to run the input through the \textit{k} hash functions and check if all bits are set to 1.
@@
 
-% FOR EACH SKETCH DO:
-% approx aggregation to compute
-% data structures
-% algorithm for insertion
-% algorithm for query
-% error bound related with parameters
-% time complexity analysis
-% space complexity analysis
-% Applicability to our Hypothesis
 \subsection*{HyperLogLog}
@@
 The authors claim that the proposed sliding version remains as accurate as of the original HLL aggregator.
-With an additional memory of only 35kB it is possible to estimate the number of distinct elements in the set with a standard error of 3\% in the presence of up to a million distinct elements.
+With an additional memory of only 35kB, it is possible to estimate the number of distinct elements in the set with a standard error of 3\% in the presence of up to a million distinct elements.
 
@@
 Despite the fact memory consumption dynamically changes during runtime, this is a valid approach to test our Hypothesis due to the worst-case scenario being a logarithmic space complexity.
-However, in Section \ref{sec:hll-drv} we analyze an implementation that is not dynamic in memory during run time making it a preferred approach.
+However, in Section \ref{sec:hll-drv} we analyze an implementation that is not dynamic in memory during runtime making it a preferred approach.
 
@@
 In Section \ref{sec:outliers}, we reviewed outlier detection algorithms.
-However, we found most of them inapplicable in our scenario, because they were resource intensive (both time and memory wise), provided no measure of how divergent and outlier was or because they evolved their reference period.
+However, we found most of them inapplicable in our scenario, because they were resource-intensive (both time and memory-wise), provided no measure of how divergent and outlier was or because they evolved their reference period.
 
@@
 We then discussed probabilistic data structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}).
-Despite being efficient, both time and memory wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use to us.
+Despite being efficient, both time and memory-wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use to us.
+
+In the next chapter, we formally define the problem we aim to solve and the issues with current state of the art solutions.
+Then, we formulate our proposal along with the assumptions we make when approaching this problem.
+Finally, we pose some research questions we aim to answer by the end of this thesis.

### 2020-07-04 chapters/validation.tex
Small corrections

@@
 In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
-In this section we present summary statistics for the synthetic datasets created and our generated alarms.
+In this section, we present summary statistics for the synthetic datasets created and our generated alarms.
 
@@
 As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
-In this section we validate our results for single-feature synthetic datasets.
+In this section, we validate our results for single-feature synthetic datasets.
 
@@
 
-We began by performing single-feature analysis on synthetic data.
+We began by performing a single-feature analysis of synthetic data.
 The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported.
@@
 In our last experiment, we attempted to reduce concept drift to a maximum by using the same period as reference and target.
-To our surprise, we still produced a large number of long lasting alarms.
-We defer the investigation of the root cause of the problem to future work and leave a few untested hypothesis as solutions for this issue, like the possibility that the EMA-like histograms in batch work with subsets of data and the streaming ones work with infinite amounts of data.
+To our surprise, we still produced a large number of long-lasting alarms.
+We defer the investigation of the root cause of the problem to future work and leave a few untested hypotheses as solutions for this issue, like the possibility that the EMA-like histograms in batch work with subsets of data and the streaming ones work with infinite amounts of data.
 In other words, for the streaming EMAs, the previous events never have their contribution set to zero, even though the weights tend to zero, whereas the batch EMAs worked with a finite dataset.
 Additionally, our sampling may have been insufficient.
-To test this, an experiment with lower $\gamma$ and hence higher number of samples should be carried out.
+To test this, an experiment with a lower $\gamma$ and hence a higher number of samples should be carried out.

### 2020-07-05 chapters/background.tex
Multiple grammar corrections

@@
 using a reducer function.
-To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms).
+To do this, the distributed computation framework reads and writes from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.}, iterative Machine Learning algorithms).
 Nearly a decade later, Apache Spark \cite{ApacheSpark, Spark} was released as the industry accepted successor.
@@
 Assume that temperatures measured across all racks are sent to a monitoring system in the form of a data stream.
-The average temperature of a machine over an infinite period (\textit{i.e.} since the machine was deployed) is of little value for detecting if the equipment is currently overheating, and to trigger the appropriate action to prevent a failure.
+The average temperature of a machine over an infinite period (\textit{i.e.}, since the machine was deployed) is of little value for detecting if the equipment is currently overheating, and to trigger the appropriate action to prevent a failure.
 In this scenario, an average of the temperatures per rack and for the last 5 minutes allows for a more actionable context, in this case identifying a likely-to-fail rack.
-To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.} all the data between the current timestamp in minutes \textit{t} and \textit{t-5}.
+To obtain such insight requires computing the average over the last 5 minutes' worth of data, effectively defining a begin and end --- \textit{i.e.}, all the data between the current timestamp in minutes \textit{t} and \textit{t-5}.
 By creating such a conceptual range, we realize that in practice we need a bounded dataset.
@@
 
-Botan et. al \cite{Botan-SECRET} define a window as:
+Botan \emph{et al.} \cite{Botan-SECRET} define a window as:
 \begin{definition}
@@
 The three most common ways are: true sliding, stepping or tumbling windows.
-A true sliding window is a window with a unitary step, i.e., where $S = 1$.
+A true sliding window is a window with a unitary step, \textit{i.e.}, where $S = 1$.
 At each step, the window advances one unit, which may be one event for tuple-based steps, or one time-unit whether that is in milliseconds, seconds, days, weeks or another magnitude of time, measured based on the system clock or event time.
@@
 
-For the scope of this Thesis, we will work under a tuple-based true sliding window framework.
+For the scope of this thesis, we will work under a tuple-based true sliding window framework.
 However, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.
@@
 When doing so, we obtain a finite dataset denoted as a sliding window and compute aggregations over it.
-Tangwongsan \emph{et.
-al} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+Tangwongsan \emph{et al.} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
 As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}.
@@
 Note that this is very similar to any aggregation made in a standard Database Management System (DBMS).
-The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.} in this case, a tumbling window).
+The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.}, in this case, a tumbling window).
 
@@
 \begin{algorithm}
-    \caption{Recalculate-From-Scratch insert, evict and query methods}
+    \caption[Recalculate-From-Scratch]{Recalculate-From-Scratch insert, evict and query methods}
     \label{pseudo:rfs}
@@
 \begin{algorithm}
-    \caption{Subtract-On-Evict insert, evict and query methods}
+    \caption[Subtract-On-Evict]{Subtract-On-Evict insert, evict and query methods}
     \label{pseudo:soe}
@@
 As it turns out, besides \textit{invertibility}, other properties allow us to group aggregations in different families.
-Tangwongsan et al.
-construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+Tangwongsan \emph{et al.} construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
 
@@
 For instance, consider once again the use case of computing a sum aggregation over a window of data.
-The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.} the current sum value --- with the new incoming event.
-The \textit{query} function would return the current aggregation state --- \textit{i.e.} return the total sum computed.
+The \textit{combine} function would be the arithmetic operator \textit{+} and would merge the current aggregation state --- \textit{i.e.}, the current sum value --- with the new incoming event.
+The \textit{query} function would return the current aggregation state --- \textit{i.e.}, return the total sum computed.
 
@@
 
-    \item  a \textbf{\textit{unary query}} if the query does not need any parameters --- \textit{e.g.} a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
+    \item  a \textbf{\textit{unary query}} if the query does not need any parameters --- \textit{e.g.}, a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
 \end{enumerate}
@@
 
-According to these properties, Tangwongsan et al.
-define five aggregation families \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
+According to these properties, Tangwongsan \emph{et al.} define five aggregation families \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
 These are the sum, collect, median, max and sketch-like families.
@@
 For example, keeping a string concatenation aggregation is invertible and associative but not commutative nor size-preserving.
-While in the string concatenation case you do not need to pass any arguments to the \textit{query} function, in the case of another collect-like aggregation, for instance, keeping track of the \textit{i-th} youngest element, you need to specify in your query the parameter \textit{i} (\textit{e.g.} the 5th youngest element).
+While in the string concatenation case you do not need to pass any arguments to the \textit{query} function, in the case of another collect-like aggregation, for instance, keeping track of the \textit{i-th} youngest element, you need to specify in your query the parameter \textit{i} (\textit{e.g.}, the 5th youngest element).
 Hence, their \textit{query} function is not always unary.
 The median-like family of aggregations contains for example the percentile operation.
-For example, the percentile aggregation is invertible, associative and commutative but not size-preserving and requires an argument to the \textit{query} function: the percentile to compute (\textit{e.g.} percentile 95 or 99).
+For example, the percentile aggregation is invertible, associative and commutative but not size-preserving and requires an argument to the \textit{query} function: the percentile to compute (\textit{e.g.}, percentile 95 or 99).
 The max-like family comprises aggregations like the maximum or minimum, operations that are not invertible but associative and size-preserving, as well as unary from a \textit{query} function perspective.
@@
 
-The aggregation properties analyzed in this Section will be used to evaluate SWAG algorithms.
+The aggregation properties analyzed in this section will be used to evaluate SWAG algorithms.
 Different algorithms might only apply to certain families of aggregations.
-This way, the multiple SWAG algorithms presented throughout this Thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.
+This way, the multiple SWAG algorithms presented throughout this thesis will be analyzed not only on their time and space complexity but also on the restrictions they impose regarding the aggregation' properties required.
 For example, under this framework, to implement Subtract-On-Evict (SOE), both the \textit{combine} $\oplus$ and its inverse $\ominus$ functions must be implemented so that SOE applies $\ominus$ to expire the evicted element from the aggregation state.
@@
 Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones.
-In this Thesis, we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.
+In this thesis, we chose to work with tuple-based EMAs due to its simplicity, since it does not require keeping track of timestamp variations, but our approach can be adjusted to use time-based ones.
 

### 2020-07-05 chapters/introduction.tex
Multiple grammar corrections

@@
 Cybersecurity applications are yet another example of time-sensitive, data-driven applications.
-The entities that work in these use-cases need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the underlying system.
+The entities that work in these use cases need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the underlying system.
 
@@
 This data is characterized by its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems.
-Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams.
+Additionally, the large amounts of information produced by such systems and use cases lead to the creation of multiple real-time unbounded datasets, also known as data streams.
 The information that flows through such a stream can be analyzed on-the-fly or stored for later processing.
@@
 However, due to the increasing volume of data and the need for timely processing to allow companies to react to changing conditions in real-time, the demand for stream processing systems is increasing.
-Unlike batch processing of static data sets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery} where the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}.
+Unlike batch processing of static datasets, where assumptions on the underlying data distributions can be made, streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery} where the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}.
 \textit{Recency} is measured under different scales for different use cases.
@@
 In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements.
-Operations that produce a single result when applied to data sets are commonly known as aggregation operations.
+Operations that produce a single result when applied to datasets are commonly known as aggregation operations.
 In stream processing, we apply these aggregations to data streams to obtain valuable information.
-However, since a data stream is, effectively speaking, an unbounded data set, if our aggregation requires a finite domain, there is the need to apply windowing techniques.
+However, since a data stream is, effectively speaking, an unbounded dataset, if our aggregation requires a finite domain, there is the need to apply windowing techniques.
 A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size --- \textit{e.g.}, events from the last three days or the last million events, respectively.
 
-Consider the use-case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day.
+Consider the use case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day.
 In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one.
@@
 Many market solutions monitor streams of credit card transactions with the intent of fraud detection.
-In the context of this Thesis, we work side by side with Feedzai\footnote{https://feedzai.com}.
+In the context of this thesis, we work side by side with Feedzai\footnote{https://feedzai.com}.
 In brief, the company's solution for fraud detection consists of a workflow of different components.
@@
 Ensuring constant time and space complexity means that the system will remain usable for the ever-growing volumes of data.
-Since such a monitoring system is not mission-critical --- \textit{i.e.} the provided services do not rely on it --- the operational cost associated with it must be kept to a minimum.
+Since such a monitoring system is not mission-critical --- \textit{i.e.}, the provided services do not rely on it --- the operational cost associated with it must be kept to a minimum.
 Having a real-time system monitoring other real-time systems consuming as many resources as the latter would be far too costly.
@@
 \end{itemize}
-The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic data sets and a real dataset from the financial fraud space.
+The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic datasets and a real dataset from the financial fraud space.
 

### 2020-07-05 chapters/method.tex
Multiple grammar corrections

@@
 \subsection*{Challenges}
-The analyzed outlier detection algorithms in Section \ref{sec:outliers} were deemed unfit for our use-case due to one or more of the following: \textit{(a)} high memory consumption, \textit{(b)} high time complexity, \textit{(c)} "memory loss" and \textit{(d)} obfuscated insights.
+The analyzed outlier detection algorithms in Section \ref{sec:outliers} were deemed unfit for our use case due to one or more of the following: \textit{(a)} high memory consumption, \textit{(b)} high time complexity, \textit{(c)} "memory loss" and \textit{(d)} obfuscated insights.
 
-For this Thesis, we need constant time complexity to process events one by one in real-time.
+For this thesis, we need constant time complexity to process events one by one in real-time.
 Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit.
@@
 In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream.
-While in some scenarios this is a much-needed feature, in our use-case it is not.
+While in some scenarios this is a much-needed feature, in our use case it is not.
 How clear the alerts and insights retrieved from the methods are, is also an important property.
@@
 Each event is timestamped and contains multiple fields.
-In the context of this Thesis, each of the event's fields are named features.
+In the context of this thesis, each of the event's fields are named features.
 Our goal is to detect data pattern shifts for each feature in a streaming fashion.
@@
       \includegraphics[scale=0.7]{figures/compute-sample-distances.png}
-      \caption[Compute as many distance values as samples]{Compute as many distance values as samples, one for each sample-reference histogram pair}
+      \caption[Compute sample's distance values]{Compute as many distance values as samples, one for each sample-reference histogram pair}
       \label{fig:compute-sample-distances}

### 2020-07-05 chapters/problem-statement.tex
Multiple grammar corrections

@@
 In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods.
-However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the desired characteristics, namely:
+However, the reviewed methods were deemed unfit for our use case because none of them simultaneously showed all of the desired characteristics, namely:
 
@@
 \section{Research Questions} \label{sec:rqs}
-The goal of this Thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time.
-To that end, we devise a series of research questions to be answered by the end of this Thesis:
+The goal of this thesis is to develop a lightweight stream monitoring system that detects changes in features distribution between a reference period and a sliding window in real-time.
+To that end, we devise a series of research questions to be answered by the end of this thesis:
 

### 2020-07-05 chapters/sota.tex
Multiple grammar corrections

@@
 \section{Outlier Detection in Time-Series Data} \label{sec:outliers}
-Blázquez-García \emph{et al.} review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
+Blázquez-García \emph{et al.} review state of the art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
 From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey, Aguinis-Outlier-survey, Chandola-Outlier-survey-2009, Hodge-Outlier-survey, Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.
@@
 generated by a different mechanism" \cite{Hawkins-Outliers}.
-Furthermore, Blázquez-García \emph{et al.} allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers.
+Furthermore, Blázquez-García \emph{et al.} allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this thesis all refer to outliers.
 In summary, outlier detection refers to the process of finding anomalies in data.
@@
 Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}.
-Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.
-
-Quoting the authors, histogram-based univariate point outlier detection methods are \textit{"based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original"}.
+Two points are said to be neighbors if they are close to each other --- \textit{i.e.}, the distance between them is less than or equal to a given threshold $\alpha$.
+
+The authors provide little detail on the last category, the histogram-based one.
+The authors state that these methods are ``based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original''.
 
@@
 
-Yang et al.
-propose \emph{Extra-N} \cite{Yang-Neighbor-Based-Pattern-Detection}, a method for incremental detection of point outliers in sliding window scenarios.
+Yang \emph{et al.} propose \emph{Extra-N} \cite{Yang-Neighbor-Based-Pattern-Detection}, a method for incremental detection of point outliers in sliding window scenarios.
 Data points are considered outliers if they have few neighbors, \textit{i.e.}, less than a user-defined threshold $\alpha$ of neighbors within a user-defined range $\theta$.
@@
     \includegraphics[scale=0.6]{figures/point-outlier.pdf}
-    \caption{Point outlier for $\alpha=2$ and a given $\theta$}
+    \caption[Density-based point outlier]{Density-based point outlier for $\alpha=2$ and a given $\theta$}
     \label{fig:point-outlier}
@@
 In other words, random variations in the underlying distribution of incoming data that do not persist throughout time will be ignored.
-Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.} changes in magnitudes of data or unusual timing for particular patterns, respectively.
+Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.}, changes in magnitudes of data or unusual timing for particular patterns, respectively.
 
@@
 \subsubsection{DeepAnT}
-Munir et.
-al present DeepAnT, a deep learning-based approach for the detection of anomalies in time-series data \cite{Munir-DeepAnT}.
+Munir \emph{et al.} present DeepAnT, a deep learning-based approach for the detection of anomalies in time-series data \cite{Munir-DeepAnT}.
 DeepAnT uses unlabeled data to capture and learn the distribution of the data stream used to forecast the normal behavior of the time-series.
-This method is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known.
+This method is unsupervised - \textit{i.e.}, it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known.
 It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.
@@
 Before monitoring a data stream, the CNN model must be trained.
-The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN.
-Additionally, the data set used to train the model does not require labels.
+The model can be trained with a small dataset while achieving good generalization capabilities due to the effective parameter sharing of the CNN.
+Additionally, the dataset used to train the model does not require labels.
 
 \subsubsection*{Method classification}
-DeepAnT takes as input a uni or multivariate time-series but is a univariate method by nature --- \textit{i.e.} it does not find correlations between multiple time-dependent variables.
+DeepAnT takes as input a uni or multivariate time-series but is a univariate method by nature --- \textit{i.e.}, it does not find correlations between multiple time-dependent variables.
 The model used by DeepAnT is a Convolutional Neural Network (CNN), used to make predictions on the future time-series values and measure differences with the actual observed values.
@@
 According to the authors, \textit{"I(S) is computed taking into account the number of times the symbols within S are repeated through the series"}.
-The authors explain that if \textit{F(S)} is large --- \textit{i.e.} S occurs frequently --- then \textit{I(S)} will be closer to 0.
+The authors explain that if \textit{F(S)} is large --- \textit{i.e.}, S occurs frequently --- then \textit{I(S)} will be closer to 0.
 
@@
 \subsubsection*{Applicability to our Hypothesis}
-In this Thesis, we want to detect and alert data pattern shifts with a low memory footprint.
-The method proposed by Kifer et.
-al grows linearly with window size, making it unfit for large windows, our use-case.
+In this thesis, we want to detect and alert data pattern shifts with a low memory footprint.
+The method proposed by Kifer \emph{et al.} grows linearly with window size, making it unfit for large windows, our use case.
 
 However, this two-window paradigm encodes a static normal or reference period in one window and the observed streaming values in a sliding one.
-Since in this Thesis we want to report streaming data pattern shifts relatively to a reference period, we use a similar two-windowed model, but use lightweight and incrementally maintainable aggregations over both windows.
+Since in this thesis we want to report streaming data pattern shifts relatively to a reference period, we use a similar two-windowed model, but use lightweight and incrementally maintainable aggregations over both windows.
 
@@
 
-Sampaio et al.
-present SAMM \cite{SAMM}, an automatic model monitoring system for data streams.
+Sampaio \emph{et al.} present SAMM \cite{SAMM}, an automatic model monitoring system for data streams.
 Similarly to the method presented in Section \ref{kifer}, SAMM uses two windows: the reference window R and the target window T.
@@
 We have presented Recalculate-From-Scratch (RFS) and Subtract-On-Evict (SOE) as the two most basic generic stream processing algorithms.
-In this Section we analyze other more sophisticated SWAG algorithms that fit different use-cases, restraining the set of computable aggregations to certain families, as defined in Table \ref{tbl:aggregations-properties}.
+In this section, we analyze other more sophisticated SWAG algorithms that fit different use cases, restraining the set of computable aggregations to certain families, as defined in Table \ref{tbl:aggregations-properties}.
 
 \subsection{Two-Stacks} \label{sec:2stacks}
-The Two-Stacks SWAG algorithm works for \textit{associative} aggregations (recall Table \ref{tbl:aggregations-properties}) in scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.} elements are evicted in the same order they were inserted.
+The Two-Stacks SWAG algorithm works for \textit{associative} aggregations (recall Table \ref{tbl:aggregations-properties}) in scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.}, elements are evicted in the same order they were inserted.
 Two-Stacks \cite{Tangwongsan-DABA} implements the FIFO queue using two stacks, namely the front stack F and the back stack B.
@@
 \begin{algorithm}
-    \caption{Two-Stacks insert, evict and query methods}
+    \caption[Two-Stacks]{Two-Stacks insert, evict and query methods}
     \label{pseudo:2-stacks}
@@
 Inserting a new element corresponds to pushing a new value \textit{val} in \textit{vals} and pushing the new aggregation value \textit{agg} in \textit{aggs}.
-The new aggregation value \textit{agg} is computed by applying the \textit{combine} function $\oplus$ to the back element of \textit{aggs} and the new value \textit{val} --- \textit{i.e.} pushing \textit{val} to \textit{vals} and $\textit{aggs[E]} \oplus \textit{val}$ to \textit{aggs}.
+The new aggregation value \textit{agg} is computed by applying the \textit{combine} function $\oplus$ to the back element of \textit{aggs} and the new value \textit{val} --- \textit{i.e.}, pushing \textit{val} to \textit{vals} and $\textit{aggs[E]} \oplus \textit{val}$ to \textit{aggs}.
 
@@
 
-\textit{l\textsubscript{F}} and \textit{l\textsubscript{B}} combined cover the whole \textit{vals} queue --- \textit{i.e.} every element in the sliding window.
+\textit{l\textsubscript{F}} and \textit{l\textsubscript{B}} combined cover the whole \textit{vals} queue --- \textit{i.e.}, every element in the sliding window.
 \textit{l\textsubscript{F}} is aggregated to the left, thus \textit{aggs[F]} contains the aggregation value of everything in \textit{l\textsubscript{F}}.
@@
 Algorithms \ref{pseudo:daba-helpers} and \ref{pseudo:daba} show how such reversal is done but we will not elaborate on such procedure.
-However, it is important to stress out that this is the core idea of DABA: avoid occasional expensive operations, like Two-Stacks' \textit{flip}, by spreading them out across run time, in the form of DABA's \textit{fixup}.
+However, it is important to stress out that this is the core idea of DABA: avoid occasional expensive operations, like Two-Stacks' \textit{flip}, by spreading them out across runtime, in the form of DABA's \textit{fixup}.
 
 \begin{algorithm}
-    \caption{DABA helper functions}
+    \caption[De-Amortized Banker’s Aggregator (DABA) helpers]{DABA helper functions}
     \label{pseudo:daba-helpers}
@@
 \begin{algorithm}
-    \caption{DABA insert, evict and query methods}
+    \caption[De-Amortized Banker’s Aggregator (DABA)]{DABA insert, evict and query methods}
     \label{pseudo:daba}
@@
 \section{Probabilistic Data Structures} \label{sec:pds}
-In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs).
+In Section \ref{sec:sota-swag-algs} we presented some state of the art algorithms for computing sliding window aggregations (SWAGs).
 However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods.
@@
 A Bloom Filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements.
-Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?".
-Being a probabilistic data structure and approximate aggregator, Bloom filters' answers have an error associated.
-\textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is.
-In other words, a Bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set.
+Bloom Filters answer the query: "Is element \textit{el} in the set of seen elements so far?".
+Being a probabilistic data structure and approximate aggregator, Bloom Filters' answers have an error associated.
+\textit{False positive} matches are possible --- \textit{i.e.}, saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.}, saying \textit{el} is not in the set but it actually is.
+In other words, a Bloom Filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set.
 Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.
@@
 \subsection*{Data structure}
-A Bloom filter is an array of bits of size \textit{m}, initially all set to 0.
-Figure \ref{fig:initial-bloom-filter} represents a Bloom filter with a bit array of size \textit{m}=7 in its initial state.
+A Bloom Filter is an array of bits of size \textit{m}, initially all set to 0.
+Figure \ref{fig:initial-bloom-filter} represents a Bloom Filter with a bit array of size \textit{m}=7 in its initial state.
 
@@
       \includegraphics[scale=0.9]{figures/initial-bloom.png}
-      \caption[Bloom filter initial state]{Bloom filter bit array initial state of \textit{m}=7}
+      \caption[Bloom Filter initial state]{Bloom Filter bit array initial state of \textit{m}=7}
       \label{fig:initial-bloom-filter}
@@
 The resultant \textit{k} outputs are used as the \textit{k} positions of the array to be set to 1.
-Figure \ref{fig:insertion-bloom-filter} shows the insertion of element \textit{el} in the Bloom filter, making use of three hash functions (\textit{k}=3), \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}}.
+Figure \ref{fig:insertion-bloom-filter} shows the insertion of element \textit{el} in the Bloom Filter, making use of three hash functions (\textit{k}=3), \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}}.
 The outputs of the hash functions correspond to the array indexes to set to 1.
@@
       \includegraphics[scale=0.3]{figures/insert-bloom.png}
-      \caption[Bloom filter insertion]{Insertion of element \textit{el} into a Bloom filter with \textit{m}=7 and \textit{k}=3}
+      \caption[Bloom Filter insertion]{Insertion of element \textit{el} into a Bloom Filter with \textit{m}=7 and \textit{k}=3}
       \label{fig:insertion-bloom-filter}
@@
 \subsection*{Membership Query}
-Performing a membership query --- \textit{i.e.} testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions.
+Performing a membership query --- \textit{i.e.}, testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions.
 If \textit{el} was already in the set, all \textit{k} positions should be 1.
@@
       \includegraphics[scale=0.3]{figures/query-bloom.png}
-      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
+      \caption[Bloom Filter membership query]{Bloom Filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
       \label{fig:bloom-filter}
@@
 If all positions are 1 then the element is said to be \textit{possibly in the set}.
-The reason a Bloom filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements.
+The reason a Bloom Filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements.
 For example, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ and an element \textit{el2} to positions $[4,5,6]$.
@@
 When querying about the membership of a new element \textit{el3}, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map it to positions $[1,3,5]$ as exemplified in Figure \ref{fig:bloom-filter-fp}.
-Since all of these positions are 1 because of the previous elements inserted, the Bloom filter returns \textit{"possibly in set"} when in reality \textit{el3} was not in the set.
+Since all of these positions are 1 because of the previous elements inserted, the Bloom Filter returns \textit{"possibly in set"} when in reality \textit{el3} was not in the set.
 This is considered a \textit{false positive}.
@@
       \includegraphics[scale=0.4]{figures/fp-bloom-3.png}
-      \caption{Bloom filter false positive}
+      \caption{Bloom Filter false positive}
       \label{fig:bloom-filter-fp}
@@
 
-Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula. The probability of one arbitrary bit being set to 1 is:
+Mullin presents an analysis of the Bloom Filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula. The probability of one arbitrary bit being set to 1 is:
 \begin{equation}
@@
 We want to choose the number of hash functions \textit{k} in order to minimize the false positive rate.
-In order to do so, we must first allocate \textit{m} bits for the Bloom filter, choosing \textit{m} based on available memory.
+In order to do so, we must first allocate \textit{m} bits for the Bloom Filter, choosing \textit{m} based on available memory.
 The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by:
@@
 
-Note that the time complexity does not at all depend on the number of elements processed by the Bloom filter and that \textit{k} will be a rather small constant.
-Hence, the Bloom filter is said to have constant time complexity.
+Note that the time complexity does not at all depend on the number of elements processed by the Bloom Filter and that \textit{k} will be a rather small constant.
+Hence, the Bloom Filter is said to have constant time complexity.
 
 \subsection*{Space complexity analysis}
-Considering a Bloom filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity.
+Considering a Bloom Filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity.
 Similarly to the time complexity analysis, we point out that \textit{m} will be constant and that each of the array elements occupies 1 bit.
-Thus the Bloom filter has a constant space complexity.
+Thus the Bloom Filter has a constant space complexity.
 
 \subsection*{Applicability to our Hypothesis}
-Bloom filters exhibit constant space and time complexity making them possible methods to test our hypothesis.
+Bloom Filters exhibit constant space and time complexity making them possible methods to test our hypothesis.
 Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}.
-However, we need to evict old events, and for that reason, bloom filters are of no use to us.
+However, we need to evict old events, and for that reason, Bloom Filters are of no use to us.
 
@@
 % approx aggregation to compute
-The Count-Min Sketch (CMS) uses two principles from the discussed Bloom filters in Section \ref{sec:bloom}.
-First, Bloom filters show that precision can be sacrificed to achieve space savings.
+The Count-Min Sketch (CMS) uses two principles from the discussed Bloom Filters in Section \ref{sec:bloom}.
+First, Bloom Filters show that precision can be sacrificed to achieve space savings.
 Second, they are very similar at a technical level.
 
-A Count–Min Sketch \cite{Cormode-CMS} is an approximate aggregator that estimates the frequency of each element in the data set.
+A Count–Min Sketch \cite{Cormode-CMS} is an approximate aggregator that estimates the frequency of each element in the dataset.
 It is named after the two basic operations used to provide the estimate: counting (\textit{count}) and computing the minimum (\textit{min}).
@@
 \subsection*{Applicability to our Hypothesis}
-Count-Min Sketches are constant in both time and space. Similarly to Bloom filters, this makes them ideal candidates for our solution.
+Count-Min Sketches are constant in both time and space. Similarly to Bloom Filters, this makes them ideal candidates for our solution.
 
@@
 
-A set of probabilistic counting algorithms is introduced by Flajolet et al.
-in \cite{Flajolet-PCA} to estimate the number of distinct elements of a set --- also known as the cardinality of the set.
+A set of probabilistic counting algorithms is introduced by Flajolet \emph{et al.} in \cite{Flajolet-PCA} to estimate the number of distinct elements of a set --- also known as the cardinality of the set.
 These techniques are based on \textit{"bit-pattern observables"} in the binary representation of the hashed values.
@@
 In LogLog, the bit-pattern observable recorded for each item is the position of the leftmost 1-bit, denoted as \textit{P}.
-The authors claim it is related to the total number of distinct elements in the data set in that it is \textit{"more or less a likely indication that the cardinality of S is at least $2^P$"}.
+The authors claim it is related to the total number of distinct elements in the dataset in that it is \textit{"more or less a likely indication that the cardinality of S is at least $2^P$"}.
 
@@
 \subsection*{Space complexity analysis}
-Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in the original HyperLogLog paper \cite{Flajolet-HLL}: 
+Quoting Flajolet \emph{et al.} about the memory consumption and error in the approximations obtained in their experiments in the original HyperLogLog paper \cite{Flajolet-HLL}: 
 \say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}
@@
 Similar to the previously studied approximate aggregators, HyperLogLog (HLL) has constant space and time complexity and provides an estimate of a set's cardinality.
-Hence, HLLs are considered valuable aggregators for our Thesis.
-
-\iffalse
-%TODO: Worth to show? HLL uses only one hash, only CMS and Bloom benefit from this.
-\subsection{Hash improvement}
-An improvement to this structure was defined in \cite{Kirsch-Better-Bloom} where the authors attempt to reduce the overhead of bloom filters operations such as adding a new element or answering a query by focusing on the hash functions.
-The proposed method is based on hashing literature.
-It has been shown that multiple hash functions can be created from the combination of only two hash functions.
-This means that a new hash function \textit{g} can be produced based on  two other independent and uniform existing hash functions, \textit{h1} and \textit{h2}.
-The new hash function will map elements to a universe ranging from \textit{0} to \textit{p-1} and will be defined as \textit{$g(x) = h1(x) + ih2(x) mod p$}.
-Since the only nontrivial computations performed by a Bloom filter are the evaluations of pseudo-random hash functions, any reduction in the required number of pseudo-random hash functions yields a reduction in the time required to add an element or query a Bloom filter.
-
-Bloom filters are not the only approximate aggregators that use hash functions as a basis of their algorithm.
-All of the studied probabilistic data structures do so.
-Therefore, the technique proposed is beneficial in Count-Min Sketches and Bloom filters.
-\fi
+Hence, HLLs are considered valuable aggregators for our thesis.
 
@@
 However, when working with sliding windows there is the need not only to insert new events but to evict old ones.
-The probabilistic data structures described until now can aggregate new incoming data but do not address the type of temporality needed for sliding windows --- \textit{i.e.} the ability to expire old elements from the aggregated state.
+The probabilistic data structures described until now can aggregate new incoming data but do not address the type of temporality needed for sliding windows --- \textit{i.e.}, the ability to expire old elements from the aggregated state.
 
@@
 With this approach, despite computing the aggregation state in constant time, we have a linear in window size space complexity solution.
-Thus, in this Section, we explore sliding window implementations of the previously discussed probabilistic data structures that keep the space and time complexity at the very least logarithmic.
-
-%\subsection{Sliding Bloom filter}
+Thus, in this section, we explore sliding window implementations of the previously discussed probabilistic data structures that keep the space and time complexity at the very least logarithmic.
+
+%\subsection{Sliding Bloom Filter}
 %\subsection*{Data Structure}
@@
 
-The work of Chabchoub et al. adapts the original HLL aggregator from Flajolet et al. to work under a sliding window framework  \cite{Chabchoub-Sliding-HLL}.
+The work of Chabchoub \emph{et al.} adapts the original HLL aggregator from Flajolet \emph{et al.} to work under a sliding window framework  \cite{Chabchoub-Sliding-HLL}.
 The main idea behind this implementation is to replace every HLL counter with a \textit{List of Future Possible Maxima (LFPM)}.
@@
 \begin{itemize}
-    \item Delete \textit{old packets}: packets that do not belong in the window anymore --- \textit{i.e.} packets <t\textsubscript{i}, p\textsubscript{i}> with t\textsubscript{i} $\leq$ t\textsubscript{k} - WindowSize
-
-    \item Delete less or equally valued packets: packets that will never be a maximum --- \textit{i.e.} packets <t\textsubscript{i}, p\textsubscript{i}> with p\textsubscript{i} $\leq$ p\textsubscript{k}
+    \item Delete \textit{old packets}: packets that do not belong in the window anymore --- \textit{i.e.}, packets <t\textsubscript{i}, p\textsubscript{i}> with t\textsubscript{i} $\leq$ t\textsubscript{k} - WindowSize
+
+    \item Delete less or equally valued packets: packets that will never be a maximum --- \textit{i.e.}, packets <t\textsubscript{i}, p\textsubscript{i}> with p\textsubscript{i} $\leq$ p\textsubscript{k}
 
@@
 Estimation of cardinality is done similarly to the original HLL cardinality estimation algorithm.
-To estimate the cardinality at a timestamp \textit{t}, from all \textit{m} buckets and all \textit{m} lists, select packets that are in the window --- \textit{i.e.} packets <t\textsubscript{i}, p\textsubscript{i}> where t\textsubscript{i} > \textit{t} - WindowSize.
+To estimate the cardinality at a timestamp \textit{t}, from all \textit{m} buckets and all \textit{m} lists, select packets that are in the window --- \textit{i.e.}, packets <t\textsubscript{i}, p\textsubscript{i}> where t\textsubscript{i} > \textit{t} - WindowSize.
 This filtering is necessary since the proposed algorithm evicts old packets only at element insertion and at the time of the query some packets may no longer belong to the window.
@@
 
-The lists are dynamic thus they shrink and grow at run time.
+The lists are dynamic thus they shrink and grow at runtime.
 However, \textit{ln(n/m)} was proved to be an upper bound for \textit{L\textsubscript{n}}.
@@
 In Section \ref{sec:hll-lfpm} we have analyzed a Sliding HLL implementation where each of the \textit{m} buckets holds a list of values (LFPM) instead of a single integer count.
-However, the LFPM structure is dynamic and its size varies at run time.
-
-The approach proposed by Xu uses a smaller amount of memory that does not change at run time \cite{Xu-hll-sliding-drv}.
-The author proposes a special counter: a \textit{Distance Recorder Vector (DRV)}.
+However, the LFPM structure is dynamic and its size varies at runtime.
+
+The approach proposed by Xu uses a smaller amount of memory that does not change at runtime \cite{Xu-hll-sliding-drv}. The author proposes a special counter: a \textit{Distance Recorder Vector (DRV)}.
 
@@
 For the DRV of that bucket, we increment the value of all records by one.
-Next, we set position \textit{p\textsubscript{new}} in the DRV to 0 --- \textit{i.e.} $\textrm{DRV}[p_{new}] = 0$.
+Next, we set position \textit{p\textsubscript{new}} in the DRV to 0 --- \textit{i.e.}, $\textrm{DRV}[p_{new}] = 0$.
 
@@
 \section{Summary}
-
-In this chapter, we presented state-of-the-art outlier detection methods and classified them according to Blázquez-García \emph{et al.} taxonomy \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
+In this chapter, we presented state of the art outlier detection methods and classified them according to Blázquez-García \emph{et al.} taxonomy \cite{Blazquez-Garcia-Review-Anomaly-Detection}.
 Additionally, we presented sliding window aggregation algorithms, probabilistic data structures and their respective sliding window implementations, with the intent of researching how to incrementally maintain an aggregation in a streaming environment.

### 2020-07-05 chapters/validation.tex
Multiple grammar corrections

@@
 
-All the synthetic datasets used throughout this section are a time-series where at each time-step we sample uniformly from a fixed distribution, \textit{i.e.} there are no time correlations in the used datasets.
+All the synthetic datasets used throughout this section are a time-series where at each time-step we sample uniformly from a fixed distribution, \textit{i.e.}, there are no time correlations in the used datasets.
 The reference dataset R1 used in this first experiment contained a single feature that followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$ for the 5 million rows of the dataset.
 Figure \ref{fig:timeseries-r1} shows the time-series of the single feature in dataset R1.
-The dataset processed in the streaming phase (\textit{i.e.} the target dataset) was generated with a size of 5 million events and a single feature.
+The dataset processed in the streaming phase (\textit{i.e.}, the target dataset) was generated with a size of 5 million events and a single feature.
 For the first half, \textit{i.e.}, for the first 2.5 million events, the feature followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$, like the reference dataset.
@@
 
-In this experiment, we used a half-life $n_{1/2}=62500$ and corresponding sample size of 250,000 events, like in the previous Section' experiment 02.
+In this experiment, we used a half-life $n_{1/2}=62500$ and corresponding sample size of 250,000 events, like in the previous section's experiment 02.
 Figure \ref{fig:JSD-signal-test02} shows the computed JSD signal for the target dataset and Figure \ref{fig:JSD-signal-zoom-test02} zooms in on the first 2 millionth events, before any distribution is changed.
@@
 
-We used the reference dataset R5 (Figure \ref{fig:timeseries-r5}) which had 5 million events, with a single-feature that followed a gaussian with  $\mu=100$ and $\sigma=10$.
+We used the reference dataset R5 (Figure \ref{fig:timeseries-r5}) which had 5 million events, with a single feature that followed a gaussian with  $\mu=100$ and $\sigma=10$.
 Our target dataset T5 is represented in Figure \ref{fig:timeseries-t5}.
@@
         \textbf{Feature} & \textbf{Reference Distribution} & \textbf{Parameters} &  &  \\ \cline{1-3}
-        x1               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
-        x2               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
-        x3               & Uniform                         & $a=200$, $b=250$       &  &  \\ \cline{1-3}
-        x4               & Uniform                         & $a=50$, $b=125$        &  &  \\ \cline{1-3}
+        $x_1$               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
+        $x_2$               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
+        $x_3$               & Uniform                         & $a=200$, $b=250$       &  &  \\ \cline{1-3}
+        $x_4$               & Uniform                         & $a=50$, $b=125$        &  &  \\ \cline{1-3}
         \end{tabular}
@@
     \includegraphics[width=1\linewidth]{figures/timeseries-r6-x1.pdf} 
-    \caption{Feature x1 reference time-series} 
+    \caption{Feature $x_1$ reference time-series} 
     \label{fig:timeseries-r6-x1} 
@@
     \includegraphics[width=1\linewidth]{figures/timeseries-r6-x2.pdf} 
-    \caption{Feature x2 reference time-series} 
+    \caption{Feature $x_2$ reference time-series} 
     \label{fig:timeseries-r6-x2} 
@@
     \includegraphics[width=1\linewidth]{figures/timeseries-r6-x3.pdf} 
-    \caption{Feature x3 reference time-series} 
+    \caption{Feature $x_3$ reference time-series} 
     \label{fig:timeseries-r6-x3} 
@@
     \includegraphics[width=1\linewidth]{figures/timeseries-r6-x4.pdf} 
-    \caption{Feature x4 reference time-series} 
+    \caption{Feature $x_4$ reference time-series} 
     \label{fig:timeseries-r6-x4} 
@@
 We used a target dataset T6 with 5 million events and the same four features ($x_1$, $x_2$, $x_3$ and $x_4$).
-Feature x1 followed the reference distribution for the 5 million events of the target dataset.
+Feature $x_1$ followed the reference distribution for the 5 million events of the target dataset.
 We did not change this feature underlying distribution to serve as a control, \textit{i.e.}, we expect no alerts for this one.
@@
 
-Feature x2's underlying distribution changed four times and returned to normal, so we expect to see alarms with increasing divergence values but eventually no more alarms.
+Feature $x_2$'s underlying distribution changed four times and returned to normal, so we expect to see alarms with increasing divergence values but eventually no more alarms.
 Figure \ref{fig:timeseries-t6-x2} shows the time-series for the target dataset T6 and feature $x_2$.
 Feature $x_2$ followed a gaussian with $\mu=100$ and $\sigma=5$ like in the reference period up to the 1 millionth event mark, then a gaussian with $\mu=100$ and $\sigma=10$ up to the 2.5 millionth event, then a gaussian $\mu=100$ and $\sigma=15$ until the 3 millionth event, then a gaussian $\mu=100$ and $\sigma=20$ until the 4 millionth event and finally returned to the original gaussian with $\mu=100$ and $\sigma=5$ until the last event.
-Table \ref{tbl:multi-feat-x2-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature x2.
+Table \ref{tbl:multi-feat-x2-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_2$.
 \begin{table}[!htb]
@@
     \end{center}
-    \caption{Feature x2 generating distributions}
+    \caption{Feature $x_2$ generating distributions}
     \label{tbl:multi-feat-x2-changes}
@@
 
-Feature x3's underlying distribution changed three times and never returned to normal, so we expect to see alarms with increasing divergence values up until the end of the stream analysis.
-Figure \ref{fig:timeseries-t6-x3} shows the time-series for the target dataset T6 and feature x3.
-Feature x3 followed a uniform distribution with $a=200$ and $b=250$ like in the reference period up to the 1.5 millionth event mark, then a uniform distribution with $a=200$ and $b=220$ up to the 2 millionth event, then a uniform distribution with $a=180$ and $b=200$ until the 3.5 millionth event and finally a gaussian with $\mu=50$ and $\sigma=20$ until the last event.
-Table \ref{tbl:multi-feat-x3-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature x3.
+Feature $x_3$'s underlying distribution changed three times and never returned to normal, so we expect to see alarms with increasing divergence values up until the end of the stream analysis.
+Figure \ref{fig:timeseries-t6-x3} shows the time-series for the target dataset T6 and feature $x_3$.
+Feature $x_3$ followed a uniform distribution with $a=200$ and $b=250$ like in the reference period up to the 1.5 millionth event mark, then a uniform distribution with $a=200$ and $b=220$ up to the 2 millionth event, then a uniform distribution with $a=180$ and $b=200$ until the 3.5 millionth event and finally a gaussian with $\mu=50$ and $\sigma=20$ until the last event.
+Table \ref{tbl:multi-feat-x3-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_3$.
 \begin{table}[!htb]
@@
     \end{center}
-    \caption{Feature x3 generating distributions}
+    \caption{Feature $x_3$ generating distributions}
     \label{tbl:multi-feat-x3-changes}
@@
 
-Feature x4's underlying distribution changed two times and returned to normal, so we expect to see no alarms towards the end of the stream analysis.
-Figure \ref{fig:timeseries-t6-x4} shows the time-series for the target dataset T6 and feature x4.
-Feature x4 followed a uniform distribution like the reference one, with $a=50$ and $b=125$ up until the 1.5 millionth event.
+Feature $x_4$'s underlying distribution changed two times and returned to normal, so we expect to see no alarms towards the end of the stream analysis.
+Figure \ref{fig:timeseries-t6-x4} shows the time-series for the target dataset T6 and feature $x_4$.
+Feature $x_4$ followed a uniform distribution like the reference one, with $a=50$ and $b=125$ up until the 1.5 millionth event.
 After that, it followed a uniform distribution with $a=125$ and $b=200$ until the 3 millionth event where it resumed its original uniform distribution (with $a=50$ and $b=125$) until the end.
-Table \ref{tbl:multi-feat-x4-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature x4.
+Table \ref{tbl:multi-feat-x4-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_4$.
 \begin{table}[!htb]
@@
     \end{center}
-    \caption{Feature x4 generating distributions}
+    \caption{Feature $x_4$ generating distributions}
     \label{tbl:multi-feat-x4-changes}
@@
     \includegraphics[width=1\linewidth]{figures/timeseries-t6-x1.pdf} 
-    \caption{Feature x1 target time-series} 
+    \caption{Feature $x_1$ target time-series} 
     \label{fig:timeseries-t6-x1} 
@@
     \includegraphics[width=1\linewidth]{figures/timeseries-t6-x2.pdf} 
-    \caption{Feature x2 target time-series} 
+    \caption{Feature $x_2$ target time-series} 
     \label{fig:timeseries-t6-x2} 
@@
     \includegraphics[width=1\linewidth]{figures/timeseries-t6-x3.pdf} 
-    \caption{Feature x3 target time-series} 
+    \caption{Feature $x_3$ target time-series} 
     \label{fig:timeseries-t6-x3} 
@@
     \includegraphics[width=1\linewidth]{figures/timeseries-t6-x4.pdf} 
-    \caption{Feature x4 target time-series} 
+    \caption{Feature $x_4$ target time-series} 
     \label{fig:timeseries-t6-x4} 
@@
 
-In this experiment, we used a tuple-based half-life $n_{1/2}=62500$, \textit{i.e.} sampling and target windows of 250,000 tuples.
+In this experiment, we used a tuple-based half-life $n_{1/2}=62500$, \textit{i.e.}, sampling and target windows of 250,000 tuples.
 We used 100 bins for the reference histogram plus the two special infinity bins.
@@
 In Section \ref{sec:stream-phase}, we mention that, in our stream analysis, we perform our multiple hypothesis divergence tests on a fixed user-defined frequency.
-Finding an optimal frequency value for the test falls out of scope for this Thesis, but ideally, we set it to be large enough so that we do fewer computations and achieve higher throughput but still get alerts in useful time.
+Finding an optimal frequency value for the test falls out of scope for this thesis, but ideally, we set it to be large enough so that we do fewer computations and achieve higher throughput but still get alerts in useful time.
 In this experiment, we performed the divergence test with the Holm-Bonferroni correction \ref{sec:holmbonferroni} for every new event (frequency of 1 event).
@@
 
-In Figure \ref{fig:x1-corrected-pvalues} we see the plot of corrected p-values for feature x1 and the probability threshold black horizontal line at the bottom.
-Recall that feature x1' values were generated by a Gaussian distribution for both the reference period (Figure \ref{fig:timeseries-r6-x1}) and the target period (Figure \ref{fig:timeseries-t6-x1}), with the same parameters.
+In Figure \ref{fig:x1-corrected-pvalues} we see the plot of corrected p-values for feature $x_1$ and the probability threshold black horizontal line at the bottom.
+Recall that feature $x_1$'s values were generated by a Gaussian distribution for both the reference period (Figure \ref{fig:timeseries-r6-x1}) and the target period (Figure \ref{fig:timeseries-t6-x1}), with the same parameters.
 We observe that the corrected p-values never pass below the threshold, as we expected.
@@
       \includegraphics[scale=0.8]{figures/x1-corrected-pvalues.png}
-      \caption{Feature x1 corrected p-values and threshold}
+      \caption{Feature $x_1$ corrected p-values and threshold}
       \label{fig:x1-corrected-pvalues}
@@
 
-Figure \ref{fig:x2-corrected-pvalues} shows a plot of feature x2's corrected p-values and the FWER threshold.
-Feature x2 underlying distribution was changed four times (Table \ref{tbl:multi-feat-x2-changes}): at the 1, 2.5, 3 and 4 millionth events.
+Figure \ref{fig:x2-corrected-pvalues} shows a plot of feature $x_2$'s corrected p-values and the FWER threshold.
+Feature $x_2$ underlying distribution was changed four times (Table \ref{tbl:multi-feat-x2-changes}): at the 1, 2.5, 3 and 4 millionth events.
 The first change is consistent with the sudden drop in corrected p-value for this feature, as we cross below the threshold and enter a divergent state.
@@
       \includegraphics[scale=0.8]{figures/x2-corrected-pvalues.png}
-      \caption{Feature x2 corrected p-values and threshold}
+      \caption{Feature $x_2$ corrected p-values and threshold}
       \label{fig:x2-corrected-pvalues}
@@
 We resume the reference or original distribution at the 4 millionth event.
-However, as we can see from a close-up on the original plot (Figure \ref{fig:x2-corrected-pvalues-zoom}), it takes around 600,000 events for the target sliding histogram aggregation to once again resemble the reference one, resulting in a high enough p-value and thus stop alerting x2 as divergent.
+However, as we can see from a close-up on the original plot (Figure \ref{fig:x2-corrected-pvalues-zoom}), it takes around 600,000 events for the target sliding histogram aggregation to once again resemble the reference one, resulting in a high enough p-value and thus stop alerting $x_2$ as divergent.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.8]{figures/x2-corrected-pvalues-zoom2.png}
-      \caption{Feature x2 zoomed in corrected p-values and threshold}
+      \caption{Feature $x_2$ zoomed in corrected p-values and threshold}
       \label{fig:x2-corrected-pvalues-zoom}
@@
 
-Feature x3 was the only feature that never returned to its original distribution and is the only one that should be reported until the end as divergent.
+Feature $x_3$ was the only feature that never returned to its original distribution and is the only one that should be reported until the end as divergent.
 This is in fact the case as illustrated in Figure \ref{fig:x3-corrected-pvalues}.
-The first change to x3's generating distribution was introduced at the 1.5 millionth event and we clearly see the corrected p-value rapidly approaching values close to zero.
-For the rest of the dataset, this x3's p-value remains below the threshold, as we never swap to its original distribution.
+The first change to $x_3$'s generating distribution was introduced at the 1.5 millionth event and we clearly see the corrected p-value rapidly approaching values close to zero.
+For the rest of the dataset, this $x_3$'s p-value remains below the threshold, as we never swap to its original distribution.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.8]{figures/x3-corrected-pvalues.png}
-      \caption{Feature x3 corrected p-values and threshold}
+      \caption{Feature $x_3$ corrected p-values and threshold}
       \label{fig:x3-corrected-pvalues}
@@
 
-Finally, feature x4's corrected p-value plot is given by Figure \ref{fig:x4-corrected-pvalues}.
-Similarly to feature x2, we resume the original distribution at a certain point, in this case at the 3-millionth event.
-However, as we can see from a close-up on the original plot (Figure \ref{fig:x4-corrected-pvalues-zoom}), it takes around 810,000 events for the target sliding window aggregation to be considered similar enough to the reference one and stop alerting x4 as divergent.
+Finally, feature $x_4$'s corrected p-value plot is given by Figure \ref{fig:x4-corrected-pvalues}.
+Similarly to feature $x_2$, we resume the original distribution at a certain point, in this case at the 3-millionth event.
+However, as we can see from a close-up on the original plot (Figure \ref{fig:x4-corrected-pvalues-zoom}), it takes around 810,000 events for the target sliding window aggregation to be considered similar enough to the reference one and stop alerting $x_4$ as divergent.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.8]{figures/x4-corrected-pvalues.png}
-      \caption{Feature x4 corrected p-values and threshold}
+      \caption{Feature $x_4$ corrected p-values and threshold}
       \label{fig:x4-corrected-pvalues}
@@
       \includegraphics[scale=0.8]{figures/x4-corrected-pvalues-zoom2.png}
-      \caption{Feature x4 zoomed in corrected p-values and threshold}
+      \caption{Feature $x_4$ zoomed in corrected p-values and threshold}
       \label{fig:x4-corrected-pvalues-zoom}
@@
 With this experiment, we further validate that the p-value is a measure of how divergent a feature is: the closer to zero, the lower the probability of the statistical test result and hence the more divergent a feature is considered to be.
-At the 1.5 millionth event, we swap both x3 and x4's distribution.
-For the first 2,535,000 events, x4 has a lower p-value than x3 and hence is considered more divergent.
-However, after that, we change x3's distribution so aggressively that x3 starts being reported as most divergent, with a lower p-value than x4.
+At the 1.5 millionth event, we swap both $x_3$ and $x_4$'s distribution.
+For the first 2,535,000 events, $x_4$ has a lower p-value than $x_3$ and hence is considered more divergent.
+However, after that, we change $x_3$'s distribution so aggressively that $x_3$ starts being reported as most divergent, with a lower p-value than $x_4$.
 
@@
 In this experiment, we used the older one as the reference dataset and the more recent one as the target period.
-Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus numerically encoded categorical features (\textit{e.g.} one-hot encoded or ordinal encoded \cite{categoricalencoding}).
+Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus numerically encoded categorical features (\textit{e.g.}, one-hot encoded or ordinal encoded \cite{categoricalencoding}).
 Table \ref{tbl:merchant1-datasets-summary} shows summary statistics for both datasets.
@@
 We set the half-life to be $n_{1/2}=37,803$ and discard events whose EMA weight is 6\% or less, which is equivalent to using sampling windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).
-Hence, our samples' windows had 151,212 events.
+Hence, our samples's windows had 151,212 events.
 Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:nsamples}) of 10\% we needed to make a minimum of 42,137 samples.
@@
       \includegraphics[scale=0.5]{figures/merchant-x1-reference.pdf}
-      \caption{Merchant feature x1 reference time-series}
+      \caption{Merchant feature $x_1$ reference time-series}
       \label{fig:merchant-x1-reference}
@@
       \includegraphics[scale=0.5]{figures/merchant-x1-target.pdf}
-      \caption{Merchant feature x1 target time-series}
+      \caption{Merchant feature $x_1$ target time-series}
       \label{fig:merchant-x1-target}
@@
       \includegraphics[scale=0.5]{figures/merchant-x1-correctedpvalues.pdf}
-      \caption{Merchant feature x1 corrected p-values}
+      \caption{Merchant feature $x_1$ corrected p-values}
       \label{fig:merchant-x1-correctedpvalues}
@@
 Now let us look at features $x_2$ and $x_3$.
-Figures \ref{fig:merchant-x2-reference} and \ref{fig:merchant-x2-target} show feature $x_2$' time-series plots for the reference and target periods, respectively.
+Figures \ref{fig:merchant-x2-reference} and \ref{fig:merchant-x2-target} show feature $x_2$'s time-series plots for the reference and target periods, respectively.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.5]{figures/merchant-x2-reference.pdf}
-      \caption{Merchant feature x2 reference time-series}
+      \caption{Merchant feature $x_2$ reference time-series}
       \label{fig:merchant-x2-reference}
@@
       \includegraphics[scale=0.5]{figures/merchant-x2-target.pdf}
-      \caption{Merchant feature x2 target time-series}
+      \caption{Merchant feature $x_2$ target time-series}
       \label{fig:merchant-x2-target}
@@
 \end{figure}
-Figures \ref{fig:merchant-x3-reference} and \ref{fig:merchant-x3-target} show feature $x_3$' time-series plots for the reference and target periods, respectively.
+Figures \ref{fig:merchant-x3-reference} and \ref{fig:merchant-x3-target} show feature $x_3$'s time-series plots for the reference and target periods, respectively.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.5]{figures/merchant-x3-reference.pdf}
-      \caption{Merchant feature x3 reference time-series}
+      \caption{Merchant feature $x_3$ reference time-series}
       \label{fig:merchant-x3-reference}
@@
       \includegraphics[scale=0.5]{figures/merchant-x3-target.pdf}
-      \caption{Merchant feature x3 target time-series}
+      \caption{Merchant feature $x_3$ target time-series}
       \label{fig:merchant-x3-target}
@@
 Relative to $x_1$, features $x_2$ and $x_3$ also have very similar reference and target periods.
-Just like feature x1, we expect to have some alarms but still not too many.
+Just like feature $x_1$, we expect to have some alarms but still not too many.
 Once again, to visualize the alarms, we plot the corrected p-values for each event of the target time-series for each feature.
-Figure \ref{fig:merchant-x2-correctedpvalues} is a plot of feature x2' corrected p-values and the FWER threshold of 1\%.
-Likewise, Figure \ref{fig:merchant-x3-correctedpvalues} shows the same plot but for feature x3.
+Figure \ref{fig:merchant-x2-correctedpvalues} is a plot of feature $x_2$'s corrected p-values and the FWER threshold of 1\%.
+Likewise, Figure \ref{fig:merchant-x3-correctedpvalues} shows the same plot but for feature $x_3$.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.5]{figures/merchant-x2-correctedpvalues.pdf}
-      \caption{Merchant feature x2 corrected p-values}
+      \caption{Merchant feature $x_2$ corrected p-values}
       \label{fig:merchant-x2-correctedpvalues}
@@
       \includegraphics[scale=0.5]{figures/merchant-x3-correctedpvalues.pdf}
-      \caption{Merchant feature x3 corrected p-values}
+      \caption{Merchant feature $x_3$ corrected p-values}
       \label{fig:merchant-x3-correctedpvalues}
@@
 Next, we analyze a feature with very different reference and target time-series.
-Feature x4' reference time-series is presented in Figure \ref{fig:merchant-x4-reference} and its target period in Figure \ref{fig:merchant-x4-target}.
+Feature $x_4$'s reference time-series is presented in Figure \ref{fig:merchant-x4-reference} and its target period in Figure \ref{fig:merchant-x4-target}.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.5]{figures/merchant-x4-reference.pdf}
-      \caption{Merchant feature x4 reference time-series}
+      \caption{Merchant feature $x_4$ reference time-series}
       \label{fig:merchant-x4-reference}
@@
       \includegraphics[scale=0.5]{figures/merchant-x4-target.pdf}
-      \caption{Merchant feature x4 target time-series}
+      \caption{Merchant feature $x_4$ target time-series}
       \label{fig:merchant-x4-target}
@@
 Notice that even if for the most part both reference and target periods values belong to a domain between 0 and 1, at a certain point in time, in the target time-series, a lot of -1 values appear.
-In Figure \ref{fig:merchant-x4-correctedpvalues} we plot the corrected p-values for feature x4 target time-series.
+In Figure \ref{fig:merchant-x4-correctedpvalues} we plot the corrected p-values for feature $x_4$ target time-series.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.5]{figures/merchant-x4-correctedpvalues.pdf}
-      \caption{Merchant feature x4 corrected p-values}
+      \caption{Merchant feature $x_4$ corrected p-values}
       \label{fig:merchant-x4-correctedpvalues}
@@
 \end{figure}
-Like the previous analyzed features, x4 also drops below the threshold some times.
-However, unlike the other features, where feature values return to normal and alarms stopped being raised, x4 remains below the threshold line until the end of the target period, where a small spike occurs.
-The last time feature x4 was ever above the threshold line (and hence considered normal) was before the appearance of the -1 values in the target period.
-This seems to work as expected: -1 values were so off the reference domain that for here on out, as long as these values keep appearing, we will not stop reporting x4 as divergent.
+Like the previous analyzed features, $x_4$ also drops below the threshold some times.
+However, unlike the other features, where feature values return to normal and alarms stopped being raised, $x_4$ remains below the threshold line until the end of the target period, where a small spike occurs.
+The last time feature $x_4$ was ever above the threshold line (and hence considered normal) was before the appearance of the -1 values in the target period.
+This seems to work as expected: -1 values were so off the reference domain that for here on out, as long as these values keep appearing, we will not stop reporting $x_4$ as divergent.
 
-The -1 value is often used as a placeholder for missing values, \textit{i.e.} if some problem occurs and the value for a feature is missing at a given timestamp or event, -1 is used.
+The -1 value is often used as a placeholder for missing values, \textit{i.e.}, if some problem occurs and the value for a feature is missing at a given timestamp or event, -1 is used.
 A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
@@
 
-Feature x1 has less alerts in this experiment where we use equal reference and target periods.
+Feature $x_1$ has less alerts in this experiment where we use equal reference and target periods.
 However, the alarms last for about two months.
 This is unexpected as we predicted we would see no alarms or few ones with short duration.
-Likewise, we observe that for feature x2 (Figure \ref{fig:merchant2-x2-correctedpvalues}) and x3 (Figure \ref{fig:merchant2-x3-correctedpvalues}) we still have many alarms and some lasting up to one month.
+Likewise, we observe that for feature $x_2$ (Figure \ref{fig:merchant2-x2-correctedpvalues}) and $x_3$ (Figure \ref{fig:merchant2-x3-correctedpvalues}) we still have many alarms and some lasting up to one month.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.5]{figures/merchant2-x1-correctedpvalues.pdf}
-      \caption{Merchant feature x1 corrected p-values with equal reference and target periods}
+      \caption{Merchant feature $x_1$ corrected p-values with equal reference and target periods}
       \label{fig:merchant2-x1-correctedpvalues}
@@
       \includegraphics[scale=0.5]{figures/merchant2-x2-correctedpvalues.pdf}
-      \caption{Merchant feature x2 corrected p-values with equal reference and target periods}
+      \caption{Merchant feature $x_2$ corrected p-values with equal reference and target periods}
       \label{fig:merchant2-x2-correctedpvalues}
@@
       \includegraphics[scale=0.5]{figures/merchant2-x3-correctedpvalues.pdf}
-      \caption{Merchant feature x3 corrected p-values with equal reference and target periods}
+      \caption{Merchant feature $x_3$ corrected p-values with equal reference and target periods}
       \label{fig:merchant2-x3-correctedpvalues}
@@
 
-Now we analyze feature x4 from the previous experiment, as this one was the one with the highest measured drift.
-Recall that feature x4' target time-series (Figure \ref{fig:merchant-x4-target}) was significantly different from the reference one (Figure \ref{fig:merchant-x4-reference}).
+Now we analyze feature $x_4$ from the previous experiment, as this one was the one with the highest measured drift.
+Recall that feature $x_4$'s target time-series (Figure \ref{fig:merchant-x4-target}) was significantly different from the reference one (Figure \ref{fig:merchant-x4-reference}).
 Because of this, the corrected p-values plot from the previous experiment (Figure \ref{fig:merchant-x4-correctedpvalues}) crossed our threshold line multiple times and remained in an alert state for most of the target time-series.
@@
 Because of this, the corrected p-values plot is now different.
-Figure \ref{fig:merchant2-x4-correctedpvalues} shows the corrected p-values plot for feature x4 when we used the reference period for both phases.
+Figure \ref{fig:merchant2-x4-correctedpvalues} shows the corrected p-values plot for feature $x_4$ when we used the reference period for both phases.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.5]{figures/merchant2-x4-correctedpvalues.pdf}
-      \caption{Merchant feature x4 corrected p-values with equal reference and target periods}
+      \caption{Merchant feature $x_4$ corrected p-values with equal reference and target periods}
       \label{fig:merchant2-x4-correctedpvalues}
@@
 Figure \ref{fig:merchant2-x4-correctedpvalues} shows that we have a few alerts but significantly less than in Figure \ref{fig:merchant-x4-correctedpvalues}.
-Furthermore, in this new experiment, x4 eventually stops being reported at all until the end of the dataset and ends at a high p-value.
+Furthermore, in this new experiment, $x_4$ eventually stops being reported at all until the end of the dataset and ends at a high p-value.
 Recall that the higher the p-value the less divergent a feature is.
@@
 
-We now introduce a new feature, x5.
-Feature x5 also suffers from a large amount of drift.
-In the reference and target datasets presented in Table \ref{tbl:merchant1-datasets-summary}, feature x5 reference time-series plot is given by Figure \ref{fig:merchant-x5-reference} and the target time-series by Figure \ref{fig:merchant-x5-target}.
+We now introduce a new feature, $x_5$.
+Feature $x_5$ also suffers from a large amount of drift.
+In the reference and target datasets presented in Table \ref{tbl:merchant1-datasets-summary}, feature $x_5$ reference time-series plot is given by Figure \ref{fig:merchant-x5-reference} and the target time-series by Figure \ref{fig:merchant-x5-target}.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.5]{figures/merchant-x5-reference.pdf}
-      \caption{Merchant feature x5 reference time-series}
+      \caption{Merchant feature $x_5$ reference time-series}
       \label{fig:merchant-x5-reference}
@@
       \includegraphics[scale=0.5]{figures/merchant-x5-target.pdf}
-      \caption{Merchant feature x5 target time-series}
+      \caption{Merchant feature $x_5$ target time-series}
       \label{fig:merchant-x5-target}
@@
       \includegraphics[scale=0.5]{figures/merchant-x5-correctedpvalues.pdf}
-      \caption{Merchant feature x5 corrected p-values}
+      \caption{Merchant feature $x_5$ corrected p-values}
       \label{fig:merchant-x5-correctedpvalues}
@@
       \includegraphics[scale=0.5]{figures/merchant2-x5-correctedpvalues.pdf}
-      \caption{Merchant feature x5 corrected p-values with equal reference and target periods}
+      \caption{Merchant feature $x_5$ corrected p-values with equal reference and target periods}
       \label{fig:merchant2-x5-correctedpvalues}

### 2020-07-05 chapters/background.tex
Abbrevs

@@
 
-Two of the most basic SWAG algorithms are \textit{Recalculate-From-Scratch (RFS)} and \textit{Subtract-On-Evict (SOE)}.
+Two of the most basic SWAG algorithms are Recalculate-From-Scratch (RFS) and Subtract-On-Evict (SOE).
 Recalculate-From-Scratch (RFS) is the simplest and most general SWAG algorithm.
@@
 Recall that the combine function is represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.
-Pseudocode for the Recalculate-From-Scratch algorithm is shown in Algorithm \ref{pseudo:rfs}.
+Pseudocode for the Recalculate-From-Scratch algorithm is shown in Algorithm \ref{pseudo:rfs}, assuming all functions have access to a \textit{vals} array which holds the window contents.
 
@@
     \begin{algorithmic}[1]
-        \Function{query}{vals}
+        \Function{query}{}
             \State $agg\gets\overline{\theta}$
@@
 
-        \Function{insert}{vals, v}
-
+        \Function{insert}{v}
             \State $vals.pushBack(v)$
@@
 
-        \Function{evict}{vals}
-
+        \Function{evict}{}
             \State $vals.popFront()$
@@
 However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable.
-Algorithm \ref{pseudo:soe} shows pseudocode for SOE.
+Algorithm \ref{pseudo:soe} shows pseudocode for SOE assuming all functions have access to the window contents through the \textit{vals} array and access to the current aggregation state \textit{agg}.
 
@@
         \Function{query}{}
-
             \State \Return $agg$
@@
 
-        \Function{evict}{vals}
-
+        \Function{evict}{}
             \State $v \gets vals.popFront()$

### 2020-07-05 chapters/sota.tex
Abbrevs

@@
 Each element in each of the stacks is a pair \textit{<val, agg>} where \textit{val} is the item's value and \textit{agg} is the aggregation value of everything below it on the stack.
-The pseudocode for Two-Stacks can be seen in Algorithm \ref{pseudo:2-stacks}, with $\oplus$ as the \textit{combine} function and $\overline{\theta}$ as the identity value of the \textit{combine} function, such that $\overline{\theta} \oplus x$ = $x \oplus \overline{\theta}$ = x.
+The pseudocode for Two-Stacks can be seen in Algorithm \ref{pseudo:2-stacks}, assuming all functions have access to stacks F and B, with $\oplus$ as the \textit{combine} function and $\overline{\theta}$ as the identity value of the \textit{combine} function, such that $\overline{\theta} \oplus x$ = $x \oplus \overline{\theta}$ = x.
 For example, if $\oplus$ is the arithmetic sum then $\overline{\theta}$ is 0.
@@
 
-        \Function{query}{F, B}
+        \Function{query}{}
             \State \Return{getStackAggregationValue(F) $\oplus$ getStackAggregationValue(B)}
@@
 
-        \Function{insert}{B, v}
+        \Function{insert}{v}
             \State{B.push( \{val: v, agg: getStackAggregationValue(B) $\oplus$ v\} }
@@
 
-        \Function{evict}{F, B}
-
+        \Function{evict}{}
             \If{F.isEmpty()} \Comment{// Flip B onto F and update top aggregation value}
@@
 Packets are only kept in the list if they are possible maxima over future windows.
-Because of that, the authors name this list a \textit{"List of Future Possible Maxima (LFPM)"}.
+Because of that, the authors name this list a ``List of Future Possible Maxima (LFPM)''.
 

### 2020-07-05 chapters/abstract.tex
Resumo tuga

@@
 We develop a two-phased and two-windowed method: first, we perform batch analysis on a reference window and then a stream analysis over a sliding streaming window.
-We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages.
-We do this for each event and each of its fields also denominated features in our context.
-We measure the divergence between both reference and sliding window histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used.
+We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs).
+We do this for each event and each of its fields, also denominated features in our context.
+We measure the divergence between both reference and sliding window EMA-based histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used.
 For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and their associated features.
@@
 Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next.
-Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity regarding window size.
+Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
 
 \vspace*{10mm}\noindent
-\textbf{Keywords}: streaming, real-time, lightweight, monitoring, concept drift, anomaly detection
+\textbf{Keywords}: data streams, monitoring, real-time, lightweight, concept drift

### 2020-07-05 chapters/resumo.tex
Resumo tuga

@@
 \chapter*{Resumo}
+Muitos sistemas de monitorização de \textit{streams} de dados em tempo real são estáticos uma vez colocados em produção.
+Os engenheiros desses sistemas configuram-nos sob o pressuposto de que dados futuros que fluem pelo sistema seguirão aproximadamente a mesma distribuição que os dados vistos anteriormente.
+Fazem-no conscientemente, com as ferramentas disponíveis, tendo em mente que no futuro poderão precisar de reconfigurar o sistema.
+Assim, mesmo que a configuração inicial do sistema seja uma das melhores, ao longo do tempo, devido a mudanças nos padrões de dados, o desempenho do sistema em produção decai gradualmente.
+A detecção de mudança de padrões de dados refere-se ao processo de encontrar padrões em dados que não estão em conformidade com o comportamento esperado ou usual.
+Uma detecção oportuna e precisa dos desvios dos padrões de dados permite que medidas imediatas sejam tomadas.
+Assim, o problema em questão é determinar quando reconfigurar o sistema, \textit{e.g.}, um modelo de \textit{Machine Learning}, com base numa análise dos desvios de padrões da \textit{stream} de dados.
+
+%Approach/Solution and %Main contributions
+Nesta tese, projetamos um método que utiliza agregações de \textit{streaming} leves e funções de divergência de distribuição para alertar sobre desvios nos padrões de dados em tempo real, na presença de grandes volumes de dados a alta velocidade, distorcidos e sazonais.
+Desenvolvemos um método de duas fases e duas janelas: primeiro, realizamos uma análise \textit{batch} numa janela de referência e, em seguida, uma análise \textit{streaming} sobre uma janela deslizante.
+Agregamos o conteúdo de ambas as janelas usando um histograma aproximado baseado em Médias Móveis Exponenciais (MMEs).
+Fazemos isto para cada evento e cada um dos seus campos, também denominados \textit{features} no nosso contexto.
+Medimos a divergência entre os histogramas de referência e os das janelas deslizantes baseados em MMEs para cada \textit{feature} com a função de divergência de \textit{Jensen-Shannon}, mas outras funções podem ser usadas.
+Para cada \textit{feature}, descobrimos o valor de probabilidade para cada medida de divergência e aplicamos a correção de múltiplo teste de Holm-Bonferroni para descobrir quais os valores de probabilidade mais baixos e as \textit{features} associados.
+Geramos alertas  para cada \textit{feature} com base num limite de probabilidade definido pelo administrador.
+
+Avaliamos o nosso método através de uma série de testes, alguns com conjuntos de dados sintéticos e outros com dados reais.
+Dividimos ainda os testes em testes de análise a uma única \textit{feature} e tests de análise com múltiplas \textit{features}.
+O nosso método detectou com precisão as anomalias introduzidas nos testes com conjuntos de dados sintéticos, mantendo um alto \textit{throughput}, tanto para análises com uma única ou múltiplas \textit{features}.
+No entanto, os testes com dados reais não foram tão precisos.
+Apesar de não sabermos as razões específicas para isso, passamos os próximos testes para trabalhos futuros e formulamos um conjunto de hipóteses que pretendem explicar o sucedido e, portanto, são dignas de serem exploradas em seguida.
+O nosso conjunto de testes apoia a afirmação de que métodos de agregação de janelas deslizantes e métodos de divergência de distribuição podem ser combinados para detectar alterações nos padrões de dados em cenários de \textit{streaming} com complexidade temporal e de memória constante.
 
 \vspace*{10mm}\noindent
-\textbf{Keywords}:
+\textbf{Keywords}: data streams, monitoring, real-time, lightweight, concept drift

### 2020-07-05 chapters/conclusions.tex
Correct wassertein vs jsd example

@@
 We believe that future work comprises testing new distances in the literature.
-In particular, we believe the Wasserstein distance is worth trying next, and illustrate why in Figure \ref{fig:wasserteinvsjsd}.
+In particular, we believe the Wasserstein distance is worth trying next, and illustrate why with Figure \ref{fig:wasserteinvsjsd} and Table \ref{tbl:wassertein-vs-jsd}.
 \begin{figure}[!htb]
     \begin{center}
-      \includegraphics[scale=0.6]{figures/wasserstein-vs-jsd.pdf}
+      \includegraphics[scale=0.55]{figures/wasserstein-vs-jsd.pdf}
       \caption{Wassertein vs Jensen–Shannon distances}
@@
 \end{figure}
-The Wasserstein distance measures the work required to transport the probability mass from one histogram to the other.
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|c|}
+        \hline
+        \textbf{Distribution \#1} & \textbf{Distribution \#2} & \textbf{Wasserstein Output} & \textbf{Jensen–Shannon Output} \\ \hline
+        a                         & a'                        & 2.0                         & 0.4451                         \\ \hline
+        b                         & b'                        & 0.5                         & 0.4451                         \\ \hline
+        \end{tabular}
+    \end{center}
+    \caption[Wasserstein and Jensen-Shannon divergences output comparison]{Wasserstein and Jensen-Shannon divergences for both distribution pairs}
+    \label{tbl:wassertein-vs-jsd}
+\end{table}
+The Wasserstein distance measures the work required to transport the probability mass from one histogram to another.
 This measure will be larger the further way the probability mass is.
-In Figure \ref{fig:wasserteinvsjsd} scenario, while in the JSD measure is the same for both cases, the Wasserstein distance is larger in case 2 because the probability mass has to be transported further.
+In Figure \ref{fig:wasserteinvsjsd}, we show two pairs of histograms, \textit{Histogram (a)} paired with \textit{Histogram (a')} and \textit{Histogram (b)} paired with \textit{Histogram (b')}.
+Table \ref{tbl:wassertein-vs-jsd} shows the output values for both pairs by each divergence function.
+The JSD measure is the same for both cases, whereas the Wasserstein distance is larger for the first pair because the probability mass has to be transported further.
 

### 2020-07-05 chapters/resumo.tex
Correct wassertein vs jsd example

@@
 %Approach/Solution and %Main contributions
-Nesta tese, projetamos um método que utiliza agregações de \textit{streaming} leves e funções de divergência de distribuição para alertar sobre desvios nos padrões de dados em tempo real, na presença de grandes volumes de dados a alta velocidade, distorcidos e sazonais.
+Nesta tese, projetamos um método que utiliza agregações de \textit{streaming} leves e funções de divergência para alertar sobre desvios nos padrões de dados em tempo real, na presença de grandes volumes de dados que fluem a alta velocidade, são altamente \textit{skewed} e sazonais.
 Desenvolvemos um método de duas fases e duas janelas: primeiro, realizamos uma análise \textit{batch} numa janela de referência e, em seguida, uma análise \textit{streaming} sobre uma janela deslizante.
@@
 O nosso conjunto de testes apoia a afirmação de que métodos de agregação de janelas deslizantes e métodos de divergência de distribuição podem ser combinados para detectar alterações nos padrões de dados em cenários de \textit{streaming} com complexidade temporal e de memória constante.
-
-\vspace*{10mm}\noindent
-\textbf{Keywords}: data streams, monitoring, real-time, lightweight, concept drift

### 2020-07-06 chapters/abstract.tex
Final changes and submission

@@
 In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data.
-We develop a two-phased and two-windowed method: first, we perform batch analysis on a reference window and then a stream analysis over a sliding streaming window.
+We develop a two-phased and two-windowed method: first, we perform a batch analysis on a reference window and then a stream analysis over a sliding streaming window.
 We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs).
@@
 We further split the tests into single and multi-feature analysis.
-Our method accurately detected the introduced anomalies in the synthetic dataset experiments while maintaining high throughput, for single and multi-feature analysis.
+Our method accurately detected the introduced anomalies in the experiments using synthetic datasets while maintaining high throughput, for single and multi-feature analysis.
 However, experiments with real data were not as accurate.

### 2020-07-06 chapters/background.tex
Final changes and submission

@@
 In this chapter, we discuss fundamental concepts that will help us identify the set of relevant techniques that are suitable to address the problem to be formulated and tackled in this thesis.
-In Section \ref{sec:stream-superset} we introduce batch and stream processing as two large scale data processing techniques.
-In Section \ref{sec:windows} we review windowing mechanisms that essentially turn unbounded data streams to finite datasets.
-In Section \ref{sec:aggregations} we introduce aggregations as a whole.
+In Section \ref{sec:stream-superset}, we introduce batch and stream processing as two large scale data processing techniques.
+In Section \ref{sec:windows}, we review windowing mechanisms that essentially turn unbounded data streams to finite datasets.
+In Section \ref{sec:aggregations}, we introduce aggregations as a whole.
 Then, we specifically discuss sliding window aggregations, algorithms used for its computation and aggregation properties.
-In Section \ref{sec:emas} we present Exponential Moving Averages as aggregators that replace the sliding window contents for a lighter representation, providing an approximate estimation of the aggregation value.
+In Section \ref{sec:emas}, we present Exponential Moving Averages as aggregators that replace the sliding window contents for a lighter representation, providing an approximate estimation of the aggregation value.
 
@@
 Batch datasets are known as static data \cite{Martin-Batch-Defin}.
-An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes).
+An example application of such a technique would be processing all the keywords searched for in Google's search engine\footnote{https://www.google.com}, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes).
 Between batch and stream processing, batch processing is the older paradigm.
@@
 For example, if we use a batch size of 100 events, the system would only produce accurate results every 100 events.
-To handle system inactivity usually these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed.
+Usually, to handle system inactivity, these micro-batch systems also have the notion of updating the aggregations with incomplete batches, if no event has entered the system after a pre-specified time period has elapsed.
 As such, micro-batching was created as a natural implementation of stream processing using the available batch processing systems of the time, but can not be considered true stream processing.
@@
 The Venn diagram in Figure \ref{fig:stream-superset} \cite{batch-is-a-special-case-of-streaming} illustrates the relation of both processing techniques.
-A data stream is unbounded or infinite while a bounded data stream has a beginning and an end.
+A data stream is unbounded or infinite whereas a bounded dataset has a beginning and an end.
 Using these definitions, the concept of an unbounded data stream contains that of a bounded dataset.
-That is, from an infinite stream, it is possible to fix a left and right limits to obtain a bounded and finite dataset.
+That is, from an infinite stream, it is possible to fix left and right limits to obtain a bounded and finite dataset.
 This relation is visible in Figure \ref{fig:unbounded-bounded-superset} from the same blog post.
@@
 
-In Section \ref{sec:windows} we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.
+In the next section, we will cover exactly how an unbounded dataset (also known as a data stream) can be converted into a bounded dataset with the use of windows.
 
@@
 
-In Section \ref{sec:stream-superset} we have defined a data stream as an unbounded dataset.
+In Section \ref{sec:stream-superset}, we have defined a data stream as an unbounded dataset.
 But what exactly are the issues of working with an infinite stream of data?
@@
 \begin{definition}
-A window W over a stream $\mathbb{S}$ is a finite subset of $\mathbb{S}$
+A window over a stream $\mathbb{S}$ is a finite subset of $\mathbb{S}$
 \end{definition}
@@
 An example of a tumbling window can be seen in Figure \ref{fig:tumbling-window} with a tuple-based window of size \textit{W} of three and an equally large step size \textit{S}.
-
 \begin{figure}[!htb]
@@
 Thus, approximate aggregators trade precision of results for efficient time and space usage.
-In Section \ref{sec:pds} we explore approximate aggregators that make such a trade-off.
+In Section \ref{sec:pds}, we explore approximate aggregators that make such a trade-off.
 
@@
 Doing this recursively for each item in the collection results in a single value: the aggregate value.
-The combine function is from here onwards represented as $\oplus$ and the identity or neutral element of the aggregation as $\overline{\theta}$.
+The \textit{combine} function is from here onwards represented as $\oplus$ and the identity or neutral element of the aggregation as $\overline{\theta}$.
 
@@
 If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:
-
 \begin{verbatim}
@@
 \end{verbatim}
-
 In the query above, the
@@
 Finally, for each rack and its list of temperatures, we compute an \texttt{AVG} aggregation, resulting in the average temperature per rack in the past 5 minutes.
-Note that this is very similar to any aggregation made in a standard Database Management System (DBMS).
+Note that this is very similar to any aggregation made in a standard Relational Database Management System (RDBMS).
 The main difference is that in standard SQL the \texttt{GROUP BY} clause includes only the columns on which the data will be grouped while in streaming SQL the \texttt{GROUP BY} also specifies the type of window used (\textit{e.g.}, in this case, a tumbling window).
@@
 However, given \textit{n} as the window size, this algorithm takes \textit{O(n)} space and time complexity which makes it unfit in most scenarios of large scale data processing.
-Recall that the combine function is represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.
-Pseudocode for the Recalculate-From-Scratch algorithm is shown in Algorithm \ref{pseudo:rfs}, assuming all functions have access to a \textit{vals} array which holds the window contents.
+Recall that the \textit{combine} function is represented by $\oplus$ and the identity or neutral element of the aggregation by $\overline{\theta}$.
+Pseudocode for the Recalculate-From-Scratch algorithm is shown in Algorithm \ref{pseudo:rfs}, assuming all functions have access to the \textit{vals} queue which holds the window contents.
 
@@
 However, aggregations are not always invertible thus making this algorithm very restrictive and not always applicable.
-Algorithm \ref{pseudo:soe} shows pseudocode for SOE assuming all functions have access to the window contents through the \textit{vals} array and access to the current aggregation state \textit{agg}.
+Algorithm \ref{pseudo:soe} shows pseudocode for SOE assuming all functions have access to the window contents through the \textit{vals} queue and access to the current aggregation state \textit{agg}.
 
@@
 Tangwongsan \emph{et al.} construct Table \ref{tbl:aggregations-properties} to summarize the aggregation families and satisfying properties \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}.
-
 To understand these properties, we must be familiar with the \textit{combine} and \textit{query} functions.
@@
 The \textit{query} function would return the current aggregation state --- \textit{i.e.}, return the total sum computed.
-
-Based on the \textit{combine} and \textit{query} functions, the following are formal definitions of these properties. An aggregation is said to have:
+Based on the \textit{combine} and \textit{query} functions, the following are formal definitions of these properties.
+An aggregation is said to have:
 
 \begin{enumerate}
-    \item  an \textbf{\textit{invertible combine}} if a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$
+    \item  an \textbf{\textit{invertible combine}}, if a function $\ominus$ exists such that $(x \oplus y) \ominus y = x$, for all $x$ and $y$;
 
-    \item  an \textbf{\textit{associative combine}} if $(x \oplus y) \oplus z = x \oplus (y \oplus z)$, for all $x$ and $y$
+    \item  an \textbf{\textit{associative combine}}, if $(x \oplus y) \oplus z = x \oplus (y \oplus z)$, for all $x$ and $y$;
 
-    \item  a \textbf{\textit{commutative combine}} if $x \oplus y = y \oplus x$, for all $x$ and $y$
+    \item  a \textbf{\textit{commutative combine}}, if $x \oplus y = y \oplus x$, for all $x$ and $y$;
 
-    \item  a \textbf{\textit{size-preserving combine}} if the aggregation state resulting from $\oplus$ always occupies the same space in memory
+    \item  a \textbf{\textit{size-preserving combine}}, if the aggregation state resulting from $\oplus$ always occupies the same space in memory;
 
-    \item  a \textbf{\textit{unary query}} if the query does not need any parameters --- \textit{e.g.}, a membership test with a Bloom filter would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters
+    \item  a \textbf{\textit{unary query}}, if the query does not need any parameters --- \textit{e.g.}, a membership test with a Bloom Filter (detailed in Section \ref{sec:bloom}) would require the element to test membership for as a parameter so it is not unary; returning the total sum is unary since it does not require any parameters.
 \end{enumerate}
@@
     \caption[Aggregation properties]{Aggregation properties.
-Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all, or false for some of a given group of like operations, respectively}
+Checkmarks (\checkmark), crosses ($\times$), and question marks (?) indicate a property is true for all, false for all or false for only some aggregations of that family, respectively \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}}
   \label{tbl:aggregations-properties}
@@
 
-In Section \ref{sec:windows} we discussed windowing mechanisms that allowed us to convert unbounded data streams (or infinite datasets) to finite datasets.
+In Section \ref{sec:windows}, we discussed windowing mechanisms that allowed us to convert unbounded data streams (or infinite datasets) to finite datasets.
 
-In Section \ref{sec:aggregations} we introduced aggregations over datasets.
+In Section \ref{sec:aggregations}, we introduced aggregations over datasets.
 We gave examples of computable aggregations over unbounded datasets at first, but then discussed sliding window aggregations, which make use of windowing techniques to obtain a finite dataset, on which the aggregation is computed.

### 2020-07-06 chapters/introduction.tex
Final changes and submission

@@
 Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously.
-Social networks are responsible for generating large volumes of data of which,  Twitter\footnote{https://twitter.com} for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
+Social networks are responsible for generating large volumes of data.
+Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
 \textit{Tweets} may contain text, media content or both.
@@
 In either processing scenario, we usually want to apply operations on data, for example, to compute the maximum value, an average or a count of distinct elements.
-Operations that produce a single result when applied to datasets are commonly known as aggregation operations.
+Operations that produce a single result when applied to datasets are commonly known as aggregations.
 In stream processing, we apply these aggregations to data streams to obtain valuable information.
 However, since a data stream is, effectively speaking, an unbounded dataset, if our aggregation requires a finite domain, there is the need to apply windowing techniques.
-A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size --- \textit{e.g.}, events from the last three days or the last million events, respectively.
+A thorough analysis of stream windowing techniques is done in Chapter \ref{chap:background} of this document but for now it suffices to say that windowing is a mechanism to select a fine portion of an unbounded data stream, typically by defining a time or tuple-based window size, \textit{e.g.}, events from the last three days or the last million events, respectively.
 
@@
 Therefore, it is common for companies to have auxiliary systems monitoring their services, testing their usage rate, availability and performance, just to name a few.
-The motivation behind this thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in the performance of the system analyzing said data stream.
+The motivation behind this thesis is to produce a system capable of real-time monitoring a stream of data so that changes in the underlying stream are caught early, preventing a fall in the performance of another system analyzing said data stream.
 
@@
 \end{itemize}
-The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic datasets and a real dataset from the financial fraud space.
+The hypothesis will be tested and evaluated on the previous criteria. Experiments will be made using multiple synthetic datasets and real data from the financial fraud space, provided by Feedzai.
 
@@
 
-In Chapter \ref{chap:statement}, we define the problem this thesis aims to solve, the issues with state of the art solutions, our proposed solution, the assumptions made on it and the research questions we intend to answer.
+In Chapter \ref{chap:statement}, we define the problem this thesis aims to solve, the issues with state of the art solutions, our proposed solution, the assumptions made with it and the research questions we intend to answer.
 
-In Chapter \ref{chap:my-work}, we discuss initial approaches and reasons to discard them before presenting our proposed solution.
+In Chapter \ref{chap:my-work}, we discuss initial approaches and the reasons that led us to discard them, before presenting our proposed solution.
 

### 2020-07-06 chapters/method.tex
Final changes and submission

@@
 
-\subsection{Exponential Moving Average Based Histogram} \label{sec:ema-hist}
+\subsection{Exponential Moving Average Histogram} \label{sec:ema-hist}
 As stated, we compute histogram aggregations for the reference and target sliding windows.

### 2020-07-06 chapters/problem-statement.tex
Final changes and submission

@@
 \section{Summary}
-In Section \ref{sec:theproblem}, we formally defined the problem we aim to solve with this thesis: alert streaming data pattern shifts in regards to a static reference period with the intent of helping system administrators reconfigure their systems before their performance declines.
-Furthermore, we make explicit we work under time and space constraints, inherent to the streaming paradigm.
+In Section \ref{sec:theproblem}, we formally defined the problem we aim to solve with this thesis: alert streaming data pattern shifts in regards to a static reference period with the intent of helping system administrators reconfigure their systems before their performance declines due to data pattern shifts.
+Furthermore, we make explicit we work under time and space constraints, inherent to the streaming paradigm and to the non-critical nature of our monitoring system.
 

### 2020-07-06 chapters/resumo.tex
Final changes and submission

@@
 %Approach/Solution and %Main contributions
-Nesta tese, projetamos um método que utiliza agregações de \textit{streaming} leves e funções de divergência para alertar sobre desvios nos padrões de dados em tempo real, na presença de grandes volumes de dados que fluem a alta velocidade, são altamente \textit{skewed} e sazonais.
+Nesta tese, desenvolvemos um método que utiliza agregações de \textit{streaming} leves e funções de divergência para alertar sobre desvios nos padrões de dados em tempo real, na presença de grandes volumes de dados que fluem a alta velocidade, são altamente \textit{skewed} e sazonais.
 Desenvolvemos um método de duas fases e duas janelas: primeiro, realizamos uma análise \textit{batch} numa janela de referência e, em seguida, uma análise \textit{streaming} sobre uma janela deslizante.
@@
 Medimos a divergência entre os histogramas de referência e os das janelas deslizantes baseados em MMEs para cada \textit{feature} com a função de divergência de \textit{Jensen-Shannon}, mas outras funções podem ser usadas.
-Para cada \textit{feature}, descobrimos o valor de probabilidade para cada medida de divergência e aplicamos a correção de múltiplo teste de Holm-Bonferroni para descobrir quais os valores de probabilidade mais baixos e as \textit{features} associados.
+Para cada \textit{feature}, descobrimos o valor de probabilidade para cada medida de divergência e aplicamos a correção de múltiplo teste de \textit{Holm-Bonferroni} para descobrir quais os valores de probabilidade mais baixos e as \textit{features} associados.
 Geramos alertas  para cada \textit{feature} com base num limite de probabilidade definido pelo administrador.
@@
 Dividimos ainda os testes em testes de análise a uma única \textit{feature} e tests de análise com múltiplas \textit{features}.
-O nosso método detectou com precisão as anomalias introduzidas nos testes com conjuntos de dados sintéticos, mantendo um alto \textit{throughput}, tanto para análises com uma única ou múltiplas \textit{features}.
+O nosso método detectou com precisão as anomalias introduzidas nos testes com conjuntos de dados sintéticos, mantendo um alto \textit{throughput}, tanto para análises com uma única ou com múltiplas \textit{features}.
 No entanto, os testes com dados reais não foram tão precisos.
-Apesar de não sabermos as razões específicas para isso, passamos os próximos testes para trabalhos futuros e formulamos um conjunto de hipóteses que pretendem explicar o sucedido e, portanto, são dignas de serem exploradas em seguida.
+Apesar de não sabermos as razões específicas para tal, passamos os próximos testes para trabalhos futuros e formulamos um conjunto de hipóteses que pretendem explicar o sucedido e, portanto, são dignas de serem exploradas em seguida.
 O nosso conjunto de testes apoia a afirmação de que métodos de agregação de janelas deslizantes e métodos de divergência de distribuição podem ser combinados para detectar alterações nos padrões de dados em cenários de \textit{streaming} com complexidade temporal e de memória constante.

### 2020-07-06 chapters/sota.tex
Final changes and submission

@@
 In Section \ref{sec:outliers}, we reviewed outlier detection algorithms.
-However, we found most of them inapplicable in our scenario, because they were resource-intensive (both time and memory-wise), provided no measure of how divergent and outlier was or because they evolved their reference period.
+However, we found most of them inapplicable in our scenario, because they were resource-intensive (both time and memory-wise), provided no measure of how divergent an outlier was or because they evolved their reference period.
 

### 2020-07-09 chapters/method.tex
EMA hist pseudocode and Final remarks analysis

@@
 where $n_{1/2}$ is commonly denoted as a tuple-based half-life.
+
+Algorithm \ref{pseudo:ema-hist} shows how we update our EMA histogram.
+We assume we have access to global \textit{counts} and \textit{bins} arrays and use a suppression factor of $2^{- \frac{1}{n_{1/2}}}$ with $n_{1/2}$ as the tuple-based half-life.
+The histogram aggregation is updated for each new streaming value \texttt{val} using the \texttt{processValue(val)} function, that increments the correct histogram bin (note that the \textit{bins} array defines the bin boundaries while the \textit{counts} array actually holds each bin's count) and applies the EMA suppression to every bin.
+\begin{algorithm}[!htb]
+    \caption[EMA histogram]{EMA histogram sliding window emulation}
+    \label{pseudo:ema-hist}
+    \begin{algorithmic}[1]
+        \Function{getBin}{val}
+            \State \Return{binarySearch(bins, val)}
+        \EndFunction
+
+        \Function{processValue}{val}
+            \State $bin \gets getBin(val)$
+            \State $counts[bin] \gets counts[bin] + 1$ \Comment{// Increment correct bin}
+
+            \ForEach {$count \in counts$} \Comment{// Apply EMA suppression to all bins}
+                \State $count \gets count \times 2^{- \frac{1}{n_{1/2}}}$
+            \EndFor
+        \EndFunction
+    \end{algorithmic}
+\end{algorithm}
 
@@
 When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives.
-In other words, when repeating a test multiple times the multiple test or simultaneous statistical inference problem \cite{MultiTestProblem-Pubmed}
-\cite{MultiTestProblem-Dickhaus2014}
-\cite{MultiTestProblem-Miller1966}  \cite{multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
+In other words, when repeating a test multiple times the multiple test or simultaneous statistical inference problem \cite{MultiTestProblem-Pubmed, MultiTestProblem-Dickhaus2014, MultiTestProblem-Miller1966, multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
 In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives.
@@
 Therefore we leave as future work the testing of other correction methods.
-In statistics, Family-Wise Error Rate (FWER) \cite{MultivariateMT} \cite{MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not.
+In statistics, Family-Wise Error Rate (FWER) \cite{MultivariateMT, MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not.
 The Holm-Bonferroni method \cite{HolmBonferroni} is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
@@
 \end{equation}
+
+\subsubsection*{Method classification}
+Similarly to the methods presented in Sections \ref{kifer} and \ref{sec:SAMM}, we use a two-windowed model that can take uni or multivariate input.
+However, we perform a multivariate analysis of all feature's time-series with the use of the Holm-Bonferroni correction.
+Hence, we say our method is multivariate.
+We also alert entire windows or subsequences as anomalous.
+Thus, according to the taxonomy presented in Section \ref{sec:outliers}, our method is a dissimilarity-based multivariate subsequence outlier detection method.
+
+\subsubsection*{Time complexity of the streaming phase}
+The critical phase of our method is the stream analysis because it has to process new events and alert for possible deviations in real-time.
+
+For each event and for each feature, we update our EMA histogram using Algorithm \ref{pseudo:ema-hist} which has to perform a binary search to find the correct bin to insert our new value and traverses all bins to apply a suppression factor.
+Using $n$ as the number of bins, we have a time complexity of $O(log(n) + n)$, for each feature.
+However, since $n$ is a fixed and small constant, \textit{i.e.}, more or less 100 bins, we say that this algorithm has constant time complexity.
+In other words, our time complexity does not depend on the size of the emulated streaming windows.
+
+\subsubsection*{Space complexity of the streaming phase}
+We store $n$ bins per feature, so we have a space complexity of $O(n)$ per feature.
+Again, since $n$ is a small and fixed constant, we claim to have constant space complexity for ever-growing volumes of data.

### 2020-07-09 paper.tex
Initial paper boilerplate

%%%% Proceedings format for most of ACM conferences (with the exceptions listed below) and all ICPS volumes.
\documentclass[sigconf]{acmart}
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[group-separator={,}]{siunitx}

%%%% As of March 2017, [siggraph] is no longer used. Please use sigconf (above) for SIGGRAPH conferences.

%%%% Proceedings format for SIGPLAN conferences 
% \documentclass[sigplan, anonymous, review]{acmart}

%%%% Proceedings format for SIGCHI conferences
% \documentclass[sigchi, review]{acmart}

%%%% To use the SIGCHI extended abstract template, please visit
% https://www.overleaf.com/read/zzzfqvkmrfzn

%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% Rights management information. 
% This information is sent to you when you complete the rights form.
% These commands have SAMPLE values in them; it is your responsibility as an author to replace
% the commands and values with those provided to you when you complete the rights form.
%
% These commands are for a PROCEEDINGS abstract or paper.
\copyrightyear{}
\acmYear{}
\setcopyright{none}
\acmConference[KDD-ADF-2019]{2nd KDD Workshop on Anomaly Detection in Finance}{August 5, 2019}{Anchorage, Alaska, USA}
\acmBooktitle{}
\acmPrice{}
\acmDOI{}
\acmISBN{}
\acmISBN{}

%
% These commands are for a JOURNAL article.
%\setcopyright{none}
%\acmJournal{TOG}
%\acmYear{2018}\acmVolume{37}\acmNumber{4}\acmArticle{111}\acmMonth{8}
%\acmDOI{10.1145/1122445.1122456}

%
% Submission ID. 
% Use this when submitting an article to a sponsored event. You'll receive a unique submission ID from the organizers
% of the event, and this ID should be used as the parameter to this command.
%\acmSubmissionID{123-A56-BU3}

%
% The majority of ACM publications use numbered citations and references. If you are preparing content for an event
% sponsored by ACM SIGGRAPH, you must use the "author year" style of citations and references. Uncommenting
% the next command will enable that style.
%\citestyle{acmauthoryear}

%
% end of the preamble, start of the body of the document source.
\begin{document}

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Lightweight Real-Time Feature Monitoring}

%
% The "author" command and its associated commands are used to define the authors and their affiliations.
% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
% used to denote shared contribution to the research.

\author[Conde]{João Conde}
\email{joao.conde@feedzai.com}
\affiliation{%
  \institution{Feedzai}
}

\author[Sampaio]{Marco O. P. Sampaio}
\email{marco.sampaio@feedzai.com}
\affiliation{%
  \institution{Feedzai}
}

\author[Cardoso]{Pedro Cardoso}
\email{pedro.cardoso@feedzai.com}
\affiliation{%
	\institution{Feedzai}
}

\author[Ribeiro]{Pedro Manuel Pinto Ribeiro}
\email{pribeiro@dcc.fc.up.pt}
\affiliation{%
  \institution{FCUP}
}

\author[Restivo]{André Monteiro de Oliveira Restivo}
\email{arestivo@fe.up.pt}
\affiliation{%
	\institution{FEUP}
}

%
% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
% other information printed in the page headers. This command allows the author to define a more concise list
% of authors' names for this purpose.

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
% The problem
Many real-time stream monitoring systems are static once deployed in a production environment. The engineers of those systems configure them under the assumption that future data flowing through the system roughly follows the same distribution as previously seen data. They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system. Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates. Data pattern shift detection refers to the process of finding patterns in data that do not conform to expected or usual behavior. Accurate and timely detection of data pattern deviations allows for immediate measures to be taken. Thus, the problem at hand is to determine when to reconfigure the system, \textit{e.g.}, a Machine Learning model, based on an analysis of the drifts in the stream of data.

%Approach/Solution and %Main contributions
In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data. We develop a two-phased and two-windowed method: first, we perform a batch analysis on a reference window and then a stream analysis over a sliding streaming window. We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs). We do this for each event and each of its fields, also denominated features in our context. We measure the divergence between both reference and sliding window EMA-based histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used. For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and their associated features. We generate alerts based on a user-defined probability threshold, for each feature.

We evaluate our method through a series of tests, some with synthetic datasets and some with real data. We further split the tests into single and multi-feature analysis. Our method accurately detected the introduced anomalies in the experiments using synthetic datasets while maintaining high throughput, for single and multi-feature analysis. However, experiments with real data were not as accurate. Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next. Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
\end{abstract}

%
% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

%\ccsdesc[500]{Computer systems organization~Embedded systems}
%\ccsdesc[300]{Computer systems organization~Redundancy}
%\ccsdesc{Computer systems organization~Robotics}
%\ccsdesc[100]{Networks~Network reliability}

%
% Keywords. The author(s) should pick words that accurately describe the work being
% presented. Separate the keywords with commas.
\keywords{data streams, monitoring, real-time, lightweight, concept drift}

%
% This command processes the author and affiliation and title information and builds
% the first part of the formatted document.
\maketitle

\section{Introduction}
Nice Intro

\section{Related Work}
\label{sec:RelatedWork}
Nice related work

\section{Method}
\label{sec:Method}

Such method wow

\section{Experiments}
\label{sec:Experiments}

Much scientist

\section{Discussion}
\label{sec:Discussion}

Fight

\section{Conclusions}
\label{sec:Conclusions}
\cite{ApacheFlink}
Thesis done, PogU

%
% The next two lines define the bibliography style to be used, and the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{myrefs.bib}

\end{document}

### 2020-07-13 paper.tex
Changed paper template

@@
-%%%% Proceedings format for most of ACM conferences (with the exceptions listed below) and all ICPS volumes.
+% The first command in your LaTeX source must be the \documentclass command.
 \documentclass[sigconf]{acmart}
-\usepackage{graphicx}
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following line
-% to display URLs in blue roman font according to Springer's eBook style:
-% \renewcommand\UrlFont{\color{blue}\rmfamily}
+\usepackage{listings}
+\usepackage{dirtytalk}
+\usepackage[shortlabels]{enumitem}
+\usepackage{amssymb}
+\usepackage{minitoc}
 \usepackage{algorithm}
 \usepackage[noend]{algpseudocode}
-\usepackage{amsmath}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage[group-separator={,}]{siunitx}
+\usepackage{mathtools}
 
-%%%% As of March 2017, [siggraph] is no longer used. Please use sigconf (above) for SIGGRAPH conferences.
+% pseudocode
+\algnewcommand\algorithmicforeach{\textbf{for each}}
+\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
 
-%%%% Proceedings format for SIGPLAN conferences 
-% \documentclass[sigplan, anonymous, review]{acmart}
-
-%%%% Proceedings format for SIGCHI conferences
-% \documentclass[sigchi, review]{acmart}
-
-%%%% To use the SIGCHI extended abstract template, please visit
-% https://www.overleaf.com/read/zzzfqvkmrfzn
+\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
 
 %
-% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+% \BibTeX command to typeset BibTeX logo in the docs
+\AtBeginDocument{%
+  \providecommand\BibTeX{{%
+    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
 
@@
 % These commands are for a PROCEEDINGS abstract or paper.
-\copyrightyear{}
-\acmYear{}
-\setcopyright{none}
-\acmConference[KDD-ADF-2019]{2nd KDD Workshop on Anomaly Detection in Finance}{August 5, 2019}{Anchorage, Alaska, USA}
-\acmBooktitle{}
-\acmPrice{}
-\acmDOI{}
-\acmISBN{}
-\acmISBN{}
+\copyrightyear{2020}
+\acmYear{2020}
+\setcopyright{acmlicensed}
+\acmConference[MileTS '20]{MileTS '20: 6th KDD Workshop on Mining and Learning from Time Series}{August 24th, 2020}{San Diego, California, USA}
+\acmBooktitle{MileTS '20: 6th KDD Workshop on Mining and Learning from Time Series, August 24th, 2020, San Diego, Claifornia, USA}
+\acmPrice{15.00}
+\acmDOI{10.1145/1122445.1122456}
+\acmISBN{978-1-4503-9999-9/18/06}
 
@@
 % These commands are for a JOURNAL article.
-%\setcopyright{none}
+%\setcopyright{acmcopyright}
 %\acmJournal{TOG}
@@
 % used to denote shared contribution to the research.
-
 \author[Conde]{João Conde}
@@
 
-\author[Ribeiro]{Pedro Manuel Pinto Ribeiro}
+\author[Ribeiro]{Pedro Ribeiro}
 \email{pribeiro@dcc.fc.up.pt}
 \affiliation{%
-  \institution{FCUP}
+  \institution{University of Porto}
 }
 
-\author[Restivo]{André Monteiro de Oliveira Restivo}
+\author[Restivo]{André Restivo}
 \email{arestivo@fe.up.pt}
 \affiliation{%
-	\institution{FEUP}
+	\institution{University of Porto}
 }
@@
 % of authors' names for this purpose.
+\renewcommand{\shortauthors}{Trovato and Tobin, et al.}
 
@@
 \begin{abstract}
-% The problem
 Many real-time stream monitoring systems are static once deployed in a production environment.
-The engineers of those systems configure them under the assumption that future data flowing through the system roughly follows the same distribution as previously seen data.
-They do so consciously, to the best of their knowledge and available tools, keeping in mind that in the future they may need to reconfigure the system.
-Thus, even though the initial configuration of the system may be one of the best fits, over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.
+Over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.
 Data pattern shift detection refers to the process of finding patterns in data that do not conform to expected or usual behavior.
@@
 Thus, the problem at hand is to determine when to reconfigure the system, \textit{e.g.}, a Machine Learning model, based on an analysis of the drifts in the stream of data.
-
-%Approach/Solution and %Main contributions
-In this thesis, we design a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time, while in the presence of large volumes of high velocity, highly skewed and seasonal data.
-We develop a two-phased and two-windowed method: first, we perform a batch analysis on a reference window and then a stream analysis over a sliding streaming window.
-We aggregate the contents of both windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs).
-We do this for each event and each of its fields, also denominated features in our context.
-We measure the divergence between both reference and sliding window EMA-based histograms for each feature with the Jensen–Shannon Divergence function, but other distance functions could be used.
-For each feature, we find out the probability value of each divergence measure and apply the Holm-Bonferroni multiple test correction to find out which probability values are lowest and their associated features.
-We generate alerts based on a user-defined probability threshold, for each feature.
-
+In this paper, we present a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time.
+We generate alerts based on a user-defined probability threshold, for each of the event's fields, also denominated features in our context.
 We evaluate our method through a series of tests, some with synthetic datasets and some with real data.
@@
    <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+       <concept_id>10002951.10003227.10003351.10003446</concept_id>
+       <concept_desc>Information systems~Data stream mining</concept_desc>
        <concept_significance>500</concept_significance>
@@
    <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
+       <concept_id>10010520.10010570.10010574</concept_id>
+       <concept_desc>Computer systems organization~Real-time system architecture</concept_desc>
+       <concept_significance>500</concept_significance>
        </concept>
@@
 
-%\ccsdesc[500]{Computer systems organization~Embedded systems}
-%\ccsdesc[300]{Computer systems organization~Redundancy}
-%\ccsdesc{Computer systems organization~Robotics}
-%\ccsdesc[100]{Networks~Network reliability}
+\ccsdesc[500]{Information systems~Data stream mining}
+\ccsdesc[500]{Computer systems organization~Real-time system architecture}
 
@@
 \section{Introduction}
-Nice Intro
+In the past decade, applications have become increasingly data-driven, placing data at the center of application design.
+For different use cases, the data is processed in different ways for different purposes.
+For instance, e-commerce platforms such as Alibaba\footnote{https://alibaba.com} need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes.
+Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously.
+Social networks are responsible for generating large volumes of data.
+Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
+\textit{Tweets} may contain text, media content or both.
+Cybersecurity applications are yet another example of time-sensitive, data-driven applications.
+The entities that work in these use cases need to monitor user accesses and their actions in the network, where timely detection of intruders is critical to prevent them from tampering with the underlying system.
+
+All of these applications generate large volumes of data which lead to the creation of multiple real-time unbounded datasets, also known as data streams.
+The information that flows through such a stream can be analyzed on-the-fly or stored for later processing.
+The former is best known as stream processing, whereas the latter is known as batch processing.
+Streaming data is known for being non-stationary \cite{Gama-Knowledge-Discovery} where the value of the produced information lies in its recency \cite{Kolajo-Big-data-stream-SLR}.
+\textit{Recency} is measured under different scales for different use cases.
+For instance, consider the monitoring of geological data to forecast possible natural disasters such as earthquakes versus the monitoring of a computer network for intruder detection.
+In the former case, after the forecast of a future earthquake, governmental authorities need a couple of days to launch an evacuation plan and keep the population safe.
+However, in the intruder detection scenario, the decision of removing access to a user must be done as soon as possible, preferably in a few seconds.
+Hence, while in the first scenario information retrieved within a day would still be recent, in the second scenario, information is considered recent and relevant if delivered within seconds (or even milliseconds).
+
+Consider the use case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day.
+In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one.
+Such a system would have to process all incoming transactions and decide, in a fraction of a second, if a transaction is fraudulent or not.
+Unlike the fraud detection system, our system is not mission-critical.
+Hence, it must monitor the data stream looking for data pattern deviations in real-time but with a low memory footprint.
+
+To summarize, in this paper we design a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.
+The motivation to detect data stream pattern deviations in real-time is to know when to reconfigure the mission-critical system (\textit{e.g.}, retrain the Machine Learning model) before its performance decays.
+We want to build a lightweight enough solution that can be integrated into existing workflows since having a real-time system monitoring other real-time systems consuming as many resources as the latter would be far too costly.
 
@@
 \label{sec:RelatedWork}
-Nice related work
 
@@
 \label{sec:Method}
+A data stream is a continuous collection of events, each timestamped and with multiple fields.
+We denote each of the fields as features.
+Our goal is to detect data pattern deviations for each feature in a streaming fashion.
+Such shifts will be measured between a static in time reference period versus the observed reality during the streaming phase.
+Figure \ref{fig:timelines} shows the separated reference and streaming timelines.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/timeperiods.png}
+      \caption[]{Reference and streaming timelines}
+      \label{fig:timelines}
+    \end{center}
+\end{figure}
+To detect data pattern shifts, we do not need to directly compare the entire window contents of the reference and sliding windows.
+This approach would require storing all window contents hence making the memory consumption grow linearly with target sliding window and/or reference period size.
+Instead, we want to use a \textit{(a)} fixed cost in-memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
+Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.
 
-Such method wow
+One aggregation that encodes distributions properly are histograms.
+Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}.
+Histograms require storing a fixed-size list of bins and corresponding counts per bin regardless of the volume of data, satisfying property \textit{(a)}.
+Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and incrementing or decrementing the associated count, respectively, thus satisfying property \textit{(b)}.
+Recall Section \ref{sec:back-swag-algs} where we discussed generic sliding window aggregation algorithms and remember the Subtract-On-Evict (SOE) algorithm (Algorithm \ref{pseudo:soe}).
+Computing a histogram aggregation over a sliding window of size $W$ using SOE implies storing all $W$ elements in memory to know the element to remove after a new one is inserted.
+This violates our constant in-memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}.
+EMA-like histograms computed over sliding windows do not require the storage of window contents.
+An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values.
+These EMA-like histograms are further explained in Section \ref{sec:ema-hist}.
+
+Since the reference window is static, \textit{i.e.}, it is not a sliding window, we can use exact histogram aggregations.
+Hence, for each feature, we build an exact reference histogram aggregation.
+On the other hand, in streaming, we use one EMA-like approximate histogram aggregation per feature (Figure \ref{fig:timelines-hists}).
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/timelines-hists.png}
+      \caption[]{Reference and Streaming timelines and corresponding histogram aggregations}
+      \label{fig:timelines-hists}
+    \end{center}
+\end{figure}
+
+To compare each feature's reference and target sliding window histogram aggregation we need a measure of divergence.
+We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein \cite{EncyclopaediaMathematics}.
+The JSD takes as input two histograms and outputs a value between 0 and 1.
+The closer this value is to 0 the more similar the distributions are.
+On the other hand, a JSD value of 1 represents two different distributions.
+We chose to use the JSD metric due to its simplicity and because it was used in previous work successfully \cite{SAMM}, leaving as future work other possible tests.
+
+Given a divergence or distance value for the two aggregations (reference and target) we need a measure of normal statistical fluctuations.
+Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
+This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand-picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).
+
+To implement the proposed methodology, we devised a two-phase method.
+The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}).
+The second phase (detailed in Section \ref{sec:stream-phase}) focuses on efficiently updating the target sliding window histogram for each feature and periodically performing a divergence test between the static reference aggregation and the target sliding one, alerting for changes if need be.
+The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.
+
+\subsection{Exponential Moving Average Histogram} \label{sec:ema-hist}
+As stated, we compute histogram aggregations for the reference and target sliding windows.
+The computation of the reference histogram is done once in batch.
+Since the reference data is fixed and does not change over time, computing this histogram in batch is more efficient and easier to implement in a distributed system like Spark.
+In addition, a batch computation also allows us to compute an exact reference histogram.
+However, the target sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.
+
+In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they compute sliding window aggregations without actually storing the window contents.
+Definition \ref{def:tuple-ema} shows how to compute an EMA using its recursive formula.
+An EMA state or value depends only on the previous one multiplied by an exponential decay.
+Recall that a tuple-based decay is given by:
+\[ 2^{- \frac{1}{n_{1/2}}} \]
+where $n_{1/2}$ is commonly denoted as a tuple-based half-life.
+
+Algorithm \ref{pseudo:ema-hist} shows how we update our EMA histogram.
+We assume we have access to global \textit{counts} and \textit{bins} arrays and use a suppression factor of $2^{- \frac{1}{n_{1/2}}}$ with $n_{1/2}$ as the tuple-based half-life.
+The histogram aggregation is updated for each new streaming value \texttt{val} using the \texttt{processValue(val)} function, that increments the correct histogram bin (note that the \textit{bins} array defines the bin boundaries while the \textit{counts} array actually holds each bin's count) and applies the EMA suppression to every bin.
+\begin{algorithm}[!htb]
+    \caption[EMA histogram]{EMA histogram sliding window emulation}
+    \label{pseudo:ema-hist}
+    \begin{algorithmic}[1]
+        \Function{getBin}{val}
+            \State \Return{binarySearch(bins, val)}
+        \EndFunction
+
+        \Function{processValue}{val}
+            \State $bin \gets getBin(val)$
+            \State $counts[bin] \gets counts[bin] + 1$ \Comment{// Increment correct bin}
+
+            \ForEach {$count \in counts$} \Comment{// Apply EMA suppression to all bins}
+                \State $count \gets count \times 2^{- \frac{1}{n_{1/2}}}$
+            \EndFor
+        \EndFunction
+    \end{algorithmic}
+\end{algorithm}
+
+After processing \textit{x} tuples the oldest tuple is \emph{smoothed} by a factor of $2^{- \frac{x}{n_{1/2}}}$.
+If we process $x = 4n_{1/2}$ tuples the oldest tuple would have its contribution reduced by
+\[ 2^{- \dfrac{4n_{1/2}}{n_{1/2}}} = 2^{-4} \]
+which is approximately 0.06. In other words, the \textit{x-th} event would have 6\% contribution towards the EMA state or value meaning older events would have even less weight.
+
+We choose to emulate windows with four times the half-life size (or $4n_{1/2}$), meaning we discard events whose contribution is less than 6\%.
+We set the half-life to be one-quarter of the size of the window we want to emulate.
+The size of the window we want to monitor varies from situation to situation, but in the limited scope of this thesis we set to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data.
+This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.
+
+\subsection{Batch Analysis Phase} \label{sec:batch-phase}
+The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and EMA-like aggregations.
+In other words, we:
+
+\begin{enumerate}
+    \item Obtain a reference histogram for each feature using all data in the reference period
+
+    \item Compute the distribution of distances between the reference histogram and random windows of data and corresponding histograms
+\end{enumerate}
+
+This distribution of values can then later be compared to a given value of divergence for a single sample to find out how probable that value is to occur (if it is very low an alert may be raised).
+
+\subsubsection*{Building the Reference Histogram for each Feature}
+
+First, for each feature, we compute the reference histogram from the reference period dataset.
+This is an exact histogram aggregation.
+When building it we ensure the histogram has equal counts (or equal height) for all bins which often results in different sized bins.
+We make no assumption on the dataset distributions and hence use equal height histograms because they adjust better to wildly varying ones.
+Equal height means we have more bins covering very dense regions (regions with many data points) and fewer bins in lower density regions.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.3]{figures/ref-hist.png}
+      \caption[]{Reference equal-height histogram}
+      \label{fig:ref-hist}
+    \end{center}
+\end{figure}
+
+\subsubsection*{Finding the Distribution of Distances for each Feature} \label{sec:sampling-batch}
+
+Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase.
+To that end, we make $S$ samples of transactions (in Section \ref{sec:nsamples} we discuss the minimum number $S$ of samples to make), each with the same tuple-based size of the target sliding window we will use in streaming.
+Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
+
+For each sample, and each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}).
+We also add two extra bins that will cover the region to the left and the right of our histogram, respectively.
+In other words, we add a bin that covers the region between $-\infty$ and our first bin and another that covers values from the last bin to $+\infty$.
+This way, observed values outside of the reference period will be placed in these special bins, either left or right.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.3]{figures/ref-bins.png}
+      \caption[]{Reference equal-height histogram}
+      \label{fig:ref-hist-bins}
+    \end{center}
+\end{figure}
+
+Each of the histograms is an EMA-like histogram.
+An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}.
+Each sample is processed event by event and a discount factor is applied per event to each bin (EMA expiration factor), thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample.
+This procedure is illustrated in Figure \ref{fig:EMA-hist-build}.
+\begin{figure}[!htb]
+    \begin{center}
+    %\hspace{2cm}
+      \includegraphics[scale=0.8]{figures/ema-hist.png}
+      \caption[]{Reference equal-height histogram}
+      \label{fig:EMA-hist-build}
+    \end{center}
+\end{figure}
+We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period.
+We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.
+
+For each feature, we have now \textit{S} histograms, one for each sample.
+For each of the \textit{S} histograms, we compute the distance between it and the reference histogram.
+Figure \ref{fig:compute-sample-distances} illustrates this process as we apply our distance function for each sample histogram and the reference histogram, obtaining \textit{S} distance measurements.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.4]{figures/compute-sample-distances.png}
+      \caption[Compute sample's distance values]{Compute as many distance values as samples, one for each sample-reference histogram pair}
+      \label{fig:compute-sample-distances}
+    \end{center}
+\end{figure}
+
+For each feature, we end up with \textit{S} distance values.
+These $S$ measurements are distance measurements between random samples of data and the reference period.
+Hence, we claim we now know the distribution of expected distance values.
+Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if the probability is below a certain threshold.
+Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the target sliding window histogram.
+This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.
+
+It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself.
+For instance, given that JSD values vary between 0 and 1, why not just set to alert if the measured JSD value is 0.7?
+The reason is that a divergence value may fall on different percentiles for different features.
+In other words, the same \textit{i-th} percentile will correspond to a different divergence value for different features.
+Consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.2]{figures/skewed-left-distro.png}
+      \caption[]{Left skewed distribution of distance values}
+      \label{fig:skewed-left-distro}
+    \end{center}
+\end{figure}
+
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.2]{figures/skewed-right-distro.png}
+      \caption[]{Right skewed distribution of distance values}
+      \label{fig:skewed-right-distro}
+    \end{center}
+\end{figure}
+
+Let's assume that the histogram from Figure \ref{fig:skewed-left-distro} encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}.
+Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results.
+For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}.
+Instead, we define our threshold as a percentile of the distribution and not as a hard-coded distance constant.
+For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
+
+Hard-coding a threshold value to be used for all features is cumbersome, impractical and error-prone. The distance threshold must be computed based on the distribution of distances.
+
+\subsubsection*{Burn-In Period at System Initialization}
+When we boot up our system in the streaming phase we initially have empty histogram aggregations.
+The period of time where you discard the alerts produced by the system until it processes enough data to produce accurate reports is commonly denoted as the burn-in period.
+To avoid this burn-in period we propose that the target sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
+
+\subsubsection*{Batch Phase Artifacts} \label{sec:batch-artifacts-summary}
+By the end of the batch phase, we should have, for each feature: 
+\begin{itemize}
+    \item a reference histogram
+    \item a list of distance values that represent its distribution
+    \item the last sample's histogram, to be used as burn-in period for the target sliding window histogram
+\end{itemize}
+
+\subsection{Streaming Phase} \label{sec:stream-phase}
+In the streaming or online phase, the system will process event by event in a true sliding window fashion (recall Section \ref{sec:windows}) and perform a change detection test periodically.
+This phase makes use of all of the batch phase artifacts, namely, for each feature, the reference histogram, the distribution of distance values and the last sample's histogram.
+
+For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase.
+Then, for each incoming event and each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
+This is all we do, regarding event processing.
+
+In order to obtain alerts, we periodically perform a multi-feature change detection test.
+The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the target sliding window histogram of the same feature.
+This distance value is then tested against the known distribution of distance values for feature \textit{x}.
+In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile, \textit{e.g.}, the 99th-percentile.
+When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and target sliding window histogram aggregations.
+The probability value or \textit{p-value} of a distance value is:
+
+\[\textit{p-value\textsubscript{distance\textsubscript{x}}} = 1 - percentile(\textit{distance\textsubscript{x}})\]
+
+For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01.
+Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
+\[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]
+or:
+\begin{equation}
+    \label{eq:alert-test}
+    \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha
+\end{equation}
+
+\subsubsection*{Multiple Test Correction} \label{sec:multi-test}
+Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis.
+When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives.
+In other words, when repeating a test multiple times the multiple test or simultaneous statistical inference problem \cite{MultiTestProblem-Pubmed, MultiTestProblem-Dickhaus2014, MultiTestProblem-Miller1966, multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
+In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives.
+Furthermore, it does not require to assume any independence between the tested hypothesis (as other methods do) which would almost surely be a wrong assumption for realistic datasets.
+Therefore we leave as future work the testing of other correction methods.
+In statistics, Family-Wise Error Rate (FWER) \cite{MultivariateMT, MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not.
+The Holm-Bonferroni method \cite{HolmBonferroni} is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
+
+\subsubsection*{Holm-Bonferroni Correction} \label{sec:holmbonferroni}
+The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.
+This correction is done by dividing the threshold $\alpha$ by a correction factor, as described next.
+
+First, we compute the \textit{p-value} of each feature.
+We then order our features by \textit{p-value}, in ascending order.
+Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if:
+\begin{equation}
+    p-value_{distance_{x}} < \alpha / (m - k)
+    \label{eq:corrected-pvalue}
+\end{equation}
+with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature.
+We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.
+
+\subsection{Final Remarks}
+
+\subsubsection*{Minimum Number of Samples} \label{sec:nsamples}
+In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase.
+But what is the optimal number of samples to make?
+How can we mathematically derive a formula for the number of samples to make and avoid \textit{ad hoc} user-defined constants?
+
+When checking for alerts we are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests.
+We apply a multiple test correction on the tests performed and want to set the FWER to $\alpha$.
+In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.
+
+We use the Holm-Bonferroni correction and, for each feature, we need to reliably estimate the upper tail of the distribution which corresponds to the \textit{$\alpha$ / n\textsubscript{features}} region.
+Each sample we do translates into a distance value and the collection of these data points builds the distribution of distance values.
+We want to be sure that the number of samples \textit{S} that we do generates enough distance points to have a high probability of having one or more points in the upper tail region.
+To that end, we define the probability of not having at least one point in the upper tail of the distance distribution as $\gamma$.
+
+In \cite{SAMM}, the authors prove that the minimum number of samples $S$ needed to satisfy these probability requirements is:
+\begin{equation}
+    \label{eq:optimal-n-samples}
+    \textit{S} = \ceil[\bigg]{\frac{\log\gamma}{log(1 - \frac{\alpha}{n_{features}})}}
+\end{equation}
+
+\subsubsection*{Method classification}
+Similarly to the methods presented in Sections \ref{kifer} and \ref{sec:SAMM}, we use a two-windowed model that can take uni or multivariate input.
+However, we perform a multivariate analysis of all feature's time-series with the use of the Holm-Bonferroni correction.
+Hence, we say our method is multivariate.
+We also alert entire windows or subsequences as anomalous.
+Thus, according to the taxonomy presented in Section \ref{sec:outliers}, our method is a dissimilarity-based multivariate subsequence outlier detection method.
+
+\subsubsection*{Time complexity of the streaming phase}
+The critical phase of our method is the stream analysis because it has to process new events and alert for possible deviations in real-time.
+
+For each event and for each feature, we update our EMA histogram using Algorithm \ref{pseudo:ema-hist} which has to perform a binary search to find the correct bin to insert our new value and traverses all bins to apply a suppression factor.
+Using $n$ as the number of bins, we have a time complexity of $O(log(n) + n)$, for each feature.
+However, since $n$ is a fixed and small constant, \textit{i.e.}, more or less 100 bins, we say that this algorithm has constant time complexity.
+In other words, our time complexity does not depend on the size of the emulated streaming windows.
+
+\subsubsection*{Space complexity of the streaming phase}
+We store $n$ bins per feature, so we have a space complexity of $O(n)$ per feature.
+Again, since $n$ is a small and fixed constant, we claim to have constant space complexity for ever-growing volumes of data.
 
@@
 
-%
-% The next two lines define the bibliography style to be used, and the bibliography file.
+\section{CCS Concepts and User-Defined Keywords}
+
+Two elements of the ``acmart'' document class provide powerful taxonomic tools for you to help readers find your work in an online search.
+
+The ACM Computing Classification System --- \url{https://www.acm.org/publications/class-2012} --- is a set of classifiers and concepts that describe the computing discipline.
+Authors can select entries from this classification system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the commands to be included in the \LaTeX\ source.
+
+User-defined keywords are a comma-separated list of words and phrases of the authors' choosing, providing a more flexible way of describing the research being presented.
+
+CCS concepts and user-defined keywords are required for all short- and full-length articles, and optional for two-page abstracts.
+
+\section{Tables}
+
+The ``\verb|acmart|'' document class includes the ``\verb|booktabs|'' package --- \url{https://ctan.org/pkg/booktabs} --- for preparing high-quality tables.
+
+Table captions are placed {\it above} the table.
+
+Because tables cannot be split across pages, the best placement for them is typically the top of the page nearest their initial cite.
+To ensure this proper ``floating'' placement of tables, use the environment \textbf{table} to enclose the table's contents and the table caption.
+The contents of the table itself must go in the \textbf{tabular} environment, to be aligned properly in rows and columns, with the desired horizontal and vertical rules.
+Again, detailed instructions on \textbf{tabular} material are found in the \textit{\LaTeX\ User's Guide}.
+
+Immediately following this sentence is the point at which Table~\ref{tab:freq} is included in the input file; compare the placement of the table here with the table in the printed output of this document.
+
+\begin{table}
+  \caption{Frequency of Special Characters}
+  \label{tab:freq}
+  \begin{tabular}{ccl}
+    \toprule
+    Non-English or Math&Frequency&Comments\\
+    \midrule
+    \O & 1 in 1,000& For Swedish names\\
+    $\pi$ & 1 in 5& Common in math\\
+    \$ & 4 in 5 & Used in business\\
+    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
+  \bottomrule
+\end{tabular}
+\end{table}
+
+To set a wider table, which takes up the whole width of the page's live area, use the environment \textbf{table*} to enclose the table's contents and the table caption.
+As with a single-column table, this wide table will ``float'' to a location deemed more desirable.
+Immediately following this sentence is the point at which Table~\ref{tab:commands} is included in the input file; again, it is instructive to compare the placement of the table here with the table in the printed output of this document.
+
+\begin{table*}
+  \caption{Some Typical Commands}
+  \label{tab:commands}
+  \begin{tabular}{ccl}
+    \toprule
+    Command &A Number & Comments\\
+    \midrule
+    \texttt{{\char'134}author} & 100& Author \\
+    \texttt{{\char'134}table}& 300 & For tables\\
+    \texttt{{\char'134}table*}& 400& For wider tables\\
+    \bottomrule
+  \end{tabular}
+\end{table*}
+
+\section{Figures}
+
+The ``\verb|figure|'' environment should be used for figures.
+One or more images can be placed within a figure.
+If your figure contains third-party material, you must clearly identify it as such, as shown in the example below.
+\iffalse
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{sample-franklin}
+  \caption{1907 Franklin Model D roadster. Photograph by Harris \& Ewing, Inc. [Public domain], via Wikimedia Commons. (\url{https://goo.gl/VLCRBB}).}
+  \Description{The 1907 Franklin Model D roadster.}
+\end{figure}
+\fi 
+Your figures should contain a caption which describes the figure to the reader.
+Figure captions go below the figure.
+Your figures should {\bf also} include a description suitable for screen readers, to assist the visually-challenged to better understand your work.
+
+Figure captions are placed {\it below} the figure.
+
 \bibliographystyle{ACM-Reference-Format}

### 2020-07-14 paper.tex
New paper template, new experiment figures

@@
 Data pattern shift detection refers to the process of finding patterns in data that do not conform to expected or usual behavior.
-Accurate and timely detection of data pattern deviations allows for immediate measures to be taken.
+Accurate and timely detection of data pattern deviations allow for immediate measures to be taken.
 Thus, the problem at hand is to determine when to reconfigure the system, \textit{e.g.}, a Machine Learning model, based on an analysis of the drifts in the stream of data.
-In this paper, we present a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time.
-We generate alerts based on a user-defined probability threshold, for each of the event's fields, also denominated features in our context.
-We evaluate our method through a series of tests, some with synthetic datasets and some with real data.
+In this paper, we present a method that makes use of constant time & memory streaming aggregations and distribution divergence functions to alert deviations in data patterns in real-time.
+We generate alerts based on a user-defined probability threshold, for each of an event's fields, denominated features in our context.
+We evaluate our method through a series of tests, some with synthetic data sets and some with real data.
 We further split the tests into single and multi-feature analysis.
@@
 However, experiments with real data were not as accurate.
-Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next.
-Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
+Despite not knowing the specific reasons for that, we detail further investigation to future work and formulate a set of hypotheses that might explain the latter results and hence are worthy of pursuing next.
+Moreover, our set of experiments support the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
 \end{abstract}
@@
 A data stream is a continuous collection of events, each timestamped and with multiple fields.
-We denote each of the fields as features.
+In this paper, we refer to the event's fields as features.
 Our goal is to detect data pattern deviations for each feature in a streaming fashion.
-Such shifts will be measured between a static in time reference period versus the observed reality during the streaming phase.
-Figure \ref{fig:timelines} shows the separated reference and streaming timelines.
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/timeperiods.png}
-      \caption[]{Reference and streaming timelines}
-      \label{fig:timelines}
-    \end{center}
-\end{figure}
-To detect data pattern shifts, we do not need to directly compare the entire window contents of the reference and sliding windows.
-This approach would require storing all window contents hence making the memory consumption grow linearly with target sliding window and/or reference period size.
-Instead, we want to use a \textit{(a)} fixed cost in-memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase.
-Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.
-
-One aggregation that encodes distributions properly are histograms.
-Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}.
-Histograms require storing a fixed-size list of bins and corresponding counts per bin regardless of the volume of data, satisfying property \textit{(a)}.
-Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and incrementing or decrementing the associated count, respectively, thus satisfying property \textit{(b)}.
-Recall Section \ref{sec:back-swag-algs} where we discussed generic sliding window aggregation algorithms and remember the Subtract-On-Evict (SOE) algorithm (Algorithm \ref{pseudo:soe}).
-Computing a histogram aggregation over a sliding window of size $W$ using SOE implies storing all $W$ elements in memory to know the element to remove after a new one is inserted.
-This violates our constant in-memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}.
-EMA-like histograms computed over sliding windows do not require the storage of window contents.
-An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values.
-These EMA-like histograms are further explained in Section \ref{sec:ema-hist}.
-
-Since the reference window is static, \textit{i.e.}, it is not a sliding window, we can use exact histogram aggregations.
-Hence, for each feature, we build an exact reference histogram aggregation.
-On the other hand, in streaming, we use one EMA-like approximate histogram aggregation per feature (Figure \ref{fig:timelines-hists}).
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/timelines-hists.png}
-      \caption[]{Reference and Streaming timelines and corresponding histogram aggregations}
-      \label{fig:timelines-hists}
-    \end{center}
-\end{figure}
-
-To compare each feature's reference and target sliding window histogram aggregation we need a measure of divergence.
-We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein \cite{EncyclopaediaMathematics}.
-The JSD takes as input two histograms and outputs a value between 0 and 1.
-The closer this value is to 0 the more similar the distributions are.
-On the other hand, a JSD value of 1 represents two different distributions.
-We chose to use the JSD metric due to its simplicity and because it was used in previous work successfully \cite{SAMM}, leaving as future work other possible tests.
-
-Given a divergence or distance value for the two aggregations (reference and target) we need a measure of normal statistical fluctuations.
-Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis.
-This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand-picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).
-
-To implement the proposed methodology, we devised a two-phase method.
-The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}).
-The second phase (detailed in Section \ref{sec:stream-phase}) focuses on efficiently updating the target sliding window histogram for each feature and periodically performing a divergence test between the static reference aggregation and the target sliding one, alerting for changes if need be.
-The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.
-
-\subsection{Exponential Moving Average Histogram} \label{sec:ema-hist}
-As stated, we compute histogram aggregations for the reference and target sliding windows.
-The computation of the reference histogram is done once in batch.
-Since the reference data is fixed and does not change over time, computing this histogram in batch is more efficient and easier to implement in a distributed system like Spark.
-In addition, a batch computation also allows us to compute an exact reference histogram.
-However, the target sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.
-
-In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they compute sliding window aggregations without actually storing the window contents.
-Definition \ref{def:tuple-ema} shows how to compute an EMA using its recursive formula.
-An EMA state or value depends only on the previous one multiplied by an exponential decay.
-Recall that a tuple-based decay is given by:
-\[ 2^{- \frac{1}{n_{1/2}}} \]
-where $n_{1/2}$ is commonly denoted as a tuple-based half-life.
-
-Algorithm \ref{pseudo:ema-hist} shows how we update our EMA histogram.
-We assume we have access to global \textit{counts} and \textit{bins} arrays and use a suppression factor of $2^{- \frac{1}{n_{1/2}}}$ with $n_{1/2}$ as the tuple-based half-life.
-The histogram aggregation is updated for each new streaming value \texttt{val} using the \texttt{processValue(val)} function, that increments the correct histogram bin (note that the \textit{bins} array defines the bin boundaries while the \textit{counts} array actually holds each bin's count) and applies the EMA suppression to every bin.
-\begin{algorithm}[!htb]
-    \caption[EMA histogram]{EMA histogram sliding window emulation}
-    \label{pseudo:ema-hist}
-    \begin{algorithmic}[1]
-        \Function{getBin}{val}
-            \State \Return{binarySearch(bins, val)}
-        \EndFunction
-
-        \Function{processValue}{val}
-            \State $bin \gets getBin(val)$
-            \State $counts[bin] \gets counts[bin] + 1$ \Comment{// Increment correct bin}
-
-            \ForEach {$count \in counts$} \Comment{// Apply EMA suppression to all bins}
-                \State $count \gets count \times 2^{- \frac{1}{n_{1/2}}}$
-            \EndFor
-        \EndFunction
-    \end{algorithmic}
-\end{algorithm}
-
-After processing \textit{x} tuples the oldest tuple is \emph{smoothed} by a factor of $2^{- \frac{x}{n_{1/2}}}$.
-If we process $x = 4n_{1/2}$ tuples the oldest tuple would have its contribution reduced by
-\[ 2^{- \dfrac{4n_{1/2}}{n_{1/2}}} = 2^{-4} \]
-which is approximately 0.06. In other words, the \textit{x-th} event would have 6\% contribution towards the EMA state or value meaning older events would have even less weight.
-
-We choose to emulate windows with four times the half-life size (or $4n_{1/2}$), meaning we discard events whose contribution is less than 6\%.
-We set the half-life to be one-quarter of the size of the window we want to emulate.
-The size of the window we want to monitor varies from situation to situation, but in the limited scope of this thesis we set to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data.
-This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.
+To that end, we devise a method that works with two windows and comprises two phases, a batch analysis phase and a stream analysis one.
+The batch analysis is performed on a static in time window which we call the reference window.
+The stream analysis is performed over a sliding window that groups data stream events.
+We aggregate the contents of the reference and sliding windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB}.
+We then apply a statistical distribution divergence function to both aggregations to obtain a divergence value.
+In the batch phase, we build the distribution of these values so we can define a percentile threshold for newly measured divergence values.
 
 \subsection{Batch Analysis Phase} \label{sec:batch-phase}
-The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and EMA-like aggregations.
-In other words, we:
-
+The goal of the batch phase is to build our reference aggregation for each feature and get to know the distribution of divergence or distance values. In other words, for each feature, we:
 \begin{enumerate}
-    \item Obtain a reference histogram for each feature using all data in the reference period
-
-    \item Compute the distribution of distances between the reference histogram and random windows of data and corresponding histograms
+    \item Compute the reference histogram using all data in the reference period
+    \item Compute the distribution of distances between the reference histogram and random samples of data
 \end{enumerate}
@@
     \begin{center}
-      \includegraphics[scale=0.3]{figures/ref-hist.png}
+      \includegraphics[scale=0.4]{figures/ref-hist.png}
       \caption[]{Reference equal-height histogram}
@@
     \begin{center}
-      \includegraphics[scale=0.3]{figures/ref-bins.png}
+      \includegraphics[scale=0.4]{figures/ref-bins.png}
       \caption[]{Reference equal-height histogram}
@@
     \begin{center}
-    %\hspace{2cm}
+    \hspace{2cm}
       \includegraphics[scale=0.8]{figures/ema-hist.png}
@@
     \begin{center}
-      \includegraphics[scale=0.4]{figures/compute-sample-distances.png}
+      \includegraphics[scale=0.7]{figures/compute-sample-distances.png}
       \caption[Compute sample's distance values]{Compute as many distance values as samples, one for each sample-reference histogram pair}
@@
     \begin{center}
-      \includegraphics[scale=0.2]{figures/skewed-left-distro.png}
+      \includegraphics[scale=0.4]{figures/skewed-left-distro.png}
       \caption[]{Left skewed distribution of distance values}
@@
     \begin{center}
-      \includegraphics[scale=0.2]{figures/skewed-right-distro.png}
+      \includegraphics[scale=0.4]{figures/skewed-right-distro.png}
       \caption[]{Right skewed distribution of distance values}
@@
 \end{itemize}
-
-\subsection{Streaming Phase} \label{sec:stream-phase}
-In the streaming or online phase, the system will process event by event in a true sliding window fashion (recall Section \ref{sec:windows}) and perform a change detection test periodically.
-This phase makes use of all of the batch phase artifacts, namely, for each feature, the reference histogram, the distribution of distance values and the last sample's histogram.
-
-For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase.
-Then, for each incoming event and each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
-This is all we do, regarding event processing.
-
-In order to obtain alerts, we periodically perform a multi-feature change detection test.
-The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the target sliding window histogram of the same feature.
-This distance value is then tested against the known distribution of distance values for feature \textit{x}.
-In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile, \textit{e.g.}, the 99th-percentile.
-When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and target sliding window histogram aggregations.
-The probability value or \textit{p-value} of a distance value is:
-
-\[\textit{p-value\textsubscript{distance\textsubscript{x}}} = 1 - percentile(\textit{distance\textsubscript{x}})\]
-
-For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01.
-Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
-\[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]
-or:
-\begin{equation}
-    \label{eq:alert-test}
-    \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha
-\end{equation}
-
-\subsubsection*{Multiple Test Correction} \label{sec:multi-test}
-Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis.
-When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives.
-In other words, when repeating a test multiple times the multiple test or simultaneous statistical inference problem \cite{MultiTestProblem-Pubmed, MultiTestProblem-Dickhaus2014, MultiTestProblem-Miller1966, multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
-In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives.
-Furthermore, it does not require to assume any independence between the tested hypothesis (as other methods do) which would almost surely be a wrong assumption for realistic datasets.
-Therefore we leave as future work the testing of other correction methods.
-In statistics, Family-Wise Error Rate (FWER) \cite{MultivariateMT, MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not.
-The Holm-Bonferroni method \cite{HolmBonferroni} is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
-
-\subsubsection*{Holm-Bonferroni Correction} \label{sec:holmbonferroni}
-The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.
-This correction is done by dividing the threshold $\alpha$ by a correction factor, as described next.
-
-First, we compute the \textit{p-value} of each feature.
-We then order our features by \textit{p-value}, in ascending order.
-Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if:
-\begin{equation}
-    p-value_{distance_{x}} < \alpha / (m - k)
-    \label{eq:corrected-pvalue}
-\end{equation}
-with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature.
-We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.
-
-\subsection{Final Remarks}
-
-\subsubsection*{Minimum Number of Samples} \label{sec:nsamples}
-In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase.
-But what is the optimal number of samples to make?
-How can we mathematically derive a formula for the number of samples to make and avoid \textit{ad hoc} user-defined constants?
-
-When checking for alerts we are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests.
-We apply a multiple test correction on the tests performed and want to set the FWER to $\alpha$.
-In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.
-
-We use the Holm-Bonferroni correction and, for each feature, we need to reliably estimate the upper tail of the distribution which corresponds to the \textit{$\alpha$ / n\textsubscript{features}} region.
-Each sample we do translates into a distance value and the collection of these data points builds the distribution of distance values.
-We want to be sure that the number of samples \textit{S} that we do generates enough distance points to have a high probability of having one or more points in the upper tail region.
-To that end, we define the probability of not having at least one point in the upper tail of the distance distribution as $\gamma$.
-
-In \cite{SAMM}, the authors prove that the minimum number of samples $S$ needed to satisfy these probability requirements is:
-\begin{equation}
-    \label{eq:optimal-n-samples}
-    \textit{S} = \ceil[\bigg]{\frac{\log\gamma}{log(1 - \frac{\alpha}{n_{features}})}}
-\end{equation}
-
-\subsubsection*{Method classification}
-Similarly to the methods presented in Sections \ref{kifer} and \ref{sec:SAMM}, we use a two-windowed model that can take uni or multivariate input.
-However, we perform a multivariate analysis of all feature's time-series with the use of the Holm-Bonferroni correction.
-Hence, we say our method is multivariate.
-We also alert entire windows or subsequences as anomalous.
-Thus, according to the taxonomy presented in Section \ref{sec:outliers}, our method is a dissimilarity-based multivariate subsequence outlier detection method.
-
-\subsubsection*{Time complexity of the streaming phase}
-The critical phase of our method is the stream analysis because it has to process new events and alert for possible deviations in real-time.
-
-For each event and for each feature, we update our EMA histogram using Algorithm \ref{pseudo:ema-hist} which has to perform a binary search to find the correct bin to insert our new value and traverses all bins to apply a suppression factor.
-Using $n$ as the number of bins, we have a time complexity of $O(log(n) + n)$, for each feature.
-However, since $n$ is a fixed and small constant, \textit{i.e.}, more or less 100 bins, we say that this algorithm has constant time complexity.
-In other words, our time complexity does not depend on the size of the emulated streaming windows.
-
-\subsubsection*{Space complexity of the streaming phase}
-We store $n$ bins per feature, so we have a space complexity of $O(n)$ per feature.
-Again, since $n$ is a small and fixed constant, we claim to have constant space complexity for ever-growing volumes of data.
 

### 2020-07-15 chapters/conclusions.tex
Experiment results for last test

@@
 In the real data experiments, once again, we managed to achieve high throughput but the produced alerts were not very accurate.
-We defer further investigation to future work because of the time constraints of this thesis but leave open the possibility that the approximate aggregation value is more inaccurate than what we thought it would be since the EMA tail holds all past events.
+We defer further investigation to future work because of the time constraints of this thesis but leave open the possibilities that our divergence measure distribution was poorly built and hence more samples need to be made in the batch phase or that the approximate EMA histogram aggregation value is more inaccurate than what we thought it would be since the EMA tail holds all past events and higher EMA smoothing factors should be used.
 
@@
 
-    \item experiments on real data that showed unexpected results but where we provide insights and possible hypothesis to test for future work
+    \item experiments on real data that showed unexpected results but where we provide insights and possible hypotheses to test for future work
 \end{enumerate}
@@
 First and foremost, we believe the next steps are to reason about the obtained results in the real data experiments and find the root cause for the existence of many and large alert periods.
-As a possible hypothesis to test, we suggest testing with EMAs with higher smoothing factors, so that past events are forgotten faster; additionally, we suggest decreasing $\gamma$, which further increases the number of samples to make in the batch phase.
+As a possible hypothesis to test, we suggest testing with EMAs with higher smoothing factors, so that past events are forgotten faster; additionally, we suggest decreasing $\gamma$, which further increases the number of samples to make in the batch phase, improving the accuracy of the divergence measure distribution built.
 
@@
 Lastly, we propose another type of validation method.
-We mentioned in the beginning that our objective is to detect data pattern shifts before the performance of the systems relying on the data declines.
+We mentioned in the beginning that our objective is to detect data pattern shifts before the performance of the systems that rely on the data declines.
 We propose a test where we train a Machine Learning model on a reference period and measure its prediction accuracy on a target period.
@@
 We propose that the model is retrained at these timestamps (tantamount to system reconfiguration) and then used to predict for the remaining target period.
-Finally, we propose that the accuracy should be compared to the previously observed model performance, expecting it to be better after retraining in one of the high divergent timestamps.
+Finally, we propose that the accuracy should be compared to the previously observed model performance, expecting it to be higher after retraining in one of the high divergent timestamps.

### 2020-07-15 chapters/validation.tex
Final changes and submission
Experiment results for last test

@@
     \end{center}
-    \caption{Multi feature reference dataset and feature distributions}
+    \caption{Multi-feature reference dataset and feature distributions}
     \label{tbl:multi-feat-ref-dataset-distros}
@@
 
+\subsection{Building an Accurate Divergence Measure Distribution}
+Our method comprises two-phases: the batch and the streaming analysis.
+We can not conclude in which phase an error that could possibly explain the last experiment's results exists.
+To isolate both phases, we conduct this last experiment where instead of using the distribution of divergence measures built in the batch phase we use a distribution we know to be the best fit for the streaming period.
+
+Our best fit divergence distribution was built over the streaming period.
+To do this, we processed the streaming events once and saved all the JSD measures.
+We binned these JSD values with 50,000 bins all with equal heights (or counts).
+Then we performed once again the streaming analysis over the same period but this time using the previously built distribution for which we know it is the best possible fit, since it was computed over the exact same period, built event by event.
+We set the FWER to be 1\% so we expect to see some alarms, \textit{i.e.}, for JSD measures above the 99th percentile.
+
+Recall that in the last experiment, for feature $x_1$, using the same target and reference periods, we had two large alarms that lasted months (Figure \ref{fig:merchant2-x1-correctedpvalues}).
+Using the same periods but with the newly computed distribution of divergence measures we obtain the corrected p-values plot shown in Figure \ref{fig:merchant3-x1-correctedpvalues}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x1-correctedpvalues.pdf}
+      \caption{Merchant feature $x_1$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x1-correctedpvalues}
+    \end{center}
+\end{figure}
+This plot actually makes sense and represents expected behavior.
+Notice that we have only a few short-lasting alarms, which correspond to the JSD measures above the 99th percentile.
+These can be thought of as false positives.
+In reality, the definition of the multi-test we perform is exactly this: alert events above the 99th percentile (FWER=1\%).
+
+Figures \ref{fig:merchant3-x2-correctedpvalues}, \ref{fig:merchant3-x3-correctedpvalues}, \ref{fig:merchant3-x4-correctedpvalues} and \ref{fig:merchant3-x5-correctedpvalues} show the corrected p-value plots for this experiment for features $x_2$, $x_3$, $x_4$ and $x_5$, respectively.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x2-correctedpvalues.pdf}
+      \caption{Merchant feature $x_2$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x2-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x3-correctedpvalues.pdf}
+      \caption{Merchant feature $x_3$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x3-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x4-correctedpvalues.pdf}
+      \caption{Merchant feature $x_4$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x4-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x5-correctedpvalues.pdf}
+      \caption{Merchant feature $x_5$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x5-correctedpvalues}
+    \end{center}
+\end{figure}
+The analysis performed for feature $x_1$ applies for all of these features.
+Notice that for all plots we have very few alarms of very short durations.
+This coherency in results across multiple features seems to indicate that our streaming analysis is accurate and detects the outliers accurately if the divergence measure distribution used is well built.
+
+\newpage
 \section{Conclusions}
@@
 
-We began by performing a single-feature analysis of synthetic data.
+We began by performing single-feature analysis of synthetic data.
 The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported.
@@
 
-Finally, we moved on to real data.
+We then moved on to real data.
 Due to time constraints and the scope of this thesis, we experimented only with data from one merchant.
@@
 A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift.
-In our last experiment, we attempted to reduce concept drift to a maximum by using the same period as reference and target.
+
+In our second experiment with real data, we attempted to reduce concept drift to a maximum by using the same period as reference and target.
 To our surprise, we still produced a large number of long-lasting alarms.
@@
 In other words, for the streaming EMAs, the previous events never have their contribution set to zero, even though the weights tend to zero, whereas the batch EMAs worked with a finite dataset.
-Additionally, our sampling may have been insufficient.
-To test this, an experiment with a lower $\gamma$ and hence a higher number of samples should be carried out.
+Additionally, our sampling may have been insufficient and hence the divergence measure distribution was inaccurate.
+To test the first hypothesis, experiments with higher smoothing factors for streaming EMAs should be carried out.
+To test the second one, a lower $\gamma$ parameter should be used in the batch phase, hence increasing the number of samples to be made, potentially building a more accurate distribution of JSD values.
+
+In our final experiment, we aim to build the most accurate as possible distribution of divergence measures by building it based on the measured divergences observed in streaming.
+In our second streaming run, we used this distribution and observed that the alarms produced were very few and short-lasting ones.
+The few alarms raised correspond to divergence measures above the 99th percentile.
+With this experiment we conclude our streaming analysis is correctly built but depends a lot on how accurate the divergence measure distribution is.
+As discussed, future tests should be carried out where more samples are made, to ensure a more accurate divergence measure distribution is computed.

### 2020-07-15 chapters/conclusions.tex
Added Sliding Bloom and Sliding CMS to references, mention only

@@
 
+Besides testing new methods for distribution divergence measurement or multiple test correction methods, we have a set of parameters to work with.
+Experiments where only one of the parameters is changed should be carried out.
+Our method has lots of parameters to fine tune and experiment with, as defined in Section \ref{sec:batch-phase}, such as the tuple-based half-life, the size of the samples we make, the $\gamma$ probability (which defines the probability of not having at least one sample divergence measure in the upper tail of the distribution) and the Family-Wise Error Rate.
+
 Lastly, we propose another type of validation method.

### 2020-07-15 chapters/sota.tex
EMA hist pseudocode and Final remarks analysis
Added Sliding Bloom and Sliding CMS to references, mention only

@@
 
-\subsubsection{SAMM: Automatic Model Monitoring for Data Streams}
+\subsubsection{SAMM: Automatic Model Monitoring for Data Streams} \label{sec:SAMM}
 
@@
 
-%\subsection{Sliding Bloom Filter}
-%\subsection*{Data Structure}
-%\subsection*{Element Insertion}
-%\subsection*{Element Eviction}
-%\subsection*{Time complexity analysis}
-%\subsection*{Space complexity analysis}
-%\subsection*{Applicability to our Hypothesis}
-
-%\subsection{Sliding Count-Min Sketch}
-%\subsection*{Data Structure}
-%\subsection*{Element Insertion}
-%\subsection*{Element Eviction}
-%\subsection*{Time complexity analysis}
-%\subsection*{Space complexity analysis}
-%\subsection*{Applicability to our Hypothesis}
-
 \subsection{Sliding HyperLogLog}
@@
 
-We then discussed probabilistic data structures (Section \ref{sec:pds}) and their sliding window implementations (Section \ref{sec:sliding-pds}).
+In Section \ref{sec:pds}, we discussed probabilistic data structures.
+Then, in Section \ref{sec:sliding-pds}, we explored some probabilistic data structure sliding window implementations.
+We explored only sliding window versions of the HyperLogLog, but sliding versions of other probabilistic data structures exist, such as the Sliding Bloom Filter \cite{SlidingBloom} and the Sliding Count-Min Sketch \cite{SlidingCMS}.
 Despite being efficient, both time and memory-wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use to us.

### 2020-07-16 chapters/conclusions.tex
Added final thesis pdf

--- 2020-07-15 chapters/conclusions.tex
@@
 
-\subsubsection*{(RQ2) How can we maintain such an aggregation in a sliding window fashion in real-time with constant memory usage?}
+\subsubsection*{(RQ2) How can we maintain such aggregation in a sliding window fashion in real-time with constant memory usage?}
 
@@
 In Section \ref{sec:sampling-batch}, we explain how we define the threshold and hence define when we raise an alert.
-In our batch phase, we make a number of samples and measure the JSD distance between the histogram aggregations of those samples and our reference histogram aggregation.
+In our batch phase, we make several samples and measure the JSD distance between the histogram aggregations of those samples and our reference histogram aggregation.
 We end up with a list of JSD values.
@@
 
+Besides testing new methods for distribution divergence measurement or multiple test correction methods, we have a set of parameters to work with.
+Experiments where only one of the parameters is changed should be carried out.
+Our method has lots of parameters to fine-tune and experiment with, as defined in Section \ref{sec:batch-phase}, such as the tuple-based half-life, the size of the samples we make, the $\gamma$ probability (which defines the probability of not having at least one sample divergence measure in the upper tail of the distribution) and the Family-Wise Error Rate.
+
 Lastly, we propose another type of validation method.

### 2020-07-16 chapters/problem-statement.tex
Added final thesis pdf

@@
 These systems are configured \textit{a priori} and make assumptions about future streaming data based on the observable past events.
-The problem is that if these assumptions no longer hold in the future, the stream monitoring systems' performance decays.
+The problem is that if these assumptions no longer hold in the future, the systems' performance decays.
 Thus, the problem at hand is to determine when to reconfigure the system based on an analysis of the drifts in the stream of data.
@@
 Our objective is to alert data pattern shifts.
-Our motivation is doing it so that system administrators can reconfigure the system before the system's performance decays.
+Our motivation is doing it so that system administrators can reconfigure the system before its performance decays.
 In this example, system reconfiguration is tantamount to model retraining.
@@
 We devise a two-phased and two-windowed method: the first phase is a batch analysis that makes use of a reference window; and the second phase is a streaming analysis that makes use of a target sliding window.
-The batch analysis main goal is to compute the reference aggregation for the reference window.
+The batch analysis main goals are to compute the reference aggregation for the reference window and the divergence measures distribution.
 The stream analysis main goals are to incrementally maintain the same type of aggregation but over the streaming data and alert when the difference between the reference aggregation and the target sliding one is high.

### 2020-07-16 chapters/sota.tex
Added final thesis pdf

@@
 We explored only sliding window versions of the HyperLogLog, but sliding versions of other probabilistic data structures exist, such as the Sliding Bloom Filter \cite{SlidingBloom} and the Sliding Count-Min Sketch \cite{SlidingCMS}.
-Despite being efficient, both time and memory-wise, each probabilistic data structure allows for one type of aggregation, and none of them happened to be of particular use to us.
-
-In the next chapter, we formally define the problem we aim to solve and the issues with current state of the art solutions.
+Despite being efficient, both time and memory-wise, each probabilistic data structure only allows for one type of aggregation, and none of them happened to be of particular use to us.
+
+In the next chapter, we formally define the problem we aim to solve and the issues with the current state of the art solutions.
 Then, we formulate our proposal along with the assumptions we make when approaching this problem.

### 2020-07-16 chapters/validation.tex
Added final thesis pdf

@@
 \subsection{Building an Accurate Divergence Measure Distribution}
-Our method comprises two-phases: the batch and the streaming analysis.
-We can not conclude in which phase an error that could possibly explain the last experiment's results exists.
+Our method comprises two phases: the batch and the streaming analysis.
+We can not conclude in which phase an error that could explain the last experiment's results exists.
 To isolate both phases, we conduct this last experiment where instead of using the distribution of divergence measures built in the batch phase we use a distribution we know to be the best fit for the streaming period.
@@
 We binned these JSD values with 50,000 bins all with equal heights (or counts).
-Then we performed once again the streaming analysis over the same period but this time using the previously built distribution for which we know it is the best possible fit, since it was computed over the exact same period, built event by event.
+Then we performed once again the streaming analysis over the same period but this time using the previously built distribution for which we know it is the best possible fit since it was computed over the same period, built event by event.
 We set the FWER to be 1\% so we expect to see some alarms, \textit{i.e.}, for JSD measures above the 99th percentile.
@@
 The few alarms raised correspond to divergence measures above the 99th percentile.
-With this experiment we conclude our streaming analysis is correctly built but depends a lot on how accurate the divergence measure distribution is.
+With this experiment, we conclude our streaming analysis is correctly built but depends a lot on how accurate the divergence measure distribution is.
 As discussed, future tests should be carried out where more samples are made, to ensure a more accurate divergence measure distribution is computed.

### 2020-07-16 paper.tex
Added final thesis pdf

@@
 
-To summarize, in this paper we design a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.
+In this paper, we present a lightweight real-time system that monitors a stream of high volumes of high velocity, highly skewed and seasonal data and detects pattern shifts relative to a reference period.
 The motivation to detect data stream pattern deviations in real-time is to know when to reconfigure the mission-critical system (\textit{e.g.}, retrain the Machine Learning model) before its performance decays.
@@
 The stream analysis is performed over a sliding window that groups data stream events.
-We aggregate the contents of the reference and sliding windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB}.
-We then apply a statistical distribution divergence function to both aggregations to obtain a divergence value.
-In the batch phase, we build the distribution of these values so we can define a percentile threshold for newly measured divergence values.
+We aggregate the contents of the reference window with an exact histogram and the sliding window with an approximate histogram aggregation based on Exponential Moving Averages (EMAs) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB}.
+We then apply a statistical distribution divergence function to both histogram aggregations to obtain a divergence value.
+In the batch phase, we build the distribution of these values so we can define a percentile threshold for the newly measured divergence values in streaming.
 
@@
 \end{enumerate}
-
 This distribution of values can then later be compared to a given value of divergence for a single sample to find out how probable that value is to occur (if it is very low an alert may be raised).
 
-\subsubsection*{Building the Reference Histogram for each Feature}
-
+\subsubsection{Building the Reference Histogram}
 First, for each feature, we compute the reference histogram from the reference period dataset.
 This is an exact histogram aggregation.
-When building it we ensure the histogram has equal counts (or equal height) for all bins which often results in different sized bins.
-We make no assumption on the dataset distributions and hence use equal height histograms because they adjust better to wildly varying ones.
-Equal height means we have more bins covering very dense regions (regions with many data points) and fewer bins in lower density regions.
+When building it we ensure the histogram has equal counts for all bins which often results in different sized bins.
+Equal height histograms adjust better to wildly varying distributions.
+Equal height means we have more bins covering very dense regions and fewer bins in lower density regions.
 
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/ref-hist.png}
-      \caption[]{Reference equal-height histogram}
-      \label{fig:ref-hist}
-    \end{center}
-\end{figure}
-
-\subsubsection*{Finding the Distribution of Distances for each Feature} \label{sec:sampling-batch}
-
-Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase.
-To that end, we make $S$ samples of transactions (in Section \ref{sec:nsamples} we discuss the minimum number $S$ of samples to make), each with the same tuple-based size of the target sliding window we will use in streaming.
+\subsubsection{Finding the Distribution of Divergence Values} \label{sec:sampling-batch}
+Secondly, we aim to build the histogram that encodes the distribution of expected distance or divergence values, to later on threshold the observed distance values during the streaming phase.
+To that end, we make $S$ samples of transactions, each with the same tuple-based size of the sliding window we emulate in streaming.
 Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
 
-For each sample, and each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}).
+\textcolor{red}{DEFINE S}
+
+For each sample, and each feature, we compute the approximated EMA histogram using the bins from the reference histogram.
 We also add two extra bins that will cover the region to the left and the right of our histogram, respectively.
@@
 
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/ref-bins.png}
-      \caption[]{Reference equal-height histogram}
-      \label{fig:ref-hist-bins}
-    \end{center}
-\end{figure}
-
-Each of the histograms is an EMA-like histogram.
+Each of the sample's histograms is an EMA-like histogram.
 An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}.
@@
     \begin{center}
-    \hspace{2cm}
-      \includegraphics[scale=0.8]{figures/ema-hist.png}
+    %\hspace{6cm}
+      \includegraphics[scale=0.6]{figures/ema-hist.png}
       \caption[]{Reference equal-height histogram}
@@
 \end{figure}
+\textcolor{red}{DEFINE EMA LIKE HIST}
+
 We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period.
@@
 For each feature, we have now \textit{S} histograms, one for each sample.
-For each of the \textit{S} histograms, we compute the distance between it and the reference histogram.
-Figure \ref{fig:compute-sample-distances} illustrates this process as we apply our distance function for each sample histogram and the reference histogram, obtaining \textit{S} distance measurements.
-
+We compute the divergence or distance between each of the \textit{S} histograms and the reference histogram.
+Figure \ref{fig:compute-sample-distances} illustrates this process as we apply our distance function for each sample histogram and the reference histogram, obtaining \textit{S} divergence measurements.
 \begin{figure}[!htb]
     \begin{center}
-      \includegraphics[scale=0.7]{figures/compute-sample-distances.png}
+      \includegraphics[scale=0.4]{figures/compute-sample-distances.png}
       \caption[Compute sample's distance values]{Compute as many distance values as samples, one for each sample-reference histogram pair}
@@
 
-For each feature, we end up with \textit{S} distance values.
+For each feature, we end up with \textit{S} distance or divergence values.
 These $S$ measurements are distance measurements between random samples of data and the reference period.
@@
 
-It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself.
-For instance, given that JSD values vary between 0 and 1, why not just set to alert if the measured JSD value is 0.7?
-The reason is that a divergence value may fall on different percentiles for different features.
-In other words, the same \textit{i-th} percentile will correspond to a different divergence value for different features.
-Consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/skewed-left-distro.png}
-      \caption[]{Left skewed distribution of distance values}
-      \label{fig:skewed-left-distro}
-    \end{center}
-\end{figure}
-
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.4]{figures/skewed-right-distro.png}
-      \caption[]{Right skewed distribution of distance values}
-      \label{fig:skewed-right-distro}
-    \end{center}
-\end{figure}
-
-Let's assume that the histogram from Figure \ref{fig:skewed-left-distro} encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}.
-Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results.
-For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}.
-Instead, we define our threshold as a percentile of the distribution and not as a hard-coded distance constant.
-For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.
-
-Hard-coding a threshold value to be used for all features is cumbersome, impractical and error-prone. The distance threshold must be computed based on the distribution of distances.
-
-\subsubsection*{Burn-In Period at System Initialization}
+\subsubsection{Burn-In Period at System Initialization}
 When we boot up our system in the streaming phase we initially have empty histogram aggregations.
 The period of time where you discard the alerts produced by the system until it processes enough data to produce accurate reports is commonly denoted as the burn-in period.
-To avoid this burn-in period we propose that the target sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
+To avoid this burn-in period we propose that the sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
 
-\subsubsection*{Batch Phase Artifacts} \label{sec:batch-artifacts-summary}
-By the end of the batch phase, we should have, for each feature: 
-\begin{itemize}
+\subsubsection{Batch Phase Artifacts} \label{sec:batch-artifacts-summary}
+By the end of the batch phase, for each feature, we have: 
+\begin{enumerate}
     \item a reference histogram
-    \item a list of distance values that represent its distribution
-    \item the last sample's histogram, to be used as burn-in period for the target sliding window histogram
-\end{itemize}
+    \item a distribution of divergence values
+    \item the last sample's histogram
+\end{enumerate}
+
+\subsection{Streaming Phase} \label{sec:stream-phase}
+In the streaming phase, the system will process event by event and perform a change detection test periodically.
+For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase.
+Then, for each incoming event and each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
+This is all we do, regarding event processing.
+
+\textcolor{red}{DEFINE EMA HIST}
+
+We periodically perform a multi-feature change detection test.
+The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the target sliding window histogram of the same feature.
+This distance value is then tested against the known distribution of distance values for feature \textit{x}.
+In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile.
+When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and target sliding window histogram aggregations.
+The probability value or \textit{p-value} of a distance value is:
+
+\[\textit{p-value\textsubscript{distance\textsubscript{x}}} = 1 - percentile(\textit{distance\textsubscript{x}})\]
+
+For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01.
+Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
+\[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]
+or:
+\begin{equation}
+    \label{eq:alert-test}
+    \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha
+\end{equation}
+
+\subsubsection{Multiple Test Correction} \label{sec:multi-test}
+Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis.
+When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives.
+In other words, when repeating a test multiple times the multiple test or simultaneous statistical inference problem \cite{MultiTestProblem-Pubmed, MultiTestProblem-Dickhaus2014, MultiTestProblem-Miller1966, multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
+In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives.
+Furthermore, it does not require to assume any independence between the tested hypothesis (as other methods do) which would almost surely be a wrong assumption for realistic datasets.
+In statistics, the Family-Wise Error Rate (FWER) \cite{MultivariateMT, MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not.
+The Holm-Bonferroni method \cite{HolmBonferroni} is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
+
+\subsubsection*{Holm-Bonferroni Correction} \label{sec:holmbonferroni}
+The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.
+This correction is done by dividing the threshold $\alpha$ by a correction factor, as described next.
+
+First, we compute the \textit{p-value} of each feature.
+We then order our features by \textit{p-value}, in ascending order.
+Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if:
+\begin{equation}
+    p-value_{distance_{x}} < \alpha / (m - k)
+    \label{eq:corrected-pvalue}
+\end{equation}
+with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature.
+We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.
 
@@
 \label{sec:Conclusions}
-\cite{ApacheFlink}
 Thesis done, PogU
-
-\section{CCS Concepts and User-Defined Keywords}
-
-Two elements of the ``acmart'' document class provide powerful taxonomic tools for you to help readers find your work in an online search.
-
-The ACM Computing Classification System --- \url{https://www.acm.org/publications/class-2012} --- is a set of classifiers and concepts that describe the computing discipline.
-Authors can select entries from this classification system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the commands to be included in the \LaTeX\ source.
-
-User-defined keywords are a comma-separated list of words and phrases of the authors' choosing, providing a more flexible way of describing the research being presented.
-
-CCS concepts and user-defined keywords are required for all short- and full-length articles, and optional for two-page abstracts.
-
-\section{Tables}
-
-The ``\verb|acmart|'' document class includes the ``\verb|booktabs|'' package --- \url{https://ctan.org/pkg/booktabs} --- for preparing high-quality tables.
-
-Table captions are placed {\it above} the table.
-
-Because tables cannot be split across pages, the best placement for them is typically the top of the page nearest their initial cite.
-To ensure this proper ``floating'' placement of tables, use the environment \textbf{table} to enclose the table's contents and the table caption.
-The contents of the table itself must go in the \textbf{tabular} environment, to be aligned properly in rows and columns, with the desired horizontal and vertical rules.
-Again, detailed instructions on \textbf{tabular} material are found in the \textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which Table~\ref{tab:freq} is included in the input file; compare the placement of the table here with the table in the printed output of this document.
-
-\begin{table}
-  \caption{Frequency of Special Characters}
-  \label{tab:freq}
-  \begin{tabular}{ccl}
-    \toprule
-    Non-English or Math&Frequency&Comments\\
-    \midrule
-    \O & 1 in 1,000& For Swedish names\\
-    $\pi$ & 1 in 5& Common in math\\
-    \$ & 4 in 5 & Used in business\\
-    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-  \bottomrule
-\end{tabular}
-\end{table}
-
-To set a wider table, which takes up the whole width of the page's live area, use the environment \textbf{table*} to enclose the table's contents and the table caption.
-As with a single-column table, this wide table will ``float'' to a location deemed more desirable.
-Immediately following this sentence is the point at which Table~\ref{tab:commands} is included in the input file; again, it is instructive to compare the placement of the table here with the table in the printed output of this document.
-
-\begin{table*}
-  \caption{Some Typical Commands}
-  \label{tab:commands}
-  \begin{tabular}{ccl}
-    \toprule
-    Command &A Number & Comments\\
-    \midrule
-    \texttt{{\char'134}author} & 100& Author \\
-    \texttt{{\char'134}table}& 300 & For tables\\
-    \texttt{{\char'134}table*}& 400& For wider tables\\
-    \bottomrule
-  \end{tabular}
-\end{table*}
-
-\section{Figures}
-
-The ``\verb|figure|'' environment should be used for figures.
-One or more images can be placed within a figure.
-If your figure contains third-party material, you must clearly identify it as such, as shown in the example below.
-\iffalse
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{sample-franklin}
-  \caption{1907 Franklin Model D roadster. Photograph by Harris \& Ewing, Inc. [Public domain], via Wikimedia Commons. (\url{https://goo.gl/VLCRBB}).}
-  \Description{The 1907 Franklin Model D roadster.}
-\end{figure}
-\fi 
-Your figures should contain a caption which describes the figure to the reader.
-Figure captions go below the figure.
-Your figures should {\bf also} include a description suitable for screen readers, to assist the visually-challenged to better understand your work.
-
-Figure captions are placed {\it below} the figure.
 

### 2020-07-16 paper.tex
Paper shortening

@@
 
-\author[Ribeiro]{Pedro Ribeiro}
-\email{pribeiro@dcc.fc.up.pt}
+\author[Restivo]{André Restivo}
+\email{arestivo@fe.up.pt}
 \affiliation{%
@@
 
-\author[Restivo]{André Restivo}
-\email{arestivo@fe.up.pt}
+\author[Ribeiro]{Pedro Ribeiro}
+\email{pribeiro@dcc.fc.up.pt}
 \affiliation{%
@@
 % of authors' names for this purpose.
-\renewcommand{\shortauthors}{Trovato and Tobin, et al.}
+%\renewcommand{\shortauthors}{Trovato and Tobin, et al.}
 
@@
 Over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates.
-Data pattern shift detection refers to the process of finding patterns in data that do not conform to expected or usual behavior.
 Accurate and timely detection of data pattern deviations allow for immediate measures to be taken.
 Thus, the problem at hand is to determine when to reconfigure the system, \textit{e.g.}, a Machine Learning model, based on an analysis of the drifts in the stream of data.
-In this paper, we present a method that makes use of constant time & memory streaming aggregations and distribution divergence functions to alert deviations in data patterns in real-time.
-We generate alerts based on a user-defined probability threshold, for each of an event's fields, denominated features in our context.
+In this paper, we present a method that makes use of constant time and memory streaming aggregations and distribution divergence functions to alert deviations in data patterns in real-time.
+We generate alerts based on a user-defined probability threshold, for each of an event's numerical fields, denominated features in our context.
 We evaluate our method through a series of tests, some with synthetic data sets and some with real data.
@@
 Despite not knowing the specific reasons for that, we detail further investigation to future work and formulate a set of hypotheses that might explain the latter results and hence are worthy of pursuing next.
-Moreover, our set of experiments support the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
 \end{abstract}
@@
 For different use cases, the data is processed in different ways for different purposes.
-For instance, e-commerce platforms such as Alibaba\footnote{https://alibaba.com} need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes.
-Streaming services, like YouTube\footnote{https://youtube.com} and Netflix\footnote{https://netflix.com}, try to guarantee that media content reaches up to millions of users simultaneously.
+For instance, e-commerce platforms such as Alibaba need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes.
+Streaming services, like YouTube and Netflix, try to guarantee that media content reaches up to millions of users simultaneously.
 Social networks are responsible for generating large volumes of data.
-Twitter\footnote{https://twitter.com}, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day\footnote{https://blog.twitter.com/engineering/en\_us/a/2013/new-tweets-per-second-record-and-how.html}.
+Twitter, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day.
 \textit{Tweets} may contain text, media content or both.
@@
 
-Consider the use case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon\footnote{https://amazon.com} by every US citizen on a single day.
+Consider the use case where we want to detect credit card fraud on the data stream of all the transactions made on Amazon by every US citizen on a single day.
 In this case, Amazon's transaction stream will need to be monitored by another system, a fraud detection one.
@@
 The batch analysis is performed on a static in time window which we call the reference window.
-The stream analysis is performed over a sliding window that groups data stream events.
-We aggregate the contents of the reference window with an exact histogram and the sliding window with an approximate histogram aggregation based on Exponential Moving Averages (EMAs) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB}.
+The stream analysis is performed over an emulated sliding window that groups data stream events.
+We aggregate the contents of the reference window with an exact histogram and the sliding window with an approximate histogram aggregation based on Exponential Moving Averages (EMAs).
 We then apply a statistical distribution divergence function to both histogram aggregations to obtain a divergence value.
 In the batch phase, we build the distribution of these values so we can define a percentile threshold for the newly measured divergence values in streaming.
+We raise alerts for features above certain percentiles, \textit{e.g.}, the 99th percentile.
 
-\subsection{Batch Analysis Phase} \label{sec:batch-phase}
-The goal of the batch phase is to build our reference aggregation for each feature and get to know the distribution of divergence or distance values. In other words, for each feature, we:
-\begin{enumerate}
-    \item Compute the reference histogram using all data in the reference period
-    \item Compute the distribution of distances between the reference histogram and random samples of data
-\end{enumerate}
-This distribution of values can then later be compared to a given value of divergence for a single sample to find out how probable that value is to occur (if it is very low an alert may be raised).
+\subsection{EMA-based histogram} \label{sec:ema-hist}
+Exponential Moving Averages (EMAs) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB} are a type of reduce functions applied to a sequence of elements in an orderly fashion to compute a weighted average, using different weights for each element.
+These weights decrease exponentially with the age of the observations.
+EMAs are recursive: $EMA_t$ depends only on the $val_t$ plus a smoothed $EMA_{t-1}$.
+This recursive property allows us to store only one EMA value and continuously update it for each new value $val_t$, based on the following recursion formula:
+\begin{definition}
+EMA\textsubscript{t} = EMA\textsubscript{t-1} * $2^{-\frac{1}{n_{1/2}}}$ + $val_t$
+\label{def:tuple-ema}
+\end{definition} where $n_{1/2}$ is the tuple-based half-life of the EMA. 
+The half-life $n_{1/2}$ controls how fast old events are forgotten.
+The tuple-based half-life is the amount of tuples or events the EMA must process before a certain element has its weight reduced by half.
+For example, with a tuple-based half-life of $n_{1/2}=1$ tuple, for each new tuple, all weights are reduced by half.
 
-\subsubsection{Building the Reference Histogram}
-First, for each feature, we compute the reference histogram from the reference period dataset.
-This is an exact histogram aggregation.
-When building it we ensure the histogram has equal counts for all bins which often results in different sized bins.
-Equal height histograms adjust better to wildly varying distributions.
-Equal height means we have more bins covering very dense regions and fewer bins in lower density regions.
+Each count of our EMA-based histogram is essentially an EMA-count.
+Algorithm \ref{pseudo:ema-hist} shows how we update our EMA histogram.
+We assume we have access to global \textit{counts} and \textit{bins} arrays and use a suppression factor of $2^{- \frac{1}{n_{1/2}}}$.
+The histogram aggregation is updated for each new streaming value \texttt{val} using the \texttt{processValue(val)} function, that increments the correct histogram bin (note that the \textit{bins} array defines the bin boundaries while the \textit{counts} array actually holds each bin's count) and applies the EMA suppression to every bin.
+\begin{algorithm}[!htb]
+    \caption[EMA histogram]{EMA histogram sliding window emulation}
+    \label{pseudo:ema-hist}
+    \begin{algorithmic}[1]
+        \Function{getBin}{val}
+            \State \Return{binarySearch(bins, val)}
+        \EndFunction
 
-\subsubsection{Finding the Distribution of Divergence Values} \label{sec:sampling-batch}
-Secondly, we aim to build the histogram that encodes the distribution of expected distance or divergence values, to later on threshold the observed distance values during the streaming phase.
-To that end, we make $S$ samples of transactions, each with the same tuple-based size of the sliding window we emulate in streaming.
-Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
+        \Function{processValue}{val}
+            \State $bin \gets getBin(val)$
+            \State $counts[bin] \gets counts[bin] + 1$ \Comment{//Increment correct bin}
 
-\textcolor{red}{DEFINE S}
-
-For each sample, and each feature, we compute the approximated EMA histogram using the bins from the reference histogram.
-We also add two extra bins that will cover the region to the left and the right of our histogram, respectively.
-In other words, we add a bin that covers the region between $-\infty$ and our first bin and another that covers values from the last bin to $+\infty$.
-This way, observed values outside of the reference period will be placed in these special bins, either left or right.
-
-Each of the sample's histograms is an EMA-like histogram.
-An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}.
-Each sample is processed event by event and a discount factor is applied per event to each bin (EMA expiration factor), thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample.
+            \ForEach {$count \in counts$} \Comment{//EMA suppression}
+                \State $count \gets count \times 2^{- \frac{1}{n_{1/2}}}$
+            \EndFor
+        \EndFunction
+    \end{algorithmic}
+\end{algorithm}
 This procedure is illustrated in Figure \ref{fig:EMA-hist-build}.
@@
 \end{figure}
-\textcolor{red}{DEFINE EMA LIKE HIST}
 
+The EMA-based approximate histogram emulates a sliding window as an exponential tail.
+This aggregation shows constant memory and time complexity regarding window size, making it ideal for our lightweight real-time system.
+
+\subsection{Batch Analysis Phase} \label{sec:batch-phase}
+The main goals of the batch phase are to build our reference aggregation and the distribution of divergence values, for each feature.
+Streaming divergence values will be compared to this distribution of values later to find out how probable a value is to occur (if it is very low, an alert is raised).
+
+\subsubsection{Building the Reference Histogram}
+First, for each feature, we compute the reference histogram from the reference period.
+When building it we ensure the histogram has equal counts for all bins which results in different sized bins.
+Equal height histograms adjust better to wildly varying distributions.
+Equal height means we have more bins covering very dense regions and fewer bins in lower density regions.
+
+\subsubsection{Building the Distribution of Divergence Values} \label{sec:sampling-batch}
+Secondly, we aim to build the distribution of expected distance or divergence values, to later on threshold the observed distance values during the streaming phase.
+To that end, we make $S$ samples of transactions, each with the same tuple-based size of the sliding window we emulate in streaming.
+Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
+With this sampling we aim to build a distribution of divergence values and reliably estimate the upper tail of that distribution.
+In \cite{SAMM}, the authors show how many samples must be made to have at least one value in the upper tail of the distribution.
+Defining $\alpha$ as our threshold and $\frac{\alpha}{n_{features}}$ as our threshold corrected by the Holm Bonferroni multiple test correction, $\gamma$ is defined as the probability of \textbf{not} having at least one point in the $\frac{\alpha}{n_{features}}$ upper tail of the distribution and the minimum number of samples $S$ to make is given by:
+\begin{equation*}
+    \label{eq:optimal-n-samples}
+    \textit{S} = \ceil[\bigg]{\frac{\log\gamma}{log(1 - \frac{\alpha}{n_{features}})}}
+\end{equation*}
+For each sample, and each feature, we compute the approximated EMA histogram using the bins from the reference histogram.
+We also add two extra bins that will cover the region to the left and the right of our histogram, respectively.
+This way, observed values outside of the reference period will be placed in these special bins, either left or right.
 We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period.
@@
 For each feature, we have now \textit{S} histograms, one for each sample.
-We compute the divergence or distance between each of the \textit{S} histograms and the reference histogram.
-Figure \ref{fig:compute-sample-distances} illustrates this process as we apply our distance function for each sample histogram and the reference histogram, obtaining \textit{S} divergence measurements.
+We compute the divergence between each of the \textit{S} histograms and the reference histogram.
+Figure \ref{fig:compute-sample-distances} illustrates this process as we apply our divergence or distance function for each sample's EMA-based histogram and the reference histogram, obtaining \textit{S} divergence measurements.
 \begin{figure}[!htb]
@@
       \includegraphics[scale=0.4]{figures/compute-sample-distances.png}
-      \caption[Compute sample's distance values]{Compute as many distance values as samples, one for each sample-reference histogram pair}
+      \caption[Compute sample's distance values]{Compute divergence values between reference and samples}
       \label{fig:compute-sample-distances}
@@
 \end{figure}
-
-For each feature, we end up with \textit{S} distance or divergence values.
-These $S$ measurements are distance measurements between random samples of data and the reference period.
+These $S$ measurements are divergence measurements between random samples of data and the reference period.
 Hence, we claim we now know the distribution of expected distance values.
 Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if the probability is below a certain threshold.
-Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the target sliding window histogram.
+Note that the histograms for each sample are EMA-based histograms just like the emulated sliding window histogram.
 This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.
@@
 When we boot up our system in the streaming phase we initially have empty histogram aggregations.
-The period of time where you discard the alerts produced by the system until it processes enough data to produce accurate reports is commonly denoted as the burn-in period.
-To avoid this burn-in period we propose that the sliding window approximate histogram aggregation for each feature is initialized using the last sample's histogram for that feature.
+To avoid this burn-in period, we propose saving the last sample's histogram for each feature so that the sliding window EMA-based histogram aggregation is initialized with it.
 
@@
 
-\textcolor{red}{DEFINE EMA HIST}
-
 We periodically perform a multi-feature change detection test.
-The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the target sliding window histogram of the same feature.
-This distance value is then tested against the known distribution of distance values for feature \textit{x}.
-In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile.
-When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and target sliding window histogram aggregations.
-The probability value or \textit{p-value} of a distance value is:
-
-\[\textit{p-value\textsubscript{distance\textsubscript{x}}} = 1 - percentile(\textit{distance\textsubscript{x}})\]
-
+The test is done by first computing the divergence between the reference histogram for a feature \textit{x} and the sliding window EMA-based histogram of the same feature.
+This value is tested against the known distribution of distance values for feature \textit{x}.
+We test to see if the percentile of the computed distance relative to the distribution of divergence is above a certain percentile.
+The probability of a divergence value $val_x$ (or \textit{p-value}) is:
+\[pvalue(val_x) = 1 - percentile(val_x)\]
 For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01.
-Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
-\[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]
+We alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
+\[ percentile(val_x) > 1 - \alpha \]
 or:
@@
     \label{eq:alert-test}
-    \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha
+    pvalue(val_x) < \alpha
 \end{equation}
@@
 Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis.
-When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives.
-In other words, when repeating a test multiple times the multiple test or simultaneous statistical inference problem \cite{MultiTestProblem-Pubmed, MultiTestProblem-Dickhaus2014, MultiTestProblem-Miller1966, multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur.
+When considering several hypotheses, the problem of multiplicity arises \cite{MultiTestProblem-Pubmed, MultiTestProblem-Dickhaus2014, MultiTestProblem-Miller1966, multiple-test-correction-geoffrey}: the more hypotheses are checked, the higher the probability of obtaining false positives.
 In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives.
@@
 The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.
-This correction is done by dividing the threshold $\alpha$ by a correction factor, as described next.
-
+This correction is done by dividing the threshold $\alpha$ by a correction factor.
 First, we compute the \textit{p-value} of each feature.
@@
 \begin{equation}
-    p-value_{distance_{x}} < \alpha / (m - k)
+    pvalue(val_x) < \alpha / (m - k)
     \label{eq:corrected-pvalue}
 \end{equation}
-with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature.
+with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test (in our case the number of features) and \textit{k} as the 0-based index of the current feature.
 We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.

### 2020-07-16 paper.tex
New conference which allows for 9 pages

@@
 \setcopyright{acmlicensed}
-\acmConference[MileTS '20]{MileTS '20: 6th KDD Workshop on Mining and Learning from Time Series}{August 24th, 2020}{San Diego, California, USA}
-\acmBooktitle{MileTS '20: 6th KDD Workshop on Mining and Learning from Time Series, August 24th, 2020, San Diego, Claifornia, USA}
+\acmConference[KDD-ADS '20]{KDD-ADS '20: Knowledge Discovery and Data Mining - Applied Data Science 2020}{August 22th, 2020}{San Diego, California, USA}
+\acmBooktitle{KDD-ADS '20: Knowledge Discovery and Data Mining - Applied Data Science 2020, August 22th, 2020, San Diego, California, USA}
 \acmPrice{15.00}

### 2020-07-17 paper.tex
....

@@
 \usepackage{mathtools}
+\usepackage{subcaption}
 
@@
 The batch analysis is performed on a static in time window which we call the reference window.
-The stream analysis is performed over an emulated sliding window that groups data stream events.
+The stream analysis is performed over a exponential tail which emulates a sliding window that groups data stream events.
 We aggregate the contents of the reference window with an exact histogram and the sliding window with an approximate histogram aggregation based on Exponential Moving Averages (EMAs).
@@
 \end{figure}
-
-The EMA-based approximate histogram emulates a sliding window as an exponential tail.
+The EMA-based approximate histogram emulates a sliding window as an exponential tail, seen in Figure \ref{fig:EMA-tail}.
 This aggregation shows constant memory and time complexity regarding window size, making it ideal for our lightweight real-time system.
+\begin{figure}[!htb]
+    \begin{center}
+    %\hspace{6cm}
+      \includegraphics[scale=0.35]{figures/emavswindow.pdf}
+      \caption[]{EMA tail versus sliding window weights}
+      \label{fig:EMA-tail}
+    \end{center}
+\end{figure}
 
@@
 In \cite{SAMM}, the authors show how many samples must be made to have at least one value in the upper tail of the distribution.
-Defining $\alpha$ as our threshold and $\frac{\alpha}{n_{features}}$ as our threshold corrected by the Holm Bonferroni multiple test correction, $\gamma$ is defined as the probability of \textbf{not} having at least one point in the $\frac{\alpha}{n_{features}}$ upper tail of the distribution and the minimum number of samples $S$ to make is given by:
+Defining $\alpha$ as our threshold (and $\frac{\alpha}{n_{features}}$ as our threshold corrected by the Holm-Bonferroni multiple test correction \cite{HolmBonferroni}, detailed in Section \ref{sec:multi-test}), $\gamma$ is defined as the probability of \textbf{not} having at least one point in the $\frac{\alpha}{n_{features}}$ upper tail of the distribution and the minimum number of samples $S$ to make is given by:
 \begin{equation*}
@@
 \end{equation*}
-For each sample, and each feature, we compute the approximated EMA histogram using the bins from the reference histogram.
+
+For each sample, and each feature, we compute the EMA histogram using the bins from the reference histogram.
 We also add two extra bins that will cover the region to the left and the right of our histogram, respectively.
@@
 We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period.
-We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.
+We use an EMA histogram to mimic the target EMA histogram we will incrementally maintain in the streaming environment.
 
@@
 Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if the probability is below a certain threshold.
-Note that the histograms for each sample are EMA-based histograms just like the emulated sliding window histogram.
+Note that the histograms for each sample are EMA-based histograms just like the streaming histogram.
 This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.
@@
 When we boot up our system in the streaming phase we initially have empty histogram aggregations.
-To avoid this burn-in period, we propose saving the last sample's histogram for each feature so that the sliding window EMA-based histogram aggregation is initialized with it.
+To avoid this burn-in period, we propose saving the last sample's histogram for each feature so that the streaming EMA-based histogram aggregation is initialized with it.
 
@@
     \item a reference histogram
-    \item a distribution of divergence values
+    \item the distribution of divergence values
     \item the last sample's histogram
@@
 We periodically perform a multi-feature change detection test.
-The test is done by first computing the divergence between the reference histogram for a feature \textit{x} and the sliding window EMA-based histogram of the same feature.
+The test is done by first computing the divergence (here denoted as $div_x$) between the reference histogram for a feature \textit{x} and the current streaming EMA histogram for $x$.
 This value is tested against the known distribution of distance values for feature \textit{x}.
 We test to see if the percentile of the computed distance relative to the distribution of divergence is above a certain percentile.
-The probability of a divergence value $val_x$ (or \textit{p-value}) is:
-\[pvalue(val_x) = 1 - percentile(val_x)\]
-For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01.
+The probability of a divergence value $div_x$ is:
+\[pvalue(div_x) = 1 - percentile(div_x)\]
+For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or below a probability of \textit{p-value} = 0.01.
 We alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
-\[ percentile(val_x) > 1 - \alpha \]
+\[ percentile(div_x) > 1 - \alpha \]
 or:
@@
     \label{eq:alert-test}
-    pvalue(val_x) < \alpha
+    pvalue(div_x) < \alpha
 \end{equation}
@@
 When considering several hypotheses, the problem of multiplicity arises \cite{MultiTestProblem-Pubmed, MultiTestProblem-Dickhaus2014, MultiTestProblem-Miller1966, multiple-test-correction-geoffrey}: the more hypotheses are checked, the higher the probability of obtaining false positives.
-In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives.
+In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni \cite{HolmBonferroni} multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives.
 Furthermore, it does not require to assume any independence between the tested hypothesis (as other methods do) which would almost surely be a wrong assumption for realistic datasets.
 In statistics, the Family-Wise Error Rate (FWER) \cite{MultivariateMT, MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not.
-The Holm-Bonferroni method \cite{HolmBonferroni} is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
-
-\subsubsection*{Holm-Bonferroni Correction} \label{sec:holmbonferroni}
+The Holm-Bonferroni method is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
 The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.
@@
 with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test (in our case the number of features) and \textit{k} as the 0-based index of the current feature.
-We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.
+We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and an alert per feature is logged with corresponding probability value.
 
-\section{Experiments}
-\label{sec:Experiments}
+\section{Experiments with Synthetic Data}
+\label{sec:SyntheticExperiments}
+In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data.
 
-Much scientist
+\subsection{Single-Feature Analysis}
+As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
+
+\subsubsection{Varying Sample Sizes}
+We began with single-feature scenarios and by experimenting with different tuple-based half-lives $n_{1/2}$.
+The half-life $n_{1/2}$ parameter in turn controls the size of the samples we make in the batch phase and the streaming EMA tail size.
+We set a half-life $n_{1/2}$ value and then multiply it by a constant factor of 4 to get the size of the samples to make.
+We process $4n_{1/2}$ events, meaning any further data we processed would make our oldest events have less than 6\% weight, following Algorithm \ref{pseudo:ema-hist}'s successively applied suppression factor.
+Processing $4n_{1/2}$ events means the oldest event's contribution is $({2^{- \frac{1}{n_{1/2}}}})^{4n_{1/2}} = {2^{- \frac{4n_{1/2}}{n_{1/2}}}} = 2^{-4} \approx 0.06 = 6\%$.
+
+All the synthetic datasets used throughout this section are a time-series where at each time-step we sample uniformly from a fixed distribution, \textit{i.e.}, there are no time correlations in the used datasets.
+The reference dataset R1 used in this first experiment contained a single feature that followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$ for the 5 million rows of the dataset.
+Figure \ref{fig:timeseries-r1} shows the time-series of the single feature in dataset R1.
+The dataset processed in the streaming phase (\textit{i.e.}, the target dataset) was generated with a size of 5 million events and a single feature.
+For the first half, \textit{i.e.}, for the first 2.5 million events, the feature followed a normal distribution with mean $\mu=10$ and standard deviation $\sigma=2$, like the reference dataset.
+For the other half, we changed the generating distribution to be a continuous uniform one with lower bound $a=100$ and upper bound $b=200$.
+Figure \ref{fig:timeseries-t1} shows the time-series of the single feature in the target dataset T1 where we see the abrupt change in feature values.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/01-reference.pdf}
+      \caption{Time-series for dataset R1}
+      \label{fig:timeseries-r1}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/01-target.pdf}
+      \caption{Time-series for dataset T1}
+      \label{fig:timeseries-t1}
+    \end{center}
+\end{figure}
+We conducted three different experiments varying the tuple-based half-life $n_{1/2}$, using the reference dataset R1 and the target dataset T1.
+The experiment identifiers, the corresponding half-lives and the execution times are shown in Table \ref{tbl:tests-half-life}.
+Notice that $Sample Size$ corresponds to the size of each sample made in the batch phase and so it is four times the half-life.
+We used 100 bins (plus the two special infinity bins) for the reference histogram and made 1000 samples.
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|c|c|}
+        \hline
+        \multicolumn{1}{|l|}{\textbf{Experiment}} & \multicolumn{1}{l|}{\textbf{Half-life}} & \multicolumn{1}{l|}{\textbf{Sample Size}} & \multicolumn{1}{l|}{\textbf{Batch execution time}} \\ \hline
+        01  & 625  & 2500 & 152                                                          \\ \hline
+        02                                        & 62,500                                  & 250,000                               & 2,822                                                        \\ \hline
+        03                                        & 250,000                                 & 1,000,000                           & 10,082                                                       \\ \hline
+        \end{tabular}
+    \end{center}
+    \caption{Experiments with varying half-lives}
+    \label{tbl:tests-half-life}
+\end{table}
+Table \ref{tbl:tests-half-life} shows the execution time for the batch phase in seconds.
+All experiments used 300 Spark executor instances.
+Since our sample size depends on the half-life (four times larger), the larger the half-life, the larger the sample size.
+Larger samples imply longer processing times when computing the EMA-like histogram for each sample.
+With no surprise, we see that the larger the sample the longer it takes to finish the batch phase.
+
+From a functional point of view, were we able to detect the abrupt change introduced in the target dataset T1?
+Recall that from the 2.5 millionth event forward we change the distribution from a gaussian to an uniform one (Figure \ref{fig:timeseries-t1}).
+Figure \ref{fig:JSD-signal-01} shows the distance (JSD) values computed during streaming at each event of the 5 million events from dataset T1, using the outputs from the batch phase of experiment 01.
+We clearly see a change in JSD values, and we move from values close to 0 to values close to 1 (or even 1).
+Figure \ref{fig:JSD-signal-zoom-01} shows the same signal but for the first half of the streaming period (essentially a zoom-in).
+We see our JSD values before the change were very low (magnitudes of $10^{-2}$).
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-625.pdf}
+      \caption{Whole time-series JSD signal and threshold}
+      \label{fig:JSD-signal-01}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-zoom-625.pdf}
+      \caption{Zoom-in on JSD signal and threshold}
+     \label{fig:JSD-signal-zoom-01}
+    \end{center}
+\end{figure}
+However, the change in JSD value is very abrupt as well.
+Almost instantly we go from close to 0 values to 1.
+We also had some false-positives, \textit{i.e.}, some JSD values above the threshold.
+This indicates to us that the window used is too small and sensitive to small changes, or that we can choose a more strict threshold.
+
+In experiment 02, we use larger sampling windows.
+Remember our target sliding window in streaming is of the same size as the sampling windows.
+In the second experiment, we increase the size of the sampling windows by 100 times.
+Figure \ref{fig:JSD-signal-02} shows the JSD signal during streaming for dataset T1, using the outputs from the batch phase of experiment 02.
+Again, we see a change in signal from the middle of the streaming period forward.
+Figure \ref{fig:JSD-signal-zoom-02} shows the same signal for the first half of the streaming period.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-62500.pdf}
+      \caption{Whole time-series JSD signal and threshold}
+      \label{fig:JSD-signal-02}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-zoom-62500.pdf}
+      \caption{Zoom-in on JSD signal and threshold}
+     \label{fig:JSD-signal-zoom-02}
+    \end{center}
+\end{figure}
+Notice that this time the growth of our signal is slower.
+Notice also that we do not have false positives in the first half of the signal.
+This is because we use a larger window, which means each event has a smaller contribution to the aggregation state.
+Hence, outliers will have less impact on the aggregation and corresponding signal value.
+
+In experiment 03, we increase the sampling and target windows size once again, this time four times larger than in experiment 02.
+We would expect an even smoother growth of our signal and, again, no false positives.
+And indeed, judging by the corresponding signal computation and zoom-in, Figures \ref{fig:JSD-signal-03} and \ref{fig:JSD-signal-zoom-03}, respectively, the signal looks smoother and we see no false positives.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-250000.pdf}
+      \caption{Whole time-series JSD signal and threshold}
+      \label{fig:JSD-signal-03}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-zoom-250000.pdf}
+      \caption{Zoom-in on JSD signal and threshold}
+     \label{fig:JSD-signal-zoom-03}
+    \end{center}
+\end{figure}
+
+\subsubsection{Multiple Distribution Changes}
+In this experiment, we wanted to see how our signal behaved when there were multiple change points, where each gaussian distribution was a scaled down version of the previous one.
+Figure \ref{fig:timeseries-r2} shows the time-series for the reference dataset R2 used in this experiment.
+It contained 5 million events with a single feature following a gaussian distribution with mean $\mu=100$ and standard deviation $\sigma=50$.
+The time-series for target dataset T2 is shown in Figure \ref{fig:timeseries-t2}.
+Notice that we change distributions at 2, 3 and 4 millionth events.
+Notice also that each subsequent distribution produces values whose domain is a subset of the previous distribution.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-r2.pdf}
+      \caption{Time-series for dataset R2}
+      \label{fig:timeseries-r2}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-t2.pdf}
+      \caption{Time-series for dataset T2}
+     \label{fig:timeseries-t2}
+    \end{center}
+\end{figure}
+
+The target dataset T2 begins with the same distribution as the reference dataset R2, a gaussian distribution with mean $\mu=100$ and standard deviation $\sigma=50$.
+At the 2 millionth event mark, we use another Gaussian distribution, this time with mean $\mu=100$ and standard deviation $\sigma=30$.
+At the 3 millionth event, we change the standard deviation to be $\sigma=10$.
+From the 4 millionth event to the end, we use a standard deviation of $\sigma=5$.
+
+In this experiment, we used a half-life $n_{1/2}=62500$ and corresponding sample size of 250,000 events, like in the previous section's experiment 02.
+Figure \ref{fig:JSD-signal-test02} shows the computed JSD signal for the target dataset and Figure \ref{fig:JSD-signal-zoom-test02} zooms in on the first 2 millionth events, before any distribution is changed.
+We observe that once again we have no false positives in the first 2 million events, which is ideal.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-test02.pdf}
+      \caption{Whole time-series JSD signal and threshold}
+      \label{fig:JSD-signal-test02}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-zoom-test02.pdf}
+      \caption{Zoom-in on first 2 million events}
+     \label{fig:JSD-signal-zoom-test02}
+    \end{center}
+\end{figure}
+Furthermore, we observe that the computed JSD signal varies accordingly with the changes in distribution.
+Despite us changing the distributions to subsets of the previous ones to try and deceive our method, our signal accurately represents these changes.
+In Figure \ref{fig:JSD-signal-test02}, we see several growth points that correspond to the points where the generating gaussian distribution parameters changed.
+Hence, we are able to detect changes in distribution parameter changes and quantify how big the change is (the closer to 1 our signal is).
+
+\subsubsection{Multiple Distribution Changes, each a Superset of the Previous One}
+In this experiment, we wanted to do the reverse of the previous one.
+In this one, there are multiple change points but each new distribution spans a domain that is a superset of the previous one.
+Figure \ref{fig:timeseries-r3} shows the reference period used, with 5 million events and one feature following a gaussian distribution of mean $\mu=100$ and standard deviation $\sigma=10$.
+Figure \ref{fig:timeseries-t3} shows the target period used, also with 5 million events and one feature.
+Like in the previous experiment, we change distributions at the 2, 3 and 4 millionth event mark.
+On the other hand, unlike the previous experiment, we now make subsequent distributions cover the previous ones.
+We change the reference gaussian used with mean $\mu=100$ and standard deviation $\sigma=10$ at the 2, 3 and 4 millionth mark, to use standard deviations $\sigma=30$, $\sigma=40$ and $\sigma=50$, respectively.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-r3.pdf}
+      \caption{Time-series for dataset R3}
+      \label{fig:timeseries-r3}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-t3.pdf}
+      \caption{Time-series for dataset T3}
+     \label{fig:timeseries-t3}
+    \end{center}
+\end{figure}
+
+Figure \ref{fig:JSD-signal-test03} shows the JSD signal for the target dataset in this experiment.
+Once again, we were able to detect the changes introduced at the 2, 3 and 4 millionth events.
+We are also able to identify bigger distribution changes, relative to the reference, by observing bigger JSD values.
+Figure \ref{fig:JSD-signal-zoom-test03} shows a zoomed in portion of the original signal, where we can verify that there are no false positives.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-test03.pdf}
+      \caption{Whole time-series JSD signal and threshold}
+      \label{fig:JSD-signal-test03}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-zoom-test03.pdf}
+      \caption{Zoom-in on first 2 million events}
+     \label{fig:JSD-signal-zoom-test03}
+    \end{center}
+\end{figure}
+
+\subsubsection{Return to Normality After Change}
+Our goal with this experiment was to check if our signal returned to its normal values (below the alert threshold) if we changed our distribution to the original reference one, as intended.
+
+In this experiment, we used reference dataset R4 (Figure \ref{fig:timeseries-r4}) and target dataset T4 (Figure \ref{fig:timeseries-t4}). 
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-r4.pdf}
+      \caption{Time-series for dataset R4}
+      \label{fig:timeseries-r4}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-t4.pdf}
+      \caption{Time-series for dataset T4}
+     \label{fig:timeseries-t4}
+    \end{center}
+\end{figure}
+
+The reference dataset R4 had a single feature that followed a Gaussian distribution of $\mu=100$ and $\sigma=5$ for all 5 million events.
+Up until the 2 millionth event, the target dataset T4 followed a Gaussian distribution of $\mu=100$ and $\sigma=5$, similarly to the reference dataset R4.
+At the 2 millionth event, we change the Gaussian parameters to $\mu=100$ and $\sigma=30$.
+At the 3 millionth mark, we once again change the Gaussian parameters to $\mu=100$ and $\sigma=50$.
+Finally, on the 4 millionth event, we return to our reference distribution, a gaussian with $\mu=100$ and $\sigma=5$.
+
+Figure \ref{fig:JSD-signal-test04} shows the computed JSD signal for the target period.
+As expected, our JSD signal increases at the 2 millionth and 3 millionth event mark, because the distribution parameters change.
+At the 4 millionth event, we use the original reference distribution and notice that the JSD signal returns to values below the threshold, considering it a normal state, as we intended.
+Figure \ref{fig:JSD-signal-zoom-test04} is a zoom in on the first 2 millionth events before any change occurs, and we see no false positives.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-test04.pdf}
+      \caption{Whole time-series JSD signal and threshold}
+      \label{fig:JSD-signal-test04}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-zoom-test04.pdf}
+      \caption{Zoom-in on first 2 million events}
+    \label{fig:JSD-signal-zoom-test04}
+    \end{center}
+\end{figure}
+
+\subsubsection{An Attempt to Deceive our Method}
+In this experiment, we attempt to deceive our method by using a different distribution from the reference period but ensuring its domain is roughly the same as the reference one.
+
+We used the reference dataset R5 (Figure \ref{fig:timeseries-r5}) which had 5 million events, with a single feature that followed a gaussian with  $\mu=100$ and $\sigma=10$.
+Our target dataset T5 is represented in Figure \ref{fig:timeseries-t5}.
+For the first 2 million events, the generating distribution used was similar to the one used in the reference dataset, a gaussian with $\mu=100$ and $\sigma=10$.
+However, at the 2 millionth event, we change the gaussian standard deviation to $\sigma=30$.
+Then, at the 4 millionth event, we change the distribution type.
+We use an uniform distribution with lower bound $a=80$ and upper bound $b=120$.
+Note in Figure \ref{fig:timeseries-t5} that this last portion resembles the first one, the reference.
+\begin{center}
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-r5.pdf}
+  \captionof{figure}{Time-series for dataset R5}
+  \label{fig:timeseries-r5}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/timeseries-t5.pdf}
+  \captionof{figure}{Time-series for dataset T5}
+  \label{fig:timeseries-t5}
+\end{minipage}
+\end{center}
+
+Let's take a look at the signal computed for the target period.
+Figure \ref{fig:JSD-signal-test05} shows us the JSD signal for dataset T5 (Figure \ref{fig:timeseries-t5}).
+As expected, we see a rapid growth of JSD value after the 2 millionth event, when we change the Gaussian parameters.
+We then change the distribution type, from a gaussian to a uniform one.
+Despite the latter having a domain contained within the former, we see that our signal is accurate and does not drop below the alert threshold.
+This is still a different distribution that we want to alert.
+Figure \ref{fig:JSD-signal-zoom-test05} is a zoom in of the first 2 millionth events and once again we have no false positives.
+\begin{figure}[!htb]
+\centering
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test05.pdf}
+  \captionof{figure}{JSD signal and threshold}
+  \label{fig:JSD-signal-test05}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test05.pdf}
+  \caption{Zoom in on the first 2 million events}
+  \label{fig:JSD-signal-zoom-test05}
+\end{subfigure}
+\caption{JSD signal and threshold}
+\end{figure}
+
+\subsection{Multi-Feature Analysis}
+Before moving to real data, we performed one last experiment, using synthetic data but this time performing a multi-feature analysis.
+Table \ref{tbl:multi-feat-ref-dataset-distros} shows the underlying generating distributions and parameters for each of the four features ($x_1$, $x_2$, $x_3$ and $x_4$) in the reference multi-feature dataset R6.
+Figures \ref{fig:timeseries-r6-x1}, \ref{fig:timeseries-r6-x2}, \ref{fig:timeseries-r6-x3} and \ref{fig:timeseries-r6-x4} show the reference time-series for features $x_1$, $x_2$, $x_3$ and $x_4$ on the reference dataset R6, respectively.
+
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|ll}
+        \cline{1-3}
+        \textbf{Feature} & \textbf{Reference Distribution} & \textbf{Parameters} &  &  \\ \cline{1-3}
+        $x_1$               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
+        $x_2$               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
+        $x_3$               & Uniform                         & $a=200$, $b=250$       &  &  \\ \cline{1-3}
+        $x_4$               & Uniform                         & $a=50$, $b=125$        &  &  \\ \cline{1-3}
+        \end{tabular}
+    \end{center}
+    \caption{Multi-feature reference dataset and feature distributions}
+    \label{tbl:multi-feat-ref-dataset-distros}
+\end{table}
+
+\begin{figure}[!htb] 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x1.pdf} 
+    \caption{Feature $x_1$ reference time-series} 
+    \label{fig:timeseries-r6-x1} 
+    \vspace{4ex}
+  \end{minipage}%%
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x2.pdf} 
+    \caption{Feature $x_2$ reference time-series} 
+    \label{fig:timeseries-r6-x2} 
+    \vspace{4ex}
+  \end{minipage} 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x3.pdf} 
+    \caption{Feature $x_3$ reference time-series} 
+    \label{fig:timeseries-r6-x3} 
+    \vspace{4ex}
+  \end{minipage}%% 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x4.pdf} 
+    \caption{Feature $x_4$ reference time-series} 
+    \label{fig:timeseries-r6-x4} 
+    \vspace{4ex}
+  \end{minipage} 
+\end{figure}
+
+We used a target dataset T6 with 5 million events and the same four features ($x_1$, $x_2$, $x_3$ and $x_4$).
+Feature $x_1$ followed the reference distribution for the 5 million events of the target dataset.
+We did not change this feature underlying distribution to serve as a control, \textit{i.e.}, we expect no alerts for this one.
+Figure \ref{fig:timeseries-t6-x1} shows the time-series for the target dataset T6 and feature $x_1$.
+
+Feature $x_2$'s underlying distribution changed four times and returned to normal, so we expect to see alarms with increasing divergence values but eventually no more alarms.
+Figure \ref{fig:timeseries-t6-x2} shows the time-series for the target dataset T6 and feature $x_2$.
+Feature $x_2$ followed a gaussian with $\mu=100$ and $\sigma=5$ like in the reference period up to the 1 millionth event mark, then a gaussian with $\mu=100$ and $\sigma=10$ up to the 2.5 millionth event, then a gaussian $\mu=100$ and $\sigma=15$ until the 3 millionth event, then a gaussian $\mu=100$ and $\sigma=20$ until the 4 millionth event and finally returned to the original gaussian with $\mu=100$ and $\sigma=5$ until the last event.
+Table \ref{tbl:multi-feat-x2-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_2$.
+\begin{table}[!htb]
+    \begin{center}
+    \begin{tabular}{|c|c|c|c|l}
+    \cline{1-4}
+    \textbf{From Event} & \textbf{To Event} & \textbf{Distribution} & \multicolumn{1}{l|}{\textbf{Parameters}} &  \\ \cline{1-4}
+    0                   & 1,000,000         & Gaussian              & $\mu=100, \sigma=5$                      &  \\ \cline{1-4}
+    1,000,000           & 2,500,000         & Gaussian              & $\mu=100, \sigma=10$                     &  \\ \cline{1-4}
+    2,500,000           & 3,000,000         & Gaussian              & $\mu=100, \sigma=15$                     &  \\ \cline{1-4}
+    3,000,000           & 4,000,000         & Gaussian              & $\mu=100, \sigma=20$                      &  \\ \cline{1-4}
+    4,000,000           & 5,000,000         & Gaussian              & $\mu=100, \sigma=5$                      &  \\ \cline{1-4}
+    \end{tabular}
+    \end{center}
+    \caption{Feature $x_2$ generating distributions}
+    \label{tbl:multi-feat-x2-changes}
+\end{table}
+
+Feature $x_3$'s underlying distribution changed three times and never returned to normal, so we expect to see alarms with increasing divergence values up until the end of the stream analysis.
+Figure \ref{fig:timeseries-t6-x3} shows the time-series for the target dataset T6 and feature $x_3$.
+Feature $x_3$ followed a uniform distribution with $a=200$ and $b=250$ like in the reference period up to the 1.5 millionth event mark, then a uniform distribution with $a=200$ and $b=220$ up to the 2 millionth event, then a uniform distribution with $a=180$ and $b=200$ until the 3.5 millionth event and finally a gaussian with $\mu=50$ and $\sigma=20$ until the last event.
+Table \ref{tbl:multi-feat-x3-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_3$.
+\begin{table}[!htb]
+    \begin{center}
+    \begin{tabular}{|c|c|c|c|l}
+    \cline{1-4}
+    \textbf{From Event} & \textbf{To Event} & \textbf{Distribution} & \multicolumn{1}{l|}{\textbf{Parameters}} &  \\ \cline{1-4}
+    0                   & 1,500,000         & Uniform               & $a=200, b=250$                           &  \\ \cline{1-4}
+    1,500,000           & 2,000,000         & Uniform               & $a=200, b=220$                           &  \\ \cline{1-4}
+    2,000,000            & 3,500,000         & Uniform               & $a=180, b=200$                           &  \\ \cline{1-4}
+    3,500,000           & 5,000,000         & Gaussian              & $\mu=50, \sigma=20$                      &  \\ \cline{1-4}
+    \end{tabular}
+    \end{center}
+    \caption{Feature $x_3$ generating distributions}
+    \label{tbl:multi-feat-x3-changes}
+\end{table}
+
+Feature $x_4$'s underlying distribution changed two times and returned to normal, so we expect to see no alarms towards the end of the stream analysis.
+Figure \ref{fig:timeseries-t6-x4} shows the time-series for the target dataset T6 and feature $x_4$.
+Feature $x_4$ followed a uniform distribution like the reference one, with $a=50$ and $b=125$ up until the 1.5 millionth event.
+After that, it followed a uniform distribution with $a=125$ and $b=200$ until the 3 millionth event where it resumed its original uniform distribution (with $a=50$ and $b=125$) until the end.
+Table \ref{tbl:multi-feat-x4-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_4$.
+\begin{table}[!htb]
+    \begin{center}
+    \begin{tabular}{|c|c|c|c|l}
+    \cline{1-4}
+    \textbf{From Event} & \textbf{To Event} & \textbf{Distribution} & \multicolumn{1}{l|}{\textbf{Parameters}} &  \\ \cline{1-4}
+    0                   & 1,500,000         & Uniform               & $a=50, b=125$                            &  \\ \cline{1-4}
+    1,500,000           & 3,000,000         & Uniform               & $a=125, b=200$                           &  \\ \cline{1-4}
+    3,000,000           & 5,000,000         & Uniform               & $a=50, b=125$                            &  \\ \cline{1-4}
+    \end{tabular}
+    \end{center}
+    \caption{Feature $x_4$ generating distributions}
+    \label{tbl:multi-feat-x4-changes}
+\end{table}
+
+\begin{figure}[!htb] 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x1.pdf} 
+    \caption{Feature $x_1$ target time-series} 
+    \label{fig:timeseries-t6-x1} 
+    \vspace{4ex}
+  \end{minipage}%%
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x2.pdf} 
+    \caption{Feature $x_2$ target time-series} 
+    \label{fig:timeseries-t6-x2} 
+    \vspace{4ex}
+  \end{minipage} 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x3.pdf} 
+    \caption{Feature $x_3$ target time-series} 
+    \label{fig:timeseries-t6-x3} 
+    \vspace{4ex}
+  \end{minipage}%% 
+  \begin{minipage}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x4.pdf} 
+    \caption{Feature $x_4$ target time-series} 
+    \label{fig:timeseries-t6-x4} 
+    \vspace{4ex}
+  \end{minipage} 
+\end{figure}
+
+In this experiment, we used a tuple-based half-life $n_{1/2}=62500$, \textit{i.e.}, sampling and target windows of 250,000 tuples.
+We used 100 bins for the reference histogram plus the two special infinity bins.
+In this multi-feature scenario, we used the Holm-Bonferroni correction discussed in Section \ref{sec:holmbonferroni} and set the Family-Wise Error Rate (FWER) to 1\% (FWER = 0.01).
+Furthermore, instead of doing 1000 samples as we did previously in an ad-hoc way, we use Equation \ref{eq:optimal-n-samples} to determine the minimum number of samples to make while ensuring $\gamma=1\%$ (recall Section \ref{sec:nsamples}).
+For our four feature analysis scenario, using as inputs  $\gamma=0.01$ and $FWER=0.01$ in Equation \ref{eq:optimal-n-samples} tells us we need to make 1840 samples minimum.
+Using 300 Spark executors, each given 3G of memory, we finished the batch phase in 418 seconds.
+
+In Section \ref{sec:stream-phase}, we mention that, in our stream analysis, we perform our multiple hypothesis divergence tests on a fixed user-defined frequency.
+Finding an optimal frequency value for the test falls out of scope for this thesis, but ideally, we set it to be large enough so that we do fewer computations and achieve higher throughput but still get alerts in useful time.
+In this experiment, we performed the divergence test with the Holm-Bonferroni correction \ref{sec:holmbonferroni} for every new event (frequency of 1 event).
+We did this to collect a list of ordered corrected p-values for each feature.
+A corrected p-value (derived from Equation \ref{eq:corrected-pvalue}) is given by:
+\begin{equation}
+    p-value_{corrected} = p-value_{distance_{x}} * (m - k)
+    \label{for:corrected-pvalue}
+\end{equation}
+With this list, for each feature, we can plot the corrected p-values on each event and compare it with the FWER threshold $\alpha$.
+Note that despite performing the test at the maximum frequency (in real-time) we still achieved a throughput of 17,921 transactions per second (TPS), even if only for four features.
+
+In the following plots, unlike the JSD signal plots presented before, a feature is in a divergent state if its p-value is below the threshold, according to Equation \ref{eq:corrected-pvalue}.
+Working with corrected p-values (Formula \ref{for:corrected-pvalue}), a feature is considered divergent if its corrected p-value is below the threshold.
+
+In Figure \ref{fig:x1-corrected-pvalues} we see the plot of corrected p-values for feature $x_1$ and the probability threshold black horizontal line at the bottom.
+Recall that feature $x_1$'s values were generated by a Gaussian distribution for both the reference period (Figure \ref{fig:timeseries-r6-x1}) and the target period (Figure \ref{fig:timeseries-t6-x1}), with the same parameters.
+We observe that the corrected p-values never pass below the threshold, as we expected.
+Hence, this feature is never alerted as divergent.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x1-corrected-pvalues.png}
+      \caption{Feature $x_1$ corrected p-values and threshold}
+      \label{fig:x1-corrected-pvalues}
+    \end{center}
+\end{figure}
+
+Figure \ref{fig:x2-corrected-pvalues} shows a plot of feature $x_2$'s corrected p-values and the FWER threshold.
+Feature $x_2$ underlying distribution was changed four times (Table \ref{tbl:multi-feat-x2-changes}): at the 1, 2.5, 3 and 4 millionth events.
+The first change is consistent with the sudden drop in corrected p-value for this feature, as we cross below the threshold and enter a divergent state.
+Around a thousand events after the change we saw p-values of 0.2\% (or 0.002).
+The values keep dropping as time passes and as we introduce bigger changes in distribution.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x2-corrected-pvalues.png}
+      \caption{Feature $x_2$ corrected p-values and threshold}
+      \label{fig:x2-corrected-pvalues}
+    \end{center}
+\end{figure}
+We resume the reference or original distribution at the 4 millionth event.
+However, as we can see from a close-up on the original plot (Figure \ref{fig:x2-corrected-pvalues-zoom}), it takes around 600,000 events for the target sliding histogram aggregation to once again resemble the reference one, resulting in a high enough p-value and thus stop alerting $x_2$ as divergent.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x2-corrected-pvalues-zoom2.png}
+      \caption{Feature $x_2$ zoomed in corrected p-values and threshold}
+      \label{fig:x2-corrected-pvalues-zoom}
+    \end{center}
+\end{figure}
+
+Feature $x_3$ was the only feature that never returned to its original distribution and is the only one that should be reported until the end as divergent.
+This is in fact the case as illustrated in Figure \ref{fig:x3-corrected-pvalues}.
+The first change to $x_3$'s generating distribution was introduced at the 1.5 millionth event and we clearly see the corrected p-value rapidly approaching values close to zero.
+For the rest of the dataset, this $x_3$'s p-value remains below the threshold, as we never swap to its original distribution.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x3-corrected-pvalues.png}
+      \caption{Feature $x_3$ corrected p-values and threshold}
+      \label{fig:x3-corrected-pvalues}
+    \end{center}
+\end{figure}
+
+Finally, feature $x_4$'s corrected p-value plot is given by Figure \ref{fig:x4-corrected-pvalues}.
+Similarly to feature $x_2$, we resume the original distribution at a certain point, in this case at the 3-millionth event.
+However, as we can see from a close-up on the original plot (Figure \ref{fig:x4-corrected-pvalues-zoom}), it takes around 810,000 events for the target sliding window aggregation to be considered similar enough to the reference one and stop alerting $x_4$ as divergent.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x4-corrected-pvalues.png}
+      \caption{Feature $x_4$ corrected p-values and threshold}
+      \label{fig:x4-corrected-pvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.8]{figures/x4-corrected-pvalues-zoom2.png}
+      \caption{Feature $x_4$ zoomed in corrected p-values and threshold}
+      \label{fig:x4-corrected-pvalues-zoom}
+    \end{center}
+\end{figure}
+
+With this experiment, we further validate that the p-value is a measure of how divergent a feature is: the closer to zero, the lower the probability of the statistical test result and hence the more divergent a feature is considered to be.
+At the 1.5 millionth event, we swap both $x_3$ and $x_4$'s distribution.
+For the first 2,535,000 events, $x_4$ has a lower p-value than $x_3$ and hence is considered more divergent.
+However, after that, we change $x_3$'s distribution so aggressively that $x_3$ starts being reported as most divergent, with a lower p-value than $x_4$.
+
+\clearpage
+\section{Experiments with Real Data}
+
+Due to time constraints, we experimented with data from only one merchant. The merchant's entity as well as feature names will remain anonymous due to confidentiality reasons.
+
+Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that, combined, covered a period of roughly 1 year and 9 months.
+In this experiment, we used the older one as the reference dataset and the more recent one as the target period.
+Both datasets were already feature-engineered \cite{Domingos-ML-Feat-Eng} and had a total of 183 numerical plus numerically encoded categorical features (\textit{e.g.}, one-hot encoded or ordinal encoded \cite{categoricalencoding}).
+Table \ref{tbl:merchant1-datasets-summary} shows summary statistics for both datasets.
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|c|}
+        \hline
+        \textbf{Dataset} & \textbf{From}              & \textbf{To}           & \multicolumn{1}{l|}{\textbf{Transactions}} \\ \hline
+        Reference        & Tuesday, September 5, 2017 & Friday, June 29, 2018 & 1,604,509                                  \\ \hline
+        Target           & Friday, June 29, 2018      & Sunday, June 30, 2019 & 4,032,505                                  \\ \hline
+        \end{tabular}
+        \caption{Summary statistics for reference and target merchant datasets}
+        \label{tbl:merchant1-datasets-summary}
+    \end{center}
+\end{table}
+
+In the batch phase, we used 100 Spark executors, each with 12G of memory.
+We determined that one week's worth of data for this merchant was on average equivalent to 37,803 events.
+We set the half-life to be $n_{1/2}=37,803$ and discard events whose EMA weight is 6\% or less, which is equivalent to using sampling windows of data of four times the half-life (as detailed in Section \ref{sec:ema-hist}).
+Hence, our samples's windows had 151,212 events.
+Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:nsamples}) of 10\% we needed to make a minimum of 42,137 samples.
+Our batch analysis took 5 hours and 34 minutes.
+
+After the batch phase, we had the necessary inputs for the streaming phase (detailed in Section \ref{sec:batch-artifacts-summary}), namely, for each feature, the reference histograms, the list of observed divergence values and the last sample's histogram.
+We then used the target dataset, simulating a stream of ordered events.
+Our monitoring system processed the dataset event by event in sequential, time-ordered fashion, emulating a stream computation and reported divergent features with associated p-values.
+We performed our multiple hypotheses with the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for every 1000 events.
+It took us 382,363 milliseconds (roughly 6 minutes) to process the 4,032,505 transactions, amounting to a throughput of 10,546 transactions per second (TPS).
+
+This merchant and in particular these datasets were known to suffer a lot from concept drift.
+Our system produced a large list of alerts, from nearly the beginning of the target period until the end.
+Due to a large number of features (183), we only show some plots.
+Furthermore, we refer to features with anonymized identifiers, like feature $x_j$.
+In the following plots, we show the corrected p-values (Formula \ref{for:corrected-pvalue}) for some features at each event of the time-series and the FWER threshold set at 1\%.
+A feature is considered divergent at a given timestamp if below that threshold line.
+
+Figure \ref{fig:merchant-x1-reference} is a plot of the reference time-series of feature $x_1$ and Figure \ref{fig:merchant-x1-target} the corresponding target time-series.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x1-reference.pdf}
+      \caption{Merchant feature $x_1$ reference time-series}
+      \label{fig:merchant-x1-reference}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x1-target.pdf}
+      \caption{Merchant feature $x_1$ target time-series}
+      \label{fig:merchant-x1-target}
+    \end{center}
+\end{figure}
+In Figure \ref{fig:merchant-x1-correctedpvalues} we plot the corrected p-values for the each event of the target period.
+The reference and target periods look very similar and in fact we have very few alarms for this feature, \textit{i.e.}, the number of times the line crosses to the bottom half of the plot, imposed by the threshold line.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x1-correctedpvalues.pdf}
+      \caption{Merchant feature $x_1$ corrected p-values}
+      \label{fig:merchant-x1-correctedpvalues}
+    \end{center}
+\end{figure}
+Furthermore, the alarms appear to be roughly in the middle portion of the time-series, where in fact the target period appears to have less points.
+We can see roughly three alarms.
+The first one in early January 2019 for a few days, the second one lasting about a week at the end of January 2019 and the last one in the second half of February 2019 again for a few days.
+
+Now let us look at features $x_2$ and $x_3$.
+Figures \ref{fig:merchant-x2-reference} and \ref{fig:merchant-x2-target} show feature $x_2$'s time-series plots for the reference and target periods, respectively.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x2-reference.pdf}
+      \caption{Merchant feature $x_2$ reference time-series}
+      \label{fig:merchant-x2-reference}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x2-target.pdf}
+      \caption{Merchant feature $x_2$ target time-series}
+      \label{fig:merchant-x2-target}
+    \end{center}
+\end{figure}
+Figures \ref{fig:merchant-x3-reference} and \ref{fig:merchant-x3-target} show feature $x_3$'s time-series plots for the reference and target periods, respectively.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x3-reference.pdf}
+      \caption{Merchant feature $x_3$ reference time-series}
+      \label{fig:merchant-x3-reference}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x3-target.pdf}
+      \caption{Merchant feature $x_3$ target time-series}
+      \label{fig:merchant-x3-target}
+    \end{center}
+\end{figure}
+\clearpage
+Relative to $x_1$, features $x_2$ and $x_3$ also have very similar reference and target periods.
+Just like feature $x_1$, we expect to have some alarms but still not too many.
+Once again, to visualize the alarms, we plot the corrected p-values for each event of the target time-series for each feature.
+Figure \ref{fig:merchant-x2-correctedpvalues} is a plot of feature $x_2$'s corrected p-values and the FWER threshold of 1\%.
+Likewise, Figure \ref{fig:merchant-x3-correctedpvalues} shows the same plot but for feature $x_3$.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x2-correctedpvalues.pdf}
+      \caption{Merchant feature $x_2$ corrected p-values}
+      \label{fig:merchant-x2-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x3-correctedpvalues.pdf}
+      \caption{Merchant feature $x_3$ corrected p-values}
+      \label{fig:merchant-x3-correctedpvalues}
+    \end{center}
+\end{figure}
+The corrected p-values plot shows us once again that both $x_2$ and $x_3$ features had some alarms but eventually rose above the threshold and resumed their reference distribution.
+This is expected from features whose reference and target time-series plots look so similar.
+
+Next, we analyze a feature with very different reference and target time-series.
+Feature $x_4$'s reference time-series is presented in Figure \ref{fig:merchant-x4-reference} and its target period in Figure \ref{fig:merchant-x4-target}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x4-reference.pdf}
+      \caption{Merchant feature $x_4$ reference time-series}
+      \label{fig:merchant-x4-reference}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x4-target.pdf}
+      \caption{Merchant feature $x_4$ target time-series}
+      \label{fig:merchant-x4-target}
+    \end{center}
+\end{figure}
+Notice that even if for the most part both reference and target periods values belong to a domain between 0 and 1, at a certain point in time, in the target time-series, a lot of -1 values appear.
+In Figure \ref{fig:merchant-x4-correctedpvalues} we plot the corrected p-values for feature $x_4$ target time-series.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x4-correctedpvalues.pdf}
+      \caption{Merchant feature $x_4$ corrected p-values}
+      \label{fig:merchant-x4-correctedpvalues}
+    \end{center}
+\end{figure}
+Like the previous analyzed features, $x_4$ also drops below the threshold some times.
+However, unlike the other features, where feature values return to normal and alarms stopped being raised, $x_4$ remains below the threshold line until the end of the target period, where a small spike occurs.
+The last time feature $x_4$ was ever above the threshold line (and hence considered normal) was before the appearance of the -1 values in the target period.
+This seems to work as expected: -1 values were so off the reference domain that for here on out, as long as these values keep appearing, we will not stop reporting $x_4$ as divergent.
+
+The -1 value is often used as a placeholder for missing values, \textit{i.e.}, if some problem occurs and the value for a feature is missing at a given timestamp or event, -1 is used.
+A follow-up question that is raised is if we should include some mechanism of dealing with missing values in our method.
+However, we discard this idea and defend that we should not: missing values are indeed something we want to detect.
+For instance, if a certain device responsible for reporting one feature stops working, we want to know that.
+This scenario is likely to happen in a distributed Internet-of-Things (IoT) system with hundreds of devices transmitting data.
+
+%feats = [max_amt_rate_per_location_72h, num_preauths_per_card_1w, user_avg_amount_completions_1h, completions_rate_per_location_zip3_72h, completion_rate_per_card_72h]
+%low drift:
+    %x1 - max_amt_rate_per_location_72h
+    %x2 - num_preauths_per_card_1w
+    %x3 - user_avg_amount_completions_1h
+%high drift:
+    %x4 - completion_rate_per_card_72h
+    %x5 - dif_last_and_current_trx
+\subsection{Same Reference and Target Periods}
+In this experiment, we set out to test our system in a low drift scenario.
+To guarantee the reference and target periods are as similar as possible, we decided to use the same dataset to work as reference and target.
+We chose to use the reference dataset from the previous experiment (Reference of Table \ref{tbl:merchant1-datasets-summary}) and feed it as input for both our batch and stream analysis as the reference and target periods, respectively.
+We expect to have less alarms for each feature but not necessarily zero.
+
+We first show the corrected p-value plots when both reference and target periods are the same, for the previous features $x_1$, $x_2$ and $x_3$ in Figures \ref{fig:merchant2-x1-correctedpvalues}, \ref{fig:merchant2-x2-correctedpvalues} and \ref{fig:merchant2-x3-correctedpvalues}, respectively.
+
+Feature $x_1$ has less alerts in this experiment where we use equal reference and target periods.
+However, the alarms last for about two months.
+This is unexpected as we predicted we would see no alarms or few ones with short duration.
+Likewise, we observe that for feature $x_2$ (Figure \ref{fig:merchant2-x2-correctedpvalues}) and $x_3$ (Figure \ref{fig:merchant2-x3-correctedpvalues}) we still have many alarms and some lasting up to one month.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x1-correctedpvalues.pdf}
+      \caption{Merchant feature $x_1$ corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x1-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x2-correctedpvalues.pdf}
+      \caption{Merchant feature $x_2$ corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x2-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x3-correctedpvalues.pdf}
+      \caption{Merchant feature $x_3$ corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x3-correctedpvalues}
+    \end{center}
+\end{figure}
+We do not have a clear explanation for this and due to time constraints we leave further tests to future work.
+We suspect the issue may be that there is a difference between the batch and streaming phases.
+In batch we have subsets of events and compute EMA-like histogram aggregations based on those.
+In streaming, however, we have a continuous stream of data and never expire the 6\% of data that in batch is not present, because it is a finite window.
+
+Now we analyze feature $x_4$ from the previous experiment, as this one was the one with the highest measured drift.
+Recall that feature $x_4$'s target time-series (Figure \ref{fig:merchant-x4-target}) was significantly different from the reference one (Figure \ref{fig:merchant-x4-reference}).
+Because of this, the corrected p-values plot from the previous experiment (Figure \ref{fig:merchant-x4-correctedpvalues}) crossed our threshold line multiple times and remained in an alert state for most of the target time-series.
+In this new experiment, we used the reference period twice.
+Because of this, the corrected p-values plot is now different.
+Figure \ref{fig:merchant2-x4-correctedpvalues} shows the corrected p-values plot for feature $x_4$ when we used the reference period for both phases.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x4-correctedpvalues.pdf}
+      \caption{Merchant feature $x_4$ corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x4-correctedpvalues}
+    \end{center}
+\end{figure}
+Using the same period twice theoretically means we have no drift at all.
+Hence, we expected to see no alerts.
+Figure \ref{fig:merchant2-x4-correctedpvalues} shows that we have a few alerts but significantly less than in Figure \ref{fig:merchant-x4-correctedpvalues}.
+Furthermore, in this new experiment, $x_4$ eventually stops being reported at all until the end of the dataset and ends at a high p-value.
+Recall that the higher the p-value the less divergent a feature is.
+Overall, we see that this feature has fewer alarms and stabilizes well above our alarm threshold which is coherent since we used the same dataset as reference and target periods.
+
+We now introduce a new feature, $x_5$.
+Feature $x_5$ also suffers from a large amount of drift.
+In the reference and target datasets presented in Table \ref{tbl:merchant1-datasets-summary}, feature $x_5$ reference time-series plot is given by Figure \ref{fig:merchant-x5-reference} and the target time-series by Figure \ref{fig:merchant-x5-target}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x5-reference.pdf}
+      \caption{Merchant feature $x_5$ reference time-series}
+      \label{fig:merchant-x5-reference}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x5-target.pdf}
+      \caption{Merchant feature $x_5$ target time-series}
+      \label{fig:merchant-x5-target}
+    \end{center}
+\end{figure}
+The density in the upper domain of values is clearly different in both reference and target periods.
+Hence, in our first experiment, this feature had a corrected p-values plot with two alarm states, where the second one lasted until the end of the target period, with the lowest possible p-value (zero), represented in Figure \ref{fig:merchant-x5-correctedpvalues}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant-x5-correctedpvalues.pdf}
+      \caption{Merchant feature $x_5$ corrected p-values}
+      \label{fig:merchant-x5-correctedpvalues}
+    \end{center}
+\end{figure}
+
+However, when using the same reference and target periods, in this case using the time-series in Figure \ref{fig:merchant-x5-reference} twice, we achieve our lowest possible drift scenario.
+The difference in corrected p-values is clear.
+Figure \ref{fig:merchant2-x5-correctedpvalues} shows this plot when using equal reference and target periods.
+We observe that our corrected p-values never go below the threshold line, in fact, they stay well above it.
+This is expected behavior when using equal reference and target periods.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant2-x5-correctedpvalues.pdf}
+      \caption{Merchant feature $x_5$ corrected p-values with equal reference and target periods}
+      \label{fig:merchant2-x5-correctedpvalues}
+    \end{center}
+\end{figure}
+
+\subsection{Building an Accurate Divergence Measure Distribution}
+Our method comprises two phases: the batch and the streaming analysis.
+We can not conclude in which phase an error that could explain the last experiment's results exists.
+To isolate both phases, we conduct this last experiment where instead of using the distribution of divergence measures built in the batch phase we use a distribution we know to be the best fit for the streaming period.
+
+Our best fit divergence distribution was built over the streaming period.
+To do this, we processed the streaming events once and saved all the JSD measures.
+We binned these JSD values with 50,000 bins all with equal heights (or counts).
+Then we performed once again the streaming analysis over the same period but this time using the previously built distribution for which we know it is the best possible fit since it was computed over the same period, built event by event.
+We set the FWER to be 1\% so we expect to see some alarms, \textit{i.e.}, for JSD measures above the 99th percentile.
+
+Recall that in the last experiment, for feature $x_1$, using the same target and reference periods, we had two large alarms that lasted months (Figure \ref{fig:merchant2-x1-correctedpvalues}).
+Using the same periods but with the newly computed distribution of divergence measures we obtain the corrected p-values plot shown in Figure \ref{fig:merchant3-x1-correctedpvalues}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x1-correctedpvalues.pdf}
+      \caption{Merchant feature $x_1$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x1-correctedpvalues}
+    \end{center}
+\end{figure}
+This plot actually makes sense and represents expected behavior.
+Notice that we have only a few short-lasting alarms, which correspond to the JSD measures above the 99th percentile.
+These can be thought of as false positives.
+In reality, the definition of the multi-test we perform is exactly this: alert events above the 99th percentile (FWER=1\%).
+
+Figures \ref{fig:merchant3-x2-correctedpvalues}, \ref{fig:merchant3-x3-correctedpvalues}, \ref{fig:merchant3-x4-correctedpvalues} and \ref{fig:merchant3-x5-correctedpvalues} show the corrected p-value plots for this experiment for features $x_2$, $x_3$, $x_4$ and $x_5$, respectively.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x2-correctedpvalues.pdf}
+      \caption{Merchant feature $x_2$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x2-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x3-correctedpvalues.pdf}
+      \caption{Merchant feature $x_3$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x3-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x4-correctedpvalues.pdf}
+      \caption{Merchant feature $x_4$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x4-correctedpvalues}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/merchant3-x5-correctedpvalues.pdf}
+      \caption{Merchant feature $x_5$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
+      \label{fig:merchant3-x5-correctedpvalues}
+    \end{center}
+\end{figure}
+The analysis performed for feature $x_1$ applies for all of these features.
+Notice that for all plots we have very few alarms of very short durations.
+This coherency in results across multiple features seems to indicate that our streaming analysis is accurate and detects the outliers accurately if the divergence measure distribution used is well built.
+
+\newpage
+\section{Conclusions}
+In this section, we summarize our experimental tests, results and uncovered issues.
+
+We began by performing single-feature analysis of synthetic data.
+The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported.
+Still working with synthetic data, we performed our first multi-feature analysis.
+In this scenario, we applied the Holm-Bonferroni multiple test correction to the p-values of each hypothesis.
+Once again, each feature was correctly reported as divergent at the timestamps where we changed the underlying generating distribution.
+
+We then moved on to real data.
+Due to time constraints and the scope of this thesis, we experimented only with data from one merchant.
+In our first experiment with real data, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms.
+A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift.
+
+In our second experiment with real data, we attempted to reduce concept drift to a maximum by using the same period as reference and target.
+To our surprise, we still produced a large number of long-lasting alarms.
+We defer the investigation of the root cause of the problem to future work and leave a few untested hypotheses as solutions for this issue, like the possibility that the EMA-like histograms in batch work with subsets of data and the streaming ones work with infinite amounts of data.
+In other words, for the streaming EMAs, the previous events never have their contribution set to zero, even though the weights tend to zero, whereas the batch EMAs worked with a finite dataset.
+Additionally, our sampling may have been insufficient and hence the divergence measure distribution was inaccurate.
+To test the first hypothesis, experiments with higher smoothing factors for streaming EMAs should be carried out.
+To test the second one, a lower $\gamma$ parameter should be used in the batch phase, hence increasing the number of samples to be made, potentially building a more accurate distribution of JSD values.
+
+In our final experiment, we aim to build the most accurate as possible distribution of divergence measures by building it based on the measured divergences observed in streaming.
+In our second streaming run, we used this distribution and observed that the alarms produced were very few and short-lasting ones.
+The few alarms raised correspond to divergence measures above the 99th percentile.
+With this experiment, we conclude our streaming analysis is correctly built but depends a lot on how accurate the divergence measure distribution is.
+As discussed, future tests should be carried out where more samples are made, to ensure a more accurate divergence measure distribution is computed.
 
@@
 \label{sec:Discussion}
-
+\textcolor{red}{TODO}
 Fight

### 2020-07-19 chapters/validation.tex
....
Experiments, images scaled to two-column format

@@
     \end{center}
-    \caption{Experiments with varying half-life}
+    \caption{Experiments with varying half-lives}
     \label{tbl:tests-half-life}
@@
 All experiments used 300 Spark executor instances.
-From the Table we conclude that the larger the half-life, the larger the sample size.
+Since our sample size depends on the half-life (four times larger), the larger the half-life, the larger the sample size.
 Larger samples imply longer processing times when computing the EMA-like histogram for each sample.
@@
 Finding an optimal frequency value for the test falls out of scope for this thesis, but ideally, we set it to be large enough so that we do fewer computations and achieve higher throughput but still get alerts in useful time.
-In this experiment, we performed the divergence test with the Holm-Bonferroni correction \ref{sec:holmbonferroni} for every new event (frequency of 1 event).
+In this experiment, we performed the divergence test with the Holm-Bonferroni correction for every new event (frequency of 1 event).
 We did this to collect a list of ordered corrected p-values for each feature.

### 2020-07-19 paper.tex
Experiments, images scaled to two-column format

@@
 We generate alerts based on a user-defined probability threshold, for each of an event's numerical fields, denominated features in our context.
-We evaluate our method through a series of tests, some with synthetic data sets and some with real data.
+We evaluate our method through a series of tests, some with synthetic datasets and some with real data.
 We further split the tests into single and multi-feature analysis.
@@
 In the past decade, applications have become increasingly data-driven, placing data at the center of application design.
-For different use cases, the data is processed in different ways for different purposes.
+For different use cases, data is processed in different ways for different purposes.
 For instance, e-commerce platforms such as Alibaba need to process a large number of daily transactions while ensuring that sales run smoothly and that products are delivered to customers' homes.
-Streaming services, like YouTube and Netflix, try to guarantee that media content reaches up to millions of users simultaneously.
-Social networks are responsible for generating large volumes of data.
-Twitter, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day.
+Live streaming services, like YouTube and Netflix, try to guarantee that media content reaches up to millions of users simultaneously.
+Social networks are responsible for generating large volumes of data: Twitter, for example, generates more than 500 million \textit{tweets} (posts in Twitter) per day.
 \textit{Tweets} may contain text, media content or both.
@@
 The batch analysis is performed on a static in time window which we call the reference window.
-The stream analysis is performed over a exponential tail which emulates a sliding window that groups data stream events.
+The stream analysis is performed over an exponential tail aggregation which emulates a sliding window that groups data stream events.
 We aggregate the contents of the reference window with an exact histogram and the sliding window with an approximate histogram aggregation based on Exponential Moving Averages (EMAs).
@@
 \subsection{Batch Analysis Phase} \label{sec:batch-phase}
-The main goals of the batch phase are to build our reference aggregation and the distribution of divergence values, for each feature.
+The main goals of the batch phase are to build the reference histogram and the distribution of divergence values, for each feature.
 Streaming divergence values will be compared to this distribution of values later to find out how probable a value is to occur (if it is very low, an alert is raised).
@@
 Secondly, we aim to build the distribution of expected distance or divergence values, to later on threshold the observed distance values during the streaming phase.
-To that end, we make $S$ samples of transactions, each with the same tuple-based size of the sliding window we emulate in streaming.
+To that end, we make $S$ samples of transactions of $K$ tuples ($S$ and $K$ detailed later).
 Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.
@@
 Hence, we claim we now know the distribution of expected distance values.
-Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if the probability is below a certain threshold.
+Given a new distance value measured between another random histogram and the reference one, we can compute its probability and produce alerts if the probability is below a certain threshold.
 Note that the histograms for each sample are EMA-based histograms just like the streaming histogram.
@@
 When we boot up our system in the streaming phase we initially have empty histogram aggregations.
-To avoid this burn-in period, we propose saving the last sample's histogram for each feature so that the streaming EMA-based histogram aggregation is initialized with it.
+To avoid this burn-in period, we propose saving the last sample's histogram for each feature so that the streaming EMA-based histogram aggregation is initialized with its counts.
 
@@
 \subsection{Streaming Phase} \label{sec:stream-phase}
-In the streaming phase, the system will process event by event and perform a change detection test periodically.
+In the streaming phase, the system will process event by event, update the streaming aggregations and periodically perform a change detection test.
+
 For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase.
-Then, for each incoming event and each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
+Then, for each incoming event and each of the event's features, we update our streaming histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}.
 This is all we do, regarding event processing.
@@
 We periodically perform a multi-feature change detection test.
-The test is done by first computing the divergence (here denoted as $div_x$) between the reference histogram for a feature \textit{x} and the current streaming EMA histogram for $x$.
+The test is done by first computing the divergence (here denoted as $div_x$) between the reference histogram for a feature \textit{x} and the current streaming histogram for $x$.
 This value is tested against the known distribution of distance values for feature \textit{x}.
@@
 \[pvalue(div_x) = 1 - percentile(div_x)\]
-For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or below a probability of \textit{p-value} = 0.01.
+For example, if we define the minimum probability threshold to be 1\%, it means the system will raise an alert for distance values above the 99th-percentile or below a probability \textit{p-value} = 0.01.
 We alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
@@
 
-\subsubsection{Multiple Test Correction} \label{sec:multi-test}
+\subsubsection{Multiple Test Correction with Holm-Bonferroni} \label{sec:multi-test}
 Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis.
@@
 In statistics, the Family-Wise Error Rate (FWER) \cite{MultivariateMT, MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not.
-The Holm-Bonferroni method is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.
+The Holm-Bonferroni method is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis (or test performed).
 The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature.
@@
 \subsection{Single-Feature Analysis}
-As an initial step, we started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
+We started by validating our system using single-feature datasets before moving to multi-feature scenarios to facilitate system validation, considering the complexities of multi-feature datasets and the problem of multiple test correction.
 
 \subsubsection{Varying Sample Sizes}
-We began with single-feature scenarios and by experimenting with different tuple-based half-lives $n_{1/2}$.
+We began with single-feature scenarios, experimenting with different tuple-based half-lives $n_{1/2}$.
 The half-life $n_{1/2}$ parameter in turn controls the size of the samples we make in the batch phase and the streaming EMA tail size.
@@
 
-From a functional point of view, were we able to detect the abrupt change introduced in the target dataset T1?
 Recall that from the 2.5 millionth event forward we change the distribution from a gaussian to an uniform one (Figure \ref{fig:timeseries-t1}).
@@
 In experiment 02, we use larger sampling windows.
-Remember our target sliding window in streaming is of the same size as the sampling windows.
+Remember our EMa tail size in streaming is proportional to the sampling windows.
 In the second experiment, we increase the size of the sampling windows by 100 times.
@@
 Notice also that we do not have false positives in the first half of the signal.
-This is because we use a larger window, which means each event has a smaller contribution to the aggregation state.
+This is because we use a half-life, which means each event has a smaller contribution to the aggregation state or EMA tail.
 Hence, outliers will have less impact on the aggregation and corresponding signal value.
 
-In experiment 03, we increase the sampling and target windows size once again, this time four times larger than in experiment 02.
+In experiment 03, we increase the half-life again, this time four times larger than in experiment 02.
 We would expect an even smoother growth of our signal and, again, no false positives.
@@
 \end{figure}
+However, the execution time was of course much larger.
+We will need to scale our experiments and do our batch analysis on more features and perform more samples, so we opt to use the 62,500 half-life from experiment 02 since it is a good compromise between batch execution time and EMA tail sensitivity to outlier values.
 
@@
 \end{figure}
-
 The target dataset T2 begins with the same distribution as the reference dataset R2, a gaussian distribution with mean $\mu=100$ and standard deviation $\sigma=50$.
@@
 From the 4 millionth event to the end, we use a standard deviation of $\sigma=5$.
-
 In this experiment, we used a half-life $n_{1/2}=62500$ and corresponding sample size of 250,000 events, like in the previous section's experiment 02.
@@
 In this experiment, we wanted to do the reverse of the previous one.
-In this one, there are multiple change points but each new distribution spans a domain that is a superset of the previous one.
+In this experiment, there are multiple change points but each new distribution spans a domain that is a superset of the previous one.
 Figure \ref{fig:timeseries-r3} shows the reference period used, with 5 million events and one feature following a gaussian distribution of mean $\mu=100$ and standard deviation $\sigma=10$.
@@
 \end{figure}
+Once again, we are able to detect changes in distribution parameter changes and quantify how big the changes are (the closer to 1 our signal is).
 
@@
 \end{figure}
-
 The reference dataset R4 had a single feature that followed a Gaussian distribution of $\mu=100$ and $\sigma=5$ for all 5 million events.
@@
 Note in Figure \ref{fig:timeseries-t5} that this last portion resembles the first one, the reference.
-\begin{center}
-\begin{minipage}{.5\textwidth}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-r5.pdf}
-  \captionof{figure}{Time-series for dataset R5}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-r5.pdf}
+      \caption{Time-series for dataset R5}
       \label{fig:timeseries-r5}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/timeseries-t5.pdf}
-  \captionof{figure}{Time-series for dataset T5}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-t5.pdf}
+      \caption{Time-series for dataset T5}
      \label{fig:timeseries-t5}
-\end{minipage}
-\end{center}
+    \end{center}
+\end{figure}
 
@@
 \begin{figure}[!htb]
-\centering
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-test05.pdf}
-  \captionof{figure}{JSD signal and threshold}
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-test05.pdf}
+      \caption{Whole time-series JSD signal and threshold}
       \label{fig:JSD-signal-test05}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[width=1\linewidth]{figures/stream-analysis-viz-zoom-test05.pdf}
-  \caption{Zoom in on the first 2 million events}
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/stream-analysis-viz-zoom-test05.pdf}
+      \caption{Zoom-in on first 2 million events}
     \label{fig:JSD-signal-zoom-test05}
-\end{subfigure}
-\caption{JSD signal and threshold}
+    \end{center}
 \end{figure}
@@
 Figures \ref{fig:timeseries-r6-x1}, \ref{fig:timeseries-r6-x2}, \ref{fig:timeseries-r6-x3} and \ref{fig:timeseries-r6-x4} show the reference time-series for features $x_1$, $x_2$, $x_3$ and $x_4$ on the reference dataset R6, respectively.
-
 \begin{table}[!htb]
@@
 \end{table}
-
-\begin{figure}[!htb] 
-  \begin{minipage}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x1.pdf} 
-    \caption{Feature $x_1$ reference time-series} 
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-r6-x1.pdf} 
+      \caption{$x_1$ reference time-series} 
       \label{fig:timeseries-r6-x1} 
-    \vspace{4ex}
-  \end{minipage}%%
-  \begin{minipage}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x2.pdf} 
-    \caption{Feature $x_2$ reference time-series} 
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-r6-x2.pdf} 
+      \caption{$x_2$ reference time-series} 
       \label{fig:timeseries-r6-x2} 
-    \vspace{4ex}
-  \end{minipage} 
-  \begin{minipage}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x3.pdf} 
-    \caption{Feature $x_3$ reference time-series} 
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-r6-x3.pdf} 
+      \caption{$x_3$ reference time-series} 
       \label{fig:timeseries-r6-x3} 
-    \vspace{4ex}
-  \end{minipage}%% 
-  \begin{minipage}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-r6-x4.pdf} 
-    \caption{Feature $x_4$ reference time-series} 
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-r6-x4.pdf} 
+      \caption{$x_4$ reference time-series} 
       \label{fig:timeseries-r6-x4} 
-    \vspace{4ex}
-  \end{minipage} 
-\end{figure}
-
+    \end{center}
+\end{figure}
 We used a target dataset T6 with 5 million events and the same four features ($x_1$, $x_2$, $x_3$ and $x_4$).
@@
 \end{table}
-
-\begin{figure}[!htb] 
-  \begin{minipage}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x1.pdf} 
-    \caption{Feature $x_1$ target time-series} 
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-t6-x1.pdf} 
+      \caption{$x_1$ target time-series} 
       \label{fig:timeseries-t6-x1} 
-    \vspace{4ex}
-  \end{minipage}%%
-  \begin{minipage}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x2.pdf} 
-    \caption{Feature $x_2$ target time-series} 
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-t6-x2.pdf} 
+      \caption{$x_2$ target time-series} 
       \label{fig:timeseries-t6-x2} 
-    \vspace{4ex}
-  \end{minipage} 
-  \begin{minipage}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x3.pdf} 
-    \caption{Feature $x_3$ target time-series} 
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-t6-x3.pdf} 
+      \caption{$x_3$ target time-series} 
       \label{fig:timeseries-t6-x3} 
-    \vspace{4ex}
-  \end{minipage}%% 
-  \begin{minipage}[b]{0.5\linewidth}
-    \centering
-    \includegraphics[width=1\linewidth]{figures/timeseries-t6-x4.pdf} 
-    \caption{Feature $x_4$ target time-series} 
+    \end{center}
+\end{figure}
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.5]{figures/timeseries-t6-x4.pdf} 
+      \caption{$x_4$ target time-series} 
       \label{fig:timeseries-t6-x4} 
-    \vspace{4ex}
-  \end{minipage} 
+    \end{center}
 \end{figure}
@@
 We used 100 bins for the reference histogram plus the two special infinity bins.
-In this multi-feature scenario, we used the Holm-Bonferroni correction discussed in Section \ref{sec:holmbonferroni} and set the Family-Wise Error Rate (FWER) to 1\% (FWER = 0.01).
-Furthermore, instead of doing 1000 samples as we did previously in an ad-hoc way, we use Equation \ref{eq:optimal-n-samples} to determine the minimum number of samples to make while ensuring $\gamma=1\%$ (recall Section \ref{sec:nsamples}).
+In this multi-feature scenario, we used the Holm-Bonferroni correction discussed in Section \ref{sec:multi-test} and set the Family-Wise Error Rate (FWER) to 1\% (FWER = 0.01).
+Furthermore, instead of doing 1000 samples as we did previously in an ad-hoc way, we use Equation \ref{eq:optimal-n-samples} to determine the minimum number of samples to make while ensuring $\gamma=1\%$ (recall Section \ref{sec:sampling-batch}).
 For our four feature analysis scenario, using as inputs  $\gamma=0.01$ and $FWER=0.01$ in Equation \ref{eq:optimal-n-samples} tells us we need to make 1840 samples minimum.
@@
     \begin{center}
-      \includegraphics[scale=0.8]{figures/x1-corrected-pvalues.png}
+      \includegraphics[scale=0.5]{figures/x1-corrected-pvalues.png}
       \caption{Feature $x_1$ corrected p-values and threshold}
@@
     \begin{center}
-      \includegraphics[scale=0.8]{figures/x2-corrected-pvalues.png}
+      \includegraphics[scale=0.5]{figures/x2-corrected-pvalues.png}
       \caption{Feature $x_2$ corrected p-values and threshold}
@@
     \begin{center}
-      \includegraphics[scale=0.8]{figures/x2-corrected-pvalues-zoom2.png}
+      \includegraphics[scale=0.5]{figures/x2-corrected-pvalues-zoom2.png}
       \caption{Feature $x_2$ zoomed in corrected p-values and threshold}
@@
     \begin{center}
-      \includegraphics[scale=0.8]{figures/x3-corrected-pvalues.png}
+      \includegraphics[scale=0.5]{figures/x3-corrected-pvalues.png}
       \caption{Feature $x_3$ corrected p-values and threshold}
@@
     \begin{center}
-      \includegraphics[scale=0.8]{figures/x4-corrected-pvalues.png}
+      \includegraphics[scale=0.5]{figures/x4-corrected-pvalues.png}
       \caption{Feature $x_4$ corrected p-values and threshold}
@@
     \begin{center}
-      \includegraphics[scale=0.8]{figures/x4-corrected-pvalues-zoom2.png}
+      \includegraphics[scale=0.5]{figures/x4-corrected-pvalues-zoom2.png}
       \caption{Feature $x_4$ zoomed in corrected p-values and threshold}
@@
 
-\clearpage
 \section{Experiments with Real Data}
@@
         \textbf{Dataset} & \textbf{From}              & \textbf{To}           & \multicolumn{1}{l|}{\textbf{Transactions}} \\ \hline
-        Reference        & Tuesday, September 5, 2017 & Friday, June 29, 2018 & 1,604,509                                  \\ \hline
-        Target           & Friday, June 29, 2018      & Sunday, June 30, 2019 & 4,032,505                                  \\ \hline
+        Reference        & September 5, 2017 & June 29, 2018 & 1,604,509                                  \\ \hline
+        Target           & June 29, 2018      & June 30, 2019 & 4,032,505                                  \\ \hline
         \end{tabular}
@@
 Hence, our samples's windows had 151,212 events.
-Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:nsamples}) of 10\% we needed to make a minimum of 42,137 samples.
+Using Equation \ref{eq:optimal-n-samples}, we determined that for 183 features, a FWER of 1\% and $\gamma$ (as defined in Section \ref{sec:sampling-batch}) of 10\% we needed to make a minimum of 42,137 samples.
 Our batch analysis took 5 hours and 34 minutes.
@@
 Our monitoring system processed the dataset event by event in sequential, time-ordered fashion, emulating a stream computation and reported divergent features with associated p-values.
-We performed our multiple hypotheses with the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for every 1000 events.
+We performed our multiple hypotheses with the Holm-Bonferroni correction (Section \ref{sec:multi-test}) for every 1000 events.
 It took us 382,363 milliseconds (roughly 6 minutes) to process the 4,032,505 transactions, amounting to a throughput of 10,546 transactions per second (TPS).
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x1-reference.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x1-reference.pdf}
       \caption{Merchant feature $x_1$ reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x1-target.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x1-target.pdf}
       \caption{Merchant feature $x_1$ target time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x1-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x1-correctedpvalues.pdf}
       \caption{Merchant feature $x_1$ corrected p-values}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x2-reference.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x2-reference.pdf}
       \caption{Merchant feature $x_2$ reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x2-target.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x2-target.pdf}
       \caption{Merchant feature $x_2$ target time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x3-reference.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x3-reference.pdf}
       \caption{Merchant feature $x_3$ reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x3-target.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x3-target.pdf}
       \caption{Merchant feature $x_3$ target time-series}
@@
 \end{figure}
-\clearpage
+
 Relative to $x_1$, features $x_2$ and $x_3$ also have very similar reference and target periods.
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x2-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x2-correctedpvalues.pdf}
       \caption{Merchant feature $x_2$ corrected p-values}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x3-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x3-correctedpvalues.pdf}
       \caption{Merchant feature $x_3$ corrected p-values}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x4-reference.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x4-reference.pdf}
       \caption{Merchant feature $x_4$ reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x4-target.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x4-target.pdf}
       \caption{Merchant feature $x_4$ target time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x4-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x4-correctedpvalues.pdf}
       \caption{Merchant feature $x_4$ corrected p-values}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant2-x2-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant2-x2-correctedpvalues.pdf}
       \caption{Merchant feature $x_2$ corrected p-values with equal reference and target periods}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant2-x3-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant2-x3-correctedpvalues.pdf}
       \caption{Merchant feature $x_3$ corrected p-values with equal reference and target periods}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant2-x4-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant2-x4-correctedpvalues.pdf}
       \caption{Merchant feature $x_4$ corrected p-values with equal reference and target periods}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x5-reference.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x5-reference.pdf}
       \caption{Merchant feature $x_5$ reference time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x5-target.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x5-target.pdf}
       \caption{Merchant feature $x_5$ target time-series}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant-x5-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant-x5-correctedpvalues.pdf}
       \caption{Merchant feature $x_5$ corrected p-values}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant2-x5-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant2-x5-correctedpvalues.pdf}
       \caption{Merchant feature $x_5$ corrected p-values with equal reference and target periods}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant3-x1-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant3-x1-correctedpvalues.pdf}
       \caption{Merchant feature $x_1$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant3-x2-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant3-x2-correctedpvalues.pdf}
       \caption{Merchant feature $x_2$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant3-x3-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant3-x3-correctedpvalues.pdf}
       \caption{Merchant feature $x_3$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant3-x4-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant3-x4-correctedpvalues.pdf}
       \caption{Merchant feature $x_4$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant3-x5-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant3-x5-correctedpvalues.pdf}
       \caption{Merchant feature $x_5$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
@@
 
-\newpage
 \section{Conclusions}

### 2020-07-19 paper.tex
Conclusions. Paper nearly done, only thing missing is related work and cut experiments to fit the 9 page format

@@
 Table \ref{tbl:multi-feat-ref-dataset-distros} shows the underlying generating distributions and parameters for each of the four features ($x_1$, $x_2$, $x_3$ and $x_4$) in the reference multi-feature dataset R6.
-Figures \ref{fig:timeseries-r6-x1}, \ref{fig:timeseries-r6-x2}, \ref{fig:timeseries-r6-x3} and \ref{fig:timeseries-r6-x4} show the reference time-series for features $x_1$, $x_2$, $x_3$ and $x_4$ on the reference dataset R6, respectively.
 \begin{table}[!htb]
@@
 \end{table}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.5]{figures/timeseries-r6-x1.pdf} 
-      \caption{$x_1$ reference time-series} 
-      \label{fig:timeseries-r6-x1} 
-    \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.5]{figures/timeseries-r6-x2.pdf} 
-      \caption{$x_2$ reference time-series} 
-      \label{fig:timeseries-r6-x2} 
-    \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.5]{figures/timeseries-r6-x3.pdf} 
-      \caption{$x_3$ reference time-series} 
-      \label{fig:timeseries-r6-x3} 
-    \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
-      \includegraphics[scale=0.5]{figures/timeseries-r6-x4.pdf} 
-      \caption{$x_4$ reference time-series} 
-      \label{fig:timeseries-r6-x4} 
-    \end{center}
-\end{figure}
 We used a target dataset T6 with 5 million events and the same four features ($x_1$, $x_2$, $x_3$ and $x_4$).
+
 Feature $x_1$ followed the reference distribution for the 5 million events of the target dataset.
 We did not change this feature underlying distribution to serve as a control, \textit{i.e.}, we expect no alerts for this one.
-Figure \ref{fig:timeseries-t6-x1} shows the time-series for the target dataset T6 and feature $x_1$.
 
@@
 Figure \ref{fig:timeseries-t6-x2} shows the time-series for the target dataset T6 and feature $x_2$.
-Feature $x_2$ followed a gaussian with $\mu=100$ and $\sigma=5$ like in the reference period up to the 1 millionth event mark, then a gaussian with $\mu=100$ and $\sigma=10$ up to the 2.5 millionth event, then a gaussian $\mu=100$ and $\sigma=15$ until the 3 millionth event, then a gaussian $\mu=100$ and $\sigma=20$ until the 4 millionth event and finally returned to the original gaussian with $\mu=100$ and $\sigma=5$ until the last event.
-Table \ref{tbl:multi-feat-x2-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_2$.
+Table \ref{tbl:multi-feat-x2-changes} sums up the distributions used and its parameters for each interval.
 \begin{table}[!htb]
@@
 Figure \ref{fig:timeseries-t6-x3} shows the time-series for the target dataset T6 and feature $x_3$.
-Feature $x_3$ followed a uniform distribution with $a=200$ and $b=250$ like in the reference period up to the 1.5 millionth event mark, then a uniform distribution with $a=200$ and $b=220$ up to the 2 millionth event, then a uniform distribution with $a=180$ and $b=200$ until the 3.5 millionth event and finally a gaussian with $\mu=50$ and $\sigma=20$ until the last event.
-Table \ref{tbl:multi-feat-x3-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_3$.
+Table \ref{tbl:multi-feat-x3-changes} sums up the distributions used and its parameters for each interval.
 \begin{table}[!htb]
@@
 Figure \ref{fig:timeseries-t6-x4} shows the time-series for the target dataset T6 and feature $x_4$.
-Feature $x_4$ followed a uniform distribution like the reference one, with $a=50$ and $b=125$ up until the 1.5 millionth event.
-After that, it followed a uniform distribution with $a=125$ and $b=200$ until the 3 millionth event where it resumed its original uniform distribution (with $a=50$ and $b=125$) until the end.
 Table \ref{tbl:multi-feat-x4-changes} sums up the distributions used and its parameters for each interval of the reference dataset for feature $x_4$.
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/timeseries-t6-x1.pdf} 
-      \caption{$x_1$ target time-series} 
-      \label{fig:timeseries-t6-x1} 
-    \end{center}
-\end{figure}
-\begin{figure}[!htb]
-    \begin{center}
       \includegraphics[scale=0.5]{figures/timeseries-t6-x2.pdf} 
@@
 \end{figure}
-
 In this experiment, we used a tuple-based half-life $n_{1/2}=62500$, \textit{i.e.}, sampling and target windows of 250,000 tuples.
@@
 
-In Section \ref{sec:stream-phase}, we mention that, in our stream analysis, we perform our multiple hypothesis divergence tests on a fixed user-defined frequency.
-Finding an optimal frequency value for the test falls out of scope for this thesis, but ideally, we set it to be large enough so that we do fewer computations and achieve higher throughput but still get alerts in useful time.
 In this experiment, we performed the divergence test with the Holm-Bonferroni correction for every new event (frequency of 1 event).
@@
 \begin{equation}
-    p-value_{corrected} = p-value_{distance_{x}} * (m - k)
+    pvalue_{corrected} = pvalue * (m - k)
     \label{for:corrected-pvalue}
@@
 With this list, for each feature, we can plot the corrected p-values on each event and compare it with the FWER threshold $\alpha$.
-Note that despite performing the test at the maximum frequency (in real-time) we still achieved a throughput of 17,921 transactions per second (TPS), even if only for four features.
 
@@
 In Figure \ref{fig:x1-corrected-pvalues} we see the plot of corrected p-values for feature $x_1$ and the probability threshold black horizontal line at the bottom.
-Recall that feature $x_1$'s values were generated by a Gaussian distribution for both the reference period (Figure \ref{fig:timeseries-r6-x1}) and the target period (Figure \ref{fig:timeseries-t6-x1}), with the same parameters.
+Recall that feature $x_1$'s values were generated by a Gaussian distribution for both the reference period and the target period, with the same parameters.
 We observe that the corrected p-values never pass below the threshold, as we expected.
@@
 
-\section{Conclusions}
-In this section, we summarize our experimental tests, results and uncovered issues.
-
+\section{Discussion} \label{sec:Discussion}
 We began by performing single-feature analysis of synthetic data.
@@
 We then moved on to real data.
-Due to time constraints and the scope of this thesis, we experimented only with data from one merchant.
+We experimented only with data from one merchant.
 In our first experiment with real data, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms.
-A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift.
+A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as these datasets contained a lot of drift.
 
@@
 
-\section{Discussion}
-\label{sec:Discussion}
-\textcolor{red}{TODO}
-Fight
-
-\section{Conclusions}
-\label{sec:Conclusions}
-Thesis done, PogU
+\section{Conclusions} \label{sec:Conclusions}
+We summarize our contributions as follows:
+
+\begin{enumerate}
+    \item a two-phased and two-windowed univariate subsequence outlier detection method, classified according to the taxonomy in Section \ref{sec:uni-sub-out} and detailed in Chapter \ref{chap:my-work}, that is lightweight and works for real-time stream monitoring
+
+    \item a histogram aggregation based on Exponential Moving Averages, that replaces the entire sliding window contents by a lighter representation, resulting in a constant in time and space complexity sliding window aggregation
+
+    \item a set of synthetic datasets and experiments where the method in question accurately detected anomalies
+
+    \item experiments on real data that showed unexpected results but where we provide insights and possible hypotheses to test for future work
+\end{enumerate}
+
+\subsection{Future Work}
+
+First and foremost, we believe the next steps are to reason about the obtained results in the real data experiments and find the root cause for the existence of many and large alert periods.
+As a possible hypothesis to test, we suggest testing with EMAs with higher smoothing factors, so that past events are forgotten faster; additionally, we suggest decreasing $\gamma$, which further increases the number of samples to make in the batch phase, improving the accuracy of the divergence measure distribution built.
+
+In this thesis, we used the Jensen–Shannon Divergence (JSD) as a statistical test to measure the distance or divergence between our reference and target sliding window aggregation.
+We chose to use the JSD metric due to its simplicity and because it was used in previous work successfully \cite{SAMM}.
+We believe that future work comprises testing new distances in the literature.
+In particular, we believe the Wasserstein distance is worth trying next, and illustrate why with Figure \ref{fig:wasserteinvsjsd} and Table \ref{tbl:wassertein-vs-jsd}.
+\begin{figure}[!htb]
+    \begin{center}
+      \includegraphics[scale=0.3]{figures/wasserstein-vs-jsd.pdf}
+      \caption{Wassertein vs Jensen–Shannon distances}
+      \label{fig:wasserteinvsjsd}
+    \end{center}
+\end{figure}
+\begin{table}[!htb]
+    \begin{center}
+        \begin{tabular}{|c|c|c|c|}
+        \hline
+        \textbf{Distribution \#1} & \textbf{Distribution \#2} & \textbf{Wasserstein} & \textbf{JSD} \\ \hline
+        a                         & a'                        & 2.0                         & 0.4451                         \\ \hline
+        b                         & b'                        & 0.5                         & 0.4451                         \\ \hline
+        \end{tabular}
+    \end{center}
+    \caption[Wasserstein and Jensen-Shannon divergences output comparison]{Wasserstein and Jensen-Shannon divergences for both distribution pairs}
+    \label{tbl:wassertein-vs-jsd}
+\end{table}
+The Wasserstein distance measures the work required to transport the probability mass from one histogram to another.
+This measure will be larger the further way the probability mass is.
+In Figure \ref{fig:wasserteinvsjsd}, we show two pairs of histograms, \textit{Histogram (a)} paired with \textit{Histogram (a')} and \textit{Histogram (b)} paired with \textit{Histogram (b')}.
+Table \ref{tbl:wassertein-vs-jsd} shows the output values for both pairs by each divergence function.
+The JSD measure is the same for both cases, whereas the Wasserstein distance is larger for the first pair because the probability mass has to be transported further.
+
+In this thesis, we had to deal with the multiple test problem and apply a multiple test correction.
+In our case, we used Holm-Bonferroni due to its simplicity and because it controls the family-wise error rate with larger power than some other methods (thus providing a good trade-off between simplicity and power).
+There are many other multiple test correction methods, and we leave the testing of each as future work.
+
+Besides testing new methods for distribution divergence measurement or multiple test correction methods, we have a set of parameters to work with.
+Experiments where only one of the parameters is changed should be carried out.
+Our method has lots of parameters to fine-tune and experiment with, as defined in Section \ref{sec:batch-phase}, such as the tuple-based half-life, the size of the samples we make, the $\gamma$ probability (which defines the probability of not having at least one sample divergence measure in the upper tail of the distribution) and the Family-Wise Error Rate.
+
+Lastly, we propose another type of validation method.
+We mentioned in the beginning that our objective is to detect data pattern shifts before the performance of the systems that rely on the data declines.
+We propose a test where we train a Machine Learning model on a reference period and measure its prediction accuracy on a target period.
+Then, we propose running our method on reference and target periods to find out timestamps with a high number of divergent features.
+We propose that the model is retrained at these timestamps (tantamount to system reconfiguration) and then used to predict for the remaining target period.
+Finally, we propose that the accuracy should be compared to the previously observed model performance, expecting it to be higher after retraining in one of the high divergent timestamps.
 

### 2020-07-20 paper.tex
Comment out synthetic data experiments
Final paper

@@
 
-\section{Related Work}
-\label{sec:RelatedWork}
+%\section{Related Work} \label{sec:RelatedWork}
 
@@
     \begin{center}
-    %\hspace{6cm}
-      \includegraphics[scale=0.6]{figures/ema-hist.png}
+      \includegraphics[scale=0.6]{figures/ema-hist-paper.png}
       \caption[]{Reference equal-height histogram}
@@
 
+\iffalse
 \section{Experiments with Synthetic Data}
@@
 
-\section{Experiments with Real Data}
-
-Due to time constraints, we experimented with data from only one merchant. The merchant's entity as well as feature names will remain anonymous due to confidentiality reasons.
-
+\fi
+
+\section{Experimental Results}
+We experimented with data from only one merchant and its name as well as feature names will remain anonymous due to confidentiality reasons.
 Feedzai provides a fraud detection solution for this merchant and had stored two contiguous in time datasets of credit card transactions that, combined, covered a period of roughly 1 year and 9 months.
@@
     \begin{center}
-      \includegraphics[scale=0.5]{figures/merchant2-x1-correctedpvalues.pdf}
+      \includegraphics[scale=0.25]{figures/merchant2-x1-correctedpvalues.pdf}
       \caption{Merchant feature $x_1$ corrected p-values with equal reference and target periods}
@@
 \section{Discussion} \label{sec:Discussion}
-We began by performing single-feature analysis of synthetic data.
-The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported.
-Still working with synthetic data, we performed our first multi-feature analysis.
-In this scenario, we applied the Holm-Bonferroni multiple test correction to the p-values of each hypothesis.
-Once again, each feature was correctly reported as divergent at the timestamps where we changed the underlying generating distribution.
-
-We then moved on to real data.
-We experimented only with data from one merchant.
-In our first experiment with real data, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms.
+In our first experiment, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms.
 A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as these datasets contained a lot of drift.

### 2020-07-20 paper.tex
Paper done

@@
 
-In this thesis, we used the Jensen–Shannon Divergence (JSD) as a statistical test to measure the distance or divergence between our reference and target sliding window aggregation.
+In this paper, we used the Jensen–Shannon Divergence (JSD) as a statistical test to measure the distance or divergence between our reference and target sliding window aggregation.
 We chose to use the JSD metric due to its simplicity and because it was used in previous work successfully \cite{SAMM}.
@@
 
-In this thesis, we had to deal with the multiple test problem and apply a multiple test correction.
+In our method, we had to deal with the multiple test problem and apply a multiple test correction.
 In our case, we used Holm-Bonferroni due to its simplicity and because it controls the family-wise error rate with larger power than some other methods (thus providing a good trade-off between simplicity and power).
